---
title: "Simulated data: nested validation design"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

## 1. Purpose

This module defines the nested cross-validation design used by every genomic
prediction model fitted to the simulated population. The partitions are created
**before model fitting** so model development and final evaluation use the same
individuals for GBLUP-ADE and all ANN variants.

The validation design has four fixed features:

- **Repeated outer cross-validation:** 5 repetitions of 5 outer folds produce
  25 outer held-out evaluations.
- **Nested inner cross-validation:** each outer analysis set is divided into 4
  inner folds for ANN hyperparameter selection.
- **Common partitions:** the same outer and inner assignments are reused for all
  six simulation scenarios and all prediction methods.
- **Information separation:** phenotype, true genetic value, scenario, and model
  identity do not participate in fold generation.

Two terms are used throughout the module:

- **Analysis set:** individuals available for model development at the current
  validation level.
- **Assessment set:** individuals held out from that development step and used
  only for evaluation at the same level.

The distinction is especially important in nested validation. The **outer
assessment** individuals remain untouched until final prediction, whereas the
**inner assessment** individuals are used only to compare ANN candidates inside
the corresponding outer analysis set.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)

options(stringsAsFactors = FALSE)

dir.create(
  "output/simulated/m02/figures",
  showWarnings = FALSE,
  recursive = TRUE
)
```

## 2. Packages and fixed validation settings

### 2.1 Packages

Three packages are used, each for a specific reader-facing task:

- **`knitr`** formats the validation settings, partition sizes, and design
  summaries as tables that can be inspected directly in the page.
- **`ggplot2`** constructs the diagram that shows how the 1,000 individuals are
  separated at the outer and inner validation levels.
- **`ggthemes`** applies the same Google Docs theme and color scale adopted in
  the preceding module so the project keeps a consistent visual style.

```{r packages, warning=FALSE, message=FALSE}
library(knitr)     # Format validation summaries as readable HTML tables.
library(ggplot2)   # Draw the nested cross-validation structure.
library(ggthemes)  # Apply the Google Docs theme and matching color scale.
```

### 2.2 Fixed settings

The complete partitioning procedure is controlled by four values. They are set
once here and then reused for every scenario and prediction method:

- **Global seed:** makes all random assignments reproducible.
- **Outer repetitions:** repeats the held-out evaluation five times.
- **Outer folds:** divide the 1,000 individuals into five assessment groups per
  repetition.
- **Inner folds:** divide each 800-individual outer analysis set into four groups
  for ANN tuning.

```{r validation-parameters}
SEED_GLOBAL <- 20260821L
N_OUTER_REPEATS <- 5L
N_OUTER_FOLDS <- 5L
N_INNER_FOLDS <- 4L

validation_parameters <- data.frame(
  Parameter = c(
    "Global seed",
    "Outer repetitions",
    "Outer folds per repetition",
    "Inner folds per outer split"
  ),
  Value = c(
    SEED_GLOBAL,
    N_OUTER_REPEATS,
    N_OUTER_FOLDS,
    N_INNER_FOLDS
  ),
  Role = c(
    "reproducible partition generation",
    "repeat held-out evaluation",
    "define outer assessment sets",
    "compare ANN candidates inside outer analysis"
  )
)

validation_parameters |>
  knitr::kable(
    format = "html",
    caption = "Fixed nested cross-validation settings",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The global seed is combined with repetition and fold identifiers later in the
module. This produces distinct but exactly reproducible assignments without
using any response or model information.

## 3. Individuals used to create the partitions

The shared simulated-data object created in M01 contains the aligned phenotype,
true genetic value, and genotype matrices. Before any fold is generated, the
reader needs to verify which rows define the population available for
partitioning.

The next chunk therefore does three things:

- reads the shared object produced by M01 directly from `output/`;
- compares the dimensions of the three aligned matrices;
- extracts the row identifiers that will be used to construct every fold.

```{r individual-universe}
# Read the shared simulated-data object prepared in M01.
simulated_common <- readRDS(
  "output/simulated/audit/preprocessed_simulated_common.rds"
)

input_components <- data.frame(
  Component = c("phenotype", "true_genetic_value", "genotype_m101"),
  Rows = c(
    nrow(simulated_common$phenotype),
    nrow(simulated_common$true_genetic_value),
    nrow(simulated_common$genotype_m101)
  ),
  Columns = c(
    ncol(simulated_common$phenotype),
    ncol(simulated_common$true_genetic_value),
    ncol(simulated_common$genotype_m101)
  )
)

input_components |>
  knitr::kable(
    format = "html",
    caption = "Aligned objects available before partition generation",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )

# Use only the stored row identifiers to generate the folds.
ids <- as.integer(simulated_common$metadata$row_identifier)

individual_universe <- data.frame(
  Quantity = c(
    "Individuals used to generate folds",
    "First row identifier",
    "Last row identifier"
  ),
  Observed = c(
    length(ids),
    min(ids),
    max(ids)
  )
)

individual_universe |>
  knitr::kable(
    format = "html",
    caption = "Individual universe used for partition generation",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The displayed objects establish the partitioning universe:

- **1,000 rows** are shared by phenotype, true genetic value, and genotype data;
- **row identifiers 1-1,000** are the only individual-level values supplied to
  the partitioning algorithm;
- **phenotype and true genetic value remain outside fold generation**, so the
  partition structure cannot depend on the simulated response or its known
  genetic component.

## 4. Repeated outer cross-validation

The outer level provides the final held-out evaluation. In each repetition, the
1,000 row identifiers are shuffled reproducibly and assigned cyclically to five
fold labels. Because 1,000 is divisible by five, every fold contains exactly 200
individuals.

For a given fold:

- the **200 individuals assigned to that fold** form the outer assessment set;
- the **remaining 800 individuals** form the outer analysis set;
- the two sets are disjoint and together recover all 1,000 individuals.

### 4.1 Outer fold assignments

The code below repeats the same direct assignment procedure five times. The
seed changes with the repetition number, which gives each repetition a distinct
assignment while preserving exact reproducibility.

```{r outer-folds}
outer_assignment_list <- vector(
  "list",
  N_OUTER_REPEATS
)

# Build one balanced five-fold assignment for each repetition.
for (rep_id in seq_len(N_OUTER_REPEATS)) {

  set.seed(SEED_GLOBAL + rep_id)

  shuffled_ids <- sample(
    ids,
    size = length(ids),
    replace = FALSE
  )

  assigned_folds <- rep(
    seq_len(N_OUTER_FOLDS),
    length.out = length(ids)
  )

  assignment <- data.frame(
    ID = shuffled_ids,
    Fold = assigned_folds,
    stringsAsFactors = FALSE
  )

  # Sort only the displayed/stored rows; fold membership is already fixed.
  assignment <- assignment[
    order(assignment$ID),
    ,
    drop = FALSE
  ]
  rownames(assignment) <- NULL

  outer_assignment_list[[rep_id]] <- data.frame(
    Repetition = rep_id,
    ID = assignment$ID,
    OuterFold = assignment$Fold,
    stringsAsFactors = FALSE
  )
}

outer_assignments <- do.call(
  rbind,
  outer_assignment_list
)
rownames(outer_assignments) <- NULL

outer_assignments$SplitID <- sprintf(
  "R%02d_F%02d",
  outer_assignments$Repetition,
  outer_assignments$OuterFold
)

# Count the assessment and analysis individuals in every outer split.
outer_sizes <- aggregate(
  ID ~ Repetition + OuterFold + SplitID,
  data = outer_assignments,
  FUN = length
)
names(outer_sizes)[names(outer_sizes) == "ID"] <- "Assessment_N"
outer_sizes$Analysis_N <- length(ids) - outer_sizes$Assessment_N

outer_sizes |>
  knitr::kable(
    format = "html",
    caption = "Sizes of the 25 outer validation splits",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The outer table should be read as follows:

- **25 rows:** 5 repetitions x 5 outer folds;
- **Assessment_N = 200:** individuals held out for final prediction in that
  split;
- **Analysis_N = 800:** individuals available for all model development within
  that split;
- **one assessment appearance per repetition:** each individual is assessed once
  in every repetition and therefore five times across the complete outer design.

### 4.2 Explicit outer analysis and assessment sets

The fold assignment table identifies membership, but the modeling modules need
explicit ID vectors. The next chunk converts every `SplitID` into a list with
its 800 analysis IDs and 200 assessment IDs.

An example from the first split is displayed so the reader can inspect both
sizes and confirm that the overlap is zero.

```{r outer-split-objects}
outer_split_ids <- sort(
  unique(outer_assignments$SplitID)
)

outer_splits <- setNames(
  vector("list", length(outer_split_ids)),
  outer_split_ids
)

# Store the two ID sets consumed later by GBLUP-ADE and ANN.
for (split_id in outer_split_ids) {

  assessment_ids <- outer_assignments$ID[
    outer_assignments$SplitID == split_id
  ]

  analysis_ids <- setdiff(
    ids,
    assessment_ids
  )

  outer_splits[[split_id]] <- list(
    split_id = split_id,
    analysis_ids = sort(analysis_ids),
    assessment_ids = sort(assessment_ids)
  )
}

example_outer_split <- outer_splits[[1]]

outer_object_summary <- data.frame(
  Component = c(
    "analysis_ids",
    "assessment_ids",
    "overlap"
  ),
  N = c(
    length(example_outer_split$analysis_ids),
    length(example_outer_split$assessment_ids),
    length(
      intersect(
        example_outer_split$analysis_ids,
        example_outer_split$assessment_ids
      )
    )
  ),
  First_IDs = c(
    paste(
      head(example_outer_split$analysis_ids, 6),
      collapse = ", "
    ),
    paste(
      head(example_outer_split$assessment_ids, 6),
      collapse = ", "
    ),
    "-"
  )
)

outer_object_summary |>
  knitr::kable(
    format = "html",
    caption = paste(
      "Structure of outer split",
      example_outer_split$split_id
    ),
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The example makes the outer separation explicit: the 800 analysis IDs and 200
assessment IDs do not overlap, and the same list structure is stored for all 25
outer splits.

## 5. Nested inner cross-validation

The inner level is created **inside each outer analysis set**. It is used for ANN
hyperparameter selection and therefore cannot contain any of the 200 outer
assessment individuals.

For each 800-individual outer analysis set:

- the IDs are shuffled with a seed determined by the outer repetition and fold;
- four balanced inner folds of 200 individuals are created;
- one inner fold becomes the **inner assessment** set;
- the other 600 individuals become the **inner analysis** set;
- ANN early validation is drawn later only from the current inner analysis data,
  not from the inner assessment or outer assessment sets.

### 5.1 Inner fold assignments

The code repeats this procedure for all 25 outer splits, producing 100 inner
splits in total. The outer assessment IDs never enter the inner assignment
operation because the starting vector is always `outer_split$analysis_ids`.

```{r inner-folds}
inner_assignments_list <- vector(
  "list",
  length(outer_splits)
)
names(inner_assignments_list) <- names(outer_splits)

inner_splits <- setNames(
  vector("list", length(outer_splits)),
  names(outer_splits)
)

# Generate four inner folds separately within each outer analysis set.
for (split_id in names(outer_splits)) {

  outer_split <- outer_splits[[split_id]]

  split_tokens <- strsplit(
    sub("^R", "", split_id),
    "_F"
  )[[1]]

  rep_id <- as.integer(split_tokens[1])
  outer_fold <- as.integer(split_tokens[2])

  inner_seed <-
    SEED_GLOBAL +
    100000L +
    rep_id * 100L +
    outer_fold

  set.seed(inner_seed)

  shuffled_ids <- sample(
    outer_split$analysis_ids,
    size = length(outer_split$analysis_ids),
    replace = FALSE
  )

  assigned_folds <- rep(
    seq_len(N_INNER_FOLDS),
    length.out = length(outer_split$analysis_ids)
  )

  assignment <- data.frame(
    ID = shuffled_ids,
    Fold = assigned_folds,
    stringsAsFactors = FALSE
  )

  assignment <- assignment[
    order(assignment$ID),
    ,
    drop = FALSE
  ]
  rownames(assignment) <- NULL

  inner_assignments_list[[split_id]] <- data.frame(
    SplitID = split_id,
    Repetition = rep_id,
    OuterFold = outer_fold,
    ID = assignment$ID,
    InnerFold = assignment$Fold,
    stringsAsFactors = FALSE
  )

  current_inner_splits <- setNames(
    vector("list", N_INNER_FOLDS),
    paste0("I", seq_len(N_INNER_FOLDS))
  )

  # Convert every inner fold into explicit analysis and assessment ID sets.
  for (inner_fold in seq_len(N_INNER_FOLDS)) {

    inner_assessment_ids <- assignment$ID[
      assignment$Fold == inner_fold
    ]

    inner_analysis_ids <- setdiff(
      outer_split$analysis_ids,
      inner_assessment_ids
    )

    current_inner_splits[[paste0("I", inner_fold)]] <- list(
      split_id = split_id,
      inner_fold = inner_fold,
      analysis_ids = sort(inner_analysis_ids),
      assessment_ids = sort(inner_assessment_ids)
    )
  }

  inner_splits[[split_id]] <- current_inner_splits
}

inner_assignments <- do.call(
  rbind,
  inner_assignments_list
)
rownames(inner_assignments) <- NULL

inner_sizes <- aggregate(
  ID ~ SplitID + Repetition + OuterFold + InnerFold,
  data = inner_assignments,
  FUN = length
)
names(inner_sizes)[names(inner_sizes) == "ID"] <- "Assessment_N"
inner_sizes$Analysis_N <- 800L - inner_sizes$Assessment_N
```

### 5.2 Inner split sizes

Displaying all 100 rows would add little information because every split has the
same size. The first 20 are shown as a representative sample of the stored
design.

```{r inner-sizes}
head(inner_sizes, 20) |>
  knitr::kable(
    format = "html",
    caption = "First 20 of the 100 inner validation splits",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The inner structure is constant across all outer splits:

- **Inner analysis:** 600 individuals available for candidate development.
- **Inner assessment:** 200 individuals used to compare ANN candidates.
- **Outer assessment:** the remaining 200 individuals from the original
  population are absent from inner validation and remain reserved for final
  evaluation.

### 5.3 Nested design at a glance

The diagram below places the two levels on the same 1,000-individual scale. It
shows why the outer assessment set cannot leak into ANN tuning:

- the outer level separates **800 analysis** from **200 assessment** individuals;
- the inner level subdivides only the first 800 into **600 inner analysis** and
  **200 inner assessment** individuals;
- the final 200 remain outside the inner process entirely.

```{r nested-validation-figure, fig.cap="Nested cross-validation structure for one outer split. Inner validation is restricted to the 800 outer-analysis individuals; the 200 outer-assessment individuals remain held out."}
validation_flow <- data.frame(
  Level = c(
    2, 2,
    1, 1, 1
  ),
  xmin = c(
    0, 800,
    0, 600, 800
  ),
  xmax = c(
    800, 1000,
    600, 800, 1000
  ),
  Role = c(
    "Analysis set",
    "Assessment set",
    "Analysis set",
    "Assessment set",
    "Held-out outer assessment"
  ),
  Label = c(
    "Outer analysis\n800",
    "Outer assessment\n200",
    "Inner analysis\n600",
    "Inner assessment\n200",
    "Outer assessment\nnot used"
  ),
  stringsAsFactors = FALSE
)

p_validation <- ggplot(validation_flow) +
  geom_rect(
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = Level - 0.30,
      ymax = Level + 0.30,
      fill = Role
    ),
    color = "white",
    linewidth = 0.8
  ) +
  geom_text(
    aes(
      x = (xmin + xmax) / 2,
      y = Level,
      label = Label
    ),
    size = 3.6,
    lineheight = 0.9
  ) +
  scale_fill_gdocs() +
  scale_x_continuous(
    breaks = seq(0, 1000, 200),
    limits = c(0, 1000),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = c(1, 2),
    labels = c("Inner level", "Outer level"),
    limits = c(0.5, 2.5)
  ) +
  labs(
    x = "Individuals represented in one outer split",
    y = NULL,
    fill = NULL
  ) +
  theme_gdocs() +
  theme(
    legend.position = "bottom"
  )

print(p_validation)
```

```{r save-nested-validation-tiff, include=FALSE}
ggsave(
  "output/simulated/m02/figures/nested_cross_validation_design.tiff",
  plot = p_validation,
  device = "tiff",
  width = 10,
  height = 4.5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The diagram represents one outer split, but the same hierarchy is repeated for
all 25 outer splits. This common structure is what allows every prediction
method to be compared under the same held-out evaluations.

## 6. Information excluded from partition generation

The folds are deliberately generated from a minimal set of information. This
prevents the partition structure from adapting to the response, the simulated
truth, or the prediction method.

The table distinguishes what participates in fold generation from what remains
outside it:

```{r partition-information}
partition_information <- data.frame(
  Information = c(
    "Individual row identifier",
    "Deterministic seed",
    "Simulation scenario",
    "Phenotype",
    "True genetic value",
    "Prediction model"
  ),
  Partition_role = c(
    "Used",
    "Used",
    "Not used",
    "Not used",
    "Not used",
    "Not used"
  )
)

partition_information |>
  knitr::kable(
    format = "html",
    caption = "Information used and excluded when generating validation partitions",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

This separation has direct consequences for model comparison:

- **Same across scenarios:** C1-C6 use exactly the same individual partitions.
- **Same across models:** GBLUP-ADE and all ANN variants use the same outer
  assessment individuals.
- **Phenotype not used:** response values do not influence who is assigned to
  a fold.
- **True genetic value not used:** known simulated genetic values are reserved for
  held-out evaluation and never determine partitions or ANN tuning.

## 7. Validation design summary

The complete structure can now be summarized without reference to any fitted
model. These counts define the number and size of comparisons used later by both
modeling branches.

```{r design-summary}
design_summary <- data.frame(
  Item = c(
    "Individuals",
    "Scenarios",
    "Outer repetitions",
    "Outer folds per repetition",
    "Total outer splits",
    "Outer analysis N",
    "Outer assessment N",
    "Inner folds per outer split",
    "Total inner splits",
    "Inner analysis N",
    "Inner assessment N",
    "Global seed"
  ),
  Value = c(
    length(ids),
    simulated_common$metadata$n_scenarios,
    N_OUTER_REPEATS,
    N_OUTER_FOLDS,
    N_OUTER_REPEATS * N_OUTER_FOLDS,
    800,
    200,
    N_INNER_FOLDS,
    N_OUTER_REPEATS * N_OUTER_FOLDS * N_INNER_FOLDS,
    600,
    200,
    SEED_GLOBAL
  )
)

design_summary |>
  knitr::kable(
    format = "html",
    caption = "Nested validation design used for the simulated data",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The fixed design therefore contains:

- **25 outer evaluations** = 5 repetitions x 5 folds;
- **100 inner splits** = 25 outer splits x 4 inner folds;
- **800/200 outer separation** for model development and final evaluation;
- **600/200 inner separation** inside every outer analysis set for ANN tuning.

### 7.1 Observed sizes and overlap

The next summary is calculated from the generated objects rather than from the
settings above. It verifies that the stored partitions have the intended sizes
and that analysis and assessment sets do not overlap.

```{r observed-validation-structure}
# Measure overlap directly from every stored outer split.
outer_overlap_values <- integer(
  length(outer_splits)
)

for (i in seq_along(outer_splits)) {
  outer_overlap_values[i] <- length(
    intersect(
      outer_splits[[i]]$analysis_ids,
      outer_splits[[i]]$assessment_ids
    )
  )
}

outer_overlap <- max(outer_overlap_values)

# Measure overlap directly from all 100 inner splits.
inner_overlap_values <- integer(
  length(outer_splits) * N_INNER_FOLDS
)

position <- 1L

for (split_id in names(inner_splits)) {
  for (inner_id in names(inner_splits[[split_id]])) {

    current_split <- inner_splits[[split_id]][[inner_id]]

    inner_overlap_values[position] <- length(
      intersect(
        current_split$analysis_ids,
        current_split$assessment_ids
      )
    )

    position <- position + 1L
  }
}

inner_overlap <- max(inner_overlap_values)

validation_observed <- data.frame(
  Quantity = c(
    "Outer assignment rows",
    "Outer splits",
    "Outer analysis N",
    "Outer assessment N",
    "Maximum outer overlap",
    "Inner assignment rows",
    "Inner splits",
    "Inner analysis N",
    "Inner assessment N",
    "Maximum inner overlap"
  ),
  Observed = c(
    nrow(outer_assignments),
    length(outer_splits),
    unique(outer_sizes$Analysis_N),
    unique(outer_sizes$Assessment_N),
    outer_overlap,
    nrow(inner_assignments),
    length(outer_splits) * N_INNER_FOLDS,
    unique(inner_sizes$Analysis_N),
    unique(inner_sizes$Assessment_N),
    inner_overlap
  )
)

validation_observed |>
  knitr::kable(
    format = "html",
    caption = "Observed structure of the generated nested validation design",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The observed objects confirm the intended design:

- **5,000 outer assignment rows:** 1,000 individuals x 5 repetitions;
- **25 outer splits:** each with 800 analysis and 200 assessment individuals;
- **20,000 inner assignment rows:** 800 individuals x 25 outer splits;
- **100 inner splits:** each with 600 analysis and 200 assessment individuals;
- **zero overlap:** analysis and assessment sets remain disjoint at both levels.

```{r validation-output-checks, include=FALSE}
# Read the preceding M01 status file only to verify that no blocking issue exists.
required_files <- c(
  "output/simulated/audit/preprocessed_simulated_common.rds",
  "output/simulated/audit/simulated_audit_summary.csv"
)
files_exist <- file.exists(required_files)

m01_check_summary <- read.csv(
  "output/simulated/audit/simulated_audit_summary.csv",
  stringsAsFactors = FALSE
)

n_blocked_audit <- sum(
  m01_check_summary$Status == "BLOCKED",
  na.rm = TRUE
)

required_components <- c(
  "metadata",
  "phenotype",
  "true_genetic_value",
  "genotype_m101"
)

components_ok <- all(
  required_components %in% names(simulated_common)
)

individual_universe_ok <-
  length(ids) == 1000L &&
  !anyDuplicated(ids) &&
  identical(sort(ids), seq_len(1000L)) &&
  nrow(simulated_common$phenotype) == length(ids) &&
  nrow(simulated_common$true_genetic_value) == length(ids) &&
  nrow(simulated_common$genotype_m101) == length(ids)

# Check the complete outer assignment structure.
outer_row_count_ok <-
  nrow(outer_assignments) == length(ids) * N_OUTER_REPEATS

outer_once_per_repeat <- all(
  table(
    outer_assignments$Repetition,
    outer_assignments$ID
  ) == 1L
)

outer_balanced <- all(
  outer_sizes$Assessment_N == 200L &
  outer_sizes$Analysis_N == 800L
)

outer_patterns <- matrix(
  NA_integer_,
  nrow = N_OUTER_REPEATS,
  ncol = length(ids)
)

for (rep_id in seq_len(N_OUTER_REPEATS)) {
  x <- outer_assignments[
    outer_assignments$Repetition == rep_id,
    c("ID", "OuterFold")
  ]
  x <- x[order(x$ID), ]
  outer_patterns[rep_id, ] <- x$OuterFold
}

unique_repetition_patterns <- !anyDuplicated(
  as.data.frame(outer_patterns)
)

assessment_frequency_ok <- all(
  table(outer_assignments$ID) == N_OUTER_REPEATS
)

outer_split_checks <- logical(
  length(outer_splits)
)

for (i in seq_along(outer_splits)) {
  x <- outer_splits[[i]]

  outer_split_checks[i] <-
    length(x$analysis_ids) == 800L &&
    length(x$assessment_ids) == 200L &&
    length(intersect(x$analysis_ids, x$assessment_ids)) == 0L &&
    identical(
      sort(c(x$analysis_ids, x$assessment_ids)),
      sort(ids)
    )
}

outer_split_integrity <- all(outer_split_checks)

# Check the complete inner assignment structure.
inner_row_count_ok <-
  nrow(inner_assignments) == length(outer_splits) * 800L

inner_balanced <- all(
  inner_sizes$Assessment_N == 200L &
  inner_sizes$Analysis_N == 600L
)

inner_assignment_checks <- logical(
  length(outer_splits)
)

for (i in seq_along(outer_splits)) {

  split_id <- names(outer_splits)[i]

  x <- inner_assignments[
    inner_assignments$SplitID == split_id,
    ,
    drop = FALSE
  ]

  outer_analysis_ids <- outer_splits[[split_id]]$analysis_ids

  inner_assignment_checks[i] <-
    nrow(x) == 800L &&
    !anyDuplicated(x$ID) &&
    identical(sort(x$ID), sort(outer_analysis_ids)) &&
    identical(
      sort(unique(x$InnerFold)),
      seq_len(N_INNER_FOLDS)
    )
}

inner_assignment_integrity <- all(
  inner_assignment_checks
)

inner_split_checks <- logical(
  length(outer_splits) * N_INNER_FOLDS
)
position <- 1L

for (split_id in names(inner_splits)) {

  outer_analysis_ids <- outer_splits[[split_id]]$analysis_ids

  for (inner_id in names(inner_splits[[split_id]])) {

    x <- inner_splits[[split_id]][[inner_id]]

    inner_split_checks[position] <-
      length(x$analysis_ids) == 600L &&
      length(x$assessment_ids) == 200L &&
      length(intersect(x$analysis_ids, x$assessment_ids)) == 0L &&
      identical(
        sort(c(x$analysis_ids, x$assessment_ids)),
        sort(outer_analysis_ids)
      )

    position <- position + 1L
  }
}

inner_split_integrity <- all(
  inner_split_checks
)

split_variables <- unique(
  c(
    names(outer_assignments),
    names(inner_assignments)
  )
)

forbidden_split_fields <- c(
  "Scenario",
  "Phenotype",
  "Observed",
  "TrueGeneticValue",
  "Model"
)

split_independence_ok <- !any(
  forbidden_split_fields %in% split_variables
)

# Preserve the established output log and its historical field labels.
validation_log <- data.frame(
  Check = c(
    "Canonical input files",
    "Simulated-data audit prerequisite",
    "Canonical object components",
    "Individual universe",
    "Outer assignment row count",
    "One outer assignment per ID and repetition",
    "Balanced outer folds",
    "Distinct outer repetitions",
    "Outer assessment frequency",
    "Outer split integrity",
    "Inner assignment row count",
    "Balanced inner folds",
    "Inner assignment integrity",
    "Inner split integrity",
    "Scenario-independent partitions",
    "True genetic value excluded from partitioning",
    "Phenotype excluded from partitioning"
  ),
  Status = c(
    ifelse(all(files_exist), "PASS", "BLOCKED"),
    ifelse(n_blocked_audit == 0L, "PASS", "BLOCKED"),
    ifelse(components_ok, "PASS", "BLOCKED"),
    ifelse(individual_universe_ok, "PASS", "BLOCKED"),
    ifelse(outer_row_count_ok, "PASS", "BLOCKED"),
    ifelse(outer_once_per_repeat, "PASS", "BLOCKED"),
    ifelse(outer_balanced, "PASS", "BLOCKED"),
    ifelse(unique_repetition_patterns, "PASS", "WARNING"),
    ifelse(assessment_frequency_ok, "PASS", "BLOCKED"),
    ifelse(outer_split_integrity, "PASS", "BLOCKED"),
    ifelse(inner_row_count_ok, "PASS", "BLOCKED"),
    ifelse(inner_balanced, "PASS", "BLOCKED"),
    ifelse(inner_assignment_integrity, "PASS", "BLOCKED"),
    ifelse(inner_split_integrity, "PASS", "BLOCKED"),
    ifelse(split_independence_ok, "PASS", "BLOCKED"),
    "PASS",
    "PASS"
  ),
  Details = c(
    if (all(files_exist)) {
      "Canonical simulated object and audit summary are present."
    } else {
      paste(
        "Missing:",
        paste(required_files[!files_exist], collapse = ", ")
      )
    },
    paste(
      n_blocked_audit,
      "blocking check(s) in simulated-data audit."
    ),
    if (components_ok) {
      paste(
        "Required components present:",
        paste(required_components, collapse = ", ")
      )
    } else {
      paste(
        "Missing components:",
        paste(
          setdiff(required_components, names(simulated_common)),
          collapse = ", "
        )
      )
    },
    "Expected 1000 unique row identifiers aligned with phenotype, true genetic value, and genotype matrices.",
    paste(
      "Observed rows:",
      nrow(outer_assignments),
      "; expected:",
      length(ids) * N_OUTER_REPEATS
    ),
    "Each individual must occur exactly once in every repetition.",
    "Every outer split must contain 800 analysis and 200 assessment individuals.",
    paste(
      "All",
      N_OUTER_REPEATS,
      "repetitions should use distinct fold assignments."
    ),
    paste(
      "Every individual is assigned to one assessment fold per repetition and therefore appears in exactly",
      N_OUTER_REPEATS,
      "outer assessment sets."
    ),
    paste(
      length(outer_splits),
      "outer split objects checked for size, disjointness, and complete coverage."
    ),
    paste(
      "Observed rows:",
      nrow(inner_assignments),
      "; expected:",
      length(outer_splits) * 800L
    ),
    "Every inner split must contain 600 analysis and 200 assessment individuals.",
    "Inner assignments are restricted to the corresponding outer analysis set.",
    paste(
      length(outer_splits) * N_INNER_FOLDS,
      "inner splits checked for size, disjointness, and complete outer-analysis coverage."
    ),
    "Partitions are generated from canonical individual IDs and deterministic seeds only; the same folds are reused for C1-C6 and all models.",
    "True simulated genetic values are stored in the canonical data object for final evaluation only and are not referenced when generating folds.",
    "Phenotypes are not used to construct folds. They may be used later inside training data and inner-CV tuning, never to redefine partitions."
  ),
  stringsAsFactors = FALSE
)

validation_log$Status <- factor(
  validation_log$Status,
  levels = c(
    "BLOCKED",
    "WARNING",
    "PASS"
  )
)

validation_log <- validation_log[
  order(validation_log$Status),
  ,
  drop = FALSE
]
rownames(validation_log) <- NULL
```

## 8. Saved validation object

The modeling modules do not regenerate these partitions. Instead, the complete
design is saved once and then read directly by GBLUP-ADE and ANN.

The saved files retain:

- the 5,000 outer assignment rows;
- the 20,000 inner assignment rows;
- the explicit analysis and assessment ID vectors for every split;
- the fixed validation settings and the internal consistency checks.

The output-writing chunk is hidden because it records already established
objects rather than introducing a new analytical step.

```{r save-validation-design, include=FALSE}
dir.create(
  "output/simulated/validation",
  showWarnings = FALSE,
  recursive = TRUE
)

write.csv(
  validation_log,
  "output/simulated/validation/simulated_cv_design_audit.csv",
  row.names = FALSE
)

write.csv(
  outer_assignments,
  "output/simulated/validation/simulated_cv_outer_assignments.csv",
  row.names = FALSE
)

write.csv(
  inner_assignments,
  "output/simulated/validation/simulated_cv_inner_assignments.csv",
  row.names = FALSE
)

write.csv(
  design_summary,
  "output/simulated/validation/simulated_cv_design_summary.csv",
  row.names = FALSE
)

# Preserve the established object structure consumed by M03 and M04.
cv_design <- list(
  metadata = list(
    dataset = "simulated",
    seed_global = SEED_GLOBAL,
    n_individuals = length(ids),
    n_scenarios = simulated_common$metadata$n_scenarios,
    outer_repetitions = N_OUTER_REPEATS,
    outer_folds = N_OUTER_FOLDS,
    inner_folds = N_INNER_FOLDS,
    outer_analysis_n = 800L,
    outer_assessment_n = 200L,
    inner_analysis_n = 600L,
    inner_assessment_n = 200L,
    scenario_independent = TRUE,
    model_independent = TRUE,
    phenotype_used_for_partitioning = FALSE,
    true_genetic_value_used_for_partitioning = FALSE
  ),
  outer_assignments = outer_assignments,
  outer_splits = outer_splits,
  inner_assignments = inner_assignments,
  inner_splits = inner_splits,
  audit = validation_log
)

saveRDS(
  cv_design,
  "output/simulated/validation/simulated_cv_design.rds",
  compress = "gzip"
)
```

The resulting `simulated_cv_design.rds` is the single partition object consumed
by the prediction modules. Keeping this object fixed ensures that differences
between GBLUP-ADE and ANN results reflect the models rather than different
validation samples.

## 9. Navigation

[Previous: Simulated data preparation](simulated_data_audit.html) | [GBLUP-ADE](simulated_data_gblup.html) | [Regularized ANN](simulated_data_ann.html)
