---
title: "GBLUP-ADE for simulated genomic data"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Purpose

This module fits the canonical GBLUP-ADE benchmark to all six simulated
scenarios using the 25 frozen outer validation splits. Phenotypes are used for
model fitting. True simulated genetic values are used only after prediction
for primary evaluation of genetic-value recovery.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)

options(stringsAsFactors = FALSE)

WIDE_TABLE_ATTR <- paste(
  'style="display:block;',
  'overflow-x:auto;',
  'white-space:nowrap;',
  'width:100%;"'
)

find_project_root <- function(start = getwd()) {
  path <- normalizePath(start, winslash = "/", mustWork = TRUE)

  repeat {
    if (file.exists(file.path(path, "_workflowr.yml"))) {
      return(path)
    }

    parent <- dirname(path)

    if (identical(parent, path)) {
      stop("Could not locate workflowr project root.")
    }

    path <- parent
  }
}

PROJECT_ROOT <- find_project_root()

project_path <- function(...) {
  file.path(PROJECT_ROOT, ...)
}


SEED_GLOBAL <- 20260821L
N_ITERS_SOMMER <- 20L
RUN_GBLUP <- identical(Sys.getenv("RUN_GBLUP", "0"), "1")

DATA_FILE <- project_path("output", "simulated", "audit", "preprocessed_simulated_common.rds")
CV_FILE <- project_path("output", "simulated", "validation", "simulated_cv_design.rds")
DATA_AUDIT_FILE <- project_path("output", "simulated", "audit", "simulated_audit_summary.csv")
CV_AUDIT_FILE <- project_path("output", "simulated", "validation", "simulated_cv_design_audit.csv")

RESULTS_RDS <- project_path("output", "simulated", "gblup", "simulated_gblup_ade_results.rds")
PREDICTIONS_RDS <- project_path("output", "simulated", "gblup", "simulated_gblup_ade_predictions.rds")
RESULTS_CSV <- project_path("output", "simulated", "gblup", "simulated_gblup_ade_results.csv")
PREDICTIONS_CSV <- project_path("output", "simulated", "gblup", "simulated_gblup_ade_predictions.csv")
KERNEL_CSV <- project_path("output", "simulated", "gblup", "simulated_gblup_ade_kernel_summary.csv")
RUN_RDS <- project_path("output", "simulated", "gblup", "simulated_gblup_ade_run_metadata.rds")

safe_cor <- function(pred, obs) {
  ok <- is.finite(pred) & is.finite(obs)
  if (sum(ok) < 3L || sd(pred[ok]) == 0 || sd(obs[ok]) == 0) return(NA_real_)
  cor(pred[ok], obs[ok])
}

safe_rmse <- function(pred, obs) {
  ok <- is.finite(pred) & is.finite(obs)
  if (!any(ok)) return(NA_real_)
  sqrt(mean((pred[ok] - obs[ok])^2))
}

safe_mae <- function(pred, obs) {
  ok <- is.finite(pred) & is.finite(obs)
  if (!any(ok)) return(NA_real_)
  mean(abs(pred[ok] - obs[ok]))
}

safe_bias <- function(pred, obs) {
  ok <- is.finite(pred) & is.finite(obs)
  if (!any(ok)) return(NA_real_)
  mean(pred[ok] - obs[ok])
}

safe_slope <- function(pred, obs) {
  ok <- is.finite(pred) & is.finite(obs)
  if (sum(ok) < 3L || sd(pred[ok]) == 0) return(NA_real_)
  as.numeric(coef(lm(obs[ok] ~ pred[ok]))[2])
}

safe_r2 <- function(pred, obs) {
  ok <- is.finite(pred) & is.finite(obs)
  if (sum(ok) < 2L) return(NA_real_)
  den <- sum((obs[ok] - mean(obs[ok]))^2)
  if (den == 0) return(NA_real_)
  1 - sum((obs[ok] - pred[ok])^2) / den
}

metrics <- function(pred, obs, prefix) {
  z <- c(
    safe_cor(pred, obs),
    safe_rmse(pred, obs),
    safe_mae(pred, obs),
    safe_bias(pred, obs),
    safe_slope(pred, obs),
    safe_r2(pred, obs)
  )
  names(z) <- paste0(
    prefix,
    c("_Correlation", "_RMSE", "_MAE", "_Bias", "_Slope", "_R2")
  )
  z
}

extract_u <- function(fit, dat, factor_name, exact_name) {
  unames <- names(fit$uList)
  if (exact_name %in% unames) {
    nm <- exact_name
  } else {
    token <- paste0("ism(", factor_name, ")")
    hit <- grep(token, unames, fixed = TRUE)
    if (length(hit) != 1L) {
      stop(
        paste(
          "Could not identify random component for", factor_name,
          ". Available:", paste(unames, collapse = " | ")
        )
      )
    }
    nm <- unames[hit]
  }

  u <- fit$uList[[nm]]
  if (is.data.frame(u)) u <- as.matrix(u)
  if (is.vector(u)) u <- matrix(u, ncol = 1L)

  X <- model.matrix(
    as.formula(paste0("~", factor_name, " - 1")),
    data = dat
  )

  if (ncol(X) != nrow(u)) {
    stop(
      paste(
        "Random-effect dimension mismatch for", factor_name,
        ":", ncol(X), "vs", nrow(u)
      )
    )
  }

  as.numeric(X %*% u[, 1L, drop = FALSE])
}
```

# Canonical inputs

The GBLUP-ADE branch uses the audited simulated-data object and the frozen
validation design created in the preceding modules. The canonical dataset
contains 1,000 individuals, 4,010 markers, six simulation scenarios, and 25
outer validation splits. These inputs are checked before any model result is
used.

```{r inputs, include=FALSE}
if (RUN_GBLUP) {
  if (!requireNamespace("sommer", quietly = TRUE)) {
    stop("Package 'sommer' is required when RUN_GBLUP=1.")
  }

  # sommer::mmes() evaluates the random-effects formula using helper
  # functions vsm() and ism(). Attach sommer explicitly so these helpers
  # are available in the fresh R session created by workflowr/callr.
  suppressPackageStartupMessages(
    library(sommer)
  )

  if (
    !exists("vsm", mode = "function") ||
    !exists("ism", mode = "function")
  ) {
    stop(
      paste(
        "sommer was attached, but vsm() and/or ism() are unavailable.",
        "Installed sommer version:",
        as.character(packageVersion("sommer"))
      )
    )
  }
}

required_files <- c(
  DATA_FILE,
  CV_FILE,
  DATA_AUDIT_FILE,
  CV_AUDIT_FILE
)

if (!all(file.exists(required_files))) {
  stop(
    paste(
      "Missing canonical file(s):",
      paste(required_files[!file.exists(required_files)], collapse = ", ")
    )
  )
}

data_audit <- read.csv(DATA_AUDIT_FILE, stringsAsFactors = FALSE)
cv_audit <- read.csv(CV_AUDIT_FILE, stringsAsFactors = FALSE)

if (any(data_audit$Status == "BLOCKED")) {
  stop("Simulated-data audit contains blocking checks.")
}

if (any(cv_audit$Status == "BLOCKED")) {
  stop("Validation-design audit contains blocking checks.")
}

common <- readRDS(DATA_FILE)
cv <- readRDS(CV_FILE)

phenotype <- common$phenotype
true_g <- common$true_genetic_value
geno <- common$genotype_m101
control <- common$scenario_control
ids <- as.integer(common$metadata$row_identifier)

if (is.null(ids) || length(ids) == 0L) {
  ids <- seq_len(nrow(geno))
}

id_labels <- sprintf("ID%04d", ids)
rownames(geno) <- id_labels

scenario_names <- colnames(phenotype)
if (is.null(scenario_names)) scenario_names <- paste0("C", 1:6)

input_summary <- data.frame(
  Item = c("Individuals", "Markers", "Scenarios", "Outer splits"),
  Value = c(nrow(geno), ncol(geno), ncol(phenotype), length(cv$outer_splits))
)

knitr::kable(input_summary)
```

# GBLUP-ADE model

For each scenario and frozen outer split, the phenotype is modeled as

\[
\mathbf{y} = \mathbf{1}\mu + \mathbf{a} + \mathbf{d} + \mathbf{e} + \boldsymbol{\varepsilon},
\]

where \(\mu\) is the intercept, \(\mathbf{a}\) is the additive genomic effect,
\(\mathbf{d}\) is the dominance effect, \(\mathbf{e}\) is the epistatic effect,
and \(\boldsymbol{\varepsilon}\) is the residual.

The corresponding covariance structures are

\[
\mathbf{a} \sim N(0,\mathbf{A}\sigma_a^2),\quad
\mathbf{d} \sim N(0,\mathbf{D}\sigma_d^2),
\]

\[
\mathbf{e} \sim N(0,\mathbf{E}\sigma_e^2),\quad
\boldsymbol{\varepsilon} \sim N(0,\mathbf{I}\sigma_{\varepsilon}^2).
\]

At the scientific level, \(\mathbf{A}\), \(\mathbf{D}\), and \(\mathbf{E}\)
are three genomic relationship structures built from the marker data. They
specify how covariance is shared among individuals for each random genetic
component of the same phenotype:

- \(\mathbf{A}\) represents additive genomic relationship and structures the
  covariance of the additive effects \(\mathbf{a}\);
- \(\mathbf{D}\) represents dominance genomic relationship and structures the
  covariance of the dominance effects \(\mathbf{d}\);
- \(\mathbf{E}\) represents the epistatic genomic relationship used by this
  benchmark and structures the covariance of the epistatic effects
  \(\mathbf{e}\).

These matrices are covariance structures, not additional response variables.
The model combines the three predicted genetic components to represent the
total predicted genetic value for each individual.

The genomic relationship matrices \(\mathbf{A}\), \(\mathbf{D}\), and
\(\mathbf{E}\) are constructed with `sommer::A.mat()`, `sommer::D.mat()`,
and `sommer::E.mat()`, respectively. In `sommer::mmes()`, `ism()` identifies
the individual levels associated with each random effect and `vsm()` links that
effect to its genomic covariance matrix through `Gu`.

The canonical fitting protocol uses `nIters = 20`. This value is fixed across
all scenarios and outer splits and is not a tuned model-selection parameter.

The total predicted genetic component used for evaluation is the sum of the
additive, dominance, and epistatic predictions. The model intercept is added
when predicting the phenotype.

# Pre-fit integrity

Before any saved result is summarized, the canonical phenotype, true-genetic-
value, genotype, scenario, and frozen outer-split dimensions are checked
internally. A blocking mismatch stops the module rather than allowing a result
table to be interpreted against an inconsistent input contract.

```{r prefit, include=FALSE}
checks <- data.frame(
  Check = c(
    "Phenotype dimensions",
    "True-G dimensions",
    "Genotype dimensions",
    "Genotype coding",
    "Outer split count",
    "Scenario names"
  ),
  PASS = c(
    identical(dim(phenotype), c(1000L, 6L)),
    identical(dim(true_g), c(1000L, 6L)),
    identical(dim(geno), c(1000L, 4010L)),
    identical(sort(unique(as.vector(geno))), -1:1),
    length(cv$outer_splits) == 25L,
    identical(scenario_names, paste0("C", 1:6))
  ),
  stringsAsFactors = FALSE
)

checks$Status <- ifelse(checks$PASS, "PASS", "BLOCKED")

knitr::kable(checks[, c("Check", "Status")])

if (any(!checks$PASS)) {
  stop("Blocking GBLUP-ADE pre-fit check detected.")
}

outer_ok <- all(
  vapply(
    cv$outer_splits,
    function(s) {
      length(s$analysis_ids) == 800L &&
        length(s$assessment_ids) == 200L &&
        length(intersect(s$analysis_ids, s$assessment_ids)) == 0L
    },
    logical(1)
  )
)

if (!outer_ok) stop("Frozen outer splits failed integrity check.")
```

# Analysis tasks

The six scenarios are crossed with the 25 frozen outer splits to define the
complete GBLUP-ADE workload. Each row below therefore identifies one independent
scenario-by-split prediction task; the full design contains 150 tasks.

```{r task-table}
outer_meta <- unique(
  cv$outer_assignments[, c("Repetition", "OuterFold", "SplitID")]
)

outer_meta <- outer_meta[
  order(outer_meta$Repetition, outer_meta$OuterFold),
]

task_table <- merge(
  data.frame(Scenario = scenario_names, stringsAsFactors = FALSE),
  outer_meta,
  by = NULL
)

task_table <- task_table[
  order(
    match(task_table$Scenario, scenario_names),
    task_table$Repetition,
    task_table$OuterFold
  ),
]

task_table$TaskID <- paste(
  task_table$Scenario,
  task_table$SplitID,
  sep = "_"
)

rownames(task_table) <- NULL

cat("Expected GBLUP-ADE tasks:", nrow(task_table), "\n")
knitr::kable(head(task_table, 15))

if (nrow(task_table) != 150L) {
  stop("Expected 150 scenario-by-split tasks.")
}
```

# Fit GBLUP-ADE

The A, D, and E relationship matrices are built once from the complete
genotype matrix and reused across tasks. All genotype profiles may contribute
to these relationship matrices because genomic information is available at
prediction time.

For each outer split, however, the phenotypes of the 200 assessment individuals
are replaced by `NA` before fitting. Consequently, those individuals contribute
genomic relationships but not response information. Their true genetic values
are also excluded from fitting and are accessed only after prediction for
performance evaluation.

```{r fit-gblup-ade, eval=RUN_GBLUP}
set.seed(SEED_GLOBAL)

run_start <- proc.time()[["elapsed"]]

kernel_start <- proc.time()[["elapsed"]]

A_matrix <- sommer::A.mat(geno)
D_matrix <- sommer::D.mat(geno)
E_matrix <- sommer::E.mat(geno)

kernel_seconds <- proc.time()[["elapsed"]] - kernel_start

rownames(A_matrix) <- colnames(A_matrix) <- id_labels
rownames(D_matrix) <- colnames(D_matrix) <- id_labels
rownames(E_matrix) <- colnames(E_matrix) <- id_labels

kernel_summary <- data.frame(
  Kernel = c("Additive", "Dominance", "Epistatic"),
  Rows = c(nrow(A_matrix), nrow(D_matrix), nrow(E_matrix)),
  Columns = c(ncol(A_matrix), ncol(D_matrix), ncol(E_matrix)),
  Symmetric = c(
    isTRUE(all.equal(A_matrix, t(A_matrix), tolerance = 1e-8)),
    isTRUE(all.equal(D_matrix, t(D_matrix), tolerance = 1e-8)),
    isTRUE(all.equal(E_matrix, t(E_matrix), tolerance = 1e-8))
  ),
  Finite = c(
    all(is.finite(A_matrix)),
    all(is.finite(D_matrix)),
    all(is.finite(E_matrix))
  ),
  Mean_diagonal = c(
    mean(diag(A_matrix)),
    mean(diag(D_matrix)),
    mean(diag(E_matrix))
  ),
  Kernel_build_seconds = rep(kernel_seconds, 3L)
)

if (any(!kernel_summary$Symmetric) || any(!kernel_summary$Finite)) {
  stop("A, D, or E kernel failed integrity checks.")
}

write.csv(kernel_summary, KERNEL_CSV, row.names = FALSE)

result_list <- vector("list", nrow(task_table))
prediction_list <- vector("list", nrow(task_table))

for (i in seq_len(nrow(task_table))) {
  task <- task_table[i, ]
  scenario <- task$Scenario
  split_id <- task$SplitID
  j <- match(scenario, scenario_names)

  split <- cv$outer_splits[[split_id]]

  tr_idx <- match(split$analysis_ids, ids)
  te_idx <- match(split$assessment_ids, ids)

  if (anyNA(tr_idx) || anyNA(te_idx)) {
    stop(paste("ID alignment failed for", task$TaskID))
  }

  y <- phenotype[, j]
  g <- true_g[, j]

  id_factor <- factor(id_labels, levels = id_labels)

  dat <- data.frame(
    y = y,
    id = id_factor,
    idd = id_factor,
    ide = id_factor
  )

  dat$y[te_idx] <- NA_real_

  fit_status <- "ok"
  pred_status <- "not_run"
  error_message <- NA_character_

  y_pred <- gebv <- u_add <- u_dom <- u_epi <-
    rep(NA_real_, nrow(dat))

  fit_start <- proc.time()[["elapsed"]]

  fit <- tryCatch(
    sommer::mmes(
      y ~ 1,
      random = ~
        vsm(ism(id), Gu = A_matrix) +
        vsm(ism(idd), Gu = D_matrix) +
        vsm(ism(ide), Gu = E_matrix),
      rcov = ~ units,
      nIters = N_ITERS_SOMMER,
      data = dat,
      verbose = FALSE,
      dateWarning = FALSE
    ),
    error = function(e) {
      fit_status <<- "fit_error"
      error_message <<- conditionMessage(e)
      NULL
    }
  )

  fit_seconds <- proc.time()[["elapsed"]] - fit_start
  predict_seconds <- 0

  if (!is.null(fit)) {
    pred_status <- "ok"
    predict_start <- proc.time()[["elapsed"]]

    pc <- tryCatch(
      {
        a <- extract_u(
          fit, dat, "id",
          "vsm(ism(id), Gu = A_matrix)"
        )
        d <- extract_u(
          fit, dat, "idd",
          "vsm(ism(idd), Gu = D_matrix)"
        )
        e <- extract_u(
          fit, dat, "ide",
          "vsm(ism(ide), Gu = E_matrix)"
        )

        gv <- a + d + e
        beta0 <- as.numeric(fit$b)[1L]

        list(
          additive = a,
          dominance = d,
          epistatic = e,
          gebv = gv,
          prediction = beta0 + gv
        )
      },
      error = function(e) {
        pred_status <<- "pred_error"
        error_message <<- conditionMessage(e)
        NULL
      }
    )

    predict_seconds <- proc.time()[["elapsed"]] - predict_start

    if (!is.null(pc)) {
      u_add <- pc$additive
      u_dom <- pc$dominance
      u_epi <- pc$epistatic
      gebv <- pc$gebv
      y_pred <- pc$prediction
    }
  }

  obs_test <- y[te_idx]
  g_test <- g[te_idx]
  pred_test <- y_pred[te_idx]
  gebv_test <- gebv[te_idx]

  mg <- metrics(pred_test, g_test, "Genetic")
  mp <- metrics(pred_test, obs_test, "Phenotype")

  n_finite <- sum(is.finite(pred_test))

  status <- if (
    fit_status == "ok" &&
      pred_status == "ok" &&
      n_finite == 200L
  ) {
    "ok"
  } else {
    paste(fit_status, pred_status, sep = ";")
  }

  meta <- control[control$Scenario == scenario, , drop = FALSE]

  result_list[[i]] <- data.frame(
    Dataset = "simulated",
    Scenario = scenario,
    QTL_N = meta$QTL_N,
    H2_nominal = meta$H2_nominal,
    Repetition = task$Repetition,
    Fold = task$OuterFold,
    SplitID = split_id,
    TaskID = task$TaskID,
    Model = "GBLUP_ADE",
    N_train = length(tr_idx),
    N_test = length(te_idx),
    N_pred_finite = n_finite,
    Genetic_Correlation = mg["Genetic_Correlation"],
    Genetic_RMSE = mg["Genetic_RMSE"],
    Genetic_MAE = mg["Genetic_MAE"],
    Genetic_Bias = mg["Genetic_Bias"],
    Genetic_Slope = mg["Genetic_Slope"],
    Genetic_R2 = mg["Genetic_R2"],
    GEBV_TrueG_Correlation = safe_cor(gebv_test, g_test),
    Phenotype_Correlation = mp["Phenotype_Correlation"],
    Phenotype_RMSE = mp["Phenotype_RMSE"],
    Phenotype_MAE = mp["Phenotype_MAE"],
    Phenotype_Bias = mp["Phenotype_Bias"],
    Phenotype_Slope = mp["Phenotype_Slope"],
    Phenotype_R2 = mp["Phenotype_R2"],
    Kernel_build_seconds = kernel_seconds,
    Fit_seconds = fit_seconds,
    Predict_seconds = predict_seconds,
    Status = status,
    Error = error_message,
    stringsAsFactors = FALSE
  )

  prediction_list[[i]] <- data.frame(
    Dataset = "simulated",
    Scenario = scenario,
    QTL_N = meta$QTL_N,
    H2_nominal = meta$H2_nominal,
    Repetition = task$Repetition,
    Fold = task$OuterFold,
    SplitID = split_id,
    TaskID = task$TaskID,
    Model = "GBLUP_ADE",
    ID = ids[te_idx],
    Observed_Phenotype = obs_test,
    True_Genetic_Value = g_test,
    Predicted_Value = pred_test,
    GEBV = gebv_test,
    Additive_Prediction = u_add[te_idx],
    Dominance_Prediction = u_dom[te_idx],
    Epistatic_Prediction = u_epi[te_idx],
    Status = status,
    stringsAsFactors = FALSE
  )

  cat(
    sprintf(
      "[%03d/%03d] %s | %s | rG=%.4f | RMSE_G=%.4f | fit=%.2fs\n",
      i,
      nrow(task_table),
      task$TaskID,
      status,
      result_list[[i]]$Genetic_Correlation,
      result_list[[i]]$Genetic_RMSE,
      fit_seconds
    )
  )
}

results <- do.call(rbind, result_list)
predictions <- do.call(rbind, prediction_list)

rownames(results) <- NULL
rownames(predictions) <- NULL

run_seconds <- proc.time()[["elapsed"]] - run_start

run_metadata <- list(
  dataset = "simulated",
  model = "GBLUP_ADE",
  seed_global = SEED_GLOBAL,
  sommer_version = as.character(packageVersion("sommer")),
  nIters_sommer = N_ITERS_SOMMER,
  tasks = nrow(results),
  prediction_rows = nrow(predictions),
  kernel_build_seconds = kernel_seconds,
  total_run_seconds = run_seconds,
  timestamp = as.character(Sys.time())
)

saveRDS(results, RESULTS_RDS, compress = "gzip")
saveRDS(predictions, PREDICTIONS_RDS, compress = "gzip")
write.csv(results, RESULTS_CSV, row.names = FALSE)
write.csv(predictions, PREDICTIONS_CSV, row.names = FALSE)
saveRDS(run_metadata, RUN_RDS, compress = "gzip")

cat(
  "\nGBLUP-ADE completed:",
  nrow(results), "tasks;",
  nrow(predictions), "held-out predictions;",
  round(run_seconds, 2), "seconds.\n"
)
```

# Kernel diagnostics

Numerical integrity of the additive, dominance, and epistatic relationship
matrices is checked before saved predictions are interpreted. The detailed
kernel table is retained as an output artifact; the reader-facing page reports
the scientifically relevant structural summary.

```{r kernel-diagnostics, include=FALSE}
if (file.exists(KERNEL_CSV)) {
  kernel_summary <- read.csv(KERNEL_CSV)
  knitr::kable(
    kernel_summary,
    digits = 5,
    caption = "A, D, and E kernel diagnostics"
  )
} else {
  cat("Canonical kernel diagnostics are unavailable in the current saved result set.\n")
}
```

```{r kernel-interpretation, echo=FALSE, results="asis"}
if (file.exists(KERNEL_CSV)) {
  cat("The canonical diagnostics contain three 1,000 x 1,000 relationship matrices. All three are finite and symmetric, and their mean diagonal values are close to one. These checks support the numerical integrity of the additive, dominance, and epistatic covariance structures before interpreting predictions.\n")
}
```

# Saved-result integrity

The saved canonical results are checked against the frozen validation contract
before performance aggregation. Task count, prediction count, split IDs,
successful status, and prediction finiteness remain blocking or warning checks,
but the detailed PASS table is kept out of the main narrative.

```{r result-audit, include=FALSE}
results_available <-
  file.exists(RESULTS_RDS) &&
  file.exists(PREDICTIONS_RDS) &&
  file.exists(RUN_RDS)

if (!results_available) {
  cat("Canonical fitted results are unavailable; predictive summaries cannot be displayed.\n")
} else {
  results <- readRDS(RESULTS_RDS)
  predictions <- readRDS(PREDICTIONS_RDS)
  run_metadata <- readRDS(RUN_RDS)

  result_audit <- data.frame(
    Check = c(
      "Task rows",
      "Prediction rows",
      "Unique tasks",
      "Predictions per task",
      "Successful tasks",
      "Finite predictions",
      "Frozen split IDs"
    ),
    Status = c(
      if (nrow(results) == 150L) "PASS" else "BLOCKED",
      if (nrow(predictions) == 30000L) "PASS" else "BLOCKED",
      if (length(unique(results$TaskID)) == 150L) "PASS" else "BLOCKED",
      if (all(table(predictions$TaskID) == 200L)) "PASS" else "BLOCKED",
      if (all(results$Status == "ok")) "PASS" else "WARNING",
      if (all(is.finite(predictions$Predicted_Value))) "PASS" else "WARNING",
      if (
        setequal(unique(results$SplitID), names(cv$outer_splits))
      ) "PASS" else "BLOCKED"
    ),
    stringsAsFactors = FALSE
  )

  knitr::kable(result_audit)

  if (any(result_audit$Status == "BLOCKED")) {
    stop("Saved GBLUP-ADE results failed a blocking audit check.")
  }
}
```

For the canonical saved results, the audit verifies 150 unique scenario-by-split
tasks and 30,000 held-out predictions, corresponding to 200 assessment
individuals per task. The frozen split identifiers and prediction finiteness are
checked before any performance summary is interpreted.

# Predictive performance

The primary metrics compare the held-out final prediction
(intercept + additive + dominance + epistatic components) with the known true
genetic value. Phenotype metrics are secondary.

```{r performance}
if (results_available) {
  ok_results <- results[results$Status == "ok", , drop = FALSE]

  if (nrow(ok_results) == 0L) {
    cat(
      "No successful GBLUP-ADE tasks are available for performance aggregation.\n",
      "Inspect the saved-result audit and Error column before rerunning.\n"
    )
  } else {

  scenario_performance <- aggregate(
    cbind(
      Genetic_Correlation,
      Genetic_RMSE,
      Genetic_MAE,
      Genetic_Bias,
      Genetic_Slope,
      Genetic_R2,
      Phenotype_Correlation,
      Phenotype_RMSE,
      Phenotype_MAE,
      Phenotype_Bias,
      Phenotype_Slope,
      Phenotype_R2
    ) ~ Scenario + QTL_N + H2_nominal,
    data = ok_results,
    FUN = mean,
    na.rm = TRUE
  )

  scenario_performance <- scenario_performance[
    order(match(scenario_performance$Scenario, scenario_names)),
  ]

  genetic_performance_display <- scenario_performance[
    ,
    c(
      "Scenario",
      "QTL_N",
      "H2_nominal",
      "Genetic_Correlation",
      "Genetic_RMSE",
      "Genetic_MAE",
      "Genetic_Bias",
      "Genetic_Slope",
      "Genetic_R2"
    )
  ]

  phenotype_performance_display <- scenario_performance[
    ,
    c(
      "Scenario",
      "QTL_N",
      "H2_nominal",
      "Phenotype_Correlation",
      "Phenotype_RMSE",
      "Phenotype_MAE",
      "Phenotype_Bias",
      "Phenotype_Slope",
      "Phenotype_R2"
    )
  ]

  knitr::kable(
    genetic_performance_display,
    digits = 4,
    caption = "Primary evaluation: true genetic-value prediction across the 25 outer splits",
    table.attr = WIDE_TABLE_ATTR
  )

  knitr::kable(
    phenotype_performance_display,
    digits = 4,
    caption = "Secondary evaluation: phenotype prediction across the 25 outer splits",
    table.attr = WIDE_TABLE_ATTR
  )
  }
}
```

```{r performance-interpretation, echo=FALSE, results="asis"}
if (results_available && exists("scenario_performance")) {
  cat(
    sprintf(
      "Across the six scenarios, mean held-out genetic correlation ranges from %.4f to %.4f, while mean genetic RMSE ranges from %.4f to %.4f. The C1--C6 pattern is interpreted as performance across the prespecified genetic-complexity gradient; because QTL number and nominal heritability vary together across the complete sequence, it is not an isolated causal effect of either factor.\n",
      min(scenario_performance$Genetic_Correlation),
      max(scenario_performance$Genetic_Correlation),
      min(scenario_performance$Genetic_RMSE),
      max(scenario_performance$Genetic_RMSE)
    )
  )
}
```

# Split-level variability

Scenario means can conceal sensitivity to which individuals were held out. The
next summary therefore describes the distribution of primary genetic-value
performance across the 25 frozen outer splits within each scenario.

```{r split-summary}
if (results_available && exists("ok_results") && nrow(ok_results) > 0L) {
  split_summary <- do.call(
    rbind,
    lapply(
      split(ok_results, ok_results$Scenario),
      function(x) {
        data.frame(
          Scenario = unique(x$Scenario),
          N = nrow(x),
          Cor_mean = mean(x$Genetic_Correlation, na.rm = TRUE),
          Cor_sd = sd(x$Genetic_Correlation, na.rm = TRUE),
          Cor_min = min(x$Genetic_Correlation, na.rm = TRUE),
          Cor_max = max(x$Genetic_Correlation, na.rm = TRUE),
          RMSE_mean = mean(x$Genetic_RMSE, na.rm = TRUE),
          RMSE_sd = sd(x$Genetic_RMSE, na.rm = TRUE),
          MAE_mean = mean(x$Genetic_MAE, na.rm = TRUE),
          R2_mean = mean(x$Genetic_R2, na.rm = TRUE)
        )
      }
    )
  )

  rownames(split_summary) <- NULL
  split_summary <- split_summary[
    order(match(split_summary$Scenario, scenario_names)),
  ]

  knitr::kable(
    split_summary,
    digits = 4,
    caption = "Genetic-value prediction variability across outer splits",
    table.attr = WIDE_TABLE_ATTR
  )
}
```

Variation across the 25 outer splits quantifies the sensitivity of predictive
performance to which individuals are held out. Reporting the mean, standard
deviation, minimum, and maximum genetic correlation therefore complements the
scenario averages and prevents a single partition from determining the
conclusion.

# Computational time

The timing summary separates the one-time construction of the genomic
relationship matrices from repeated model fitting and prediction across the 150
tasks. This distinction prevents kernel construction from being interpreted as
a cost paid independently for every split.

```{r time-summary}
if (results_available) {
  time_summary <- aggregate(
    cbind(Fit_seconds, Predict_seconds) ~ Scenario,
    data = results,
    FUN = mean,
    na.rm = TRUE
  )

  time_summary$Kernel_build_seconds <- run_metadata$kernel_build_seconds

  knitr::kable(
    time_summary,
    digits = 3,
    caption = "GBLUP-ADE computational time"
  )

  cat(
    "Total run time:",
    round(run_metadata$total_run_seconds, 2),
    "seconds.\n"
  )
}
```

Because the A, D, and E matrices are built once and reused across all 150
tasks, their construction time is reported separately from the mean fitting and
prediction times.

```{r output-contract, include=FALSE}
output_files <- c(
  RESULTS_RDS,
  PREDICTIONS_RDS,
  RESULTS_CSV,
  PREDICTIONS_CSV,
  KERNEL_CSV,
  RUN_RDS
)

output_table <- data.frame(
  File = output_files,
  Exists = file.exists(output_files),
  stringsAsFactors = FALSE
)

knitr::kable(output_table)
```

# Navigation

[Previous: Validation design](simulated_data_validation.html) | [Parallel branch: Regularized ANN](simulated_data_ann.html) | [Next: Model comparison](simulated_data_comparison.html)

# Reproducibility information

```{r session-info}
sessionInfo()
```
