---
title: "Canonical comparison of GBLUP-ADE and regularized ANN models"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

# Purpose

This module provides the terminal comparison between GBLUP-ADE and the four ANN variants under the frozen outer-CV design. All competing methods are evaluated on identical outer-assessment individuals. Prediction of the true simulated genetic value is the primary outcome, with phenotype prediction retained as a secondary outcome.

Results are summarized overall, by simulation scenario, and through task-matched ANN versus GBLUP-ADE contrasts.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)

options(stringsAsFactors = FALSE)

WIDE_TABLE_ATTR <- paste(
  'style="display:block;',
  'overflow-x:auto;',
  'white-space:nowrap;',
  'width:100%;"'
)

find_project_root <- function(start = getwd()) {
  path <- normalizePath(start, winslash = "/", mustWork = TRUE)

  repeat {
    if (file.exists(file.path(path, "_workflowr.yml"))) {
      return(path)
    }

    parent <- dirname(path)

    if (identical(parent, path)) {
      stop("Could not locate workflowr project root.")
    }

    path <- parent
  }
}

PROJECT_ROOT <- find_project_root()
project_path <- function(...) file.path(PROJECT_ROOT, ...)

GBLUP_RESULTS_RDS <- project_path(
  "output", "simulated", "gblup", "simulated_gblup_ade_results.rds"
)
GBLUP_PREDICTIONS_RDS <- project_path(
  "output", "simulated", "gblup", "simulated_gblup_ade_predictions.rds"
)
ANN_RESULTS_RDS <- project_path(
  "output", "simulated", "ann", "final", "simulated_ann_results.rds"
)
ANN_PREDICTIONS_RDS <- project_path(
  "output", "simulated", "ann", "final", "simulated_ann_predictions.rds"
)

COMPARISON_DIR <- project_path("output", "simulated", "comparison")
OVERALL_CSV <- file.path(
  COMPARISON_DIR, "simulated_model_overall_summary.csv"
)
SCENARIO_CSV <- file.path(
  COMPARISON_DIR, "simulated_model_scenario_summary.csv"
)
PAIRED_CSV <- file.path(
  COMPARISON_DIR, "simulated_ann_vs_gblup_paired_summary.csv"
)
PAIRED_SCENARIO_CSV <- file.path(
  COMPARISON_DIR, "simulated_ann_vs_gblup_paired_by_scenario.csv"
)

WRITE_COMPARISON_OUTPUTS <- identical(
  Sys.getenv("WRITE_COMPARISON_OUTPUTS", "0"),
  "1"
)

metrics <- c(
  "Genetic_Correlation",
  "Genetic_RMSE",
  "Genetic_MAE",
  "Genetic_Bias",
  "Genetic_Slope",
  "Genetic_R2",
  "Phenotype_Correlation",
  "Phenotype_RMSE",
  "Phenotype_MAE",
  "Phenotype_Bias",
  "Phenotype_Slope",
  "Phenotype_R2"
)

higher_better <- c(
  "Genetic_Correlation",
  "Genetic_R2",
  "Phenotype_Correlation",
  "Phenotype_R2"
)
lower_better <- c(
  "Genetic_RMSE",
  "Genetic_MAE",
  "Phenotype_RMSE",
  "Phenotype_MAE"
)
zero_target <- c("Genetic_Bias", "Phenotype_Bias")
one_target <- c("Genetic_Slope", "Phenotype_Slope")

expected_ann_models <- c("ANN_0", "ANN_L1", "ANN_L2", "ANN_L1_L2")
```

# Canonical inputs

This terminal comparison consumes the finalized outputs from the two parallel
modeling branches. From the GBLUP-ADE module it reads the 150 task-level result
rows and 30,000 held-out predictions; from the ANN module it reads the 600
task-by-model result rows and 120,000 held-out predictions for the four ANN
variants.

These inputs are compared under the same frozen outer-CV design established
upstream. This module does not fit predictive models, tune hyperparameters,
select training duration, or modify the held-out predictions. Its role is to
verify cross-method compatibility and then summarize performance on matched
tasks and matched outer-assessment individuals.

```{r load-canonical-inputs, include=FALSE}
required_files <- c(
  GBLUP_RESULTS_RDS,
  GBLUP_PREDICTIONS_RDS,
  ANN_RESULTS_RDS,
  ANN_PREDICTIONS_RDS
)

if (!all(file.exists(required_files))) {
  stop(
    "Canonical model outputs are missing: ",
    paste(required_files[!file.exists(required_files)], collapse = ", ")
  )
}

gr <- readRDS(GBLUP_RESULTS_RDS)
gp <- readRDS(GBLUP_PREDICTIONS_RDS)
ar <- readRDS(ANN_RESULTS_RDS)
ap <- readRDS(ANN_PREDICTIONS_RDS)

cat("GBLUP result rows:", nrow(gr), "\n")
cat("GBLUP prediction rows:", nrow(gp), "\n")
cat("ANN result rows:", nrow(ar), "\n")
cat("ANN prediction rows:", nrow(ap), "\n")
cat("ANN models:", paste(sort(unique(ar$Model)), collapse = ", "), "\n")
```

# Compatibility audit

The result-level audit verifies the frozen task design. The prediction-level
audit independently verifies that the actual held-out individuals, observed
phenotypes, and true simulated genetic values are identical across methods.

```{r compatibility-audit, include=FALSE}
required_result_cols <- c(
  "Scenario", "Repetition", "Fold", "SplitID", "TaskID",
  "Model", "Status", metrics
)

required_prediction_cols <- c(
  "Scenario", "Repetition", "Fold", "SplitID", "TaskID", "Model", "ID",
  "Observed_Phenotype", "True_Genetic_Value", "Predicted_Value", "Status"
)

missing_gr <- setdiff(required_result_cols, names(gr))
missing_ar <- setdiff(required_result_cols, names(ar))
missing_gp <- setdiff(required_prediction_cols, names(gp))
missing_ap <- setdiff(required_prediction_cols, names(ap))

if (length(missing_gr)) {
  stop("Missing GBLUP result columns: ", paste(missing_gr, collapse = ", "))
}
if (length(missing_ar)) {
  stop("Missing ANN result columns: ", paste(missing_ar, collapse = ", "))
}
if (length(missing_gp)) {
  stop("Missing GBLUP prediction columns: ", paste(missing_gp, collapse = ", "))
}
if (length(missing_ap)) {
  stop("Missing ANN prediction columns: ", paste(missing_ap, collapse = ", "))
}

result_key <- function(x) {
  paste(
    x$Scenario, x$Repetition, x$Fold, x$SplitID, x$TaskID,
    sep = "|"
  )
}

assessment_key <- function(x) {
  paste(
    x$Scenario, x$Repetition, x$Fold, x$SplitID, x$TaskID,
    as.character(x$ID),
    sep = "|"
  )
}

g_result_keys <- sort(result_key(gr))
g_assessment_keys <- sort(assessment_key(gp))
gp_ordered <- gp[order(assessment_key(gp)), , drop = FALSE]

ann_result_keys_match <- vapply(
  expected_ann_models,
  function(model) {
    z <- ar[ar$Model == model, , drop = FALSE]
    nrow(z) == 150L &&
      !anyDuplicated(z$TaskID) &&
      identical(sort(result_key(z)), g_result_keys)
  },
  logical(1)
)

ann_assessment_keys_match <- vapply(
  expected_ann_models,
  function(model) {
    z <- ap[ap$Model == model, , drop = FALSE]
    nrow(z) == 30000L &&
      !anyDuplicated(z[, c("TaskID", "ID")]) &&
      identical(sort(assessment_key(z)), g_assessment_keys)
  },
  logical(1)
)

ann_phenotype_match <- vapply(
  expected_ann_models,
  function(model) {
    z <- ap[ap$Model == model, , drop = FALSE]
    z <- z[order(assessment_key(z)), , drop = FALSE]
    isTRUE(all.equal(
      z$Observed_Phenotype,
      gp_ordered$Observed_Phenotype,
      tolerance = 1e-12,
      check.attributes = FALSE
    ))
  },
  logical(1)
)

ann_true_g_match <- vapply(
  expected_ann_models,
  function(model) {
    z <- ap[ap$Model == model, , drop = FALSE]
    z <- z[order(assessment_key(z)), , drop = FALSE]
    isTRUE(all.equal(
      z$True_Genetic_Value,
      gp_ordered$True_Genetic_Value,
      tolerance = 1e-12,
      check.attributes = FALSE
    ))
  },
  logical(1)
)

compatibility_checks <- data.frame(
  Check = c(
    "GBLUP has 150 result rows",
    "ANN has 600 result rows",
    "expected ANN models are present",
    "GBLUP TaskID values are unique",
    "each ANN model has identical result-level task keys",
    "GBLUP has 30000 held-out prediction rows",
    "ANN has 120000 held-out prediction rows",
    "GBLUP has exactly 200 predictions per task",
    "ANN has exactly 200 predictions per task and model",
    "GBLUP TaskID/ID pairs are unique",
    "each ANN model has identical outer-assessment individuals",
    "observed phenotypes are identical across paired assessments",
    "true genetic values are identical across paired assessments",
    "all GBLUP result statuses are ok",
    "all ANN result statuses are ok",
    "all GBLUP prediction statuses are ok",
    "all ANN prediction statuses are ok",
    "all GBLUP comparison metrics are finite",
    "all ANN comparison metrics are finite",
    "all GBLUP held-out predictions are finite",
    "all ANN held-out predictions are finite"
  ),
  PASS = c(
    nrow(gr) == 150L,
    nrow(ar) == 600L,
    identical(sort(unique(ar$Model)), sort(expected_ann_models)),
    !anyDuplicated(gr$TaskID),
    all(ann_result_keys_match),
    nrow(gp) == 30000L,
    nrow(ap) == 120000L,
    all(table(gp$TaskID) == 200L),
    all(table(interaction(ap$TaskID, ap$Model, drop = TRUE)) == 200L),
    !anyDuplicated(gp[, c("TaskID", "ID")]),
    all(ann_assessment_keys_match),
    all(ann_phenotype_match),
    all(ann_true_g_match),
    all(gr$Status == "ok"),
    all(ar$Status == "ok"),
    all(gp$Status == "ok"),
    all(ap$Status == "ok"),
    all(vapply(
      gr[, metrics, drop = FALSE],
      function(x) all(is.finite(x)),
      logical(1)
    )),
    all(vapply(
      ar[, metrics, drop = FALSE],
      function(x) all(is.finite(x)),
      logical(1)
    )),
    all(is.finite(gp$Predicted_Value)),
    all(is.finite(ap$Predicted_Value))
  ),
  stringsAsFactors = FALSE
)

compatibility_checks$Status <- ifelse(
  compatibility_checks$PASS,
  "PASS",
  "BLOCKED"
)

knitr::kable(
  compatibility_checks[, c("Check", "Status")],
  caption = "Compatibility audit for the canonical model comparison."
)

if (any(!compatibility_checks$PASS)) {
  stop("Canonical model-comparison compatibility audit failed.")
}
```

The full compatibility audit is retained internally. For interpretation, the
fair-comparison contract can be summarized by the four conditions below.

```{r compatibility-summary}
comparison_contract <- data.frame(
  Requirement = c(
    "Same 150 outer tasks",
    "Same 200 assessment individuals per task",
    "Same phenotype and true genetic-value targets",
    "Successful finite predictions for all compared methods"
  ),
  Status = rep("CONFIRMED", 4L),
  stringsAsFactors = FALSE
)

knitr::kable(
  comparison_contract,
  caption =
    "Reader-facing contract for a fair task-matched model comparison",
  row.names = FALSE
)
```

The four conditions are prerequisites rather than performance results. Because
the hidden compatibility audit blocks execution when any condition fails, all
subsequent tables compare the same tasks, individuals, and evaluation targets.

# Interpretation of calibration metrics

The same metric definitions are applied separately to prediction of the true
simulated genetic value and to prediction of the observed phenotype.

- **Bias** is the mean prediction error, calculated as
  \(\operatorname{mean}(\text{prediction} - \text{target})\). A positive value
  indicates average overprediction, a negative value indicates average
  underprediction, and zero is the calibration target.
- **Slope** is the coefficient of the prediction term in the regression
  `lm(target ~ prediction)`. A value near one indicates appropriate
  proportional scaling of predictions relative to the target. Slope alone is
  not a complete calibration assessment because the regression intercept also
  contributes to calibration.
- **R-squared** is calculated as \(1 - SSE/SST\), using prediction errors
  relative to the target mean. It is therefore not defined here as squared
  correlation. For genuinely held-out predictions it can be negative when the
  predictions perform worse, in squared-error terms, than using the target
  mean as a reference.

Correlation, RMSE, and MAE describe complementary aspects of predictive
agreement, while Bias and Slope provide calibration diagnostics. No single
metric is interpreted as establishing universal model superiority.

The comparison code groups the metrics according to the direction of a better
value or the calibration target.

```{r comparison-metric-rules}
higher_better <- c(
  "Genetic_Correlation",
  "Genetic_R2",
  "Phenotype_Correlation",
  "Phenotype_R2"
)

lower_better <- c(
  "Genetic_RMSE",
  "Genetic_MAE",
  "Phenotype_RMSE",
  "Phenotype_MAE"
)

zero_target <- c(
  "Genetic_Bias",
  "Phenotype_Bias"
)

one_target <- c(
  "Genetic_Slope",
  "Phenotype_Slope"
)

metric_rule <- c(
  setNames(rep("higher", length(higher_better)), higher_better),
  setNames(rep("lower", length(lower_better)), lower_better),
  setNames(rep("zero", length(zero_target)), zero_target),
  setNames(rep("one", length(one_target)), one_target)
)

metric_rule_table <- data.frame(
  Metric = names(metric_rule),
  Better_value = unname(metric_rule),
  stringsAsFactors = FALSE
)

knitr::kable(
  metric_rule_table,
  caption =
    "Direction or calibration target used in paired model comparisons",
  row.names = FALSE
)
```

For `higher`, larger values are preferred; for `lower`, smaller values are
preferred. Bias is evaluated by distance from zero and slope by distance from
one.

# Overall model performance

Overall performance is shown separately for the two evaluation targets so that
the primary simulated-data criterion is not visually mixed with the secondary
phenotype criterion. Both summaries average the same 150 matched outer-CV tasks
per model.

```{r overall-comparison}
analysis_cols <- c(
  "Scenario", "Repetition", "Fold", "SplitID", "TaskID", "Model", metrics
)

all_results <- rbind(gr[, analysis_cols], ar[, analysis_cols])

overall <- aggregate(
  all_results[, metrics],
  by = list(Model = all_results$Model),
  FUN = mean
)

overall_genetic_display <- overall[, c(
  "Model",
  "Genetic_Correlation",
  "Genetic_RMSE",
  "Genetic_MAE",
  "Genetic_R2"
)]
overall_genetic_display[-1] <- lapply(
  overall_genetic_display[-1],
  round,
  4
)

overall_phenotype_display <- overall[, c(
  "Model",
  "Phenotype_Correlation",
  "Phenotype_RMSE",
  "Phenotype_MAE",
  "Phenotype_R2"
)]
overall_phenotype_display[-1] <- lapply(
  overall_phenotype_display[-1],
  round,
  4
)

knitr::kable(
  overall_genetic_display,
  caption =
    "Primary overall performance against true genetic value",
  row.names = FALSE
)

knitr::kable(
  overall_phenotype_display,
  caption =
    "Secondary overall performance against observed phenotype",
  row.names = FALSE
)
```

```{r overall-interpretation}
best_corr_model <- overall$Model[
  which.max(overall$Genetic_Correlation)
]

best_rmse_model <- overall$Model[
  which.min(overall$Genetic_RMSE)
]

best_r2_model <- overall$Model[
  which.max(overall$Genetic_R2)
]

best_overall <- overall[
  overall$Model == best_corr_model,
  ,
  drop = FALSE
]

ann_overall_now <- overall[
  overall$Model != "GBLUP_ADE",
  ,
  drop = FALSE
]

best_ann_now <- ann_overall_now[
  which.max(
    ann_overall_now$Genetic_Correlation
  ),
  ,
  drop = FALSE
]
```

Across all 150 matched outer-CV tasks, **`r best_corr_model`** has the highest
mean genetic correlation
(`r sprintf("%.4f", best_overall$Genetic_Correlation)`), the lowest mean
genetic RMSE (`r sprintf("%.4f", best_overall$Genetic_RMSE)`), and the highest
mean genetic \(R^2\) (`r sprintf("%.4f", best_overall$Genetic_R2)`). Among the
ANN variants, **`r best_ann_now$Model`** has the highest mean genetic
correlation (`r sprintf("%.4f", best_ann_now$Genetic_Correlation)`). These
statements describe the evaluated model and search space; they do not establish
universal superiority.

# Performance by simulation scenario

The next table asks whether the overall ranking persists across the six
prespecified simulation scenarios. Only the primary genetic-value metrics are
shown here; the complete metric set remains available in the versioned
comparison outputs.

```{r scenario-comparison}
scenario_summary <- aggregate(
  all_results[, metrics],
  by = list(
    Scenario = all_results$Scenario,
    Model = all_results$Model
  ),
  FUN = mean
)

scenario_display <- scenario_summary[, c(
  "Scenario", "Model", "Genetic_Correlation", "Genetic_RMSE", "Genetic_R2"
)]
scenario_display[, -(1:2)] <- lapply(
  scenario_display[, -(1:2), drop = FALSE],
  round,
  4
)

knitr::kable(
  scenario_display,
  caption =
    "Primary genetic-value prediction metrics by scenario",
  row.names = FALSE
)
```

```{r scenario-interpretation}
scenario_ids <- sort(
  unique(scenario_summary$Scenario)
)

scenario_winners <- do.call(
  rbind,
  lapply(
    scenario_ids,
    function(sc) {
      z <- scenario_summary[
        scenario_summary$Scenario == sc,
        ,
        drop = FALSE
      ]

      data.frame(
        Scenario = sc,
        Best_correlation =
          z$Model[which.max(z$Genetic_Correlation)],
        Best_RMSE =
          z$Model[which.min(z$Genetic_RMSE)],
        Best_R2 =
          z$Model[which.max(z$Genetic_R2)],
        stringsAsFactors = FALSE
      )
    }
  )
)

corr_wins <- sum(
  scenario_winners$Best_correlation == "GBLUP_ADE"
)

rmse_wins <- sum(
  scenario_winners$Best_RMSE == "GBLUP_ADE"
)

r2_wins <- sum(
  scenario_winners$Best_R2 == "GBLUP_ADE"
)
```

GBLUP-ADE has the best mean genetic correlation in
**`r corr_wins` of `r length(scenario_ids)` scenarios**, the lowest mean genetic
RMSE in **`r rmse_wins` of `r length(scenario_ids)`**, and the highest mean
genetic \(R^2\) in **`r r2_wins` of `r length(scenario_ids)`**. The C1-C6
sequence is interpreted as performance across the prespecified joint simulation
gradient, not as an isolated causal effect of QTL number or nominal
heritability.

# Paired ANN versus GBLUP-ADE comparison

For the three primary genetic-value metrics, the paired advantage is especially
simple:

- **Correlation advantage** = ANN correlation - GBLUP-ADE correlation.
- **RMSE advantage** = GBLUP-ADE RMSE - ANN RMSE.
- **\(R^2\) advantage** = ANN \(R^2\) - GBLUP-ADE \(R^2\).

Thus, a positive value always favors the ANN variant. The same convention is
extended to MAE, bias, slope, and the secondary phenotype metrics by comparing
the appropriate direction or distance from the calibration target.

```{r paired-comparison}
performance_advantage <- function(
  metric,
  ann_value,
  gblup_value
) {
  rule <- unname(
    metric_rule[metric]
  )

  candidate_advantages <- list(
    higher =
      ann_value - gblup_value,
    lower =
      gblup_value - ann_value,
    zero =
      abs(gblup_value) - abs(ann_value),
    one =
      abs(gblup_value - 1) -
      abs(ann_value - 1)
  )

  candidate_advantages[[rule]]
}

g <- gr[, c("TaskID", "SplitID", metrics)]
names(g)[-(1:2)] <- paste0(names(g)[-(1:2)], "_GBLUP")

paired_list <- list()

for (model in sort(unique(ar$Model))) {
  a <- ar[
    ar$Model == model,
    c("TaskID", "SplitID", "Scenario", metrics)
  ]

  z <- merge(a, g, by = c("TaskID", "SplitID"), all = FALSE)

  for (metric in metrics) {
    ann_value <- z[[metric]]
    gblup_value <- z[[paste0(metric, "_GBLUP")]]
    raw_difference <- ann_value - gblup_value
    advantage <- performance_advantage(metric, ann_value, gblup_value)

    paired_list[[length(paired_list) + 1L]] <- data.frame(
      Model = model,
      Metric = metric,
      N = length(raw_difference),
      Mean_ANN = mean(ann_value),
      Mean_GBLUP = mean(gblup_value),
      Mean_raw_difference = mean(raw_difference),
      Median_raw_difference = median(raw_difference),
      Mean_performance_advantage = mean(advantage),
      Median_performance_advantage = median(advantage),
      ANN_better = sum(advantage > 0),
      Equal = sum(advantage == 0),
      GBLUP_better = sum(advantage < 0),
      stringsAsFactors = FALSE
    )
  }
}

paired <- do.call(rbind, paired_list)
rownames(paired) <- NULL

paired_primary <- paired[
  paired$Metric %in% c("Genetic_Correlation", "Genetic_RMSE", "Genetic_R2"),
  c(
    "Model", "Metric", "Mean_ANN", "Mean_GBLUP",
    "Mean_performance_advantage", "ANN_better", "Equal", "GBLUP_better"
  )
]
paired_primary[, 3:5] <- lapply(
  paired_primary[, 3:5, drop = FALSE],
  round,
  4
)

knitr::kable(
  paired_primary,
  caption = paste(
    "Paired ANN versus GBLUP-ADE comparison for the primary",
    "genetic-value metrics"
  ),
  table.attr = WIDE_TABLE_ATTR,
  row.names = FALSE
)
```

The `ANN_better` and `GBLUP_better` counts refer to the same frozen
outer-assessment tasks for both methods. They therefore describe within-task
predictive wins under directly matched test samples. These counts are
descriptive summaries of repeated cross-validation and are not treated as
independent replicates for a naive fold-level significance test.

# Paired comparison by scenario

To localize the paired pattern, genetic-correlation advantage is then summarized
within each scenario over its 25 matched outer tasks. Positive values favor the
ANN variant and negative values favor GBLUP-ADE.

```{r paired-by-scenario}
paired_scenario_list <- list()

for (model in sort(unique(ar$Model))) {
  a <- ar[
    ar$Model == model,
    c("TaskID", "SplitID", "Scenario", metrics)
  ]

  z <- merge(a, g, by = c("TaskID", "SplitID"), all = FALSE)

  for (scenario in sort(unique(z$Scenario))) {
    zs <- z[z$Scenario == scenario, , drop = FALSE]

    for (metric in metrics) {
      ann_value <- zs[[metric]]
      gblup_value <- zs[[paste0(metric, "_GBLUP")]]
      advantage <- performance_advantage(metric, ann_value, gblup_value)

      paired_scenario_list[[length(paired_scenario_list) + 1L]] <- data.frame(
        Scenario = scenario,
        Model = model,
        Metric = metric,
        N = nrow(zs),
        Mean_advantage = mean(advantage),
        Median_advantage = median(advantage),
        ANN_better = sum(advantage > 0),
        Equal = sum(advantage == 0),
        GBLUP_better = sum(advantage < 0),
        stringsAsFactors = FALSE
      )
    }
  }
}

paired_scenario <- do.call(rbind, paired_scenario_list)
rownames(paired_scenario) <- NULL

paired_scenario_primary <- paired_scenario[
  paired_scenario$Metric == "Genetic_Correlation",
  c(
    "Scenario", "Model", "Mean_advantage",
    "ANN_better", "Equal", "GBLUP_better"
  )
]
paired_scenario_primary$Mean_advantage <- round(
  paired_scenario_primary$Mean_advantage,
  4
)

knitr::kable(
  paired_scenario_primary,
  caption = paste(
    "Paired genetic-correlation advantage by scenario.",
    "Positive values favor ANN."
  ),
  row.names = FALSE
)
```

The scenario-level paired summaries show whether the overall pattern is
concentrated in a subset of simulation conditions or persists across the full
C1--C6 sequence. Positive mean advantage values favor the ANN variant; negative
values favor GBLUP-ADE. Together with the scenario means above, these paired
summaries provide a task-matched description of how consistently the methods
differ.

```{r write-comparison-outputs, eval=WRITE_COMPARISON_OUTPUTS}
dir.create(COMPARISON_DIR, recursive = TRUE, showWarnings = FALSE)
write.csv(overall, OVERALL_CSV, row.names = FALSE)
write.csv(scenario_summary, SCENARIO_CSV, row.names = FALSE)
write.csv(paired, PAIRED_CSV, row.names = FALSE)
write.csv(paired_scenario, PAIRED_SCENARIO_CSV, row.names = FALSE)
```

```{r verify-comparison-outputs, include=FALSE}
canonical_files <- c(
  OVERALL_CSV,
  SCENARIO_CSV,
  PAIRED_CSV,
  PAIRED_SCENARIO_CSV
)

if (!all(file.exists(canonical_files))) {
  stop(
    "Canonical comparison CSV files are missing: ",
    paste(canonical_files[!file.exists(canonical_files)], collapse = ", ")
  )
}

frames_equal <- function(saved, computed, tolerance = 1e-12) {
  if (!identical(names(saved), names(computed)) || nrow(saved) != nrow(computed)) {
    return(FALSE)
  }

  for (name in names(saved)) {
    x <- saved[[name]]
    y <- computed[[name]]

    if (is.numeric(x) && is.numeric(y)) {
      if (!all(is.finite(x)) || !all(is.finite(y))) {
        return(FALSE)
      }
      if (!all(abs(x - y) <= tolerance)) {
        return(FALSE)
      }
    } else if (!identical(as.character(x), as.character(y))) {
      return(FALSE)
    }
  }

  TRUE
}

saved_overall <- read.csv(
  OVERALL_CSV, stringsAsFactors = FALSE, check.names = FALSE
)
saved_scenario <- read.csv(
  SCENARIO_CSV, stringsAsFactors = FALSE, check.names = FALSE
)
saved_paired <- read.csv(
  PAIRED_CSV, stringsAsFactors = FALSE, check.names = FALSE
)
saved_paired_scenario <- read.csv(
  PAIRED_SCENARIO_CSV, stringsAsFactors = FALSE, check.names = FALSE
)

output_checks <- data.frame(
  Output = basename(canonical_files),
  Status = ifelse(
    c(
      frames_equal(saved_overall, overall),
      frames_equal(saved_scenario, scenario_summary),
      frames_equal(saved_paired, paired),
      frames_equal(saved_paired_scenario, paired_scenario)
    ),
    "PASS",
    "BLOCKED"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  output_checks,
  caption = "Agreement between recomputed and versioned comparison tables."
)

if (any(output_checks$Status != "PASS")) {
  stop(
    "Canonical comparison exports do not match the recomputed results. ",
    "Review model outputs before using WRITE_COMPARISON_OUTPUTS=1."
  )
}
```

# Overall conclusion

Within the prespecified canonical model and ANN architecture search space,
GBLUP-ADE provides the strongest overall predictive performance and leads the
primary genetic-value metrics across all six simulation scenarios. Among the
neural networks, ANN-L1+L2 has the best overall averages by a very small margin
over ANN-L1, while ANN-L2 remains close to the unregularized ANN.

These results support a specific conclusion for the frozen design evaluated
here; they do not establish a universal ranking between genomic mixed models
and all possible neural-network architectures.

# Navigation

[GBLUP-ADE](simulated_data_gblup.html) | [Regularized ANN](simulated_data_ann.html) | [Home](index.html)

# Reproducibility information

`sessionInfo()` records the R version, platform, and loaded packages used to
render this terminal comparison tutorial.

```{r session-info}
sessionInfo()
```
