---
title: "Simulated data audit"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Purpose

This analysis performs the canonical audit and preprocessing of the simulated
genomic dataset before any genomic prediction model is fitted.

The audit evaluates file integrity, dimensional consistency, genotype coding,
genetic map structure, simulation scenarios, quantitative trait loci (QTL),
phenotypic and true genetic values, realized heritability, parental genotypes,
marker frequencies, duplicate observations, and linkage disequilibrium.

No genomic prediction model is fitted in this document.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

options(stringsAsFactors = FALSE)

find_project_root <- function(start = getwd()) {
  path <- normalizePath(start, winslash = "/", mustWork = TRUE)

  repeat {
    if (file.exists(file.path(path, "_workflowr.yml"))) {
      return(path)
    }

    parent <- dirname(path)

    if (identical(parent, path)) {
      stop("Could not locate workflowr project root.")
    }

    path <- parent
  }
}

PROJECT_ROOT <- find_project_root()

project_path <- function(...) {
  file.path(PROJECT_ROOT, ...)
}


SEED_GLOBAL <- 20260821L
set.seed(SEED_GLOBAL)

data_dir <- project_path("data", "dados_simulados")
output_dir <- project_path("output", "simulated", "audit")

dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

audit_log <- data.frame(
  Check = character(),
  Status = character(),
  Details = character(),
  stringsAsFactors = FALSE
)

add_audit <- function(check, status, details) {
  stopifnot(status %in% c("PASS", "WARNING", "BLOCKED"))

  audit_log <<- rbind(
    audit_log,
    data.frame(
      Check = check,
      Status = status,
      Details = as.character(details),
      stringsAsFactors = FALSE
    )
  )

  invisible(NULL)
}

fmt_dim <- function(x) {
  paste(dim(x), collapse = " x ")
}
```

## How to read this audit

Each section evaluates one property required before genomic prediction. A
`PASS` confirms that the observed data satisfy the prespecified requirement, a
`WARNING` records a feature that needs interpretation but does not invalidate
the dataset, and `BLOCKED` prevents creation of the canonical modeling object.

# Expected files

Before checking file availability, the table below maps each simulator input to
its scientific role in the audit. This separates file names from the biological
or analytical quantity that each file contributes to the workflow.

| File | Scientific role in this workflow |
|---|---|
| `controlegenetico#1.dat` | Simulation-control metadata, including scenario parameters and the QTL lists used to define C1-C6. |
| `DG#gen_F2_Vfen.txt` | Phenotypic values for the 1,000 individuals across the six simulated scenarios. |
| `DG#gen_F2_Vgen.dat` | True simulated genetic values for the same individuals and scenarios; retained for downstream evaluation, not model fitting. |
| `genoma.txt` | Text representation of the genetic map used to recover linkage groups, marker identities, and map positions. |
| `genoma_mapa.txt` | Genotype matrix in 0/1/2 coding; after orientation and validation, it is the source of the canonical genotype representation used downstream. |
| `genoma_mapa_tr.txt` | Alternative 11/12/22 genotype representation used to verify cell-by-cell equivalence with `genoma_mapa.txt`. |
| `genoma_pais.txt` | Parental genotype information used to audit the parental contrast underlying the simulated F2 population. |
| `map.rds` | Serialized map object supplied with the simulation data and inspected independently from the text map rather than assumed to be canonical. |

The audit first verifies that all eight inputs are present and then evaluates
their internal consistency in the sections that follow.

```{r expected-files}
expected_files <- c(
  "controlegenetico#1.dat",
  "DG#gen_F2_Vfen.txt",
  "DG#gen_F2_Vgen.dat",
  "genoma.txt",
  "genoma_mapa.txt",
  "genoma_mapa_tr.txt",
  "genoma_pais.txt",
  "map.rds"
)

file_paths <- file.path(data_dir, expected_files)

file_table <- data.frame(
  File = expected_files,
  Exists = file.exists(file_paths),
  Size_MB = round(file.info(file_paths)$size / 1024^2, 3),
  stringsAsFactors = FALSE
)

knitr::kable(file_table)

if (all(file_table$Exists)) {
  add_audit(
    "Expected files",
    "PASS",
    sprintf("All %d expected files are present.", length(expected_files))
  )
} else {
  missing_files <- paste(
    file_table$File[!file_table$Exists],
    collapse = ", "
  )

  add_audit(
    "Expected files",
    "BLOCKED",
    paste("Missing files:", missing_files)
  )

  write.csv(
    audit_log,
    file.path(output_dir, "simulated_audit_summary.csv"),
    row.names = FALSE
  )

  stop("Critical simulated-data files are missing.")
}
```

The eight expected simulation inputs are present. File size is recorded here as
part of file-level provenance; biological and numerical consistency are tested
explicitly in the sections that follow rather than inferred from file size.

# Read primary matrices

The first computational step reads the phenotype, true simulated genetic
value, and genotype matrices supplied by the simulator. The genotype file in
0/1/2 coding is transposed so that all downstream matrices use the same
orientation: **individuals in rows and variables or markers in columns**.

```{r read-primary-data}
phenotype <- as.matrix(
  read.table(
    file.path(data_dir, "DG#gen_F2_Vfen.txt"),
    header = FALSE
  )
)

true_genetic_value <- as.matrix(
  read.table(
    file.path(data_dir, "DG#gen_F2_Vgen.dat"),
    header = FALSE
  )
)

geno_raw_012 <- as.matrix(
  read.table(
    file.path(data_dir, "genoma_mapa.txt"),
    header = FALSE
  )
)

geno_012 <- t(geno_raw_012)
storage.mode(geno_012) <- "integer"

rm(geno_raw_012)
invisible(gc())

geno_1122 <- as.matrix(
  read.table(
    file.path(data_dir, "genoma_mapa_tr.txt"),
    header = FALSE
  )
)

storage.mode(geno_1122) <- "integer"
```

Before checking dimensions and coding, the small previews below make that
orientation explicit. Only the first six individuals and, for the genotype
matrix, the first ten markers are displayed; the complete matrices are retained
for all calculations.

```{r primary-matrix-previews}
phenotype_preview <- data.frame(
  ID = seq_len(6),
  phenotype[1:6, , drop = FALSE],
  check.names = FALSE
)
names(phenotype_preview)[-1] <- paste0("C", seq_len(ncol(phenotype)))

true_g_preview <- data.frame(
  ID = seq_len(6),
  true_genetic_value[1:6, , drop = FALSE],
  check.names = FALSE
)
names(true_g_preview)[-1] <- paste0(
  "C",
  seq_len(ncol(true_genetic_value))
)

genotype_preview <- data.frame(
  ID = seq_len(6),
  geno_012[1:6, 1:10, drop = FALSE],
  check.names = FALSE
)
names(genotype_preview)[-1] <- paste0("M", seq_len(10))

knitr::kable(
  phenotype_preview,
  digits = 4,
  caption = "First six rows of the phenotype matrix"
)

knitr::kable(
  true_g_preview,
  digits = 4,
  caption = "First six rows of the true genetic-value matrix"
)

knitr::kable(
  genotype_preview,
  caption = "First six individuals and first ten markers in 0/1/2 coding"
)
```

# Core dimensions

The next check verifies that the objects just loaded match the dimensions
defined by the simulation design. This is the point at which the reader can
connect the matrix previews above with the full dataset used downstream.

```{r core-dimensions}
core_dimensions <- data.frame(
  Object = c(
    "Phenotypic values",
    "True genetic values",
    "Genotype matrix (0/1/2)",
    "Genotype matrix (11/12/22)"
  ),
  Rows = c(
    nrow(phenotype),
    nrow(true_genetic_value),
    nrow(geno_012),
    nrow(geno_1122)
  ),
  Columns = c(
    ncol(phenotype),
    ncol(true_genetic_value),
    ncol(geno_012),
    ncol(geno_1122)
  )
)

knitr::kable(core_dimensions)

dim_ok <-
  identical(dim(phenotype), c(1000L, 6L)) &&
  identical(dim(true_genetic_value), c(1000L, 6L)) &&
  identical(dim(geno_012), c(1000L, 4010L)) &&
  identical(dim(geno_1122), c(1000L, 4010L))

add_audit(
  "Core dimensions",
  if (dim_ok) "PASS" else "BLOCKED",
  paste(
    "Expected phenotype and genetic-value matrices: 1000 x 6;",
    "expected genotype matrices: 1000 x 4010."
  )
)

finite_ok <-
  all(is.finite(phenotype)) &&
  all(is.finite(true_genetic_value))

add_audit(
  "Finite phenotype and genetic values",
  if (finite_ok) "PASS" else "BLOCKED",
  if (finite_ok) {
    "All phenotypic and true genetic values are finite."
  } else {
    "Non-finite values were detected."
  }
)

if (!dim_ok || !finite_ok) {
  write.csv(
    audit_log,
    file.path(output_dir, "simulated_audit_summary.csv"),
    row.names = FALSE
  )

  stop("Critical dimensional or numeric integrity check failed.")
}
```

The four core matrices have the dimensions required by the simulation design:
1,000 individuals by 6 scenarios for phenotypic and true genetic values, and
1,000 individuals by 4,010 markers for both genotype representations. All
phenotypic and genetic values are finite, so the dataset can proceed to the
genotype-level integrity checks.

# Genotype coding

The simulator provides the genotype matrix in two representations. The
canonical representation used in subsequent analyses is constructed from
`genoma_mapa.txt`, transposed to individuals by markers.

The following code first confirms the valid codes in both source files,
verifies their cell-by-cell equivalence after recoding 11/12/22 to 2/1/0, and
then derives the -1/0/1 representation consumed by the prediction models.

```{r genotype-coding}
codes_012 <- sort(unique(as.vector(geno_012)))
codes_1122 <- sort(unique(as.vector(geno_1122)))

geno_code_table <- data.frame(
  Matrix = c("genoma_mapa.txt", "genoma_mapa_tr.txt"),
  Codes = c(
    paste(codes_012, collapse = ", "),
    paste(codes_1122, collapse = ", ")
  )
)

knitr::kable(geno_code_table)

code_012_ok <- identical(codes_012, 0:2)
code_1122_ok <- identical(codes_1122, c(11L, 12L, 22L))

add_audit(
  "Genotype codes 0/1/2",
  if (code_012_ok) "PASS" else "BLOCKED",
  paste("Observed codes:", paste(codes_012, collapse = ", "))
)

add_audit(
  "Genotype codes 11/12/22",
  if (code_1122_ok) "PASS" else "BLOCKED",
  paste("Observed codes:", paste(codes_1122, collapse = ", "))
)

geno_1122_to_012 <- matrix(
  NA_integer_,
  nrow = nrow(geno_1122),
  ncol = ncol(geno_1122)
)

geno_1122_to_012[geno_1122 == 11L] <- 2L
geno_1122_to_012[geno_1122 == 12L] <- 1L
geno_1122_to_012[geno_1122 == 22L] <- 0L

equivalent_genotype_files <-
  identical(dim(geno_012), dim(geno_1122_to_012)) &&
  all(geno_012 == geno_1122_to_012)

add_audit(
  "Equivalence of genotype files",
  if (equivalent_genotype_files) "PASS" else "BLOCKED",
  paste(
    "Checked mapping 11 -> 2, 12 -> 1, and 22 -> 0",
    "after orienting both matrices as individuals x markers."
  )
)

rm(geno_1122, geno_1122_to_012)
invisible(gc())

geno_m101 <- geno_012 - 1L

add_audit(
  "Canonical -1/0/1 coding",
  if (identical(sort(unique(as.vector(geno_m101))), -1:1)) {
    "PASS"
  } else {
    "BLOCKED"
  },
  "Canonical coding is genotype_012 - 1."
)
```

The equivalence check compares the two genotype representations after mapping
11/12/22 to 2/1/0. When this mapping agrees across all cells, subtracting one
from the validated 0/1/2 representation produces the canonical -1/0/1 coding
used by the genomic prediction models.

To make this transformation visible, the same first six individuals and ten
markers shown above are displayed again in the canonical -1/0/1 coding.

```{r canonical-genotype-preview}
canonical_genotype_preview <- data.frame(
  ID = seq_len(6),
  geno_m101[1:6, 1:10, drop = FALSE],
  check.names = FALSE
)
names(canonical_genotype_preview)[-1] <- paste0("M", seq_len(10))

knitr::kable(
  canonical_genotype_preview,
  caption = "First six individuals and first ten markers in canonical -1/0/1 coding"
)
```

# Genetic map

The genetic-map file is parsed next to recover linkage-group membership,
marker identity, and map position. The resulting summary is used to verify the
expected chromosome structure before QTL positions or linkage disequilibrium
are interpreted.

```{r genetic-map}
map_lines <- readLines(
  file.path(data_dir, "genoma.txt"),
  warn = FALSE,
  encoding = "latin1"
)

map_parts <- strsplit(map_lines, ",", fixed = TRUE)
map_parts <- lapply(map_parts, trimws)

is_marker_line <- vapply(
  map_parts,
  function(z) {
    length(z) == 5L &&
      toupper(z[3]) == "C"
  },
  logical(1)
)

marker_matrix <- do.call(
  rbind,
  map_parts[is_marker_line]
)

if (is.null(marker_matrix) || nrow(marker_matrix) == 0L) {
  add_audit(
    "Genetic map parsing",
    "BLOCKED",
    "No marker records could be parsed from genoma.txt."
  )
  stop("Could not parse genetic map from genoma.txt.")
}

genetic_map <- data.frame(
  LG = as.integer(marker_matrix[, 1]),
  Marker_within_LG = as.integer(marker_matrix[, 2]),
  Type = marker_matrix[, 3],
  Marker = marker_matrix[, 4],
  Position_cM = as.numeric(marker_matrix[, 5]),
  stringsAsFactors = FALSE
)

genetic_map$Global_marker <- seq_len(nrow(genetic_map))

map_summary <- do.call(
  rbind,
  lapply(
    split(genetic_map, genetic_map$LG),
    function(x) {
      pos <- sort(x$Position_cM)

      data.frame(
        LG = unique(x$LG),
        Markers = nrow(x),
        Min_cM = min(pos),
        Max_cM = max(pos),
        Spacing_cM = if (length(pos) > 1L) {
          median(diff(pos))
        } else {
          NA_real_
        }
      )
    }
  )
)

rownames(map_summary) <- NULL

knitr::kable(map_summary)

map_ok <-
  nrow(genetic_map) == 4010L &&
  identical(sort(unique(genetic_map$LG)), 1:10) &&
  all(map_summary$Markers == 401L) &&
  all(abs(map_summary$Min_cM - 0) < 1e-12) &&
  all(abs(map_summary$Max_cM - 200) < 1e-12) &&
  all(abs(map_summary$Spacing_cM - 0.5) < 1e-12)

add_audit(
  "Genetic map structure",
  if (map_ok) "PASS" else "BLOCKED",
  paste(
    "Expected 10 linkage groups, 401 markers per group,",
    "200 cM per group, and 0.5 cM marker spacing."
  )
)

n_duplicate_marker_names <- sum(duplicated(genetic_map$Marker))

add_audit(
  "Unique map marker names",
  if (n_duplicate_marker_names == 0L) "PASS" else "BLOCKED",
  sprintf("%d duplicated marker names detected.", n_duplicate_marker_names)
)
```

The map audit tests the expected structure of 4,010 markers distributed across
10 linkage groups, with 401 markers per group, positions from 0 to 200 cM, and
0.5 cM spacing. It also checks marker-name uniqueness before the map is used as
the positional reference for QTL and linkage-disequilibrium analyses.

# Simulation control file

The control file defines the six simulated scenarios. The code below reads each
scenario block, extracts the prespecified metadata and QTL list, and preserves
the original numeric values while normalizing only the malformed whitespace in
one source key required for parsing.

```{r simulation-control}
control_raw <- readLines(
  file.path(data_dir, "controlegenetico#1.dat"),
  warn = FALSE
)

control_ascii <- iconv(
  control_raw,
  from = "latin1",
  to = "ASCII//TRANSLIT",
  sub = ""
)

# The original control file contains an internal tab in the C1 key
# "Escal\ta". Normalize whitespace inside keys only; numeric values are
# retained exactly as stored in the source file.
normalize_control_key <- function(x) {
  key <- sub(",.*$", "", x)
  key <- trimws(key)
  key <- gsub("[[:space:]]+", "", key)
  tolower(key)
}

control_keys <- vapply(
  control_ascii,
  normalize_control_key,
  character(1)
)

scenario_start <- which(control_keys == "variavel")

if (length(scenario_start) != 6L) {
  add_audit(
    "Simulation control blocks",
    "BLOCKED",
    paste("Expected 6 scenario blocks; found", length(scenario_start))
  )
  stop("Unexpected number of scenario blocks in control file.")
}

scenario_end <- c(
  scenario_start[-1L] - 1L,
  length(control_ascii)
)

scale_source_lines <- grep(
  "^\\s*Escal",
  control_ascii,
  value = TRUE,
  ignore.case = TRUE
)

scale_source_keys <- trimws(
  sub(",.*$", "", scale_source_lines)
)

scale_key_has_internal_whitespace <- any(
  grepl("[[:space:]]", scale_source_keys)
)

add_audit(
  "Simulation-control key formatting",
  if (scale_key_has_internal_whitespace) "WARNING" else "PASS",
  if (scale_key_has_internal_whitespace) {
    paste(
      "The source control file contains internal whitespace in the C1",
      "Scale key (Escal<tab>a). Key whitespace was normalized for parsing;",
      "the stored numeric value was not altered."
    )
  } else {
    "Control-file keys required no internal-whitespace normalization."
  }
)

extract_control_value <- function(block, key) {
  block_keys <- vapply(
    block,
    normalize_control_key,
    character(1)
  )

  target_key <- normalize_control_key(
    paste0(key, ",")
  )

  hit <- which(block_keys == target_key)

  if (length(hit) == 0L) {
    return(NA_real_)
  }

  value_text <- trimws(
    sub("^[^,]*,", "", block[hit[1L]])
  )

  suppressWarnings(
    as.numeric(value_text)
  )
}

control_rows <- vector("list", length(scenario_start))
qtl_lists <- vector("list", length(scenario_start))

for (i in seq_along(scenario_start)) {
  block <- control_ascii[
    scenario_start[i]:scenario_end[i]
  ]

  block_keys <- vapply(
    block,
    normalize_control_key,
    character(1)
  )

  selected_line <- which(
    block_keys == "locos_selecionados"
  )

  if (length(selected_line) == 0L ||
      selected_line[1L] >= length(block)) {
    add_audit(
      paste("QTL list scenario", i),
      "BLOCKED",
      "Could not locate QTL list after Locos_selecionados."
    )
    stop("Could not parse QTL list from simulation control file.")
  }

  qtl_line <- block[selected_line[1L] + 1L]

  qtl_lists[[i]] <- as.integer(
    scan(
      text = qtl_line,
      what = numeric(),
      quiet = TRUE
    )
  )

  control_rows[[i]] <- data.frame(
    Scenario = paste0("C", i),
    Variable = extract_control_value(block, "Variavel"),
    Total_loci = extract_control_value(block, "Locos_total"),
    QTL_N = extract_control_value(block, "Locos_selecionados"),
    gmd = extract_control_value(block, "gmd"),
    Model = extract_control_value(block, "Modelo"),
    H2_nominal = extract_control_value(block, "Herdabilidade"),
    Constant = extract_control_value(block, "Constante"),
    Scale = extract_control_value(block, "Escala"),
    CV_metadata = extract_control_value(block, "CV")
  )
}

control_table <- do.call(rbind, control_rows)

expected_control <- data.frame(
  Scenario = paste0("C", 1:6),
  Variable = 1:6,
  Total_loci = rep(4010, 6),
  QTL_N = c(8, 40, 80, 120, 240, 480),
  gmd = rep(0.5, 6),
  Model = rep(2, 6),
  H2_nominal = c(0.7, 0.7, 0.5, 0.5, 0.3, 0.3),
  Constant = rep(100, 6),
  Scale = rep(1, 6),
  CV_metadata = rep(12, 6)
)

knitr::kable(control_table, digits = 4)

control_ok <- isTRUE(
  all(
    control_table$Variable == expected_control$Variable,
    control_table$Total_loci == expected_control$Total_loci,
    control_table$QTL_N == expected_control$QTL_N,
    abs(control_table$gmd - expected_control$gmd) < 1e-12,
    control_table$Model == expected_control$Model,
    abs(control_table$H2_nominal - expected_control$H2_nominal) < 1e-12,
    control_table$Constant == expected_control$Constant,
    control_table$Scale == expected_control$Scale,
    control_table$CV_metadata == expected_control$CV_metadata
  )
)

add_audit(
  "Simulation scenario metadata",
  if (control_ok) "PASS" else "BLOCKED",
  paste(
    "Six scenarios expected with QTL counts",
    "8, 40, 80, 120, 240, and 480 and nominal",
    "heritabilities 0.7, 0.7, 0.5, 0.5, 0.3, and 0.3."
  )
)

qtl_count_matches <- all(
  vapply(qtl_lists, length, integer(1)) ==
    control_table$QTL_N
)

add_audit(
  "QTL-list lengths",
  if (qtl_count_matches) "PASS" else "BLOCKED",
  paste(
    "Observed QTL-list lengths:",
    paste(vapply(qtl_lists, length, integer(1)), collapse = ", ")
  )
)
```

# QTL positions

The QTL indices recorded in the control file are now mapped back to linkage
groups and genetic positions. This step checks whether all selected loci are
valid map positions and summarizes how the QTL are distributed and shared
across scenarios.

```{r qtl-audit}
qtl_rows <- do.call(
  rbind,
  lapply(
    seq_along(qtl_lists),
    function(i) {
      qtl <- qtl_lists[[i]]

      data.frame(
        Scenario = paste0("C", i),
        QTL_index = qtl,
        LG = genetic_map$LG[qtl],
        Position_cM = genetic_map$Position_cM[qtl]
      )
    }
  )
)

qtl_in_range <- all(
  qtl_rows$QTL_index >= 1L &
    qtl_rows$QTL_index <= nrow(genetic_map)
)

qtl_unique_within_scenario <- all(
  vapply(
    qtl_lists,
    function(x) length(unique(x)) == length(x),
    logical(1)
  )
)

add_audit(
  "QTL indices within map",
  if (qtl_in_range) "PASS" else "BLOCKED",
  "All QTL indices must lie between 1 and 4010."
)

add_audit(
  "Unique QTL within scenario",
  if (qtl_unique_within_scenario) "PASS" else "BLOCKED",
  "No duplicated QTL index is expected within a simulation scenario."
)

qtl_table_matrix <- table(
  factor(qtl_rows$Scenario, levels = paste0("C", 1:6)),
  factor(qtl_rows$LG, levels = 1:10)
)

qtl_by_lg <- data.frame(
  Scenario = rownames(qtl_table_matrix),
  unclass(qtl_table_matrix),
  check.names = FALSE,
  row.names = NULL
)

names(qtl_by_lg)[-1] <- paste0("LG", 1:10)

knitr::kable(qtl_by_lg)

qtl_balance_ok <- all(
  vapply(
    seq_len(nrow(control_table)),
    function(i) {
      expected_per_lg <- control_table$QTL_N[i] / 8
      observed <- as.numeric(qtl_table_matrix[i, ])

      all(observed[1:8] == expected_per_lg) &&
        all(observed[9:10] == 0)
    },
    logical(1)
  )
)

add_audit(
  "QTL distribution across linkage groups",
  if (qtl_balance_ok) "PASS" else "WARNING",
  paste(
    "The simulation design places QTL evenly across LG1-LG8;",
    "LG9 and LG10 contain no selected QTL."
  )
)

qtl_overlap <- outer(
  seq_along(qtl_lists),
  seq_along(qtl_lists),
  Vectorize(
    function(i, j) {
      length(intersect(qtl_lists[[i]], qtl_lists[[j]]))
    }
  )
)

dimnames(qtl_overlap) <- list(
  paste0("C", 1:6),
  paste0("C", 1:6)
)

knitr::kable(
  qtl_overlap,
  caption = "Number of shared QTL indices between simulation scenarios"
)
```

QTL overlap between scenarios is reported as a property of the simulation
design. Overlap is not treated as evidence of data corruption.

# Phenotypic and true genetic values

Row order is the canonical individual identifier because the simulation files
do not contain an explicit individual-ID column. All phenotype, true genetic
value, and genotype matrices therefore must contain exactly the same 1,000
rows.

```{r phenotypic-genetic-summary}
scenario_names <- paste0("C", 1:6)

colnames(phenotype) <- scenario_names
colnames(true_genetic_value) <- scenario_names

residual_value <- phenotype - true_genetic_value

scenario_summary <- data.frame(
  Scenario = scenario_names,
  QTL_N = control_table$QTL_N,
  H2_nominal = control_table$H2_nominal,
  Phenotype_mean = colMeans(phenotype),
  Phenotype_sd = apply(phenotype, 2, sd),
  Genetic_mean = colMeans(true_genetic_value),
  Genetic_sd = apply(true_genetic_value, 2, sd),
  Residual_mean = colMeans(residual_value),
  Residual_sd = apply(residual_value, 2, sd),
  Cor_P_G = vapply(
    seq_len(ncol(phenotype)),
    function(j) {
      cor(phenotype[, j], true_genetic_value[, j])
    },
    numeric(1)
  ),
  Cor_G_E = vapply(
    seq_len(ncol(phenotype)),
    function(j) {
      cor(true_genetic_value[, j], residual_value[, j])
    },
    numeric(1)
  ),
  H2_realized = vapply(
    seq_len(ncol(phenotype)),
    function(j) {
      vg <- var(true_genetic_value[, j])
      ve <- var(residual_value[, j])
      vg / (vg + ve)
    },
    numeric(1)
  ),
  VarG_over_VarP = vapply(
    seq_len(ncol(phenotype)),
    function(j) {
      var(true_genetic_value[, j]) / var(phenotype[, j])
    },
    numeric(1)
  ),
  CV_empirical_pct = 100 *
    apply(phenotype, 2, sd) /
    colMeans(phenotype),
  CV_metadata = control_table$CV_metadata
)

knitr::kable(
  scenario_summary,
  digits = 4
)

row_order_contract_ok <-
  nrow(phenotype) == nrow(true_genetic_value) &&
  nrow(phenotype) == nrow(geno_012)

add_audit(
  "Row-order contract",
  if (row_order_contract_ok) "PASS" else "BLOCKED",
  paste(
    "Phenotype, true genetic-value, and genotype matrices contain the same",
    "1000 rows. Because the simulation files do not provide explicit",
    "individual IDs, row position is adopted as the canonical individual",
    "identifier; independent identity verification is not available."
  )
)

phenotype_genetic_signal_ok <- all(
  scenario_summary$Cor_P_G > 0.5
)

add_audit(
  "Phenotype-genetic signal",
  if (phenotype_genetic_signal_ok) "PASS" else "WARNING",
  paste(
    "Phenotype-true genetic correlations:",
    paste(
      round(scenario_summary$Cor_P_G, 4),
      collapse = ", "
    )
  )
)

h2_difference <- abs(
  scenario_summary$H2_realized -
    scenario_summary$H2_nominal
)

h2_status <- if (any(h2_difference > 0.03)) {
  "WARNING"
} else {
  "PASS"
}

add_audit(
  "Nominal versus realized heritability",
  h2_status,
  paste(
    "Absolute differences:",
    paste(round(h2_difference, 4), collapse = ", "),
    ". Differences are documented rather than used to redefine the scenarios."
  )
)

cv_difference <- abs(
  scenario_summary$CV_empirical_pct -
    scenario_summary$CV_metadata
)

add_audit(
  "CV metadata versus empirical phenotype CV",
  if (any(cv_difference > 1)) "WARNING" else "PASS",
  paste(
    "The simulator reports CV metadata = 12 for all scenarios;",
    "empirical phenotype CV values are reported separately.",
    "The metadata field is therefore not interpreted automatically",
    "as the empirical coefficient of variation."
  )
)
```

Across the six scenarios, the observed phenotype--true-genetic-value
correlations range from `r sprintf("%.3f", min(scenario_summary$Cor_P_G))` to
`r sprintf("%.3f", max(scenario_summary$Cor_P_G))`. The largest absolute
difference between nominal and realized heritability is
`r sprintf("%.3f", max(abs(scenario_summary$H2_realized - scenario_summary$H2_nominal)))`.
These quantities document the realized signal in each simulated scenario;
because QTL number and nominal heritability change together across the complete
C1--C6 gradient, they are not interpreted as isolated causal effects.

# Genotype quality control

This section checks whether the simulated genotype matrix has the structural
properties required for the downstream genomic analyses. The diagnostics have
different roles:

- **missing genotypes** identify unavailable marker calls; missing cells are
  blocking here because the canonical simulated dataset is expected to be
  complete;
- **duplicate individuals** identify identical complete genotype profiles among
  progeny and are treated as blocking because they would violate the expected
  individual-level structure of this simulated population;
- **duplicate markers** identify marker columns with identical genotype profiles
  across all individuals;
- **monomorphic markers** identify loci with no observed genotype variation and
  therefore no information for distinguishing individuals;
- **minor allele frequency (MAF)** summarizes the frequency of the less common
  allele at each marker and provides a marker-level description of segregation;
- **heterozygosity** records the proportion of heterozygous individuals at each
  marker;
- the overall **0/1/2 genotype proportions** are compared with the expected
  F2 segregation pattern of approximately 1:2:1.

These checks audit the known simulated population rather than impose a generic
quality-control pipeline for empirical genomic data. The numerical criteria
below are therefore part of this simulation audit. Non-blocking departures are
reported as `WARNING` and retained explicitly instead of being filtered
silently from the canonical genotype matrix.

```{r genotype-qc}
missing_genotypes <- sum(is.na(geno_012))

duplicate_individuals <- sum(
  duplicated(geno_012, MARGIN = 1)
)

duplicate_markers <- sum(
  duplicated(geno_012, MARGIN = 2)
)

allele_frequency <- colMeans(geno_012) / 2
maf <- pmin(allele_frequency, 1 - allele_frequency)
heterozygosity <- colMeans(geno_012 == 1L)
marker_variance <- apply(geno_012, 2, var)
monomorphic_markers <- sum(marker_variance == 0)

geno_frequency <- prop.table(
  table(
    factor(
      as.vector(geno_012),
      levels = 0:2
    )
  )
)

geno_frequency_table <- data.frame(
  Genotype_code = 0:2,
  Proportion = as.numeric(geno_frequency),
  F2_expectation = c(0.25, 0.50, 0.25)
)

knitr::kable(
  geno_frequency_table,
  digits = 6
)

marker_qc_summary <- data.frame(
  Statistic = c(
    "Missing genotype cells",
    "Duplicate individuals",
    "Duplicate markers",
    "Monomorphic markers",
    "Minimum MAF",
    "Median MAF",
    "Maximum MAF",
    "Minimum heterozygosity",
    "Median heterozygosity",
    "Maximum heterozygosity"
  ),
  Value = c(
    missing_genotypes,
    duplicate_individuals,
    duplicate_markers,
    monomorphic_markers,
    min(maf),
    median(maf),
    max(maf),
    min(heterozygosity),
    median(heterozygosity),
    max(heterozygosity)
  )
)

knitr::kable(
  marker_qc_summary,
  digits = 6
)

add_audit(
  "Missing genotypes",
  if (missing_genotypes == 0L) "PASS" else "BLOCKED",
  sprintf("%d missing genotype cells.", missing_genotypes)
)

add_audit(
  "Duplicate individuals",
  if (duplicate_individuals == 0L) "PASS" else "BLOCKED",
  sprintf("%d duplicate genotype profiles.", duplicate_individuals)
)

add_audit(
  "Duplicate markers",
  if (duplicate_markers == 0L) "PASS" else "WARNING",
  sprintf("%d duplicate marker profiles.", duplicate_markers)
)

add_audit(
  "Monomorphic markers",
  if (monomorphic_markers == 0L) "PASS" else "WARNING",
  sprintf("%d monomorphic markers.", monomorphic_markers)
)

f2_frequency_ok <- max(
  abs(
    as.numeric(geno_frequency) -
      c(0.25, 0.50, 0.25)
  )
) < 0.02

add_audit(
  "Overall F2 genotype frequencies",
  if (f2_frequency_ok) "PASS" else "WARNING",
  paste(
    "Observed proportions:",
    paste(
      round(as.numeric(geno_frequency), 5),
      collapse = ", "
    )
  )
)

add_audit(
  "Minor allele frequency",
  if (min(maf) > 0.40) "PASS" else "WARNING",
  sprintf(
    "MAF range: %.4f to %.4f; median %.4f.",
    min(maf),
    max(maf),
    median(maf)
  )
)
```

The genotype matrix contains `r missing_genotypes` missing cells,
`r duplicate_individuals` duplicate individual profiles,
`r duplicate_markers` duplicate marker profiles, and
`r monomorphic_markers` monomorphic markers. Minor-allele frequency ranges from
`r sprintf("%.3f", min(maf))` to `r sprintf("%.3f", max(maf))`, with median
`r sprintf("%.3f", median(maf))`. These statistics document the marker-quality
profile used downstream; any non-blocking feature remains explicit in the audit
rather than being silently removed.

# Parental genotypes

The parental file is read separately from the F2 progeny matrix. The objective
is to confirm the expected parental contrast and linkage-group organization
that underlie the simulated segregation pattern.

```{r parents}
parents <- as.matrix(
  read.table(
    file.path(data_dir, "genoma_pais.txt"),
    header = FALSE
  )
)

parent_structure_ok <-
  nrow(parents) == 20L &&
  ncol(parents) == 405L

add_audit(
  "Parental file dimensions",
  if (parent_structure_ok) "PASS" else "BLOCKED",
  paste(
    "Expected 20 rows: two parents x 10 linkage groups,",
    "with 401 marker states per linkage group."
  )
)

if (parent_structure_ok) {
  parent_id <- parents[, 1]
  parent_lg <- parents[, 2]
  parent_nmarkers <- parents[, 3]
  parent_genotypes <- parents[, 5:405, drop = FALSE]

  parent1_ok <- all(
    parent_genotypes[parent_id == 1, ] == 1
  )

  parent2_ok <- all(
    parent_genotypes[parent_id == 2, ] == 0
  )

  lg_parent_ok <-
    identical(sort(unique(parent_lg)), 1:10) &&
    all(table(parent_lg) == 2L) &&
    all(parent_nmarkers == 401)

  add_audit(
    "Parental genotype contrast",
    if (parent1_ok && parent2_ok) "PASS" else "BLOCKED",
    paste(
      "Parent 1 is fixed for state 1 and parent 2",
      "is fixed for state 0 across all markers."
    )
  )

  add_audit(
    "Parental linkage-group structure",
    if (lg_parent_ok) "PASS" else "BLOCKED",
    "Each of the 10 linkage groups is represented for both parents."
  )
}
```

The parental file is evaluated separately from the progeny matrix to verify the
cross underlying the simulated population. The audit checks the expected
contrasting parental states across all 10 linkage groups before relating the
parental configuration to the F2 segregation pattern observed in the progeny.

# `map.rds`

The serialized map object is retained as part of the original simulation
dataset. It is inspected independently from the text map to avoid assuming its
internal structure.

```{r map-rds}
map_rds <- readRDS(
  file.path(data_dir, "map.rds")
)

map_rds_info <- data.frame(
  Property = c("Class", "Dimensions", "Names"),
  Value = c(
    paste(class(map_rds), collapse = ", "),
    if (is.null(dim(map_rds))) {
      "No dimensions"
    } else {
      paste(dim(map_rds), collapse = " x ")
    },
    if (is.null(names(map_rds))) {
      "No names"
    } else {
      paste(head(names(map_rds), 20), collapse = ", ")
    }
  )
)

knitr::kable(map_rds_info)

add_audit(
  "map.rds readability",
  "PASS",
  paste(
    "Object successfully read; class:",
    paste(class(map_rds), collapse = ", ")
  )
)

if (!is.null(dim(map_rds))) {
  map_rds_nrow <- nrow(map_rds)

  add_audit(
    "map.rds row count",
    if (map_rds_nrow == 4010L) "PASS" else "WARNING",
    paste(
      "Observed rows:",
      map_rds_nrow,
      "; text-map markers:",
      nrow(genetic_map)
    )
  )
}
```

# Linkage disequilibrium

Linkage disequilibrium (LD) is used here to describe how strongly marker
genotypes remain associated as genetic-map distance increases. For each marker
pair, the audit uses \(r^2\), the squared correlation between their standardized
0/1/2 genotype codings. Larger values indicate stronger marker association,
whereas values closer to zero indicate weaker association.

The analysis separates two complementary comparisons:

- **within-linkage-group LD** is evaluated for marker pairs separated by fixed
  map distances from 0.5 to 200 cM. In a recombining population, linked markers
  are generally expected to show stronger association at short distances and
  weaker association as map distance increases;
- an **interchromosomal reference** is obtained from 5,000 randomly sampled
  marker pairs belonging to different linkage groups. Because these pairs are
  not linked on the same chromosome, their mean \(r^2\) provides a background
  reference for interpreting residual association among widely separated
  markers within a linkage group.

The purpose is descriptive and diagnostic rather than a marker-filtering step.
In particular, persistence of elevated within-linkage-group LD at 100-200 cM
relative to the interchromosomal reference is recorded as a `WARNING` about the
simulator's long-distance recombination behavior. It is not interpreted by
itself as evidence of corrupted data, and it does not cause markers to be
removed from the canonical genotype matrix.

```{r linkage-disequilibrium, fig.cap="Linkage disequilibrium decay across fixed within-linkage-group map distances. Points and the connecting line show mean within-linkage-group r-squared; the dashed horizontal line shows the mean interchromosomal r-squared estimated from 5,000 randomly sampled marker pairs from different linkage groups."}
if (monomorphic_markers == 0L) {
  geno_scaled <- scale(
    geno_012,
    center = TRUE,
    scale = TRUE
  )

  pair_r2 <- function(i, j, z, block_size = 500L) {
    stopifnot(length(i) == length(j))

    result <- numeric(length(i))

    if (length(i) == 0L) {
      return(result)
    }

    block_start <- seq(
      1L,
      length(i),
      by = block_size
    )

    for (start in block_start) {
      end <- min(
        start + block_size - 1L,
        length(i)
      )

      b <- start:end

      r <- colSums(
        z[, i[b], drop = FALSE] *
          z[, j[b], drop = FALSE]
      ) / (nrow(z) - 1)

      result[b] <- r^2
    }

    result
  }

  ld_distances <- c(
    0.5,
    1,
    2.5,
    5,
    10,
    25,
    50,
    100,
    200
  )

  ld_rows <- vector(
    "list",
    length(ld_distances)
  )

  for (d in seq_along(ld_distances)) {
    distance <- ld_distances[d]
    r2_all <- numeric()

    for (lg in sort(unique(genetic_map$LG))) {
      idx <- which(genetic_map$LG == lg)

      idx <- idx[
        order(genetic_map$Position_cM[idx])
      ]

      pos <- genetic_map$Position_cM[idx]

      target <- match(
        pos + distance,
        pos
      )

      keep <- !is.na(target)

      if (any(keep)) {
        i <- idx[keep]
        j <- idx[target[keep]]

        r2_all <- c(
          r2_all,
          pair_r2(
            i,
            j,
            geno_scaled
          )
        )
      }
    }

    ld_rows[[d]] <- data.frame(
      Distance_cM = distance,
      Pairs = length(r2_all),
      Mean_r2 = mean(r2_all),
      Median_r2 = median(r2_all)
    )
  }

  ld_summary <- do.call(
    rbind,
    ld_rows
  )

  set.seed(SEED_GLOBAL + 101L)

  n_inter_pairs <- 5000L

  inter_i <- sample(
    seq_len(ncol(geno_012)),
    n_inter_pairs,
    replace = TRUE
  )

  inter_j <- sample(
    seq_len(ncol(geno_012)),
    n_inter_pairs,
    replace = TRUE
  )

  same_lg <- genetic_map$LG[inter_i] ==
    genetic_map$LG[inter_j]

  while (any(same_lg)) {
    inter_j[same_lg] <- sample(
      seq_len(ncol(geno_012)),
      sum(same_lg),
      replace = TRUE
    )

    same_lg <- genetic_map$LG[inter_i] ==
      genetic_map$LG[inter_j]
  }

  inter_r2 <- pair_r2(
    inter_i,
    inter_j,
    geno_scaled
  )

  interchromosomal_ld <- data.frame(
    Distance_cM = NA_real_,
    Pairs = length(inter_r2),
    Mean_r2 = mean(inter_r2),
    Median_r2 = median(inter_r2)
  )

  knitr::kable(
    ld_summary,
    digits = 5,
    caption = "Within-linkage-group LD decay"
  )

  knitr::kable(
    interchromosomal_ld,
    digits = 5,
    caption = "Interchromosomal LD reference"
  )

  plot(
    ld_summary$Distance_cM,
    ld_summary$Mean_r2,
    type = "b",
    xlab = "Map distance (cM)",
    ylab = expression(mean(r^2)),
    main = "Linkage disequilibrium decay"
  )

  abline(
    h = interchromosomal_ld$Mean_r2,
    lty = 2
  )

  short_ld <- ld_summary$Mean_r2[
    ld_summary$Distance_cM == 0.5
  ]

  medium_ld <- ld_summary$Mean_r2[
    ld_summary$Distance_cM == 10
  ]

  ld_50 <- ld_summary$Mean_r2[
    ld_summary$Distance_cM == 50
  ]

  ld_100 <- ld_summary$Mean_r2[
    ld_summary$Distance_cM == 100
  ]

  ld_200 <- ld_summary$Mean_r2[
    ld_summary$Distance_cM == 200
  ]

  ld_decay_ok <-
    short_ld > medium_ld &&
    medium_ld > ld_50 &&
    ld_50 > interchromosomal_ld$Mean_r2 &&
    interchromosomal_ld$Mean_r2 < 0.02

  add_audit(
    "Short-to-medium distance LD decay",
    if (ld_decay_ok) "PASS" else "WARNING",
    paste(
      "Mean r2 at 0.5, 10, and 50 cM:",
      paste(
        round(c(short_ld, medium_ld, ld_50), 4),
        collapse = ", "
      ),
      "; interchromosomal mean r2:",
      round(interchromosomal_ld$Mean_r2, 5)
    )
  )

  long_distance_plateau <-
    is.finite(ld_100) &&
    is.finite(ld_200) &&
    ld_100 > 0.05 &&
    ld_200 > 0.05 &&
    ld_100 > 10 * interchromosomal_ld$Mean_r2 &&
    ld_200 > 10 * interchromosomal_ld$Mean_r2

  add_audit(
    "Long-distance LD behavior",
    if (long_distance_plateau) "WARNING" else "PASS",
    paste(
      "Mean r2 at 50, 100, and 200 cM:",
      paste(
        round(c(ld_50, ld_100, ld_200), 4),
        collapse = ", "
      ),
      "; interchromosomal mean r2:",
      round(interchromosomal_ld$Mean_r2, 5),
      if (long_distance_plateau) {
        paste(
          ". LD declines strongly up to 50 cM but remains elevated at",
          "100-200 cM relative to the interchromosomal reference.",
          "This is documented as a property requiring interpretation of",
          "the simulator's long-distance recombination behavior, not as",
          "evidence of data corruption."
        )
      } else {
        ". No marked long-distance LD plateau was detected."
      }
    )
  )

  rm(geno_scaled)
  invisible(gc())
} else {
  ld_summary <- data.frame()
  interchromosomal_ld <- data.frame()

  add_audit(
    "Short-to-medium distance LD decay",
    "BLOCKED",
    "LD calculation was blocked because monomorphic markers were detected."
  )

  add_audit(
    "Long-distance LD behavior",
    "BLOCKED",
    "LD calculation was blocked because monomorphic markers were detected."
  )
}
```

The LD audit distinguishes the expected short-to-medium distance decay from
the behavior observed at the longest intrachromosomal distances. If LD remains
substantially elevated at 100-200 cM relative to the interchromosomal
reference, this is recorded as a `WARNING`. The warning does not alter or
filter the genotype matrix; it documents a feature that should be considered
when interpreting the simulator's recombination process.

# Interpretation of the simulation scenarios

The six scenarios constitute a **complexity gradient**, rather than a complete
factorial experiment between QTL number and heritability.

The nominal scenarios are:

```{r scenario-design}
scenario_design <- control_table[
  ,
  c(
    "Scenario",
    "QTL_N",
    "H2_nominal",
    "gmd",
    "Model",
    "Constant"
  )
]

knitr::kable(
  scenario_design,
  digits = 3
)
```

Direct comparisons of QTL-number complexity while nominal heritability is held
constant are therefore naturally defined within the following pairs:

- C1 versus C2 at nominal \(H^2 = 0.7\);
- C3 versus C4 at nominal \(H^2 = 0.5\);
- C5 versus C6 at nominal \(H^2 = 0.3\).

Across all six scenarios simultaneously, QTL number increases while nominal
heritability decreases. Consequently, a global trend across C1-C6 must not be
interpreted as an isolated causal effect of QTL number or heritability.

The simulator control file records `Model = 2` for all six scenarios. This
numeric field is retained here as part of the data provenance, but its
biological meaning is not inferred from the code value alone in this audit.
The downstream GBLUP-ADE benchmark is a separate prespecified modeling choice:
its additive, dominance, and epistatic components are defined and justified in
the GBLUP module rather than being established solely by `Model = 2`.

# Final audit status

```{r final-audit}
status_order <- c(
  "BLOCKED",
  "WARNING",
  "PASS"
)

audit_log$Status <- factor(
  audit_log$Status,
  levels = status_order
)

audit_log <- audit_log[
  order(audit_log$Status),
]

rownames(audit_log) <- NULL

knitr::kable(
  audit_log,
  caption = "Canonical simulated-data audit"
)

audit_counts <- as.data.frame(
  table(audit_log$Status)
)

names(audit_counts) <- c(
  "Status",
  "Checks"
)

knitr::kable(
  audit_counts,
  caption = "Audit status counts"
)

blocked_checks <- sum(
  audit_log$Status == "BLOCKED"
)

warning_checks <- sum(
  audit_log$Status == "WARNING"
)

final_classification <- if (blocked_checks > 0L) {
  "BLOCKED"
} else if (warning_checks > 0L) {
  "PASS WITH WARNINGS"
} else {
  "PASS"
}

cat(
  "\nFinal audit classification:",
  final_classification,
  "\n"
)
```

Warnings document characteristics of the simulation that require explicit
interpretation. They do not invalidate the dataset unless a corresponding
check is classified as `BLOCKED`.

# Canonical preprocessed object

The canonical preprocessed object is generated only when no blocking audit
failure is present. The saved `preprocessed_simulated_common.rds` acts as the
shared data contract for the downstream validation and modeling modules, so
those modules consume the same audited representation instead of independently
re-reading and re-interpreting the raw simulator files.

Its components have distinct scientific roles:

- `phenotype` contains the observed response values available for model fitting
  within the appropriate analysis subsets;
- `genotype_m101` is the canonical -1/0/1 genotype representation used by the
  downstream genomic prediction workflows, while `genotype_012` is retained as
  the validated source representation;
- `metadata$row_identifier` preserves the row-position identity contract used
  to align individuals across phenotypes, genotypes, validation splits, and
  predictions;
- `true_genetic_value` is retained for held-out predictive evaluation only. It
  is not used to generate validation folds, tune ANN hyperparameters, select
  training duration, or fit the predictive models;
- map, scenario, QTL, marker-QC, and LD components preserve the audited
  simulation structure and provenance for downstream checks and interpretation.
  Their presence in the common object does not make them additional response
  variables or automatic model predictors.

The next module uses this audited object to construct the frozen nested
cross-validation design. The GBLUP-ADE and ANN branches then consume that common
data contract together with the same frozen validation design, preserving
individual and task alignment across the parallel modeling workflows.

```{r save-canonical-object}
write.csv(
  audit_log,
  file.path(
    output_dir,
    "simulated_audit_summary.csv"
  ),
  row.names = FALSE
)

write.csv(
  scenario_summary,
  file.path(
    output_dir,
    "simulated_scenario_summary.csv"
  ),
  row.names = FALSE
)

write.csv(
  qtl_rows,
  file.path(
    output_dir,
    "simulated_qtl_map.csv"
  ),
  row.names = FALSE
)

if (nrow(ld_summary) > 0L) {
  write.csv(
    ld_summary,
    file.path(
      output_dir,
      "simulated_ld_summary.csv"
    ),
    row.names = FALSE
  )
}

if (blocked_checks > 0L) {
  stop(
    paste(
      blocked_checks,
      "blocking audit check(s) detected.",
      "Canonical simulated object was not created."
    )
  )
}

simulated_common <- list(
  metadata = list(
    dataset = "simulated",
    seed = SEED_GLOBAL,
    n_individuals = nrow(geno_012),
    n_markers = ncol(geno_012),
    n_scenarios = ncol(phenotype),
    row_identifier = seq_len(nrow(geno_012)),
    row_alignment = "row position",
    canonical_genotype_coding = "-1/0/1"
  ),
  phenotype = phenotype,
  true_genetic_value = true_genetic_value,
  residual_value = residual_value,
  genotype_012 = geno_012,
  genotype_m101 = geno_m101,
  genetic_map = genetic_map,
  scenario_control = control_table,
  qtl_lists = qtl_lists,
  qtl_map = qtl_rows,
  scenario_summary = scenario_summary,
  marker_qc = list(
    allele_frequency = allele_frequency,
    maf = maf,
    heterozygosity = heterozygosity
  ),
  ld_summary = ld_summary,
  interchromosomal_ld = interchromosomal_ld
)

saveRDS(
  simulated_common,
  file.path(
    output_dir,
    "preprocessed_simulated_common.rds"
  ),
  compress = "gzip"
)

cat(
  "Canonical object written to:",
  file.path(
    output_dir,
    "preprocessed_simulated_common.rds"
  ),
  "\n"
)
```

# Navigation

[Home](index.html) | [Next: Validation design](simulated_data_validation.html)

# Reproducibility information

```{r session-info}
sessionInfo()
```
