---
title: "Simulated data audit"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

## 1. Objective

This analysis performs the canonical audit and preprocessing of the simulated
genomic dataset before any genomic prediction model is fitted.

The audit evaluates file integrity, dimensional consistency, genotype coding,
genetic map structure, simulation scenarios, quantitative trait loci (QTL),
phenotypic and true genetic values, realized heritability, parental genotypes,
marker frequencies, duplicate observations, and linkage disequilibrium.

No genomic prediction model is fitted in this document.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)

options(stringsAsFactors = FALSE)
```

### 1.1 How to read this module

Each section explains a property required before genomic prediction and shows
the data summaries that are useful for scientific interpretation. The audit reports observed dimensions, codes, distributions, frequencies, and
diagnostics directly. The reader can therefore inspect what is present in the
simulated data without translating boolean guards into scientific meaning.

## 2. Packages, files, and parameters

### 2.1 Package used

This module uses base R for the audit calculations and `knitr` to format the
tables presented in the tutorial.

```{r packages, warning=FALSE, message=FALSE}
library(knitr) # Format tables and R Markdown output.
```

```{r table-formatting, include=FALSE}
format_table <- function(
  x,
  caption = NULL,
  digits = getOption("digits"),
  row.names = NA,
  ...
) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    row.names = row.names,
    table.attr = paste0(
      'class="table table-striped table-hover table-bordered" ',
      'style="width:100%; margin-bottom:0;"'
    ),
    ...
  )

  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

### 2.2 Paths and reproducibility parameters

The audit uses a single project-wide seed. The input directory contains the
original simulator files, while the output directory stores the canonical
objects created by this module.

```{r paths-and-parameters}
PROJECT_ROOT <- "."
project_path <- function(...) file.path(PROJECT_ROOT, ...)

# Reproduce stochastic diagnostics such as the interchromosomal LD sample.
SEED_GLOBAL <- 20260821L
set.seed(SEED_GLOBAL)

# Define the input and output directories used by this module.
data_dir <- project_path("data", "dados_simulados")
output_dir <- project_path("output", "simulated", "audit")
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
```

The seed affects only operations that require random sampling during the audit;
it does not modify the simulated phenotype, genotype, map, or scenario data.

### 2.3 Input files

Before checking file availability, the table below maps each simulator input to
its scientific role in the audit. This separates file names from the biological
or analytical quantity that each file contributes to the workflow.

```{r input-file-roles}
file_roles <- data.frame(
  File = c(
    "controlegenetico#1.dat",
    "DG#gen_F2_Vfen.txt",
    "DG#gen_F2_Vgen.dat",
    "genoma.txt",
    "genoma_mapa.txt",
    "genoma_mapa_tr.txt",
    "genoma_pais.txt",
    "map.rds"
  ),
  Scientific_role = c(
    "Simulation-control metadata, scenario parameters, and QTL lists.",
    "Phenotypic values for 1,000 individuals across six scenarios.",
    "True simulated genetic values retained for predictive evaluation.",
    "Text genetic map with linkage groups, markers, and positions.",
    "Genotype matrix in 0/1/2 coding used as the canonical source.",
    "Alternative 11/12/22 coding used to verify genotype equivalence.",
    "Parental genotypes used to characterize the simulated F2 cross.",
    "Serialized map object inspected as an independent map reference."
  )
)

format_table(
  file_roles,
  caption = "Scientific role of the simulator input files"
)
```

The audit first verifies that all eight inputs are present and then evaluates
their internal consistency in the sections that follow.

```{r expected-files}
expected_files <- c(
  "controlegenetico#1.dat",
  "DG#gen_F2_Vfen.txt",
  "DG#gen_F2_Vgen.dat",
  "genoma.txt",
  "genoma_mapa.txt",
  "genoma_mapa_tr.txt",
  "genoma_pais.txt",
  "map.rds"
)

file_paths <- file.path(data_dir, expected_files)
file_table <- data.frame(
  File = expected_files,
  Exists = file.exists(file_paths),
  Size_MB = round(file.info(file_paths)$size / 1024^2, 3),
  stringsAsFactors = FALSE
)

format_table(
  file_table,
  caption = "Simulator files used in the audit"
)
```

All eight simulator inputs are required before preprocessing can proceed.
Their availability and file-level provenance are checked internally; the table
above is retained as the reader-facing description of what each file contributes
scientifically to the workflow.

## 3. Reading the simulated data

### 3.1 Primary matrices

The first computational step reads the phenotype, true simulated genetic
value, and genotype matrices supplied by the simulator. The genotype file in
0/1/2 coding is transposed so that all downstream matrices use the same
orientation: **individuals in rows and variables or markers in columns**.

```{r read-primary-data}
# Read phenotypic values for the six simulation scenarios.
phenotype <- as.matrix(
  read.table(
    file.path(data_dir, "DG#gen_F2_Vfen.txt"),
    header = FALSE
  )
)

# Read the true simulated genetic values used later for evaluation.
true_genetic_value <- as.matrix(
  read.table(
    file.path(data_dir, "DG#gen_F2_Vgen.dat"),
    header = FALSE
  )
)

# Read the genotype matrix provided in 0/1/2 coding.
geno_raw_012 <- as.matrix(
  read.table(
    file.path(data_dir, "genoma_mapa.txt"),
    header = FALSE
  )
)

# Transpose the matrix so individuals are rows and markers are columns.
geno_012 <- t(geno_raw_012)
storage.mode(geno_012) <- "integer"

rm(geno_raw_012)
invisible(gc())

# Read the alternative 11/12/22 representation for coding comparison.
geno_1122 <- as.matrix(
  read.table(
    file.path(data_dir, "genoma_mapa_tr.txt"),
    header = FALSE
  )
)

storage.mode(geno_1122) <- "integer"
```

### 3.2 Data preview

Before checking dimensions and coding, the small previews below make that
orientation explicit. Only the first six individuals and, for the genotype
matrix, the first ten markers are displayed; the complete matrices are retained
for all calculations.

```{r primary-matrix-previews}
phenotype_preview <- data.frame(
  ID = seq_len(6),
  phenotype[1:6, , drop = FALSE],
  check.names = FALSE
)
names(phenotype_preview)[-1] <- paste0("C", seq_len(ncol(phenotype)))

true_g_preview <- data.frame(
  ID = seq_len(6),
  true_genetic_value[1:6, , drop = FALSE],
  check.names = FALSE
)
names(true_g_preview)[-1] <- paste0(
  "C",
  seq_len(ncol(true_genetic_value))
)

genotype_preview <- data.frame(
  ID = seq_len(6),
  geno_012[1:6, 1:10, drop = FALSE],
  check.names = FALSE
)
names(genotype_preview)[-1] <- paste0("M", seq_len(10))

format_table(
  phenotype_preview,
  digits = 4,
  caption = "First six rows of the phenotype matrix"
)

format_table(
  true_g_preview,
  digits = 4,
  caption = "First six rows of the true genetic-value matrix"
)

format_table(
  genotype_preview,
  caption = "First six individuals and first ten markers in 0/1/2 coding"
)
```

The phenotype and true-genetic-value previews contain one column for each
scenario (`C1`-`C6`), while the genotype preview contains one column for each
marker (`M1`-`M10` in this small display). In all three objects, rows represent
the same simulated individuals; only a small sample is printed, while the full
matrices remain in memory for the analyses below.

### 3.3 Matrix dimensions

The next check verifies that the objects just loaded match the dimensions
defined by the simulation design. This is the point at which the reader can
connect the matrix previews above with the full dataset used downstream.

```{r core-dimensions}
core_dimensions <- data.frame(
  Object = c(
    "Phenotypic values",
    "True genetic values",
    "Genotype matrix (0/1/2)",
    "Genotype matrix (11/12/22)"
  ),
  Rows = c(
    nrow(phenotype),
    nrow(true_genetic_value),
    nrow(geno_012),
    nrow(geno_1122)
  ),
  Columns = c(
    ncol(phenotype),
    ncol(true_genetic_value),
    ncol(geno_012),
    ncol(geno_1122)
  )
)

response_summary <- data.frame(
  Object = c("Phenotypic values", "True genetic values"),
  Missing = c(sum(is.na(phenotype)), sum(is.na(true_genetic_value))),
  Minimum = c(min(phenotype), min(true_genetic_value)),
  Median = c(median(phenotype), median(true_genetic_value)),
  Maximum = c(max(phenotype), max(true_genetic_value))
)

format_table(
  core_dimensions,
  caption = "Dimensions of the four primary matrices"
)

format_table(
  response_summary,
  digits = 4,
  caption = "Observed values and missing cells in the response matrices"
)
```

The four core matrices have the dimensions required by the simulation design:
1,000 individuals by 6 scenarios for phenotypic and true genetic values, and
1,000 individuals by 4,010 markers for both genotype representations. All
phenotypic and genetic values are finite, so the dataset can proceed to the
genotype-level integrity checks.

## 4. Genotype coding

The simulator provides the genotype matrix in two numerical representations.
This section first inspects the codes observed in each source file, then checks
their cell-by-cell equivalence, and finally derives the canonical -1/0/1 coding
used by the genomic prediction models.

### 4.1 Original genotype representations

The two source matrices encode the same three genotype classes using different
numerical labels.

```{r genotype-source-codes}
codes_012 <- sort(unique(as.vector(geno_012)))
codes_1122 <- sort(unique(as.vector(geno_1122)))

source_code_table <- data.frame(
  Representation = c("Source 0/1/2", "Source 11/12/22"),
  Observed_codes = c(
    paste(codes_012, collapse = ", "),
    paste(codes_1122, collapse = ", ")
  )
)

format_table(
  source_code_table,
  caption = "Genotype codes observed in the two source matrices"
)
```

The first source uses `0`, `1`, and `2`, whereas the alternative file uses
`11`, `12`, and `22`. These are alternative numerical representations rather
than different biological genotype calls.

### 4.2 Equivalence of genotype representations

The 11/12/22 matrix is recoded to the corresponding 2/1/0 representation and
compared directly with the original 0/1/2 matrix.

```{r genotype-equivalence}
# Translate the alternative coding to the 0/1/2 scale.
geno_1122_to_012 <- matrix(
  NA_integer_,
  nrow = nrow(geno_1122),
  ncol = ncol(geno_1122)
)

geno_1122_to_012[geno_1122 == 11L] <- 2L
geno_1122_to_012[geno_1122 == 12L] <- 1L
geno_1122_to_012[geno_1122 == 22L] <- 0L

# Count cell-level differences between the two representations.
genotype_mismatches <- sum(geno_012 != geno_1122_to_012)

format_table(
  data.frame(
    Quantity = "Cell differences after recoding 11/12/22 to 2/1/0",
    Observed = genotype_mismatches
  ),
  caption = "Equivalence of the two genotype source files"
)
```

The observed mismatch count shows whether the two files encode the same
individual-marker genotypes after accounting for their different numerical
conventions. The comparison changes only the coding scale; it does not alter
the underlying genotype calls.

### 4.3 Canonical -1/0/1 representation

The validated 0/1/2 matrix is shifted by one unit to obtain the canonical
-1/0/1 representation used by the downstream GBLUP-ADE and ANN workflows.

```{r canonical-genotype-coding}
geno_m101 <- geno_012 - 1L
codes_m101 <- sort(unique(as.vector(geno_m101)))

canonical_code_table <- data.frame(
  Representation = "Canonical -1/0/1",
  Observed_codes = paste(codes_m101, collapse = ", ")
)

format_table(
  canonical_code_table,
  caption = "Observed codes in the canonical genotype matrix"
)
```

To make the transformation concrete, the first six individuals and first ten
markers are shown below in the canonical representation.

```{r canonical-genotype-preview}
canonical_genotype_preview <- data.frame(
  ID = seq_len(6),
  geno_m101[1:6, 1:10, drop = FALSE],
  check.names = FALSE
)
names(canonical_genotype_preview)[-1] <- paste0("M", seq_len(10))

format_table(
  canonical_genotype_preview,
  caption = "First six individuals and first ten markers in canonical -1/0/1 coding"
)

rm(geno_1122, geno_1122_to_012)
invisible(gc())
```

Compared with the original 0/1/2 representation, every genotype value is
shifted by one unit: `0 -> -1`, `1 -> 0`, and `2 -> 1`. The biological genotype
calls remain unchanged; only the numerical representation used by the
prediction models changes.

## 5. Genetic map

The genetic-map file is parsed next to recover linkage-group membership,
marker identity, and map position. The resulting summary is used to verify the
expected chromosome structure before QTL positions or linkage disequilibrium
are interpreted.

```{r genetic-map, include=FALSE}
map_lines <- readLines(
  file.path(data_dir, "genoma.txt"),
  warn = FALSE,
  encoding = "latin1"
)
map_parts <- lapply(strsplit(map_lines, ",", fixed = TRUE), trimws)
is_marker_line <- vapply(
  map_parts,
  function(z) length(z) == 5L && toupper(z[3]) == "C",
  logical(1)
)
marker_matrix <- do.call(rbind, map_parts[is_marker_line])

genetic_map <- data.frame(
  LG = as.integer(marker_matrix[, 1]),
  Marker_within_LG = as.integer(marker_matrix[, 2]),
  Type = marker_matrix[, 3],
  Marker = marker_matrix[, 4],
  Position_cM = as.numeric(marker_matrix[, 5]),
  stringsAsFactors = FALSE
)
genetic_map$Global_marker <- seq_len(nrow(genetic_map))

map_summary <- do.call(
  rbind,
  lapply(split(genetic_map, genetic_map$LG), function(x) {
    pos <- sort(x$Position_cM)
    data.frame(
      LG = unique(x$LG),
      Markers = nrow(x),
      Min_cM = min(pos),
      Max_cM = max(pos),
      Spacing_cM = median(diff(pos))
    )
  })
)
rownames(map_summary) <- NULL
n_duplicate_marker_names <- sum(duplicated(genetic_map$Marker))
```

### 5.1 Map structure and marker positions

The parsed map is summarized below so the positional structure used by the QTL
and LD analyses is visible to the reader.

```{r genetic-map-summary}
format_table(
  map_summary,
  digits = 3,
  caption = "Canonical genetic-map structure",
  row.names = FALSE
)

format_table(
  data.frame(
    Quantity = c("Map markers", "Linkage groups", "Duplicate marker names"),
    Observed = c(
      nrow(genetic_map),
      length(unique(genetic_map$LG)),
      n_duplicate_marker_names
    )
  ),
  caption = "Observed map totals",
  row.names = FALSE
)
```

The map contains 10 linkage groups with 401 markers per group. Every group spans
0-200 cM with 0.5 cM spacing, giving 4,010 marker positions in total. Marker
names are unique, so the map provides an unambiguous positional reference for
the analyses that follow.

## 6. Simulation scenarios

The control file defines the six simulated scenarios. The code below reads each
scenario block, extracts the prespecified metadata and QTL list, and preserves
the original numeric values while normalizing only the malformed whitespace in
one source key required for parsing.

```{r simulation-control, include=FALSE}
control_raw <- readLines(
  file.path(data_dir, "controlegenetico#1.dat"),
  warn = FALSE
)
control_ascii <- iconv(
  control_raw,
  from = "latin1",
  to = "ASCII//TRANSLIT",
  sub = ""
)

normalize_control_key <- function(x) {
  key <- trimws(sub(",.*$", "", x))
  key <- gsub("[[:space:]]+", "", key)
  tolower(key)
}

extract_control_value <- function(block, key) {
  block_keys <- vapply(block, normalize_control_key, character(1))
  hit <- which(block_keys == normalize_control_key(paste0(key, ",")))[1L]
  value_text <- trimws(sub("^[^,]*,", "", block[hit]))
  suppressWarnings(as.numeric(value_text))
}

control_keys <- vapply(control_ascii, normalize_control_key, character(1))
scenario_start <- which(control_keys == "variavel")
scenario_end <- c(scenario_start[-1L] - 1L, length(control_ascii))

scale_source_lines <- grep(
  "^\\s*Escal",
  control_ascii,
  value = TRUE,
  ignore.case = TRUE
)
scale_source_keys <- trimws(sub(",.*$", "", scale_source_lines))

control_rows <- vector("list", length(scenario_start))
qtl_lists <- vector("list", length(scenario_start))

for (i in seq_along(scenario_start)) {
  block <- control_ascii[scenario_start[i]:scenario_end[i]]
  block_keys <- vapply(block, normalize_control_key, character(1))
  selected_line <- which(block_keys == "locos_selecionados")[1L]
  qtl_line <- block[selected_line + 1L]

  qtl_lists[[i]] <- as.integer(
    scan(text = qtl_line, what = numeric(), quiet = TRUE)
  )

  control_rows[[i]] <- data.frame(
    Scenario = paste0("C", i),
    Variable = extract_control_value(block, "Variavel"),
    Total_loci = extract_control_value(block, "Locos_total"),
    QTL_N = extract_control_value(block, "Locos_selecionados"),
    gmd = extract_control_value(block, "gmd"),
    Model = extract_control_value(block, "Modelo"),
    H2_nominal = extract_control_value(block, "Herdabilidade"),
    Constant = extract_control_value(block, "Constante"),
    Scale = extract_control_value(block, "Escala"),
    CV_metadata = extract_control_value(block, "CV")
  )
}

control_table <- do.call(rbind, control_rows)
```

### 6.1 Scenario parameters

The parsed scenario parameters are displayed separately from the loading logic.

```{r simulation-control-summary}
control_display <- control_table[
  ,
  c(
    "Scenario", "QTL_N", "H2_nominal", "gmd",
    "Model", "Constant", "Scale", "CV_metadata"
  )
]

format_table(
  control_display,
  digits = 4,
  caption = "Simulation-scenario parameters used downstream",
  row.names = FALSE
)

format_table(
  data.frame(
    Scenario = paste0("C", seq_along(qtl_lists)),
    QTL_in_file = vapply(qtl_lists, length, integer(1)),
    QTL_metadata = control_table$QTL_N
  ),
  caption = "QTL counts read from each scenario block",
  row.names = FALSE
)
```

QTL number increases from 8 in C1 to 480 in C6. Nominal heritability decreases
in pairs from 0.7 to 0.5 and then 0.3, while `gmd = 0.5`, `Model = 2`, and
`Constant = 100` remain fixed. The six scenarios therefore form a joint
complexity gradient rather than a factorial design in which QTL number and
heritability vary independently.

### 6.2 QTL positions and distribution

The QTL indices recorded in the control file are now mapped back to linkage
groups and genetic positions. This step checks whether all selected loci are
valid map positions and summarizes how the QTL are distributed and shared
across scenarios.

```{r qtl-audit}
qtl_rows <- do.call(
  rbind,
  lapply(seq_along(qtl_lists), function(i) {
    qtl <- qtl_lists[[i]]
    data.frame(
      Scenario = paste0("C", i),
      QTL_index = qtl,
      LG = genetic_map$LG[qtl],
      Position_cM = genetic_map$Position_cM[qtl]
    )
  })
)

qtl_table_matrix <- table(
  factor(qtl_rows$Scenario, levels = paste0("C", 1:6)),
  factor(qtl_rows$LG, levels = 1:10)
)
qtl_by_lg <- data.frame(
  Scenario = rownames(qtl_table_matrix),
  unclass(qtl_table_matrix),
  check.names = FALSE,
  row.names = NULL
)
names(qtl_by_lg)[-1] <- paste0("LG", 1:10)

qtl_overlap <- outer(
  seq_along(qtl_lists),
  seq_along(qtl_lists),
  Vectorize(function(i, j) length(intersect(qtl_lists[[i]], qtl_lists[[j]])))
)
dimnames(qtl_overlap) <- list(paste0("C", 1:6), paste0("C", 1:6))

qtl_observed <- data.frame(
  Scenario = paste0("C", seq_along(qtl_lists)),
  QTL_N = vapply(qtl_lists, length, integer(1)),
  Minimum_index = vapply(qtl_lists, min, integer(1)),
  Maximum_index = vapply(qtl_lists, max, integer(1)),
  Duplicated_indices = vapply(
    qtl_lists,
    function(x) sum(duplicated(x)),
    integer(1)
  )
)

format_table(
  qtl_observed,
  caption = "Observed QTL indices by scenario",
  row.names = FALSE
)
format_table(
  qtl_by_lg,
  caption = "Number of QTL per linkage group",
  row.names = FALSE
)
format_table(
  qtl_overlap,
  caption = "Number of shared QTL indices between simulation scenarios"
)
```

The QTL counts are distributed evenly across LG1-LG8 in every scenario, while
LG9 and LG10 contain no selected QTL. This distribution is part of the
simulation design, not a marker-filtering result.

QTL overlap between scenarios is reported as a property of the simulation
design. Overlap is not treated as evidence of data corruption.

### 6.3 Scenario interpretation

The six scenarios constitute a **complexity gradient**, rather than a complete
factorial experiment between QTL number and heritability.

The nominal scenarios are:

```{r scenario-design}
scenario_design <- control_table[
  ,
  c(
    "Scenario",
    "QTL_N",
    "H2_nominal",
    "gmd",
    "Model",
    "Constant"
  )
]

format_table(
  scenario_design,
  digits = 3,
  caption =
    "Joint C1-C6 simulation-complexity gradient",
  row.names = FALSE
)
```

Direct comparisons of QTL-number complexity while nominal heritability is held
constant are therefore naturally defined within the following pairs:

- C1 versus C2 at nominal \(H^2 = 0.7\);
- C3 versus C4 at nominal \(H^2 = 0.5\);
- C5 versus C6 at nominal \(H^2 = 0.3\).

Across all six scenarios simultaneously, QTL number increases while nominal
heritability decreases. Consequently, a global trend across C1-C6 must not be
interpreted as an isolated causal effect of QTL number or heritability.

The simulator control file records `Model = 2` for all six scenarios. This
numeric field is retained here as part of the data provenance, but its
biological meaning is not inferred from the code value alone in this audit.
The downstream GBLUP-ADE benchmark is a separate prespecified modeling choice:
its additive, dominance, and epistatic components are defined and justified in
the GBLUP module rather than being established solely by `Model = 2`.

## 7. Phenotypic and genetic signal

Row order is the canonical individual identifier because the simulation files
do not contain an explicit individual-ID column. All phenotype, true genetic
value, and genotype matrices therefore must contain exactly the same 1,000
rows.

### 7.1 Scenario-level signal and realized heritability

The scenario summary is displayed in two complementary tables. The first
emphasizes the simulated signal and realized heritability; the second reports
dispersion, residual association, and coefficient-of-variation diagnostics.
The complete `scenario_summary` object is still retained unchanged for the
canonical output files and downstream analyses.

```{r phenotypic-genetic-summary}
scenario_names <- paste0("C", 1:6)
colnames(phenotype) <- scenario_names
colnames(true_genetic_value) <- scenario_names
residual_value <- phenotype - true_genetic_value

scenario_summary <- data.frame(
  Scenario = scenario_names,
  QTL_N = control_table$QTL_N,
  H2_nominal = control_table$H2_nominal,
  Phenotype_mean = colMeans(phenotype),
  Phenotype_sd = apply(phenotype, 2, sd),
  Genetic_mean = colMeans(true_genetic_value),
  Genetic_sd = apply(true_genetic_value, 2, sd),
  Residual_mean = colMeans(residual_value),
  Residual_sd = apply(residual_value, 2, sd),
  Cor_P_G = vapply(
    seq_len(ncol(phenotype)),
    function(j) cor(phenotype[, j], true_genetic_value[, j]),
    numeric(1)
  ),
  Cor_G_E = vapply(
    seq_len(ncol(phenotype)),
    function(j) cor(true_genetic_value[, j], residual_value[, j]),
    numeric(1)
  ),
  H2_realized = vapply(
    seq_len(ncol(phenotype)),
    function(j) {
      vg <- var(true_genetic_value[, j])
      ve <- var(residual_value[, j])
      vg / (vg + ve)
    },
    numeric(1)
  ),
  VarG_over_VarP = vapply(
    seq_len(ncol(phenotype)),
    function(j) var(true_genetic_value[, j]) / var(phenotype[, j]),
    numeric(1)
  ),
  CV_empirical_pct = 100 * apply(phenotype, 2, sd) / colMeans(phenotype),
  CV_metadata = control_table$CV_metadata
)
scenario_display <- transform(
  scenario_summary,
  H2_difference = H2_realized - H2_nominal,
  CV_difference = CV_empirical_pct - CV_metadata
)

scenario_signal_display <- scenario_display[
  , c(
    "Scenario", "QTL_N", "H2_nominal", "H2_realized",
    "H2_difference", "Phenotype_mean", "Genetic_mean", "Cor_P_G"
  )
]
scenario_diagnostics_display <- scenario_display[
  , c(
    "Scenario", "Phenotype_sd", "Genetic_sd", "Residual_mean",
    "Residual_sd", "Cor_G_E", "VarG_over_VarP",
    "CV_empirical_pct", "CV_metadata", "CV_difference"
  )
]

format_table(
  scenario_signal_display,
  digits = 4,
  caption = "Phenotypic and true-genetic signal by simulation scenario",
  row.names = FALSE
)
format_table(
  scenario_diagnostics_display,
  digits = 4,
  caption = "Dispersion and residual diagnostics by simulation scenario",
  row.names = FALSE
)
```

Across the six scenarios, the observed phenotype--true-genetic-value
correlations range from `r sprintf("%.3f", min(scenario_summary$Cor_P_G))` to
`r sprintf("%.3f", max(scenario_summary$Cor_P_G))`. The largest absolute
difference between nominal and realized heritability is
`r sprintf("%.3f", max(abs(scenario_summary$H2_realized - scenario_summary$H2_nominal)))`.
These quantities document the realized signal in each simulated scenario;
because QTL number and nominal heritability change together across the complete
C1--C6 gradient, they are not interpreted as isolated causal effects.

## 8. Genotype diagnostics

### 8.1 Marker and individual diagnostics

This section checks whether the simulated genotype matrix has the structural
properties required for the downstream genomic analyses. The diagnostics have
different roles:

- **missing genotypes** identify unavailable marker calls; missing cells are
  blocking here because the canonical simulated dataset is expected to be
  complete;
- **duplicate individuals** identify identical complete genotype profiles among
  progeny and are treated as blocking because they would violate the expected
  individual-level structure of this simulated population;
- **duplicate markers** identify marker columns with identical genotype profiles
  across all individuals;
- **monomorphic markers** identify loci with no observed genotype variation and
  therefore no information for distinguishing individuals;
- **minor allele frequency (MAF)** summarizes the frequency of the less common
  allele at each marker and provides a marker-level description of segregation;
- **heterozygosity** records the proportion of heterozygous individuals at each
  marker;
- the overall **0/1/2 genotype proportions** are compared with the expected
  F2 segregation pattern of approximately 1:2:1.

These checks audit the known simulated population rather than impose a generic
quality-control pipeline for empirical genomic data. The numerical criteria
below are therefore part of this simulation audit. Observed departures are retained explicitly instead of being filtered
silently from the canonical genotype matrix.

```{r genotype-qc}
missing_genotypes <- sum(is.na(geno_012))
duplicate_individuals <- sum(duplicated(geno_012, MARGIN = 1))
duplicate_markers <- sum(duplicated(geno_012, MARGIN = 2))

allele_frequency <- colMeans(geno_012) / 2
maf <- pmin(allele_frequency, 1 - allele_frequency)
heterozygosity <- colMeans(geno_012 == 1L)
marker_variance <- apply(geno_012, 2, var)
monomorphic_markers <- sum(marker_variance == 0)

geno_frequency <- prop.table(
  table(factor(as.vector(geno_012), levels = 0:2))
)
geno_frequency_table <- data.frame(
  Genotype_code = 0:2,
  Proportion = as.numeric(geno_frequency),
  F2_expectation = c(0.25, 0.50, 0.25)
)
marker_qc_summary <- data.frame(
  Statistic = c(
    "Missing genotype cells", "Duplicate individuals",
    "Duplicate markers", "Monomorphic markers",
    "Minimum MAF", "Median MAF", "Maximum MAF",
    "Minimum heterozygosity", "Median heterozygosity",
    "Maximum heterozygosity"
  ),
  Value = c(
    missing_genotypes, duplicate_individuals, duplicate_markers,
    monomorphic_markers, min(maf), median(maf), max(maf),
    min(heterozygosity), median(heterozygosity), max(heterozygosity)
  )
)

format_table(
  geno_frequency_table,
  digits = 6,
  caption = "Observed 0/1/2 genotype proportions and F2 expectation",
  row.names = FALSE
)
format_table(
  marker_qc_summary,
  digits = 6,
  caption = "Marker and individual quality-control summary",
  row.names = FALSE
)
```

The genotype matrix contains `r missing_genotypes` missing cells,
`r duplicate_individuals` duplicate individual profiles,
`r duplicate_markers` duplicate marker profiles, and
`r monomorphic_markers` monomorphic markers. Minor-allele frequency ranges from
`r sprintf("%.3f", min(maf))` to `r sprintf("%.3f", max(maf))`, with median
`r sprintf("%.3f", median(maf))`. These statistics document the marker-quality
profile used downstream; any non-blocking feature remains explicit in the audit
rather than being silently removed.

## 9. Parental and map references

### 9.1 Parental genotypes

The parental file is read separately from the F2 progeny matrix. The objective
is to confirm the expected parental contrast and linkage-group organization
that underlie the simulated segregation pattern.

```{r parents, include=FALSE}
parents <- as.matrix(
  read.table(file.path(data_dir, "genoma_pais.txt"), header = FALSE)
)
parent_id <- parents[, 1]
parent_lg <- parents[, 2]
parent_nmarkers <- parents[, 3]
parent_genotypes <- parents[, 5:405, drop = FALSE]
```

The parental file is loaded and checked internally, while the contrasting
genotype states relevant to the simulated cross are summarized below.

```{r parent-summary}
parent_summary <- data.frame(
  Parent = c(1, 2),
  Linkage_groups = c(
    length(
      unique(
        parents[
          parents[, 1] == 1,
          2
        ]
      )
    ),
    length(
      unique(
        parents[
          parents[, 1] == 2,
          2
        ]
      )
    )
  ),
  Marker_state = c(
    unique(
      as.vector(
        parents[
          parents[, 1] == 1,
          5:405,
          drop = FALSE
        ]
      )
    )[1],
    unique(
      as.vector(
        parents[
          parents[, 1] == 2,
          5:405,
          drop = FALSE
        ]
      )
    )[1]
  )
)

format_table(
  parent_summary,
  caption =
    "Parental contrast underlying the simulated F2 population",
  row.names = FALSE
)
```

Both parents are represented across all 10 linkage groups. Parent 1 is fixed
for marker state 1 and parent 2 for marker state 0, providing the contrasting
parental configuration from which the simulated F2 segregation is generated.

### 9.2 Serialized map reference

The serialized `map.rds` object is retained as part of the original simulation
dataset and is checked independently from the text map. Its readability and
basic structure remain part of the internal integrity checks, but the object is
not treated as an additional response variable or as a separate modeling branch.

```{r map-rds}
map_rds <- readRDS(file.path(data_dir, "map.rds"))
map_rds_info <- data.frame(
  Property = c("Class", "Dimensions", "Names"),
  Value = c(
    paste(class(map_rds), collapse = ", "),
    if (is.null(dim(map_rds))) "No dimensions" else paste(dim(map_rds), collapse = " x "),
    if (is.null(names(map_rds))) "No names" else paste(head(names(map_rds), 20), collapse = ", ")
  )
)
format_table(map_rds_info, caption = "Serialized map object")
```

The serialized map object provides an independent representation of the genetic
map and is retained as a provenance and integrity reference. The text-derived
`genetic_map` remains the positional object used for the QTL and LD summaries
in this module.

## 10. Linkage disequilibrium

Linkage disequilibrium (LD) is used here to describe how strongly marker
genotypes remain associated as genetic-map distance increases. For each marker
pair, the audit uses \(r^2\), the squared correlation between their standardized
0/1/2 genotype codings. Larger values indicate stronger marker association,
whereas values closer to zero indicate weaker association.

### 10.1 LD comparisons

The analysis separates two complementary comparisons:

- **within-linkage-group LD** is evaluated for marker pairs separated by fixed
  map distances from 0.5 to 200 cM. In a recombining population, linked markers
  are generally expected to show stronger association at short distances and
  weaker association as map distance increases;
- an **interchromosomal reference** is obtained from 5,000 randomly sampled
  marker pairs belonging to different linkage groups. Because these pairs are
  not linked on the same chromosome, their mean \(r^2\) provides a background
  reference for interpreting residual association among widely separated
  markers within a linkage group.

The purpose is descriptive and diagnostic rather than a marker-filtering step.
In particular, persistence of elevated within-linkage-group LD at 100-200 cM
relative to the interchromosomal reference is reported explicitly as a feature of the
simulator's long-distance recombination behavior. It is not interpreted by
itself as evidence of corrupted data, and it does not cause markers to be
removed from the canonical genotype matrix.

```{r linkage-disequilibrium, fig.cap="Linkage disequilibrium decay across fixed within-linkage-group map distances. Points and the connecting line show mean within-linkage-group r-squared; the dashed horizontal line shows the mean interchromosomal r-squared estimated from 5,000 randomly sampled marker pairs from different linkage groups."}
# Standardize marker genotypes before calculating correlations.
geno_scaled <- scale(
  geno_012,
  center = TRUE,
  scale = TRUE
)

# Calculate r-squared for selected marker pairs.
pair_r2 <- function(
  i,
  j,
  z,
  block_size = 500L
) {
  result <- numeric(length(i))

  n_blocks <- ceiling(
    length(i) / block_size
  )

  for (block in seq_len(n_blocks)) {
    start <- (block - 1L) * block_size + 1L
    end <- min(
      start + block_size - 1L,
      length(i)
    )

    b <- start:end

    r <- colSums(
      z[, i[b], drop = FALSE] *
        z[, j[b], drop = FALSE]
    ) / (nrow(z) - 1)

    result[b] <- r^2
  }

  result
}

ld_distances <- c(
  0.5,
  1,
  2.5,
  5,
  10,
  25,
  50,
  100,
  200
)

ld_rows <- vector(
  "list",
  length(ld_distances)
)

# Summarize LD at each prespecified map distance.
for (d in seq_along(ld_distances)) {
  distance <- ld_distances[d]
  r2_all <- numeric()

  for (lg in sort(unique(genetic_map$LG))) {
    idx <- which(genetic_map$LG == lg)
    idx <- idx[
      order(
        genetic_map$Position_cM[idx]
      )
    ]

    pos <- genetic_map$Position_cM[idx]

    target <- match(
      pos + distance,
      pos
    )
    keep <- !is.na(target)

    i <- idx[keep]
    j <- idx[target[keep]]

    r2_all <- c(
      r2_all,
      pair_r2(
        i,
        j,
        geno_scaled
      )
    )
  }

  ld_rows[[d]] <- data.frame(
    Distance_cM = distance,
    Pairs = length(r2_all),
    Mean_r2 = mean(r2_all),
    Median_r2 = median(r2_all)
  )
}

ld_summary <- do.call(
  rbind,
  ld_rows
)

# Build a reproducible interchromosomal LD reference.
set.seed(SEED_GLOBAL + 101L)

n_inter_pairs <- 5000L

inter_i <- sample(
  seq_len(ncol(geno_012)),
  n_inter_pairs,
  replace = TRUE
)

inter_j <- sample(
  seq_len(ncol(geno_012)),
  n_inter_pairs,
  replace = TRUE
)

same_lg <-
  genetic_map$LG[inter_i] ==
  genetic_map$LG[inter_j]

# Resample only the second marker until every pair is interchromosomal.
while (any(same_lg)) {
  inter_j[same_lg] <- sample(
    seq_len(ncol(geno_012)),
    sum(same_lg),
    replace = TRUE
  )

  same_lg <-
    genetic_map$LG[inter_i] ==
    genetic_map$LG[inter_j]
}

inter_r2 <- pair_r2(
  inter_i,
  inter_j,
  geno_scaled
)

interchromosomal_ld <- data.frame(
  Distance_cM = NA_real_,
  Pairs = length(inter_r2),
  Mean_r2 = mean(inter_r2),
  Median_r2 = median(inter_r2)
)

format_table(
  ld_summary,
  digits = 5,
  caption =
    "Within-linkage-group LD decay",
  row.names = FALSE
)

format_table(
  interchromosomal_ld,
  digits = 5,
  caption =
    "Interchromosomal LD reference",
  row.names = FALSE
)

plot(
  ld_summary$Distance_cM,
  ld_summary$Mean_r2,
  type = "b",
  xlab = "Map distance (cM)",
  ylab = expression(mean(r^2)),
  main = "Linkage disequilibrium decay"
)

abline(
  h = interchromosomal_ld$Mean_r2,
  lty = 2
)

rm(geno_scaled)
invisible(gc())
```

### 10.2 LD interpretation

The LD audit distinguishes the expected short-to-medium distance decay from
the behavior observed at the longest intrachromosomal distances. If LD remains
substantially elevated at 100-200 cM relative to the interchromosomal
reference, this remains visible in the LD table and does not alter or
filter the genotype matrix; it documents a feature that should be considered
when interpreting the simulator's recombination process.

```{r audit-output-contract, include=FALSE}
compat_status <- function(condition, failure = "BLOCKED", success = "PASS") {
  c(failure, success)[1L + as.integer(isTRUE(condition))]
}

audit_log <- data.frame(
  Check = character(),
  Status = character(),
  Details = character(),
  stringsAsFactors = FALSE
)

add_compat_audit <- function(check, status, details) {
  audit_log <<- rbind(
    audit_log,
    data.frame(
      Check = check,
      Status = status,
      Details = as.character(details),
      stringsAsFactors = FALSE
    )
  )
  invisible(NULL)
}

scale_key_has_internal_whitespace <- any(grepl("[[:space:]]", scale_source_keys))
h2_difference <- abs(scenario_summary$H2_realized - scenario_summary$H2_nominal)
cv_difference <- abs(scenario_summary$CV_empirical_pct - scenario_summary$CV_metadata)

short_ld <- ld_summary$Mean_r2[ld_summary$Distance_cM == 0.5]
medium_ld <- ld_summary$Mean_r2[ld_summary$Distance_cM == 10]
ld_50 <- ld_summary$Mean_r2[ld_summary$Distance_cM == 50]
ld_100 <- ld_summary$Mean_r2[ld_summary$Distance_cM == 100]
ld_200 <- ld_summary$Mean_r2[ld_summary$Distance_cM == 200]
ld_decay_ok <-
  short_ld > medium_ld &&
  medium_ld > ld_50 &&
  ld_50 > interchromosomal_ld$Mean_r2 &&
  interchromosomal_ld$Mean_r2 < 0.02
long_distance_plateau <-
  is.finite(ld_100) && is.finite(ld_200) &&
  ld_100 > 0.05 && ld_200 > 0.05 &&
  ld_100 > 10 * interchromosomal_ld$Mean_r2 &&
  ld_200 > 10 * interchromosomal_ld$Mean_r2

dim_ok <-
  identical(dim(phenotype), c(1000L, 6L)) &&
  identical(dim(true_genetic_value), c(1000L, 6L)) &&
  identical(dim(geno_012), c(1000L, 4010L))
finite_ok <- all(is.finite(phenotype)) && all(is.finite(true_genetic_value))
code_012_ok <- identical(sort(unique(as.vector(geno_012))), 0:2)
code_1122_ok <- identical(codes_1122, c(11L, 12L, 22L))
canonical_coding_ok <- identical(sort(unique(as.vector(geno_m101))), -1:1)
equivalent_genotype_files <- genotype_mismatches == 0L

map_ok <-
  nrow(genetic_map) == 4010L &&
  length(unique(genetic_map$LG)) == 10L &&
  all(map_summary$Markers == 401L) &&
  all(abs(map_summary$Min_cM) < 1e-12) &&
  all(abs(map_summary$Max_cM - 200) < 1e-12) &&
  all(abs(map_summary$Spacing_cM - 0.5) < 1e-12)

expected_qtl_n <- c(8, 40, 80, 120, 240, 480)
expected_h2 <- c(0.7, 0.7, 0.5, 0.5, 0.3, 0.3)
scenario_metadata_ok <-
  nrow(control_table) == 6L &&
  identical(as.numeric(control_table$QTL_N), expected_qtl_n) &&
  isTRUE(all.equal(as.numeric(control_table$H2_nominal), expected_h2, tolerance = 1e-12))
qtl_lengths <- vapply(qtl_lists, length, integer(1))
qtl_lengths_ok <- identical(as.numeric(qtl_lengths), expected_qtl_n)
qtl_indices_ok <- all(unlist(qtl_lists) >= 1L & unlist(qtl_lists) <= 4010L)
qtl_unique_ok <- all(vapply(qtl_lists, function(x) !anyDuplicated(x), logical(1)))
qtl_distribution_ok <-
  all(as.matrix(qtl_by_lg[, paste0("LG", 9:10), drop = FALSE]) == 0) &&
  all(vapply(seq_len(nrow(qtl_by_lg)), function(i) {
    all(as.numeric(qtl_by_lg[i, paste0("LG", 1:8)]) == expected_qtl_n[i] / 8)
  }, logical(1)))

row_order_contract_ok <-
  nrow(phenotype) == nrow(true_genetic_value) &&
  nrow(phenotype) == nrow(geno_012)
phenotype_genetic_signal_ok <- all(scenario_summary$Cor_P_G > 0.5)
h2_ok <- !any(h2_difference > 0.03)
cv_ok <- !any(cv_difference > 1)

f2_frequency_ok <- max(abs(as.numeric(geno_frequency) - c(0.25, 0.50, 0.25))) < 0.02
maf_ok <- min(maf) > 0.40

parent_structure_ok <- nrow(parents) == 20L && ncol(parents) == 405L
parent1_ok <- all(parent_genotypes[parent_id == 1, ] == 1)
parent2_ok <- all(parent_genotypes[parent_id == 2, ] == 0)
lg_parent_ok <-
  identical(sort(unique(as.integer(parent_lg))), 1:10) &&
  all(table(parent_lg) == 2L) &&
  all(parent_nmarkers == 401)

add_compat_audit(
  "Simulation-control key formatting",
  compat_status(!scale_key_has_internal_whitespace, failure = "WARNING"),
  if (scale_key_has_internal_whitespace) {
    paste(
      "The source control file contains internal whitespace in the C1 Scale key (Escal<tab>a).",
      "Key whitespace was normalized for parsing; the stored numeric value was not altered."
    )
  } else {
    "No internal whitespace was detected in the simulation-control keys."
  }
)
add_compat_audit(
  "Nominal versus realized heritability",
  compat_status(h2_ok, failure = "WARNING"),
  paste(
    "Absolute differences:", paste(round(h2_difference, 4), collapse = ", "),
    ". Differences are documented rather than used to redefine the scenarios."
  )
)
add_compat_audit(
  "CV metadata versus empirical phenotype CV",
  compat_status(cv_ok, failure = "WARNING"),
  paste(
    "The simulator reports CV metadata = 12 for all scenarios; empirical phenotype CV values are reported separately.",
    "The metadata field is therefore not interpreted automatically as the empirical coefficient of variation."
  )
)
long_distance_detail <- if (long_distance_plateau) {
  paste(
    ". LD declines strongly up to 50 cM but remains elevated at 100-200 cM relative to the interchromosomal reference.",
    "This is documented as a property requiring interpretation of the simulator's long-distance recombination behavior, not as evidence of data corruption."
  )
} else {
  ". No marked long-distance LD plateau was detected."
}
add_compat_audit(
  "Long-distance LD behavior",
  compat_status(!long_distance_plateau, failure = "WARNING"),
  paste(
    "Mean r2 at 50, 100, and 200 cM:",
    paste(round(c(ld_50, ld_100, ld_200), 4), collapse = ", "),
    "; interchromosomal mean r2:", round(interchromosomal_ld$Mean_r2, 5),
    long_distance_detail
  )
)

add_compat_audit("Expected files", compat_status(all(file_table$Exists)), sprintf("All %d expected files are present.", length(expected_files)))
add_compat_audit("Core dimensions", compat_status(dim_ok), "Expected phenotype and genetic-value matrices: 1000 x 6; expected genotype matrices: 1000 x 4010.")
add_compat_audit("Finite phenotype and genetic values", compat_status(finite_ok), if (finite_ok) "All phenotypic and true genetic values are finite." else "Non-finite values were detected.")
add_compat_audit("Genotype codes 0/1/2", compat_status(code_012_ok), paste("Observed codes:", paste(codes_012, collapse = ", ")))
add_compat_audit("Genotype codes 11/12/22", compat_status(code_1122_ok), paste("Observed codes:", paste(codes_1122, collapse = ", ")))
add_compat_audit("Equivalence of genotype files", compat_status(equivalent_genotype_files), "Checked mapping 11 -> 2, 12 -> 1, and 22 -> 0 after orienting both matrices as individuals x markers.")
add_compat_audit("Canonical -1/0/1 coding", compat_status(canonical_coding_ok), "Canonical coding is genotype_012 - 1.")
add_compat_audit("Genetic map structure", compat_status(map_ok), "Expected 10 linkage groups, 401 markers per group, 200 cM per group, and 0.5 cM marker spacing.")
add_compat_audit("Unique map marker names", compat_status(n_duplicate_marker_names == 0L), sprintf("%d duplicated marker names detected.", n_duplicate_marker_names))
add_compat_audit("Simulation scenario metadata", compat_status(scenario_metadata_ok), "Six scenarios expected with QTL counts 8, 40, 80, 120, 240, and 480 and nominal heritabilities 0.7, 0.7, 0.5, 0.5, 0.3, and 0.3.")
add_compat_audit("QTL-list lengths", compat_status(qtl_lengths_ok), paste("Observed QTL-list lengths:", paste(qtl_lengths, collapse = ", ")))
add_compat_audit("QTL indices within map", compat_status(qtl_indices_ok), "All QTL indices must lie between 1 and 4010.")
add_compat_audit("Unique QTL within scenario", compat_status(qtl_unique_ok), "No duplicated QTL index is expected within a simulation scenario.")
add_compat_audit("QTL distribution across linkage groups", compat_status(qtl_distribution_ok), "The simulation design places QTL evenly across LG1-LG8; LG9 and LG10 contain no selected QTL.")
add_compat_audit("Row-order contract", compat_status(row_order_contract_ok), "Phenotype, true genetic-value, and genotype matrices contain the same 1000 rows. Because the simulation files do not provide explicit individual IDs, row position is adopted as the canonical individual identifier; independent identity verification is not available.")
add_compat_audit("Phenotype-genetic signal", compat_status(phenotype_genetic_signal_ok, failure = "WARNING"), paste("Phenotype-true genetic correlations:", paste(round(scenario_summary$Cor_P_G, 4), collapse = ", ")))
add_compat_audit("Missing genotypes", compat_status(missing_genotypes == 0L), sprintf("%d missing genotype cells.", missing_genotypes))
add_compat_audit("Duplicate individuals", compat_status(duplicate_individuals == 0L), sprintf("%d duplicate genotype profiles.", duplicate_individuals))
add_compat_audit("Duplicate markers", compat_status(duplicate_markers == 0L, failure = "WARNING"), sprintf("%d duplicate marker profiles.", duplicate_markers))
add_compat_audit("Monomorphic markers", compat_status(monomorphic_markers == 0L, failure = "WARNING"), sprintf("%d monomorphic markers.", monomorphic_markers))
add_compat_audit("Overall F2 genotype frequencies", compat_status(f2_frequency_ok, failure = "WARNING"), paste("Observed proportions:", paste(round(as.numeric(geno_frequency), 5), collapse = ", ")))
add_compat_audit("Minor allele frequency", compat_status(maf_ok, failure = "WARNING"), sprintf("MAF range: %.4f to %.4f; median %.4f.", min(maf), max(maf), median(maf)))
add_compat_audit("Parental file dimensions", compat_status(parent_structure_ok), "Expected 20 rows: two parents x 10 linkage groups, with 401 marker states per linkage group.")
add_compat_audit("Parental genotype contrast", compat_status(parent1_ok && parent2_ok), "Parent 1 is fixed for state 1 and parent 2 is fixed for state 0 across all markers.")
add_compat_audit("Parental linkage-group structure", compat_status(lg_parent_ok), "Each of the 10 linkage groups is represented for both parents.")
add_compat_audit("map.rds readability", "PASS", paste("Object successfully read; class:", paste(class(map_rds), collapse = ", ")))
add_compat_audit("map.rds row count", compat_status(!is.null(dim(map_rds)) && nrow(map_rds) == nrow(genetic_map)), paste("Observed rows:", nrow(map_rds), "; text-map markers:", nrow(genetic_map)))
add_compat_audit(
  "Short-to-medium distance LD decay",
  compat_status(ld_decay_ok, failure = "WARNING"),
  paste(
    "Mean r2 at 0.5, 10, and 50 cM:", paste(round(c(short_ld, medium_ld, ld_50), 4), collapse = ", "),
    "; interchromosomal mean r2:", round(interchromosomal_ld$Mean_r2, 5)
  )
)

audit_log$Status <- factor(audit_log$Status, levels = c("BLOCKED", "WARNING", "PASS"))
audit_log <- audit_log[order(audit_log$Status), , drop = FALSE]
rownames(audit_log) <- NULL
```

## 11. Audit summary

The individual audit checks are retained in the exported CSV artifact. The
tutorial summarizes their final counts and classification before creating the
shared canonical object.

```{r final-audit}
audit_overview <- data.frame(
  Quantity = c(
    "Input files present",
    "Individuals",
    "Markers",
    "Scenarios",
    "Genotype-code mismatches after recoding",
    "Missing genotype cells",
    "Duplicate individuals",
    "Duplicate markers",
    "Monomorphic markers",
    "Duplicate map marker names",
    "Minimum MAF",
    "Mean interchromosomal r2"
  ),
  Observed = c(
    paste(sum(file_table$Exists), "of", nrow(file_table)),
    nrow(geno_012),
    ncol(geno_012),
    ncol(phenotype),
    genotype_mismatches,
    missing_genotypes,
    duplicate_individuals,
    duplicate_markers,
    monomorphic_markers,
    n_duplicate_marker_names,
    round(min(maf), 4),
    round(interchromosomal_ld$Mean_r2, 5)
  ),
  stringsAsFactors = FALSE
)

format_table(
  audit_overview,
  caption = "Observed summary of the simulated-data audit",
  row.names = FALSE
)
```

The table above collects the principal observed quantities used to interpret the
simulated dataset. Features that require scientific interpretation are discussed
in the section where they arise rather than converted into execution guards.

## 12. Canonical preprocessed object

The canonical preprocessed object is generated from the data and summaries shown above. The saved `preprocessed_simulated_common.rds` acts as the
shared data contract for the downstream validation and modeling modules, so
those modules consume the same audited representation instead of independently
re-reading and re-interpreting the raw simulator files.

Its components have distinct scientific roles:

- `phenotype` contains the observed response values available for model fitting
  within the appropriate analysis subsets;
- `genotype_m101` is the canonical -1/0/1 genotype representation used by the
  downstream genomic prediction workflows, while `genotype_012` is retained as
  the validated source representation;
- `metadata$row_identifier` preserves the row-position identity contract used
  to align individuals across phenotypes, genotypes, validation splits, and
  predictions;
- `true_genetic_value` is retained for held-out predictive evaluation only. It
  is not used to generate validation folds, tune ANN hyperparameters, select
  training duration, or fit the predictive models;
- map, scenario, QTL, marker-QC, and LD components preserve the audited
  simulation structure and provenance for downstream checks and interpretation.
  Their presence in the common object does not make them additional response
  variables or automatic model predictors.

The next module uses this audited object to construct the frozen nested
cross-validation design. The GBLUP-ADE and ANN branches then consume that common
data contract together with the same frozen validation design, preserving
individual and task alignment across the parallel modeling workflows.

```{r save-canonical-object}
write.csv(
  audit_log,
  file.path(output_dir, "simulated_audit_summary.csv"),
  row.names = FALSE
)
write.csv(
  scenario_summary,
  file.path(output_dir, "simulated_scenario_summary.csv"),
  row.names = FALSE
)
write.csv(
  qtl_rows,
  file.path(output_dir, "simulated_qtl_map.csv"),
  row.names = FALSE
)
write.csv(
  ld_summary,
  file.path(output_dir, "simulated_ld_summary.csv"),
  row.names = FALSE
)

simulated_common <- list(
  metadata = list(
    dataset = "simulated",
    seed = SEED_GLOBAL,
    n_individuals = nrow(geno_012),
    n_markers = ncol(geno_012),
    n_scenarios = ncol(phenotype),
    row_identifier = seq_len(nrow(geno_012)),
    row_alignment = "row position",
    canonical_genotype_coding = "-1/0/1"
  ),
  phenotype = phenotype,
  true_genetic_value = true_genetic_value,
  residual_value = residual_value,
  genotype_012 = geno_012,
  genotype_m101 = geno_m101,
  genetic_map = genetic_map,
  scenario_control = control_table,
  qtl_lists = qtl_lists,
  qtl_map = qtl_rows,
  scenario_summary = scenario_summary,
  marker_qc = list(
    allele_frequency = allele_frequency,
    maf = maf,
    heterozygosity = heterozygosity
  ),
  ld_summary = ld_summary,
  interchromosomal_ld = interchromosomal_ld
)

saveRDS(
  simulated_common,
  file.path(output_dir, "preprocessed_simulated_common.rds"),
  compress = "gzip"
)

cat(
  "Canonical object written to:",
  file.path(output_dir, "preprocessed_simulated_common.rds"),
  "\n"
)
```

The resulting `preprocessed_simulated_common.rds` is the shared preprocessed
data contract passed to the validation module and, from there, to both parallel
modeling branches. The detailed CSV audit files remain available for
reproducibility and maintenance without occupying the main scientific narrative.

## 13. Navigation

[Home](index.html) | [Next: Validation design](simulated_data_validation.html)
