---
title: "GBLUP-ADE for simulated genomic data"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Purpose

This module fits the canonical GBLUP-ADE benchmark to all six simulated
scenarios using the 25 frozen outer validation splits. Phenotypes are used for
model fitting. True simulated genetic values are used only after prediction
for primary evaluation of genetic-value recovery.

The heavy fitting chunk is disabled during an ordinary website build. Run the
complete analysis from the project root with:

```text
RUN_GBLUP=1 Rscript -e 'workflowr::wflow_build("analysis/simulated_data_gblup.Rmd")'
```

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)

options(stringsAsFactors = FALSE)

SEED_GLOBAL <- 20260821L
N_ITERS_SOMMER <- 20L
RUN_GBLUP <- identical(Sys.getenv("RUN_GBLUP", "0"), "1")

DATA_FILE <- "output/preprocessed_simulated_common.rds"
CV_FILE <- "output/simulated_cv_design.rds"
DATA_AUDIT_FILE <- "output/simulated_audit_summary.csv"
CV_AUDIT_FILE <- "output/simulated_cv_design_audit.csv"

RESULTS_RDS <- "output/simulated_gblup_ade_results.rds"
PREDICTIONS_RDS <- "output/simulated_gblup_ade_predictions.rds"
RESULTS_CSV <- "output/simulated_gblup_ade_results.csv"
PREDICTIONS_CSV <- "output/simulated_gblup_ade_predictions.csv"
KERNEL_CSV <- "output/simulated_gblup_ade_kernel_summary.csv"
RUN_RDS <- "output/simulated_gblup_ade_run_metadata.rds"

safe_cor <- function(pred, obs) {
  ok <- is.finite(pred) & is.finite(obs)
  if (sum(ok) < 3L || sd(pred[ok]) == 0 || sd(obs[ok]) == 0) return(NA_real_)
  cor(pred[ok], obs[ok])
}

safe_rmse <- function(pred, obs) {
  ok <- is.finite(pred) & is.finite(obs)
  if (!any(ok)) return(NA_real_)
  sqrt(mean((pred[ok] - obs[ok])^2))
}

safe_mae <- function(pred, obs) {
  ok <- is.finite(pred) & is.finite(obs)
  if (!any(ok)) return(NA_real_)
  mean(abs(pred[ok] - obs[ok]))
}

safe_bias <- function(pred, obs) {
  ok <- is.finite(pred) & is.finite(obs)
  if (!any(ok)) return(NA_real_)
  mean(pred[ok] - obs[ok])
}

safe_slope <- function(pred, obs) {
  ok <- is.finite(pred) & is.finite(obs)
  if (sum(ok) < 3L || sd(pred[ok]) == 0) return(NA_real_)
  as.numeric(coef(lm(obs[ok] ~ pred[ok]))[2])
}

safe_r2 <- function(pred, obs) {
  ok <- is.finite(pred) & is.finite(obs)
  if (sum(ok) < 2L) return(NA_real_)
  den <- sum((obs[ok] - mean(obs[ok]))^2)
  if (den == 0) return(NA_real_)
  1 - sum((obs[ok] - pred[ok])^2) / den
}

metrics <- function(pred, obs, prefix) {
  z <- c(
    safe_cor(pred, obs),
    safe_rmse(pred, obs),
    safe_mae(pred, obs),
    safe_bias(pred, obs),
    safe_slope(pred, obs),
    safe_r2(pred, obs)
  )
  names(z) <- paste0(
    prefix,
    c("_Correlation", "_RMSE", "_MAE", "_Bias", "_Slope", "_R2")
  )
  z
}

extract_u <- function(fit, dat, factor_name, exact_name) {
  unames <- names(fit$uList)
  if (exact_name %in% unames) {
    nm <- exact_name
  } else {
    token <- paste0("ism(", factor_name, ")")
    hit <- grep(token, unames, fixed = TRUE)
    if (length(hit) != 1L) {
      stop(
        paste(
          "Could not identify random component for", factor_name,
          ". Available:", paste(unames, collapse = " | ")
        )
      )
    }
    nm <- unames[hit]
  }

  u <- fit$uList[[nm]]
  if (is.data.frame(u)) u <- as.matrix(u)
  if (is.vector(u)) u <- matrix(u, ncol = 1L)

  X <- model.matrix(
    as.formula(paste0("~", factor_name, " - 1")),
    data = dat
  )

  if (ncol(X) != nrow(u)) {
    stop(
      paste(
        "Random-effect dimension mismatch for", factor_name,
        ":", ncol(X), "vs", nrow(u)
      )
    )
  }

  as.numeric(X %*% u[, 1L, drop = FALSE])
}
```

# Packages and canonical inputs

```{r inputs}
if (!requireNamespace("sommer", quietly = TRUE)) {
  stop("Package 'sommer' is required.")
}

# sommer::mmes() evaluates the random-effects formula using helper
# functions vsm() and ism(). Attach sommer explicitly so these helpers
# are available in the fresh R session created by workflowr/callr.
suppressPackageStartupMessages(
  library(sommer)
)

if (
  !exists("vsm", mode = "function") ||
  !exists("ism", mode = "function")
) {
  stop(
    paste(
      "sommer was attached, but vsm() and/or ism() are unavailable.",
      "Installed sommer version:",
      as.character(packageVersion("sommer"))
    )
  )
}

required_files <- c(
  DATA_FILE,
  CV_FILE,
  DATA_AUDIT_FILE,
  CV_AUDIT_FILE
)

if (!all(file.exists(required_files))) {
  stop(
    paste(
      "Missing canonical file(s):",
      paste(required_files[!file.exists(required_files)], collapse = ", ")
    )
  )
}

data_audit <- read.csv(DATA_AUDIT_FILE, stringsAsFactors = FALSE)
cv_audit <- read.csv(CV_AUDIT_FILE, stringsAsFactors = FALSE)

if (any(data_audit$Status == "BLOCKED")) {
  stop("Simulated-data audit contains blocking checks.")
}

if (any(cv_audit$Status == "BLOCKED")) {
  stop("Validation-design audit contains blocking checks.")
}

common <- readRDS(DATA_FILE)
cv <- readRDS(CV_FILE)

phenotype <- common$phenotype
true_g <- common$true_genetic_value
geno <- common$genotype_m101
control <- common$scenario_control
ids <- as.integer(common$metadata$row_identifier)

if (is.null(ids) || length(ids) == 0L) {
  ids <- seq_len(nrow(geno))
}

id_labels <- sprintf("ID%04d", ids)
rownames(geno) <- id_labels

scenario_names <- colnames(phenotype)
if (is.null(scenario_names)) scenario_names <- paste0("C", 1:6)

input_summary <- data.frame(
  Item = c("Individuals", "Markers", "Scenarios", "Outer splits"),
  Value = c(nrow(geno), ncol(geno), ncol(phenotype), length(cv$outer_splits))
)

knitr::kable(input_summary)
```

# Pre-fit integrity

```{r prefit}
sommer_helpers <- data.frame(
  Helper = c("vsm", "ism"),
  Available = c(
    exists("vsm", mode = "function"),
    exists("ism", mode = "function")
  ),
  Sommer_version = as.character(packageVersion("sommer")),
  stringsAsFactors = FALSE
)

knitr::kable(
  sommer_helpers,
  caption = "sommer helper functions required by the random-effects formula"
)

checks <- data.frame(
  Check = c(
    "Phenotype dimensions",
    "True-G dimensions",
    "Genotype dimensions",
    "Genotype coding",
    "Outer split count",
    "Scenario names"
  ),
  PASS = c(
    identical(dim(phenotype), c(1000L, 6L)),
    identical(dim(true_g), c(1000L, 6L)),
    identical(dim(geno), c(1000L, 4010L)),
    identical(sort(unique(as.vector(geno))), -1:1),
    length(cv$outer_splits) == 25L,
    identical(scenario_names, paste0("C", 1:6))
  ),
  stringsAsFactors = FALSE
)

checks$Status <- ifelse(checks$PASS, "PASS", "BLOCKED")

knitr::kable(checks[, c("Check", "Status")])

if (any(!checks$PASS)) {
  stop("Blocking GBLUP-ADE pre-fit check detected.")
}

outer_ok <- all(
  vapply(
    cv$outer_splits,
    function(s) {
      length(s$analysis_ids) == 800L &&
        length(s$assessment_ids) == 200L &&
        length(intersect(s$analysis_ids, s$assessment_ids)) == 0L
    },
    logical(1)
  )
)

if (!outer_ok) stop("Frozen outer splits failed integrity check.")
```

# Analysis tasks

```{r task-table}
outer_meta <- unique(
  cv$outer_assignments[, c("Repetition", "OuterFold", "SplitID")]
)

outer_meta <- outer_meta[
  order(outer_meta$Repetition, outer_meta$OuterFold),
]

task_table <- merge(
  data.frame(Scenario = scenario_names, stringsAsFactors = FALSE),
  outer_meta,
  by = NULL
)

task_table <- task_table[
  order(
    match(task_table$Scenario, scenario_names),
    task_table$Repetition,
    task_table$OuterFold
  ),
]

task_table$TaskID <- paste(
  task_table$Scenario,
  task_table$SplitID,
  sep = "_"
)

rownames(task_table) <- NULL

cat("Expected GBLUP-ADE tasks:", nrow(task_table), "\n")
knitr::kable(head(task_table, 15))

if (nrow(task_table) != 150L) {
  stop("Expected 150 scenario-by-split tasks.")
}
```

# Fit GBLUP-ADE

The A, D, and E relationship matrices are built once from the complete
genotype matrix and reused across tasks. Assessment phenotypes are set to
`NA` before every fit. Genotypes of assessment individuals are allowed in the
relationship matrices; their phenotypes and true genetic values are not used
for fitting.

```{r fit-gblup-ade, eval=RUN_GBLUP}
set.seed(SEED_GLOBAL)

run_start <- proc.time()[["elapsed"]]

kernel_start <- proc.time()[["elapsed"]]

A_matrix <- sommer::A.mat(geno)
D_matrix <- sommer::D.mat(geno)
E_matrix <- sommer::E.mat(geno)

kernel_seconds <- proc.time()[["elapsed"]] - kernel_start

rownames(A_matrix) <- colnames(A_matrix) <- id_labels
rownames(D_matrix) <- colnames(D_matrix) <- id_labels
rownames(E_matrix) <- colnames(E_matrix) <- id_labels

kernel_summary <- data.frame(
  Kernel = c("Additive", "Dominance", "Epistatic"),
  Rows = c(nrow(A_matrix), nrow(D_matrix), nrow(E_matrix)),
  Columns = c(ncol(A_matrix), ncol(D_matrix), ncol(E_matrix)),
  Symmetric = c(
    isTRUE(all.equal(A_matrix, t(A_matrix), tolerance = 1e-8)),
    isTRUE(all.equal(D_matrix, t(D_matrix), tolerance = 1e-8)),
    isTRUE(all.equal(E_matrix, t(E_matrix), tolerance = 1e-8))
  ),
  Finite = c(
    all(is.finite(A_matrix)),
    all(is.finite(D_matrix)),
    all(is.finite(E_matrix))
  ),
  Mean_diagonal = c(
    mean(diag(A_matrix)),
    mean(diag(D_matrix)),
    mean(diag(E_matrix))
  ),
  Kernel_build_seconds = rep(kernel_seconds, 3L)
)

if (any(!kernel_summary$Symmetric) || any(!kernel_summary$Finite)) {
  stop("A, D, or E kernel failed integrity checks.")
}

write.csv(kernel_summary, KERNEL_CSV, row.names = FALSE)

result_list <- vector("list", nrow(task_table))
prediction_list <- vector("list", nrow(task_table))

for (i in seq_len(nrow(task_table))) {
  task <- task_table[i, ]
  scenario <- task$Scenario
  split_id <- task$SplitID
  j <- match(scenario, scenario_names)

  split <- cv$outer_splits[[split_id]]

  tr_idx <- match(split$analysis_ids, ids)
  te_idx <- match(split$assessment_ids, ids)

  if (anyNA(tr_idx) || anyNA(te_idx)) {
    stop(paste("ID alignment failed for", task$TaskID))
  }

  y <- phenotype[, j]
  g <- true_g[, j]

  id_factor <- factor(id_labels, levels = id_labels)

  dat <- data.frame(
    y = y,
    id = id_factor,
    idd = id_factor,
    ide = id_factor
  )

  dat$y[te_idx] <- NA_real_

  fit_status <- "ok"
  pred_status <- "not_run"
  error_message <- NA_character_

  y_pred <- gebv <- u_add <- u_dom <- u_epi <-
    rep(NA_real_, nrow(dat))

  fit_start <- proc.time()[["elapsed"]]

  fit <- tryCatch(
    sommer::mmes(
      y ~ 1,
      random = ~
        vsm(ism(id), Gu = A_matrix) +
        vsm(ism(idd), Gu = D_matrix) +
        vsm(ism(ide), Gu = E_matrix),
      rcov = ~ units,
      nIters = N_ITERS_SOMMER,
      data = dat,
      verbose = FALSE,
      dateWarning = FALSE
    ),
    error = function(e) {
      fit_status <<- "fit_error"
      error_message <<- conditionMessage(e)
      NULL
    }
  )

  fit_seconds <- proc.time()[["elapsed"]] - fit_start
  predict_seconds <- 0

  if (!is.null(fit)) {
    pred_status <- "ok"
    predict_start <- proc.time()[["elapsed"]]

    pc <- tryCatch(
      {
        a <- extract_u(
          fit, dat, "id",
          "vsm(ism(id), Gu = A_matrix)"
        )
        d <- extract_u(
          fit, dat, "idd",
          "vsm(ism(idd), Gu = D_matrix)"
        )
        e <- extract_u(
          fit, dat, "ide",
          "vsm(ism(ide), Gu = E_matrix)"
        )

        gv <- a + d + e
        beta0 <- as.numeric(fit$b)[1L]

        list(
          additive = a,
          dominance = d,
          epistatic = e,
          gebv = gv,
          prediction = beta0 + gv
        )
      },
      error = function(e) {
        pred_status <<- "pred_error"
        error_message <<- conditionMessage(e)
        NULL
      }
    )

    predict_seconds <- proc.time()[["elapsed"]] - predict_start

    if (!is.null(pc)) {
      u_add <- pc$additive
      u_dom <- pc$dominance
      u_epi <- pc$epistatic
      gebv <- pc$gebv
      y_pred <- pc$prediction
    }
  }

  obs_test <- y[te_idx]
  g_test <- g[te_idx]
  pred_test <- y_pred[te_idx]
  gebv_test <- gebv[te_idx]

  mg <- metrics(pred_test, g_test, "Genetic")
  mp <- metrics(pred_test, obs_test, "Phenotype")

  n_finite <- sum(is.finite(pred_test))

  status <- if (
    fit_status == "ok" &&
      pred_status == "ok" &&
      n_finite == 200L
  ) {
    "ok"
  } else {
    paste(fit_status, pred_status, sep = ";")
  }

  meta <- control[control$Scenario == scenario, , drop = FALSE]

  result_list[[i]] <- data.frame(
    Dataset = "simulated",
    Scenario = scenario,
    QTL_N = meta$QTL_N,
    H2_nominal = meta$H2_nominal,
    Repetition = task$Repetition,
    Fold = task$OuterFold,
    SplitID = split_id,
    TaskID = task$TaskID,
    Model = "GBLUP_ADE",
    N_train = length(tr_idx),
    N_test = length(te_idx),
    N_pred_finite = n_finite,
    Genetic_Correlation = mg["Genetic_Correlation"],
    Genetic_RMSE = mg["Genetic_RMSE"],
    Genetic_MAE = mg["Genetic_MAE"],
    Genetic_Bias = mg["Genetic_Bias"],
    Genetic_Slope = mg["Genetic_Slope"],
    Genetic_R2 = mg["Genetic_R2"],
    GEBV_TrueG_Correlation = safe_cor(gebv_test, g_test),
    Phenotype_Correlation = mp["Phenotype_Correlation"],
    Phenotype_RMSE = mp["Phenotype_RMSE"],
    Phenotype_MAE = mp["Phenotype_MAE"],
    Phenotype_Bias = mp["Phenotype_Bias"],
    Phenotype_Slope = mp["Phenotype_Slope"],
    Phenotype_R2 = mp["Phenotype_R2"],
    Kernel_build_seconds = kernel_seconds,
    Fit_seconds = fit_seconds,
    Predict_seconds = predict_seconds,
    Status = status,
    Error = error_message,
    stringsAsFactors = FALSE
  )

  prediction_list[[i]] <- data.frame(
    Dataset = "simulated",
    Scenario = scenario,
    QTL_N = meta$QTL_N,
    H2_nominal = meta$H2_nominal,
    Repetition = task$Repetition,
    Fold = task$OuterFold,
    SplitID = split_id,
    TaskID = task$TaskID,
    Model = "GBLUP_ADE",
    ID = ids[te_idx],
    Observed_Phenotype = obs_test,
    True_Genetic_Value = g_test,
    Predicted_Value = pred_test,
    GEBV = gebv_test,
    Additive_Prediction = u_add[te_idx],
    Dominance_Prediction = u_dom[te_idx],
    Epistatic_Prediction = u_epi[te_idx],
    Status = status,
    stringsAsFactors = FALSE
  )

  cat(
    sprintf(
      "[%03d/%03d] %s | %s | rG=%.4f | RMSE_G=%.4f | fit=%.2fs\n",
      i,
      nrow(task_table),
      task$TaskID,
      status,
      result_list[[i]]$Genetic_Correlation,
      result_list[[i]]$Genetic_RMSE,
      fit_seconds
    )
  )
}

results <- do.call(rbind, result_list)
predictions <- do.call(rbind, prediction_list)

rownames(results) <- NULL
rownames(predictions) <- NULL

run_seconds <- proc.time()[["elapsed"]] - run_start

run_metadata <- list(
  dataset = "simulated",
  model = "GBLUP_ADE",
  seed_global = SEED_GLOBAL,
  sommer_version = as.character(packageVersion("sommer")),
  nIters_sommer = N_ITERS_SOMMER,
  tasks = nrow(results),
  prediction_rows = nrow(predictions),
  kernel_build_seconds = kernel_seconds,
  total_run_seconds = run_seconds,
  timestamp = as.character(Sys.time())
)

saveRDS(results, RESULTS_RDS, compress = "gzip")
saveRDS(predictions, PREDICTIONS_RDS, compress = "gzip")
write.csv(results, RESULTS_CSV, row.names = FALSE)
write.csv(predictions, PREDICTIONS_CSV, row.names = FALSE)
saveRDS(run_metadata, RUN_RDS, compress = "gzip")

cat(
  "\nGBLUP-ADE completed:",
  nrow(results), "tasks;",
  nrow(predictions), "held-out predictions;",
  round(run_seconds, 2), "seconds.\n"
)
```

# Kernel diagnostics

```{r kernel-diagnostics}
if (file.exists(KERNEL_CSV)) {
  kernel_summary <- read.csv(KERNEL_CSV)
  knitr::kable(
    kernel_summary,
    digits = 5,
    caption = "A, D, and E kernel diagnostics"
  )
} else {
  cat("Kernel diagnostics not available yet. Run with RUN_GBLUP=1.\n")
}
```

# Saved-result audit

```{r result-audit}
results_available <-
  file.exists(RESULTS_RDS) &&
  file.exists(PREDICTIONS_RDS) &&
  file.exists(RUN_RDS)

if (!results_available) {
  cat(
    "Fitted results are not available yet.\n",
    "Run:\n",
    "RUN_GBLUP=1 Rscript -e ",
    "'workflowr::wflow_build(\"analysis/simulated_data_gblup.Rmd\")'\n",
    sep = ""
  )
} else {
  results <- readRDS(RESULTS_RDS)
  predictions <- readRDS(PREDICTIONS_RDS)
  run_metadata <- readRDS(RUN_RDS)

  result_audit <- data.frame(
    Check = c(
      "Task rows",
      "Prediction rows",
      "Unique tasks",
      "Predictions per task",
      "Successful tasks",
      "Finite predictions",
      "Frozen split IDs"
    ),
    Status = c(
      if (nrow(results) == 150L) "PASS" else "BLOCKED",
      if (nrow(predictions) == 30000L) "PASS" else "BLOCKED",
      if (length(unique(results$TaskID)) == 150L) "PASS" else "BLOCKED",
      if (all(table(predictions$TaskID) == 200L)) "PASS" else "BLOCKED",
      if (all(results$Status == "ok")) "PASS" else "WARNING",
      if (all(is.finite(predictions$Predicted_Value))) "PASS" else "WARNING",
      if (
        setequal(unique(results$SplitID), names(cv$outer_splits))
      ) "PASS" else "BLOCKED"
    ),
    stringsAsFactors = FALSE
  )

  knitr::kable(result_audit)

  if (any(result_audit$Status == "BLOCKED")) {
    stop("Saved GBLUP-ADE results failed a blocking audit check.")
  }
}
```

# Predictive performance

The primary metrics compare the held-out final prediction
(intercept + additive + dominance + epistatic components) with the known true
genetic value. Phenotype metrics are secondary.

```{r performance}
if (results_available) {
  ok_results <- results[results$Status == "ok", , drop = FALSE]

  if (nrow(ok_results) == 0L) {
    cat(
      "No successful GBLUP-ADE tasks are available for performance aggregation.\n",
      "Inspect the saved-result audit and Error column before rerunning.\n"
    )
  } else {

  scenario_performance <- aggregate(
    cbind(
      Genetic_Correlation,
      Genetic_RMSE,
      Genetic_MAE,
      Genetic_Bias,
      Genetic_Slope,
      Genetic_R2,
      Phenotype_Correlation,
      Phenotype_RMSE,
      Phenotype_MAE,
      Phenotype_Bias,
      Phenotype_Slope,
      Phenotype_R2
    ) ~ Scenario + QTL_N + H2_nominal,
    data = ok_results,
    FUN = mean,
    na.rm = TRUE
  )

  scenario_performance <- scenario_performance[
    order(match(scenario_performance$Scenario, scenario_names)),
  ]

  knitr::kable(
    scenario_performance,
    digits = 4,
    caption = "Mean predictive performance across the 25 outer splits"
  )
  }
}
```

# Split-level variability

```{r split-summary}
if (results_available && exists("ok_results") && nrow(ok_results) > 0L) {
  split_summary <- do.call(
    rbind,
    lapply(
      split(ok_results, ok_results$Scenario),
      function(x) {
        data.frame(
          Scenario = unique(x$Scenario),
          N = nrow(x),
          Cor_mean = mean(x$Genetic_Correlation, na.rm = TRUE),
          Cor_sd = sd(x$Genetic_Correlation, na.rm = TRUE),
          Cor_min = min(x$Genetic_Correlation, na.rm = TRUE),
          Cor_max = max(x$Genetic_Correlation, na.rm = TRUE),
          RMSE_mean = mean(x$Genetic_RMSE, na.rm = TRUE),
          RMSE_sd = sd(x$Genetic_RMSE, na.rm = TRUE),
          MAE_mean = mean(x$Genetic_MAE, na.rm = TRUE),
          R2_mean = mean(x$Genetic_R2, na.rm = TRUE)
        )
      }
    )
  )

  rownames(split_summary) <- NULL
  split_summary <- split_summary[
    order(match(split_summary$Scenario, scenario_names)),
  ]

  knitr::kable(
    split_summary,
    digits = 4,
    caption = "Genetic-value prediction variability across outer splits"
  )
}
```

# Computational time

```{r time-summary}
if (results_available) {
  time_summary <- aggregate(
    cbind(Fit_seconds, Predict_seconds) ~ Scenario,
    data = results,
    FUN = mean,
    na.rm = TRUE
  )

  time_summary$Kernel_build_seconds <- run_metadata$kernel_build_seconds

  knitr::kable(
    time_summary,
    digits = 3,
    caption = "GBLUP-ADE computational time"
  )

  cat(
    "Total run time:",
    round(run_metadata$total_run_seconds, 2),
    "seconds.\n"
  )
}
```

# Output contract

```{r output-contract}
output_files <- c(
  RESULTS_RDS,
  PREDICTIONS_RDS,
  RESULTS_CSV,
  PREDICTIONS_CSV,
  KERNEL_CSV,
  RUN_RDS
)

output_table <- data.frame(
  File = output_files,
  Exists = file.exists(output_files),
  stringsAsFactors = FALSE
)

knitr::kable(output_table)
```

# Reproducibility information

```{r session-info}
sessionInfo()
```
