---
title: "Regularized artificial neural networks for simulated genomic data"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

# Purpose

This module is the **canonical ANN workflow** for the simulated genomic-data
branch. ANN hyperparameter tuning and final outer-CV fitting are internal
stages of one modeling method, parallel to GBLUP in the public pipeline.

The four variants are `ANN_0`, `ANN_L1`, `ANN_L2`, and `ANN_L1_L2`
(displayed as **ANN-L1+L2**). They use the same multilayer perceptron and differ
only in L1/L2 weight penalties.

# Tuning contract

The four ANN variants use the same multilayer perceptron. They differ only in
the L1 and L2 coefficients applied to the network weights.

## Tuned hyperparameters

| Hyperparameter | Role |
|---|---|
| Architecture | number of hidden units/layers |
| Learning rate | Adam step size |
| L1 coefficient | strength of absolute-weight penalty |
| L2 coefficient | strength of squared-weight penalty |

Architecture and learning rate are selected once per outer task with `ANN_0`
and then shared by all four ANN variants. The regularized variants subsequently
select only their penalty coefficient(s).

## Fixed optimization controls

| Control | Canonical value |
|---|---:|
| Activation | ReLU |
| Optimizer | Adam |
| Dropout | 0 |
| Maximum tuning epochs | 2000 |
| Patience | 25 |
| Minimum improvement (`min_delta`) | \(10^{-6}\) |
| Early-validation fraction | 0.15 |

These values control **how** a candidate is optimized; they are not additional
hyperparameters in the scientific comparison.

## Early stopping

An epoch is one complete pass through the current fitting observations.
Early stopping monitors loss on a separate early-validation subset drawn only
from the current analysis sample. Training stops after 25 consecutive epochs
without an improvement of at least \(10^{-6}\).

The 2000-epoch value is therefore a convergence ceiling, not a fixed training
length. Epoch diagnostics are excluded from candidate ranking.

# Architecture range

The canonical architecture catalog defines a bounded small-to-moderate
feed-forward network search space. The candidate hidden-layer structures are
`2`, `4`, `8`, `16`, `32`, `8-4`, `16-8`, `32-16`, and `64-32`.

The `64-32` architecture is the prespecified upper bound of the canonical
search space. Consequently, all ANN results and the subsequent comparison with
GBLUP-ADE are interpreted **within the prespecified canonical model/architecture
search space**, rather than as a comparison against every possible neural-network
architecture.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
options(stringsAsFactors = FALSE)

WIDE_TABLE_ATTR <- paste(
  'style="display:block;',
  'overflow-x:auto;',
  'white-space:nowrap;',
  'width:100%;"'
)

find_project_root <- function(start = getwd()) {
  path <- normalizePath(start, winslash = "/", mustWork = TRUE)

  repeat {
    if (file.exists(file.path(path, "_workflowr.yml"))) {
      return(path)
    }

    parent <- dirname(path)

    if (identical(parent, path)) {
      stop("Could not locate workflowr project root.")
    }

    path <- parent
  }
}

PROJECT_ROOT <- find_project_root()

project_path <- function(...) {
  file.path(PROJECT_ROOT, ...)
}


SEED_GLOBAL <- 20260821L

RUN_ANN_TUNING <- identical(
  Sys.getenv("RUN_ANN_TUNING", "0"),
  "1"
)
RUN_ANN <- identical(
  Sys.getenv("RUN_ANN", "0"),
  "1"
)
ANN_RUNTIME_REQUIRED <- RUN_ANN_TUNING || RUN_ANN

ANN_TUNING_FORCE <- identical(
  Sys.getenv("ANN_TUNING_FORCE", "0"),
  "1"
)
ANN_TUNING_PILOT <- identical(
  Sys.getenv("ANN_TUNING_PILOT", "0"),
  "1"
)
ANN_TUNING_PILOT_TASK <- Sys.getenv(
  "ANN_TUNING_PILOT_TASK",
  "C1_R01_F01"
)

MAX_EPOCHS <- 2000L
# Compatibility token only: preserves the exact config_id of the completed
# ANN_TUNING_CANONICAL_V1 run. It does not control final refit length.
TUNING_V1_COMPATIBILITY_TOKEN <- 2000L
PATIENCE <- 25L
MIN_DELTA <- 1e-6
EARLY_VALIDATION_FRACTION <- 0.15
DROPOUT <- 0

ARCHITECTURES <- list(
  `2` = c(2L),
  `4` = c(4L),
  `8` = c(8L),
  `16` = c(16L),
  `32` = c(32L),
  `8-4` = c(8L, 4L),
  `16-8` = c(16L, 8L),
  `32-16` = c(32L, 16L),
  `64-32` = c(64L, 32L)
)

LEARNING_RATES <- c(
  1.25e-4,
  2.5e-4,
  5e-4,
  1e-3
)

L1_VALUES <- c(
  1e-5,
  1e-4,
  1e-3,
  1e-2,
  3e-2,
  1e-1
)

L2_VALUES <- c(
  1e-5,
  1e-4,
  1e-3,
  1e-2
)

# Full Cartesian grid for the simultaneous L1+L2 model.
# Both coefficients are strictly positive in ANN_L1_L2.
L1_L2_GRID <- expand.grid(
  L1 = L1_VALUES,
  L2 = L2_VALUES,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

MODEL_LEVELS <- c(
  "ANN_0",
  "ANN_L1",
  "ANN_L2",
  "ANN_L1_L2"
)

DATA_FILE <- project_path("output", "simulated", "audit", "preprocessed_simulated_common.rds")
CV_FILE <- project_path("output", "simulated", "validation", "simulated_cv_design.rds")
DATA_AUDIT_FILE <- project_path("output", "simulated", "audit", "simulated_audit_summary.csv")
CV_AUDIT_FILE <- project_path("output", "simulated", "validation", "simulated_cv_design_audit.csv")

CHECKPOINT_DIR <- project_path(".local", "checkpoints", "canonical", "ann_tuning")

OUTPUT_STEM <- if (ANN_TUNING_PILOT) {
  project_path(".local", "pilot", "ann", "tuning", "simulated_ann_tuning_pilot")
} else {
  project_path("output", "simulated", "ann", "tuning", "simulated_ann_tuning")
}

CANDIDATE_RDS <- paste0(OUTPUT_STEM, "_candidate_summary.rds")
FOLD_RDS <- paste0(OUTPUT_STEM, "_fold_results.rds")
SELECTED_RDS <- paste0(OUTPUT_STEM, "_selected_configurations.rds")
RUN_RDS <- paste0(OUTPUT_STEM, "_run_metadata.rds")
MANIFEST_RDS <- paste0(OUTPUT_STEM, "_manifest.rds")

CANDIDATE_CSV <- paste0(OUTPUT_STEM, "_candidate_summary.csv")
FOLD_CSV <- paste0(OUTPUT_STEM, "_fold_results.csv")
SELECTED_CSV <- paste0(OUTPUT_STEM, "_selected_configurations.csv")
MANIFEST_CSV <- paste0(OUTPUT_STEM, "_manifest.csv")

ANN_TUNING_CONFIG_ID <- paste(
  "ANN_TUNING_CANONICAL_V1",
  paste(names(ARCHITECTURES), collapse = "-"),
  paste(LEARNING_RATES, collapse = "-"),
  paste(L1_VALUES, collapse = "-"),
  paste(L2_VALUES, collapse = "-"),
  MAX_EPOCHS,
  PATIENCE,
  MIN_DELTA,
  EARLY_VALIDATION_FRACTION,
  TUNING_V1_COMPATIBILITY_TOKEN,
  sep = "__"
)

if (RUN_ANN_TUNING) {
  dir.create(
    CHECKPOINT_DIR,
    recursive = TRUE,
    showWarnings = FALSE
  )
  dir.create(
    dirname(CANDIDATE_RDS),
    recursive = TRUE,
    showWarnings = FALSE
  )
}
```

```{r packages, include=FALSE}
if (ANN_RUNTIME_REQUIRED) {
  if (!requireNamespace("torch", quietly = TRUE)) {
    stop(
      paste(
        "Package 'torch' is required when",
        "RUN_ANN_TUNING=1 or RUN_ANN=1."
      )
    )
  }

  suppressPackageStartupMessages(
    library(torch)
  )

  backend_ok <- tryCatch({
    z <- torch::torch_tensor(
      c(1, 2),
      dtype = torch::torch_float()
    )
    all(as.numeric(z) == c(1, 2))
  }, error = function(e) FALSE)

  if (!backend_ok) {
    stop(
      "torch is installed but the libtorch backend is unavailable."
    )
  }

  cuda_available <- isTRUE(
    tryCatch(
      torch::cuda_is_available(),
      error = function(e) FALSE
    )
  )

  requested_device <- tolower(
    Sys.getenv("ANN_DEVICE", "auto")
  )

  if (!requested_device %in% c("auto", "cpu", "cuda")) {
    stop("ANN_DEVICE must be auto, cpu, or cuda.")
  }

  if (requested_device == "cuda" && !cuda_available) {
    stop(
      "ANN_DEVICE=cuda requested but CUDA is unavailable."
    )
  }

  ANN_DEVICE_LABEL <- if (
    requested_device == "cuda" ||
    (requested_device == "auto" && cuda_available)
  ) {
    "cuda"
  } else {
    "cpu"
  }

  ANN_DEVICE <- torch::torch_device(
    ANN_DEVICE_LABEL
  )

  sync_device <- function() {
    if (ANN_DEVICE_LABEL == "cuda") {
      try(
        torch::cuda_synchronize(),
        silent = TRUE
      )
    }
    invisible(NULL)
  }

  set_ann_seed <- function(seed) {
    set.seed(as.integer(seed))
    torch::torch_manual_seed(
      as.integer(seed)
    )
    invisible(NULL)
  }

  torch_version_display <-
    as.character(packageVersion("torch"))
  backend_display <- backend_ok
  cuda_display <- cuda_available
  device_display <- ANN_DEVICE_LABEL
} else {
  backend_ok <- NA
  cuda_available <- NA
  ANN_DEVICE_LABEL <-
    "not initialized (saved-output render)"
  ANN_DEVICE <- NULL

  sync_device <- function() {
    invisible(NULL)
  }

  set_ann_seed <- function(seed) {
    stop(
      paste(
        "ANN runtime is not initialized.",
        "Set RUN_ANN_TUNING=1 or RUN_ANN=1."
      )
    )
  }

  torch_version_display <-
    "not initialized (saved-output render)"
  backend_display <-
    "not initialized (saved-output render)"
  cuda_display <-
    "not initialized (saved-output render)"
  device_display <- ANN_DEVICE_LABEL
}

environment_table <- data.frame(
  Item = c(
    "torch version",
    "backend available",
    "CUDA available",
    "device",
    "MAX_EPOCHS",
    "TUNING_V1_COMPATIBILITY_TOKEN",
    "patience",
    "minimum improvement",
    "early-validation fraction",
    "dropout"
  ),
  Value = c(
    torch_version_display,
    backend_display,
    cuda_display,
    device_display,
    MAX_EPOCHS,
    TUNING_V1_COMPATIBILITY_TOKEN,
    PATIENCE,
    MIN_DELTA,
    EARLY_VALIDATION_FRACTION,
    DROPOUT
  )
)

knitr::kable(
  environment_table,
  caption = "Fixed ANN computational protocol"
)
```

# Canonical ANN parameter values

The exact search space and optimization controls used by the canonical ANN
workflow are defined explicitly below. These objects are referenced by the
candidate-grid, training, tuning, and final-fitting chunks that follow.

```{r ann-parameter-values}
MAX_EPOCHS <- 2000L
PATIENCE <- 25L
MIN_DELTA <- 1e-6
EARLY_VALIDATION_FRACTION <- 0.15
DROPOUT <- 0

ARCHITECTURES <- list(
  `2` = c(2L),
  `4` = c(4L),
  `8` = c(8L),
  `16` = c(16L),
  `32` = c(32L),
  `8-4` = c(8L, 4L),
  `16-8` = c(16L, 8L),
  `32-16` = c(32L, 16L),
  `64-32` = c(64L, 32L)
)

LEARNING_RATES <- c(
  1.25e-4,
  2.5e-4,
  5e-4,
  1e-3
)

L1_VALUES <- c(
  1e-5,
  1e-4,
  1e-3,
  1e-2,
  3e-2,
  1e-1
)

L2_VALUES <- c(
  1e-5,
  1e-4,
  1e-3,
  1e-2
)

L1_L2_GRID <- expand.grid(
  L1 = L1_VALUES,
  L2 = L2_VALUES,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

MODEL_LEVELS <- c(
  "ANN_0",
  "ANN_L1",
  "ANN_L2",
  "ANN_L1_L2"
)
```

There are nine candidate architectures and four learning rates for the shared
structure search. `ANN_L1` evaluates six L1 values, `ANN_L2` evaluates four L2
values, and `ANN_L1_L2` evaluates the 24 positive L1 × L2 combinations.

# Canonical inputs

This module uses the audited common preprocessing object and the frozen nested
validation design established in the preceding modules. These fixed inputs keep
model tuning and outer-CV evaluation aligned across all four ANN variants.
Their dimensions, genotype coding, scenario names, and frozen split counts are
verified internally before any ANN tuning summary is interpreted.

```{r inputs, include=FALSE}
required_files <- c(
  DATA_FILE,
  CV_FILE,
  DATA_AUDIT_FILE,
  CV_AUDIT_FILE
)

if (!all(file.exists(required_files))) {
  stop(
    "Missing required file(s): ",
    paste(
      required_files[!file.exists(required_files)],
      collapse = ", "
    )
  )
}

data_audit <- read.csv(
  DATA_AUDIT_FILE,
  stringsAsFactors = FALSE
)
cv_audit <- read.csv(
  CV_AUDIT_FILE,
  stringsAsFactors = FALSE
)

if (any(data_audit$Status == "BLOCKED")) {
  stop(
    "Simulated-data audit contains BLOCKED checks."
  )
}
if (any(cv_audit$Status == "BLOCKED")) {
  stop(
    "Validation-design audit contains BLOCKED checks."
  )
}

common <- readRDS(DATA_FILE)
cv <- readRDS(CV_FILE)

phenotype <- common$phenotype
geno <- common$genotype_m101

ids <- as.integer(
  common$metadata$row_identifier
)
if (is.null(ids) || !length(ids)) {
  ids <- seq_len(nrow(geno))
}

scenario_names <- colnames(phenotype)
if (is.null(scenario_names)) {
  scenario_names <- paste0("C", 1:6)
}

input_checks <- data.frame(
  Check = c(
    "Phenotype is 1000 x 6",
    "Genotype is 1000 x 4010",
    "Genotype coding is -1/0/1",
    "25 frozen outer splits",
    "100 frozen inner splits",
    "Scenario names are C1-C6"
  ),
  PASS = c(
    identical(
      dim(phenotype),
      c(1000L, 6L)
    ),
    identical(
      dim(geno),
      c(1000L, 4010L)
    ),
    identical(
      sort(unique(as.vector(geno))),
      -1:1
    ),
    length(cv$outer_splits) == 25L,
    sum(vapply(
      cv$inner_splits,
      length,
      integer(1)
    )) == 100L,
    identical(
      scenario_names,
      paste0("C", 1:6)
    )
  )
)

input_checks$Status <- ifelse(
  input_checks$PASS,
  "PASS",
  "BLOCKED"
)

knitr::kable(
  input_checks[, c("Check", "Status")],
  caption = "Canonical ANN tuning pre-fit audit"
)

if (any(!input_checks$PASS)) {
  stop(
    "Blocking ANN tuning pre-fit check detected."
  )
}
```

# Candidate grids

Architecture and learning rate are selected **once per outer task using
`ANN_0`**. The selected structure is then shared by all four ANN variants. This
keeps the comparison between regularization variants focused on L1/L2
penalties rather than allowing each model family to change network capacity.

After structure selection:

- `ANN_L1` tunes only L1;
- `ANN_L2` tunes only L2;
- `ANN_L1_L2` tunes the full positive L1 × L2 Cartesian grid.

```{r grids}
structure_grid <- expand.grid(
  Architecture = names(ARCHITECTURES),
  Learning_rate = LEARNING_RATES,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)
structure_grid$Model <- "ANN_0"
structure_grid$L1 <- 0
structure_grid$L2 <- 0

l1_grid <- data.frame(
  Model = "ANN_L1",
  L1 = L1_VALUES,
  L2 = 0,
  stringsAsFactors = FALSE
)

l2_grid <- data.frame(
  Model = "ANN_L2",
  L1 = 0,
  L2 = L2_VALUES,
  stringsAsFactors = FALSE
)

l1l2_grid <- data.frame(
  Model = "ANN_L1_L2",
  L1 = L1_L2_GRID$L1,
  L2 = L1_L2_GRID$L2,
  stringsAsFactors = FALSE
)

grid_summary <- data.frame(
  Stage = c(
    "Shared architecture + learning rate",
    "ANN_L1 penalty",
    "ANN_L2 penalty",
    "ANN_L1_L2 penalties"
  ),
  Candidates = c(
    nrow(structure_grid),
    nrow(l1_grid),
    nrow(l2_grid),
    nrow(l1l2_grid)
  )
)

grid_summary$Inner_fits_per_outer_task <-
  4L * grid_summary$Candidates

knitr::kable(
  grid_summary,
  caption = "Canonical ANN tuning grid"
)

CANDIDATES_PER_TASK <- sum(
  grid_summary$Candidates
)
INNER_FITS_PER_TASK <- 4L *
  CANDIDATES_PER_TASK

cat(
  "Candidates per outer task:",
  CANDIDATES_PER_TASK,
  "\n"
)
cat(
  "Inner candidate-fold fits per outer task:",
  INNER_FITS_PER_TASK,
  "\n"
)
cat(
  "Planned inner fits for all 150 outer tasks:",
  150L * INNER_FITS_PER_TASK,
  "\n"
)
```

The grid contains 70 candidate configurations per outer task. With four frozen inner folds, this corresponds to 280 candidate-fold fits per task and 42,000 fits across the 150 scenario-by-split tasks. Architecture and learning rate are selected using `ANN_0` and then held common across the four ANN variants within each task, so their subsequent contrast isolates the L1/L2 penalty choice.

# Within-fold preprocessing

For every inner fit, marker means and standard deviations are estimated only
from the fitting subset. Those same parameters are then applied to the
corresponding early-validation and inner-assessment individuals. Phenotypes are
standardized from the fitting subset in the same way, and predictions are
returned to the original phenotype scale before calculating performance
metrics. Markers with effectively zero fitting-set variance receive unit scale
to avoid numerical division by zero.

This ordering prevents validation or assessment observations from influencing
the preprocessing parameters used to fit the network.

```{r support-functions}
correlation <- function(predicted, observed) {
  cor(predicted, observed)
}

rmse <- function(predicted, observed) {
  sqrt(
    mean(
      (predicted - observed)^2
    )
  )
}

mae <- function(predicted, observed) {
  mean(
    abs(
      predicted - observed
    )
  )
}

# Historical names retained for compatibility with result-building code.
safe_cor <- correlation
safe_rmse <- rmse
safe_mae <- mae

parameter_count <- function(
  input_dim,
  hidden_units
) {
  inputs <- c(input_dim, hidden_units)
  outputs <- c(hidden_units, 1L)

  sum(
    inputs * outputs +
      outputs
  )
}

make_validation_split <- function(
  x,
  frac,
  seed
) {
  set.seed(seed)

  n_validation <- as.integer(
    round(
      length(x) * frac
    )
  )

  validation <- sort(
    sample(
      x,
      n_validation
    )
  )

  list(
    train = sort(
      setdiff(
        x,
        validation
      )
    ),
    validation = validation
  )
}

fit_x_scaler <- function(x) {
  center <- colMeans(x)
  scale <- apply(x, 2, sd)

  zero_variance <- scale < 1e-8
  scale[zero_variance] <- 1

  list(
    center = center,
    scale = scale,
    zero = zero_variance
  )
}

scale_x <- function(x, scaler) {
  sweep(
    sweep(
      x,
      2,
      scaler$center,
      "-"
    ),
    2,
    scaler$scale,
    "/"
  )
}

fit_y_scaler <- function(y) {
  list(
    center = mean(y),
    scale = max(
      sd(y),
      1e-8
    )
  )
}

scale_y <- function(y, scaler) {
  (y - scaler$center) /
    scaler$scale
}

unscale_y <- function(y, scaler) {
  y * scaler$scale +
    scaler$center
}
```

# Shared ANN implementation

## Network objects

For a standardized marker vector \(\mathbf{x}\), each hidden layer applies a
linear transformation followed by ReLU activation:

\[
\mathbf{h}^{(\ell)} =
\operatorname{ReLU}
(\mathbf{W}_{\ell}\mathbf{h}^{(\ell-1)}+\mathbf{b}_{\ell}).
\]

| Symbol | Meaning |
|---|---|
| \(\mathbf{x}\) | standardized marker vector |
| \(\mathbf{W}\) | layer weights |
| \(\mathbf{b}\) | layer biases |
| \(\mathbf{h}\) | hidden-layer activations |
| \(\hat{y}\) | network prediction |

The output layer is linear because the response is continuous.

## Loss and regularization

Training minimizes

\[
\mathcal{L}=
\operatorname{MSE}(y,\hat{y})+
\lambda_1\sum_j |w_j|+
\frac{\lambda_2}{2}\sum_j w_j^2.
\]

`ANN_0` uses \(\lambda_1=\lambda_2=0\), `ANN_L1` uses only L1,
`ANN_L2` uses only L2, and `ANN_L1_L2` uses both. Bias terms are not penalized.
Adam uses `weight_decay = 0`, so L2 enters only through the explicit penalty
above.

The response is standardized from the fitting subset during optimization and
predictions are transformed back to the original phenotype scale before
performance metrics are calculated.

## Representative 64-32 network

The largest architecture in the prespecified catalog can be written directly as:

```{r ann-forward-example, eval=FALSE}
model <- nn_module(
  initialize = function(input_dim) {
    self$hidden1 <- nn_linear(input_dim, 64)
    self$hidden2 <- nn_linear(64, 32)
    self$output <- nn_linear(32, 1)
  },
  forward = function(x) {
    x <- nnf_relu(self$hidden1(x))
    x <- nnf_relu(self$hidden2(x))
    self$output(x)
  }
)
```

Smaller catalog architectures follow the same pattern with fewer hidden units
or one hidden layer. The generic implementation shown next constructs all nine
architectures from the same definition.

```{r model-definition, eval=ANN_RUNTIME_REQUIRED}
TorchGenomicMLP <- torch::nn_module(
  "TorchGenomicMLP_CanonicalTuning",

  initialize = function(
    input_dim,
    hidden_units
  ) {
    self$linears <-
      torch::nn_module_list()

    previous <-
      as.integer(input_dim)

    for (hidden in hidden_units) {
      self$linears$append(
        torch::nn_linear(
          previous,
          as.integer(hidden)
        )
      )
      previous <- as.integer(hidden)
    }

    self$output <-
      torch::nn_linear(
        previous,
        1L
      )
  },

  forward = function(x) {
    for (
      layer_index in seq_len(
        length(self$linears)
      )
    ) {
      x <- torch::nnf_relu(
        self$linears[
          [layer_index]
        ](x)
      )
    }

    self$output(x)
  }
)

weight_penalty <- function(
  model,
  l1,
  l2
) {
  weights <- lapply(
    seq_len(
      length(model$linears)
    ),
    function(layer_index) {
      model$linears[
        [layer_index]
      ]$weight
    }
  )

  weights[
    [length(weights) + 1L]
  ] <- model$output$weight

  penalty_l1 <-
    weights[[1]]$sum() * 0
  penalty_l2 <-
    weights[[1]]$sum() * 0

  for (weight in weights) {
    penalty_l1 <-
      penalty_l1 +
      weight$abs()$sum()

    penalty_l2 <-
      penalty_l2 +
      weight$pow(2)$sum()
  }

  l1 * penalty_l1 +
    (l2 / 2) * penalty_l2
}

clone_state <- function(model) {
  state <- model$state_dict()

  copied_state <- lapply(
    state,
    function(parameter) {
      parameter$detach()$clone()
    }
  )

  names(copied_state) <-
    names(state)

  copied_state
}
```

# Inner-CV training

The inner analysis set is divided deterministically into:

- a fitting subset;
- an early-validation subset.

The frozen inner assessment set remains untouched until prediction of that
candidate. Preprocessing is estimated from the fitting subset only.

Early stopping is an optimization mechanism. Its epoch diagnostics are saved
but excluded from the tuning rule.

The core optimization step itself is simple: predict the fitting subset,
calculate MSE plus the selected penalty, backpropagate, and update the weights.

```{r ann-optimization-step, eval=FALSE}
optimizer$zero_grad()

prediction <- model(x_fit)

loss <- nnf_mse_loss(
  prediction,
  y_fit
) + weight_penalty(
  model,
  l1,
  l2
)

loss$backward()
optimizer$step()
```

The complete helper shown next applies the fitting-set scalers, monitors the
early-validation loss, restores the best epoch, and predicts the untouched
inner-assessment fold. The only conditionals retained are the ones that define
the early-stopping algorithm itself.

```{r training-functions}
prepare_bundle <- function(
  fit_ids,
  validation_ids,
  assessment_ids,
  scenario_index
) {
  fit_index <- match(fit_ids, ids)
  validation_index <- match(
    validation_ids,
    ids
  )
  assessment_index <- match(
    assessment_ids,
    ids
  )

  x_scaler <- fit_x_scaler(
    geno[
      fit_index,
      ,
      drop = FALSE
    ]
  )
  y_scaler <- fit_y_scaler(
    phenotype[
      fit_index,
      scenario_index
    ]
  )

  list(
    x_fit = scale_x(
      geno[
        fit_index,
        ,
        drop = FALSE
      ],
      x_scaler
    ),
    x_validation = scale_x(
      geno[
        validation_index,
        ,
        drop = FALSE
      ],
      x_scaler
    ),
    x_assessment = scale_x(
      geno[
        assessment_index,
        ,
        drop = FALSE
      ],
      x_scaler
    ),
    y_fit = scale_y(
      phenotype[
        fit_index,
        scenario_index
      ],
      y_scaler
    ),
    y_validation = scale_y(
      phenotype[
        validation_index,
        scenario_index
      ],
      y_scaler
    ),
    y_assessment = phenotype[
      assessment_index,
      scenario_index
    ],
    y_scaler = y_scaler,
    zero_variance_markers =
      sum(x_scaler$zero)
  )
}

train_inner_candidate <- function(
  bundle,
  architecture,
  learning_rate,
  l1,
  l2,
  seed
) {
  set_ann_seed(seed)

  x_fit <- torch::torch_tensor(
    bundle$x_fit,
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  y_fit <- torch::torch_tensor(
    matrix(bundle$y_fit, ncol = 1),
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  x_validation <- torch::torch_tensor(
    bundle$x_validation,
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  y_validation <- torch::torch_tensor(
    matrix(
      bundle$y_validation,
      ncol = 1
    ),
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  x_assessment <- torch::torch_tensor(
    bundle$x_assessment,
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  model <- TorchGenomicMLP(
    ncol(bundle$x_fit),
    ARCHITECTURES[[architecture]]
  )$to(device = ANN_DEVICE)

  optimizer <- torch::optim_adam(
    params = model$parameters,
    lr = learning_rate,
    weight_decay = 0
  )

  best_validation_loss <- Inf
  checkpoint_epoch <- 1L
  checkpoint_state <- clone_state(model)
  stale_epochs <- 0L
  last_epoch <- 0L
  stopped_by_patience <- FALSE

  sync_device()
  fit_start <-
    proc.time()[["elapsed"]]

  for (epoch in seq_len(MAX_EPOCHS)) {
    last_epoch <- epoch

    model$train()
    optimizer$zero_grad()

    prediction <- model(x_fit)

    loss <-
      torch::nnf_mse_loss(
        prediction,
        y_fit
      ) +
      weight_penalty(
        model,
        l1,
        l2
      )

    loss$backward()
    optimizer$step()

    model$eval()

    validation_loss <-
      torch::with_no_grad({
        torch::nnf_mse_loss(
          model(x_validation),
          y_validation
        )
      })

    validation_value <-
      as.numeric(
        validation_loss$to(
          device =
            torch::torch_device("cpu")
        )
      )

    improved <-
      validation_value <
      best_validation_loss -
      MIN_DELTA

    if (improved) {
      best_validation_loss <-
        validation_value
      checkpoint_epoch <- epoch
      checkpoint_state <-
        clone_state(model)
      stale_epochs <- 0L
    } else {
      stale_epochs <-
        stale_epochs + 1L
    }

    if (stale_epochs >= PATIENCE) {
      stopped_by_patience <- TRUE
      break
    }
  }

  sync_device()
  fit_seconds <-
    proc.time()[["elapsed"]] -
    fit_start

  model$load_state_dict(
    checkpoint_state
  )
  model$eval()

  prediction_start <-
    proc.time()[["elapsed"]]

  prediction_scaled <-
    torch::with_no_grad({
      as.numeric(
        model(x_assessment)$to(
          device =
            torch::torch_device("cpu")
        )
      )
    })

  prediction_seconds <-
    proc.time()[["elapsed"]] -
    prediction_start

  list(
    prediction = unscale_y(
      prediction_scaled,
      bundle$y_scaler
    ),
    early_stop_epoch =
      as.integer(checkpoint_epoch),
    last_epoch =
      as.integer(last_epoch),
    stopped_by_patience =
      stopped_by_patience,
    hit_epoch_ceiling =
      last_epoch >= MAX_EPOCHS &&
      !stopped_by_patience,
    fit_seconds = fit_seconds,
    prediction_seconds =
      prediction_seconds,
    zero_variance_markers =
      bundle$zero_variance_markers
  )
}
```

# Candidate evaluation and selection

Candidate ranking is deterministic and does **not** include epoch information.

The ranking hierarchy is:

1. lower mean inner-assessment RMSE;
2. higher mean inner-assessment correlation;
3. fewer network parameters;
4. smaller total penalty coefficient;
5. smaller learning rate;
6. candidate ID.

A completed candidate table is therefore sorted by the same hierarchy:

```{r candidate-ranking-example, eval=FALSE}
candidate_summary <- candidate_summary[
  order(
    candidate_summary$Mean_RMSE,
    -candidate_summary$Mean_Correlation,
    candidate_summary$Parameters,
    candidate_summary$Penalty,
    candidate_summary$Learning_rate,
    candidate_summary$CandidateID
  ),
]
```

The first row is the selected candidate. The complete function shown next
fits the same candidate in each of the four inner folds, averages the three
prediction metrics, and then applies exactly this ranking hierarchy.

```{r candidate-functions}
evaluate_candidate <- function(
  bundles,
  model_name,
  architecture,
  learning_rate,
  l1,
  l2,
  seed_base,
  candidate_id
) {
  fold_rows <- vector("list", 4L)

  for (inner_fold in 1:4) {
    fitted <- train_inner_candidate(
      bundles[[inner_fold]],
      architecture,
      learning_rate,
      l1,
      l2,
      seed_base + inner_fold
    )

    fold_rows[[inner_fold]] <- data.frame(
      Model = model_name,
      CandidateID = candidate_id,
      Architecture = architecture,
      Learning_rate = learning_rate,
      L1 = l1,
      L2 = l2,
      InnerFold = inner_fold,
      RMSE = rmse(
        fitted$prediction,
        bundles[[inner_fold]]$y_assessment
      ),
      Correlation = correlation(
        fitted$prediction,
        bundles[[inner_fold]]$y_assessment
      ),
      MAE = mae(
        fitted$prediction,
        bundles[[inner_fold]]$y_assessment
      ),
      EarlyStop_epoch =
        fitted$early_stop_epoch,
      Last_epoch =
        fitted$last_epoch,
      Stopped_by_patience =
        fitted$stopped_by_patience,
      Hit_epoch_ceiling =
        fitted$hit_epoch_ceiling,
      Fit_seconds =
        fitted$fit_seconds,
      Prediction_seconds =
        fitted$prediction_seconds,
      Status = "ok",
      Error = NA_character_,
      stringsAsFactors = FALSE
    )
  }

  folds <- do.call(
    rbind,
    fold_rows
  )

  candidate_summary <- data.frame(
    Model = model_name,
    CandidateID = candidate_id,
    Architecture = architecture,
    Learning_rate = learning_rate,
    L1 = l1,
    L2 = l2,
    Inner_folds_ok = 4L,
    Mean_RMSE = mean(folds$RMSE),
    Mean_Correlation =
      mean(folds$Correlation),
    Mean_MAE = mean(folds$MAE),
    Median_EarlyStop_epoch =
      median(folds$EarlyStop_epoch),
    Max_Last_epoch =
      max(folds$Last_epoch),
    N_epoch_ceiling =
      sum(folds$Hit_epoch_ceiling),
    Status = "ok",
    stringsAsFactors = FALSE
  )

  list(
    summary = candidate_summary,
    folds = folds
  )
}

choose_candidate <- function(x) {
  ranked <- x

  ranked$Parameters <- vapply(
    ranked$Architecture,
    function(architecture) {
      parameter_count(
        ncol(geno),
        ARCHITECTURES[[architecture]]
      )
    },
    numeric(1)
  )

  ranked$Penalty <-
    ranked$L1 +
    ranked$L2

  ranked <- ranked[
    order(
      ranked$Mean_RMSE,
      -ranked$Mean_Correlation,
      ranked$Parameters,
      ranked$Penalty,
      ranked$Learning_rate,
      ranked$CandidateID
    ),
    ,
    drop = FALSE
  ]

  ranked[
    1,
    ,
    drop = FALSE
  ]
}
```

# Outer-task table

The six scenarios share the same 25 frozen outer splits, producing 150
`Scenario × outer split` tuning tasks.

```{r tasks}
outer_map <- unique(
  cv$outer_assignments[
    ,
    c(
      "Repetition",
      "OuterFold",
      "SplitID"
    )
  ]
)

outer_map <- outer_map[
  order(
    outer_map$Repetition,
    outer_map$OuterFold
  ),
  ,
  drop = FALSE
]

task_table <- merge(
  data.frame(
    Scenario = scenario_names,
    stringsAsFactors = FALSE
  ),
  outer_map,
  by = NULL
)

task_table <- task_table[
  order(
    match(
      task_table$Scenario,
      scenario_names
    ),
    task_table$Repetition,
    task_table$OuterFold
  ),
  ,
  drop = FALSE
]

task_table$TaskID <- paste(
  task_table$Scenario,
  task_table$SplitID,
  sep = "_"
)
task_table$Fold <-
  task_table$OuterFold

rownames(task_table) <- NULL

EXPECTED_TASKS <- nrow(task_table)
EXPECTED_SELECTED_ROWS <-
  EXPECTED_TASKS *
  length(MODEL_LEVELS)
EXPECTED_CANDIDATE_ROWS <-
  EXPECTED_TASKS *
  CANDIDATES_PER_TASK
EXPECTED_FOLD_ROWS <-
  EXPECTED_TASKS *
  INNER_FITS_PER_TASK

knitr::kable(
  head(task_table, 15),
  caption =
    "First 15 canonical ANN tuning tasks",
  row.names = FALSE
)
```



Each row represents one `Scenario × Repetition × Fold` combination. The same
`TaskID` is used for all four ANN variants, which ensures that their selected
regularization settings are compared within the same frozen outer task.

# Canonical nested tuning

This section evaluates only inner-CV candidates. It never predicts the outer
assessment set.

```{r run-tuning, eval=RUN_ANN_TUNING}
run_start <-
  proc.time()[["elapsed"]]

task_outputs <- vector(
  "list",
  nrow(task_table)
)

for (
  task_index in seq_len(
    nrow(task_table)
  )
) {
  task <- task_table[
    task_index,
    ,
    drop = FALSE
  ]

  scenario <- task$Scenario
  scenario_index <- match(
    scenario,
    scenario_names
  )
  split_id <- task$SplitID

  task_start <-
    proc.time()[["elapsed"]]

  inner <-
    cv$inner_splits[[split_id]]

  bundles <- vector(
    "list",
    4L
  )

  for (inner_fold in 1:4) {
    inner_split <-
      inner[[paste0(
        "I",
        inner_fold
      )]]

    validation_seed <-
      SEED_GLOBAL +
      scenario_index * 1000000L +
      task$Repetition * 10000L +
      task$OuterFold * 1000L +
      inner_fold * 10L

    validation <-
      make_validation_split(
        inner_split$analysis_ids,
        EARLY_VALIDATION_FRACTION,
        validation_seed
      )

    bundles[[inner_fold]] <-
      prepare_bundle(
        validation$train,
        validation$validation,
        inner_split$assessment_ids,
        scenario_index
      )
  }

  base_seed <-
    SEED_GLOBAL +
    scenario_index * 10000000L +
    task$Repetition * 100000L +
    task$OuterFold * 10000L

  # Stage 1: select architecture and learning rate using ANN_0.
  structure_summary_rows <- vector(
    "list",
    nrow(structure_grid)
  )
  structure_fold_rows <- vector(
    "list",
    nrow(structure_grid)
  )

  for (
    candidate_index in seq_len(
      nrow(structure_grid)
    )
  ) {
    candidate <-
      structure_grid[
        candidate_index,
        ,
        drop = FALSE
      ]

    evaluated <- evaluate_candidate(
      bundles = bundles,
      model_name = "ANN_0",
      architecture =
        candidate$Architecture,
      learning_rate =
        candidate$Learning_rate,
      l1 = 0,
      l2 = 0,
      seed_base =
        base_seed + 1000000L,
      candidate_id =
        candidate_index
    )

    structure_summary_rows[
      [candidate_index]
    ] <- evaluated$summary

    structure_fold_rows[
      [candidate_index]
    ] <- evaluated$folds
  }

  structure_summary <- do.call(
    rbind,
    structure_summary_rows
  )
  structure_folds <- do.call(
    rbind,
    structure_fold_rows
  )

  chosen_structure <-
    choose_candidate(
      structure_summary
    )

  selected_architecture <-
    chosen_structure$Architecture
  selected_learning_rate <-
    chosen_structure$Learning_rate

  # Stage 2: tune only the regularization coefficients.
  penalty_grids <- list(
    ANN_L1 = l1_grid,
    ANN_L2 = l2_grid,
    ANN_L1_L2 = l1l2_grid
  )

  penalty_summaries <- list()
  penalty_folds <- list()
  penalty_selected <- list()

  for (
    model_name in names(
      penalty_grids
    )
  ) {
    grid <-
      penalty_grids[[model_name]]

    candidate_summaries <- vector(
      "list",
      nrow(grid)
    )
    candidate_folds <- vector(
      "list",
      nrow(grid)
    )

    for (
      candidate_index in seq_len(
        nrow(grid)
      )
    ) {
      evaluated <- evaluate_candidate(
        bundles = bundles,
        model_name = model_name,
        architecture =
          selected_architecture,
        learning_rate =
          selected_learning_rate,
        l1 =
          grid$L1[candidate_index],
        l2 =
          grid$L2[candidate_index],
        seed_base =
          base_seed + 2000000L,
        candidate_id =
          candidate_index
      )

      candidate_summaries[
        [candidate_index]
      ] <- evaluated$summary
      candidate_folds[
        [candidate_index]
      ] <- evaluated$folds
    }

    penalty_summaries[
      [model_name]
    ] <- do.call(
      rbind,
      candidate_summaries
    )
    penalty_folds[
      [model_name]
    ] <- do.call(
      rbind,
      candidate_folds
    )
    penalty_selected[
      [model_name]
    ] <- choose_candidate(
      penalty_summaries[
        [model_name]
      ]
    )
  }

  selected_table <- rbind(
    data.frame(
      Model = "ANN_0",
      Architecture =
        selected_architecture,
      Learning_rate =
        selected_learning_rate,
      L1 = 0,
      L2 = 0,
      Inner_RMSE =
        chosen_structure$Mean_RMSE,
      Inner_Correlation =
        chosen_structure$Mean_Correlation,
      Inner_MAE =
        chosen_structure$Mean_MAE,
      stringsAsFactors = FALSE
    ),
    do.call(
      rbind,
      lapply(
        names(penalty_selected),
        function(model_name) {
          selected_candidate <-
            penalty_selected[
              [model_name]
            ]

          data.frame(
            Model = model_name,
            Architecture =
              selected_architecture,
            Learning_rate =
              selected_learning_rate,
            L1 =
              selected_candidate$L1,
            L2 =
              selected_candidate$L2,
            Inner_RMSE =
              selected_candidate$Mean_RMSE,
            Inner_Correlation =
              selected_candidate$Mean_Correlation,
            Inner_MAE =
              selected_candidate$Mean_MAE,
            stringsAsFactors = FALSE
          )
        }
      )
    )
  )

  for (
    column_name in c(
      "Scenario",
      "Repetition",
      "Fold",
      "SplitID",
      "TaskID"
    )
  ) {
    selected_table[[column_name]] <-
      task[[column_name]]
  }

  tuning_summary <- rbind(
    structure_summary,
    do.call(
      rbind,
      penalty_summaries
    )
  )
  tuning_folds <- rbind(
    structure_folds,
    do.call(
      rbind,
      penalty_folds
    )
  )

  for (
    column_name in c(
      "Scenario",
      "Repetition",
      "Fold",
      "SplitID",
      "TaskID"
    )
  ) {
    tuning_summary[[column_name]] <-
      task[[column_name]]
    tuning_folds[[column_name]] <-
      task[[column_name]]
  }

  task_seconds <-
    proc.time()[["elapsed"]] -
    task_start

  task_outputs[[task_index]] <- list(
    Status = "ok",
    ConfigID =
      ANN_TUNING_CONFIG_ID,
    TaskID = task$TaskID,
    selected = selected_table,
    tuning_summary = tuning_summary,
    tuning_folds = tuning_folds,
    task_seconds = task_seconds
  )

  cat(
    sprintf(
      "[%03d/%03d] %s | %.1f s | arch=%s lr=%g\n",
      task_index,
      nrow(task_table),
      task$TaskID,
      task_seconds,
      selected_architecture,
      selected_learning_rate
    )
  )
}

current_session_seconds <-
  proc.time()[["elapsed"]] -
  run_start

selected_configurations <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(task_output) {
      task_output$selected
    }
  )
)
candidate_summary <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(task_output) {
      task_output$tuning_summary
    }
  )
)
fold_results <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(task_output) {
      task_output$tuning_folds
    }
  )
)

task_seconds_total <- sum(
  vapply(
    task_outputs,
    function(task_output) {
      task_output$task_seconds
    },
    numeric(1)
  )
)

manifest <- list(
  analysis =
    "canonical_simulated_ann_tuning",
  dataset = "simulated",
  mode = "full",
  pilot_task = NA_character_,
  config_id =
    ANN_TUNING_CONFIG_ID,
  seed_global = SEED_GLOBAL,
  models = MODEL_LEVELS,
  architectures = ARCHITECTURES,
  learning_rates = LEARNING_RATES,
  l1_values = L1_VALUES,
  l2_values = L2_VALUES,
  l1_l2_grid = L1_L2_GRID,
  optimizer = "Adam",
  activation = "ReLU",
  dropout = DROPOUT,
  max_epochs = MAX_EPOCHS,
  final_train_epochs =
    TUNING_V1_COMPATIBILITY_TOKEN,
  patience = PATIENCE,
  min_delta = MIN_DELTA,
  early_validation_fraction =
    EARLY_VALIDATION_FRACTION,
  primary_tuning_metric =
    "mean inner-assessment RMSE",
  secondary_tuning_metric =
    "mean inner-assessment correlation",
  epoch_used_in_candidate_ranking =
    FALSE,
  outer_assessment_used =
    FALSE,
  true_genetic_value_used =
    FALSE,
  architecture_lr_shared_across_models =
    TRUE,
  candidates_per_task =
    CANDIDATES_PER_TASK,
  inner_fits_per_task =
    INNER_FITS_PER_TASK,
  number_tasks =
    EXPECTED_TASKS,
  selected_rows =
    nrow(selected_configurations),
  candidate_rows =
    nrow(candidate_summary),
  fold_rows =
    nrow(fold_results),
  torch_version =
    as.character(
      packageVersion("torch")
    ),
  device = ANN_DEVICE_LABEL,
  cuda_available =
    cuda_available,
  current_session_seconds =
    current_session_seconds,
  checkpoint_task_seconds =
    task_seconds_total,
  timestamp =
    as.character(Sys.time())
)

manifest_table <- data.frame(
  Field = c(
    "analysis",
    "mode",
    "config_id",
    "architectures",
    "learning_rates",
    "L1_values",
    "L2_values",
    "combined_grid_size",
    "MAX_EPOCHS",
    "TUNING_V1_COMPATIBILITY_TOKEN",
    "patience",
    "dropout",
    "optimizer",
    "activation",
    "candidates_per_task",
    "inner_fits_per_task",
    "tasks",
    "epoch_used_in_candidate_ranking",
    "outer_assessment_used",
    "true_genetic_value_used"
  ),
  Value = c(
    manifest$analysis,
    manifest$mode,
    manifest$config_id,
    paste(names(ARCHITECTURES), collapse = " | "),
    paste(LEARNING_RATES, collapse = " | "),
    paste(L1_VALUES, collapse = " | "),
    paste(L2_VALUES, collapse = " | "),
    nrow(L1_L2_GRID),
    MAX_EPOCHS,
    TUNING_V1_COMPATIBILITY_TOKEN,
    PATIENCE,
    DROPOUT,
    "Adam",
    "ReLU",
    CANDIDATES_PER_TASK,
    INNER_FITS_PER_TASK,
    EXPECTED_TASKS,
    FALSE,
    FALSE,
    FALSE
  ),
  stringsAsFactors = FALSE
)

saveRDS(
  selected_configurations,
  SELECTED_RDS,
  compress = "gzip"
)
saveRDS(
  candidate_summary,
  CANDIDATE_RDS,
  compress = "gzip"
)
saveRDS(
  fold_results,
  FOLD_RDS,
  compress = "gzip"
)
saveRDS(
  manifest,
  RUN_RDS,
  compress = "gzip"
)
saveRDS(
  manifest,
  MANIFEST_RDS,
  compress = "gzip"
)

write.csv(
  selected_configurations,
  SELECTED_CSV,
  row.names = FALSE
)
write.csv(
  candidate_summary,
  CANDIDATE_CSV,
  row.names = FALSE
)
write.csv(
  fold_results,
  FOLD_CSV,
  row.names = FALSE
)
write.csv(
  manifest_table,
  MANIFEST_CSV,
  row.names = FALSE
)
```

# Canonical tuning results

The selected configurations and tuning diagnostics below are generated from
the canonical saved tuning outputs. These files are loaded and checked
internally before display so the tutorial can focus on how the selected
hyperparameters are summarized and interpreted.

```{r saved-integrity, include=FALSE}
available <- all(
  file.exists(
    c(
      SELECTED_RDS,
      CANDIDATE_RDS,
      FOLD_RDS,
      RUN_RDS
    )
  )
)

if (!available) {
  cat(
    "Canonical ANN tuning outputs are unavailable; tuning summaries cannot be displayed.\n"
  )
} else {
  selected_configurations <-
    readRDS(SELECTED_RDS)
  candidate_summary <-
    readRDS(CANDIDATE_RDS)
  fold_results <-
    readRDS(FOLD_RDS)
  run_metadata <-
    readRDS(RUN_RDS)

  integrity <- data.frame(
    Check = c(
      "Configuration ID matches",
      "Selected rows",
      "Candidate rows",
      "Fold rows",
      "All candidate rows are ok",
      "All fold rows are ok",
      "All candidate RMSE values are finite",
      "All fold RMSE values are finite",
      "Epoch excluded from ranking",
      "Outer assessment excluded",
      "True genetic values excluded",
      "Architecture/LR shared within task"
    ),
    Status = c(
      if (
        identical(
          run_metadata$config_id,
          ANN_TUNING_CONFIG_ID
        )
      ) "PASS" else "BLOCKED",
      if (
        nrow(selected_configurations) ==
        EXPECTED_SELECTED_ROWS
      ) "PASS" else "BLOCKED",
      if (
        nrow(candidate_summary) ==
        EXPECTED_CANDIDATE_ROWS
      ) "PASS" else "BLOCKED",
      if (
        nrow(fold_results) ==
        EXPECTED_FOLD_ROWS
      ) "PASS" else "BLOCKED",
      if (
        all(
          candidate_summary$Status ==
            "ok"
        )
      ) "PASS" else "BLOCKED",
      if (
        all(
          fold_results$Status ==
            "ok"
        )
      ) "PASS" else "BLOCKED",
      if (
        all(
          is.finite(
            candidate_summary$Mean_RMSE
          )
        )
      ) "PASS" else "BLOCKED",
      if (
        all(
          is.finite(
            fold_results$RMSE
          )
        )
      ) "PASS" else "BLOCKED",
      if (
        identical(
          run_metadata$epoch_used_in_candidate_ranking,
          FALSE
        )
      ) "PASS" else "BLOCKED",
      if (
        identical(
          run_metadata$outer_assessment_used,
          FALSE
        )
      ) "PASS" else "BLOCKED",
      if (
        identical(
          run_metadata$true_genetic_value_used,
          FALSE
        )
      ) "PASS" else "BLOCKED",
      {
        task_structure <- aggregate(
          cbind(
            Architecture_code =
              as.numeric(
                factor(
                  selected_configurations$Architecture,
                  levels =
                    names(ARCHITECTURES)
                )
              ),
            Learning_rate =
              selected_configurations$Learning_rate
          ) ~ TaskID,
          data =
            selected_configurations,
          FUN = function(z) {
            length(unique(z))
          }
        )

        if (
          all(
            task_structure$Architecture_code ==
              1
          ) &&
          all(
            task_structure$Learning_rate ==
              1
          )
        ) {
          "PASS"
        } else {
          "BLOCKED"
        }
      }
    ),
    stringsAsFactors = FALSE
  )

  knitr::kable(
    integrity,
    caption =
      "Canonical ANN tuning saved-result integrity"
  )

  if (
    any(
      integrity$Status ==
        "BLOCKED"
    )
  ) {
    stop(
      "Canonical ANN tuning failed a blocking integrity check."
    )
  }
}
```

# Selected configurations

The summaries below show how often the 150 outer tasks selected each shared
architecture and learning rate, followed by the selected penalty values for the
regularized variants. A small configuration sample then illustrates the
task-level selections without reproducing all 600 rows.

```{r selected-summary}
task_level <- unique(
  selected_configurations[
    ,
    c(
      "TaskID",
      "Architecture",
      "Learning_rate"
    )
  ]
)

architecture_frequency <- as.data.frame(
  table(
    Architecture =
      task_level$Architecture
  ),
  stringsAsFactors = FALSE
)
names(architecture_frequency)[2] <-
  "Selected_tasks"

architecture_frequency <-
  architecture_frequency[
    order(
      match(
        architecture_frequency$Architecture,
        names(ARCHITECTURES)
      )
    ),
    ,
    drop = FALSE
  ]

learning_rate_frequency <- as.data.frame(
  table(
    Learning_rate =
      task_level$Learning_rate
  ),
  stringsAsFactors = FALSE
)
names(learning_rate_frequency)[2] <-
  "Selected_tasks"

l1_frequency <- selected_configurations[
  selected_configurations$Model %in%
    c("ANN_L1", "ANN_L1_L2"),
  c("Model", "L1"),
  drop = FALSE
]
l1_frequency$Selected_tasks <- 1L
l1_frequency <- aggregate(
  Selected_tasks ~ Model + L1,
  data = l1_frequency,
  FUN = sum
)

l2_frequency <- selected_configurations[
  selected_configurations$Model %in%
    c("ANN_L2", "ANN_L1_L2"),
  c("Model", "L2"),
  drop = FALSE
]
l2_frequency$Selected_tasks <- 1L
l2_frequency <- aggregate(
  Selected_tasks ~ Model + L2,
  data = l2_frequency,
  FUN = sum
)

knitr::kable(
  architecture_frequency,
  caption =
    "Selected shared architecture across the 150 outer tasks",
  row.names = FALSE
)
knitr::kable(
  learning_rate_frequency,
  caption =
    "Selected shared learning rate across the 150 outer tasks",
  row.names = FALSE
)
knitr::kable(
  l1_frequency,
  digits = 6,
  caption =
    "Selected L1 coefficients by regularized ANN variant",
  row.names = FALSE
)
knitr::kable(
  l2_frequency,
  digits = 6,
  caption =
    "Selected L2 coefficients by regularized ANN variant",
  row.names = FALSE
)

selected_display <- head(
  selected_configurations[
    order(
      selected_configurations$Scenario,
      selected_configurations$Repetition,
      selected_configurations$Fold,
      match(
        selected_configurations$Model,
        MODEL_LEVELS
      )
    ),
    c(
      "Scenario",
      "Repetition",
      "Fold",
      "Model",
      "Architecture",
      "Learning_rate",
      "L1",
      "L2",
      "Inner_RMSE",
      "Inner_Correlation"
    ),
    drop = FALSE
  ],
  20
)

knitr::kable(
  selected_display,
  digits = 6,
  caption =
    "First 20 selected canonical ANN configurations",
  table.attr = WIDE_TABLE_ATTR,
  row.names = FALSE
)
```

The selection frequencies summarize how often each architecture, learning rate, and penalty coefficient is preferred across the frozen outer tasks. Because architecture and learning rate are shared by the four ANN variants within each task, differences among `ANN_0`, `ANN_L1`, `ANN_L2`, and `ANN_L1_L2` are not produced by giving the regularized variants a different structural search space.

# Optimization diagnostics

These diagnostics describe how the fixed early-stopping protocol behaved over
the canonical tuning fits. They do not enter candidate ranking and do not
change the tuning grid.

```{r optimization-diagnostics}
selected_keys <- merge(
  selected_configurations[
    ,
    c(
      "TaskID",
      "Model",
      "Architecture",
      "Learning_rate",
      "L1",
      "L2"
    )
  ],
  fold_results,
  by = c(
    "TaskID",
    "Model",
    "Architecture",
    "Learning_rate",
    "L1",
    "L2"
  ),
  all.x = TRUE,
  sort = FALSE
)

optimization_summary <- data.frame(
  Diagnostic = c(
    "Inner candidate-fold fits",
    "Median best early-stop epoch",
    "Median last epoch",
    "Fits stopped by patience",
    "Fits reaching 2000-epoch ceiling",
    "Selected-candidate inner folds",
    "Selected-candidate folds reaching ceiling"
  ),
  Value = c(
    as.character(nrow(fold_results)),
    sprintf(
      "%.1f",
      median(
        fold_results$EarlyStop_epoch,
        na.rm = TRUE
      )
    ),
    sprintf(
      "%.1f",
      median(
        fold_results$Last_epoch,
        na.rm = TRUE
      )
    ),
    sprintf(
      "%d / %d",
      sum(
        fold_results$Stopped_by_patience,
        na.rm = TRUE
      ),
      nrow(fold_results)
    ),
    sprintf(
      "%d / %d",
      sum(
        fold_results$Hit_epoch_ceiling,
        na.rm = TRUE
      ),
      nrow(fold_results)
    ),
    as.character(nrow(selected_keys)),
    sprintf(
      "%d / %d",
      sum(
        selected_keys$Hit_epoch_ceiling,
        na.rm = TRUE
      ),
      nrow(selected_keys)
    )
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  optimization_summary,
  caption =
    "Canonical ANN tuning optimization diagnostics",
  row.names = FALSE
)
```

Of the `r nrow(fold_results)` inner candidate-fold fits,
`r sum(fold_results$Stopped_by_patience, na.rm = TRUE)` stopped by patience and
`r sum(fold_results$Hit_epoch_ceiling, na.rm = TRUE)` reached the 2000-epoch
ceiling. The median best epoch shows where the best early-validation checkpoint
typically occurred; these diagnostics are not used to rank candidates.

# Boundary description

Boundary occupancy is reported descriptively because the upper architecture
`64-32` is a deliberate methodological cap. Selection of `64-32` indicates that
the bounded search space was active for those tasks, but it does not trigger
post hoc enlargement of the canonical grid.

Consequently, boundary occupancy is treated as a limitation of the prespecified
canonical comparison rather than as justification for extending the search
space after observing the results.

```{r boundary-description}
task_level <- unique(
  selected_configurations[
    ,
    c(
      "TaskID",
      "Scenario",
      "Architecture",
      "Learning_rate"
    )
  ]
)

boundary_table <- data.frame(
  Boundary = c(
    "Architecture = 64-32",
    "Learning rate = 1.25e-4",
    "ANN_L1: L1 = 1e-1",
    "ANN_L1_L2: L1 = 1e-1",
    "ANN_L2: L2 = 1e-2",
    "ANN_L1_L2: L2 = 1e-2"
  ),
  Selected = c(
    sum(
      task_level$Architecture ==
        "64-32"
    ),
    sum(
      abs(
        task_level$Learning_rate -
          min(LEARNING_RATES)
      ) < 1e-14
    ),
    sum(
      selected_configurations$Model ==
        "ANN_L1" &
      abs(
        selected_configurations$L1 -
          max(L1_VALUES)
      ) < 1e-14
    ),
    sum(
      selected_configurations$Model ==
        "ANN_L1_L2" &
      abs(
        selected_configurations$L1 -
          max(L1_VALUES)
      ) < 1e-14
    ),
    sum(
      selected_configurations$Model ==
        "ANN_L2" &
      abs(
        selected_configurations$L2 -
          max(L2_VALUES)
      ) < 1e-14
    ),
    sum(
      selected_configurations$Model ==
        "ANN_L1_L2" &
      abs(
        selected_configurations$L2 -
          max(L2_VALUES)
      ) < 1e-14
    )
  ),
  Denominator = c(
    150L,
    150L,
    150L,
    150L,
    150L,
    150L
  ),
  stringsAsFactors = FALSE
)

boundary_table$Proportion <-
  boundary_table$Selected /
  boundary_table$Denominator

knitr::kable(
  boundary_table,
  digits = 4,
  caption =
    "Descriptive canonical tuning-boundary occupancy",
  row.names = FALSE
)
```

In the canonical run, the upper architecture `64-32` was selected in
`r boundary_table$Selected[boundary_table$Boundary == "Architecture = 64-32"]`
of 150 tasks. The remaining rows show how often the selected learning rate or
penalty coefficient reached the edge of its prespecified grid. These counts are
reported as search-space diagnostics, not as a reason to enlarge the grid after
observing the results.

# From tuning to final fitting

Nested tuning returns one architecture, learning rate, and applicable penalty
coefficient(s) for each `Scenario × outer split × ANN variant`.

The final stage follows one fixed sequence:

1. take the selected hyperparameters from nested tuning;
2. use only the 800 outer-analysis individuals to select `Best_epoch`;
3. allow at most 3000 epochs during this epoch-selection fit;
4. reinitialize the network;
5. refit all 800 outer-analysis individuals for exactly `Best_epoch`;
6. predict the untouched 200 outer-assessment individuals.

The 3000-epoch value is a **safety ceiling**, not the final training length.
Outer-assessment phenotypes and true genetic values are excluded from both
hyperparameter selection and `Best_epoch` selection.

| Final-fitting item | Canonical rule |
|---|---|
| Hyperparameters | inherited from nested tuning |
| Epoch-selection sample | 800 outer-analysis individuals only |
| Safety ceiling | 3000 epochs |
| Final refit | all 800 outer-analysis individuals |
| Final training length | exactly `Best_epoch` |
| Final assessment | untouched 200 individuals |

The complete final-fitting code is shown below. The table above provides the
conceptual contract first; the subsequent functions and loop implement that
contract for all 600 model-by-task fits.

```{r final-ann-contract}
FINAL_EPOCH_MAX_EPOCHS <- 3000L

FINAL_OUTPUT_STEM <- project_path(
  "output",
  "simulated",
  "ann",
  "final",
  "simulated_ann"
)

FINAL_RESULTS_RDS <-
  paste0(
    FINAL_OUTPUT_STEM,
    "_results.rds"
  )
FINAL_PREDICTIONS_RDS <-
  paste0(
    FINAL_OUTPUT_STEM,
    "_predictions.rds"
  )
FINAL_EPOCH_RDS <-
  paste0(
    FINAL_OUTPUT_STEM,
    "_epoch_selection.rds"
  )
FINAL_RUN_RDS <-
  paste0(
    FINAL_OUTPUT_STEM,
    "_run_metadata.rds"
  )

FINAL_RESULTS_CSV <-
  paste0(
    FINAL_OUTPUT_STEM,
    "_results.csv"
  )
FINAL_PREDICTIONS_CSV <-
  paste0(
    FINAL_OUTPUT_STEM,
    "_predictions.csv"
  )
FINAL_EPOCH_CSV <-
  paste0(
    FINAL_OUTPUT_STEM,
    "_epoch_selection.csv"
  )

FINAL_CONFIG_ID <- paste0(
  "ANN_FINAL_CANONICAL_V2__FROM__",
  ANN_TUNING_CONFIG_ID,
  "__FINAL_EPOCH_CAP_",
  FINAL_EPOCH_MAX_EPOCHS,
  "__BEST_EPOCH_REFIT"
)

final_contract <- data.frame(
  Item = c(
    "Tuning source",
    "Epoch ceiling",
    "Final epoch rule",
    "Outer assessment in tuning",
    "Outer assessment in epoch selection",
    "True G in selection"
  ),
  Value = c(
    "ANN_TUNING_CANONICAL_V1",
    FINAL_EPOCH_MAX_EPOCHS,
    "Best_epoch selected inside outer-analysis; refit all 800",
    FALSE,
    FALSE,
    FALSE
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  final_contract,
  caption =
    "Canonical final ANN contract",
  row.names = FALSE
)
```

The three `FALSE` entries are the key leakage controls: the outer assessment
set is excluded from tuning and epoch selection, and true genetic values are
used only after prediction for evaluation. `Best_epoch` is therefore selected
without access to either final assessment target.

<!-- Source section: final-fit helper functions -->

```{r final-ann-functions}
final_bias <- function(
  predicted,
  observed
) {
  mean(
    predicted - observed
  )
}

final_slope <- function(
  predicted,
  observed
) {
  as.numeric(
    coef(
      lm(
        observed ~ predicted
      )
    )[2]
  )
}

final_r2 <- function(
  predicted,
  observed
) {
  total_variation <- sum(
    (
      observed -
      mean(observed)
    )^2
  )

  1 -
    sum(
      (
        observed -
        predicted
      )^2
    ) /
    total_variation
}

final_metrics <- function(
  predicted,
  observed,
  prefix
) {
  values <- c(
    correlation(predicted, observed),
    rmse(predicted, observed),
    mae(predicted, observed),
    final_bias(predicted, observed),
    final_slope(predicted, observed),
    final_r2(predicted, observed)
  )

  names(values) <- paste0(
    prefix,
    c(
      "_Correlation",
      "_RMSE",
      "_MAE",
      "_Bias",
      "_Slope",
      "_R2"
    )
  )

  values
}

select_final_epoch <- function(
  analysis_ids,
  scenario_index,
  architecture,
  learning_rate,
  l1,
  l2,
  split_seed,
  fit_seed
) {
  validation_split <-
    make_validation_split(
      analysis_ids,
      EARLY_VALIDATION_FRACTION,
      split_seed
    )

  fit_index <- match(
    validation_split$train,
    ids
  )
  validation_index <- match(
    validation_split$validation,
    ids
  )

  x_scaler <- fit_x_scaler(
    geno[
      fit_index,
      ,
      drop = FALSE
    ]
  )
  y_scaler <- fit_y_scaler(
    phenotype[
      fit_index,
      scenario_index
    ]
  )

  x_fit <- scale_x(
    geno[
      fit_index,
      ,
      drop = FALSE
    ],
    x_scaler
  )
  x_validation <- scale_x(
    geno[
      validation_index,
      ,
      drop = FALSE
    ],
    x_scaler
  )
  y_fit <- scale_y(
    phenotype[
      fit_index,
      scenario_index
    ],
    y_scaler
  )
  y_validation <- scale_y(
    phenotype[
      validation_index,
      scenario_index
    ],
    y_scaler
  )

  set_ann_seed(fit_seed)

  x_fit_tensor <-
    torch::torch_tensor(
      x_fit,
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)

  y_fit_tensor <-
    torch::torch_tensor(
      matrix(y_fit, ncol = 1),
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)

  x_validation_tensor <-
    torch::torch_tensor(
      x_validation,
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)

  y_validation_tensor <-
    torch::torch_tensor(
      matrix(y_validation, ncol = 1),
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)

  model <- TorchGenomicMLP(
    ncol(x_fit),
    ARCHITECTURES[[architecture]]
  )$to(device = ANN_DEVICE)

  optimizer <- torch::optim_adam(
    params = model$parameters,
    lr = learning_rate,
    weight_decay = 0
  )

  best_validation_loss <- Inf
  best_epoch <- 1L
  best_state <- clone_state(model)
  stale_epochs <- 0L
  last_epoch <- 0L
  stopped_by_patience <- FALSE

  fit_start <-
    proc.time()[["elapsed"]]

  for (
    epoch in seq_len(
      FINAL_EPOCH_MAX_EPOCHS
    )
  ) {
    last_epoch <- epoch

    model$train()
    optimizer$zero_grad()

    prediction <-
      model(x_fit_tensor)

    loss <-
      torch::nnf_mse_loss(
        prediction,
        y_fit_tensor
      ) +
      weight_penalty(
        model,
        l1,
        l2
      )

    loss$backward()
    optimizer$step()

    model$eval()

    validation_loss <-
      torch::with_no_grad({
        torch::nnf_mse_loss(
          model(x_validation_tensor),
          y_validation_tensor
        )
      })

    validation_value <-
      as.numeric(
        validation_loss$to(
          device =
            torch::torch_device("cpu")
        )
      )

    improved <-
      validation_value <
      best_validation_loss -
      MIN_DELTA

    if (improved) {
      best_validation_loss <-
        validation_value
      best_epoch <- epoch
      best_state <-
        clone_state(model)
      stale_epochs <- 0L
    } else {
      stale_epochs <-
        stale_epochs + 1L
    }

    if (stale_epochs >= PATIENCE) {
      stopped_by_patience <- TRUE
      break
    }
  }

  fit_seconds <-
    proc.time()[["elapsed"]] -
    fit_start

  model$load_state_dict(
    best_state
  )
  model$eval()

  prediction_scaled <-
    torch::with_no_grad({
      as.numeric(
        model(
          x_validation_tensor
        )$to(
          device =
            torch::torch_device("cpu")
        )
      )
    })

  prediction <-
    unscale_y(
      prediction_scaled,
      y_scaler
    )
  observed <-
    phenotype[
      validation_index,
      scenario_index
    ]

  list(
    best_epoch =
      as.integer(best_epoch),
    last_epoch =
      as.integer(last_epoch),
    stopped_by_patience =
      stopped_by_patience,
    hit_epoch_ceiling =
      last_epoch >=
      FINAL_EPOCH_MAX_EPOCHS &&
      !stopped_by_patience,
    validation_rmse =
      rmse(prediction, observed),
    validation_correlation =
      correlation(
        prediction,
        observed
      ),
    fit_seconds = fit_seconds,
    n_fit =
      length(
        validation_split$train
      ),
    n_validation =
      length(
        validation_split$validation
      )
  )
}

train_final_fixed <- function(
  train_ids,
  assessment_ids,
  scenario_index,
  architecture,
  learning_rate,
  l1,
  l2,
  epochs,
  seed
) {
  train_index <- match(
    train_ids,
    ids
  )
  assessment_index <- match(
    assessment_ids,
    ids
  )

  x_scaler <- fit_x_scaler(
    geno[
      train_index,
      ,
      drop = FALSE
    ]
  )
  y_scaler <- fit_y_scaler(
    phenotype[
      train_index,
      scenario_index
    ]
  )

  x_train <- scale_x(
    geno[
      train_index,
      ,
      drop = FALSE
    ],
    x_scaler
  )
  x_assessment <- scale_x(
    geno[
      assessment_index,
      ,
      drop = FALSE
    ],
    x_scaler
  )
  y_train <- scale_y(
    phenotype[
      train_index,
      scenario_index
    ],
    y_scaler
  )

  set_ann_seed(seed)

  x_train_tensor <-
    torch::torch_tensor(
      x_train,
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)

  y_train_tensor <-
    torch::torch_tensor(
      matrix(y_train, ncol = 1),
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)

  x_assessment_tensor <-
    torch::torch_tensor(
      x_assessment,
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)

  model <- TorchGenomicMLP(
    ncol(x_train),
    ARCHITECTURES[[architecture]]
  )$to(device = ANN_DEVICE)

  optimizer <- torch::optim_adam(
    params = model$parameters,
    lr = learning_rate,
    weight_decay = 0
  )

  fit_start <-
    proc.time()[["elapsed"]]

  for (
    epoch in seq_len(
      as.integer(epochs)
    )
  ) {
    model$train()
    optimizer$zero_grad()

    prediction <-
      model(x_train_tensor)

    loss <-
      torch::nnf_mse_loss(
        prediction,
        y_train_tensor
      ) +
      weight_penalty(
        model,
        l1,
        l2
      )

    loss$backward()
    optimizer$step()
  }

  fit_seconds <-
    proc.time()[["elapsed"]] -
    fit_start

  model$eval()

  prediction_start <-
    proc.time()[["elapsed"]]

  prediction_scaled <-
    torch::with_no_grad({
      as.numeric(
        model(
          x_assessment_tensor
        )$to(
          device =
            torch::torch_device("cpu")
        )
      )
    })

  prediction_seconds <-
    proc.time()[["elapsed"]] -
    prediction_start

  list(
    prediction = unscale_y(
      prediction_scaled,
      y_scaler
    ),
    fit_seconds =
      fit_seconds,
    prediction_seconds =
      prediction_seconds,
    zero_variance_markers =
      sum(x_scaler$zero)
  )
}
```

<!-- Source section: final outer-CV tasks -->

```{r final-ann-tasks}
outer_map <- unique(
  cv$outer_assignments[
    ,
    c(
      "Repetition",
      "OuterFold",
      "SplitID"
    )
  ]
)

outer_map <- outer_map[
  order(
    outer_map$Repetition,
    outer_map$OuterFold
  ),
  ,
  drop = FALSE
]

final_task_table <- merge(
  data.frame(
    Scenario = scenario_names,
    stringsAsFactors = FALSE
  ),
  outer_map,
  by = NULL
)

final_task_table <- final_task_table[
  order(
    match(
      final_task_table$Scenario,
      scenario_names
    ),
    final_task_table$Repetition,
    final_task_table$OuterFold
  ),
  ,
  drop = FALSE
]

final_task_table$Fold <-
  final_task_table$OuterFold
final_task_table$TaskID <- paste(
  final_task_table$Scenario,
  final_task_table$SplitID,
  sep = "_"
)

rownames(final_task_table) <- NULL

knitr::kable(
  head(final_task_table, 15),
  caption =
    "First 15 canonical ANN final tasks",
  row.names = FALSE
)
```

Each row is one `Scenario × outer split` final-prediction task. The four ANN
variants reuse the same `TaskID`, so the final stage contains
**150 tasks × 4 variants = 600 model fits**, each evaluated on the same 200
outer-assessment individuals.

<!-- Source section: final fitting and outer prediction -->

```{r run-final-ann, eval=RUN_ANN}
dir.create(
  dirname(FINAL_RESULTS_RDS),
  recursive = TRUE,
  showWarnings = FALSE
)

true_g <-
  common$true_genetic_value
control <-
  common$scenario_control

run_start <-
  proc.time()[["elapsed"]]

task_outputs <- vector(
  "list",
  nrow(final_task_table)
)

for (
  task_index in seq_len(
    nrow(final_task_table)
  )
) {
  task <- final_task_table[
    task_index,
    ,
    drop = FALSE
  ]

  scenario_index <- match(
    task$Scenario,
    scenario_names
  )
  outer <-
    cv$outer_splits[[task$SplitID]]

  selected_for_task <-
    selected_configurations[
      selected_configurations$TaskID ==
        task$TaskID,
      ,
      drop = FALSE
    ]

  seed_base <-
    SEED_GLOBAL +
    scenario_index * 10000000L +
    task$Repetition * 100000L +
    task$OuterFold * 10000L

  split_seed <-
    seed_base + 8000000L
  epoch_seed <-
    seed_base + 8100000L
  final_fit_seed <-
    seed_base + 8200000L

  result_rows <- vector(
    "list",
    length(MODEL_LEVELS)
  )
  prediction_rows <- vector(
    "list",
    length(MODEL_LEVELS)
  )
  epoch_rows <- vector(
    "list",
    length(MODEL_LEVELS)
  )

  task_start <-
    proc.time()[["elapsed"]]

  for (
    model_index in seq_along(
      MODEL_LEVELS
    )
  ) {
    model_name <-
      MODEL_LEVELS[model_index]

    config <-
      selected_for_task[
        selected_for_task$Model ==
          model_name,
        ,
        drop = FALSE
      ]

    epoch_fit <- select_final_epoch(
      analysis_ids =
        outer$analysis_ids,
      scenario_index =
        scenario_index,
      architecture =
        config$Architecture,
      learning_rate =
        config$Learning_rate,
      l1 = config$L1,
      l2 = config$L2,
      split_seed =
        split_seed,
      fit_seed =
        epoch_seed
    )

    final_fit <- train_final_fixed(
      train_ids =
        outer$analysis_ids,
      assessment_ids =
        outer$assessment_ids,
      scenario_index =
        scenario_index,
      architecture =
        config$Architecture,
      learning_rate =
        config$Learning_rate,
      l1 = config$L1,
      l2 = config$L2,
      epochs =
        epoch_fit$best_epoch,
      seed =
        final_fit_seed
    )

    assessment_index <- match(
      outer$assessment_ids,
      ids
    )

    observed <-
      phenotype[
        assessment_index,
        scenario_index
      ]
    true_genetic_value <-
      true_g[
        assessment_index,
        scenario_index
      ]
    predicted_value <-
      final_fit$prediction

    genetic_metrics <- final_metrics(
      predicted_value,
      true_genetic_value,
      "Genetic"
    )
    phenotype_metrics <- final_metrics(
      predicted_value,
      observed,
      "Phenotype"
    )

    meta <- control[
      control$Scenario ==
        task$Scenario,
      ,
      drop = FALSE
    ]

    result_rows[[model_index]] <- data.frame(
      Dataset = "simulated",
      Scenario = task$Scenario,
      QTL_N = meta$QTL_N,
      H2_nominal = meta$H2_nominal,
      Repetition = task$Repetition,
      Fold = task$OuterFold,
      SplitID = task$SplitID,
      TaskID = task$TaskID,
      Model = model_name,
      Architecture =
        config$Architecture,
      Learning_rate =
        config$Learning_rate,
      L1 = config$L1,
      L2 = config$L2,
      Best_epoch =
        epoch_fit$best_epoch,
      Epoch_Last =
        epoch_fit$last_epoch,
      Epoch_Stopped_by_patience =
        epoch_fit$stopped_by_patience,
      Epoch_Hit_ceiling =
        epoch_fit$hit_epoch_ceiling,
      Inner_RMSE =
        config$Inner_RMSE,
      Inner_Correlation =
        config$Inner_Correlation,
      Inner_MAE =
        config$Inner_MAE,
      N_train = 800L,
      N_test = 200L,
      N_pred_finite =
        sum(
          is.finite(
            predicted_value
          )
        ),
      N_zero_variance_markers =
        final_fit$zero_variance_markers,
      Genetic_Correlation =
        genetic_metrics[
          "Genetic_Correlation"
        ],
      Genetic_RMSE =
        genetic_metrics[
          "Genetic_RMSE"
        ],
      Genetic_MAE =
        genetic_metrics[
          "Genetic_MAE"
        ],
      Genetic_Bias =
        genetic_metrics[
          "Genetic_Bias"
        ],
      Genetic_Slope =
        genetic_metrics[
          "Genetic_Slope"
        ],
      Genetic_R2 =
        genetic_metrics[
          "Genetic_R2"
        ],
      Phenotype_Correlation =
        phenotype_metrics[
          "Phenotype_Correlation"
        ],
      Phenotype_RMSE =
        phenotype_metrics[
          "Phenotype_RMSE"
        ],
      Phenotype_MAE =
        phenotype_metrics[
          "Phenotype_MAE"
        ],
      Phenotype_Bias =
        phenotype_metrics[
          "Phenotype_Bias"
        ],
      Phenotype_Slope =
        phenotype_metrics[
          "Phenotype_Slope"
        ],
      Phenotype_R2 =
        phenotype_metrics[
          "Phenotype_R2"
        ],
      Epoch_selection_seconds =
        epoch_fit$fit_seconds,
      Final_fit_seconds =
        final_fit$fit_seconds,
      Predict_seconds =
        final_fit$prediction_seconds,
      Status = "ok",
      Error = NA_character_,
      stringsAsFactors = FALSE
    )

    prediction_rows[[model_index]] <- data.frame(
      Dataset = "simulated",
      Scenario = task$Scenario,
      QTL_N = meta$QTL_N,
      H2_nominal = meta$H2_nominal,
      Repetition = task$Repetition,
      Fold = task$OuterFold,
      SplitID = task$SplitID,
      TaskID = task$TaskID,
      Model = model_name,
      Architecture =
        config$Architecture,
      Learning_rate =
        config$Learning_rate,
      L1 = config$L1,
      L2 = config$L2,
      Best_epoch =
        epoch_fit$best_epoch,
      ID = ids[assessment_index],
      Observed_Phenotype =
        observed,
      True_Genetic_Value =
        true_genetic_value,
      Predicted_Value =
        predicted_value,
      Status = "ok",
      stringsAsFactors = FALSE
    )

    epoch_rows[[model_index]] <- data.frame(
      Scenario = task$Scenario,
      Repetition = task$Repetition,
      Fold = task$OuterFold,
      SplitID = task$SplitID,
      TaskID = task$TaskID,
      Model = model_name,
      Architecture =
        config$Architecture,
      Learning_rate =
        config$Learning_rate,
      L1 = config$L1,
      L2 = config$L2,
      Best_epoch =
        epoch_fit$best_epoch,
      Last_epoch =
        epoch_fit$last_epoch,
      Stopped_by_patience =
        epoch_fit$stopped_by_patience,
      Hit_epoch_ceiling =
        epoch_fit$hit_epoch_ceiling,
      Validation_RMSE =
        epoch_fit$validation_rmse,
      Validation_Correlation =
        epoch_fit$validation_correlation,
      N_epoch_fit =
        epoch_fit$n_fit,
      N_epoch_validation =
        epoch_fit$n_validation,
      Fit_seconds =
        epoch_fit$fit_seconds,
      stringsAsFactors = FALSE
    )
  }

  task_outputs[[task_index]] <- list(
    results = do.call(
      rbind,
      result_rows
    ),
    predictions = do.call(
      rbind,
      prediction_rows
    ),
    epoch_selection = do.call(
      rbind,
      epoch_rows
    ),
    task_seconds =
      proc.time()[["elapsed"]] -
      task_start
  )

  cat(
    sprintf(
      "[%03d/%03d] %s | final %.1f min\n",
      task_index,
      nrow(final_task_table),
      task$TaskID,
      task_outputs[
        [task_index]
      ]$task_seconds / 60
    )
  )
}

results <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(task_output) {
      task_output$results
    }
  )
)
predictions <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(task_output) {
      task_output$predictions
    }
  )
)
epoch_selection <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(task_output) {
      task_output$epoch_selection
    }
  )
)

rownames(results) <- NULL
rownames(predictions) <- NULL
rownames(epoch_selection) <- NULL

run_seconds <-
  proc.time()[["elapsed"]] -
  run_start

task_seconds_total <- sum(
  vapply(
    task_outputs,
    function(task_output) {
      task_output$task_seconds
    },
    numeric(1)
  )
)

final_meta <- list(
  dataset = "simulated",
  analysis =
    "canonical_simulated_ann",
  mode = "full",
  pilot_task = NA_character_,
  config_id = FINAL_CONFIG_ID,
  tuning_config_id =
    ANN_TUNING_CONFIG_ID,
  seed_global = SEED_GLOBAL,
  models = MODEL_LEVELS,
  tuning_max_epochs =
    MAX_EPOCHS,
  final_epoch_max_epochs =
    FINAL_EPOCH_MAX_EPOCHS,
  patience = PATIENCE,
  min_delta = MIN_DELTA,
  early_validation_fraction =
    EARLY_VALIDATION_FRACTION,
  final_refit_rule =
    "Best_epoch selected inside outer-analysis, then refit on all 800",
  outer_assessment_used_in_tuning =
    FALSE,
  outer_assessment_used_in_epoch_selection =
    FALSE,
  true_genetic_value_used_in_selection =
    FALSE,
  number_tasks =
    nrow(final_task_table),
  model_task_rows =
    nrow(results),
  prediction_rows =
    nrow(predictions),
  epoch_selection_rows =
    nrow(epoch_selection),
  torch_version =
    as.character(
      packageVersion("torch")
    ),
  device = ANN_DEVICE_LABEL,
  cuda_available =
    cuda_available,
  current_session_seconds =
    run_seconds,
  checkpoint_task_seconds =
    task_seconds_total,
  timestamp =
    as.character(Sys.time())
)

saveRDS(
  results,
  FINAL_RESULTS_RDS,
  compress = "gzip"
)
saveRDS(
  predictions,
  FINAL_PREDICTIONS_RDS,
  compress = "gzip"
)
saveRDS(
  epoch_selection,
  FINAL_EPOCH_RDS,
  compress = "gzip"
)
saveRDS(
  final_meta,
  FINAL_RUN_RDS,
  compress = "gzip"
)

write.csv(
  results,
  FINAL_RESULTS_CSV,
  row.names = FALSE
)
write.csv(
  predictions,
  FINAL_PREDICTIONS_CSV,
  row.names = FALSE
)
write.csv(
  epoch_selection,
  FINAL_EPOCH_CSV,
  row.names = FALSE
)
```

# Canonical final results

The final performance tables use the canonical 600 model-task results,
120,000 held-out predictions, and 600 epoch-selection records. These saved
objects are loaded and checked internally before the reader-facing summaries are
constructed.

```{r final-ann-integrity, include=FALSE}
final_available<-all(file.exists(c(FINAL_RESULTS_RDS,FINAL_PREDICTIONS_RDS,FINAL_EPOCH_RDS,FINAL_RUN_RDS)))
if(!final_available){cat("Canonical final ANN outputs are unavailable; final summaries cannot be displayed.\n")}else{
  results<-readRDS(FINAL_RESULTS_RDS);predictions<-readRDS(FINAL_PREDICTIONS_RDS);epoch_selection<-readRDS(FINAL_EPOCH_RDS);final_meta<-readRDS(FINAL_RUN_RDS);nt<-if(identical(final_meta$mode,"pilot"))1L else 150L;nr<-nt*4L;np<-nr*200L
  audit<-data.frame(Check=c("Final config ID","Tuning config ID","Model-task rows","Prediction rows","Epoch-selection rows","4 models/task","200 predictions/model-task","All model tasks ok","All predictions finite","Best_epoch in range","Outer excluded from tuning","Outer excluded from epoch selection","True G excluded from selection"),Status=c(if(identical(final_meta$config_id,FINAL_CONFIG_ID))"PASS"else"BLOCKED",if(identical(final_meta$tuning_config_id,ANN_TUNING_CONFIG_ID))"PASS"else"BLOCKED",if(nrow(results)==nr)"PASS"else"BLOCKED",if(nrow(predictions)==np)"PASS"else"BLOCKED",if(nrow(epoch_selection)==nr)"PASS"else"BLOCKED",if(all(table(results$TaskID)==4L))"PASS"else"BLOCKED",if(all(table(interaction(predictions$TaskID,predictions$Model,drop=TRUE))==200L))"PASS"else"BLOCKED",if(all(results$Status=="ok"))"PASS"else"BLOCKED",if(all(is.finite(predictions$Predicted_Value)))"PASS"else"BLOCKED",if(all(epoch_selection$Best_epoch>=1L&epoch_selection$Best_epoch<=FINAL_EPOCH_MAX_EPOCHS))"PASS"else"BLOCKED",if(identical(final_meta$outer_assessment_used_in_tuning,FALSE))"PASS"else"BLOCKED",if(identical(final_meta$outer_assessment_used_in_epoch_selection,FALSE))"PASS"else"BLOCKED",if(identical(final_meta$true_genetic_value_used_in_selection,FALSE))"PASS"else"BLOCKED"))
  knitr::kable(audit,caption="Canonical final ANN saved-result integrity");if(any(audit$Status=="BLOCKED"))stop("Canonical final ANN failed integrity audit.");cat("\nFinal Best_epoch summary:\n");print(summary(epoch_selection$Best_epoch));cat("\nFinal epoch-selection fits hitting",FINAL_EPOCH_MAX_EPOCHS,"ceiling:",sum(epoch_selection$Hit_epoch_ceiling,na.rm=TRUE),"/",nrow(epoch_selection),"\n")
}
```

Across the `r nrow(epoch_selection)` final epoch-selection fits,
`r sum(epoch_selection$Stopped_by_patience, na.rm = TRUE)` stopped by patience
and `r sum(epoch_selection$Hit_epoch_ceiling, na.rm = TRUE)` reached the
`r FINAL_EPOCH_MAX_EPOCHS`-epoch ceiling. Thus, the safety ceiling did not define
a common training length; each final network was subsequently refit for its own
selected `Best_epoch`.


# Final ANN predictive performance

Final performance is summarized separately for the two evaluation targets.
Recovery of the known **true genetic value** is the primary simulated-data
criterion; prediction of the **observed phenotype** is secondary. Both tables
average the same 25 frozen outer splits for each scenario and ANN variant.

```{r final-ann-performance}
ok <- results[
  results$Status == "ok",
  ,
  drop = FALSE
]

perf <- aggregate(
  cbind(
    Genetic_Correlation,
    Genetic_RMSE,
    Genetic_MAE,
    Genetic_Bias,
    Genetic_Slope,
    Genetic_R2,
    Phenotype_Correlation,
    Phenotype_RMSE,
    Phenotype_MAE,
    Phenotype_Bias,
    Phenotype_Slope,
    Phenotype_R2
  ) ~ Scenario + Model + QTL_N + H2_nominal,
  data = ok,
  FUN = mean,
  na.rm = TRUE
)

perf <- perf[
  order(
    match(
      perf$Scenario,
      scenario_names
    ),
    match(
      perf$Model,
      MODEL_LEVELS
    )
  ),
  ,
  drop = FALSE
]

rownames(perf) <- NULL

knitr::kable(
  perf[
    ,
    c(
      "Scenario",
      "Model",
      "QTL_N",
      "H2_nominal",
      "Genetic_Correlation",
      "Genetic_RMSE",
      "Genetic_MAE",
      "Genetic_Bias",
      "Genetic_Slope",
      "Genetic_R2"
    )
  ],
  digits = 4,
  caption =
    "Primary evaluation: mean outer-CV ANN performance against true genetic value",
  table.attr = WIDE_TABLE_ATTR,
  row.names = FALSE
)

knitr::kable(
  perf[
    ,
    c(
      "Scenario",
      "Model",
      "QTL_N",
      "H2_nominal",
      "Phenotype_Correlation",
      "Phenotype_RMSE",
      "Phenotype_MAE",
      "Phenotype_Bias",
      "Phenotype_Slope",
      "Phenotype_R2"
    )
  ],
  digits = 4,
  caption =
    "Secondary evaluation: mean outer-CV ANN performance against observed phenotype",
  table.attr = WIDE_TABLE_ATTR,
  row.names = FALSE
)

overall_ann <- aggregate(
  cbind(
    Genetic_Correlation,
    Genetic_RMSE,
    Genetic_R2
  ) ~ Model,
  data = ok,
  FUN = mean,
  na.rm = TRUE
)

overall_ann <- overall_ann[
  match(
    c(
      "ANN_0",
      "ANN_L1",
      "ANN_L2",
      "ANN_L1_L2"
    ),
    overall_ann$Model
  ),
  ,
  drop = FALSE
]
```

Across all outer-CV tasks, ANN-L1+L2 has the highest mean genetic
correlation
(`r sprintf("%.4f", overall_ann$Genetic_Correlation[4])`) and lowest mean
genetic RMSE
(`r sprintf("%.4f", overall_ann$Genetic_RMSE[4])`) among the four ANN variants.
ANN-L1 is effectively tied (correlation
`r sprintf("%.4f", overall_ann$Genetic_Correlation[2])`; RMSE
`r sprintf("%.4f", overall_ann$Genetic_RMSE[2])`), whereas ANN-L2 remains close
to the unregularized ANN (correlations
`r sprintf("%.4f", overall_ann$Genetic_Correlation[3])` and
`r sprintf("%.4f", overall_ann$Genetic_Correlation[1])`, respectively).
Thus, within the prespecified canonical model/architecture search space, the
observable regularization gain is driven mainly by the L1 component. The formal
comparison with GBLUP-ADE is presented in the next module.

# Navigation

[Previous: Validation design](simulated_data_validation.html) | [Parallel branch: GBLUP-ADE](simulated_data_gblup.html) | [Next: Model comparison](simulated_data_comparison.html)

# Reproducibility information

```{r session-info}
sessionInfo()
```
