---
title: "Simulated data validation design"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Purpose

This analysis freezes the validation design used by all genomic prediction
models fitted to the simulated dataset.

The design is created **before model fitting** and is independent of phenotype,
true genetic value, scenario, and model. Consequently, GBLUP-ADE and all ANN
variants are evaluated on exactly the same outer assessment individuals.

The canonical design uses:

- 5 outer folds;
- 5 outer repetitions;
- 4 inner folds within each outer analysis set;
- deterministic seeds derived from the project-wide seed;
- the same outer and inner partitions for all six simulation scenarios.

The outer assessment sets are reserved exclusively for final model evaluation.
The inner folds are reserved for ANN hyperparameter tuning. True simulated
genetic values are never used to generate partitions or tune models.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

options(stringsAsFactors = FALSE)

find_project_root <- function(start = getwd()) {
  path <- normalizePath(start, winslash = "/", mustWork = TRUE)

  repeat {
    if (file.exists(file.path(path, "_workflowr.yml"))) {
      return(path)
    }

    parent <- dirname(path)

    if (identical(parent, path)) {
      stop("Could not locate workflowr project root.")
    }

    path <- parent
  }
}

PROJECT_ROOT <- find_project_root()

project_path <- function(...) {
  file.path(PROJECT_ROOT, ...)
}


SEED_GLOBAL <- 20260821L
N_OUTER_FOLDS <- 5L
N_OUTER_REPEATS <- 5L
N_INNER_FOLDS <- 4L

input_file <- project_path("output", "simulated", "audit", "preprocessed_simulated_common.rds")
audit_file <- project_path("output", "simulated", "audit", "simulated_audit_summary.csv")
output_dir <- project_path("output", "simulated", "validation")

dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

validation_log <- data.frame(
  Check = character(),
  Status = character(),
  Details = character(),
  stringsAsFactors = FALSE
)

add_check <- function(check, status, details) {
  stopifnot(status %in% c("PASS", "WARNING", "BLOCKED"))

  validation_log <<- rbind(
    validation_log,
    data.frame(
      Check = check,
      Status = status,
      Details = as.character(details),
      stringsAsFactors = FALSE
    )
  )

  invisible(NULL)
}

balanced_fold_assignment <- function(ids, k, seed) {
  stopifnot(
    length(ids) >= k,
    !anyDuplicated(ids),
    k >= 2L
  )

  set.seed(seed)

  shuffled_ids <- sample(
    ids,
    size = length(ids),
    replace = FALSE
  )

  assigned_folds <- rep(
    seq_len(k),
    length.out = length(ids)
  )

  out <- data.frame(
    ID = shuffled_ids,
    Fold = assigned_folds,
    stringsAsFactors = FALSE
  )

  out <- out[
    order(out$ID),
    ,
    drop = FALSE
  ]

  rownames(out) <- NULL
  out
}
```

## How to read the validation design

The validation scheme separates model development from final predictive
evaluation. An **analysis set** contains the individuals available for fitting
or tuning, whereas the corresponding **assessment set** remains held out for
evaluation at that level of cross-validation. Outer assessment individuals are
therefore never used to choose ANN hyperparameters or training duration.

Internal integrity checks still verify file availability, ID alignment,
partition sizes, disjointness, and coverage. Those checks remain executable and
are exported with the validation outputs, but the public page emphasizes the
validation logic that determines how models are trained and evaluated.

# Canonical data prerequisite

The validation design is built only after the simulated-data audit has
completed without any blocking check.

```{r prerequisites, include=FALSE}
required_files <- c(
  input_file,
  audit_file
)

files_exist <- file.exists(required_files)

add_check(
  "Canonical input files",
  if (all(files_exist)) "PASS" else "BLOCKED",
  if (all(files_exist)) {
    "Canonical simulated object and audit summary are present."
  } else {
    paste(
      "Missing:",
      paste(required_files[!files_exist], collapse = ", ")
    )
  }
)

if (!all(files_exist)) {
  stop("Canonical simulated-data inputs are missing.")
}

audit_summary <- read.csv(
  audit_file,
  stringsAsFactors = FALSE
)

n_blocked_audit <- sum(
  audit_summary$Status == "BLOCKED",
  na.rm = TRUE
)

add_check(
  "Simulated-data audit prerequisite",
  if (n_blocked_audit == 0L) "PASS" else "BLOCKED",
  paste(
    n_blocked_audit,
    "blocking check(s) in simulated-data audit."
  )
)

if (n_blocked_audit > 0L) {
  stop("Simulated-data audit contains blocking checks.")
}

simulated_common <- readRDS(input_file)

required_components <- c(
  "metadata",
  "phenotype",
  "true_genetic_value",
  "genotype_m101"
)

components_ok <- all(
  required_components %in% names(simulated_common)
)

add_check(
  "Canonical object components",
  if (components_ok) "PASS" else "BLOCKED",
  if (components_ok) {
    paste(
      "Required components present:",
      paste(required_components, collapse = ", ")
    )
  } else {
    paste(
      "Missing components:",
      paste(
        setdiff(required_components, names(simulated_common)),
        collapse = ", "
      )
    )
  }
)

if (!components_ok) {
  stop("Canonical simulated object is incomplete.")
}
```

# Individual universe

Partitions are generated from the 1,000 canonical row identifiers only. No
phenotype, true genetic value, scenario label, or model information is used to
define the partitions. Alignment with the canonical phenotype, true-genetic-
value, and genotype matrices is verified internally before the folds are built.

```{r individual-universe, include=FALSE}
ids <- simulated_common$metadata$row_identifier

if (is.null(ids)) {
  ids <- seq_len(
    simulated_common$metadata$n_individuals
  )
}

ids <- as.integer(ids)

individual_universe_ok <-
  length(ids) == 1000L &&
  !anyDuplicated(ids) &&
  identical(sort(ids), seq_len(1000L)) &&
  nrow(simulated_common$phenotype) == length(ids) &&
  nrow(simulated_common$true_genetic_value) == length(ids) &&
  nrow(simulated_common$genotype_m101) == length(ids)

add_check(
  "Individual universe",
  if (individual_universe_ok) "PASS" else "BLOCKED",
  paste(
    "Expected 1000 unique row identifiers aligned with phenotype,",
    "true genetic value, and genotype matrices."
  )
)

if (!individual_universe_ok) {
  stop("Canonical individual universe is inconsistent.")
}
```

# Outer repeated cross-validation

For each repetition, `sample()` randomly permutes the 1,000 canonical individual
IDs using a deterministic seed. `rep()` then assigns the shuffled IDs across
five fold labels as evenly as possible. Because 1,000 is exactly divisible by
five, every outer assessment fold contains 200 individuals and the remaining
800 form its outer analysis set.

The same procedure is repeated five times with different deterministic seeds,
producing 25 outer train--test comparisons while preserving exact reproducibility.

```{r outer-folds}
outer_assignments <- do.call(
  rbind,
  lapply(
    seq_len(N_OUTER_REPEATS),
    function(rep_id) {
      assignment <- balanced_fold_assignment(
        ids = ids,
        k = N_OUTER_FOLDS,
        seed = SEED_GLOBAL + rep_id
      )

      data.frame(
        Repetition = rep_id,
        ID = assignment$ID,
        OuterFold = assignment$Fold,
        stringsAsFactors = FALSE
      )
    }
  )
)

rownames(outer_assignments) <- NULL

outer_assignments$SplitID <- sprintf(
  "R%02d_F%02d",
  outer_assignments$Repetition,
  outer_assignments$OuterFold
)

outer_sizes <- aggregate(
  ID ~ Repetition + OuterFold + SplitID,
  data = outer_assignments,
  FUN = length
)

names(outer_sizes)[names(outer_sizes) == "ID"] <- "Assessment_N"
outer_sizes$Analysis_N <- length(ids) - outer_sizes$Assessment_N

knitr::kable(
  outer_sizes,
  caption = "Outer repeated cross-validation split sizes"
)

outer_row_count_ok <-
  nrow(outer_assignments) ==
    length(ids) * N_OUTER_REPEATS

add_check(
  "Outer assignment row count",
  if (outer_row_count_ok) "PASS" else "BLOCKED",
  paste(
    "Observed rows:",
    nrow(outer_assignments),
    "; expected:",
    length(ids) * N_OUTER_REPEATS
  )
)

outer_once_per_repeat <- all(
  table(
    outer_assignments$Repetition,
    outer_assignments$ID
  ) == 1L
)

add_check(
  "One outer assignment per ID and repetition",
  if (outer_once_per_repeat) "PASS" else "BLOCKED",
  "Each individual must occur exactly once in every repetition."
)

outer_balanced <- all(
  outer_sizes$Assessment_N == 200L &
    outer_sizes$Analysis_N == 800L
)

add_check(
  "Balanced outer folds",
  if (outer_balanced) "PASS" else "BLOCKED",
  "Every outer split must contain 800 analysis and 200 assessment individuals."
)

outer_patterns <- do.call(
  rbind,
  lapply(
    seq_len(N_OUTER_REPEATS),
    function(rep_id) {
      x <- outer_assignments[
        outer_assignments$Repetition == rep_id,
        c("ID", "OuterFold")
      ]

      x <- x[order(x$ID), ]
      x$OuterFold
    }
  )
)

unique_repetition_patterns <-
  !anyDuplicated(as.data.frame(outer_patterns))

add_check(
  "Distinct outer repetitions",
  if (unique_repetition_patterns) "PASS" else "WARNING",
  paste(
    "All",
    N_OUTER_REPEATS,
    "repetitions should use distinct fold assignments."
  )
)

assessment_frequency <- table(
  outer_assignments$ID
)

assessment_frequency_ok <- all(
  assessment_frequency == N_OUTER_REPEATS
)

add_check(
  "Outer assessment frequency",
  if (assessment_frequency_ok) "PASS" else "BLOCKED",
  paste(
    "Every individual is assigned to one assessment fold per repetition",
    "and therefore appears in exactly",
    N_OUTER_REPEATS,
    "outer assessment sets."
  )
)
```

# Explicit outer split objects

The assignment table is then converted into explicit analysis and assessment
ID sets for each outer split. For a given split, `setdiff()` removes the 200
assessment IDs from the 1,000-individual universe, leaving the corresponding
800-individual analysis set. This representation is the object consumed by the
modeling modules.

```{r outer-split-objects}
outer_split_ids <- unique(
  outer_assignments$SplitID
)

outer_split_ids <- outer_split_ids[
  order(outer_split_ids)
]

outer_splits <- setNames(
  vector("list", length(outer_split_ids)),
  outer_split_ids
)

for (split_id in outer_split_ids) {
  assessment_ids <- outer_assignments$ID[
    outer_assignments$SplitID == split_id
  ]

  analysis_ids <- setdiff(
    ids,
    assessment_ids
  )

  outer_splits[[split_id]] <- list(
    split_id = split_id,
    analysis_ids = sort(analysis_ids),
    assessment_ids = sort(assessment_ids)
  )
}

outer_split_integrity <- all(
  vapply(
    outer_splits,
    function(x) {
      length(x$analysis_ids) == 800L &&
        length(x$assessment_ids) == 200L &&
        length(intersect(
          x$analysis_ids,
          x$assessment_ids
        )) == 0L &&
        identical(
          sort(c(
            x$analysis_ids,
            x$assessment_ids
          )),
          sort(ids)
        )
    },
    logical(1)
  )
)

add_check(
  "Outer split integrity",
  if (outer_split_integrity) "PASS" else "BLOCKED",
  paste(
    length(outer_splits),
    "outer split objects checked for size, disjointness, and complete coverage."
  )
)
```

# Nested inner folds

Each outer analysis set is partitioned again for ANN hyperparameter selection.
Within a given outer split, one inner fold acts as the assessment subset and the
remaining inner individuals form the analysis subset. The operation `setdiff()`
constructs the analysis IDs by removing the current assessment IDs, which makes
the two sets disjoint by construction.

With 800 individuals in each outer analysis set and four inner folds, every
inner assessment set contains 200 individuals and every inner analysis set
contains 600.

```{r inner-folds}
inner_assignments_list <- vector(
  "list",
  length(outer_splits)
)

names(inner_assignments_list) <- names(outer_splits)

inner_splits <- setNames(
  vector("list", length(outer_splits)),
  names(outer_splits)
)

for (split_id in names(outer_splits)) {
  outer_split <- outer_splits[[split_id]]

  split_tokens <- strsplit(
    sub("^R", "", split_id),
    "_F"
  )[[1]]

  rep_id <- as.integer(split_tokens[1])
  outer_fold <- as.integer(split_tokens[2])

  inner_seed <-
    SEED_GLOBAL +
    100000L +
    rep_id * 100L +
    outer_fold

  assignment <- balanced_fold_assignment(
    ids = outer_split$analysis_ids,
    k = N_INNER_FOLDS,
    seed = inner_seed
  )

  inner_assignments_list[[split_id]] <- data.frame(
    SplitID = split_id,
    Repetition = rep_id,
    OuterFold = outer_fold,
    ID = assignment$ID,
    InnerFold = assignment$Fold,
    stringsAsFactors = FALSE
  )

  current_inner_splits <- setNames(
    vector("list", N_INNER_FOLDS),
    paste0("I", seq_len(N_INNER_FOLDS))
  )

  for (inner_fold in seq_len(N_INNER_FOLDS)) {
    inner_assessment_ids <- assignment$ID[
      assignment$Fold == inner_fold
    ]

    inner_analysis_ids <- setdiff(
      outer_split$analysis_ids,
      inner_assessment_ids
    )

    current_inner_splits[[paste0("I", inner_fold)]] <- list(
      split_id = split_id,
      inner_fold = inner_fold,
      analysis_ids = sort(inner_analysis_ids),
      assessment_ids = sort(inner_assessment_ids)
    )
  }

  inner_splits[[split_id]] <- current_inner_splits
}

inner_assignments <- do.call(
  rbind,
  inner_assignments_list
)

rownames(inner_assignments) <- NULL

inner_sizes <- aggregate(
  ID ~ SplitID + Repetition + OuterFold + InnerFold,
  data = inner_assignments,
  FUN = length
)

names(inner_sizes)[names(inner_sizes) == "ID"] <- "Assessment_N"
inner_sizes$Analysis_N <- 800L - inner_sizes$Assessment_N

knitr::kable(
  head(inner_sizes, 20),
  caption = "First 20 nested inner split sizes"
)

inner_row_count_ok <-
  nrow(inner_assignments) ==
    length(outer_splits) * 800L

add_check(
  "Inner assignment row count",
  if (inner_row_count_ok) "PASS" else "BLOCKED",
  paste(
    "Observed rows:",
    nrow(inner_assignments),
    "; expected:",
    length(outer_splits) * 800L
  )
)

inner_balanced <- all(
  inner_sizes$Assessment_N == 200L &
    inner_sizes$Analysis_N == 600L
)

add_check(
  "Balanced inner folds",
  if (inner_balanced) "PASS" else "BLOCKED",
  "Every inner split must contain 600 analysis and 200 assessment individuals."
)

inner_assignment_integrity <- all(
  vapply(
    names(outer_splits),
    function(split_id) {
      x <- inner_assignments[
        inner_assignments$SplitID == split_id,
      ]

      outer_analysis_ids <-
        outer_splits[[split_id]]$analysis_ids

      nrow(x) == 800L &&
        !anyDuplicated(x$ID) &&
        identical(
          sort(x$ID),
          sort(outer_analysis_ids)
        ) &&
        all(sort(unique(x$InnerFold)) ==
              seq_len(N_INNER_FOLDS))
    },
    logical(1)
  )
)

add_check(
  "Inner assignment integrity",
  if (inner_assignment_integrity) "PASS" else "BLOCKED",
  "Inner assignments are restricted to the corresponding outer analysis set."
)

inner_split_integrity <- all(
  vapply(
    names(inner_splits),
    function(split_id) {
      outer_analysis_ids <-
        outer_splits[[split_id]]$analysis_ids

      all(
        vapply(
          inner_splits[[split_id]],
          function(x) {
            length(x$analysis_ids) == 600L &&
              length(x$assessment_ids) == 200L &&
              length(intersect(
                x$analysis_ids,
                x$assessment_ids
              )) == 0L &&
              identical(
                sort(c(
                  x$analysis_ids,
                  x$assessment_ids
                )),
                sort(outer_analysis_ids)
              )
          },
          logical(1)
        )
      )
    },
    logical(1)
  )
)

add_check(
  "Inner split integrity",
  if (inner_split_integrity) "PASS" else "BLOCKED",
  paste(
    length(outer_splits) * N_INNER_FOLDS,
    "inner splits checked for size, disjointness, and complete outer-analysis coverage."
  )
)
```

# Scenario and model independence

The split object contains no scenario, phenotype, true genetic value, or model
information. The same partitions are therefore reused for C1-C6 and for every
prediction method. This independence is verified internally from the fields
stored in the frozen split objects.

```{r independence-contract, include=FALSE}
split_variables <- unique(
  c(
    names(outer_assignments),
    names(inner_assignments)
  )
)

forbidden_split_fields <- c(
  "Scenario",
  "Phenotype",
  "Observed",
  "TrueGeneticValue",
  "Model"
)

split_independence_ok <- !any(
  forbidden_split_fields %in% split_variables
)

add_check(
  "Scenario-independent partitions",
  if (split_independence_ok) "PASS" else "BLOCKED",
  paste(
    "Partitions are generated from canonical individual IDs and deterministic",
    "seeds only; the same folds are reused for C1-C6 and all models."
  )
)

add_check(
  "True genetic value excluded from partitioning",
  "PASS",
  paste(
    "True simulated genetic values are stored in the canonical data object",
    "for final evaluation only and are not referenced when generating folds."
  )
)

add_check(
  "Phenotype excluded from partitioning",
  "PASS",
  paste(
    "Phenotypes are not used to construct folds. They may be used later",
    "inside training data and inner-CV tuning, never to redefine partitions."
  )
)
```

# Validation contract

The frozen validation protocol is:

1. For each of the 5 repetitions, every individual belongs to exactly one of
   the 5 outer assessment folds.
2. The same 25 outer splits are used for C1-C6 and for GBLUP-ADE, ANN-0,
   ANN-L1, ANN-L2, and ANN-L1+L2.
3. Outer assessment phenotypes and true genetic values are not used for model
   fitting or hyperparameter tuning.
4. ANN hyperparameters are selected using the 4 inner folds defined within the
   corresponding outer analysis set.
5. Early stopping, if used during inner ANN fitting, must use data drawn only
   from the current inner analysis set. The inner assessment set must remain
   reserved for candidate evaluation.
6. After ANN hyperparameters are selected, the final outer model must be fitted
   using only the outer analysis individuals. The outer assessment individuals
   remain untouched until final prediction.
7. Simulated true genetic values are never used to tune ANN architecture,
   learning rate, L1, L2, epoch count, or any other model choice.
8. Primary simulated-data evaluation compares final predictions with the true
   genetic value only after the model for that outer split has been frozen.
9. Secondary phenotype-prediction metrics may be computed on the same outer
   assessment individuals, but cannot feed back into model selection.

# Design summary

The complete nested design can now be summarized independently of any
prediction model. These counts define the sample sizes and number of repeated
comparisons used by both modeling branches.

```{r design-summary}
design_summary <- data.frame(
  Item = c(
    "Individuals",
    "Scenarios",
    "Outer repetitions",
    "Outer folds per repetition",
    "Total outer splits",
    "Outer analysis N",
    "Outer assessment N",
    "Inner folds per outer split",
    "Total inner splits",
    "Inner analysis N",
    "Inner assessment N",
    "Global seed"
  ),
  Value = c(
    length(ids),
    simulated_common$metadata$n_scenarios,
    N_OUTER_REPEATS,
    N_OUTER_FOLDS,
    N_OUTER_REPEATS * N_OUTER_FOLDS,
    800,
    200,
    N_INNER_FOLDS,
    N_OUTER_REPEATS * N_OUTER_FOLDS * N_INNER_FOLDS,
    600,
    200,
    SEED_GLOBAL
  )
)

knitr::kable(
  design_summary,
  caption = "Frozen simulated-data validation design"
)
```

# Internal validation status

The detailed integrity checklist is evaluated before the frozen design is
written, but it is retained as a validation artifact rather than reproduced as
a reader-facing table.

```{r final-validation-audit, include=FALSE}
status_order <- c(
  "BLOCKED",
  "WARNING",
  "PASS"
)

validation_log$Status <- factor(
  validation_log$Status,
  levels = status_order
)

validation_log <- validation_log[
  order(validation_log$Status),
  ,
  drop = FALSE
]

rownames(validation_log) <- NULL

knitr::kable(
  validation_log,
  caption = "Validation-design integrity checks"
)

validation_counts <- as.data.frame(
  table(validation_log$Status)
)

names(validation_counts) <- c(
  "Status",
  "Checks"
)

knitr::kable(
  validation_counts,
  caption = "Validation-design status counts"
)

blocked_checks <- sum(
  validation_log$Status == "BLOCKED"
)

warning_checks <- sum(
  validation_log$Status == "WARNING"
)

final_classification <- if (blocked_checks > 0L) {
  "BLOCKED"
} else if (warning_checks > 0L) {
  "PASS WITH WARNINGS"
} else {
  "PASS"
}

cat(
  "\nFinal validation-design classification:",
  final_classification,
  "\n"
)
```

The internal checks classify the frozen validation design as
**`r final_classification`**. A blocking condition would prevent creation of the
canonical cross-validation object.

# Frozen validation object

The validated design is stored only after all blocking conditions have been
excluded. The saved object contains the outer and inner assignments together
with their explicit analysis/assessment ID sets and is the single validation
contract consumed by GBLUP-ADE and ANN.

```{r save-validation-design, include=FALSE}
write.csv(
  validation_log,
  file.path(
    output_dir,
    "simulated_cv_design_audit.csv"
  ),
  row.names = FALSE
)

write.csv(
  outer_assignments,
  file.path(
    output_dir,
    "simulated_cv_outer_assignments.csv"
  ),
  row.names = FALSE
)

write.csv(
  inner_assignments,
  file.path(
    output_dir,
    "simulated_cv_inner_assignments.csv"
  ),
  row.names = FALSE
)

write.csv(
  design_summary,
  file.path(
    output_dir,
    "simulated_cv_design_summary.csv"
  ),
  row.names = FALSE
)

if (blocked_checks > 0L) {
  stop(
    paste(
      blocked_checks,
      "blocking validation-design check(s) detected.",
      "The canonical CV object was not created."
    )
  )
}

cv_design <- list(
  metadata = list(
    dataset = "simulated",
    seed_global = SEED_GLOBAL,
    n_individuals = length(ids),
    n_scenarios = simulated_common$metadata$n_scenarios,
    outer_repetitions = N_OUTER_REPEATS,
    outer_folds = N_OUTER_FOLDS,
    inner_folds = N_INNER_FOLDS,
    outer_analysis_n = 800L,
    outer_assessment_n = 200L,
    inner_analysis_n = 600L,
    inner_assessment_n = 200L,
    scenario_independent = TRUE,
    model_independent = TRUE,
    phenotype_used_for_partitioning = FALSE,
    true_genetic_value_used_for_partitioning = FALSE
  ),
  outer_assignments = outer_assignments,
  outer_splits = outer_splits,
  inner_assignments = inner_assignments,
  inner_splits = inner_splits,
  audit = validation_log
)

saveRDS(
  cv_design,
  file.path(
    output_dir,
    "simulated_cv_design.rds"
  ),
  compress = "gzip"
)

cat(
  "Frozen validation design written to:",
  file.path(
    output_dir,
    "simulated_cv_design.rds"
  ),
  "\n"
)
```

The detailed assignment tables and validation audit remain available as
versioned outputs for reproducibility and maintenance, while the modeling
modules consume `simulated_cv_design.rds` directly.

# Navigation

[Previous: Data audit](simulated_data_audit.html) | [GBLUP-ADE](simulated_data_gblup.html) | [Regularized ANN](simulated_data_ann.html)

# Reproducibility information

```{r session-info}
sessionInfo()
```
