---
title: "Simulated data validation design"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

# Purpose

This analysis freezes the validation design used by all genomic prediction
models fitted to the simulated dataset.

The design is created **before model fitting** and is independent of phenotype,
true genetic value, scenario, and model. Consequently, GBLUP-ADE and all ANN
variants are evaluated on exactly the same outer assessment individuals.

The canonical design uses:

- 5 outer folds;
- 5 outer repetitions;
- 4 inner folds within each outer analysis set;
- deterministic seeds derived from the project-wide seed;
- the same outer and inner partitions for all six simulation scenarios.

The outer assessment sets are reserved exclusively for final model evaluation.
The inner folds are reserved for ANN hyperparameter tuning. True simulated
genetic values are never used to generate partitions or tune models.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
options(stringsAsFactors = FALSE)
PROJECT_ROOT <- "."
project_path <- function(...) file.path(PROJECT_ROOT, ...)
input_file <- project_path(
  "output", "simulated", "audit", "preprocessed_simulated_common.rds"
)
output_dir <- project_path("output", "simulated", "validation")
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
```

## How to read the validation design

The validation scheme separates model development from final predictive
evaluation. An **analysis set** contains the individuals available for fitting
or tuning, whereas the corresponding **assessment set** remains held out for
evaluation at that level of cross-validation. Outer assessment individuals are
therefore never used to choose ANN hyperparameters or training duration.

The sections below show the individual universe, partition sizes, overlap, and
coverage directly. The validation structure can therefore be inspected from the
displayed objects rather than inferred from boolean checks.

# Validation parameters

The complete nested design is controlled by four explicit values.

```{r validation-parameters}
SEED_GLOBAL <- 20260821L
N_OUTER_REPEATS <- 5L
N_OUTER_FOLDS <- 5L
N_INNER_FOLDS <- 4L

validation_parameters <- data.frame(
  Parameter = c(
    "Global seed",
    "Outer repetitions",
    "Outer folds per repetition",
    "Inner folds per outer split"
  ),
  Value = c(
    SEED_GLOBAL,
    N_OUTER_REPEATS,
    N_OUTER_FOLDS,
    N_INNER_FOLDS
  ),
  Role = c(
    "reproducible partition generation",
    "repeat held-out evaluation",
    "define outer assessment sets",
    "compare ANN candidates inside outer analysis"
  )
)

knitr::kable(
  validation_parameters,
  caption =
    "Canonical nested cross-validation parameters",
  row.names = FALSE
)
```

The same global seed is combined with the repetition and fold identifiers so
every partition can be regenerated exactly.

## Balanced fold assignment

The helper below randomizes the supplied IDs with `sample()` and then cycles
through the fold labels with `rep()`. Sorting by ID at the end changes only the
display/order of the assignment table, not the randomized fold membership.

```{r balanced-fold-assignment}
balanced_fold_assignment <- function(
  ids,
  k,
  seed
) {
  set.seed(seed)

  shuffled_ids <- sample(
    ids,
    size = length(ids),
    replace = FALSE
  )

  assigned_folds <- rep(
    seq_len(k),
    length.out = length(ids)
  )

  out <- data.frame(
    ID = shuffled_ids,
    Fold = assigned_folds,
    stringsAsFactors = FALSE
  )

  out <- out[
    order(out$ID),
    ,
    drop = FALSE
  ]

  rownames(out) <- NULL
  out
}
```

For 1,000 individuals and five outer folds, this produces 200 assessment
individuals per fold. Applied to an 800-individual outer-analysis set with four
inner folds, it produces 200 inner-assessment individuals per fold.

# Canonical data prerequisite

The validation design is built from the canonical simulated-data object created
in the preceding module.

```{r prerequisites}
simulated_common <- readRDS(input_file)
canonical_components <- data.frame(
  Component = c("phenotype", "true_genetic_value", "genotype_m101"),
  Rows = c(
    nrow(simulated_common$phenotype),
    nrow(simulated_common$true_genetic_value),
    nrow(simulated_common$genotype_m101)
  ),
  Columns = c(
    ncol(simulated_common$phenotype),
    ncol(simulated_common$true_genetic_value),
    ncol(simulated_common$genotype_m101)
  )
)
knitr::kable(
  canonical_components,
  caption = "Canonical objects used to construct the validation design",
  row.names = FALSE
)
```

# Individual universe

Partitions are generated from the 1,000 canonical row identifiers only. No
phenotype, true genetic value, scenario label, or model information is used to
define the partitions. Alignment with the canonical phenotype, true-genetic-
value, and genotype matrices is verified internally before the folds are built.

```{r individual-universe}
ids <- as.integer(simulated_common$metadata$row_identifier)
individual_universe <- data.frame(
  Quantity = c(
    "Individuals used to generate folds",
    "First row identifier",
    "Last row identifier"
  ),
  Observed = c(length(ids), min(ids), max(ids))
)
knitr::kable(
  individual_universe,
  caption = "Individual universe used for fold generation",
  row.names = FALSE
)
```

# Outer repeated cross-validation

For each repetition, `sample()` randomly permutes the 1,000 canonical individual
IDs using a deterministic seed. `rep()` then assigns the shuffled IDs across
five fold labels as evenly as possible. Because 1,000 is exactly divisible by
five, every outer assessment fold contains 200 individuals and the remaining
800 form its outer analysis set.

The same procedure is repeated five times with different deterministic seeds,
producing 25 outer train--test comparisons while preserving exact reproducibility.

```{r outer-folds}
outer_assignments <- do.call(
  rbind,
  lapply(seq_len(N_OUTER_REPEATS), function(rep_id) {
    assignment <- balanced_fold_assignment(
      ids = ids,
      k = N_OUTER_FOLDS,
      seed = SEED_GLOBAL + rep_id
    )
    data.frame(
      Repetition = rep_id,
      ID = assignment$ID,
      OuterFold = assignment$Fold,
      stringsAsFactors = FALSE
    )
  })
)
rownames(outer_assignments) <- NULL
outer_assignments$SplitID <- sprintf(
  "R%02d_F%02d",
  outer_assignments$Repetition,
  outer_assignments$OuterFold
)

outer_sizes <- aggregate(
  ID ~ Repetition + OuterFold + SplitID,
  data = outer_assignments,
  FUN = length
)
names(outer_sizes)[names(outer_sizes) == "ID"] <- "Assessment_N"
outer_sizes$Analysis_N <- length(ids) - outer_sizes$Assessment_N

knitr::kable(
  outer_sizes,
  caption = "Outer repeated cross-validation split sizes",
  row.names = FALSE
)
```

Every one of the 25 outer splits therefore contains **800 analysis** and
**200 assessment** individuals. The five repetitions use distinct randomized
assignments, while every individual appears exactly once in an assessment fold
within each repetition.

# Explicit outer split objects

The assignment table is then converted into explicit analysis and assessment
ID sets for each outer split. For a given split, `setdiff()` removes the 200
assessment IDs from the 1,000-individual universe, leaving the corresponding
800-individual analysis set. This representation is the object consumed by the
modeling modules.

```{r outer-split-objects}
outer_split_ids <- sort(unique(outer_assignments$SplitID))
outer_splits <- setNames(
  vector("list", length(outer_split_ids)),
  outer_split_ids
)

for (split_id in outer_split_ids) {
  assessment_ids <- outer_assignments$ID[
    outer_assignments$SplitID == split_id
  ]
  analysis_ids <- setdiff(ids, assessment_ids)
  outer_splits[[split_id]] <- list(
    split_id = split_id,
    analysis_ids = sort(analysis_ids),
    assessment_ids = sort(assessment_ids)
  )
}

example_outer_split <- outer_splits[[1]]
outer_object_summary <- data.frame(
  Component = c("analysis_ids", "assessment_ids", "overlap"),
  N = c(
    length(example_outer_split$analysis_ids),
    length(example_outer_split$assessment_ids),
    length(intersect(
      example_outer_split$analysis_ids,
      example_outer_split$assessment_ids
    ))
  ),
  First_IDs = c(
    paste(head(example_outer_split$analysis_ids, 6), collapse = ", "),
    paste(head(example_outer_split$assessment_ids, 6), collapse = ", "),
    "-"
  )
)
knitr::kable(
  outer_object_summary,
  caption = paste("Structure of outer split", example_outer_split$split_id),
  row.names = FALSE
)
```

Each explicit split object therefore stores the 800 IDs available to model
development and the 200 IDs reserved for final assessment. `setdiff()` makes
the two sets disjoint while preserving complete coverage of the 1,000
individuals.

# Nested inner folds

Each outer analysis set is partitioned again for ANN hyperparameter selection.
Within a given outer split, one inner fold acts as the assessment subset and the
remaining inner individuals form the analysis subset. The operation `setdiff()`
constructs the analysis IDs by removing the current assessment IDs, which makes
the two sets disjoint by construction.

With 800 individuals in each outer analysis set and four inner folds, every
inner assessment set contains 200 individuals and every inner analysis set
contains 600.

```{r inner-folds}
inner_assignments_list <- vector("list", length(outer_splits))
names(inner_assignments_list) <- names(outer_splits)
inner_splits <- setNames(
  vector("list", length(outer_splits)),
  names(outer_splits)
)

for (split_id in names(outer_splits)) {
  outer_split <- outer_splits[[split_id]]
  split_tokens <- strsplit(sub("^R", "", split_id), "_F")[[1]]
  rep_id <- as.integer(split_tokens[1])
  outer_fold <- as.integer(split_tokens[2])
  inner_seed <- SEED_GLOBAL + 100000L + rep_id * 100L + outer_fold

  assignment <- balanced_fold_assignment(
    ids = outer_split$analysis_ids,
    k = N_INNER_FOLDS,
    seed = inner_seed
  )

  inner_assignments_list[[split_id]] <- data.frame(
    SplitID = split_id,
    Repetition = rep_id,
    OuterFold = outer_fold,
    ID = assignment$ID,
    InnerFold = assignment$Fold,
    stringsAsFactors = FALSE
  )

  current_inner_splits <- setNames(
    vector("list", N_INNER_FOLDS),
    paste0("I", seq_len(N_INNER_FOLDS))
  )

  for (inner_fold in seq_len(N_INNER_FOLDS)) {
    inner_assessment_ids <- assignment$ID[assignment$Fold == inner_fold]
    inner_analysis_ids <- setdiff(
      outer_split$analysis_ids,
      inner_assessment_ids
    )
    current_inner_splits[[paste0("I", inner_fold)]] <- list(
      split_id = split_id,
      inner_fold = inner_fold,
      analysis_ids = sort(inner_analysis_ids),
      assessment_ids = sort(inner_assessment_ids)
    )
  }
  inner_splits[[split_id]] <- current_inner_splits
}

inner_assignments <- do.call(rbind, inner_assignments_list)
rownames(inner_assignments) <- NULL
inner_sizes <- aggregate(
  ID ~ SplitID + Repetition + OuterFold + InnerFold,
  data = inner_assignments,
  FUN = length
)
names(inner_sizes)[names(inner_sizes) == "ID"] <- "Assessment_N"
inner_sizes$Analysis_N <- 800L - inner_sizes$Assessment_N

knitr::kable(
  head(inner_sizes, 20),
  caption = "First 20 nested inner split sizes",
  row.names = FALSE
)
```

Every inner split contains **600 analysis** and **200 assessment** individuals.
The 200 inner-assessment individuals are used only to compare ANN candidates
inside the corresponding 800-individual outer-analysis set; they are not part
of the early-validation subset used to train a candidate.

# Scenario and model independence

The split object contains no scenario, phenotype, true genetic value, or model
information. The same partitions are therefore reused for C1-C6 and for every
prediction method. The table below makes the fold-generation inputs explicit.

```{r independence-contract}
partition_information <- data.frame(
  Information = c(
    "Individual ID", "Deterministic seed", "Scenario",
    "Phenotype", "True genetic value", "Model"
  ),
  Used_to_generate_folds = c(TRUE, TRUE, FALSE, FALSE, FALSE, FALSE)
)
knitr::kable(
  partition_information,
  caption = "Information used to generate the frozen validation partitions",
  row.names = FALSE
)
```

No response variable participates in fold generation. This is the core reason
the same frozen partitions can be reused fairly by GBLUP-ADE and all four ANN
variants.

# Validation contract

## Outer evaluation

- Five repetitions of five outer folds produce **25 outer splits**.
- Each split contains 800 analysis and 200 assessment individuals.
- Outer assessment data remain untouched until the final prediction step.

## Inner ANN tuning

- Each 800-individual outer-analysis set is divided into four inner folds.
- Each inner split contains 600 analysis and 200 inner-assessment individuals.
- ANN architecture, learning rate, L1, and L2 are selected only within these
  inner splits; early validation is drawn only from the current inner-analysis
  subset.

## Final evaluation

- The same outer splits are used for C1-C6 and every prediction method.
- True genetic values never enter partitioning, tuning, or epoch selection.
- Primary true-G and secondary phenotype metrics are calculated only after the
  final model has predicted the untouched outer-assessment individuals.

# Design summary

The complete nested design can now be summarized independently of any
prediction model. These counts define the sample sizes and number of repeated
comparisons used by both modeling branches.

```{r design-summary}
design_summary <- data.frame(
  Item = c(
    "Individuals",
    "Scenarios",
    "Outer repetitions",
    "Outer folds per repetition",
    "Total outer splits",
    "Outer analysis N",
    "Outer assessment N",
    "Inner folds per outer split",
    "Total inner splits",
    "Inner analysis N",
    "Inner assessment N",
    "Global seed"
  ),
  Value = c(
    length(ids),
    simulated_common$metadata$n_scenarios,
    N_OUTER_REPEATS,
    N_OUTER_FOLDS,
    N_OUTER_REPEATS * N_OUTER_FOLDS,
    800,
    200,
    N_INNER_FOLDS,
    N_OUTER_REPEATS * N_OUTER_FOLDS * N_INNER_FOLDS,
    600,
    200,
    SEED_GLOBAL
  )
)

knitr::kable(
  design_summary,
  caption = "Frozen simulated-data validation design",
  row.names = FALSE
)
```

The frozen object therefore contains 25 outer evaluations and
`r N_OUTER_REPEATS * N_OUTER_FOLDS * N_INNER_FOLDS` inner split definitions.
These partitions are generated once and reused across all six scenarios and all
prediction methods.


```{r validation-output-contract, include=FALSE}
compat_check_status <- function(condition, failure = "BLOCKED", success = "PASS") {
  c(failure, success)[1L + as.integer(isTRUE(condition))]
}

validation_log <- data.frame(
  Check = character(), Status = character(), Details = character(),
  stringsAsFactors = FALSE
)
add_compat_check <- function(check, status, details) {
  validation_log <<- rbind(
    validation_log,
    data.frame(Check = check, Status = status, Details = as.character(details), stringsAsFactors = FALSE)
  )
  invisible(NULL)
}

audit_file <- project_path("output", "simulated", "audit", "simulated_audit_summary.csv")
required_files <- c(input_file, audit_file)
files_exist <- file.exists(required_files)
audit_summary <- read.csv(audit_file, stringsAsFactors = FALSE)
n_blocked_audit <- sum(audit_summary$Status == "BLOCKED", na.rm = TRUE)
required_components <- c("metadata", "phenotype", "true_genetic_value", "genotype_m101")
components_ok <- all(required_components %in% names(simulated_common))
individual_universe_ok <-
  length(ids) == 1000L && !anyDuplicated(ids) &&
  identical(sort(ids), seq_len(1000L)) &&
  nrow(simulated_common$phenotype) == length(ids) &&
  nrow(simulated_common$true_genetic_value) == length(ids) &&
  nrow(simulated_common$genotype_m101) == length(ids)

outer_row_count_ok <- nrow(outer_assignments) == length(ids) * N_OUTER_REPEATS
outer_once_per_repeat <- all(table(outer_assignments$Repetition, outer_assignments$ID) == 1L)
outer_balanced <- all(outer_sizes$Assessment_N == 200L & outer_sizes$Analysis_N == 800L)
outer_patterns <- do.call(rbind, lapply(seq_len(N_OUTER_REPEATS), function(rep_id) {
  x <- outer_assignments[outer_assignments$Repetition == rep_id, c("ID", "OuterFold")]
  x <- x[order(x$ID), ]
  x$OuterFold
}))
unique_repetition_patterns <- !anyDuplicated(as.data.frame(outer_patterns))
assessment_frequency_ok <- all(table(outer_assignments$ID) == N_OUTER_REPEATS)
outer_split_integrity <- all(vapply(outer_splits, function(x) {
  length(x$analysis_ids) == 800L && length(x$assessment_ids) == 200L &&
  length(intersect(x$analysis_ids, x$assessment_ids)) == 0L &&
  identical(sort(c(x$analysis_ids, x$assessment_ids)), sort(ids))
}, logical(1)))

inner_row_count_ok <- nrow(inner_assignments) == length(outer_splits) * 800L
inner_balanced <- all(inner_sizes$Assessment_N == 200L & inner_sizes$Analysis_N == 600L)
inner_assignment_integrity <- all(vapply(names(outer_splits), function(split_id) {
  x <- inner_assignments[inner_assignments$SplitID == split_id, ]
  outer_analysis_ids <- outer_splits[[split_id]]$analysis_ids
  nrow(x) == 800L && !anyDuplicated(x$ID) &&
  identical(sort(x$ID), sort(outer_analysis_ids)) &&
  identical(sort(unique(x$InnerFold)), seq_len(N_INNER_FOLDS))
}, logical(1)))
inner_split_integrity <- all(vapply(names(inner_splits), function(split_id) {
  outer_analysis_ids <- outer_splits[[split_id]]$analysis_ids
  all(vapply(inner_splits[[split_id]], function(x) {
    length(x$analysis_ids) == 600L && length(x$assessment_ids) == 200L &&
    length(intersect(x$analysis_ids, x$assessment_ids)) == 0L &&
    identical(sort(c(x$analysis_ids, x$assessment_ids)), sort(outer_analysis_ids))
  }, logical(1)))
}, logical(1)))
split_variables <- unique(c(names(outer_assignments), names(inner_assignments)))
forbidden_split_fields <- c("Scenario", "Phenotype", "Observed", "TrueGeneticValue", "Model")
split_independence_ok <- !any(forbidden_split_fields %in% split_variables)

add_compat_check("Canonical input files", compat_check_status(all(files_exist)), if (all(files_exist)) "Canonical simulated object and audit summary are present." else paste("Missing:", paste(required_files[!files_exist], collapse = ", ")))
add_compat_check("Simulated-data audit prerequisite", compat_check_status(n_blocked_audit == 0L), paste(n_blocked_audit, "blocking check(s) in simulated-data audit."))
add_compat_check("Canonical object components", compat_check_status(components_ok), if (components_ok) paste("Required components present:", paste(required_components, collapse = ", ")) else paste("Missing components:", paste(setdiff(required_components, names(simulated_common)), collapse = ", ")))
add_compat_check("Individual universe", compat_check_status(individual_universe_ok), "Expected 1000 unique row identifiers aligned with phenotype, true genetic value, and genotype matrices.")
add_compat_check("Outer assignment row count", compat_check_status(outer_row_count_ok), paste("Observed rows:", nrow(outer_assignments), "; expected:", length(ids) * N_OUTER_REPEATS))
add_compat_check("One outer assignment per ID and repetition", compat_check_status(outer_once_per_repeat), "Each individual must occur exactly once in every repetition.")
add_compat_check("Balanced outer folds", compat_check_status(outer_balanced), "Every outer split must contain 800 analysis and 200 assessment individuals.")
add_compat_check("Distinct outer repetitions", compat_check_status(unique_repetition_patterns, failure = "WARNING"), paste("All", N_OUTER_REPEATS, "repetitions should use distinct fold assignments."))
add_compat_check("Outer assessment frequency", compat_check_status(assessment_frequency_ok), paste("Every individual is assigned to one assessment fold per repetition and therefore appears in exactly", N_OUTER_REPEATS, "outer assessment sets."))
add_compat_check("Outer split integrity", compat_check_status(outer_split_integrity), paste(length(outer_splits), "outer split objects checked for size, disjointness, and complete coverage."))
add_compat_check("Inner assignment row count", compat_check_status(inner_row_count_ok), paste("Observed rows:", nrow(inner_assignments), "; expected:", length(outer_splits) * 800L))
add_compat_check("Balanced inner folds", compat_check_status(inner_balanced), "Every inner split must contain 600 analysis and 200 assessment individuals.")
add_compat_check("Inner assignment integrity", compat_check_status(inner_assignment_integrity), "Inner assignments are restricted to the corresponding outer analysis set.")
add_compat_check("Inner split integrity", compat_check_status(inner_split_integrity), paste(length(outer_splits) * N_INNER_FOLDS, "inner splits checked for size, disjointness, and complete outer-analysis coverage."))
add_compat_check("Scenario-independent partitions", compat_check_status(split_independence_ok), "Partitions are generated from canonical individual IDs and deterministic seeds only; the same folds are reused for C1-C6 and all models.")
add_compat_check("True genetic value excluded from partitioning", "PASS", "True simulated genetic values are stored in the canonical data object for final evaluation only and are not referenced when generating folds.")
add_compat_check("Phenotype excluded from partitioning", "PASS", "Phenotypes are not used to construct folds. They may be used later inside training data and inner-CV tuning, never to redefine partitions.")

validation_log$Status <- factor(validation_log$Status, levels = c("BLOCKED", "WARNING", "PASS"))
validation_log <- validation_log[order(validation_log$Status), , drop = FALSE]
rownames(validation_log) <- NULL
```

# Observed validation structure

The detailed checklist is retained in the exported validation artifact. The
tutorial summarizes the status counts and final classification before saving the
frozen design.

```{r final-validation-audit}
outer_overlap <- max(vapply(
  outer_splits,
  function(x) length(intersect(x$analysis_ids, x$assessment_ids)),
  integer(1)
))
inner_overlap <- max(unlist(lapply(
  inner_splits,
  function(split) vapply(
    split,
    function(x) length(intersect(x$analysis_ids, x$assessment_ids)),
    integer(1)
  )
)))

validation_observed <- data.frame(
  Quantity = c(
    "Outer assignment rows", "Outer splits", "Outer analysis N",
    "Outer assessment N", "Maximum outer overlap",
    "Inner assignment rows", "Inner splits", "Inner analysis N",
    "Inner assessment N", "Maximum inner overlap"
  ),
  Observed = c(
    nrow(outer_assignments),
    length(outer_splits),
    unique(outer_sizes$Analysis_N),
    unique(outer_sizes$Assessment_N),
    outer_overlap,
    nrow(inner_assignments),
    sum(vapply(inner_splits, length, integer(1))),
    unique(inner_sizes$Analysis_N),
    unique(inner_sizes$Assessment_N),
    inner_overlap
  )
)
knitr::kable(
  validation_observed,
  caption = "Observed structure of the frozen nested cross-validation design",
  row.names = FALSE
)
```

The table above shows the sizes and overlap of the frozen outer and inner
partitions before they are saved for the modeling modules.

# Frozen validation object

The design is stored after its observed partition structure has been displayed. The saved object contains the outer and inner assignments together
with their explicit analysis/assessment ID sets and is the single validation
contract consumed by GBLUP-ADE and ANN.

```{r save-validation-design}
write.csv(
  validation_log,
  file.path(output_dir, "simulated_cv_design_audit.csv"),
  row.names = FALSE
)
write.csv(
  outer_assignments,
  file.path(output_dir, "simulated_cv_outer_assignments.csv"),
  row.names = FALSE
)
write.csv(
  inner_assignments,
  file.path(output_dir, "simulated_cv_inner_assignments.csv"),
  row.names = FALSE
)
write.csv(
  design_summary,
  file.path(output_dir, "simulated_cv_design_summary.csv"),
  row.names = FALSE
)

cv_design <- list(
  metadata = list(
    dataset = "simulated",
    seed_global = SEED_GLOBAL,
    n_individuals = length(ids),
    n_scenarios = simulated_common$metadata$n_scenarios,
    outer_repetitions = N_OUTER_REPEATS,
    outer_folds = N_OUTER_FOLDS,
    inner_folds = N_INNER_FOLDS,
    outer_analysis_n = 800L,
    outer_assessment_n = 200L,
    inner_analysis_n = 600L,
    inner_assessment_n = 200L,
    scenario_independent = TRUE,
    model_independent = TRUE,
    phenotype_used_for_partitioning = FALSE,
    true_genetic_value_used_for_partitioning = FALSE
  ),
  outer_assignments = outer_assignments,
  outer_splits = outer_splits,
  inner_assignments = inner_assignments,
  inner_splits = inner_splits,
  audit = validation_log
)

saveRDS(
  cv_design,
  file.path(output_dir, "simulated_cv_design.rds"),
  compress = "gzip"
)
cat(
  "Frozen validation design written to:",
  file.path(output_dir, "simulated_cv_design.rds"),
  "
"
)
```

The detailed assignment tables and validation audit remain available as
versioned outputs for reproducibility and maintenance, while the modeling
modules consume `simulated_cv_design.rds` directly.

# Navigation

[Previous: Data audit](simulated_data_audit.html) | [GBLUP-ADE](simulated_data_gblup.html) | [Regularized ANN](simulated_data_ann.html)

# Reproducibility information

`sessionInfo()` records the R version, platform, and loaded packages used to
render this tutorial.

```{r session-info}
sessionInfo()
```
