---
title: "GBLUP-ADE for simulated genomic data"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

## 1. Objective

This module fits the canonical GBLUP-ADE benchmark to all six simulated
scenarios using the 25 frozen outer validation splits. Phenotypes are used for
model fitting. True simulated genetic values are used only after prediction
for primary evaluation of genetic-value recovery.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)

options(stringsAsFactors = FALSE)
```

### 1.1 How to read this module

The module first identifies the canonical inputs and fixed fitting controls,
then defines the GBLUP-ADE model and evaluation metrics. The six scenarios are
crossed with the 25 frozen outer splits to create 150 fitting tasks. Saved
results are then summarized with true genetic value as the primary evaluation
target and phenotype as the secondary target.

The outer assessment phenotypes remain unavailable to model fitting. Genomic
relationships can include all individuals because marker information is
available at prediction time, whereas true simulated genetic values are accessed
only after held-out predictions have been generated.

## 2. Packages, inputs, and fitting parameters

### 2.1 Packages

This module uses `knitr` to format the tutorial. `sommer` is loaded only when a
new GBLUP-ADE run is explicitly requested, allowing the page to be rebuilt from
canonical saved results without refitting the 150 models.

```{r packages, warning=FALSE, message=FALSE}
RUN_GBLUP <- identical(Sys.getenv("RUN_GBLUP", "0"), "1")

library(knitr) # Format tables and R Markdown output.

if (RUN_GBLUP) {
  suppressPackageStartupMessages(library(sommer)) # Fit GBLUP-ADE and build genomic kernels.
}
```

```{r table-formatting, include=FALSE}
format_table <- function(
  x,
  caption = NULL,
  digits = getOption("digits"),
  row.names = NA,
  ...
) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    row.names = row.names,
    table.attr = paste0(
      'class="table table-striped table-hover table-bordered" ',
      'style="width:100%; margin-bottom:0;"'
    ),
    ...
  )

  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

### 2.2 Canonical input files

The GBLUP-ADE branch consumes only the audited common data object and the frozen
validation design. The audit CSV files are retained as provenance references for
the same upstream contracts.

```{r input-paths}
PROJECT_ROOT <- "."
project_path <- function(...) file.path(PROJECT_ROOT, ...)

DATA_FILE <- project_path(
  "output", "simulated", "audit", "preprocessed_simulated_common.rds"
)
CV_FILE <- project_path(
  "output", "simulated", "validation", "simulated_cv_design.rds"
)
DATA_AUDIT_FILE <- project_path(
  "output", "simulated", "audit", "simulated_audit_summary.csv"
)
CV_AUDIT_FILE <- project_path(
  "output", "simulated", "validation", "simulated_cv_design_audit.csv"
)

input_files <- data.frame(
  File = basename(c(DATA_FILE, CV_FILE, DATA_AUDIT_FILE, CV_AUDIT_FILE)),
  Role = c(
    "Audited phenotype, true genetic value, genotype, and scenario objects.",
    "Frozen outer and inner validation partitions.",
    "Data-audit provenance and compatibility record.",
    "Validation-design provenance and compatibility record."
  )
)

format_table(
  input_files,
  caption = "Canonical files available to the GBLUP-ADE branch"
)
```

### 2.3 Output files

The canonical task-level results, held-out predictions, kernel diagnostics, and
run metadata are written to the GBLUP output directory when `RUN_GBLUP=1`.
When fitting is disabled, these same saved artifacts are read for the tutorial
summaries below.

```{r output-paths}
RESULTS_RDS <- project_path(
  "output", "simulated", "gblup", "simulated_gblup_ade_results.rds"
)
PREDICTIONS_RDS <- project_path(
  "output", "simulated", "gblup", "simulated_gblup_ade_predictions.rds"
)
RESULTS_CSV <- project_path(
  "output", "simulated", "gblup", "simulated_gblup_ade_results.csv"
)
PREDICTIONS_CSV <- project_path(
  "output", "simulated", "gblup", "simulated_gblup_ade_predictions.csv"
)
KERNEL_CSV <- project_path(
  "output", "simulated", "gblup", "simulated_gblup_ade_kernel_summary.csv"
)
RUN_RDS <- project_path(
  "output", "simulated", "gblup", "simulated_gblup_ade_run_metadata.rds"
)

output_files <- data.frame(
  File = basename(c(
    RESULTS_RDS,
    PREDICTIONS_RDS,
    RESULTS_CSV,
    PREDICTIONS_CSV,
    KERNEL_CSV,
    RUN_RDS
  )),
  Role = c(
    "Task-level GBLUP-ADE metrics.",
    "Held-out individual predictions and genomic components.",
    "CSV copy of task-level metrics.",
    "CSV copy of held-out predictions.",
    "Additive, dominance, and epistatic kernel diagnostics.",
    "Execution metadata and computational timing."
  )
)

format_table(
  output_files,
  caption = "Canonical GBLUP-ADE output artifacts"
)
```

### 2.4 Execution mode and fitting control

The project-wide seed is retained for reproducibility. The number of
`sommer::mmes()` iterations is fixed at 20 for every scenario and outer split;
it is part of the benchmark specification and is not a tuned hyperparameter.

```{r gblup-control}
SEED_GLOBAL <- 20260821L
N_ITERS_SOMMER <- 20L

fitting_control <- data.frame(
  Parameter = c("Global seed", "Run new GBLUP fits", "sommer iterations per fit"),
  Value = c(SEED_GLOBAL, RUN_GBLUP, N_ITERS_SOMMER),
  Role = c(
    "Reproduce the canonical fitting workflow.",
    "Control whether the 150 GBLUP-ADE fits are executed.",
    "Use the same fixed fitting control for every task."
  )
)

format_table(
  fitting_control,
  caption = "GBLUP-ADE execution and fitting controls"
)
```

## 3. GBLUP-ADE model

### 3.1 Statistical model

The benchmark predicts each phenotype from additive, dominance, and epistatic
genomic components:

\[
\mathbf{y} = \mathbf{1}\mu + \mathbf{a} + \mathbf{d} + \mathbf{e} +
\boldsymbol{\varepsilon}.
\]

The three genetic components are modeled as random effects:

\[
\mathbf{a}\sim N(0,\mathbf{A}\sigma_a^2),\qquad
\mathbf{d}\sim N(0,\mathbf{D}\sigma_d^2),
\]

\[
\mathbf{e}\sim N(0,\mathbf{E}\sigma_e^2),\qquad
\boldsymbol{\varepsilon}\sim N(0,\mathbf{I}\sigma_\varepsilon^2).
\]

```{r model-objects}
model_objects <- data.frame(
  Object = c("y", "mu", "a", "d", "e", "epsilon"),
  Meaning = c(
    "Phenotype vector for one simulation scenario.",
    "Overall intercept.",
    "Additive genomic effect.",
    "Dominance genomic effect.",
    "Epistatic genomic effect.",
    "Residual."
  )
)

format_table(
  model_objects,
  caption = "Objects in the GBLUP-ADE model"
)
```

### 3.2 Additive relationship matrix

The additive component uses the genomic relationship matrix \(\mathbf{A}\),
constructed from the marker matrix with `sommer::A.mat()`. It represents
additive genomic covariance among the 1,000 individuals.

### 3.3 Dominance relationship matrix

The dominance component uses \(\mathbf{D}\), constructed with
`sommer::D.mat()`. It represents covariance associated with dominance genomic
effects.

### 3.4 Epistatic relationship matrix

The epistatic component uses \(\mathbf{E}\), constructed with
`sommer::E.mat()`. It represents the epistatic genomic covariance structure used
by this prespecified benchmark.

```{r relationship-matrix-roles}
relationship_matrices <- data.frame(
  Matrix = c("A", "D", "E"),
  Role = c(
    "Additive genomic relationships.",
    "Dominance genomic relationships.",
    "Epistatic genomic relationships."
  ),
  Constructor = c("sommer::A.mat()", "sommer::D.mat()", "sommer::E.mat()")
)

format_table(
  relationship_matrices,
  caption = "Genomic relationship matrices used by GBLUP-ADE"
)
```

These matrices describe covariance among individuals; they are not additional
response variables.

### 3.5 Predicted genetic value

In `sommer::mmes()`, `ism()` identifies the individual levels associated with
each random effect and `vsm()` connects that effect to the corresponding genomic
relationship matrix through `Gu`.

The predicted genetic value is

\[
\widehat{G}=\widehat{a}+\widehat{d}+\widehat{e},
\]

and the final phenotype-scale prediction adds the fitted intercept to this
genetic component.

## 4. Evaluation metrics

The held-out predictions are summarized with correlation, RMSE, MAE, bias,
calibration slope, and \(R^2\). The functions below assume the audited canonical
assessment vectors are finite.

```{r gblup-metrics}
safe_cor <- function(pred, obs) {
  cor(pred, obs)
}

safe_rmse <- function(pred, obs) {
  sqrt(
    mean(
      (pred - obs)^2
    )
  )
}

safe_mae <- function(pred, obs) {
  mean(
    abs(
      pred - obs
    )
  )
}

safe_bias <- function(pred, obs) {
  mean(
    pred - obs
  )
}

safe_slope <- function(pred, obs) {
  as.numeric(
    coef(
      lm(
        obs ~ pred
      )
    )[2]
  )
}

safe_r2 <- function(pred, obs) {
  total_variation <- sum(
    (
      obs -
      mean(obs)
    )^2
  )

  1 -
    sum(
      (
        obs -
        pred
      )^2
    ) /
    total_variation
}

metrics <- function(
  pred,
  obs,
  prefix
) {
  values <- c(
    safe_cor(pred, obs),
    safe_rmse(pred, obs),
    safe_mae(pred, obs),
    safe_bias(pred, obs),
    safe_slope(pred, obs),
    safe_r2(pred, obs)
  )

  names(values) <- paste0(
    prefix,
    c(
      "_Correlation",
      "_RMSE",
      "_MAE",
      "_Bias",
      "_Slope",
      "_R2"
    )
  )

  values
}
```

For bias, positive values indicate overprediction because the definition is
`mean(predicted - observed)`. The calibration slope is obtained from
`observed ~ predicted`, and \(R^2\) is computed as \(1-SSE/SST\).

## 5. Canonical input objects

The GBLUP-ADE branch now reads the audited simulated-data object and frozen
validation design. The same 1,000 individuals occur in the phenotype,
true-genetic-value, and genotype objects, and the same 25 outer splits are used
throughout the benchmark.

```{r inputs}
common <- readRDS(DATA_FILE)
cv <- readRDS(CV_FILE)

phenotype <- common$phenotype
true_g <- common$true_genetic_value
geno <- common$genotype_m101
control <- common$scenario_control
ids <- seq_len(nrow(geno))
id_labels <- sprintf("ID%04d", ids)
rownames(geno) <- id_labels
scenario_names <- colnames(phenotype)

input_summary <- data.frame(
  Object = c(
    "Phenotype",
    "True genetic value",
    "Genotype",
    "Outer validation splits"
  ),
  Structure = c(
    paste(dim(phenotype), collapse = " x "),
    paste(dim(true_g), collapse = " x "),
    paste(dim(geno), collapse = " x "),
    length(cv$outer_splits)
  ),
  Role = c(
    "Response used for model fitting.",
    "Primary evaluation target only.",
    "Marker information used to build A, D, and E.",
    "Frozen held-out evaluation design."
  ),
  stringsAsFactors = FALSE
)

phenotype_sample <- data.frame(
  ID = ids[1:6],
  phenotype[1:6, , drop = FALSE],
  check.names = FALSE
)
genotype_sample <- data.frame(
  ID = ids[1:6],
  geno[1:6, 1:10, drop = FALSE],
  check.names = FALSE
)

format_table(
  input_summary,
  caption = "Canonical inputs used by GBLUP-ADE"
)
format_table(
  phenotype_sample,
  digits = 4,
  caption = "First six phenotype rows"
)
format_table(
  genotype_sample,
  caption = "First six individuals and first ten markers"
)
```

The phenotype and true-genetic-value matrices contain the same 1,000
individuals and six scenarios. The genotype matrix contains the 4,010 markers
used to construct the genomic relationship matrices, while the 25 frozen outer
splits define which 200 individuals are evaluated in each fit.

## 6. Frozen analysis tasks

The six scenarios are crossed with the 25 frozen outer splits to define the
complete GBLUP-ADE workload. Each row identifies one independent
scenario-by-split prediction task; the full design contains 150 tasks.

```{r task-table}
outer_meta <- unique(
  cv$outer_assignments[, c("Repetition", "OuterFold", "SplitID")]
)

outer_meta <- outer_meta[
  order(outer_meta$Repetition, outer_meta$OuterFold),
]

task_table <- merge(
  data.frame(Scenario = scenario_names, stringsAsFactors = FALSE),
  outer_meta,
  by = NULL
)

task_table <- task_table[
  order(
    match(task_table$Scenario, scenario_names),
    task_table$Repetition,
    task_table$OuterFold
  ),
]

task_table$TaskID <- paste(
  task_table$Scenario,
  task_table$SplitID,
  sep = "_"
)

rownames(task_table) <- NULL

format_table(
  head(task_table, 15),
  caption = "First 15 GBLUP-ADE scenario-by-split tasks"
)
```

Each row represents one `Scenario x outer split` combination. Because there are
6 scenarios and 25 frozen outer splits, the complete benchmark contains
**6 x 25 = 150 GBLUP-ADE fits**. `TaskID` is the stable identifier used to link
each fit to its held-out predictions and later to the ANN comparison.

## 7. Fitting procedure

The A, D, and E relationship matrices are built once from the complete genotype
matrix and reused across tasks. All genotype profiles may contribute to these
relationship matrices because genomic information is available at prediction
time.

For each outer split, the phenotypes of the 200 assessment individuals are
replaced by `NA` before fitting. Consequently, those individuals contribute
genomic relationships but not response information. Their true genetic values
are accessed only after prediction for performance evaluation.

`sommer::mmes()` stores the fitted additive, dominance, and epistatic random
effects in `fit$uList`. The helper below maps each fitted random-effect vector
back to the 1,000 individuals through the corresponding model matrix.

```{r extract-genomic-component}
extract_u <- function(
  fit,
  dat,
  factor_name,
  effect_name
) {
  random_effect <- as.matrix(
    fit$uList[[effect_name]]
  )

  incidence_matrix <- model.matrix(
    as.formula(
      paste0(
        "~",
        factor_name,
        " - 1"
      )
    ),
    data = dat
  )

  as.numeric(
    incidence_matrix %*%
      random_effect[
        ,
        1L,
        drop = FALSE
      ]
  )
}
```

The three calls in the fitting loop differ only in the factor and
relationship-matrix names (`id/A`, `idd/D`, and `ide/E`).

### 7.1 One representative fit

The essential model call for one scenario and one outer split is:

```{r gblup-fit-example, eval=FALSE}
fit <- sommer::mmes(
  y ~ 1,
  random = ~
    vsm(ism(id),  Gu = A_matrix) +
    vsm(ism(idd), Gu = D_matrix) +
    vsm(ism(ide), Gu = E_matrix),
  rcov = ~ units,
  nIters = 20,
  data = dat,
  verbose = FALSE
)
```

Here `dat$y` contains phenotype values for the analysis individuals and `NA` for
the 200 outer-assessment individuals. The matrices `A_matrix`, `D_matrix`, and
`E_matrix` are shared across tasks; only the scenario and frozen assessment set
change.

### 7.2 Complete 150-task workflow

The code below repeats the canonical fitting logic over all 150 frozen tasks:
identify the split, mask the assessment phenotypes, fit GBLUP-ADE, recover the
three genomic components, predict the held-out individuals, and store the
metrics.

```{r fit-gblup-ade, eval=RUN_GBLUP}
set.seed(SEED_GLOBAL)

run_start <- proc.time()[["elapsed"]]

# Build the three genomic relationship matrices once for all tasks.
kernel_start <- proc.time()[["elapsed"]]

A_matrix <- sommer::A.mat(geno)
D_matrix <- sommer::D.mat(geno)
E_matrix <- sommer::E.mat(geno)

kernel_seconds <- proc.time()[["elapsed"]] - kernel_start

rownames(A_matrix) <- colnames(A_matrix) <- id_labels
rownames(D_matrix) <- colnames(D_matrix) <- id_labels
rownames(E_matrix) <- colnames(E_matrix) <- id_labels

kernel_summary <- data.frame(
  Kernel = c("Additive", "Dominance", "Epistatic"),
  Rows = c(nrow(A_matrix), nrow(D_matrix), nrow(E_matrix)),
  Columns = c(ncol(A_matrix), ncol(D_matrix), ncol(E_matrix)),
  Symmetric = c(
    isTRUE(all.equal(A_matrix, t(A_matrix), tolerance = 1e-8)),
    isTRUE(all.equal(D_matrix, t(D_matrix), tolerance = 1e-8)),
    isTRUE(all.equal(E_matrix, t(E_matrix), tolerance = 1e-8))
  ),
  Finite = c(
    all(is.finite(A_matrix)),
    all(is.finite(D_matrix)),
    all(is.finite(E_matrix))
  ),
  Mean_diagonal = c(
    mean(diag(A_matrix)),
    mean(diag(D_matrix)),
    mean(diag(E_matrix))
  ),
  Kernel_build_seconds = rep(kernel_seconds, 3L),
  stringsAsFactors = FALSE
)

write.csv(
  kernel_summary,
  KERNEL_CSV,
  row.names = FALSE
)

result_list <- vector("list", nrow(task_table))
prediction_list <- vector("list", nrow(task_table))

for (i in seq_len(nrow(task_table))) {
  task <- task_table[i, ]
  scenario <- task$Scenario
  scenario_index <- match(
    scenario,
    scenario_names
  )

  # Identify the frozen analysis and assessment individuals for this task.
  split <- cv$outer_splits[[task$SplitID]]

  analysis_index <- match(
    split$analysis_ids,
    ids
  )
  assessment_index <- match(
    split$assessment_ids,
    ids
  )

  phenotype_vector <- phenotype[
    ,
    scenario_index
  ]
  true_genetic_vector <- true_g[
    ,
    scenario_index
  ]

  id_factor <- factor(
    id_labels,
    levels = id_labels
  )

  dat <- data.frame(
    y = phenotype_vector,
    id = id_factor,
    idd = id_factor,
    ide = id_factor
  )

  # Withhold outer-assessment phenotypes before fitting the model.
  dat$y[assessment_index] <- NA_real_

  fit_start <- proc.time()[["elapsed"]]

  # Fit the prespecified additive + dominance + epistatic mixed model.
  fit <- sommer::mmes(
    y ~ 1,
    random = ~
      vsm(ism(id), Gu = A_matrix) +
      vsm(ism(idd), Gu = D_matrix) +
      vsm(ism(ide), Gu = E_matrix),
    rcov = ~ units,
    nIters = N_ITERS_SOMMER,
    data = dat,
    verbose = FALSE,
    dateWarning = FALSE
  )

  fit_seconds <- proc.time()[["elapsed"]] - fit_start

  prediction_start <- proc.time()[["elapsed"]]

  # Recover the three fitted genomic components for all individuals.
  additive <- extract_u(
    fit,
    dat,
    "id",
    "vsm(ism(id), Gu = A_matrix)"
  )
  dominance <- extract_u(
    fit,
    dat,
    "idd",
    "vsm(ism(idd), Gu = D_matrix)"
  )
  epistatic <- extract_u(
    fit,
    dat,
    "ide",
    "vsm(ism(ide), Gu = E_matrix)"
  )

  gebv <- additive + dominance + epistatic
  intercept <- as.numeric(fit$b)[1L]
  predicted_value <- intercept + gebv

  predict_seconds <- proc.time()[["elapsed"]] - prediction_start

  # Evaluate predictions only on the untouched outer-assessment individuals.
  observed_test <- phenotype_vector[
    assessment_index
  ]
  true_g_test <- true_genetic_vector[
    assessment_index
  ]
  predicted_test <- predicted_value[
    assessment_index
  ]
  gebv_test <- gebv[
    assessment_index
  ]

  genetic_metrics <- metrics(
    predicted_test,
    true_g_test,
    "Genetic"
  )
  phenotype_metrics <- metrics(
    predicted_test,
    observed_test,
    "Phenotype"
  )

  meta <- control[
    control$Scenario == scenario,
    ,
    drop = FALSE
  ]

  result_list[[i]] <- data.frame(
    Dataset = "simulated",
    Scenario = scenario,
    QTL_N = meta$QTL_N,
    H2_nominal = meta$H2_nominal,
    Repetition = task$Repetition,
    Fold = task$OuterFold,
    SplitID = task$SplitID,
    TaskID = task$TaskID,
    Model = "GBLUP_ADE",
    N_train = length(analysis_index),
    N_test = length(assessment_index),
    N_pred_finite = sum(is.finite(predicted_test)),
    Genetic_Correlation =
      genetic_metrics["Genetic_Correlation"],
    Genetic_RMSE =
      genetic_metrics["Genetic_RMSE"],
    Genetic_MAE =
      genetic_metrics["Genetic_MAE"],
    Genetic_Bias =
      genetic_metrics["Genetic_Bias"],
    Genetic_Slope =
      genetic_metrics["Genetic_Slope"],
    Genetic_R2 =
      genetic_metrics["Genetic_R2"],
    GEBV_TrueG_Correlation =
      safe_cor(gebv_test, true_g_test),
    Phenotype_Correlation =
      phenotype_metrics["Phenotype_Correlation"],
    Phenotype_RMSE =
      phenotype_metrics["Phenotype_RMSE"],
    Phenotype_MAE =
      phenotype_metrics["Phenotype_MAE"],
    Phenotype_Bias =
      phenotype_metrics["Phenotype_Bias"],
    Phenotype_Slope =
      phenotype_metrics["Phenotype_Slope"],
    Phenotype_R2 =
      phenotype_metrics["Phenotype_R2"],
    Kernel_build_seconds = kernel_seconds,
    Fit_seconds = fit_seconds,
    Predict_seconds = predict_seconds,
    Status = "ok",
    Error = NA_character_,
    stringsAsFactors = FALSE
  )

  prediction_list[[i]] <- data.frame(
    Dataset = "simulated",
    Scenario = scenario,
    QTL_N = meta$QTL_N,
    H2_nominal = meta$H2_nominal,
    Repetition = task$Repetition,
    Fold = task$OuterFold,
    SplitID = task$SplitID,
    TaskID = task$TaskID,
    Model = "GBLUP_ADE",
    ID = ids[assessment_index],
    Observed_Phenotype = observed_test,
    True_Genetic_Value = true_g_test,
    Predicted_Value = predicted_test,
    GEBV = gebv_test,
    Additive_Prediction = additive[assessment_index],
    Dominance_Prediction = dominance[assessment_index],
    Epistatic_Prediction = epistatic[assessment_index],
    Status = "ok",
    stringsAsFactors = FALSE
  )

  cat(
    sprintf(
      "[%03d/%03d] %s | rG=%.4f | RMSE_G=%.4f | fit=%.2fs\n",
      i,
      nrow(task_table),
      task$TaskID,
      result_list[[i]]$Genetic_Correlation,
      result_list[[i]]$Genetic_RMSE,
      fit_seconds
    )
  )
}

results <- do.call(rbind, result_list)
predictions <- do.call(rbind, prediction_list)

rownames(results) <- NULL
rownames(predictions) <- NULL

run_seconds <- proc.time()[["elapsed"]] - run_start

run_metadata <- list(
  dataset = "simulated",
  model = "GBLUP_ADE",
  seed_global = SEED_GLOBAL,
  sommer_version =
    as.character(packageVersion("sommer")),
  nIters_sommer = N_ITERS_SOMMER,
  tasks = nrow(results),
  prediction_rows = nrow(predictions),
  kernel_build_seconds = kernel_seconds,
  total_run_seconds = run_seconds,
  timestamp = as.character(Sys.time())
)

saveRDS(results, RESULTS_RDS, compress = "gzip")
saveRDS(predictions, PREDICTIONS_RDS, compress = "gzip")
write.csv(results, RESULTS_CSV, row.names = FALSE)
write.csv(predictions, PREDICTIONS_CSV, row.names = FALSE)
saveRDS(run_metadata, RUN_RDS, compress = "gzip")
```

## 8. Genomic-kernel diagnostics

Numerical integrity of the additive, dominance, and epistatic relationship
matrices is checked before saved predictions are interpreted. The saved kernel
diagnostics report their dimensions, symmetry, finiteness, mean diagonal, and
one-time construction cost.

```{r kernel-diagnostics}
kernel_summary <- read.csv(
  KERNEL_CSV,
  stringsAsFactors = FALSE
)

kernel_display <- kernel_summary[
  , c("Kernel", "Rows", "Columns", "Symmetric", "Finite", "Mean_diagonal")
]

format_table(
  kernel_display,
  digits = 4,
  caption = "Numerical structure of the GBLUP-ADE genomic kernels"
)
```

The saved diagnostics contain three **1,000 x 1,000** genomic relationship
matrices. All three are finite and symmetric, with mean diagonal values close
to one. This confirms that the additive, dominance, and epistatic covariance
structures are numerically suitable for interpreting the fitted predictions.

## 9. Canonical saved results

The performance summaries below are generated from the canonical saved
GBLUP-ADE results. Their row counts and task structure are displayed directly
before the performance tables.

```{r result-audit}
results <- readRDS(RESULTS_RDS)
predictions <- readRDS(PREDICTIONS_RDS)
run_metadata <- readRDS(RUN_RDS)

result_structure <- data.frame(
  Object = c("Task-level results", "Held-out predictions"),
  Rows = c(nrow(results), nrow(predictions)),
  Unique_tasks = c(
    length(unique(results$TaskID)),
    length(unique(predictions$TaskID))
  ),
  Rows_per_task = c(
    1,
    paste(range(as.integer(table(predictions$TaskID))), collapse = " to ")
  )
)

format_table(
  result_structure,
  caption = "Observed structure of the canonical GBLUP-ADE results"
)
```

The displayed result structure shows the number of task-level fits and held-out
predictions used by the performance summaries.

## 10. Predictive performance

The primary metrics compare the held-out final prediction (intercept + additive
+ dominance + epistatic components) with the known true genetic value.
Phenotype metrics are retained as the secondary evaluation target.

```{r performance-summary}
ok_results <- results[
  results$Status == "ok",
  ,
  drop = FALSE
]

scenario_performance <- aggregate(
  cbind(
    Genetic_Correlation,
    Genetic_RMSE,
    Genetic_MAE,
    Genetic_Bias,
    Genetic_Slope,
    Genetic_R2,
    Phenotype_Correlation,
    Phenotype_RMSE,
    Phenotype_MAE,
    Phenotype_Bias,
    Phenotype_Slope,
    Phenotype_R2
  ) ~ Scenario + QTL_N + H2_nominal,
  data = ok_results,
  FUN = mean,
  na.rm = TRUE
)

scenario_performance <- scenario_performance[
  order(match(scenario_performance$Scenario, scenario_names)),
]

genetic_performance_display <- scenario_performance[
  ,
  c(
    "Scenario",
    "QTL_N",
    "H2_nominal",
    "Genetic_Correlation",
    "Genetic_RMSE",
    "Genetic_MAE",
    "Genetic_Bias",
    "Genetic_Slope",
    "Genetic_R2"
  )
]

phenotype_performance_display <- scenario_performance[
  ,
  c(
    "Scenario",
    "QTL_N",
    "H2_nominal",
    "Phenotype_Correlation",
    "Phenotype_RMSE",
    "Phenotype_MAE",
    "Phenotype_Bias",
    "Phenotype_Slope",
    "Phenotype_R2"
  )
]
```

### 10.1 True genetic value

```{r genetic-performance}
format_table(
  genetic_performance_display,
  digits = 4,
  caption = "Primary evaluation: true genetic-value prediction across the 25 outer splits"
)
```

Across the six scenarios, mean held-out genetic correlation ranges from
`r sprintf("%.4f", min(scenario_performance$Genetic_Correlation))` to
`r sprintf("%.4f", max(scenario_performance$Genetic_Correlation))`, while mean
genetic RMSE ranges from
`r sprintf("%.4f", min(scenario_performance$Genetic_RMSE))` to
`r sprintf("%.4f", max(scenario_performance$Genetic_RMSE))`. The C1-C6 pattern
is interpreted as performance across the prespecified genetic-complexity
gradient; because QTL number and nominal heritability vary together across the
complete sequence, it is not an isolated causal effect of either factor.

### 10.2 Phenotype

```{r phenotype-performance}
format_table(
  phenotype_performance_display,
  digits = 4,
  caption = "Secondary evaluation: phenotype prediction across the 25 outer splits"
)
```

Phenotype prediction is retained as the secondary evaluation because phenotype
contains both genetic and residual variation. The true simulated genetic value
remains the primary target for comparing recovery of the simulated genetic
signal.

## 11. Split-level variability

Scenario means can conceal sensitivity to which individuals were held out. The
next summary describes the distribution of primary genetic-value performance
across the 25 frozen outer splits within each scenario.

```{r split-summary}
split_summary <- do.call(
  rbind,
  lapply(
    split(ok_results, ok_results$Scenario),
    function(x) {
      data.frame(
        Scenario = unique(x$Scenario),
        N = nrow(x),
        Cor_mean = mean(x$Genetic_Correlation, na.rm = TRUE),
        Cor_sd = sd(x$Genetic_Correlation, na.rm = TRUE),
        Cor_min = min(x$Genetic_Correlation, na.rm = TRUE),
        Cor_max = max(x$Genetic_Correlation, na.rm = TRUE),
        RMSE_mean = mean(x$Genetic_RMSE, na.rm = TRUE),
        RMSE_sd = sd(x$Genetic_RMSE, na.rm = TRUE),
        MAE_mean = mean(x$Genetic_MAE, na.rm = TRUE),
        R2_mean = mean(x$Genetic_R2, na.rm = TRUE)
      )
    }
  )
)

rownames(split_summary) <- NULL
split_summary <- split_summary[
  order(match(split_summary$Scenario, scenario_names)),
]

format_table(
  split_summary,
  digits = 4,
  caption = "Genetic-value prediction variability across outer splits"
)
```

Variation across the 25 outer splits quantifies the sensitivity of predictive
performance to which individuals are held out. Reporting the mean, standard
deviation, minimum, and maximum genetic correlation therefore complements the
scenario averages and prevents a single partition from determining the
conclusion.

## 12. Computational time

The timing summary separates the one-time construction of the genomic
relationship matrices from repeated model fitting and prediction across the 150
tasks. This distinction prevents kernel construction from being interpreted as
a cost paid independently for every split.

```{r time-summary}
time_summary <- aggregate(
  cbind(Fit_seconds, Predict_seconds) ~ Scenario,
  data = results,
  FUN = mean,
  na.rm = TRUE
)

time_summary$Kernel_build_seconds <- run_metadata$kernel_build_seconds

format_table(
  time_summary,
  digits = 3,
  caption = "GBLUP-ADE computational time"
)

cat(
  "Total run time:",
  round(run_metadata$total_run_seconds, 2),
  "seconds.\n"
)
```

Because the A, D, and E matrices are built once and reused across all 150
tasks, their construction time is reported separately from the mean fitting and
prediction times.

## 13. Navigation

[Previous: Validation design](simulated_data_validation.html) | [Parallel branch: Regularized ANN](simulated_data_ann.html) | [Next: Model comparison](simulated_data_comparison.html)
