---
title: "Regularized artificial neural networks for simulated genomic data"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

# Purpose

This module is the **canonical ANN workflow** for the simulated genomic-data
branch. ANN hyperparameter tuning and final outer-CV fitting are internal
stages of one modeling method, parallel to GBLUP in the public pipeline.

The four variants are `ANN_0`, `ANN_L1`, `ANN_L2`, and `ANN_L1_L2`
(displayed as **ANN-L1+L2**). They use the same multilayer perceptron and differ
only in L1/L2 weight penalties.

The page has two independent execution modes:

- `RUN_ANN_TUNING=1`: run or refresh nested inner-CV tuning only;
- `RUN_ANN=1`: consume the saved canonical tuning selections, select
  `Best_epoch` using only the 800 outer-analysis individuals, refit from
  scratch on all 800 for that epoch count, and only then predict the untouched
  200 outer-assessment individuals.

Using both flags intentionally runs the full ANN workflow. Ordinary builds run
neither expensive stage and only render saved outputs.

# Tuning contract

Only four quantities are treated as hyperparameters:

1. hidden-layer architecture;
2. Adam learning rate;
3. L1 penalty coefficient;
4. L2 penalty coefficient.

The following are **fixed optimization controls**, not tuned quantities:

- activation: ReLU;
- optimizer: Adam;
- dropout: 0;
- maximum epochs available to an early-stopped fit: 2000;
- patience: 25;
- minimum validation improvement: \(10^{-6}\);
- early-validation fraction: 0.15.

The 2000-epoch value is a **convergence ceiling**, not a fixed final training
length. The saved `ANN_TUNING_CANONICAL_V1` metadata still contains the
historical field `final_train_epochs = 2000` because that value belongs to the
completed tuning run's immutable provenance/configuration identifier; the
canonical final-fitting stage does **not** use that field as a refit rule. Epoch diagnostics are excluded from hyperparameter ranking. After
hyperparameters are selected, a new early-stopping fit uses only the
outer-analysis sample to select `Best_epoch`; the final network is then refit
from scratch on all 800 outer-analysis individuals for exactly `Best_epoch`.

# Architecture range

The canonical architecture catalog deliberately defines a **bounded
small-to-moderate neural-network search space**, ranging from two hidden neurons
to `64-32`.

Earlier development experiments explored larger networks, including
architectures up to `256-128`, and showed that increased capacity could improve
inner-validation performance in some tasks. Those larger architectures were
not carried forward into the canonical grid. Instead, `64-32` was retained as
a prespecified computational and methodological upper bound for the final
comparison.

This restriction is therefore part of the estimand of the canonical analysis:
the reported comparison evaluates GBLUP-ADE against regularized and
unregularized ANN models **within this bounded architecture class**. It should
not be interpreted as demonstrating that GBLUP-ADE outperforms every possible
higher-capacity neural-network architecture.

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
options(stringsAsFactors = FALSE)

find_project_root <- function(start = getwd()) {
  path <- normalizePath(start, winslash = "/", mustWork = TRUE)

  repeat {
    if (file.exists(file.path(path, "_workflowr.yml"))) {
      return(path)
    }

    parent <- dirname(path)

    if (identical(parent, path)) {
      stop("Could not locate workflowr project root.")
    }

    path <- parent
  }
}

PROJECT_ROOT <- find_project_root()

project_path <- function(...) {
  file.path(PROJECT_ROOT, ...)
}


SEED_GLOBAL <- 20260821L

RUN_ANN_TUNING <- identical(
  Sys.getenv("RUN_ANN_TUNING", "0"),
  "1"
)
ANN_TUNING_FORCE <- identical(
  Sys.getenv("ANN_TUNING_FORCE", "0"),
  "1"
)
ANN_TUNING_PILOT <- identical(
  Sys.getenv("ANN_TUNING_PILOT", "0"),
  "1"
)
ANN_TUNING_PILOT_TASK <- Sys.getenv(
  "ANN_TUNING_PILOT_TASK",
  "C1_R01_F01"
)

MAX_EPOCHS <- 2000L
# Compatibility token only: preserves the exact config_id of the completed
# ANN_TUNING_CANONICAL_V1 run. It does not control final refit length.
TUNING_V1_COMPATIBILITY_TOKEN <- 2000L
PATIENCE <- 25L
MIN_DELTA <- 1e-6
EARLY_VALIDATION_FRACTION <- 0.15
DROPOUT <- 0

ARCHITECTURES <- list(
  `2` = c(2L),
  `4` = c(4L),
  `8` = c(8L),
  `16` = c(16L),
  `32` = c(32L),
  `8-4` = c(8L, 4L),
  `16-8` = c(16L, 8L),
  `32-16` = c(32L, 16L),
  `64-32` = c(64L, 32L)
)

LEARNING_RATES <- c(
  1.25e-4,
  2.5e-4,
  5e-4,
  1e-3
)

L1_VALUES <- c(
  1e-5,
  1e-4,
  1e-3,
  1e-2,
  3e-2,
  1e-1
)

L2_VALUES <- c(
  1e-5,
  1e-4,
  1e-3,
  1e-2
)

# Full Cartesian grid for the simultaneous L1+L2 model.
# Both coefficients are strictly positive in ANN_L1_L2.
L1_L2_GRID <- expand.grid(
  L1 = L1_VALUES,
  L2 = L2_VALUES,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

MODEL_LEVELS <- c(
  "ANN_0",
  "ANN_L1",
  "ANN_L2",
  "ANN_L1_L2"
)

DATA_FILE <- project_path("output", "simulated", "audit", "preprocessed_simulated_common.rds")
CV_FILE <- project_path("output", "simulated", "validation", "simulated_cv_design.rds")
DATA_AUDIT_FILE <- project_path("output", "simulated", "audit", "simulated_audit_summary.csv")
CV_AUDIT_FILE <- project_path("output", "simulated", "validation", "simulated_cv_design_audit.csv")

CHECKPOINT_DIR <- project_path(".local", "checkpoints", "canonical", "ann_tuning")

OUTPUT_STEM <- if (ANN_TUNING_PILOT) {
  project_path(".local", "pilot", "ann", "tuning", "simulated_ann_tuning_pilot")
} else {
  project_path("output", "simulated", "ann", "tuning", "simulated_ann_tuning")
}

CANDIDATE_RDS <- paste0(OUTPUT_STEM, "_candidate_summary.rds")
FOLD_RDS <- paste0(OUTPUT_STEM, "_fold_results.rds")
SELECTED_RDS <- paste0(OUTPUT_STEM, "_selected_configurations.rds")
RUN_RDS <- paste0(OUTPUT_STEM, "_run_metadata.rds")
MANIFEST_RDS <- paste0(OUTPUT_STEM, "_manifest.rds")

CANDIDATE_CSV <- paste0(OUTPUT_STEM, "_candidate_summary.csv")
FOLD_CSV <- paste0(OUTPUT_STEM, "_fold_results.csv")
SELECTED_CSV <- paste0(OUTPUT_STEM, "_selected_configurations.csv")
MANIFEST_CSV <- paste0(OUTPUT_STEM, "_manifest.csv")

ANN_TUNING_CONFIG_ID <- paste(
  "ANN_TUNING_CANONICAL_V1",
  paste(names(ARCHITECTURES), collapse = "-"),
  paste(LEARNING_RATES, collapse = "-"),
  paste(L1_VALUES, collapse = "-"),
  paste(L2_VALUES, collapse = "-"),
  MAX_EPOCHS,
  PATIENCE,
  MIN_DELTA,
  EARLY_VALIDATION_FRACTION,
  TUNING_V1_COMPATIBILITY_TOKEN,
  sep = "__"
)

if (RUN_ANN_TUNING) {
  dir.create(
    CHECKPOINT_DIR,
    recursive = TRUE,
    showWarnings = FALSE
  )
  dir.create(
    dirname(CANDIDATE_RDS),
    recursive = TRUE,
    showWarnings = FALSE
  )
}
```

# Packages and computational backend

```{r packages}
if (!requireNamespace("torch", quietly = TRUE)) {
  stop("Package 'torch' is required.")
}

suppressPackageStartupMessages(
  library(torch)
)

backend_ok <- tryCatch({
  z <- torch::torch_tensor(
    c(1, 2),
    dtype = torch::torch_float()
  )
  all(as.numeric(z) == c(1, 2))
}, error = function(e) FALSE)

if (!backend_ok) {
  stop(
    "torch is installed but the libtorch backend is unavailable."
  )
}

cuda_available <- isTRUE(
  tryCatch(
    torch::cuda_is_available(),
    error = function(e) FALSE
  )
)

requested_device <- tolower(
  Sys.getenv("ANN_DEVICE", "auto")
)

if (!requested_device %in% c("auto", "cpu", "cuda")) {
  stop("ANN_DEVICE must be auto, cpu, or cuda.")
}

if (requested_device == "cuda" && !cuda_available) {
  stop(
    "ANN_DEVICE=cuda requested but CUDA is unavailable."
  )
}

ANN_DEVICE_LABEL <- if (
  requested_device == "cuda" ||
  (requested_device == "auto" && cuda_available)
) {
  "cuda"
} else {
  "cpu"
}

ANN_DEVICE <- torch::torch_device(
  ANN_DEVICE_LABEL
)

sync_device <- function() {
  if (ANN_DEVICE_LABEL == "cuda") {
    try(
      torch::cuda_synchronize(),
      silent = TRUE
    )
  }
  invisible(NULL)
}

set_ann_seed <- function(seed) {
  set.seed(as.integer(seed))
  torch::torch_manual_seed(
    as.integer(seed)
  )
  invisible(NULL)
}

environment_table <- data.frame(
  Item = c(
    "torch version",
    "backend available",
    "CUDA available",
    "device",
    "MAX_EPOCHS",
    "TUNING_V1_COMPATIBILITY_TOKEN",
    "patience",
    "minimum improvement",
    "early-validation fraction",
    "dropout"
  ),
  Value = c(
    as.character(packageVersion("torch")),
    backend_ok,
    cuda_available,
    ANN_DEVICE_LABEL,
    MAX_EPOCHS,
    TUNING_V1_COMPATIBILITY_TOKEN,
    PATIENCE,
    MIN_DELTA,
    EARLY_VALIDATION_FRACTION,
    DROPOUT
  )
)

knitr::kable(
  environment_table,
  caption = "Fixed ANN computational protocol"
)
```

# Canonical inputs

This module depends only on the frozen common preprocessing object and frozen
nested validation design. It does not depend on V2, V3, V4, V4C, or their
outer-assessment results.

```{r inputs}
required_files <- c(
  DATA_FILE,
  CV_FILE,
  DATA_AUDIT_FILE,
  CV_AUDIT_FILE
)

if (!all(file.exists(required_files))) {
  stop(
    "Missing required file(s): ",
    paste(
      required_files[!file.exists(required_files)],
      collapse = ", "
    )
  )
}

data_audit <- read.csv(
  DATA_AUDIT_FILE,
  stringsAsFactors = FALSE
)
cv_audit <- read.csv(
  CV_AUDIT_FILE,
  stringsAsFactors = FALSE
)

if (any(data_audit$Status == "BLOCKED")) {
  stop(
    "Simulated-data audit contains BLOCKED checks."
  )
}
if (any(cv_audit$Status == "BLOCKED")) {
  stop(
    "Validation-design audit contains BLOCKED checks."
  )
}

common <- readRDS(DATA_FILE)
cv <- readRDS(CV_FILE)

phenotype <- common$phenotype
geno <- common$genotype_m101

ids <- as.integer(
  common$metadata$row_identifier
)
if (is.null(ids) || !length(ids)) {
  ids <- seq_len(nrow(geno))
}

scenario_names <- colnames(phenotype)
if (is.null(scenario_names)) {
  scenario_names <- paste0("C", 1:6)
}

input_checks <- data.frame(
  Check = c(
    "Phenotype is 1000 x 6",
    "Genotype is 1000 x 4010",
    "Genotype coding is -1/0/1",
    "25 frozen outer splits",
    "100 frozen inner splits",
    "Scenario names are C1-C6"
  ),
  PASS = c(
    identical(
      dim(phenotype),
      c(1000L, 6L)
    ),
    identical(
      dim(geno),
      c(1000L, 4010L)
    ),
    identical(
      sort(unique(as.vector(geno))),
      -1:1
    ),
    length(cv$outer_splits) == 25L,
    sum(vapply(
      cv$inner_splits,
      length,
      integer(1)
    )) == 100L,
    identical(
      scenario_names,
      paste0("C", 1:6)
    )
  )
)

input_checks$Status <- ifelse(
  input_checks$PASS,
  "PASS",
  "BLOCKED"
)

knitr::kable(
  input_checks[, c("Check", "Status")],
  caption = "Canonical ANN tuning pre-fit audit"
)

if (any(!input_checks$PASS)) {
  stop(
    "Blocking ANN tuning pre-fit check detected."
  )
}
```

# Candidate grids

Architecture and learning rate are selected **once per outer task using
`ANN_0`**. The selected structure is then shared by all four ANN variants. This
keeps the comparison between regularization variants focused on L1/L2
penalties rather than allowing each model family to change network capacity.

After structure selection:

- `ANN_L1` tunes only L1;
- `ANN_L2` tunes only L2;
- `ANN_L1_L2` tunes the full positive L1 × L2 Cartesian grid.

```{r grids}
structure_grid <- expand.grid(
  Architecture = names(ARCHITECTURES),
  Learning_rate = LEARNING_RATES,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)
structure_grid$Model <- "ANN_0"
structure_grid$L1 <- 0
structure_grid$L2 <- 0

l1_grid <- data.frame(
  Model = "ANN_L1",
  L1 = L1_VALUES,
  L2 = 0,
  stringsAsFactors = FALSE
)

l2_grid <- data.frame(
  Model = "ANN_L2",
  L1 = 0,
  L2 = L2_VALUES,
  stringsAsFactors = FALSE
)

l1l2_grid <- data.frame(
  Model = "ANN_L1_L2",
  L1 = L1_L2_GRID$L1,
  L2 = L1_L2_GRID$L2,
  stringsAsFactors = FALSE
)

grid_summary <- data.frame(
  Stage = c(
    "Shared architecture + learning rate",
    "ANN_L1 penalty",
    "ANN_L2 penalty",
    "ANN_L1_L2 penalties"
  ),
  Candidates = c(
    nrow(structure_grid),
    nrow(l1_grid),
    nrow(l2_grid),
    nrow(l1l2_grid)
  )
)

grid_summary$Inner_fits_per_outer_task <-
  4L * grid_summary$Candidates

knitr::kable(
  grid_summary,
  caption = "Canonical ANN tuning grid"
)

CANDIDATES_PER_TASK <- sum(
  grid_summary$Candidates
)
INNER_FITS_PER_TASK <- 4L *
  CANDIDATES_PER_TASK

cat(
  "Candidates per outer task:",
  CANDIDATES_PER_TASK,
  "\n"
)
cat(
  "Inner candidate-fold fits per outer task:",
  INNER_FITS_PER_TASK,
  "\n"
)
cat(
  "Planned inner fits for all 150 outer tasks:",
  150L * INNER_FITS_PER_TASK,
  "\n"
)
```

# Support functions

```{r support-functions}
safe_cor <- function(p, o) {
  z <- is.finite(p) & is.finite(o)

  if (
    sum(z) < 3L ||
    sd(p[z]) == 0 ||
    sd(o[z]) == 0
  ) {
    return(NA_real_)
  }

  cor(p[z], o[z])
}

safe_rmse <- function(p, o) {
  z <- is.finite(p) & is.finite(o)

  if (!any(z)) {
    return(NA_real_)
  }

  sqrt(
    mean(
      (p[z] - o[z])^2
    )
  )
}

safe_mae <- function(p, o) {
  z <- is.finite(p) & is.finite(o)

  if (!any(z)) {
    return(NA_real_)
  }

  mean(
    abs(
      p[z] - o[z]
    )
  )
}

parameter_count <- function(
  input_dim,
  hidden_units
) {
  inputs <- c(
    input_dim,
    hidden_units
  )
  outputs <- c(
    hidden_units,
    1L
  )

  sum(
    inputs * outputs +
      outputs
  )
}

make_validation_split <- function(
  x,
  frac,
  seed
) {
  set.seed(seed)

  n_validation <- max(
    1L,
    as.integer(
      round(
        length(x) * frac
      )
    )
  )

  validation <- sort(
    sample(
      x,
      n_validation
    )
  )

  list(
    train = sort(
      setdiff(
        x,
        validation
      )
    ),
    validation = validation
  )
}

fit_x_scaler <- function(x) {
  center <- colMeans(x)
  scale <- apply(
    x,
    2,
    sd
  )

  zero <- !is.finite(scale) |
    scale < 1e-8

  scale[zero] <- 1

  list(
    center = center,
    scale = scale,
    zero = zero
  )
}

scale_x <- function(x, scaler) {
  sweep(
    sweep(
      x,
      2,
      scaler$center,
      "-"
    ),
    2,
    scaler$scale,
    "/"
  )
}

fit_y_scaler <- function(y) {
  scale <- sd(y)

  if (
    !is.finite(scale) ||
    scale < 1e-8
  ) {
    scale <- 1
  }

  list(
    center = mean(y),
    scale = scale
  )
}

scale_y <- function(y, scaler) {
  (
    y - scaler$center
  ) / scaler$scale
}

unscale_y <- function(y, scaler) {
  y * scaler$scale +
    scaler$center
}
```

# Shared ANN implementation

```{r model-definition}
TorchGenomicMLP <- torch::nn_module(
  "TorchGenomicMLP_CanonicalTuning",

  initialize = function(
    input_dim,
    hidden_units
  ) {
    self$linears <-
      torch::nn_module_list()

    previous <-
      as.integer(input_dim)

    for (h in hidden_units) {
      self$linears$append(
        torch::nn_linear(
          previous,
          as.integer(h)
        )
      )
      previous <- as.integer(h)
    }

    self$output <-
      torch::nn_linear(
        previous,
        1L
      )
  },

  forward = function(x) {
    if (length(self$linears) > 0L) {
      for (
        i in seq_len(
          length(self$linears)
        )
      ) {
        x <- torch::nnf_relu(
          self$linears[[i]](x)
        )
      }
    }

    self$output(x)
  }
)

weight_penalty <- function(
  model,
  l1,
  l2
) {
  weights <- lapply(
    seq_len(
      length(model$linears)
    ),
    function(i) {
      model$linears[[i]]$weight
    }
  )

  weights[[length(weights) + 1L]] <-
    model$output$weight

  zero <- weights[[1]]$sum() * 0
  penalty_l1 <- zero
  penalty_l2 <- zero

  for (w in weights) {
    penalty_l1 <-
      penalty_l1 +
      w$abs()$sum()

    penalty_l2 <-
      penalty_l2 +
      w$pow(2)$sum()
  }

  l1 * penalty_l1 +
    (l2 / 2) * penalty_l2
}

clone_state <- function(model) {
  state <- model$state_dict()

  copy <- lapply(
    state,
    function(x) {
      x$detach()$clone()
    }
  )

  names(copy) <- names(state)

  copy
}
```

# Inner-CV training

The inner analysis set is divided deterministically into:

- a fitting subset;
- an early-validation subset.

The frozen inner assessment set remains untouched until prediction of that
candidate. Preprocessing is estimated from the fitting subset only.

Early stopping is an optimization mechanism. Its epoch diagnostics are saved
but excluded from the tuning rule.

```{r training-functions}
prepare_bundle <- function(
  fit_ids,
  validation_ids,
  assessment_ids,
  scenario_index
) {
  fit_index <- match(
    fit_ids,
    ids
  )
  validation_index <- match(
    validation_ids,
    ids
  )
  assessment_index <- match(
    assessment_ids,
    ids
  )

  if (
    anyNA(
      c(
        fit_index,
        validation_index,
        assessment_index
      )
    )
  ) {
    stop(
      "ANN ID alignment failed."
    )
  }

  x_scaler <- fit_x_scaler(
    geno[
      fit_index,
      ,
      drop = FALSE
    ]
  )

  y_scaler <- fit_y_scaler(
    phenotype[
      fit_index,
      scenario_index
    ]
  )

  list(
    x_fit = scale_x(
      geno[
        fit_index,
        ,
        drop = FALSE
      ],
      x_scaler
    ),
    x_validation = scale_x(
      geno[
        validation_index,
        ,
        drop = FALSE
      ],
      x_scaler
    ),
    x_assessment = scale_x(
      geno[
        assessment_index,
        ,
        drop = FALSE
      ],
      x_scaler
    ),
    y_fit = scale_y(
      phenotype[
        fit_index,
        scenario_index
      ],
      y_scaler
    ),
    y_validation = scale_y(
      phenotype[
        validation_index,
        scenario_index
      ],
      y_scaler
    ),
    y_assessment = phenotype[
      assessment_index,
      scenario_index
    ],
    y_scaler = y_scaler,
    zero_variance_markers =
      sum(x_scaler$zero)
  )
}

train_inner_candidate <- function(
  bundle,
  architecture,
  learning_rate,
  l1,
  l2,
  seed
) {
  set_ann_seed(seed)

  x_fit <- torch::torch_tensor(
    bundle$x_fit,
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  y_fit <- torch::torch_tensor(
    matrix(
      bundle$y_fit,
      ncol = 1
    ),
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  x_validation <- torch::torch_tensor(
    bundle$x_validation,
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  y_validation <- torch::torch_tensor(
    matrix(
      bundle$y_validation,
      ncol = 1
    ),
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  x_assessment <- torch::torch_tensor(
    bundle$x_assessment,
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  model <- TorchGenomicMLP(
    ncol(bundle$x_fit),
    ARCHITECTURES[[architecture]]
  )$to(device = ANN_DEVICE)

  optimizer <- torch::optim_adam(
    params = model$parameters,
    lr = learning_rate,
    weight_decay = 0
  )

  best_validation_loss <- Inf
  checkpoint_epoch <- NA_integer_
  checkpoint_state <- NULL
  stale <- 0L
  last_epoch <- 0L
  stopped_by_patience <- FALSE

  sync_device()
  fit_start <-
    proc.time()[["elapsed"]]

  for (
    epoch in seq_len(MAX_EPOCHS)
  ) {
    last_epoch <- epoch

    model$train()
    optimizer$zero_grad()

    prediction <-
      model(x_fit)

    loss <-
      torch::nnf_mse_loss(
        prediction,
        y_fit
      ) +
      weight_penalty(
        model,
        l1,
        l2
      )

    loss$backward()
    optimizer$step()

    model$eval()

    validation_loss <-
      torch::with_no_grad({
        torch::nnf_mse_loss(
          model(x_validation),
          y_validation
        )
      })

    validation_value <-
      as.numeric(
        validation_loss$to(
          device =
            torch::torch_device("cpu")
        )
      )

    if (
      is.finite(validation_value) &&
      validation_value <
        best_validation_loss -
        MIN_DELTA
    ) {
      best_validation_loss <-
        validation_value

      checkpoint_epoch <-
        epoch

      checkpoint_state <-
        clone_state(model)

      stale <- 0L
    } else {
      stale <- stale + 1L
    }

    if (stale >= PATIENCE) {
      stopped_by_patience <- TRUE
      break
    }
  }

  sync_device()
  fit_seconds <-
    proc.time()[["elapsed"]] -
    fit_start

  if (
    is.null(checkpoint_state) ||
    is.na(checkpoint_epoch)
  ) {
    stop(
      "No finite early-validation checkpoint."
    )
  }

  model$load_state_dict(
    checkpoint_state
  )
  model$eval()

  sync_device()
  prediction_start <-
    proc.time()[["elapsed"]]

  prediction_scaled <-
    torch::with_no_grad({
      as.numeric(
        model(x_assessment)$to(
          device =
            torch::torch_device("cpu")
        )
      )
    })

  sync_device()
  prediction_seconds <-
    proc.time()[["elapsed"]] -
    prediction_start

  list(
    prediction = unscale_y(
      prediction_scaled,
      bundle$y_scaler
    ),
    early_stop_epoch =
      as.integer(checkpoint_epoch),
    last_epoch =
      as.integer(last_epoch),
    stopped_by_patience =
      stopped_by_patience,
    hit_epoch_ceiling =
      last_epoch >= MAX_EPOCHS &&
      !stopped_by_patience,
    fit_seconds = fit_seconds,
    prediction_seconds =
      prediction_seconds,
    zero_variance_markers =
      bundle$zero_variance_markers
  )
}
```

# Candidate evaluation and selection

Candidate ranking is deterministic and does **not** include epoch information.

The ranking hierarchy is:

1. lower mean inner-assessment RMSE;
2. higher mean inner-assessment correlation;
3. fewer network parameters;
4. smaller total penalty coefficient;
5. smaller learning rate;
6. candidate ID.

```{r candidate-functions}
evaluate_candidate <- function(
  bundles,
  model_name,
  architecture,
  learning_rate,
  l1,
  l2,
  seed_base,
  candidate_id
) {
  rows <- vector(
    "list",
    4L
  )

  for (k in 1:4) {
    fitted <- tryCatch(
      train_inner_candidate(
        bundles[[k]],
        architecture,
        learning_rate,
        l1,
        l2,
        seed_base + k
      ),
      error = function(e) e
    )

    if (
      inherits(
        fitted,
        "error"
      )
    ) {
      rows[[k]] <- data.frame(
        Model = model_name,
        CandidateID = candidate_id,
        Architecture = architecture,
        Learning_rate = learning_rate,
        L1 = l1,
        L2 = l2,
        InnerFold = k,
        RMSE = NA_real_,
        Correlation = NA_real_,
        MAE = NA_real_,
        EarlyStop_epoch = NA_integer_,
        Last_epoch = NA_integer_,
        Stopped_by_patience = NA,
        Hit_epoch_ceiling = NA,
        Fit_seconds = NA_real_,
        Prediction_seconds = NA_real_,
        Status = "error",
        Error = conditionMessage(fitted),
        stringsAsFactors = FALSE
      )
    } else {
      rows[[k]] <- data.frame(
        Model = model_name,
        CandidateID = candidate_id,
        Architecture = architecture,
        Learning_rate = learning_rate,
        L1 = l1,
        L2 = l2,
        InnerFold = k,
        RMSE = safe_rmse(
          fitted$prediction,
          bundles[[k]]$y_assessment
        ),
        Correlation = safe_cor(
          fitted$prediction,
          bundles[[k]]$y_assessment
        ),
        MAE = safe_mae(
          fitted$prediction,
          bundles[[k]]$y_assessment
        ),
        EarlyStop_epoch =
          fitted$early_stop_epoch,
        Last_epoch =
          fitted$last_epoch,
        Stopped_by_patience =
          fitted$stopped_by_patience,
        Hit_epoch_ceiling =
          fitted$hit_epoch_ceiling,
        Fit_seconds =
          fitted$fit_seconds,
        Prediction_seconds =
          fitted$prediction_seconds,
        Status = "ok",
        Error = NA_character_,
        stringsAsFactors = FALSE
      )
    }
  }

  folds <- do.call(
    rbind,
    rows
  )

  ok <- folds$Status == "ok"

  summary <- data.frame(
    Model = model_name,
    CandidateID = candidate_id,
    Architecture = architecture,
    Learning_rate = learning_rate,
    L1 = l1,
    L2 = l2,
    Inner_folds_ok = sum(ok),
    Mean_RMSE = if (all(ok)) {
      mean(folds$RMSE)
    } else {
      NA_real_
    },
    Mean_Correlation = if (all(ok)) {
      mean(folds$Correlation)
    } else {
      NA_real_
    },
    Mean_MAE = if (all(ok)) {
      mean(folds$MAE)
    } else {
      NA_real_
    },
    Median_EarlyStop_epoch = if (all(ok)) {
      median(folds$EarlyStop_epoch)
    } else {
      NA_real_
    },
    Max_Last_epoch = if (all(ok)) {
      max(folds$Last_epoch)
    } else {
      NA_real_
    },
    N_epoch_ceiling = if (all(ok)) {
      sum(folds$Hit_epoch_ceiling)
    } else {
      NA_integer_
    },
    Status = if (all(ok)) {
      "ok"
    } else {
      "incomplete"
    },
    stringsAsFactors = FALSE
  )

  list(
    summary = summary,
    folds = folds
  )
}

choose_candidate <- function(x) {
  eligible <- x[
    x$Status == "ok" &
      is.finite(x$Mean_RMSE),
    ,
    drop = FALSE
  ]

  if (!nrow(eligible)) {
    stop(
      "No complete ANN candidate is available."
    )
  }

  eligible$Parameters <- vapply(
    eligible$Architecture,
    function(architecture) {
      parameter_count(
        ncol(geno),
        ARCHITECTURES[[architecture]]
      )
    },
    numeric(1)
  )

  eligible$Penalty <-
    eligible$L1 +
    eligible$L2

  eligible <- eligible[
    order(
      eligible$Mean_RMSE,
      -eligible$Mean_Correlation,
      eligible$Parameters,
      eligible$Penalty,
      eligible$Learning_rate,
      eligible$CandidateID
    ),
    ,
    drop = FALSE
  ]

  # Epoch diagnostics are intentionally absent from the ranking above.
  eligible[
    1,
    ,
    drop = FALSE
  ]
}
```

# Outer-task table

The six scenarios share the same 25 frozen outer splits, producing 150
`Scenario × outer split` tuning tasks.

```{r tasks}
outer_map <- unique(
  cv$outer_assignments[
    ,
    c(
      "Repetition",
      "OuterFold",
      "SplitID"
    )
  ]
)

outer_map <- outer_map[
  order(
    outer_map$Repetition,
    outer_map$OuterFold
  ),
  ,
  drop = FALSE
]

task_table <- merge(
  data.frame(
    Scenario = scenario_names,
    stringsAsFactors = FALSE
  ),
  outer_map,
  by = NULL
)

task_table <- task_table[
  order(
    match(
      task_table$Scenario,
      scenario_names
    ),
    task_table$Repetition,
    task_table$OuterFold
  ),
  ,
  drop = FALSE
]

task_table$TaskID <- paste(
  task_table$Scenario,
  task_table$SplitID,
  sep = "_"
)

task_table$Fold <-
  task_table$OuterFold

rownames(task_table) <- NULL

if (nrow(task_table) != 150L) {
  stop(
    "Expected exactly 150 ANN tuning tasks."
  )
}

full_task_table <- task_table

if (ANN_TUNING_PILOT) {
  if (
    !ANN_TUNING_PILOT_TASK %in%
      full_task_table$TaskID
  ) {
    stop(
      paste(
        "Unknown ANN_TUNING_PILOT_TASK:",
        ANN_TUNING_PILOT_TASK
      )
    )
  }

  task_table <- full_task_table[
    full_task_table$TaskID ==
      ANN_TUNING_PILOT_TASK,
    ,
    drop = FALSE
  ]
}

EXPECTED_TASKS <- nrow(task_table)
EXPECTED_SELECTED_ROWS <-
  EXPECTED_TASKS *
  length(MODEL_LEVELS)
EXPECTED_CANDIDATE_ROWS <-
  EXPECTED_TASKS *
  CANDIDATES_PER_TASK
EXPECTED_FOLD_ROWS <-
  EXPECTED_TASKS *
  INNER_FITS_PER_TASK

cat(
  "Mode:",
  if (ANN_TUNING_PILOT) {
    "PILOT"
  } else {
    "FULL"
  },
  "\n"
)

cat(
  "Tasks:",
  EXPECTED_TASKS,
  "\n"
)

if (ANN_TUNING_PILOT) {
  cat(
    "Pilot task:",
    ANN_TUNING_PILOT_TASK,
    "\n"
  )
}

cat(
  "Configuration ID:",
  ANN_TUNING_CONFIG_ID,
  "\n"
)

cat(
  "Expected selected rows:",
  EXPECTED_SELECTED_ROWS,
  "\n"
)
cat(
  "Expected candidate rows:",
  EXPECTED_CANDIDATE_ROWS,
  "\n"
)
cat(
  "Expected fold rows:",
  EXPECTED_FOLD_ROWS,
  "\n"
)

knitr::kable(
  head(
    task_table,
    15
  ),
  caption = if (
    ANN_TUNING_PILOT
  ) {
    "Canonical ANN tuning pilot task"
  } else {
    "First 15 canonical ANN tuning tasks"
  }
)
```

# Canonical nested tuning

This section evaluates only inner-CV candidates. It never predicts the outer
assessment set.

```{r run-tuning, eval=RUN_ANN_TUNING}
run_start <-
  proc.time()[["elapsed"]]

task_outputs <- vector(
  "list",
  nrow(task_table)
)

for (
  task_index in seq_len(
    nrow(task_table)
  )
) {
  task <- task_table[
    task_index,
    ,
    drop = FALSE
  ]

  scenario <- task$Scenario
  scenario_index <- match(
    scenario,
    scenario_names
  )
  split_id <- task$SplitID

  checkpoint_file <- file.path(
    CHECKPOINT_DIR,
    paste0(
      task$TaskID,
      ".rds"
    )
  )

  if (
    file.exists(checkpoint_file) &&
    !ANN_TUNING_FORCE
  ) {
    checkpoint <-
      readRDS(checkpoint_file)

    if (
      identical(
        checkpoint$Status,
        "ok"
      ) &&
      identical(
        checkpoint$ConfigID,
        ANN_TUNING_CONFIG_ID
      )
    ) {
      task_outputs[[task_index]] <-
        checkpoint

      cat(
        sprintf(
          "[%03d/%03d] %s | checkpoint reused\n",
          task_index,
          nrow(task_table),
          task$TaskID
        )
      )

      next
    } else {
      cat(
        sprintf(
          "[%03d/%03d] %s | incompatible checkpoint ignored\n",
          task_index,
          nrow(task_table),
          task$TaskID
        )
      )
    }
  }

  task_start <-
    proc.time()[["elapsed"]]

  inner <- cv$inner_splits[[split_id]]

  if (length(inner) != 4L) {
    stop(
      "Expected four frozen inner folds."
    )
  }

  bundles <- vector(
    "list",
    4L
  )

  for (k in 1:4) {
    inner_split <- inner[[paste0("I", k)]]

    validation_seed <-
      SEED_GLOBAL +
      scenario_index * 1000000L +
      task$Repetition * 10000L +
      task$OuterFold * 1000L +
      k * 10L

    validation <- make_validation_split(
      inner_split$analysis_ids,
      EARLY_VALIDATION_FRACTION,
      validation_seed
    )

    bundles[[k]] <- prepare_bundle(
      validation$train,
      validation$validation,
      inner_split$assessment_ids,
      scenario_index
    )
  }

  base_seed <-
    SEED_GLOBAL +
    scenario_index * 10000000L +
    task$Repetition * 100000L +
    task$OuterFold * 10000L

  structure_summary_rows <- vector(
    "list",
    nrow(structure_grid)
  )
  structure_fold_rows <- vector(
    "list",
    nrow(structure_grid)
  )

  for (
    candidate_index in seq_len(
      nrow(structure_grid)
    )
  ) {
    candidate <-
      structure_grid[
        candidate_index,
        ,
        drop = FALSE
      ]

    evaluated <- evaluate_candidate(
      bundles = bundles,
      model_name = "ANN_0",
      architecture =
        candidate$Architecture,
      learning_rate =
        candidate$Learning_rate,
      l1 = 0,
      l2 = 0,
      seed_base =
        base_seed + 1000000L,
      candidate_id =
        candidate_index
    )

    structure_summary_rows[[candidate_index]] <-
      evaluated$summary

    structure_fold_rows[[candidate_index]] <-
      evaluated$folds
  }

  structure_summary <- do.call(
    rbind,
    structure_summary_rows
  )
  structure_folds <- do.call(
    rbind,
    structure_fold_rows
  )

  chosen_structure <-
    choose_candidate(
      structure_summary
    )

  selected_architecture <-
    chosen_structure$Architecture

  selected_learning_rate <-
    chosen_structure$Learning_rate

  penalty_grids <- list(
    ANN_L1 = l1_grid,
    ANN_L2 = l2_grid,
    ANN_L1_L2 = l1l2_grid
  )

  penalty_summaries <- list()
  penalty_folds <- list()
  penalty_selected <- list()

  for (
    model_index in seq_along(
      penalty_grids
    )
  ) {
    model_name <-
      names(penalty_grids)[
        model_index
      ]

    grid <-
      penalty_grids[[model_index]]

    candidate_summaries <- vector(
      "list",
      nrow(grid)
    )
    candidate_folds <- vector(
      "list",
      nrow(grid)
    )

    for (
      candidate_index in seq_len(
        nrow(grid)
      )
    ) {
      evaluated <- evaluate_candidate(
        bundles = bundles,
        model_name = model_name,
        architecture =
          selected_architecture,
        learning_rate =
          selected_learning_rate,
        l1 =
          grid$L1[candidate_index],
        l2 =
          grid$L2[candidate_index],
        seed_base =
          base_seed + 2000000L,
        candidate_id =
          candidate_index
      )

      candidate_summaries[[candidate_index]] <-
        evaluated$summary

      candidate_folds[[candidate_index]] <-
        evaluated$folds
    }

    penalty_summaries[[model_name]] <-
      do.call(
        rbind,
        candidate_summaries
      )

    penalty_folds[[model_name]] <-
      do.call(
        rbind,
        candidate_folds
      )

    penalty_selected[[model_name]] <-
      choose_candidate(
        penalty_summaries[[model_name]]
      )
  }

  selected_table <- rbind(
    data.frame(
      Model = "ANN_0",
      Architecture =
        selected_architecture,
      Learning_rate =
        selected_learning_rate,
      L1 = 0,
      L2 = 0,
      Inner_RMSE =
        chosen_structure$Mean_RMSE,
      Inner_Correlation =
        chosen_structure$Mean_Correlation,
      Inner_MAE =
        chosen_structure$Mean_MAE,
      stringsAsFactors = FALSE
    ),
    do.call(
      rbind,
      lapply(
        names(penalty_selected),
        function(model_name) {
          selected_candidate <-
            penalty_selected[[model_name]]

          data.frame(
            Model = model_name,
            Architecture =
              selected_architecture,
            Learning_rate =
              selected_learning_rate,
            L1 =
              selected_candidate$L1,
            L2 =
              selected_candidate$L2,
            Inner_RMSE =
              selected_candidate$Mean_RMSE,
            Inner_Correlation =
              selected_candidate$Mean_Correlation,
            Inner_MAE =
              selected_candidate$Mean_MAE,
            stringsAsFactors = FALSE
          )
        }
      )
    )
  )

  for (
    column_name in c(
      "Scenario",
      "Repetition",
      "Fold",
      "SplitID",
      "TaskID"
    )
  ) {
    selected_table[[column_name]] <-
      task[[column_name]]
  }

  tuning_summary <- rbind(
    structure_summary,
    do.call(
      rbind,
      penalty_summaries
    )
  )

  tuning_folds <- rbind(
    structure_folds,
    do.call(
      rbind,
      penalty_folds
    )
  )

  for (
    column_name in c(
      "Scenario",
      "Repetition",
      "Fold",
      "SplitID",
      "TaskID"
    )
  ) {
    tuning_summary[[column_name]] <-
      task[[column_name]]

    tuning_folds[[column_name]] <-
      task[[column_name]]
  }

  task_seconds <-
    proc.time()[["elapsed"]] -
    task_start

  checkpoint <- list(
    Status = "ok",
    ConfigID =
      ANN_TUNING_CONFIG_ID,
    TaskID = task$TaskID,
    selected = selected_table,
    tuning_summary = tuning_summary,
    tuning_folds = tuning_folds,
    task_seconds = task_seconds
  )

  saveRDS(
    checkpoint,
    checkpoint_file,
    compress = "gzip"
  )

  task_outputs[[task_index]] <-
    checkpoint

  cat(
    sprintf(
      "[%03d/%03d] %s | %.1f s | arch=%s lr=%g\n",
      task_index,
      nrow(task_table),
      task$TaskID,
      task_seconds,
      selected_architecture,
      selected_learning_rate
    )
  )

  gc()
}

current_session_seconds <-
  proc.time()[["elapsed"]] -
  run_start

selected_configurations <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(x) {
      x$selected
    }
  )
)

candidate_summary <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(x) {
      x$tuning_summary
    }
  )
)

fold_results <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(x) {
      x$tuning_folds
    }
  )
)

checkpoint_seconds <- sum(
  vapply(
    task_outputs,
    function(x) {
      x$task_seconds
    },
    numeric(1)
  )
)

manifest <- list(
  analysis =
    "canonical_simulated_ann_tuning",
  dataset = "simulated",
  mode = if (
    ANN_TUNING_PILOT
  ) {
    "pilot"
  } else {
    "full"
  },
  pilot_task = if (
    ANN_TUNING_PILOT
  ) {
    ANN_TUNING_PILOT_TASK
  } else {
    NA_character_
  },
  config_id =
    ANN_TUNING_CONFIG_ID,
  seed_global = SEED_GLOBAL,
  models = MODEL_LEVELS,
  architectures = ARCHITECTURES,
  learning_rates =
    LEARNING_RATES,
  l1_values = L1_VALUES,
  l2_values = L2_VALUES,
  l1_l2_grid = L1_L2_GRID,
  optimizer = "Adam",
  activation = "ReLU",
  dropout = DROPOUT,
  max_epochs = MAX_EPOCHS,
  final_train_epochs =
    TUNING_V1_COMPATIBILITY_TOKEN,
  patience = PATIENCE,
  min_delta = MIN_DELTA,
  early_validation_fraction =
    EARLY_VALIDATION_FRACTION,
  primary_tuning_metric =
    "mean inner-assessment RMSE",
  secondary_tuning_metric =
    "mean inner-assessment correlation",
  epoch_used_in_candidate_ranking =
    FALSE,
  outer_assessment_used =
    FALSE,
  true_genetic_value_used =
    FALSE,
  architecture_lr_shared_across_models =
    TRUE,
  candidates_per_task =
    CANDIDATES_PER_TASK,
  inner_fits_per_task =
    INNER_FITS_PER_TASK,
  number_tasks =
    EXPECTED_TASKS,
  selected_rows =
    nrow(selected_configurations),
  candidate_rows =
    nrow(candidate_summary),
  fold_rows =
    nrow(fold_results),
  torch_version =
    as.character(
      packageVersion("torch")
    ),
  device = ANN_DEVICE_LABEL,
  cuda_available =
    cuda_available,
  current_session_seconds =
    current_session_seconds,
  checkpoint_task_seconds =
    checkpoint_seconds,
  timestamp =
    as.character(Sys.time())
)

manifest_table <- data.frame(
  Field = c(
    "analysis",
    "mode",
    "config_id",
    "architectures",
    "learning_rates",
    "L1_values",
    "L2_values",
    "combined_grid_size",
    "MAX_EPOCHS",
    "TUNING_V1_COMPATIBILITY_TOKEN",
    "patience",
    "dropout",
    "optimizer",
    "activation",
    "candidates_per_task",
    "inner_fits_per_task",
    "tasks",
    "epoch_used_in_candidate_ranking",
    "outer_assessment_used",
    "true_genetic_value_used"
  ),
  Value = c(
    manifest$analysis,
    manifest$mode,
    manifest$config_id,
    paste(
      names(ARCHITECTURES),
      collapse = " | "
    ),
    paste(
      LEARNING_RATES,
      collapse = " | "
    ),
    paste(
      L1_VALUES,
      collapse = " | "
    ),
    paste(
      L2_VALUES,
      collapse = " | "
    ),
    nrow(L1_L2_GRID),
    MAX_EPOCHS,
    TUNING_V1_COMPATIBILITY_TOKEN,
    PATIENCE,
    DROPOUT,
    "Adam",
    "ReLU",
    CANDIDATES_PER_TASK,
    INNER_FITS_PER_TASK,
    EXPECTED_TASKS,
    FALSE,
    FALSE,
    FALSE
  ),
  stringsAsFactors = FALSE
)

saveRDS(
  selected_configurations,
  SELECTED_RDS,
  compress = "gzip"
)
saveRDS(
  candidate_summary,
  CANDIDATE_RDS,
  compress = "gzip"
)
saveRDS(
  fold_results,
  FOLD_RDS,
  compress = "gzip"
)
saveRDS(
  manifest,
  RUN_RDS,
  compress = "gzip"
)
saveRDS(
  manifest,
  MANIFEST_RDS,
  compress = "gzip"
)

write.csv(
  selected_configurations,
  SELECTED_CSV,
  row.names = FALSE
)
write.csv(
  candidate_summary,
  CANDIDATE_CSV,
  row.names = FALSE
)
write.csv(
  fold_results,
  FOLD_CSV,
  row.names = FALSE
)
write.csv(
  manifest_table,
  MANIFEST_CSV,
  row.names = FALSE
)
```

# Saved-result integrity

```{r saved-integrity}
available <- all(
  file.exists(
    c(
      SELECTED_RDS,
      CANDIDATE_RDS,
      FOLD_RDS,
      RUN_RDS
    )
  )
)

if (!available) {
  cat(
    "Canonical ANN tuning outputs are not available yet. ",
    "Run with RUN_ANN_TUNING=1.\n"
  )
} else {
  selected_configurations <-
    readRDS(SELECTED_RDS)
  candidate_summary <-
    readRDS(CANDIDATE_RDS)
  fold_results <-
    readRDS(FOLD_RDS)
  run_metadata <-
    readRDS(RUN_RDS)

  integrity <- data.frame(
    Check = c(
      "Configuration ID matches",
      "Selected rows",
      "Candidate rows",
      "Fold rows",
      "All candidate rows are ok",
      "All fold rows are ok",
      "All candidate RMSE values are finite",
      "All fold RMSE values are finite",
      "Epoch excluded from ranking",
      "Outer assessment excluded",
      "True genetic values excluded",
      "Architecture/LR shared within task"
    ),
    Status = c(
      if (
        identical(
          run_metadata$config_id,
          ANN_TUNING_CONFIG_ID
        )
      ) "PASS" else "BLOCKED",
      if (
        nrow(selected_configurations) ==
        EXPECTED_SELECTED_ROWS
      ) "PASS" else "BLOCKED",
      if (
        nrow(candidate_summary) ==
        EXPECTED_CANDIDATE_ROWS
      ) "PASS" else "BLOCKED",
      if (
        nrow(fold_results) ==
        EXPECTED_FOLD_ROWS
      ) "PASS" else "BLOCKED",
      if (
        all(
          candidate_summary$Status ==
            "ok"
        )
      ) "PASS" else "BLOCKED",
      if (
        all(
          fold_results$Status ==
            "ok"
        )
      ) "PASS" else "BLOCKED",
      if (
        all(
          is.finite(
            candidate_summary$Mean_RMSE
          )
        )
      ) "PASS" else "BLOCKED",
      if (
        all(
          is.finite(
            fold_results$RMSE
          )
        )
      ) "PASS" else "BLOCKED",
      if (
        identical(
          run_metadata$epoch_used_in_candidate_ranking,
          FALSE
        )
      ) "PASS" else "BLOCKED",
      if (
        identical(
          run_metadata$outer_assessment_used,
          FALSE
        )
      ) "PASS" else "BLOCKED",
      if (
        identical(
          run_metadata$true_genetic_value_used,
          FALSE
        )
      ) "PASS" else "BLOCKED",
      {
        task_structure <- aggregate(
          cbind(
            Architecture_code =
              as.numeric(
                factor(
                  selected_configurations$Architecture,
                  levels =
                    names(ARCHITECTURES)
                )
              ),
            Learning_rate =
              selected_configurations$Learning_rate
          ) ~ TaskID,
          data =
            selected_configurations,
          FUN = function(z) {
            length(unique(z))
          }
        )

        if (
          all(
            task_structure$Architecture_code ==
              1
          ) &&
          all(
            task_structure$Learning_rate ==
              1
          )
        ) {
          "PASS"
        } else {
          "BLOCKED"
        }
      }
    ),
    stringsAsFactors = FALSE
  )

  knitr::kable(
    integrity,
    caption =
      "Canonical ANN tuning saved-result integrity"
  )

  if (
    any(
      integrity$Status ==
        "BLOCKED"
    )
  ) {
    stop(
      "Canonical ANN tuning failed a blocking integrity check."
    )
  }
}
```

# Selected configurations

```{r selected-summary}
if (available) {
  cat(
    "Selected architecture frequencies:\n"
  )
  print(
    table(
      unique(
        selected_configurations[
          ,
          c(
            "TaskID",
            "Architecture"
          )
        ]
      )$Architecture
    )
  )

  cat(
    "\nSelected learning-rate frequencies:\n"
  )
  print(
    table(
      unique(
        selected_configurations[
          ,
          c(
            "TaskID",
            "Learning_rate"
          )
        ]
      )$Learning_rate
    )
  )

  cat(
    "\nSelected L1 values:\n"
  )
  print(
    table(
      selected_configurations$Model,
      selected_configurations$L1
    )
  )

  cat(
    "\nSelected L2 values:\n"
  )
  print(
    table(
      selected_configurations$Model,
      selected_configurations$L2
    )
  )

  knitr::kable(
    head(
      selected_configurations[
        order(
          selected_configurations$Scenario,
          selected_configurations$Repetition,
          selected_configurations$Fold,
          match(
            selected_configurations$Model,
            MODEL_LEVELS
          )
        ),
        ,
        drop = FALSE
      ],
      20
    ),
    digits = 6,
    caption =
      "First selected canonical ANN configurations"
  )
}
```

# Optimization diagnostics

These diagnostics describe the fixed optimization protocol. They do not change
the tuning grid.

```{r optimization-diagnostics}
if (available) {
  cat(
    "EarlyStop_epoch summary:\n"
  )
  print(
    summary(
      fold_results$EarlyStop_epoch
    )
  )

  cat(
    "\nLast_epoch summary:\n"
  )
  print(
    summary(
      fold_results$Last_epoch
    )
  )

  cat(
    "\nFits stopped by patience:",
    sum(
      fold_results$Stopped_by_patience,
      na.rm = TRUE
    ),
    "/",
    nrow(fold_results),
    "\n"
  )

  cat(
    "Fits reaching the 2000-epoch ceiling without patience stop:",
    sum(
      fold_results$Hit_epoch_ceiling,
      na.rm = TRUE
    ),
    "/",
    nrow(fold_results),
    "\n"
  )

  selected_keys <- merge(
    selected_configurations[
      ,
      c(
        "TaskID",
        "Model",
        "Architecture",
        "Learning_rate",
        "L1",
        "L2"
      )
    ],
    fold_results,
    by = c(
      "TaskID",
      "Model",
      "Architecture",
      "Learning_rate",
      "L1",
      "L2"
    ),
    all.x = TRUE,
    sort = FALSE
  )

  cat(
    "\nSelected-candidate inner folds reaching ceiling:",
    sum(
      selected_keys$Hit_epoch_ceiling,
      na.rm = TRUE
    ),
    "/",
    nrow(selected_keys),
    "\n"
  )
}
```

# Boundary description

Boundary occupancy is reported descriptively because the upper architecture
`64-32` is a deliberate methodological cap. Selection of `64-32` indicates that
the bounded search space was active for those tasks, but it does not trigger
post hoc enlargement of the canonical grid.

Consequently, boundary occupancy is treated as a limitation of the prespecified
canonical comparison rather than as justification for extending the search
space after observing the results.

```{r boundary-description}
if (available) {
  task_level <- unique(
    selected_configurations[
      ,
      c(
        "TaskID",
        "Scenario",
        "Architecture",
        "Learning_rate"
      )
    ]
  )

  boundary_table <- data.frame(
    Boundary = c(
      "Architecture = 64-32",
      "Learning rate = 1.25e-4",
      "ANN_L1: L1 = 1e-1",
      "ANN_L1_L2: L1 = 1e-1",
      "ANN_L2: L2 = 1e-2",
      "ANN_L1_L2: L2 = 1e-2"
    ),
    Selected = c(
      sum(
        task_level$Architecture ==
          "64-32"
      ),
      sum(
        abs(
          task_level$Learning_rate -
            min(LEARNING_RATES)
        ) < 1e-14
      ),
      sum(
        selected_configurations$Model ==
          "ANN_L1" &
        abs(
          selected_configurations$L1 -
            max(L1_VALUES)
        ) < 1e-14
      ),
      sum(
        selected_configurations$Model ==
          "ANN_L1_L2" &
        abs(
          selected_configurations$L1 -
            max(L1_VALUES)
        ) < 1e-14
      ),
      sum(
        selected_configurations$Model ==
          "ANN_L2" &
        abs(
          selected_configurations$L2 -
            max(L2_VALUES)
        ) < 1e-14
      ),
      sum(
        selected_configurations$Model ==
          "ANN_L1_L2" &
        abs(
          selected_configurations$L2 -
            max(L2_VALUES)
        ) < 1e-14
      )
    ),
    Denominator = c(
      150L,
      150L,
      150L,
      150L,
      150L,
      150L
    ),
    stringsAsFactors = FALSE
  )

  boundary_table$Proportion <-
    boundary_table$Selected /
    boundary_table$Denominator

  knitr::kable(
    boundary_table,
    digits = 4,
    caption =
      "Descriptive canonical tuning-boundary occupancy"
  )
}
```

# Reproducibility outputs

The canonical tuning module produces:

- candidate-level inner-CV summaries;
- fold-level candidate results;
- selected configuration for each `Scenario × outer split × ANN variant`;
- a tuning manifest documenting every prespecified grid and fixed training
  control.

The final-fitting stage below consumes the selected configuration artifact and
the tuning manifest. It does not repeat hyperparameter tuning.

# Execution

Routine website build without model fitting:

```text
Rscript -e 'workflowr::wflow_build("analysis/simulated_data_ann.Rmd")'
```

One-task pilot:

```text
ANN_TUNING_PILOT=1 RUN_ANN_TUNING=1 \
Rscript -e 'workflowr::wflow_build("analysis/simulated_data_ann.Rmd")'
```

Complete canonical tuning:

```text
RUN_ANN_TUNING=1 \
Rscript -e 'workflowr::wflow_build("analysis/simulated_data_ann.Rmd")'
```

Completed tasks are checkpointed in
`.local/checkpoints/canonical/ann_tuning/`. Use `ANN_TUNING_FORCE=1` only when
an intentional full recomputation is required.

# Final ANN fitting

The canonical tuning stage retains `MAX_EPOCHS = 2000`. The final epoch-selection
stage uses the separate `FINAL_EPOCH_MAX_EPOCHS = 3000` safety ceiling. This
extension was introduced after an internal, outer-assessment-free convergence
diagnostic reproduced the single 2000-epoch ceiling case and reached natural
patience stopping at epoch 2044 (`Best_epoch = 2019`). The value 3000 is a
safety ceiling only; final networks are still refit for their selected
`Best_epoch`.

This stage consumes the saved `ANN_TUNING_CANONICAL_V1` selections. It never
repeats hyperparameter tuning when only `RUN_ANN=1` is set.

For each `Scenario × outer split × ANN variant`, `Best_epoch` is selected with
early stopping using only the 800 outer-analysis individuals. The network is
then reinitialized, refit on all 800 for exactly `Best_epoch`, and only then
applied to the untouched 200 outer-assessment individuals.

```{r final-ann-contract}
RUN_ANN <- identical(Sys.getenv("RUN_ANN", "0"), "1")
ANN_FORCE <- identical(Sys.getenv("ANN_FORCE", "0"), "1")
ANN_PILOT <- identical(Sys.getenv("ANN_PILOT", "0"), "1")
ANN_PILOT_TASK <- Sys.getenv("ANN_PILOT_TASK", "C1_R01_F01")

# The tuning ceiling remains MAX_EPOCHS = 2000. This separate ceiling applies
# only to Best_epoch selection inside the outer-analysis sample. A targeted
# residual-convergence diagnostic reproduced the unique 2000-epoch ceiling
# case exactly and reached natural patience stopping at epoch 2044, with
# Best_epoch = 2019, under a diagnostic ceiling of 3000.
FINAL_EPOCH_MAX_EPOCHS <- 3000L

FINAL_CHECKPOINT_DIR <- project_path(".local", "checkpoints", "canonical", "ann_final")

FINAL_OUTPUT_STEM <- if (ANN_PILOT) {
  project_path(".local", "pilot", "ann", "final", "simulated_ann_pilot")
} else {
  project_path("output", "simulated", "ann", "final", "simulated_ann")
}

FINAL_RESULTS_RDS <- paste0(FINAL_OUTPUT_STEM, "_results.rds")
FINAL_PREDICTIONS_RDS <- paste0(FINAL_OUTPUT_STEM, "_predictions.rds")
FINAL_EPOCH_RDS <- paste0(FINAL_OUTPUT_STEM, "_epoch_selection.rds")
FINAL_RUN_RDS <- paste0(FINAL_OUTPUT_STEM, "_run_metadata.rds")

FINAL_RESULTS_CSV <- paste0(FINAL_OUTPUT_STEM, "_results.csv")
FINAL_PREDICTIONS_CSV <- paste0(FINAL_OUTPUT_STEM, "_predictions.csv")
FINAL_EPOCH_CSV <- paste0(FINAL_OUTPUT_STEM, "_epoch_selection.csv")

if (RUN_ANN) {
  dir.create(
    FINAL_CHECKPOINT_DIR,
    recursive = TRUE,
    showWarnings = FALSE
  )
  dir.create(
    dirname(FINAL_RESULTS_RDS),
    recursive = TRUE,
    showWarnings = FALSE
  )
}

FINAL_CONFIG_ID <- paste0(
  "ANN_FINAL_CANONICAL_V2__FROM__",
  ANN_TUNING_CONFIG_ID,
  "__FINAL_EPOCH_CAP_",
  FINAL_EPOCH_MAX_EPOCHS,
  "__BEST_EPOCH_REFIT"
)

knitr::kable(data.frame(
  Item=c("Tuning source","Epoch ceiling","Final epoch rule","Outer assessment in tuning","Outer assessment in epoch selection","True G in selection"),
  Value=c("ANN_TUNING_CANONICAL_V1",FINAL_EPOCH_MAX_EPOCHS,"Best_epoch selected inside outer-analysis; refit all 800",FALSE,FALSE,FALSE)
), caption="Canonical final ANN contract")
```

# Final-fit helper functions

```{r final-ann-functions}
final_safe_bias <- function(p,o){z<-is.finite(p)&is.finite(o);if(!any(z))return(NA_real_);mean(p[z]-o[z])}
final_safe_slope <- function(p,o){z<-is.finite(p)&is.finite(o);if(sum(z)<3L||sd(p[z])==0)return(NA_real_);as.numeric(coef(lm(o[z]~p[z]))[2])}
final_safe_r2 <- function(p,o){z<-is.finite(p)&is.finite(o);if(sum(z)<2L)return(NA_real_);d<-sum((o[z]-mean(o[z]))^2);if(d==0)return(NA_real_);1-sum((o[z]-p[z])^2)/d}
final_metrics <- function(p,o,prefix){z<-c(safe_cor(p,o),safe_rmse(p,o),safe_mae(p,o),final_safe_bias(p,o),final_safe_slope(p,o),final_safe_r2(p,o));names(z)<-paste0(prefix,c("_Correlation","_RMSE","_MAE","_Bias","_Slope","_R2"));z}

select_final_epoch <- function(analysis_ids,scenario_index,architecture,learning_rate,l1,l2,split_seed,fit_seed){
  iv<-make_validation_split(analysis_ids,EARLY_VALIDATION_FRACTION,split_seed)
  fi<-match(iv$train,ids);vi<-match(iv$validation,ids);if(anyNA(c(fi,vi)))stop("Final epoch-selection ID alignment failed.")
  xs<-fit_x_scaler(geno[fi,,drop=FALSE]);ys<-fit_y_scaler(phenotype[fi,scenario_index])
  xf<-scale_x(geno[fi,,drop=FALSE],xs);xv<-scale_x(geno[vi,,drop=FALSE],xs);yf<-scale_y(phenotype[fi,scenario_index],ys);yv<-scale_y(phenotype[vi,scenario_index],ys)
  set_ann_seed(fit_seed)
  xt<-torch::torch_tensor(xf,dtype=torch::torch_float())$to(device=ANN_DEVICE);yt<-torch::torch_tensor(matrix(yf,ncol=1),dtype=torch::torch_float())$to(device=ANN_DEVICE)
  xv_t<-torch::torch_tensor(xv,dtype=torch::torch_float())$to(device=ANN_DEVICE);yv_t<-torch::torch_tensor(matrix(yv,ncol=1),dtype=torch::torch_float())$to(device=ANN_DEVICE)
  model<-TorchGenomicMLP(ncol(xf),ARCHITECTURES[[architecture]])$to(device=ANN_DEVICE);opt<-torch::optim_adam(params=model$parameters,lr=learning_rate,weight_decay=0)
  best<-Inf;best_epoch<-NA_integer_;best_state<-NULL;stale<-0L;last<-0L;stopped<-FALSE;sync_device();t0<-proc.time()[["elapsed"]]
  for(epoch in seq_len(FINAL_EPOCH_MAX_EPOCHS)){last<-epoch;model$train();opt$zero_grad();pr<-model(xt);loss<-torch::nnf_mse_loss(pr,yt)+weight_penalty(model,l1,l2);loss$backward();opt$step();model$eval();vl<-torch::with_no_grad({torch::nnf_mse_loss(model(xv_t),yv_t)});v<-as.numeric(vl$to(device=torch::torch_device("cpu")));if(is.finite(v)&&v<best-MIN_DELTA){best<-v;best_epoch<-epoch;best_state<-clone_state(model);stale<-0L}else stale<-stale+1L;if(stale>=PATIENCE){stopped<-TRUE;break}}
  sync_device();secs<-proc.time()[["elapsed"]]-t0;if(is.null(best_state)||is.na(best_epoch))stop("No finite final epoch-selection solution.")
  model$load_state_dict(best_state);model$eval();pv_s<-torch::with_no_grad({as.numeric(model(xv_t)$to(device=torch::torch_device("cpu")))});pv<-unscale_y(pv_s,ys);ov<-phenotype[vi,scenario_index]
  list(best_epoch=as.integer(best_epoch),last_epoch=as.integer(last),stopped_by_patience=stopped,hit_epoch_ceiling=last>=FINAL_EPOCH_MAX_EPOCHS&&!stopped,validation_rmse=safe_rmse(pv,ov),validation_correlation=safe_cor(pv,ov),fit_seconds=secs,n_fit=length(iv$train),n_validation=length(iv$validation))
}

train_final_fixed <- function(train_ids,assessment_ids,scenario_index,architecture,learning_rate,l1,l2,epochs,seed){
  tr<-match(train_ids,ids);te<-match(assessment_ids,ids);if(anyNA(c(tr,te)))stop("Final ANN ID alignment failed.")
  xs<-fit_x_scaler(geno[tr,,drop=FALSE]);ys<-fit_y_scaler(phenotype[tr,scenario_index]);xtr<-scale_x(geno[tr,,drop=FALSE],xs);xte<-scale_x(geno[te,,drop=FALSE],xs);ytr<-scale_y(phenotype[tr,scenario_index],ys)
  set_ann_seed(seed);xt<-torch::torch_tensor(xtr,dtype=torch::torch_float())$to(device=ANN_DEVICE);yt<-torch::torch_tensor(matrix(ytr,ncol=1),dtype=torch::torch_float())$to(device=ANN_DEVICE);xe<-torch::torch_tensor(xte,dtype=torch::torch_float())$to(device=ANN_DEVICE)
  model<-TorchGenomicMLP(ncol(xtr),ARCHITECTURES[[architecture]])$to(device=ANN_DEVICE);opt<-torch::optim_adam(params=model$parameters,lr=learning_rate,weight_decay=0);sync_device();t0<-proc.time()[["elapsed"]]
  for(epoch in seq_len(as.integer(epochs))){model$train();opt$zero_grad();pr<-model(xt);loss<-torch::nnf_mse_loss(pr,yt)+weight_penalty(model,l1,l2);loss$backward();opt$step()}
  sync_device();fit_seconds<-proc.time()[["elapsed"]]-t0;model$eval();sync_device();t1<-proc.time()[["elapsed"]];ps<-torch::with_no_grad({as.numeric(model(xe)$to(device=torch::torch_device("cpu")))});sync_device();pred_seconds<-proc.time()[["elapsed"]]-t1
  list(prediction=unscale_y(ps,ys),fit_seconds=fit_seconds,prediction_seconds=pred_seconds,zero_variance_markers=sum(xs$zero))
}
```

# Final outer-CV tasks

```{r final-ann-tasks}
om<-unique(cv$outer_assignments[,c("Repetition","OuterFold","SplitID")]);om<-om[order(om$Repetition,om$OuterFold),]
final_task_table<-merge(data.frame(Scenario=scenario_names),om,by=NULL);final_task_table<-final_task_table[order(match(final_task_table$Scenario,scenario_names),final_task_table$Repetition,final_task_table$OuterFold),];final_task_table$Fold<-final_task_table$OuterFold;final_task_table$TaskID<-paste(final_task_table$Scenario,final_task_table$SplitID,sep="_");rownames(final_task_table)<-NULL
if(nrow(final_task_table)!=150L)stop("Expected 150 canonical ANN outer tasks.")
if(ANN_PILOT){if(!ANN_PILOT_TASK%in%final_task_table$TaskID)stop("Unknown ANN_PILOT_TASK: ",ANN_PILOT_TASK);final_task_table<-final_task_table[final_task_table$TaskID==ANN_PILOT_TASK,,drop=FALSE]}
knitr::kable(head(final_task_table,15),caption=if(ANN_PILOT)"Canonical ANN final pilot task"else"First 15 canonical ANN final tasks")
```

# Final fitting and outer prediction

```{r run-final-ann, eval=RUN_ANN}
req<-c(project_path("output","simulated","ann","tuning","simulated_ann_tuning_selected_configurations.rds"),project_path("output","simulated","ann","tuning","simulated_ann_tuning_run_metadata.rds"),project_path("output","simulated","ann","tuning","simulated_ann_tuning_manifest.rds"));if(!all(file.exists(req)))stop("Canonical tuning outputs are missing. Run RUN_ANN_TUNING=1 first. Missing: ",paste(req[!file.exists(req)],collapse=", "))
sel<-readRDS(req[1]);tm<-readRDS(req[2]);if(!identical(tm$config_id,ANN_TUNING_CONFIG_ID))stop("Saved tuning config_id is incompatible with canonical ANN page.");if(!identical(tm$outer_assessment_used,FALSE)||!identical(tm$true_genetic_value_used,FALSE))stop("Saved tuning provenance failed.");if(nrow(sel)!=600L)stop("Expected 600 canonical selected configurations.")
true_g<-common$true_genetic_value;control<-common$scenario_control;if(!identical(dim(true_g),c(1000L,6L)))stop("True G must be 1000 x 6.")
run_start<-proc.time()[["elapsed"]];outs<-vector("list",nrow(final_task_table))
for(ti in seq_len(nrow(final_task_table))){
  task<-final_task_table[ti,,drop=FALSE];sj<-match(task$Scenario,scenario_names);outer<-cv$outer_splits[[task$SplitID]];if(length(outer$analysis_ids)!=800L||length(outer$assessment_ids)!=200L||length(intersect(outer$analysis_ids,outer$assessment_ids)))stop("Outer split integrity failure: ",task$TaskID)
  st<-sel[sel$TaskID==task$TaskID,,drop=FALSE];if(nrow(st)!=4L||!setequal(st$Model,MODEL_LEVELS))stop("Expected one config/model: ",task$TaskID)
  cf<-file.path(FINAL_CHECKPOINT_DIR,paste0(task$TaskID,".rds"));if(file.exists(cf)&&!ANN_FORCE){cp<-readRDS(cf);if(identical(cp$Status,"ok")&&identical(cp$ConfigID,FINAL_CONFIG_ID)&&identical(cp$TuningConfigID,ANN_TUNING_CONFIG_ID)){outs[[ti]]<-cp;cat(sprintf("[%03d/%03d] %s | final checkpoint reused\n",ti,nrow(final_task_table),task$TaskID));next}}
  t0<-proc.time()[["elapsed"]];base<-SEED_GLOBAL+sj*10000000L+task$Repetition*100000L+task$OuterFold*10000L;rr<-pp<-ee<-vector("list",4L);split_seed<-base+8000000L;epoch_seed<-base+8100000L;fit_seed<-base+8200000L
  for(mi in seq_along(MODEL_LEVELS)){
    ml<-MODEL_LEVELS[mi];cfg<-st[st$Model==ml,,drop=FALSE];ef<-select_final_epoch(outer$analysis_ids,sj,cfg$Architecture,cfg$Learning_rate,cfg$L1,cfg$L2,split_seed,epoch_seed);ff<-train_final_fixed(outer$analysis_ids,outer$assessment_ids,sj,cfg$Architecture,cfg$Learning_rate,cfg$L1,cfg$L2,ef$best_epoch,fit_seed)
    ai<-match(outer$assessment_ids,ids);obs<-phenotype[ai,sj];gt<-true_g[ai,sj];pred<-ff$prediction;mg<-final_metrics(pred,gt,"Genetic");mp<-final_metrics(pred,obs,"Phenotype");meta<-control[control$Scenario==task$Scenario,,drop=FALSE]
    rr[[mi]]<-data.frame(Dataset="simulated",Scenario=task$Scenario,QTL_N=meta$QTL_N,H2_nominal=meta$H2_nominal,Repetition=task$Repetition,Fold=task$OuterFold,SplitID=task$SplitID,TaskID=task$TaskID,Model=ml,Architecture=cfg$Architecture,Learning_rate=cfg$Learning_rate,L1=cfg$L1,L2=cfg$L2,Best_epoch=ef$best_epoch,Epoch_Last=ef$last_epoch,Epoch_Stopped_by_patience=ef$stopped_by_patience,Epoch_Hit_ceiling=ef$hit_epoch_ceiling,Inner_RMSE=cfg$Inner_RMSE,Inner_Correlation=cfg$Inner_Correlation,Inner_MAE=cfg$Inner_MAE,N_train=800L,N_test=200L,N_pred_finite=sum(is.finite(pred)),N_zero_variance_markers=ff$zero_variance_markers,Genetic_Correlation=mg["Genetic_Correlation"],Genetic_RMSE=mg["Genetic_RMSE"],Genetic_MAE=mg["Genetic_MAE"],Genetic_Bias=mg["Genetic_Bias"],Genetic_Slope=mg["Genetic_Slope"],Genetic_R2=mg["Genetic_R2"],Phenotype_Correlation=mp["Phenotype_Correlation"],Phenotype_RMSE=mp["Phenotype_RMSE"],Phenotype_MAE=mp["Phenotype_MAE"],Phenotype_Bias=mp["Phenotype_Bias"],Phenotype_Slope=mp["Phenotype_Slope"],Phenotype_R2=mp["Phenotype_R2"],Epoch_selection_seconds=ef$fit_seconds,Final_fit_seconds=ff$fit_seconds,Predict_seconds=ff$prediction_seconds,Status=if(all(is.finite(pred)))"ok"else"prediction_nonfinite",Error=NA_character_)
    pp[[mi]]<-data.frame(Dataset="simulated",Scenario=task$Scenario,QTL_N=meta$QTL_N,H2_nominal=meta$H2_nominal,Repetition=task$Repetition,Fold=task$OuterFold,SplitID=task$SplitID,TaskID=task$TaskID,Model=ml,Architecture=cfg$Architecture,Learning_rate=cfg$Learning_rate,L1=cfg$L1,L2=cfg$L2,Best_epoch=ef$best_epoch,ID=ids[ai],Observed_Phenotype=obs,True_Genetic_Value=gt,Predicted_Value=pred,Status="ok")
    ee[[mi]]<-data.frame(Scenario=task$Scenario,Repetition=task$Repetition,Fold=task$OuterFold,SplitID=task$SplitID,TaskID=task$TaskID,Model=ml,Architecture=cfg$Architecture,Learning_rate=cfg$Learning_rate,L1=cfg$L1,L2=cfg$L2,Best_epoch=ef$best_epoch,Last_epoch=ef$last_epoch,Stopped_by_patience=ef$stopped_by_patience,Hit_epoch_ceiling=ef$hit_epoch_ceiling,Validation_RMSE=ef$validation_rmse,Validation_Correlation=ef$validation_correlation,N_epoch_fit=ef$n_fit,N_epoch_validation=ef$n_validation,Fit_seconds=ef$fit_seconds)
  }
  cp<-list(Status="ok",ConfigID=FINAL_CONFIG_ID,TuningConfigID=ANN_TUNING_CONFIG_ID,TaskID=task$TaskID,results=do.call(rbind,rr),predictions=do.call(rbind,pp),epoch_selection=do.call(rbind,ee),task_seconds=proc.time()[["elapsed"]]-t0);saveRDS(cp,cf,compress="gzip");outs[[ti]]<-cp;cat(sprintf("[%03d/%03d] %s | final %.1f min\n",ti,nrow(final_task_table),task$TaskID,cp$task_seconds/60));gc(verbose=FALSE)
}
results<-do.call(rbind,lapply(outs,`[[`,"results"));predictions<-do.call(rbind,lapply(outs,`[[`,"predictions"));epoch_selection<-do.call(rbind,lapply(outs,`[[`,"epoch_selection"));rownames(results)<-rownames(predictions)<-rownames(epoch_selection)<-NULL
run_seconds<-proc.time()[["elapsed"]]-run_start;checkpoint_seconds<-sum(vapply(outs,function(z)as.numeric(z$task_seconds),numeric(1)),na.rm=TRUE)
final_meta<-list(dataset="simulated",analysis="canonical_simulated_ann",mode=if(ANN_PILOT)"pilot"else"full",pilot_task=if(ANN_PILOT)ANN_PILOT_TASK else NA_character_,config_id=FINAL_CONFIG_ID,tuning_config_id=ANN_TUNING_CONFIG_ID,seed_global=SEED_GLOBAL,models=MODEL_LEVELS,tuning_max_epochs=MAX_EPOCHS,final_epoch_max_epochs=FINAL_EPOCH_MAX_EPOCHS,patience=PATIENCE,min_delta=MIN_DELTA,early_validation_fraction=EARLY_VALIDATION_FRACTION,final_refit_rule="Best_epoch selected inside outer-analysis, then refit on all 800",outer_assessment_used_in_tuning=FALSE,outer_assessment_used_in_epoch_selection=FALSE,true_genetic_value_used_in_selection=FALSE,number_tasks=nrow(final_task_table),model_task_rows=nrow(results),prediction_rows=nrow(predictions),epoch_selection_rows=nrow(epoch_selection),torch_version=as.character(packageVersion("torch")),device=ANN_DEVICE_LABEL,cuda_available=cuda_available,current_session_seconds=run_seconds,checkpoint_task_seconds=checkpoint_seconds,timestamp=as.character(Sys.time()))
saveRDS(results,FINAL_RESULTS_RDS,compress="gzip");saveRDS(predictions,FINAL_PREDICTIONS_RDS,compress="gzip");saveRDS(epoch_selection,FINAL_EPOCH_RDS,compress="gzip");saveRDS(final_meta,FINAL_RUN_RDS,compress="gzip");write.csv(results,FINAL_RESULTS_CSV,row.names=FALSE);write.csv(predictions,FINAL_PREDICTIONS_CSV,row.names=FALSE);write.csv(epoch_selection,FINAL_EPOCH_CSV,row.names=FALSE)
```

# Saved final-result integrity

```{r final-ann-integrity}
final_available<-all(file.exists(c(FINAL_RESULTS_RDS,FINAL_PREDICTIONS_RDS,FINAL_EPOCH_RDS,FINAL_RUN_RDS)))
if(!final_available){cat("Canonical final ANN outputs are not available yet. Run with RUN_ANN=1 after tuning is available.\n")}else{
  results<-readRDS(FINAL_RESULTS_RDS);predictions<-readRDS(FINAL_PREDICTIONS_RDS);epoch_selection<-readRDS(FINAL_EPOCH_RDS);final_meta<-readRDS(FINAL_RUN_RDS);nt<-if(identical(final_meta$mode,"pilot"))1L else 150L;nr<-nt*4L;np<-nr*200L
  audit<-data.frame(Check=c("Final config ID","Tuning config ID","Model-task rows","Prediction rows","Epoch-selection rows","4 models/task","200 predictions/model-task","All model tasks ok","All predictions finite","Best_epoch in range","Outer excluded from tuning","Outer excluded from epoch selection","True G excluded from selection"),Status=c(if(identical(final_meta$config_id,FINAL_CONFIG_ID))"PASS"else"BLOCKED",if(identical(final_meta$tuning_config_id,ANN_TUNING_CONFIG_ID))"PASS"else"BLOCKED",if(nrow(results)==nr)"PASS"else"BLOCKED",if(nrow(predictions)==np)"PASS"else"BLOCKED",if(nrow(epoch_selection)==nr)"PASS"else"BLOCKED",if(all(table(results$TaskID)==4L))"PASS"else"BLOCKED",if(all(table(interaction(predictions$TaskID,predictions$Model,drop=TRUE))==200L))"PASS"else"BLOCKED",if(all(results$Status=="ok"))"PASS"else"BLOCKED",if(all(is.finite(predictions$Predicted_Value)))"PASS"else"BLOCKED",if(all(epoch_selection$Best_epoch>=1L&epoch_selection$Best_epoch<=FINAL_EPOCH_MAX_EPOCHS))"PASS"else"BLOCKED",if(identical(final_meta$outer_assessment_used_in_tuning,FALSE))"PASS"else"BLOCKED",if(identical(final_meta$outer_assessment_used_in_epoch_selection,FALSE))"PASS"else"BLOCKED",if(identical(final_meta$true_genetic_value_used_in_selection,FALSE))"PASS"else"BLOCKED"))
  knitr::kable(audit,caption="Canonical final ANN saved-result integrity");if(any(audit$Status=="BLOCKED"))stop("Canonical final ANN failed integrity audit.");cat("\nFinal Best_epoch summary:\n");print(summary(epoch_selection$Best_epoch));cat("\nFinal epoch-selection fits hitting",FINAL_EPOCH_MAX_EPOCHS,"ceiling:",sum(epoch_selection$Hit_epoch_ceiling,na.rm=TRUE),"/",nrow(epoch_selection),"\n")
}
```

# Final ANN predictive performance

```{r final-ann-performance}
if(final_available){ok<-results[results$Status=="ok",,drop=FALSE];perf<-aggregate(cbind(Genetic_Correlation,Genetic_RMSE,Genetic_MAE,Genetic_Bias,Genetic_Slope,Genetic_R2,Phenotype_Correlation,Phenotype_RMSE,Phenotype_MAE,Phenotype_Bias,Phenotype_Slope,Phenotype_R2)~Scenario+Model+QTL_N+H2_nominal,data=ok,FUN=mean,na.rm=TRUE);perf<-perf[order(match(perf$Scenario,scenario_names),match(perf$Model,MODEL_LEVELS)),];knitr::kable(perf,digits=4,caption="Mean outer-CV ANN predictive performance")}
```

# Reproducibility information

```{r session-info}
sessionInfo()
```
