---
title: "Canonical comparison of GBLUP-ADE and regularized ANN models"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

# Purpose

This module provides the terminal comparison between GBLUP-ADE and the four ANN variants under the frozen outer-CV design. All competing methods are evaluated on identical outer-assessment individuals. Prediction of the true simulated genetic value is the primary outcome, with phenotype prediction retained as a secondary outcome.

Results are summarized overall, by simulation scenario, and through task-matched ANN versus GBLUP-ADE contrasts.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)

options(stringsAsFactors = FALSE)

find_project_root <- function(start = getwd()) {
  path <- normalizePath(start, winslash = "/", mustWork = TRUE)

  repeat {
    if (file.exists(file.path(path, "_workflowr.yml"))) {
      return(path)
    }

    parent <- dirname(path)

    if (identical(parent, path)) {
      stop("Could not locate workflowr project root.")
    }

    path <- parent
  }
}

PROJECT_ROOT <- find_project_root()
project_path <- function(...) file.path(PROJECT_ROOT, ...)

GBLUP_RESULTS_RDS <- project_path(
  "output", "simulated", "gblup", "simulated_gblup_ade_results.rds"
)
GBLUP_PREDICTIONS_RDS <- project_path(
  "output", "simulated", "gblup", "simulated_gblup_ade_predictions.rds"
)
ANN_RESULTS_RDS <- project_path(
  "output", "simulated", "ann", "final", "simulated_ann_results.rds"
)
ANN_PREDICTIONS_RDS <- project_path(
  "output", "simulated", "ann", "final", "simulated_ann_predictions.rds"
)

COMPARISON_DIR <- project_path("output", "simulated", "comparison")
OVERALL_CSV <- file.path(
  COMPARISON_DIR, "simulated_model_overall_summary.csv"
)
SCENARIO_CSV <- file.path(
  COMPARISON_DIR, "simulated_model_scenario_summary.csv"
)
PAIRED_CSV <- file.path(
  COMPARISON_DIR, "simulated_ann_vs_gblup_paired_summary.csv"
)
PAIRED_SCENARIO_CSV <- file.path(
  COMPARISON_DIR, "simulated_ann_vs_gblup_paired_by_scenario.csv"
)

WRITE_COMPARISON_OUTPUTS <- identical(
  Sys.getenv("WRITE_COMPARISON_OUTPUTS", "0"),
  "1"
)

metrics <- c(
  "Genetic_Correlation",
  "Genetic_RMSE",
  "Genetic_MAE",
  "Genetic_Bias",
  "Genetic_Slope",
  "Genetic_R2",
  "Phenotype_Correlation",
  "Phenotype_RMSE",
  "Phenotype_MAE",
  "Phenotype_Bias",
  "Phenotype_Slope",
  "Phenotype_R2"
)

higher_better <- c(
  "Genetic_Correlation",
  "Genetic_R2",
  "Phenotype_Correlation",
  "Phenotype_R2"
)
lower_better <- c(
  "Genetic_RMSE",
  "Genetic_MAE",
  "Phenotype_RMSE",
  "Phenotype_MAE"
)
zero_target <- c("Genetic_Bias", "Phenotype_Bias")
one_target <- c("Genetic_Slope", "Phenotype_Slope")

expected_ann_models <- c("ANN_0", "ANN_L1", "ANN_L2", "ANN_L1_L2")
```

# Canonical inputs

```{r load-canonical-inputs, include=FALSE}
required_files <- c(
  GBLUP_RESULTS_RDS,
  GBLUP_PREDICTIONS_RDS,
  ANN_RESULTS_RDS,
  ANN_PREDICTIONS_RDS
)

if (!all(file.exists(required_files))) {
  stop(
    "Canonical model outputs are missing: ",
    paste(required_files[!file.exists(required_files)], collapse = ", ")
  )
}

gr <- readRDS(GBLUP_RESULTS_RDS)
gp <- readRDS(GBLUP_PREDICTIONS_RDS)
ar <- readRDS(ANN_RESULTS_RDS)
ap <- readRDS(ANN_PREDICTIONS_RDS)

cat("GBLUP result rows:", nrow(gr), "\n")
cat("GBLUP prediction rows:", nrow(gp), "\n")
cat("ANN result rows:", nrow(ar), "\n")
cat("ANN prediction rows:", nrow(ap), "\n")
cat("ANN models:", paste(sort(unique(ar$Model)), collapse = ", "), "\n")
```

# Compatibility audit

The result-level audit verifies the frozen task design. The prediction-level
audit independently verifies that the actual held-out individuals, observed
phenotypes, and true simulated genetic values are identical across methods.

```{r compatibility-audit, echo=FALSE}
required_result_cols <- c(
  "Scenario", "Repetition", "Fold", "SplitID", "TaskID",
  "Model", "Status", metrics
)

required_prediction_cols <- c(
  "Scenario", "Repetition", "Fold", "SplitID", "TaskID", "Model", "ID",
  "Observed_Phenotype", "True_Genetic_Value", "Predicted_Value", "Status"
)

missing_gr <- setdiff(required_result_cols, names(gr))
missing_ar <- setdiff(required_result_cols, names(ar))
missing_gp <- setdiff(required_prediction_cols, names(gp))
missing_ap <- setdiff(required_prediction_cols, names(ap))

if (length(missing_gr)) {
  stop("Missing GBLUP result columns: ", paste(missing_gr, collapse = ", "))
}
if (length(missing_ar)) {
  stop("Missing ANN result columns: ", paste(missing_ar, collapse = ", "))
}
if (length(missing_gp)) {
  stop("Missing GBLUP prediction columns: ", paste(missing_gp, collapse = ", "))
}
if (length(missing_ap)) {
  stop("Missing ANN prediction columns: ", paste(missing_ap, collapse = ", "))
}

result_key <- function(x) {
  paste(
    x$Scenario, x$Repetition, x$Fold, x$SplitID, x$TaskID,
    sep = "|"
  )
}

assessment_key <- function(x) {
  paste(
    x$Scenario, x$Repetition, x$Fold, x$SplitID, x$TaskID,
    as.character(x$ID),
    sep = "|"
  )
}

g_result_keys <- sort(result_key(gr))
g_assessment_keys <- sort(assessment_key(gp))
gp_ordered <- gp[order(assessment_key(gp)), , drop = FALSE]

ann_result_keys_match <- vapply(
  expected_ann_models,
  function(model) {
    z <- ar[ar$Model == model, , drop = FALSE]
    nrow(z) == 150L &&
      !anyDuplicated(z$TaskID) &&
      identical(sort(result_key(z)), g_result_keys)
  },
  logical(1)
)

ann_assessment_keys_match <- vapply(
  expected_ann_models,
  function(model) {
    z <- ap[ap$Model == model, , drop = FALSE]
    nrow(z) == 30000L &&
      !anyDuplicated(z[, c("TaskID", "ID")]) &&
      identical(sort(assessment_key(z)), g_assessment_keys)
  },
  logical(1)
)

ann_phenotype_match <- vapply(
  expected_ann_models,
  function(model) {
    z <- ap[ap$Model == model, , drop = FALSE]
    z <- z[order(assessment_key(z)), , drop = FALSE]
    isTRUE(all.equal(
      z$Observed_Phenotype,
      gp_ordered$Observed_Phenotype,
      tolerance = 1e-12,
      check.attributes = FALSE
    ))
  },
  logical(1)
)

ann_true_g_match <- vapply(
  expected_ann_models,
  function(model) {
    z <- ap[ap$Model == model, , drop = FALSE]
    z <- z[order(assessment_key(z)), , drop = FALSE]
    isTRUE(all.equal(
      z$True_Genetic_Value,
      gp_ordered$True_Genetic_Value,
      tolerance = 1e-12,
      check.attributes = FALSE
    ))
  },
  logical(1)
)

compatibility_checks <- data.frame(
  Check = c(
    "GBLUP has 150 result rows",
    "ANN has 600 result rows",
    "expected ANN models are present",
    "GBLUP TaskID values are unique",
    "each ANN model has identical result-level task keys",
    "GBLUP has 30000 held-out prediction rows",
    "ANN has 120000 held-out prediction rows",
    "GBLUP has exactly 200 predictions per task",
    "ANN has exactly 200 predictions per task and model",
    "GBLUP TaskID/ID pairs are unique",
    "each ANN model has identical outer-assessment individuals",
    "observed phenotypes are identical across paired assessments",
    "true genetic values are identical across paired assessments",
    "all GBLUP result statuses are ok",
    "all ANN result statuses are ok",
    "all GBLUP prediction statuses are ok",
    "all ANN prediction statuses are ok",
    "all GBLUP comparison metrics are finite",
    "all ANN comparison metrics are finite",
    "all GBLUP held-out predictions are finite",
    "all ANN held-out predictions are finite"
  ),
  PASS = c(
    nrow(gr) == 150L,
    nrow(ar) == 600L,
    identical(sort(unique(ar$Model)), sort(expected_ann_models)),
    !anyDuplicated(gr$TaskID),
    all(ann_result_keys_match),
    nrow(gp) == 30000L,
    nrow(ap) == 120000L,
    all(table(gp$TaskID) == 200L),
    all(table(interaction(ap$TaskID, ap$Model, drop = TRUE)) == 200L),
    !anyDuplicated(gp[, c("TaskID", "ID")]),
    all(ann_assessment_keys_match),
    all(ann_phenotype_match),
    all(ann_true_g_match),
    all(gr$Status == "ok"),
    all(ar$Status == "ok"),
    all(gp$Status == "ok"),
    all(ap$Status == "ok"),
    all(vapply(
      gr[, metrics, drop = FALSE],
      function(x) all(is.finite(x)),
      logical(1)
    )),
    all(vapply(
      ar[, metrics, drop = FALSE],
      function(x) all(is.finite(x)),
      logical(1)
    )),
    all(is.finite(gp$Predicted_Value)),
    all(is.finite(ap$Predicted_Value))
  ),
  stringsAsFactors = FALSE
)

compatibility_checks$Status <- ifelse(
  compatibility_checks$PASS,
  "PASS",
  "BLOCKED"
)

knitr::kable(
  compatibility_checks[, c("Check", "Status")],
  caption = "Compatibility audit for the canonical model comparison."
)

if (any(!compatibility_checks$PASS)) {
  stop("Canonical model-comparison compatibility audit failed.")
}
```

# Overall model performance

```{r overall-comparison, echo=FALSE}
analysis_cols <- c(
  "Scenario", "Repetition", "Fold", "SplitID", "TaskID", "Model", metrics
)

all_results <- rbind(gr[, analysis_cols], ar[, analysis_cols])

overall <- aggregate(
  all_results[, metrics],
  by = list(Model = all_results$Model),
  FUN = mean
)

overall_display <- overall[, c(
  "Model",
  "Genetic_Correlation",
  "Genetic_RMSE",
  "Genetic_MAE",
  "Genetic_R2",
  "Phenotype_Correlation",
  "Phenotype_RMSE",
  "Phenotype_MAE",
  "Phenotype_R2"
)]
overall_display[-1] <- lapply(overall_display[-1], round, 4)

knitr::kable(
  overall_display,
  caption = paste(
    "Overall held-out predictive performance.",
    "Genetic-value prediction is the primary outcome."
  )
)
```

```{r overall-interpretation, echo=FALSE, results="asis"}
best_corr_model <- overall$Model[which.max(overall$Genetic_Correlation)]
best_rmse_model <- overall$Model[which.min(overall$Genetic_RMSE)]
best_r2_model <- overall$Model[which.max(overall$Genetic_R2)]

if (!identical(best_corr_model, best_rmse_model) ||
    !identical(best_corr_model, best_r2_model)) {
  stop("Primary overall metrics do not identify the same best model.")
}

best_overall <- overall[
  overall$Model == best_corr_model,
  ,
  drop = FALSE
]

ann_overall_now <- overall[
  overall$Model != "GBLUP_ADE",
  ,
  drop = FALSE
]

best_ann_now <- ann_overall_now[
  which.max(ann_overall_now$Genetic_Correlation),
  ,
  drop = FALSE
]

cat(sprintf(
  paste0(
    "Across all 150 outer-CV tasks, %s has the highest mean genetic ",
    "correlation (%.4f), the lowest mean genetic RMSE (%.4f), and the ",
    "highest mean genetic R2 (%.4f). Among the ANN variants, %s has the ",
    "highest mean genetic correlation (%.4f). ANN_L1 is nearly tied with ",
    "ANN_L1_L2, whereas ANN_L2 remains close to ANN_0, indicating that ",
    "the observed regularization gain is driven mainly by the L1 component ",
    "within the prespecified canonical model/architecture search space.\n"
  ),
  best_overall$Model,
  best_overall$Genetic_Correlation,
  best_overall$Genetic_RMSE,
  best_overall$Genetic_R2,
  best_ann_now$Model,
  best_ann_now$Genetic_Correlation
))
```

# Performance by simulation scenario

```{r scenario-comparison, echo=FALSE}
scenario_summary <- aggregate(
  all_results[, metrics],
  by = list(
    Scenario = all_results$Scenario,
    Model = all_results$Model
  ),
  FUN = mean
)

scenario_display <- scenario_summary[, c(
  "Scenario", "Model", "Genetic_Correlation", "Genetic_RMSE", "Genetic_R2"
)]
scenario_display[, -(1:2)] <- lapply(
  scenario_display[, -(1:2), drop = FALSE],
  round,
  4
)

knitr::kable(
  scenario_display,
  caption = "Primary genetic-value prediction metrics by scenario."
)
```

```{r scenario-interpretation, echo=FALSE, results="asis"}
scenario_ids <- sort(unique(scenario_summary$Scenario))

corr_wins <- sum(vapply(
  scenario_ids,
  function(sc) {
    z <- scenario_summary[
      scenario_summary$Scenario == sc,
      ,
      drop = FALSE
    ]
    z$Model[which.max(z$Genetic_Correlation)] == "GBLUP_ADE"
  },
  logical(1)
))

rmse_wins <- sum(vapply(
  scenario_ids,
  function(sc) {
    z <- scenario_summary[
      scenario_summary$Scenario == sc,
      ,
      drop = FALSE
    ]
    z$Model[which.min(z$Genetic_RMSE)] == "GBLUP_ADE"
  },
  logical(1)
))

r2_wins <- sum(vapply(
  scenario_ids,
  function(sc) {
    z <- scenario_summary[
      scenario_summary$Scenario == sc,
      ,
      drop = FALSE
    ]
    z$Model[which.max(z$Genetic_R2)] == "GBLUP_ADE"
  },
  logical(1)
))

cat(sprintf(
  paste0(
    "GBLUP-ADE has the best mean genetic correlation in %d of %d scenarios, ",
    "the lowest mean genetic RMSE in %d of %d scenarios, and the highest ",
    "mean genetic R2 in %d of %d scenarios. The C1--C6 sequence should be ",
    "interpreted as performance across the prespecified simulation gradient ",
    "rather than as an isolated causal effect of QTL number or nominal ",
    "heritability, because those scenario characteristics change jointly ",
    "across the sequence.\n"
  ),
  corr_wins, length(scenario_ids),
  rmse_wins, length(scenario_ids),
  r2_wins, length(scenario_ids)
))
```

# Paired ANN versus GBLUP-ADE comparison

A positive `performance_advantage` always means that the ANN variant performed
better than GBLUP-ADE for the metric under consideration. Correlations and
\(R^2\) are maximized, RMSE and MAE are minimized, bias is evaluated by distance
from zero, and slope is evaluated by distance from one.

```{r paired-comparison, echo=FALSE}
performance_advantage <- function(metric, ann_value, gblup_value) {
  if (metric %in% higher_better) {
    return(ann_value - gblup_value)
  }
  if (metric %in% lower_better) {
    return(gblup_value - ann_value)
  }
  if (metric %in% zero_target) {
    return(abs(gblup_value) - abs(ann_value))
  }
  if (metric %in% one_target) {
    return(abs(gblup_value - 1) - abs(ann_value - 1))
  }

  stop("Unclassified metric: ", metric)
}

g <- gr[, c("TaskID", "SplitID", metrics)]
names(g)[-(1:2)] <- paste0(names(g)[-(1:2)], "_GBLUP")

paired_list <- list()

for (model in sort(unique(ar$Model))) {
  a <- ar[
    ar$Model == model,
    c("TaskID", "SplitID", "Scenario", metrics)
  ]

  z <- merge(a, g, by = c("TaskID", "SplitID"), all = FALSE)

  if (nrow(z) != 150L) {
    stop(model, ": expected 150 paired tasks, got ", nrow(z))
  }

  for (metric in metrics) {
    ann_value <- z[[metric]]
    gblup_value <- z[[paste0(metric, "_GBLUP")]]
    raw_difference <- ann_value - gblup_value
    advantage <- performance_advantage(metric, ann_value, gblup_value)

    paired_list[[length(paired_list) + 1L]] <- data.frame(
      Model = model,
      Metric = metric,
      N = length(raw_difference),
      Mean_ANN = mean(ann_value),
      Mean_GBLUP = mean(gblup_value),
      Mean_raw_difference = mean(raw_difference),
      Median_raw_difference = median(raw_difference),
      Mean_performance_advantage = mean(advantage),
      Median_performance_advantage = median(advantage),
      ANN_better = sum(advantage > 0),
      Equal = sum(advantage == 0),
      GBLUP_better = sum(advantage < 0),
      stringsAsFactors = FALSE
    )
  }
}

paired <- do.call(rbind, paired_list)
rownames(paired) <- NULL

paired_primary <- paired[
  paired$Metric %in% c("Genetic_Correlation", "Genetic_RMSE", "Genetic_R2"),
  c(
    "Model", "Metric", "Mean_ANN", "Mean_GBLUP",
    "Mean_performance_advantage", "ANN_better", "Equal", "GBLUP_better"
  )
]
paired_primary[, 3:5] <- lapply(
  paired_primary[, 3:5, drop = FALSE],
  round,
  4
)

knitr::kable(
  paired_primary,
  caption = paste(
    "Paired ANN versus GBLUP-ADE comparison for the primary",
    "genetic-value metrics."
  )
)
```

The `ANN_better` and `GBLUP_better` counts refer to the same frozen
outer-assessment tasks for both methods. They therefore describe within-task
predictive wins under directly matched test samples. These counts are
descriptive summaries of repeated cross-validation and are not treated as
independent replicates for a naive fold-level significance test.

# Paired comparison by scenario

```{r paired-by-scenario, echo=FALSE}
paired_scenario_list <- list()

for (model in sort(unique(ar$Model))) {
  a <- ar[
    ar$Model == model,
    c("TaskID", "SplitID", "Scenario", metrics)
  ]

  z <- merge(a, g, by = c("TaskID", "SplitID"), all = FALSE)

  for (scenario in sort(unique(z$Scenario))) {
    zs <- z[z$Scenario == scenario, , drop = FALSE]

    if (nrow(zs) != 25L) {
      stop(
        model, " / ", scenario,
        ": expected 25 paired tasks, got ", nrow(zs)
      )
    }

    for (metric in metrics) {
      ann_value <- zs[[metric]]
      gblup_value <- zs[[paste0(metric, "_GBLUP")]]
      advantage <- performance_advantage(metric, ann_value, gblup_value)

      paired_scenario_list[[length(paired_scenario_list) + 1L]] <- data.frame(
        Scenario = scenario,
        Model = model,
        Metric = metric,
        N = nrow(zs),
        Mean_advantage = mean(advantage),
        Median_advantage = median(advantage),
        ANN_better = sum(advantage > 0),
        Equal = sum(advantage == 0),
        GBLUP_better = sum(advantage < 0),
        stringsAsFactors = FALSE
      )
    }
  }
}

paired_scenario <- do.call(rbind, paired_scenario_list)
rownames(paired_scenario) <- NULL

paired_scenario_primary <- paired_scenario[
  paired_scenario$Metric == "Genetic_Correlation",
  c(
    "Scenario", "Model", "Mean_advantage",
    "ANN_better", "Equal", "GBLUP_better"
  )
]
paired_scenario_primary$Mean_advantage <- round(
  paired_scenario_primary$Mean_advantage,
  4
)

knitr::kable(
  paired_scenario_primary,
  caption = paste(
    "Paired genetic-correlation advantage by scenario.",
    "Positive values favor ANN."
  )
)
```

The scenario-level paired summaries show whether the overall pattern is
concentrated in a subset of simulation conditions or persists across the full
C1--C6 sequence. Positive mean advantage values favor the ANN variant; negative
values favor GBLUP-ADE. Together with the scenario means above, these paired
summaries provide a task-matched description of how consistently the methods
differ.

```{r write-comparison-outputs, eval=WRITE_COMPARISON_OUTPUTS, include=FALSE}
dir.create(COMPARISON_DIR, recursive = TRUE, showWarnings = FALSE)
write.csv(overall, OVERALL_CSV, row.names = FALSE)
write.csv(scenario_summary, SCENARIO_CSV, row.names = FALSE)
write.csv(paired, PAIRED_CSV, row.names = FALSE)
write.csv(paired_scenario, PAIRED_SCENARIO_CSV, row.names = FALSE)
```

```{r verify-comparison-outputs, include=FALSE}
canonical_files <- c(
  OVERALL_CSV,
  SCENARIO_CSV,
  PAIRED_CSV,
  PAIRED_SCENARIO_CSV
)

if (!all(file.exists(canonical_files))) {
  stop(
    "Canonical comparison CSV files are missing: ",
    paste(canonical_files[!file.exists(canonical_files)], collapse = ", ")
  )
}

frames_equal <- function(saved, computed, tolerance = 1e-12) {
  if (!identical(names(saved), names(computed)) || nrow(saved) != nrow(computed)) {
    return(FALSE)
  }

  for (name in names(saved)) {
    x <- saved[[name]]
    y <- computed[[name]]

    if (is.numeric(x) && is.numeric(y)) {
      if (!all(is.finite(x)) || !all(is.finite(y))) {
        return(FALSE)
      }
      if (!all(abs(x - y) <= tolerance)) {
        return(FALSE)
      }
    } else if (!identical(as.character(x), as.character(y))) {
      return(FALSE)
    }
  }

  TRUE
}

saved_overall <- read.csv(
  OVERALL_CSV, stringsAsFactors = FALSE, check.names = FALSE
)
saved_scenario <- read.csv(
  SCENARIO_CSV, stringsAsFactors = FALSE, check.names = FALSE
)
saved_paired <- read.csv(
  PAIRED_CSV, stringsAsFactors = FALSE, check.names = FALSE
)
saved_paired_scenario <- read.csv(
  PAIRED_SCENARIO_CSV, stringsAsFactors = FALSE, check.names = FALSE
)

output_checks <- data.frame(
  Output = basename(canonical_files),
  Status = ifelse(
    c(
      frames_equal(saved_overall, overall),
      frames_equal(saved_scenario, scenario_summary),
      frames_equal(saved_paired, paired),
      frames_equal(saved_paired_scenario, paired_scenario)
    ),
    "PASS",
    "BLOCKED"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  output_checks,
  caption = "Agreement between recomputed and versioned comparison tables."
)

if (any(output_checks$Status != "PASS")) {
  stop(
    "Canonical comparison exports do not match the recomputed results. ",
    "Review model outputs before using WRITE_COMPARISON_OUTPUTS=1."
  )
}
```

# Overall conclusion

Within the prespecified canonical model and ANN architecture search space,
GBLUP-ADE provides the strongest overall predictive performance and leads the
primary genetic-value metrics across all six simulation scenarios. Among the
neural networks, ANN-L1+L2 has the best overall averages by a very small margin
over ANN-L1, while ANN-L2 remains close to the unregularized ANN.

These results support a specific conclusion for the frozen design evaluated
here; they do not establish a universal ranking between genomic mixed models
and all possible neural-network architectures.

# Navigation

[GBLUP-ADE](simulated_data_gblup.html) | [Regularized ANN](simulated_data_ann.html) | [Home](index.html)

# Reproducibility information

```{r session-info}
sessionInfo()
```
