---
title: "Regularized artificial neural networks for simulated genomic data"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

## 1. Objective

This module is the **canonical ANN workflow** for the simulated genomic-data
branch. ANN hyperparameter tuning and final outer-CV fitting are internal
stages of one modeling method, parallel to GBLUP-ADE in the public pipeline.

The four variants are `ANN_0`, `ANN_L1`, `ANN_L2`, and `ANN_L1_L2`
(displayed as **ANN-L1+L2**). They use the same multilayer perceptron and differ
only in the L1 and L2 penalties applied to network weights.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)

options(stringsAsFactors = FALSE)
```

```{r table-formatting, include=FALSE}
format_table <- function(
  x,
  caption = NULL,
  digits = getOption("digits"),
  row.names = NA,
  ...
) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    row.names = row.names,
    table.attr = paste0(
      'class="table table-striped table-hover table-bordered" ',
      'style="width:100%; margin-bottom:0;"'
    ),
    ...
  )

  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

### 1.1 How to read this module

The module separates three stages that must not be conflated:

1. **nested tuning**, which uses only the frozen inner cross-validation folds;
2. **epoch selection**, which uses only the 800 outer-analysis individuals;
3. **final refitting and assessment**, which refits on all 800 outer-analysis
   individuals and predicts the untouched 200 outer-assessment individuals.

True simulated genetic values are not used in tuning or epoch selection. They
are accessed only after held-out predictions have been generated, as the
primary target for evaluating recovery of the simulated genetic signal.

## 2. ANN variants and tuning strategy

### 2.1 Models compared

All four ANN variants use the same feed-forward network definition. Their only
difference is the regularization applied to the network weights.

```{r ann-variants}
ann_variants <- data.frame(
  Model = c("ANN_0", "ANN_L1", "ANN_L2", "ANN_L1_L2"),
  L1 = c("No", "Yes", "No", "Yes"),
  L2 = c("No", "No", "Yes", "Yes"),
  Interpretation = c(
    "Unregularized reference ANN.",
    "ANN with L1 weight regularization.",
    "ANN with L2 weight regularization.",
    "ANN with simultaneous L1 and L2 weight regularization."
  ),
  stringsAsFactors = FALSE
)

format_table(
  ann_variants,
  caption = "ANN variants compared in the canonical workflow"
)
```

### 2.2 Tuned hyperparameters

Architecture and learning rate are selected once per outer task using `ANN_0`
and are then shared by all four variants. After that shared structural choice,
the regularized variants tune only the penalty coefficient or coefficients that
define their model.

```{r tuned-hyperparameters}
tuned_hyperparameters <- data.frame(
  Hyperparameter = c(
    "Architecture",
    "Learning rate",
    "L1 coefficient",
    "L2 coefficient"
  ),
  Role = c(
    "Number and size of hidden layers.",
    "Adam optimization step size.",
    "Strength of absolute-weight regularization.",
    "Strength of squared-weight regularization."
  ),
  Selection = c(
    "Selected with ANN_0 and shared across all variants.",
    "Selected with ANN_0 and shared across all variants.",
    "Selected only for ANN_L1 and ANN_L1_L2.",
    "Selected only for ANN_L2 and ANN_L1_L2."
  ),
  stringsAsFactors = FALSE
)

format_table(
  tuned_hyperparameters,
  caption = "Hyperparameters considered in ANN tuning"
)
```

### 2.3 Fixed optimization controls

ReLU activation, Adam optimization, zero dropout, the tuning epoch ceiling,
patience, minimum improvement, and early-validation fraction are fixed
optimization controls. They determine **how** each candidate is optimized and
are not additional dimensions of the scientific hyperparameter search.

```{r fixed-optimization-controls}
fixed_optimization_controls <- data.frame(
  Control = c(
    "Activation",
    "Optimizer",
    "Dropout",
    "Maximum tuning epochs",
    "Patience",
    "Minimum improvement (min_delta)",
    "Early-validation fraction"
  ),
  Canonical_value = c(
    "ReLU",
    "Adam",
    "0",
    "2000",
    "25",
    "1e-6",
    "0.15"
  ),
  stringsAsFactors = FALSE
)

format_table(
  fixed_optimization_controls,
  caption = "Fixed ANN optimization controls"
)
```

### 2.4 Early stopping

An epoch is one complete pass through the current fitting observations. During
candidate training, loss is monitored on a separate early-validation subset
drawn only from the current analysis sample.

Training stops after 25 consecutive epochs without an improvement of at least
\(10^{-6}\). The 2000-epoch value is therefore a convergence ceiling, not a
fixed training length. Epoch diagnostics are retained for optimization
diagnostics but are excluded from candidate ranking.

### 2.5 Bounded architecture search

The canonical architecture catalog is deliberately bounded to
small-to-moderate feed-forward networks. Its upper boundary is `64-32`.

Selection of a boundary architecture is interpreted as evidence that the
prespecified search space was active for that task. It does not trigger a
post-hoc enlargement of the canonical search space. Consequently, ANN versus
GBLUP-ADE conclusions apply to this prespecified ANN search space, not to every
possible neural-network architecture.

## 3. Packages and computational environment

### 3.1 Packages

`knitr` formats the reader-facing tables. `torch` is initialized only when
tuning or final ANN fitting is explicitly requested; saved-output renders do not
require a new ANN runtime.

```{r packages-base, warning=FALSE, message=FALSE}
library(knitr) # Format reader-facing tables and R Markdown output.
```

```{r legacy-table-formatting, include=FALSE}
# Temporary compatibility helper retained until later M04 table stages.
WIDE_TABLE_ATTR <- paste(
  'style="display:block;',
  'overflow-x:auto;',
  'white-space:nowrap;',
  'width:100%;"'
)
```

### 3.2 Execution mode

ANN tuning and final fitting are opt-in computational stages. With both runtime
flags disabled, the page is rebuilt from the canonical saved outputs.

```{r execution-mode}
PROJECT_ROOT <- "."
project_path <- function(...) file.path(PROJECT_ROOT, ...)

RUN_ANN_TUNING <- identical(
  Sys.getenv("RUN_ANN_TUNING", "0"),
  "1"
)
RUN_ANN <- identical(
  Sys.getenv("RUN_ANN", "0"),
  "1"
)
ANN_RUNTIME_REQUIRED <- RUN_ANN_TUNING || RUN_ANN

# Legacy/auxiliary execution flags are retained unchanged for compatibility.
ANN_TUNING_FORCE <- identical(
  Sys.getenv("ANN_TUNING_FORCE", "0"),
  "1"
)
ANN_TUNING_PILOT <- identical(
  Sys.getenv("ANN_TUNING_PILOT", "0"),
  "1"
)
ANN_TUNING_PILOT_TASK <- Sys.getenv(
  "ANN_TUNING_PILOT_TASK",
  "C1_R01_F01"
)

execution_mode <- data.frame(
  Item = c(
    "Run nested tuning",
    "Run final ANN fitting",
    "ANN runtime required"
  ),
  Value = c(
    RUN_ANN_TUNING,
    RUN_ANN,
    ANN_RUNTIME_REQUIRED
  ),
  stringsAsFactors = FALSE
)

format_table(
  execution_mode,
  caption = "ANN execution mode for the current render"
)
```

The auxiliary `ANN_TUNING_FORCE`, `ANN_TUNING_PILOT`, and
`ANN_TUNING_PILOT_TASK` flags are retained unchanged for compatibility. Their
historical behavior is not redefined in this editorial stage.

### 3.3 CPU/CUDA device

When an ANN runtime is requested, `torch` detects CUDA availability and selects
the requested `ANN_DEVICE` (`cpu`, `cuda`, or automatic selection). With saved
outputs only, the runtime remains uninitialized.

```{r packages-device, warning=FALSE, message=FALSE}
if (ANN_RUNTIME_REQUIRED) {
  suppressPackageStartupMessages(library(torch))
  cuda_available <- torch::cuda_is_available()
  requested_device <- tolower(Sys.getenv("ANN_DEVICE", "auto"))

  ANN_DEVICE_LABEL <- if (requested_device == "cuda") {
    "cuda"
  } else if (requested_device == "cpu") {
    "cpu"
  } else if (cuda_available) {
    "cuda"
  } else {
    "cpu"
  }

  ANN_DEVICE <- torch::torch_device(ANN_DEVICE_LABEL)

  sync_device <- function() {
    if (ANN_DEVICE_LABEL == "cuda") torch::cuda_synchronize()
    invisible(NULL)
  }

  set_ann_seed <- function(seed) {
    set.seed(as.integer(seed))
    torch::torch_manual_seed(as.integer(seed))
    invisible(NULL)
  }

  torch_version_display <- as.character(packageVersion("torch"))
  cuda_display <- cuda_available
  device_display <- ANN_DEVICE_LABEL
} else {
  cuda_available <- NA
  ANN_DEVICE_LABEL <- "not initialized (saved-output render)"
  ANN_DEVICE <- NULL
  sync_device <- function() invisible(NULL)
  set_ann_seed <- function(seed) invisible(seed)
  torch_version_display <- "not initialized (saved-output render)"
  cuda_display <- "not initialized (saved-output render)"
  device_display <- ANN_DEVICE_LABEL
}

environment_table <- data.frame(
  Item = c("torch version", "CUDA available", "device"),
  Value = c(torch_version_display, cuda_display, device_display),
  stringsAsFactors = FALSE
)

format_table(
  environment_table,
  caption = "ANN computational environment"
)
```

## 4. Canonical inputs and outputs

### 4.1 Input files

The ANN branch consumes the audited common simulated-data object and the frozen
nested validation design produced by the preceding modules. The two audit CSV
files are retained as provenance references for those upstream contracts.

```{r ann-input-paths}
DATA_FILE <- project_path(
  "output", "simulated", "audit", "preprocessed_simulated_common.rds"
)
CV_FILE <- project_path(
  "output", "simulated", "validation", "simulated_cv_design.rds"
)
DATA_AUDIT_FILE <- project_path(
  "output", "simulated", "audit", "simulated_audit_summary.csv"
)
CV_AUDIT_FILE <- project_path(
  "output", "simulated", "validation", "simulated_cv_design_audit.csv"
)

input_files <- data.frame(
  File = basename(c(
    DATA_FILE,
    CV_FILE,
    DATA_AUDIT_FILE,
    CV_AUDIT_FILE
  )),
  Role = c(
    "Audited phenotype, genotype, true-G, and scenario objects.",
    "Frozen outer and inner validation partitions.",
    "Data-audit provenance record.",
    "Validation-design provenance record."
  ),
  stringsAsFactors = FALSE
)

format_table(
  input_files,
  caption = "Canonical files available to the ANN branch"
)
```

### 4.2 Canonical data objects

The same 1,000 individuals, 4,010 markers, six scenarios, 25 outer splits, and
100 inner splits established upstream are reused here. Small previews make the
reader-facing inputs explicit before tuning logic is introduced.

```{r inputs}
common <- readRDS(DATA_FILE)
cv <- readRDS(CV_FILE)

phenotype <- common$phenotype
geno <- common$genotype_m101
ids <- seq_len(nrow(geno))
scenario_names <- colnames(phenotype)

ann_input_summary <- data.frame(
  Object = c(
    "Phenotype",
    "Genotype",
    "Outer validation splits",
    "Inner validation splits"
  ),
  Rows_or_count = c(
    nrow(phenotype),
    nrow(geno),
    length(cv$outer_splits),
    sum(vapply(cv$inner_splits, length, integer(1)))
  ),
  Columns = c(
    ncol(phenotype),
    ncol(geno),
    NA,
    NA
  ),
  Role = c(
    "Response used for ANN fitting.",
    "Marker predictors used by all ANN variants.",
    "Frozen final assessment partitions.",
    "Frozen candidate-evaluation partitions."
  ),
  stringsAsFactors = FALSE
)

phenotype_sample <- data.frame(
  ID = ids[1:6],
  phenotype[1:6, , drop = FALSE],
  check.names = FALSE
)
genotype_sample <- data.frame(
  ID = ids[1:6],
  geno[1:6, 1:10, drop = FALSE],
  check.names = FALSE
)

format_table(
  ann_input_summary,
  caption = "Canonical ANN input objects"
)
format_table(
  phenotype_sample,
  digits = 4,
  caption = "First six phenotype rows"
)
format_table(
  genotype_sample,
  caption = "First six individuals and first ten markers"
)
```

The validation objects are reused exactly as frozen in M02. No ANN variant
constructs its own outer or inner cross-validation partitions.

### 4.3 Tuning output files

Nested tuning writes task-level candidate summaries, inner-fold results,
selected configurations, and run metadata. The canonical output stem is
unchanged; the legacy pilot stem is retained when its existing flag is active.

```{r ann-tuning-output-paths}
CHECKPOINT_DIR <- project_path(
  ".local", "checkpoints", "canonical", "ann_tuning"
)

OUTPUT_STEM <- if (ANN_TUNING_PILOT) {
  project_path(
    ".local", "pilot", "ann", "tuning", "simulated_ann_tuning_pilot"
  )
} else {
  project_path(
    "output", "simulated", "ann", "tuning", "simulated_ann_tuning"
  )
}

CANDIDATE_RDS <- paste0(OUTPUT_STEM, "_candidate_summary.rds")
FOLD_RDS <- paste0(OUTPUT_STEM, "_fold_results.rds")
SELECTED_RDS <- paste0(OUTPUT_STEM, "_selected_configurations.rds")
RUN_RDS <- paste0(OUTPUT_STEM, "_run_metadata.rds")
MANIFEST_RDS <- paste0(OUTPUT_STEM, "_manifest.rds")

CANDIDATE_CSV <- paste0(OUTPUT_STEM, "_candidate_summary.csv")
FOLD_CSV <- paste0(OUTPUT_STEM, "_fold_results.csv")
SELECTED_CSV <- paste0(OUTPUT_STEM, "_selected_configurations.csv")
MANIFEST_CSV <- paste0(OUTPUT_STEM, "_manifest.csv")

tuning_output_files <- data.frame(
  File = basename(c(
    CANDIDATE_RDS,
    FOLD_RDS,
    SELECTED_RDS,
    RUN_RDS,
    MANIFEST_RDS,
    CANDIDATE_CSV,
    FOLD_CSV,
    SELECTED_CSV,
    MANIFEST_CSV
  )),
  Role = c(
    "Candidate-level tuning summaries.",
    "Inner candidate-fold results.",
    "Selected ANN configurations by outer task and model.",
    "Canonical tuning run metadata.",
    "Canonical tuning manifest.",
    "CSV copy of candidate summaries.",
    "CSV copy of inner candidate-fold results.",
    "CSV copy of selected configurations.",
    "CSV representation of the tuning manifest."
  ),
  stringsAsFactors = FALSE
)

format_table(
  tuning_output_files,
  caption = "Canonical ANN tuning output artifacts"
)
```

## 5. ANN parameter values

The scientific tuning space is defined once below and reused by candidate
construction, inner-CV training, nested tuning, and final fitting.

```{r ann-parameter-values}
SEED_GLOBAL <- 20260821L

MAX_EPOCHS <- 2000L

# Compatibility token only: preserves the exact config_id of the completed
# ANN_TUNING_CANONICAL_V1 run. It does not control final refit length.
TUNING_V1_COMPATIBILITY_TOKEN <- 2000L

PATIENCE <- 25L
MIN_DELTA <- 1e-6
EARLY_VALIDATION_FRACTION <- 0.15
DROPOUT <- 0

ARCHITECTURES <- list(
  `2` = c(2L),
  `4` = c(4L),
  `8` = c(8L),
  `16` = c(16L),
  `32` = c(32L),
  `8-4` = c(8L, 4L),
  `16-8` = c(16L, 8L),
  `32-16` = c(32L, 16L),
  `64-32` = c(64L, 32L)
)

LEARNING_RATES <- c(
  1.25e-4,
  2.5e-4,
  5e-4,
  1e-3
)

L1_VALUES <- c(
  1e-5,
  1e-4,
  1e-3,
  1e-2,
  3e-2,
  1e-1
)

L2_VALUES <- c(
  1e-5,
  1e-4,
  1e-3,
  1e-2
)

# Full Cartesian grid for the simultaneous L1+L2 model.
# Both coefficients are strictly positive in ANN_L1_L2.
L1_L2_GRID <- expand.grid(
  L1 = L1_VALUES,
  L2 = L2_VALUES,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

MODEL_LEVELS <- c(
  "ANN_0",
  "ANN_L1",
  "ANN_L2",
  "ANN_L1_L2"
)

ANN_TUNING_CONFIG_ID <- paste(
  "ANN_TUNING_CANONICAL_V1",
  paste(names(ARCHITECTURES), collapse = "-"),
  paste(LEARNING_RATES, collapse = "-"),
  paste(L1_VALUES, collapse = "-"),
  paste(L2_VALUES, collapse = "-"),
  MAX_EPOCHS,
  PATIENCE,
  MIN_DELTA,
  EARLY_VALIDATION_FRACTION,
  TUNING_V1_COMPATIBILITY_TOKEN,
  sep = "__"
)

if (RUN_ANN_TUNING) {
  dir.create(
    CHECKPOINT_DIR,
    recursive = TRUE,
    showWarnings = FALSE
  )
  dir.create(
    dirname(CANDIDATE_RDS),
    recursive = TRUE,
    showWarnings = FALSE
  )
}
```

`TUNING_V1_COMPATIBILITY_TOKEN` is retained solely to preserve the historical
`ANN_TUNING_CANONICAL_V1` configuration identifier. It does **not** determine
the final ANN refit length.

### 5.1 Architectures

```{r architecture-values}
architecture_table <- data.frame(
  Architecture = names(ARCHITECTURES),
  Hidden_layers = vapply(
    ARCHITECTURES,
    function(x) paste(x, collapse = " - "),
    character(1)
  ),
  Number_of_hidden_layers = vapply(
    ARCHITECTURES,
    length,
    integer(1)
  ),
  stringsAsFactors = FALSE
)

format_table(
  architecture_table,
  caption = "Prespecified ANN architecture catalog"
)
```

The `64-32` architecture is the prespecified upper bound of the canonical
search space.

### 5.2 Learning rates

```{r learning-rate-values}
learning_rate_table <- data.frame(
  Learning_rate = LEARNING_RATES
)

format_table(
  learning_rate_table,
  digits = 6,
  caption = "Candidate Adam learning rates"
)
```

### 5.3 L1 penalties

```{r l1-values}
l1_table <- data.frame(
  L1 = L1_VALUES
)

format_table(
  l1_table,
  digits = 6,
  caption = "Candidate L1 penalty coefficients"
)
```

### 5.4 L2 penalties

```{r l2-values}
l2_table <- data.frame(
  L2 = L2_VALUES
)

format_table(
  l2_table,
  digits = 6,
  caption = "Candidate L2 penalty coefficients"
)
```

### 5.5 Combined L1 + L2 grid

`ANN_L1_L2` uses the full Cartesian product of the six positive L1 values and
four positive L2 values, yielding 24 combined penalty configurations.

```{r l1-l2-values}
format_table(
  L1_L2_GRID,
  digits = 6,
  caption = "Positive L1 x L2 grid for ANN-L1+L2"
)
```

## 6. Candidate grids

Architecture and learning rate are selected **once per outer task using
`ANN_0`**. The selected structure is then shared by all four ANN variants. This
keeps the comparison between regularization variants focused on L1/L2
penalties rather than allowing each model family to change network capacity.

After structure selection:

- `ANN_L1` tunes only L1;
- `ANN_L2` tunes only L2;
- `ANN_L1_L2` tunes the full positive L1 x L2 Cartesian grid.

```{r grids}
structure_grid <- expand.grid(
  Architecture = names(ARCHITECTURES),
  Learning_rate = LEARNING_RATES,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)
structure_grid$Model <- "ANN_0"
structure_grid$L1 <- 0
structure_grid$L2 <- 0

l1_grid <- data.frame(
  Model = "ANN_L1",
  L1 = L1_VALUES,
  L2 = 0,
  stringsAsFactors = FALSE
)

l2_grid <- data.frame(
  Model = "ANN_L2",
  L1 = 0,
  L2 = L2_VALUES,
  stringsAsFactors = FALSE
)

l1l2_grid <- data.frame(
  Model = "ANN_L1_L2",
  L1 = L1_L2_GRID$L1,
  L2 = L1_L2_GRID$L2,
  stringsAsFactors = FALSE
)

grid_summary <- data.frame(
  Stage = c(
    "Shared architecture + learning rate",
    "ANN_L1 penalty",
    "ANN_L2 penalty",
    "ANN_L1_L2 penalties"
  ),
  Candidates = c(
    nrow(structure_grid),
    nrow(l1_grid),
    nrow(l2_grid),
    nrow(l1l2_grid)
  )
)

grid_summary$Inner_fits_per_outer_task <-
  4L * grid_summary$Candidates

format_table(
  grid_summary,
  caption = "Canonical ANN tuning grid"
)

CANDIDATES_PER_TASK <- sum(
  grid_summary$Candidates
)
INNER_FITS_PER_TASK <- 4L *
  CANDIDATES_PER_TASK

cat(
  "Candidates per outer task:",
  CANDIDATES_PER_TASK,
  "\n"
)
cat(
  "Inner candidate-fold fits per outer task:",
  INNER_FITS_PER_TASK,
  "\n"
)
cat(
  "Planned inner fits for all 150 outer tasks:",
  150L * INNER_FITS_PER_TASK,
  "\n"
)
```

The grid contains 70 candidate configurations per outer task. With four frozen
inner folds, this corresponds to 280 candidate-fold fits per task and 42,000
fits across the 150 scenario-by-split tasks. Architecture and learning rate are
selected using `ANN_0` and then held common across all four ANN variants within
each task, so their subsequent contrast isolates the L1/L2 penalty choice.

## 7. Data preprocessing

Preprocessing is estimated independently inside each fitting subset. The same
fitted transformation is then applied to the corresponding early-validation and
assessment observations. This ordering prevents validation information from
influencing the parameters used to train the network.

### 7.1 Marker standardization

For each fit, marker means and standard deviations are estimated only from the
current fitting individuals. The fitted center and scale are subsequently
applied unchanged to early-validation and assessment individuals.

Markers with effectively zero fitting-set variance receive unit scale. This
avoids numerical division by zero without introducing information from outside
the fitting subset.

### 7.2 Response standardization

Phenotypes are centered and scaled using only the fitting subset. Network
predictions are transformed back to the original phenotype scale before RMSE,
correlation, and MAE are calculated.

### 7.3 Early-validation split

The early-validation subset is drawn deterministically from the current
analysis sample using a prespecified seed and fraction. It is distinct from the
frozen inner-assessment fold, which remains untouched until candidate
prediction.

The helper functions below implement these transformations and retain the
historical metric aliases required by the result-building code.

```{r support-functions}
correlation <- function(predicted, observed) {
  cor(predicted, observed)
}

rmse <- function(predicted, observed) {
  sqrt(
    mean(
      (predicted - observed)^2
    )
  )
}

mae <- function(predicted, observed) {
  mean(
    abs(
      predicted - observed
    )
  )
}

# Historical names retained for compatibility with result-building code.
safe_cor <- correlation
safe_rmse <- rmse
safe_mae <- mae

parameter_count <- function(
  input_dim,
  hidden_units
) {
  inputs <- c(input_dim, hidden_units)
  outputs <- c(hidden_units, 1L)

  sum(
    inputs * outputs +
      outputs
  )
}

make_validation_split <- function(
  x,
  frac,
  seed
) {
  set.seed(seed)

  n_validation <- as.integer(
    round(
      length(x) * frac
    )
  )

  validation <- sort(
    sample(
      x,
      n_validation
    )
  )

  list(
    train = sort(
      setdiff(
        x,
        validation
      )
    ),
    validation = validation
  )
}

fit_x_scaler <- function(x) {
  center <- colMeans(x)
  scale <- apply(x, 2, sd)

  zero_variance <- scale < 1e-8
  scale[zero_variance] <- 1

  list(
    center = center,
    scale = scale,
    zero = zero_variance
  )
}

scale_x <- function(x, scaler) {
  sweep(
    sweep(
      x,
      2,
      scaler$center,
      "-"
    ),
    2,
    scaler$scale,
    "/"
  )
}

fit_y_scaler <- function(y) {
  list(
    center = mean(y),
    scale = max(
      sd(y),
      1e-8
    )
  )
}

scale_y <- function(y, scaler) {
  (y - scaler$center) /
    scaler$scale
}

unscale_y <- function(y, scaler) {
  y * scaler$scale +
    scaler$center
}
```

## 8. Neural-network definition

### 8.1 Hidden layers and ReLU

For a standardized marker vector \(\mathbf{x}\), each hidden layer applies a
linear transformation followed by ReLU activation:

\[
\mathbf{h}^{(\ell)} =
\operatorname{ReLU}
(\mathbf{W}_{\ell}\mathbf{h}^{(\ell-1)}+\mathbf{b}_{\ell}).
\]

The symbols used in the network equations have the following roles:

```{r network-symbols}
network_symbols <- data.frame(
  Symbol = c("x", "W", "b", "h", "y-hat"),
  Meaning = c(
    "Standardized marker vector.",
    "Layer weights.",
    "Layer biases.",
    "Hidden-layer activations.",
    "Network prediction."
  ),
  stringsAsFactors = FALSE
)

format_table(
  network_symbols,
  caption = "Symbols used in the ANN definition"
)
```

### 8.2 Linear output

The output layer is linear because the response is continuous. No output
activation is applied after the last linear transformation.

### 8.3 L1 and L2 penalties

Training minimizes

\[
\mathcal{L}=
\operatorname{MSE}(y,\hat{y})+
\lambda_1\sum_j |w_j|+
\frac{\lambda_2}{2}\sum_j w_j^2.
\]

`ANN_0` uses \(\lambda_1=\lambda_2=0\), `ANN_L1` uses only L1,
`ANN_L2` uses only L2, and `ANN_L1_L2` uses both. Bias terms are not penalized.
Adam uses `weight_decay = 0`, so L2 enters only through the explicit penalty
above.

The response is standardized from the fitting subset during optimization and
predictions are transformed back to the original phenotype scale before
performance metrics are calculated.

The largest `64-32` architecture in the prespecified catalog illustrates
the same hidden-layer and linear-output structure used throughout the search:

```{r ann-forward-example, eval=FALSE}
model <- nn_module(
  initialize = function(input_dim) {
    self$hidden1 <- nn_linear(input_dim, 64)
    self$hidden2 <- nn_linear(64, 32)
    self$output <- nn_linear(32, 1)
  },
  forward = function(x) {
    x <- nnf_relu(self$hidden1(x))
    x <- nnf_relu(self$hidden2(x))
    self$output(x)
  }
)
```

Smaller catalog architectures follow the same pattern with fewer hidden units
or one hidden layer. The generic implementation shown next constructs all nine
architectures from the same definition.

```{r model-definition, eval=ANN_RUNTIME_REQUIRED}
TorchGenomicMLP <- torch::nn_module(
  "TorchGenomicMLP_CanonicalTuning",

  initialize = function(
    input_dim,
    hidden_units
  ) {
    self$linears <-
      torch::nn_module_list()

    previous <-
      as.integer(input_dim)

    for (hidden in hidden_units) {
      self$linears$append(
        torch::nn_linear(
          previous,
          as.integer(hidden)
        )
      )
      previous <- as.integer(hidden)
    }

    self$output <-
      torch::nn_linear(
        previous,
        1L
      )
  },

  forward = function(x) {
    for (
      layer_index in seq_len(
        length(self$linears)
      )
    ) {
      x <- torch::nnf_relu(
        self$linears[[layer_index]](x)
      )
    }

    self$output(x)
  }
)

weight_penalty <- function(
  model,
  l1,
  l2
) {
  weights <- lapply(
    seq_len(
      length(model$linears)
    ),
    function(layer_index) {
      model$linears[[layer_index]]$weight
    }
  )

  weights[[length(weights) + 1L]] <- model$output$weight

  penalty_l1 <-
    weights[[1]]$sum() * 0
  penalty_l2 <-
    weights[[1]]$sum() * 0

  for (weight in weights) {
    penalty_l1 <-
      penalty_l1 +
      weight$abs()$sum()

    penalty_l2 <-
      penalty_l2 +
      weight$pow(2)$sum()
  }

  l1 * penalty_l1 +
    (l2 / 2) * penalty_l2
}

clone_state <- function(model) {
  state <- model$state_dict()

  copied_state <- lapply(
    state,
    function(parameter) {
      parameter$detach()$clone()
    }
  )

  names(copied_state) <-
    names(state)

  copied_state
}
```

## 9. Inner-CV training

For every frozen inner fold, the inner-analysis individuals are divided
deterministically into a fitting subset and an early-validation subset. The
frozen inner-assessment individuals remain untouched until the candidate has
been trained and its best checkpoint restored.

### 9.1 One optimization step

Each epoch predicts the fitting subset, calculates mean squared error plus the
selected weight penalty, backpropagates the loss, and updates the network with
Adam.

```{r ann-optimization-step, eval=FALSE}
optimizer$zero_grad()

prediction <- model(x_fit)

loss <- nnf_mse_loss(
  prediction,
  y_fit
) + weight_penalty(
  model,
  l1,
  l2
)

loss$backward()
optimizer$step()
```

### 9.2 Complete candidate training

The complete helper shown next applies the fitting-set scalers, trains the
candidate, monitors early-validation loss, restores the best checkpoint, and
predicts the untouched inner-assessment fold.

The conditionals inside this function are part of the ANN optimization
algorithm itself rather than defensive execution guards.

```{r training-functions}
prepare_bundle <- function(
  fit_ids,
  validation_ids,
  assessment_ids,
  scenario_index
) {
  fit_index <- match(fit_ids, ids)
  validation_index <- match(
    validation_ids,
    ids
  )
  assessment_index <- match(
    assessment_ids,
    ids
  )

  x_scaler <- fit_x_scaler(
    geno[
      fit_index,
      ,
      drop = FALSE
    ]
  )
  y_scaler <- fit_y_scaler(
    phenotype[
      fit_index,
      scenario_index
    ]
  )

  list(
    x_fit = scale_x(
      geno[
        fit_index,
        ,
        drop = FALSE
      ],
      x_scaler
    ),
    x_validation = scale_x(
      geno[
        validation_index,
        ,
        drop = FALSE
      ],
      x_scaler
    ),
    x_assessment = scale_x(
      geno[
        assessment_index,
        ,
        drop = FALSE
      ],
      x_scaler
    ),
    y_fit = scale_y(
      phenotype[
        fit_index,
        scenario_index
      ],
      y_scaler
    ),
    y_validation = scale_y(
      phenotype[
        validation_index,
        scenario_index
      ],
      y_scaler
    ),
    y_assessment = phenotype[
      assessment_index,
      scenario_index
    ],
    y_scaler = y_scaler,
    zero_variance_markers =
      sum(x_scaler$zero)
  )
}

train_inner_candidate <- function(
  bundle,
  architecture,
  learning_rate,
  l1,
  l2,
  seed
) {
  set_ann_seed(seed)

  x_fit <- torch::torch_tensor(
    bundle$x_fit,
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  y_fit <- torch::torch_tensor(
    matrix(bundle$y_fit, ncol = 1),
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  x_validation <- torch::torch_tensor(
    bundle$x_validation,
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  y_validation <- torch::torch_tensor(
    matrix(
      bundle$y_validation,
      ncol = 1
    ),
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  x_assessment <- torch::torch_tensor(
    bundle$x_assessment,
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  model <- TorchGenomicMLP(
    ncol(bundle$x_fit),
    ARCHITECTURES[[architecture]]
  )$to(device = ANN_DEVICE)

  optimizer <- torch::optim_adam(
    params = model$parameters,
    lr = learning_rate,
    weight_decay = 0
  )

  best_validation_loss <- Inf
  checkpoint_epoch <- 1L
  checkpoint_state <- clone_state(model)
  stale_epochs <- 0L
  last_epoch <- 0L
  stopped_by_patience <- FALSE

  sync_device()
  fit_start <-
    proc.time()[["elapsed"]]

  for (epoch in seq_len(MAX_EPOCHS)) {
    last_epoch <- epoch

    model$train()
    optimizer$zero_grad()

    prediction <- model(x_fit)

    loss <-
      torch::nnf_mse_loss(
        prediction,
        y_fit
      ) +
      weight_penalty(
        model,
        l1,
        l2
      )

    loss$backward()
    optimizer$step()

    model$eval()

    validation_loss <-
      torch::with_no_grad({
        torch::nnf_mse_loss(
          model(x_validation),
          y_validation
        )
      })

    validation_value <-
      as.numeric(
        validation_loss$to(
          device =
            torch::torch_device("cpu")
        )
      )

    improved <-
      validation_value <
      best_validation_loss -
      MIN_DELTA

    if (improved) {
      best_validation_loss <-
        validation_value
      checkpoint_epoch <- epoch
      checkpoint_state <-
        clone_state(model)
      stale_epochs <- 0L
    } else {
      stale_epochs <-
        stale_epochs + 1L
    }

    if (stale_epochs >= PATIENCE) {
      stopped_by_patience <- TRUE
      break
    }
  }

  sync_device()
  fit_seconds <-
    proc.time()[["elapsed"]] -
    fit_start

  model$load_state_dict(
    checkpoint_state
  )
  model$eval()

  prediction_start <-
    proc.time()[["elapsed"]]

  prediction_scaled <-
    torch::with_no_grad({
      as.numeric(
        model(x_assessment)$to(
          device =
            torch::torch_device("cpu")
        )
      )
    })

  prediction_seconds <-
    proc.time()[["elapsed"]] -
    prediction_start

  list(
    prediction = unscale_y(
      prediction_scaled,
      bundle$y_scaler
    ),
    early_stop_epoch =
      as.integer(checkpoint_epoch),
    last_epoch =
      as.integer(last_epoch),
    stopped_by_patience =
      stopped_by_patience,
    hit_epoch_ceiling =
      last_epoch >= MAX_EPOCHS &&
      !stopped_by_patience,
    fit_seconds = fit_seconds,
    prediction_seconds =
      prediction_seconds,
    zero_variance_markers =
      bundle$zero_variance_markers
  )
}
```

### 9.3 Early stopping

Within `train_inner_candidate()`, an improvement requires the current
early-validation loss to be at least `MIN_DELTA` below the best loss observed so
far. An improved epoch replaces the stored checkpoint and resets the stale-epoch
counter.

Training stops when `stale_epochs >= PATIENCE`. The stored best checkpoint is
then restored before the untouched inner-assessment fold is predicted. Thus,
the epoch at which training stops and the epoch retained for prediction are not
necessarily the same.

Early-stopping diagnostics are saved for optimization assessment but do not
participate in candidate ranking.


## 10. Candidate evaluation and ranking

Each candidate is fitted independently in the four frozen inner folds. The
candidate-level summary averages the held-out inner-assessment metrics across
those four fits.

Candidate ranking is deterministic and does **not** include epoch information.

The ranking hierarchy is:

1. lower mean inner-assessment RMSE;
2. higher mean inner-assessment correlation;
3. fewer network parameters;
4. smaller total penalty coefficient;
5. smaller learning rate;
6. candidate ID.

A completed candidate table is therefore sorted by the same hierarchy:

```{r candidate-ranking-example, eval=FALSE}
candidate_summary <- candidate_summary[
  order(
    candidate_summary$Mean_RMSE,
    -candidate_summary$Mean_Correlation,
    candidate_summary$Parameters,
    candidate_summary$Penalty,
    candidate_summary$Learning_rate,
    candidate_summary$CandidateID
  ),
]
```

The first row is the selected candidate. The complete function shown next
fits the same candidate in each of the four inner folds, averages the three
prediction metrics, and then applies exactly this ranking hierarchy.

```{r candidate-functions}
evaluate_candidate <- function(
  bundles,
  model_name,
  architecture,
  learning_rate,
  l1,
  l2,
  seed_base,
  candidate_id
) {
  fold_rows <- vector("list", 4L)

  for (inner_fold in 1:4) {
    fitted <- train_inner_candidate(
      bundles[[inner_fold]],
      architecture,
      learning_rate,
      l1,
      l2,
      seed_base + inner_fold
    )

    fold_rows[[inner_fold]] <- data.frame(
      Model = model_name,
      CandidateID = candidate_id,
      Architecture = architecture,
      Learning_rate = learning_rate,
      L1 = l1,
      L2 = l2,
      InnerFold = inner_fold,
      RMSE = rmse(
        fitted$prediction,
        bundles[[inner_fold]]$y_assessment
      ),
      Correlation = correlation(
        fitted$prediction,
        bundles[[inner_fold]]$y_assessment
      ),
      MAE = mae(
        fitted$prediction,
        bundles[[inner_fold]]$y_assessment
      ),
      EarlyStop_epoch =
        fitted$early_stop_epoch,
      Last_epoch =
        fitted$last_epoch,
      Stopped_by_patience =
        fitted$stopped_by_patience,
      Hit_epoch_ceiling =
        fitted$hit_epoch_ceiling,
      Fit_seconds =
        fitted$fit_seconds,
      Prediction_seconds =
        fitted$prediction_seconds,
      Status = "ok",
      Error = NA_character_,
      stringsAsFactors = FALSE
    )
  }

  folds <- do.call(
    rbind,
    fold_rows
  )

  candidate_summary <- data.frame(
    Model = model_name,
    CandidateID = candidate_id,
    Architecture = architecture,
    Learning_rate = learning_rate,
    L1 = l1,
    L2 = l2,
    Inner_folds_ok = 4L,
    Mean_RMSE = mean(folds$RMSE),
    Mean_Correlation =
      mean(folds$Correlation),
    Mean_MAE = mean(folds$MAE),
    Median_EarlyStop_epoch =
      median(folds$EarlyStop_epoch),
    Max_Last_epoch =
      max(folds$Last_epoch),
    N_epoch_ceiling =
      sum(folds$Hit_epoch_ceiling),
    Status = "ok",
    stringsAsFactors = FALSE
  )

  list(
    summary = candidate_summary,
    folds = folds
  )
}

choose_candidate <- function(x) {
  ranked <- x

  ranked$Parameters <- vapply(
    ranked$Architecture,
    function(architecture) {
      parameter_count(
        ncol(geno),
        ARCHITECTURES[[architecture]]
      )
    },
    numeric(1)
  )

  ranked$Penalty <-
    ranked$L1 +
    ranked$L2

  ranked <- ranked[
    order(
      ranked$Mean_RMSE,
      -ranked$Mean_Correlation,
      ranked$Parameters,
      ranked$Penalty,
      ranked$Learning_rate,
      ranked$CandidateID
    ),
    ,
    drop = FALSE
  ]

  ranked[
    1,
    ,
    drop = FALSE
  ]
}
```

## 11. Hyperparameter tuning

Nested tuning uses the frozen validation design from M02. The outer assessment
set is not predicted anywhere in this stage; it remains reserved for final
model evaluation after hyperparameter and epoch selection are complete.

### 11.1 Frozen outer tasks

The six scenarios share the same 25 frozen outer splits, producing 150
`Scenario × outer split` tuning tasks.

```{r tasks}
outer_map <- unique(
  cv$outer_assignments[
    ,
    c(
      "Repetition",
      "OuterFold",
      "SplitID"
    )
  ]
)

outer_map <- outer_map[
  order(
    outer_map$Repetition,
    outer_map$OuterFold
  ),
  ,
  drop = FALSE
]

task_table <- merge(
  data.frame(
    Scenario = scenario_names,
    stringsAsFactors = FALSE
  ),
  outer_map,
  by = NULL
)

task_table <- task_table[
  order(
    match(
      task_table$Scenario,
      scenario_names
    ),
    task_table$Repetition,
    task_table$OuterFold
  ),
  ,
  drop = FALSE
]

task_table$TaskID <- paste(
  task_table$Scenario,
  task_table$SplitID,
  sep = "_"
)
task_table$Fold <-
  task_table$OuterFold

rownames(task_table) <- NULL

EXPECTED_TASKS <- nrow(task_table)
EXPECTED_SELECTED_ROWS <-
  EXPECTED_TASKS *
  length(MODEL_LEVELS)
EXPECTED_CANDIDATE_ROWS <-
  EXPECTED_TASKS *
  CANDIDATES_PER_TASK
EXPECTED_FOLD_ROWS <-
  EXPECTED_TASKS *
  INNER_FITS_PER_TASK

format_table(
  head(task_table, 15),
  caption =
    "First 15 canonical ANN tuning tasks",
  row.names = FALSE
)
```



Each row represents one `Scenario × Repetition × Fold` combination. The same
`TaskID` is used for all four ANN variants, which ensures that their selected
regularization settings are compared within the same frozen outer task.

### 11.2 Architecture and learning rate

For each outer task, the 36 architecture-by-learning-rate combinations are
evaluated with `ANN_0`. The deterministic candidate ranking described in
Section 10 selects one shared architecture and one shared learning rate.

### 11.3 L1

With architecture and learning rate fixed, `ANN_L1` evaluates the six
prespecified positive L1 coefficients. No additional architecture search is
performed for the L1-regularized model.

### 11.4 L2

`ANN_L2` reuses the same selected architecture and learning rate and evaluates
only the four prespecified positive L2 coefficients.

### 11.5 L1 + L2

`ANN_L1_L2` also reuses the shared architecture and learning rate, evaluating
the complete 6 x 4 positive L1-by-L2 grid. This produces 24 combined-penalty
candidates.

The chunk below implements all four tuning stages for the 150 frozen outer
tasks. It evaluates only inner-CV candidates and never predicts the outer
assessment set.

```{r run-tuning, eval=RUN_ANN_TUNING}
run_start <-
  proc.time()[["elapsed"]]

task_outputs <- vector(
  "list",
  nrow(task_table)
)

for (
  task_index in seq_len(
    nrow(task_table)
  )
) {
  task <- task_table[
    task_index,
    ,
    drop = FALSE
  ]

  scenario <- task$Scenario
  scenario_index <- match(
    scenario,
    scenario_names
  )
  split_id <- task$SplitID

  task_start <-
    proc.time()[["elapsed"]]

  inner <-
    cv$inner_splits[[split_id]]

  bundles <- vector(
    "list",
    4L
  )

  for (inner_fold in 1:4) {
    inner_split <-
      inner[[paste0(
        "I",
        inner_fold
      )]]

    validation_seed <-
      SEED_GLOBAL +
      scenario_index * 1000000L +
      task$Repetition * 10000L +
      task$OuterFold * 1000L +
      inner_fold * 10L

    validation <-
      make_validation_split(
        inner_split$analysis_ids,
        EARLY_VALIDATION_FRACTION,
        validation_seed
      )

    bundles[[inner_fold]] <-
      prepare_bundle(
        validation$train,
        validation$validation,
        inner_split$assessment_ids,
        scenario_index
      )
  }

  base_seed <-
    SEED_GLOBAL +
    scenario_index * 10000000L +
    task$Repetition * 100000L +
    task$OuterFold * 10000L

  # Stage 1: select architecture and learning rate using ANN_0.
  structure_summary_rows <- vector(
    "list",
    nrow(structure_grid)
  )
  structure_fold_rows <- vector(
    "list",
    nrow(structure_grid)
  )

  for (
    candidate_index in seq_len(
      nrow(structure_grid)
    )
  ) {
    candidate <-
      structure_grid[
        candidate_index,
        ,
        drop = FALSE
      ]

    evaluated <- evaluate_candidate(
      bundles = bundles,
      model_name = "ANN_0",
      architecture =
        candidate$Architecture,
      learning_rate =
        candidate$Learning_rate,
      l1 = 0,
      l2 = 0,
      seed_base =
        base_seed + 1000000L,
      candidate_id =
        candidate_index
    )

    structure_summary_rows[[candidate_index]] <- evaluated$summary

    structure_fold_rows[[candidate_index]] <- evaluated$folds
  }

  structure_summary <- do.call(
    rbind,
    structure_summary_rows
  )
  structure_folds <- do.call(
    rbind,
    structure_fold_rows
  )

  chosen_structure <-
    choose_candidate(
      structure_summary
    )

  selected_architecture <-
    chosen_structure$Architecture
  selected_learning_rate <-
    chosen_structure$Learning_rate

  # Stage 2: tune only the regularization coefficients.
  penalty_grids <- list(
    ANN_L1 = l1_grid,
    ANN_L2 = l2_grid,
    ANN_L1_L2 = l1l2_grid
  )

  penalty_summaries <- list()
  penalty_folds <- list()
  penalty_selected <- list()

  for (
    model_name in names(
      penalty_grids
    )
  ) {
    grid <-
      penalty_grids[[model_name]]

    candidate_summaries <- vector(
      "list",
      nrow(grid)
    )
    candidate_folds <- vector(
      "list",
      nrow(grid)
    )

    for (
      candidate_index in seq_len(
        nrow(grid)
      )
    ) {
      evaluated <- evaluate_candidate(
        bundles = bundles,
        model_name = model_name,
        architecture =
          selected_architecture,
        learning_rate =
          selected_learning_rate,
        l1 =
          grid$L1[candidate_index],
        l2 =
          grid$L2[candidate_index],
        seed_base =
          base_seed + 2000000L,
        candidate_id =
          candidate_index
      )

      candidate_summaries[[candidate_index]] <- evaluated$summary
      candidate_folds[[candidate_index]] <- evaluated$folds
    }

    penalty_summaries[[model_name]] <- do.call(
      rbind,
      candidate_summaries
    )
    penalty_folds[[model_name]] <- do.call(
      rbind,
      candidate_folds
    )
    penalty_selected[[model_name]] <- choose_candidate(
      penalty_summaries[[model_name]]
    )
  }

  selected_table <- rbind(
    data.frame(
      Model = "ANN_0",
      Architecture =
        selected_architecture,
      Learning_rate =
        selected_learning_rate,
      L1 = 0,
      L2 = 0,
      Inner_RMSE =
        chosen_structure$Mean_RMSE,
      Inner_Correlation =
        chosen_structure$Mean_Correlation,
      Inner_MAE =
        chosen_structure$Mean_MAE,
      stringsAsFactors = FALSE
    ),
    do.call(
      rbind,
      lapply(
        names(penalty_selected),
        function(model_name) {
          selected_candidate <-
            penalty_selected[[model_name]]

          data.frame(
            Model = model_name,
            Architecture =
              selected_architecture,
            Learning_rate =
              selected_learning_rate,
            L1 =
              selected_candidate$L1,
            L2 =
              selected_candidate$L2,
            Inner_RMSE =
              selected_candidate$Mean_RMSE,
            Inner_Correlation =
              selected_candidate$Mean_Correlation,
            Inner_MAE =
              selected_candidate$Mean_MAE,
            stringsAsFactors = FALSE
          )
        }
      )
    )
  )

  for (
    column_name in c(
      "Scenario",
      "Repetition",
      "Fold",
      "SplitID",
      "TaskID"
    )
  ) {
    selected_table[[column_name]] <-
      task[[column_name]]
  }

  tuning_summary <- rbind(
    structure_summary,
    do.call(
      rbind,
      penalty_summaries
    )
  )
  tuning_folds <- rbind(
    structure_folds,
    do.call(
      rbind,
      penalty_folds
    )
  )

  for (
    column_name in c(
      "Scenario",
      "Repetition",
      "Fold",
      "SplitID",
      "TaskID"
    )
  ) {
    tuning_summary[[column_name]] <-
      task[[column_name]]
    tuning_folds[[column_name]] <-
      task[[column_name]]
  }

  task_seconds <-
    proc.time()[["elapsed"]] -
    task_start

  task_outputs[[task_index]] <- list(
    Status = "ok",
    ConfigID =
      ANN_TUNING_CONFIG_ID,
    TaskID = task$TaskID,
    selected = selected_table,
    tuning_summary = tuning_summary,
    tuning_folds = tuning_folds,
    task_seconds = task_seconds
  )

  cat(
    sprintf(
      "[%03d/%03d] %s | %.1f s | arch=%s lr=%g\n",
      task_index,
      nrow(task_table),
      task$TaskID,
      task_seconds,
      selected_architecture,
      selected_learning_rate
    )
  )
}

current_session_seconds <-
  proc.time()[["elapsed"]] -
  run_start

selected_configurations <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(task_output) {
      task_output$selected
    }
  )
)
candidate_summary <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(task_output) {
      task_output$tuning_summary
    }
  )
)
fold_results <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(task_output) {
      task_output$tuning_folds
    }
  )
)

task_seconds_total <- sum(
  vapply(
    task_outputs,
    function(task_output) {
      task_output$task_seconds
    },
    numeric(1)
  )
)

manifest <- list(
  analysis =
    "canonical_simulated_ann_tuning",
  dataset = "simulated",
  mode = "full",
  pilot_task = NA_character_,
  config_id =
    ANN_TUNING_CONFIG_ID,
  seed_global = SEED_GLOBAL,
  models = MODEL_LEVELS,
  architectures = ARCHITECTURES,
  learning_rates = LEARNING_RATES,
  l1_values = L1_VALUES,
  l2_values = L2_VALUES,
  l1_l2_grid = L1_L2_GRID,
  optimizer = "Adam",
  activation = "ReLU",
  dropout = DROPOUT,
  max_epochs = MAX_EPOCHS,
  final_train_epochs =
    TUNING_V1_COMPATIBILITY_TOKEN,
  patience = PATIENCE,
  min_delta = MIN_DELTA,
  early_validation_fraction =
    EARLY_VALIDATION_FRACTION,
  primary_tuning_metric =
    "mean inner-assessment RMSE",
  secondary_tuning_metric =
    "mean inner-assessment correlation",
  epoch_used_in_candidate_ranking =
    FALSE,
  outer_assessment_used =
    FALSE,
  true_genetic_value_used =
    FALSE,
  architecture_lr_shared_across_models =
    TRUE,
  candidates_per_task =
    CANDIDATES_PER_TASK,
  inner_fits_per_task =
    INNER_FITS_PER_TASK,
  number_tasks =
    EXPECTED_TASKS,
  selected_rows =
    nrow(selected_configurations),
  candidate_rows =
    nrow(candidate_summary),
  fold_rows =
    nrow(fold_results),
  torch_version =
    as.character(
      packageVersion("torch")
    ),
  device = ANN_DEVICE_LABEL,
  cuda_available =
    cuda_available,
  current_session_seconds =
    current_session_seconds,
  checkpoint_task_seconds =
    task_seconds_total,
  timestamp =
    as.character(Sys.time())
)

manifest_table <- data.frame(
  Field = c(
    "analysis",
    "mode",
    "config_id",
    "architectures",
    "learning_rates",
    "L1_values",
    "L2_values",
    "combined_grid_size",
    "MAX_EPOCHS",
    "TUNING_V1_COMPATIBILITY_TOKEN",
    "patience",
    "dropout",
    "optimizer",
    "activation",
    "candidates_per_task",
    "inner_fits_per_task",
    "tasks",
    "epoch_used_in_candidate_ranking",
    "outer_assessment_used",
    "true_genetic_value_used"
  ),
  Value = c(
    manifest$analysis,
    manifest$mode,
    manifest$config_id,
    paste(names(ARCHITECTURES), collapse = " | "),
    paste(LEARNING_RATES, collapse = " | "),
    paste(L1_VALUES, collapse = " | "),
    paste(L2_VALUES, collapse = " | "),
    nrow(L1_L2_GRID),
    MAX_EPOCHS,
    TUNING_V1_COMPATIBILITY_TOKEN,
    PATIENCE,
    DROPOUT,
    "Adam",
    "ReLU",
    CANDIDATES_PER_TASK,
    INNER_FITS_PER_TASK,
    EXPECTED_TASKS,
    FALSE,
    FALSE,
    FALSE
  ),
  stringsAsFactors = FALSE
)

saveRDS(
  selected_configurations,
  SELECTED_RDS,
  compress = "gzip"
)
saveRDS(
  candidate_summary,
  CANDIDATE_RDS,
  compress = "gzip"
)
saveRDS(
  fold_results,
  FOLD_RDS,
  compress = "gzip"
)
saveRDS(
  manifest,
  RUN_RDS,
  compress = "gzip"
)
saveRDS(
  manifest,
  MANIFEST_RDS,
  compress = "gzip"
)

write.csv(
  selected_configurations,
  SELECTED_CSV,
  row.names = FALSE
)
write.csv(
  candidate_summary,
  CANDIDATE_CSV,
  row.names = FALSE
)
write.csv(
  fold_results,
  FOLD_CSV,
  row.names = FALSE
)
write.csv(
  manifest_table,
  MANIFEST_CSV,
  row.names = FALSE
)
```

## 12. Selected configurations

The reader-facing summaries below use the canonical saved tuning outputs. This
allows the page to be rebuilt without rerunning the 42,000 inner candidate-fold
fits while preserving the exact configurations selected by the canonical run.

```{r saved-integrity}
selected_configurations <- readRDS(SELECTED_RDS)
candidate_summary <- readRDS(CANDIDATE_RDS)
fold_results <- readRDS(FOLD_RDS)
run_metadata <- readRDS(RUN_RDS)

tuning_output_summary <- data.frame(
  Object = c(
    "Selected configurations", "Candidate summaries",
    "Inner candidate-fold results"
  ),
  Rows = c(
    nrow(selected_configurations), nrow(candidate_summary), nrow(fold_results)
  )
)
format_table(
  tuning_output_summary,
  caption = "Observed size of the canonical ANN tuning outputs",
  row.names = FALSE
)
```

### 12.1 Architecture and learning-rate frequencies

Architecture and learning rate are shared by all four ANN variants within each
outer task. Their selection frequencies therefore summarize the structural
choices made before any model-specific penalty tuning.

```{r selected-structure-summary}
task_level <- unique(
  selected_configurations[
    ,
    c(
      "TaskID",
      "Architecture",
      "Learning_rate"
    )
  ]
)

architecture_frequency <- as.data.frame(
  table(
    Architecture =
      task_level$Architecture
  ),
  stringsAsFactors = FALSE
)
names(architecture_frequency)[2] <-
  "Selected_tasks"

architecture_frequency <-
  architecture_frequency[
    order(
      match(
        architecture_frequency$Architecture,
        names(ARCHITECTURES)
      )
    ),
    ,
    drop = FALSE
  ]

learning_rate_frequency <- as.data.frame(
  table(
    Learning_rate =
      task_level$Learning_rate
  ),
  stringsAsFactors = FALSE
)
names(learning_rate_frequency)[2] <-
  "Selected_tasks"

format_table(
  architecture_frequency,
  caption =
    "Selected shared architecture across the 150 outer tasks",
  row.names = FALSE
)
format_table(
  learning_rate_frequency,
  caption =
    "Selected shared learning rate across the 150 outer tasks",
  row.names = FALSE
)
```

### 12.2 Penalty selections

The regularized variants subsequently select only their applicable L1 and/or L2
coefficients. The frequency tables show how often each penalty value is selected,
and the final table gives a small task-level sample of the canonical choices.

```{r selected-penalty-summary}
l1_frequency <- selected_configurations[
  selected_configurations$Model %in%
    c("ANN_L1", "ANN_L1_L2"),
  c("Model", "L1"),
  drop = FALSE
]
l1_frequency$Selected_tasks <- 1L
l1_frequency <- aggregate(
  Selected_tasks ~ Model + L1,
  data = l1_frequency,
  FUN = sum
)

l2_frequency <- selected_configurations[
  selected_configurations$Model %in%
    c("ANN_L2", "ANN_L1_L2"),
  c("Model", "L2"),
  drop = FALSE
]
l2_frequency$Selected_tasks <- 1L
l2_frequency <- aggregate(
  Selected_tasks ~ Model + L2,
  data = l2_frequency,
  FUN = sum
)

format_table(
  l1_frequency,
  digits = 6,
  caption =
    "Selected L1 coefficients by regularized ANN variant",
  row.names = FALSE
)
format_table(
  l2_frequency,
  digits = 6,
  caption =
    "Selected L2 coefficients by regularized ANN variant",
  row.names = FALSE
)

selected_display <- head(
  selected_configurations[
    order(
      selected_configurations$Scenario,
      selected_configurations$Repetition,
      selected_configurations$Fold,
      match(
        selected_configurations$Model,
        MODEL_LEVELS
      )
    ),
    c(
      "Scenario",
      "Repetition",
      "Fold",
      "Model",
      "Architecture",
      "Learning_rate",
      "L1",
      "L2",
      "Inner_RMSE",
      "Inner_Correlation"
    ),
    drop = FALSE
  ],
  20
)

format_table(
  selected_display,
  digits = 6,
  caption =
    "First 20 selected canonical ANN configurations",
  row.names = FALSE
)
```

Because architecture and learning rate are shared by the four ANN variants within each task, differences among `ANN_0`, `ANN_L1`, `ANN_L2`, and `ANN_L1_L2` are not produced by giving the regularized variants a different structural search space.

### 12.3 Optimization diagnostics

These diagnostics describe how the fixed early-stopping protocol behaved over
the canonical tuning fits. They do not enter candidate ranking and do not
change the tuning grid.

```{r optimization-diagnostics}
selected_keys <- merge(
  selected_configurations[
    ,
    c(
      "TaskID",
      "Model",
      "Architecture",
      "Learning_rate",
      "L1",
      "L2"
    )
  ],
  fold_results,
  by = c(
    "TaskID",
    "Model",
    "Architecture",
    "Learning_rate",
    "L1",
    "L2"
  ),
  all.x = TRUE,
  sort = FALSE
)

optimization_summary <- data.frame(
  Diagnostic = c(
    "Inner candidate-fold fits",
    "Median best early-stop epoch",
    "Median last epoch",
    "Fits stopped by patience",
    "Fits reaching 2000-epoch ceiling",
    "Selected-candidate inner folds",
    "Selected-candidate folds reaching ceiling"
  ),
  Value = c(
    as.character(nrow(fold_results)),
    sprintf(
      "%.1f",
      median(
        fold_results$EarlyStop_epoch,
        na.rm = TRUE
      )
    ),
    sprintf(
      "%.1f",
      median(
        fold_results$Last_epoch,
        na.rm = TRUE
      )
    ),
    sprintf(
      "%d / %d",
      sum(
        fold_results$Stopped_by_patience,
        na.rm = TRUE
      ),
      nrow(fold_results)
    ),
    sprintf(
      "%d / %d",
      sum(
        fold_results$Hit_epoch_ceiling,
        na.rm = TRUE
      ),
      nrow(fold_results)
    ),
    as.character(nrow(selected_keys)),
    sprintf(
      "%d / %d",
      sum(
        selected_keys$Hit_epoch_ceiling,
        na.rm = TRUE
      ),
      nrow(selected_keys)
    )
  ),
  stringsAsFactors = FALSE
)

format_table(
  optimization_summary,
  caption =
    "Canonical ANN tuning optimization diagnostics",
  row.names = FALSE
)
```

Of the `r nrow(fold_results)` inner candidate-fold fits,
`r sum(fold_results$Stopped_by_patience, na.rm = TRUE)` stopped by patience and
`r sum(fold_results$Hit_epoch_ceiling, na.rm = TRUE)` reached the 2000-epoch
ceiling. The median best epoch shows where the best early-validation checkpoint
typically occurred; these diagnostics are not used to rank candidates.

### 12.4 Search-boundary diagnostics

Boundary occupancy is reported descriptively because `64-32` and the extreme
learning-rate and penalty values are deliberate limits of the prespecified
search space. Reaching one of these limits identifies where the bounded search
was active; it does not trigger post-hoc enlargement of the canonical grid.

These frequencies are therefore interpreted as search-space diagnostics and as
a limitation of the prespecified comparison, not as a new tuning rule.

```{r boundary-description}
task_level <- unique(
  selected_configurations[
    ,
    c(
      "TaskID",
      "Scenario",
      "Architecture",
      "Learning_rate"
    )
  ]
)

boundary_table <- data.frame(
  Boundary = c(
    "Architecture = 64-32",
    "Learning rate = 1.25e-4",
    "ANN_L1: L1 = 1e-1",
    "ANN_L1_L2: L1 = 1e-1",
    "ANN_L2: L2 = 1e-2",
    "ANN_L1_L2: L2 = 1e-2"
  ),
  Selected = c(
    sum(
      task_level$Architecture ==
        "64-32"
    ),
    sum(
      abs(
        task_level$Learning_rate -
          min(LEARNING_RATES)
      ) < 1e-14
    ),
    sum(
      selected_configurations$Model ==
        "ANN_L1" &
      abs(
        selected_configurations$L1 -
          max(L1_VALUES)
      ) < 1e-14
    ),
    sum(
      selected_configurations$Model ==
        "ANN_L1_L2" &
      abs(
        selected_configurations$L1 -
          max(L1_VALUES)
      ) < 1e-14
    ),
    sum(
      selected_configurations$Model ==
        "ANN_L2" &
      abs(
        selected_configurations$L2 -
          max(L2_VALUES)
      ) < 1e-14
    ),
    sum(
      selected_configurations$Model ==
        "ANN_L1_L2" &
      abs(
        selected_configurations$L2 -
          max(L2_VALUES)
      ) < 1e-14
    )
  ),
  Denominator = c(
    150L,
    150L,
    150L,
    150L,
    150L,
    150L
  ),
  stringsAsFactors = FALSE
)

boundary_table$Proportion <-
  boundary_table$Selected /
  boundary_table$Denominator

format_table(
  boundary_table,
  digits = 4,
  caption =
    "Descriptive canonical tuning-boundary occupancy",
  row.names = FALSE
)
```

In the canonical run, the upper architecture `64-32` was selected in
`r boundary_table$Selected[boundary_table$Boundary == "Architecture = 64-32"]`
of 150 tasks. The remaining rows show how often the selected learning rate or
penalty coefficient reached the edge of its prespecified grid. These counts are
reported as search-space diagnostics, not as a reason to enlarge the grid after
observing the results.

## 13. Final model fitting

Nested tuning returns one architecture, learning rate, and applicable penalty
coefficient(s) for each `Scenario × outer split × ANN variant`. Final fitting
then uses those choices without reopening the hyperparameter search.

The final stage follows one fixed sequence:

1. take the selected hyperparameters from nested tuning;
2. use only the 800 outer-analysis individuals to select `Best_epoch`;
3. allow at most 3000 epochs during this epoch-selection fit;
4. reinitialize the network;
5. refit all 800 outer-analysis individuals for exactly `Best_epoch`;
6. predict the untouched 200 outer-assessment individuals.

The 3000-epoch value is a **safety ceiling**, not the final training length.
Outer-assessment phenotypes and true genetic values are excluded from both
hyperparameter selection and `Best_epoch` selection.

### 13.1 Final fitting contract

The contract below makes the separation between tuning, epoch selection, final
refitting, and held-out assessment explicit.

```{r final-ann-contract}
FINAL_EPOCH_MAX_EPOCHS <- 3000L

FINAL_OUTPUT_STEM <- project_path(
  "output",
  "simulated",
  "ann",
  "final",
  "simulated_ann"
)

FINAL_RESULTS_RDS <-
  paste0(
    FINAL_OUTPUT_STEM,
    "_results.rds"
  )
FINAL_PREDICTIONS_RDS <-
  paste0(
    FINAL_OUTPUT_STEM,
    "_predictions.rds"
  )
FINAL_EPOCH_RDS <-
  paste0(
    FINAL_OUTPUT_STEM,
    "_epoch_selection.rds"
  )
FINAL_RUN_RDS <-
  paste0(
    FINAL_OUTPUT_STEM,
    "_run_metadata.rds"
  )

FINAL_RESULTS_CSV <-
  paste0(
    FINAL_OUTPUT_STEM,
    "_results.csv"
  )
FINAL_PREDICTIONS_CSV <-
  paste0(
    FINAL_OUTPUT_STEM,
    "_predictions.csv"
  )
FINAL_EPOCH_CSV <-
  paste0(
    FINAL_OUTPUT_STEM,
    "_epoch_selection.csv"
  )

FINAL_CONFIG_ID <- paste0(
  "ANN_FINAL_CANONICAL_V2__FROM__",
  ANN_TUNING_CONFIG_ID,
  "__FINAL_EPOCH_CAP_",
  FINAL_EPOCH_MAX_EPOCHS,
  "__BEST_EPOCH_REFIT"
)

final_contract <- data.frame(
  Item = c(
    "Tuning source",
    "Epoch ceiling",
    "Final epoch rule",
    "Outer assessment in tuning",
    "Outer assessment in epoch selection",
    "True G in selection"
  ),
  Value = c(
    "ANN_TUNING_CANONICAL_V1",
    FINAL_EPOCH_MAX_EPOCHS,
    "Best_epoch selected inside outer-analysis; refit all 800",
    FALSE,
    FALSE,
    FALSE
  ),
  stringsAsFactors = FALSE
)

format_table(
  final_contract,
  caption =
    "Canonical final ANN contract",
  row.names = FALSE
)
```

The three `FALSE` entries are the key leakage controls: the outer assessment
set is excluded from tuning and epoch selection, and true genetic values are
used only after prediction for evaluation. `Best_epoch` is therefore selected
without access to either final assessment target.

### 13.2 Best-epoch selection

`select_final_epoch()` uses only the current 800-individual outer-analysis set.
That set is divided deterministically into fitting and early-validation
subsets. Scaling is fitted on the fitting subset, and early stopping is allowed
to search up to `FINAL_EPOCH_MAX_EPOCHS = 3000`.

The best early-validation checkpoint determines `Best_epoch`. The outer
assessment set is not used in this decision.

```{r final-epoch-selection-functions}
final_bias <- function(
  predicted,
  observed
) {
  mean(
    predicted - observed
  )
}

final_slope <- function(
  predicted,
  observed
) {
  as.numeric(
    coef(
      lm(
        observed ~ predicted
      )
    )[2]
  )
}

final_r2 <- function(
  predicted,
  observed
) {
  total_variation <- sum(
    (
      observed -
      mean(observed)
    )^2
  )

  1 -
    sum(
      (
        observed -
        predicted
      )^2
    ) /
    total_variation
}

final_metrics <- function(
  predicted,
  observed,
  prefix
) {
  values <- c(
    correlation(predicted, observed),
    rmse(predicted, observed),
    mae(predicted, observed),
    final_bias(predicted, observed),
    final_slope(predicted, observed),
    final_r2(predicted, observed)
  )

  names(values) <- paste0(
    prefix,
    c(
      "_Correlation",
      "_RMSE",
      "_MAE",
      "_Bias",
      "_Slope",
      "_R2"
    )
  )

  values
}

select_final_epoch <- function(
  analysis_ids,
  scenario_index,
  architecture,
  learning_rate,
  l1,
  l2,
  split_seed,
  fit_seed
) {
  validation_split <-
    make_validation_split(
      analysis_ids,
      EARLY_VALIDATION_FRACTION,
      split_seed
    )

  fit_index <- match(
    validation_split$train,
    ids
  )
  validation_index <- match(
    validation_split$validation,
    ids
  )

  x_scaler <- fit_x_scaler(
    geno[
      fit_index,
      ,
      drop = FALSE
    ]
  )
  y_scaler <- fit_y_scaler(
    phenotype[
      fit_index,
      scenario_index
    ]
  )

  x_fit <- scale_x(
    geno[
      fit_index,
      ,
      drop = FALSE
    ],
    x_scaler
  )
  x_validation <- scale_x(
    geno[
      validation_index,
      ,
      drop = FALSE
    ],
    x_scaler
  )
  y_fit <- scale_y(
    phenotype[
      fit_index,
      scenario_index
    ],
    y_scaler
  )
  y_validation <- scale_y(
    phenotype[
      validation_index,
      scenario_index
    ],
    y_scaler
  )

  set_ann_seed(fit_seed)

  x_fit_tensor <-
    torch::torch_tensor(
      x_fit,
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)

  y_fit_tensor <-
    torch::torch_tensor(
      matrix(y_fit, ncol = 1),
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)

  x_validation_tensor <-
    torch::torch_tensor(
      x_validation,
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)

  y_validation_tensor <-
    torch::torch_tensor(
      matrix(y_validation, ncol = 1),
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)

  model <- TorchGenomicMLP(
    ncol(x_fit),
    ARCHITECTURES[[architecture]]
  )$to(device = ANN_DEVICE)

  optimizer <- torch::optim_adam(
    params = model$parameters,
    lr = learning_rate,
    weight_decay = 0
  )

  best_validation_loss <- Inf
  best_epoch <- 1L
  best_state <- clone_state(model)
  stale_epochs <- 0L
  last_epoch <- 0L
  stopped_by_patience <- FALSE

  fit_start <-
    proc.time()[["elapsed"]]

  for (
    epoch in seq_len(
      FINAL_EPOCH_MAX_EPOCHS
    )
  ) {
    last_epoch <- epoch

    model$train()
    optimizer$zero_grad()

    prediction <-
      model(x_fit_tensor)

    loss <-
      torch::nnf_mse_loss(
        prediction,
        y_fit_tensor
      ) +
      weight_penalty(
        model,
        l1,
        l2
      )

    loss$backward()
    optimizer$step()

    model$eval()

    validation_loss <-
      torch::with_no_grad({
        torch::nnf_mse_loss(
          model(x_validation_tensor),
          y_validation_tensor
        )
      })

    validation_value <-
      as.numeric(
        validation_loss$to(
          device =
            torch::torch_device("cpu")
        )
      )

    improved <-
      validation_value <
      best_validation_loss -
      MIN_DELTA

    if (improved) {
      best_validation_loss <-
        validation_value
      best_epoch <- epoch
      best_state <-
        clone_state(model)
      stale_epochs <- 0L
    } else {
      stale_epochs <-
        stale_epochs + 1L
    }

    if (stale_epochs >= PATIENCE) {
      stopped_by_patience <- TRUE
      break
    }
  }

  fit_seconds <-
    proc.time()[["elapsed"]] -
    fit_start

  model$load_state_dict(
    best_state
  )
  model$eval()

  prediction_scaled <-
    torch::with_no_grad({
      as.numeric(
        model(
          x_validation_tensor
        )$to(
          device =
            torch::torch_device("cpu")
        )
      )
    })

  prediction <-
    unscale_y(
      prediction_scaled,
      y_scaler
    )
  observed <-
    phenotype[
      validation_index,
      scenario_index
    ]

  list(
    best_epoch =
      as.integer(best_epoch),
    last_epoch =
      as.integer(last_epoch),
    stopped_by_patience =
      stopped_by_patience,
    hit_epoch_ceiling =
      last_epoch >=
      FINAL_EPOCH_MAX_EPOCHS &&
      !stopped_by_patience,
    validation_rmse =
      rmse(prediction, observed),
    validation_correlation =
      correlation(
        prediction,
        observed
      ),
    fit_seconds = fit_seconds,
    n_fit =
      length(
        validation_split$train
      ),
    n_validation =
      length(
        validation_split$validation
      )
  )
}
```

### 13.3 Refit on all outer-analysis individuals

After `Best_epoch` has been selected, `train_final_fixed()` reinitializes the
network and re-estimates marker and phenotype scaling from **all 800
outer-analysis individuals**. The network is then trained for exactly the
selected `Best_epoch`.

No early stopping is applied during this refit: the selected epoch count is
already fixed before the 200 outer-assessment individuals are predicted.

```{r final-fixed-refit-function}
train_final_fixed <- function(
  train_ids,
  assessment_ids,
  scenario_index,
  architecture,
  learning_rate,
  l1,
  l2,
  epochs,
  seed
) {
  train_index <- match(
    train_ids,
    ids
  )
  assessment_index <- match(
    assessment_ids,
    ids
  )

  x_scaler <- fit_x_scaler(
    geno[
      train_index,
      ,
      drop = FALSE
    ]
  )
  y_scaler <- fit_y_scaler(
    phenotype[
      train_index,
      scenario_index
    ]
  )

  x_train <- scale_x(
    geno[
      train_index,
      ,
      drop = FALSE
    ],
    x_scaler
  )
  x_assessment <- scale_x(
    geno[
      assessment_index,
      ,
      drop = FALSE
    ],
    x_scaler
  )
  y_train <- scale_y(
    phenotype[
      train_index,
      scenario_index
    ],
    y_scaler
  )

  set_ann_seed(seed)

  x_train_tensor <-
    torch::torch_tensor(
      x_train,
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)

  y_train_tensor <-
    torch::torch_tensor(
      matrix(y_train, ncol = 1),
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)

  x_assessment_tensor <-
    torch::torch_tensor(
      x_assessment,
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)

  model <- TorchGenomicMLP(
    ncol(x_train),
    ARCHITECTURES[[architecture]]
  )$to(device = ANN_DEVICE)

  optimizer <- torch::optim_adam(
    params = model$parameters,
    lr = learning_rate,
    weight_decay = 0
  )

  fit_start <-
    proc.time()[["elapsed"]]

  for (
    epoch in seq_len(
      as.integer(epochs)
    )
  ) {
    model$train()
    optimizer$zero_grad()

    prediction <-
      model(x_train_tensor)

    loss <-
      torch::nnf_mse_loss(
        prediction,
        y_train_tensor
      ) +
      weight_penalty(
        model,
        l1,
        l2
      )

    loss$backward()
    optimizer$step()
  }

  fit_seconds <-
    proc.time()[["elapsed"]] -
    fit_start

  model$eval()

  prediction_start <-
    proc.time()[["elapsed"]]

  prediction_scaled <-
    torch::with_no_grad({
      as.numeric(
        model(
          x_assessment_tensor
        )$to(
          device =
            torch::torch_device("cpu")
        )
      )
    })

  prediction_seconds <-
    proc.time()[["elapsed"]] -
    prediction_start

  list(
    prediction = unscale_y(
      prediction_scaled,
      y_scaler
    ),
    fit_seconds =
      fit_seconds,
    prediction_seconds =
      prediction_seconds,
    zero_variance_markers =
      sum(x_scaler$zero)
  )
}
```

### 13.4 Held-out predictions

The 150 final tasks reuse the same frozen outer splits and task identifiers used
during tuning. Each task is fitted for all four ANN variants, producing
**600 model-by-task fits**. Every fitted model predicts the same 200 untouched
outer-assessment individuals for its task.

```{r final-ann-tasks}
outer_map <- unique(
  cv$outer_assignments[
    ,
    c(
      "Repetition",
      "OuterFold",
      "SplitID"
    )
  ]
)

outer_map <- outer_map[
  order(
    outer_map$Repetition,
    outer_map$OuterFold
  ),
  ,
  drop = FALSE
]

final_task_table <- merge(
  data.frame(
    Scenario = scenario_names,
    stringsAsFactors = FALSE
  ),
  outer_map,
  by = NULL
)

final_task_table <- final_task_table[
  order(
    match(
      final_task_table$Scenario,
      scenario_names
    ),
    final_task_table$Repetition,
    final_task_table$OuterFold
  ),
  ,
  drop = FALSE
]

final_task_table$Fold <-
  final_task_table$OuterFold
final_task_table$TaskID <- paste(
  final_task_table$Scenario,
  final_task_table$SplitID,
  sep = "_"
)

rownames(final_task_table) <- NULL

format_table(
  head(final_task_table, 15),
  caption =
    "First 15 canonical ANN final tasks",
  row.names = FALSE
)
```

Each row is one `Scenario × outer split` final-prediction task. The four ANN
variants reuse the same `TaskID`. The final loop below first selects
`Best_epoch`, then refits the corresponding model on all 800 outer-analysis
individuals, and only afterward predicts the 200 outer-assessment individuals.

True genetic values and observed phenotypes from the outer assessment are
accessed only after prediction, when the held-out evaluation metrics are
constructed.

```{r run-final-ann, eval=RUN_ANN}
dir.create(
  dirname(FINAL_RESULTS_RDS),
  recursive = TRUE,
  showWarnings = FALSE
)

true_g <-
  common$true_genetic_value
control <-
  common$scenario_control

run_start <-
  proc.time()[["elapsed"]]

task_outputs <- vector(
  "list",
  nrow(final_task_table)
)

for (
  task_index in seq_len(
    nrow(final_task_table)
  )
) {
  task <- final_task_table[
    task_index,
    ,
    drop = FALSE
  ]

  scenario_index <- match(
    task$Scenario,
    scenario_names
  )
  outer <-
    cv$outer_splits[[task$SplitID]]

  selected_for_task <-
    selected_configurations[
      selected_configurations$TaskID ==
        task$TaskID,
      ,
      drop = FALSE
    ]

  seed_base <-
    SEED_GLOBAL +
    scenario_index * 10000000L +
    task$Repetition * 100000L +
    task$OuterFold * 10000L

  split_seed <-
    seed_base + 8000000L
  epoch_seed <-
    seed_base + 8100000L
  final_fit_seed <-
    seed_base + 8200000L

  result_rows <- vector(
    "list",
    length(MODEL_LEVELS)
  )
  prediction_rows <- vector(
    "list",
    length(MODEL_LEVELS)
  )
  epoch_rows <- vector(
    "list",
    length(MODEL_LEVELS)
  )

  task_start <-
    proc.time()[["elapsed"]]

  for (
    model_index in seq_along(
      MODEL_LEVELS
    )
  ) {
    model_name <-
      MODEL_LEVELS[model_index]

    config <-
      selected_for_task[
        selected_for_task$Model ==
          model_name,
        ,
        drop = FALSE
      ]

    epoch_fit <- select_final_epoch(
      analysis_ids =
        outer$analysis_ids,
      scenario_index =
        scenario_index,
      architecture =
        config$Architecture,
      learning_rate =
        config$Learning_rate,
      l1 = config$L1,
      l2 = config$L2,
      split_seed =
        split_seed,
      fit_seed =
        epoch_seed
    )

    final_fit <- train_final_fixed(
      train_ids =
        outer$analysis_ids,
      assessment_ids =
        outer$assessment_ids,
      scenario_index =
        scenario_index,
      architecture =
        config$Architecture,
      learning_rate =
        config$Learning_rate,
      l1 = config$L1,
      l2 = config$L2,
      epochs =
        epoch_fit$best_epoch,
      seed =
        final_fit_seed
    )

    assessment_index <- match(
      outer$assessment_ids,
      ids
    )

    observed <-
      phenotype[
        assessment_index,
        scenario_index
      ]
    true_genetic_value <-
      true_g[
        assessment_index,
        scenario_index
      ]
    predicted_value <-
      final_fit$prediction

    genetic_metrics <- final_metrics(
      predicted_value,
      true_genetic_value,
      "Genetic"
    )
    phenotype_metrics <- final_metrics(
      predicted_value,
      observed,
      "Phenotype"
    )

    meta <- control[
      control$Scenario ==
        task$Scenario,
      ,
      drop = FALSE
    ]

    result_rows[[model_index]] <- data.frame(
      Dataset = "simulated",
      Scenario = task$Scenario,
      QTL_N = meta$QTL_N,
      H2_nominal = meta$H2_nominal,
      Repetition = task$Repetition,
      Fold = task$OuterFold,
      SplitID = task$SplitID,
      TaskID = task$TaskID,
      Model = model_name,
      Architecture =
        config$Architecture,
      Learning_rate =
        config$Learning_rate,
      L1 = config$L1,
      L2 = config$L2,
      Best_epoch =
        epoch_fit$best_epoch,
      Epoch_Last =
        epoch_fit$last_epoch,
      Epoch_Stopped_by_patience =
        epoch_fit$stopped_by_patience,
      Epoch_Hit_ceiling =
        epoch_fit$hit_epoch_ceiling,
      Inner_RMSE =
        config$Inner_RMSE,
      Inner_Correlation =
        config$Inner_Correlation,
      Inner_MAE =
        config$Inner_MAE,
      N_train = 800L,
      N_test = 200L,
      N_pred_finite =
        sum(
          is.finite(
            predicted_value
          )
        ),
      N_zero_variance_markers =
        final_fit$zero_variance_markers,
      Genetic_Correlation =
        genetic_metrics[
          "Genetic_Correlation"
        ],
      Genetic_RMSE =
        genetic_metrics[
          "Genetic_RMSE"
        ],
      Genetic_MAE =
        genetic_metrics[
          "Genetic_MAE"
        ],
      Genetic_Bias =
        genetic_metrics[
          "Genetic_Bias"
        ],
      Genetic_Slope =
        genetic_metrics[
          "Genetic_Slope"
        ],
      Genetic_R2 =
        genetic_metrics[
          "Genetic_R2"
        ],
      Phenotype_Correlation =
        phenotype_metrics[
          "Phenotype_Correlation"
        ],
      Phenotype_RMSE =
        phenotype_metrics[
          "Phenotype_RMSE"
        ],
      Phenotype_MAE =
        phenotype_metrics[
          "Phenotype_MAE"
        ],
      Phenotype_Bias =
        phenotype_metrics[
          "Phenotype_Bias"
        ],
      Phenotype_Slope =
        phenotype_metrics[
          "Phenotype_Slope"
        ],
      Phenotype_R2 =
        phenotype_metrics[
          "Phenotype_R2"
        ],
      Epoch_selection_seconds =
        epoch_fit$fit_seconds,
      Final_fit_seconds =
        final_fit$fit_seconds,
      Predict_seconds =
        final_fit$prediction_seconds,
      Status = "ok",
      Error = NA_character_,
      stringsAsFactors = FALSE
    )

    prediction_rows[[model_index]] <- data.frame(
      Dataset = "simulated",
      Scenario = task$Scenario,
      QTL_N = meta$QTL_N,
      H2_nominal = meta$H2_nominal,
      Repetition = task$Repetition,
      Fold = task$OuterFold,
      SplitID = task$SplitID,
      TaskID = task$TaskID,
      Model = model_name,
      Architecture =
        config$Architecture,
      Learning_rate =
        config$Learning_rate,
      L1 = config$L1,
      L2 = config$L2,
      Best_epoch =
        epoch_fit$best_epoch,
      ID = ids[assessment_index],
      Observed_Phenotype =
        observed,
      True_Genetic_Value =
        true_genetic_value,
      Predicted_Value =
        predicted_value,
      Status = "ok",
      stringsAsFactors = FALSE
    )

    epoch_rows[[model_index]] <- data.frame(
      Scenario = task$Scenario,
      Repetition = task$Repetition,
      Fold = task$OuterFold,
      SplitID = task$SplitID,
      TaskID = task$TaskID,
      Model = model_name,
      Architecture =
        config$Architecture,
      Learning_rate =
        config$Learning_rate,
      L1 = config$L1,
      L2 = config$L2,
      Best_epoch =
        epoch_fit$best_epoch,
      Last_epoch =
        epoch_fit$last_epoch,
      Stopped_by_patience =
        epoch_fit$stopped_by_patience,
      Hit_epoch_ceiling =
        epoch_fit$hit_epoch_ceiling,
      Validation_RMSE =
        epoch_fit$validation_rmse,
      Validation_Correlation =
        epoch_fit$validation_correlation,
      N_epoch_fit =
        epoch_fit$n_fit,
      N_epoch_validation =
        epoch_fit$n_validation,
      Fit_seconds =
        epoch_fit$fit_seconds,
      stringsAsFactors = FALSE
    )
  }

  task_outputs[[task_index]] <- list(
    results = do.call(
      rbind,
      result_rows
    ),
    predictions = do.call(
      rbind,
      prediction_rows
    ),
    epoch_selection = do.call(
      rbind,
      epoch_rows
    ),
    task_seconds =
      proc.time()[["elapsed"]] -
      task_start
  )

  cat(
    sprintf(
      "[%03d/%03d] %s | final %.1f min\n",
      task_index,
      nrow(final_task_table),
      task$TaskID,
      task_outputs[[task_index]]$task_seconds / 60
    )
  )
}

results <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(task_output) {
      task_output$results
    }
  )
)
predictions <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(task_output) {
      task_output$predictions
    }
  )
)
epoch_selection <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(task_output) {
      task_output$epoch_selection
    }
  )
)

rownames(results) <- NULL
rownames(predictions) <- NULL
rownames(epoch_selection) <- NULL

run_seconds <-
  proc.time()[["elapsed"]] -
  run_start

task_seconds_total <- sum(
  vapply(
    task_outputs,
    function(task_output) {
      task_output$task_seconds
    },
    numeric(1)
  )
)

final_meta <- list(
  dataset = "simulated",
  analysis =
    "canonical_simulated_ann",
  mode = "full",
  pilot_task = NA_character_,
  config_id = FINAL_CONFIG_ID,
  tuning_config_id =
    ANN_TUNING_CONFIG_ID,
  seed_global = SEED_GLOBAL,
  models = MODEL_LEVELS,
  tuning_max_epochs =
    MAX_EPOCHS,
  final_epoch_max_epochs =
    FINAL_EPOCH_MAX_EPOCHS,
  patience = PATIENCE,
  min_delta = MIN_DELTA,
  early_validation_fraction =
    EARLY_VALIDATION_FRACTION,
  final_refit_rule =
    "Best_epoch selected inside outer-analysis, then refit on all 800",
  outer_assessment_used_in_tuning =
    FALSE,
  outer_assessment_used_in_epoch_selection =
    FALSE,
  true_genetic_value_used_in_selection =
    FALSE,
  number_tasks =
    nrow(final_task_table),
  model_task_rows =
    nrow(results),
  prediction_rows =
    nrow(predictions),
  epoch_selection_rows =
    nrow(epoch_selection),
  torch_version =
    as.character(
      packageVersion("torch")
    ),
  device = ANN_DEVICE_LABEL,
  cuda_available =
    cuda_available,
  current_session_seconds =
    run_seconds,
  checkpoint_task_seconds =
    task_seconds_total,
  timestamp =
    as.character(Sys.time())
)

saveRDS(
  results,
  FINAL_RESULTS_RDS,
  compress = "gzip"
)
saveRDS(
  predictions,
  FINAL_PREDICTIONS_RDS,
  compress = "gzip"
)
saveRDS(
  epoch_selection,
  FINAL_EPOCH_RDS,
  compress = "gzip"
)
saveRDS(
  final_meta,
  FINAL_RUN_RDS,
  compress = "gzip"
)

write.csv(
  results,
  FINAL_RESULTS_CSV,
  row.names = FALSE
)
write.csv(
  predictions,
  FINAL_PREDICTIONS_CSV,
  row.names = FALSE
)
write.csv(
  epoch_selection,
  FINAL_EPOCH_CSV,
  row.names = FALSE
)
```

# Canonical final results

The final performance tables use the canonical 600 model-task results,
120,000 held-out predictions, and 600 epoch-selection records. These saved
objects are loaded and checked internally before the reader-facing summaries are
constructed.

```{r final-saved-results}
results <- readRDS(FINAL_RESULTS_RDS)
predictions <- readRDS(FINAL_PREDICTIONS_RDS)
epoch_selection <- readRDS(FINAL_EPOCH_RDS)
final_meta <- readRDS(FINAL_RUN_RDS)

final_output_summary <- data.frame(
  Object = c("Model-task results", "Held-out predictions", "Epoch selections"),
  Rows = c(nrow(results), nrow(predictions), nrow(epoch_selection)),
  Unique_tasks = c(
    length(unique(results$TaskID)),
    length(unique(predictions$TaskID)),
    length(unique(epoch_selection$TaskID))
  )
)
knitr::kable(
  final_output_summary,
  caption = "Observed size of the canonical final ANN outputs",
  row.names = FALSE
)
cat("\nFinal Best_epoch summary:\n")
print(summary(epoch_selection$Best_epoch))
cat(
  "\nFinal epoch-selection fits reaching the",
  FINAL_EPOCH_MAX_EPOCHS,
  "epoch ceiling:",
  sum(epoch_selection$Hit_epoch_ceiling, na.rm = TRUE),
  "/", nrow(epoch_selection), "\n"
)
```

Across the `r nrow(epoch_selection)` final epoch-selection fits,
`r sum(epoch_selection$Stopped_by_patience, na.rm = TRUE)` stopped by patience
and `r sum(epoch_selection$Hit_epoch_ceiling, na.rm = TRUE)` reached the
`r FINAL_EPOCH_MAX_EPOCHS`-epoch ceiling. Thus, the safety ceiling did not define
a common training length; each final network was subsequently refit for its own
selected `Best_epoch`.


# Final ANN predictive performance

Final performance is summarized separately for the two evaluation targets.
Recovery of the known **true genetic value** is the primary simulated-data
criterion; prediction of the **observed phenotype** is secondary. Both tables
average the same 25 frozen outer splits for each scenario and ANN variant.

```{r final-ann-performance}
ok <- results[
  results$Status == "ok",
  ,
  drop = FALSE
]

perf <- aggregate(
  cbind(
    Genetic_Correlation,
    Genetic_RMSE,
    Genetic_MAE,
    Genetic_Bias,
    Genetic_Slope,
    Genetic_R2,
    Phenotype_Correlation,
    Phenotype_RMSE,
    Phenotype_MAE,
    Phenotype_Bias,
    Phenotype_Slope,
    Phenotype_R2
  ) ~ Scenario + Model + QTL_N + H2_nominal,
  data = ok,
  FUN = mean,
  na.rm = TRUE
)

perf <- perf[
  order(
    match(
      perf$Scenario,
      scenario_names
    ),
    match(
      perf$Model,
      MODEL_LEVELS
    )
  ),
  ,
  drop = FALSE
]

rownames(perf) <- NULL

knitr::kable(
  perf[
    ,
    c(
      "Scenario",
      "Model",
      "QTL_N",
      "H2_nominal",
      "Genetic_Correlation",
      "Genetic_RMSE",
      "Genetic_MAE",
      "Genetic_Bias",
      "Genetic_Slope",
      "Genetic_R2"
    )
  ],
  digits = 4,
  caption =
    "Primary evaluation: mean outer-CV ANN performance against true genetic value",
  table.attr = WIDE_TABLE_ATTR,
  row.names = FALSE
)

knitr::kable(
  perf[
    ,
    c(
      "Scenario",
      "Model",
      "QTL_N",
      "H2_nominal",
      "Phenotype_Correlation",
      "Phenotype_RMSE",
      "Phenotype_MAE",
      "Phenotype_Bias",
      "Phenotype_Slope",
      "Phenotype_R2"
    )
  ],
  digits = 4,
  caption =
    "Secondary evaluation: mean outer-CV ANN performance against observed phenotype",
  table.attr = WIDE_TABLE_ATTR,
  row.names = FALSE
)

overall_ann <- aggregate(
  cbind(
    Genetic_Correlation,
    Genetic_RMSE,
    Genetic_R2
  ) ~ Model,
  data = ok,
  FUN = mean,
  na.rm = TRUE
)

overall_ann <- overall_ann[
  match(
    c(
      "ANN_0",
      "ANN_L1",
      "ANN_L2",
      "ANN_L1_L2"
    ),
    overall_ann$Model
  ),
  ,
  drop = FALSE
]
```

Across all outer-CV tasks, ANN-L1+L2 has the highest mean genetic
correlation
(`r sprintf("%.4f", overall_ann$Genetic_Correlation[4])`) and lowest mean
genetic RMSE
(`r sprintf("%.4f", overall_ann$Genetic_RMSE[4])`) among the four ANN variants.
ANN-L1 is effectively tied (correlation
`r sprintf("%.4f", overall_ann$Genetic_Correlation[2])`; RMSE
`r sprintf("%.4f", overall_ann$Genetic_RMSE[2])`), whereas ANN-L2 remains close
to the unregularized ANN (correlations
`r sprintf("%.4f", overall_ann$Genetic_Correlation[3])` and
`r sprintf("%.4f", overall_ann$Genetic_Correlation[1])`, respectively).
Thus, within the prespecified canonical model/architecture search space, the
observable regularization gain is driven mainly by the L1 component. The formal
comparison with GBLUP-ADE is presented in the next module.

# Navigation

[Previous: Validation design](simulated_data_validation.html) | [Parallel branch: GBLUP-ADE](simulated_data_gblup.html) | [Next: Model comparison](simulated_data_comparison.html)

# Reproducibility information

```{r session-info}
sessionInfo()
```
