---
title: "Simulated genomic data: structure and preparation"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

## 1. Purpose

This module introduces the simulated genomic data used throughout the prediction
workflow and prepares the objects required by the later modules. The data were
generated with the **GENES** software, developed for experimental statistics and
quantitative genetics (Cruz, 2013, 2016).

The analysis follows the information in the same order in which a reader needs
it: phenotypic values, true genetic values, genotypes, the genetic map, the
simulation settings, QTL positions, the genetic signal, genotype properties,
parental genotypes, and linkage disequilibrium (LD).

The goal here is to understand the simulated population and create one shared
data object for the prediction analyses. No genomic prediction model is fitted
in this module.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)

options(stringsAsFactors = FALSE)

dir.create(
  "output/simulated/m01/figures",
  showWarnings = FALSE,
  recursive = TRUE
)
```

## 2. Packages and reproducibility

### 2.1 Packages

Only three packages are needed. The comments beside each package indicate its
role in the document.

```{r packages, warning=FALSE, message=FALSE}
library(knitr)     # Format the tables shown to the reader.
library(ggplot2)   # Build all figures in this module.
library(ggthemes)  # Apply the Google Docs theme and color scales.
```

All figures use `theme_gdocs()` together with `scale_fill_gdocs()` or
`scale_color_gdocs()` when a grouping variable is represented by color. A
300-dpi TIFF copy of every figure is also saved in
`output/simulated/m01/figures/`.

### 2.2 Random seed

The simulated data themselves are fixed. A seed is used only when 5,000 marker
pairs from different linkage groups are sampled later to provide an LD
reference.

```{r random-seed}
SEED_GLOBAL <- 20260821L
set.seed(SEED_GLOBAL)
```

## 3. Phenotypic and true genetic values

The first two files contain the quantities that define the prediction target.
Rows correspond to the same 1,000 simulated individuals, while the six columns
correspond to scenarios C1-C6.

### 3.1 Phenotypic values

We first read the phenotypic values. A boxplot is useful here because it shows
the center, spread, and possible extreme values of each scenario immediately
after the file is loaded.

```{r phenotype, fig.cap="Distribution of phenotypic values across the six simulation scenarios."}
# Read the phenotype matrix: individuals in rows and scenarios in columns.
phenotype <- as.matrix(
  read.table(
    "data/dados_simulados/DG#gen_F2_Vfen.txt",
    header = FALSE
  )
)

colnames(phenotype) <- paste0("C", 1:6)

# Convert the matrix to long format only for plotting.
phenotype_long <- stack(as.data.frame(phenotype))
names(phenotype_long) <- c("Phenotype", "Scenario")

p_phenotype <- ggplot(
  phenotype_long,
  aes(
    x = Scenario,
    y = Phenotype,
    fill = Scenario
  )
) +
  geom_boxplot(
    width = 0.65,
    show.legend = FALSE
  ) +
  scale_fill_gdocs() +
  labs(
    x = "Simulation scenario",
    y = "Phenotypic value"
  ) +
  theme_gdocs()

print(p_phenotype)

ggsave(
  "output/simulated/m01/figures/phenotype_distribution.tiff",
  plot = p_phenotype,
  device = "tiff",
  width = 8,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The file contains `r nrow(phenotype)` individuals and
`r ncol(phenotype)` scenarios. These phenotypic values are the responses
available to the prediction models within each training set.

### 3.2 True genetic values

The true genetic values are read separately because they play a different role.
They are not used to create validation partitions or to fit the prediction
models. Instead, they provide the known simulated genetic signal against which
predictions can later be evaluated.

```{r true-genetic-values, fig.cap="Distribution of true genetic values across the six simulation scenarios."}
# Read the true simulated genetic values.
true_genetic_value <- as.matrix(
  read.table(
    "data/dados_simulados/DG#gen_F2_Vgen.dat",
    header = FALSE
  )
)

colnames(true_genetic_value) <- paste0("C", 1:6)

# Reshape only for the reader-facing figure.
true_g_long <- stack(as.data.frame(true_genetic_value))
names(true_g_long) <- c("True_genetic_value", "Scenario")

p_true_g <- ggplot(
  true_g_long,
  aes(
    x = Scenario,
    y = True_genetic_value,
    fill = Scenario
  )
) +
  geom_boxplot(
    width = 0.65,
    show.legend = FALSE
  ) +
  scale_fill_gdocs() +
  labs(
    x = "Simulation scenario",
    y = "True genetic value"
  ) +
  theme_gdocs()

print(p_true_g)

ggsave(
  "output/simulated/m01/figures/true_genetic_value_distribution.tiff",
  plot = p_true_g,
  device = "tiff",
  width = 8,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The matrix has the same `r nrow(true_genetic_value)` individuals and six
scenarios as the phenotype matrix.

## 4. Genotypic data

### 4.1 Genotype matrix

The GENES genotype file stores markers in rows and individuals in columns. We
transpose it once so that each row represents one individual, which is the
orientation used by the later prediction modules.

A boxplot is not informative for a marker matrix because the values are discrete
genotype calls. Instead, the table below shows an actual sample of individuals
and markers. The selected markers are distributed across the complete matrix so
that the reader can see the original `0/1/2` coding directly.

```{r genotype-sample}
# Read the marker matrix and orient it as individuals x markers.
geno_012 <- t(
  as.matrix(
    read.table(
      "data/dados_simulados/genoma_mapa.txt",
      header = FALSE
    )
  )
)

# Select twelve markers spread across the 4,010-marker matrix.
marker_sample <- unique(
  round(
    seq(
      1,
      ncol(geno_012),
      length.out = 12
    )
  )
)

# Display the first ten individuals for those markers.
genotype_sample <- data.frame(
  Individual = 1:10,
  geno_012[
    1:10,
    marker_sample,
    drop = FALSE
  ],
  check.names = FALSE
)

names(genotype_sample)[-1] <- paste0("M", marker_sample)

genotype_sample |>
  knitr::kable(
    format = "html",
    caption = "Sample of the genotype matrix using 10 individuals and 12 markers distributed across the genome",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The complete matrix contains `r nrow(geno_012)` individuals and
`r ncol(geno_012)` markers. The observed genotype states are
`r paste(sort(unique(as.vector(geno_012))), collapse = ", ")`.

### 4.2 Genotype coding used in the prediction models

For model fitting, the same genotype calls are shifted from `0/1/2` to
`-1/0/1`. This changes only their numerical representation; it does not change
the individuals, markers, or genotype states.

```{r genotype-coding}
geno_m101 <- geno_012 - 1L

data.frame(
  Original = c(0, 1, 2),
  Analysis_coding = c(-1, 0, 1)
) |>
  knitr::kable(
    format = "html",
    caption = "Genotype coding used in the prediction analyses",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The transformation is therefore `0 -> -1`, `1 -> 0`, and `2 -> 1`.

## 5. Genetic map

The genetic map is read directly from the GENES text file. The first 12 lines
describe the map as a whole, whereas the marker-level records begin on line 13.
Each marker record identifies its linkage group and its position in
centimorgans (cM).

```{r genetic-map-data}
# Read the marker-level portion of the GENES map file.
genetic_map <- read.table(
  "data/dados_simulados/genoma.txt",
  header = FALSE,
  sep = ",",
  skip = 12,
  fill = TRUE,
  strip.white = TRUE,
  fileEncoding = "latin1",
  stringsAsFactors = FALSE
)

genetic_map <- genetic_map[, 1:5]

names(genetic_map) <- c(
  "LG",
  "Marker_within_LG",
  "Type",
  "Marker",
  "Position_cM"
)

# Keep the variables required for plotting and later QTL matching.
genetic_map$LG <- as.integer(genetic_map$LG)
genetic_map$Marker_within_LG <- as.integer(genetic_map$Marker_within_LG)
genetic_map$Position_cM <- as.numeric(genetic_map$Position_cM)
genetic_map$Global_marker <- seq_len(nrow(genetic_map))
genetic_map$LG_label <- factor(
  paste0("LG", genetic_map$LG),
  levels = paste0("LG", 1:10)
)

map_summary <- data.frame(
  Linkage_group = paste0("LG", 1:10),
  Markers = as.integer(table(genetic_map$LG)),
  Start_cM = as.numeric(tapply(genetic_map$Position_cM, genetic_map$LG, min)),
  End_cM = as.numeric(tapply(genetic_map$Position_cM, genetic_map$LG, max))
)

map_summary |>
  knitr::kable(
    format = "html",
    caption = "Structure of the simulated genetic map",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The map contains 10 linkage groups with 401 markers in each group. Positions
range from 0 to 200 cM with 0.5-cM spacing, giving 4,010 markers in total. The
QTL are added to this map after the scenario file is read in the next section.

## 6. Simulation settings and QTL architecture

### 6.1 GENES software and genetic model

The six scenarios were generated with the **GENES** software (Cruz, 2013,
2016). The control file identifies the simulation model as `Model = 2`. Using
the model expression supplied with the simulation, the phenotype of individual
\(i\) is represented as

\[
Y_i =
\mu +
\sum_j a_j +
\sum_j \sum_{j'} \alpha_j\alpha_{j'} +
\varepsilon_i .
\]

where:

- \(Y_i\) is the simulated phenotype of individual \(i\);
- \(\mu\) is the overall mean;
- \(\sum_j a_j\) represents the contributions associated with loci;
- \(\sum_j\sum_{j'}\alpha_j\alpha_{j'}\) represents pairwise interaction
  contributions between loci;
- \(\varepsilon_i\) is the residual effect.

This equation describes the model used to generate the simulated data. It
should not be confused with the GBLUP-ADE or ANN models fitted in later
modules.

### 6.2 Scenario settings

The GENES control file contains six repeated blocks, one for each scenario.
Because the fields occur in a fixed order, we read the required values directly
from their positions in each block. This keeps the code linear and makes the
connection between the file and the table below explicit.

```{r simulation-scenarios}
# Read the six scenario blocks from the GENES control file.
control_raw <- readLines(
  "data/dados_simulados/controlegenetico#1.dat",
  warn = FALSE,
  encoding = "latin1"
)

# Normalize accented text only so the fixed field positions can be read safely.
control_ascii <- iconv(
  control_raw,
  from = "latin1",
  to = "ASCII//TRANSLIT",
  sub = ""
)

scenario_start <- grep(
  "^Variavel",
  control_ascii
)

control_table <- data.frame(
  Scenario = paste0("C", 1:6),
  Variable = NA_real_,
  Total_loci = NA_real_,
  QTL_N = NA_real_,
  gmd = NA_real_,
  Model = NA_real_,
  H2_nominal = NA_real_,
  Constant = NA_real_,
  Scale = NA_real_,
  CV_setting = NA_real_
)

qtl_lists <- vector(
  "list",
  length(scenario_start)
)

# Recover the values stored in each scenario block.
for (i in seq_along(scenario_start)) {

  s <- scenario_start[i]

  control_table$Variable[i] <- as.numeric(trimws(sub("^[^,]*,", "", control_ascii[s])))
  control_table$Total_loci[i] <- as.numeric(trimws(sub("^[^,]*,", "", control_ascii[s + 1])))
  control_table$QTL_N[i] <- as.numeric(trimws(sub("^[^,]*,", "", control_ascii[s + 2])))

  # QTL entries are marker indices, so they are stored as integers.
  qtl_lists[[i]] <- as.integer(
    scan(
      text = control_ascii[s + 3],
      what = numeric(),
      quiet = TRUE
    )
  )

  control_table$gmd[i] <- as.numeric(trimws(sub("^[^,]*,", "", control_ascii[s + 5])))
  control_table$Model[i] <- as.numeric(trimws(sub("^[^,]*,", "", control_ascii[s + 6])))
  control_table$H2_nominal[i] <- as.numeric(trimws(sub("^[^,]*,", "", control_ascii[s + 7])))
  control_table$Constant[i] <- as.numeric(trimws(sub("^[^,]*,", "", control_ascii[s + 8])))
  control_table$Scale[i] <- as.numeric(trimws(sub("^[^,]*,", "", control_ascii[s + 9])))
  control_table$CV_setting[i] <- as.numeric(trimws(sub("^[^,]*,", "", control_ascii[s + 10])))
}

control_table |>
  knitr::kable(
    format = "html",
    digits = 3,
    caption = "Settings used to generate the six simulation scenarios in GENES",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The number of QTL increases from 8 in C1 to 480 in C6. The specified
heritability is 0.7 in C1-C2, 0.5 in C3-C4, and 0.3 in C5-C6. The values
`gmd = 0.5`, `Model = 2`, `Constant = 100`, `Scale = 1`, and `CV = 12` are
fixed across scenarios.

The six scenarios therefore form a joint complexity gradient: QTL number
increases while the specified heritability decreases across the sequence. These
two characteristics should consequently be interpreted together rather than as
independent factors.

### 6.3 QTL positions on the genetic map

The QTL entries in the control file are marker indices. We match each index to
the previously read genetic map to recover its linkage group and position in
cM. The first figure combines the 4,010-marker map with all QTL positions used
across the six scenarios. Gray marks represent the marker grid, while colored
points identify positions that act as QTL in at least one scenario.

```{r genetic-map-qtl, fig.cap="Genetic map of the 4,010 markers with the simulated QTL positions. Colored points identify positions used as QTL in at least one scenario; LG9 and LG10 contain no simulated QTL."}
# Match every QTL index to its linkage group and map position.
qtl_rows <- data.frame(
  Scenario = character(),
  QTL_index = integer(),
  LG = integer(),
  Position_cM = numeric()
)

for (i in seq_along(qtl_lists)) {

  qtl <- qtl_lists[[i]]

  qtl_rows <- rbind(
    qtl_rows,
    data.frame(
      Scenario = paste0("C", i),
      QTL_index = qtl,
      LG = genetic_map$LG[qtl],
      Position_cM = genetic_map$Position_cM[qtl]
    )
  )
}

qtl_rows$LG_label <- factor(
  paste0("LG", qtl_rows$LG),
  levels = paste0("LG", 1:10)
)

# Keep each QTL position only once for the combined map.
qtl_union <- unique(
  qtl_rows[
    ,
    c("LG", "Position_cM", "LG_label")
  ]
)

map_axes <- data.frame(
  LG_label = factor(
    paste0("LG", 1:10),
    levels = paste0("LG", 1:10)
  ),
  Start = 0,
  End = 200
)

p_genetic_map <- ggplot() +
  geom_segment(
    data = map_axes,
    aes(
      x = LG_label,
      xend = LG_label,
      y = Start,
      yend = End,
      color = LG_label
    ),
    linewidth = 1.3,
    show.legend = FALSE
  ) +
  geom_point(
    data = genetic_map,
    aes(
      x = LG_label,
      y = Position_cM
    ),
    shape = "_",
    size = 1.4,
    color = "grey70"
  ) +
  geom_point(
    data = qtl_union,
    aes(
      x = LG_label,
      y = Position_cM,
      fill = LG_label
    ),
    shape = 21,
    size = 2.2,
    color = "black",
    stroke = 0.25,
    show.legend = FALSE
  ) +
  scale_color_gdocs() +
  scale_fill_gdocs() +
  scale_y_reverse(
    breaks = seq(0, 200, 50)
  ) +
  labs(
    x = "Linkage group",
    y = "Position (cM)"
  ) +
  theme_gdocs()

print(p_genetic_map)

ggsave(
  "output/simulated/m01/figures/genetic_map_with_qtl.tiff",
  plot = p_genetic_map,
  device = "tiff",
  width = 9,
  height = 6,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The absence of colored QTL points in **LG9 and LG10 is intentional**. These two
linkage groups were simulated without QTL so that they provide genomic regions
where a method should not find a true simulated QTL signal. They are therefore
useful reference groups when evaluating whether a methodology prioritizes the
regions that actually contain simulated QTL.

### 6.4 QTL distribution across scenarios

The combined map shows the overall QTL architecture, but it does not show how
the number of QTL changes from C1 to C6. The faceted figure below therefore
displays each scenario separately while retaining the same linkage-group
layout and color scheme.

```{r qtl-map, fig.cap="Positions of the simulated QTL across linkage groups for each scenario. LG9 and LG10 were intentionally simulated without QTL and remain empty in every panel."}
# Create the ten linkage-group axes in every scenario panel.
qtl_reference <- expand.grid(
  Scenario = paste0("C", 1:6),
  LG_label = paste0("LG", 1:10),
  stringsAsFactors = FALSE
)

qtl_reference$LG_label <- factor(
  qtl_reference$LG_label,
  levels = paste0("LG", 1:10)
)

qtl_reference$Start <- 0
qtl_reference$End <- 200

p_qtl <- ggplot() +
  geom_segment(
    data = qtl_reference,
    aes(
      x = LG_label,
      xend = LG_label,
      y = Start,
      yend = End,
      color = LG_label
    ),
    linewidth = 0.8,
    show.legend = FALSE
  ) +
  geom_point(
    data = qtl_rows,
    aes(
      x = LG_label,
      y = Position_cM,
      fill = LG_label
    ),
    shape = 21,
    size = 1.5,
    color = "black",
    stroke = 0.2,
    show.legend = FALSE
  ) +
  facet_wrap(
    ~Scenario,
    ncol = 2
  ) +
  scale_color_gdocs() +
  scale_fill_gdocs() +
  scale_y_reverse(
    breaks = seq(0, 200, 50)
  ) +
  labs(
    x = "Linkage group",
    y = "Position (cM)"
  ) +
  theme_gdocs() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

print(p_qtl)

ggsave(
  "output/simulated/m01/figures/qtl_positions_by_scenario.tiff",
  plot = p_qtl,
  device = "tiff",
  width = 10,
  height = 8,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The increasing point density from C1 to C6 reflects the scenario design rather
than a plotting artifact. LG9 and LG10 remain without QTL in all panels.

```{r qtl-counts}
qtl_table <- table(
  factor(qtl_rows$Scenario, levels = paste0("C", 1:6)),
  factor(qtl_rows$LG, levels = 1:10)
)

qtl_by_lg <- data.frame(
  Scenario = rownames(qtl_table),
  unclass(qtl_table),
  check.names = FALSE,
  row.names = NULL
)

names(qtl_by_lg)[-1] <- paste0("LG", 1:10)

qtl_by_lg |>
  knitr::kable(
    format = "html",
    caption = "Number of simulated QTL per linkage group",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The table provides the exact QTL counts behind the figure and makes the zeros
for LG9 and LG10 explicit.

## 7. Phenotypic and genetic signal

The simulated phenotype can be separated into its known genetic component and
the remaining component:

\[
E = Y - G.
\]

Using these quantities, we calculate the phenotype-genetic correlation,
genetic-residual correlation, realized heritability, and basic summaries for
each scenario. These values describe how much genetic signal is present in the
data that the prediction methods will later analyze.

```{r phenotypic-genetic-signal}
scenario_names <- paste0("C", 1:6)
residual_value <- phenotype - true_genetic_value

cor_p_g <- numeric(6)
cor_g_e <- numeric(6)
h2_realized <- numeric(6)
var_g_over_var_p <- numeric(6)

# Calculate the genetic-signal summaries scenario by scenario.
for (j in 1:6) {

  cor_p_g[j] <- cor(
    phenotype[, j],
    true_genetic_value[, j]
  )

  cor_g_e[j] <- cor(
    true_genetic_value[, j],
    residual_value[, j]
  )

  vg <- var(true_genetic_value[, j])
  ve <- var(residual_value[, j])

  h2_realized[j] <- vg / (vg + ve)
  var_g_over_var_p[j] <- vg / var(phenotype[, j])
}

# Gather the quantities used for interpretation and later saving.
scenario_summary <- data.frame(
  Scenario = scenario_names,
  QTL_N = control_table$QTL_N,
  H2_nominal = control_table$H2_nominal,
  Phenotype_mean = colMeans(phenotype),
  Phenotype_sd = apply(phenotype, 2, sd),
  Genetic_mean = colMeans(true_genetic_value),
  Genetic_sd = apply(true_genetic_value, 2, sd),
  Residual_mean = colMeans(residual_value),
  Residual_sd = apply(residual_value, 2, sd),
  Cor_P_G = cor_p_g,
  Cor_G_E = cor_g_e,
  H2_realized = h2_realized,
  VarG_over_VarP = var_g_over_var_p,
  CV_empirical_pct = 100 * apply(phenotype, 2, sd) / colMeans(phenotype),
  CV_setting = control_table$CV_setting
)

scenario_summary[
  ,
  c(
    "Scenario",
    "QTL_N",
    "H2_nominal",
    "H2_realized",
    "Cor_P_G",
    "Phenotype_mean",
    "Genetic_mean",
    "Residual_sd"
  )
] |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Phenotypic and genetic signal by simulation scenario",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The phenotype-true-genetic-value correlation ranges from
`r sprintf("%.3f", min(scenario_summary$Cor_P_G))` to
`r sprintf("%.3f", max(scenario_summary$Cor_P_G))`, showing that the six
scenarios differ in the strength of the genetic signal available for
prediction.

### 7.1 Specified and realized heritability

The heritability entered in GENES is a simulation setting. The realized
heritability is calculated from the genetic and residual variances observed in
the finite simulated population. Comparing them shows how closely each
generated scenario reflects the value specified during simulation.

```{r heritability-comparison, fig.cap="Heritability specified in GENES and realized heritability in the generated population."}
heritability_plot <- rbind(
  data.frame(
    Scenario = scenario_summary$Scenario,
    Heritability = scenario_summary$H2_nominal,
    Type = "GENES setting"
  ),
  data.frame(
    Scenario = scenario_summary$Scenario,
    Heritability = scenario_summary$H2_realized,
    Type = "Realized"
  )
)

h2_difference <- scenario_summary$H2_realized -
  scenario_summary$H2_nominal

p_heritability <- ggplot(
  heritability_plot,
  aes(
    x = Scenario,
    y = Heritability,
    group = Type,
    color = Type,
    shape = Type
  )
) +
  geom_line(
    linewidth = 0.8
  ) +
  geom_point(
    size = 2.5
  ) +
  scale_color_gdocs() +
  labs(
    x = "Simulation scenario",
    y = expression(H^2),
    color = NULL,
    shape = NULL
  ) +
  theme_gdocs()

print(p_heritability)

ggsave(
  "output/simulated/m01/figures/heritability_specified_vs_realized.tiff",
  plot = p_heritability,
  device = "tiff",
  width = 8,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The absolute differences for C1-C6 are
`r paste(sprintf("%.4f", abs(h2_difference)), collapse = ", ")`. The largest
difference occurs in
`r scenario_summary$Scenario[which.max(abs(h2_difference))]`, with an absolute
difference of `r sprintf("%.4f", max(abs(h2_difference)))`. These differences
describe the generated population and do not redefine the six scenarios.

### 7.2 GENES CV setting and empirical phenotype CV

The control file records `CV = 12` for every scenario. We compare that value
with the coefficient of variation calculated directly from the generated
phenotypic values. This distinction is useful because the value supplied during
simulation does not have to equal the empirical value in one finite simulated
sample.

```{r cv-comparison, fig.cap="CV specified in GENES and empirical coefficient of variation calculated from the simulated phenotypes."}
cv_plot <- rbind(
  data.frame(
    Scenario = scenario_summary$Scenario,
    CV = scenario_summary$CV_setting,
    Type = "GENES setting"
  ),
  data.frame(
    Scenario = scenario_summary$Scenario,
    CV = scenario_summary$CV_empirical_pct,
    Type = "Empirical phenotype"
  )
)

cv_difference <- scenario_summary$CV_empirical_pct -
  scenario_summary$CV_setting

p_cv <- ggplot(
  cv_plot,
  aes(
    x = Scenario,
    y = CV,
    group = Type,
    color = Type,
    shape = Type
  )
) +
  geom_line(
    linewidth = 0.8
  ) +
  geom_point(
    size = 2.5
  ) +
  scale_color_gdocs() +
  labs(
    x = "Simulation scenario",
    y = "Coefficient of variation (%)",
    color = NULL,
    shape = NULL
  ) +
  theme_gdocs()

print(p_cv)

ggsave(
  "output/simulated/m01/figures/cv_genes_vs_empirical.tiff",
  plot = p_cv,
  device = "tiff",
  width = 8,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The empirical phenotype CV values are
`r paste(sprintf("%.2f", scenario_summary$CV_empirical_pct), collapse = ", ")`%.
Their differences from the value entered in GENES are
`r paste(sprintf("%.2f", cv_difference), collapse = ", ")` percentage points.
The value from the control file is therefore treated as a simulation setting,
while the empirical CV is calculated from the generated phenotype matrix.

## 8. Genotype properties

### 8.1 F2 segregation

An F2 population generated from two contrasting homozygous parents is expected
to approach a `1:2:1` distribution for the three genotype classes. We compare
that expectation with all genotype calls in the simulated marker matrix.

```{r genotype-segregation, fig.cap="Observed genotype proportions compared with the expected 1:2:1 segregation of an F2 population."}
# Calculate the overall observed frequencies of genotype classes 0, 1, and 2.
geno_frequency <- prop.table(
  table(
    factor(
      as.vector(geno_012),
      levels = 0:2
    )
  )
)

segregation_plot <- rbind(
  data.frame(
    Genotype = factor(0:2, levels = 0:2),
    Proportion = as.numeric(geno_frequency),
    Type = "Observed"
  ),
  data.frame(
    Genotype = factor(0:2, levels = 0:2),
    Proportion = c(0.25, 0.50, 0.25),
    Type = "F2 expectation"
  )
)

p_segregation <- ggplot(
  segregation_plot,
  aes(
    x = Genotype,
    y = Proportion,
    fill = Type
  )
) +
  geom_col(
    position = "dodge"
  ) +
  scale_fill_gdocs() +
  labs(
    x = "Genotype coding",
    y = "Proportion",
    fill = NULL
  ) +
  theme_gdocs()

print(p_segregation)

ggsave(
  "output/simulated/m01/figures/f2_genotype_proportions.tiff",
  plot = p_segregation,
  device = "tiff",
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The observed proportions can be compared directly with 0.25, 0.50, and 0.25.
This provides a simple population-level description of segregation before the
markers are used by the prediction models.

### 8.2 Marker and individual checks

We next summarize properties that help the reader understand the matrix:
missing calls, repeated genotype profiles, monomorphic markers, minor allele
frequency (MAF), and heterozygosity. These values describe the simulated data;
they are not used here to construct a generic filtering procedure.

```{r genotype-checks}
# Count missing values and repeated rows or columns.
missing_genotypes <- sum(is.na(geno_012))
duplicate_individuals <- sum(duplicated(geno_012, MARGIN = 1))
duplicate_markers <- sum(duplicated(geno_012, MARGIN = 2))

# Calculate allele-frequency and heterozygosity summaries marker by marker.
allele_frequency <- colMeans(geno_012) / 2
maf <- pmin(allele_frequency, 1 - allele_frequency)
heterozygosity <- colMeans(geno_012 == 1)
marker_variance <- apply(geno_012, 2, var)
monomorphic_markers <- sum(marker_variance == 0)

marker_summary <- data.frame(
  Statistic = c(
    "Missing genotype cells",
    "Duplicate individuals",
    "Duplicate markers",
    "Monomorphic markers",
    "Minimum MAF",
    "Median MAF",
    "Maximum MAF",
    "Minimum heterozygosity",
    "Median heterozygosity",
    "Maximum heterozygosity"
  ),
  Value = c(
    missing_genotypes,
    duplicate_individuals,
    duplicate_markers,
    monomorphic_markers,
    min(maf),
    median(maf),
    max(maf),
    min(heterozygosity),
    median(heterozygosity),
    max(heterozygosity)
  )
)

marker_summary |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Summary of marker and individual properties",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The genotype matrix is subsequently passed to the prediction workflow with the
same individuals and markers.

## 9. Parental genotypes

The parental file describes the two contrasting parents used by GENES to
generate the F2 population. We read the parent identifier, linkage group,
number of markers, and marker states so that the population structure can be
connected to its simulated origin.

```{r parental-genotypes}
# Read the two parental genotypes across the ten linkage groups.
parents <- as.matrix(
  read.table(
    "data/dados_simulados/genoma_pais.txt",
    header = FALSE
  )
)

parent_id <- parents[, 1]
parent_lg <- parents[, 2]
parent_nmarkers <- parents[, 3]
parent_genotypes <- parents[, 5:405, drop = FALSE]

# Summarize the linkage groups and fixed state represented by each parent.
parent_summary <- data.frame(
  Parent = c(1, 2),
  Linkage_groups = c(
    length(unique(parent_lg[parent_id == 1])),
    length(unique(parent_lg[parent_id == 2]))
  ),
  Marker_state = c(
    unique(as.vector(parent_genotypes[parent_id == 1, , drop = FALSE]))[1],
    unique(as.vector(parent_genotypes[parent_id == 2, , drop = FALSE]))[1]
  )
)

parent_summary |>
  knitr::kable(
    format = "html",
    caption = "Parental genotypes used to generate the simulated F2 population",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

Both parents occur in all 10 linkage groups. Parent 1 is fixed for marker state
1 and parent 2 for marker state 0, providing the contrasting parental states
from which the F2 genotypes were generated.

## 10. Linkage disequilibrium

Linkage disequilibrium is summarized with \(r^2\), the squared correlation
between marker genotypes. We calculate \(r^2\) for marker pairs from the same
linkage group at several genetic distances and compare those values with marker
pairs drawn from different linkage groups.

### 10.1 LD decay with map distance

The map has a regular spacing of 0.5 cM, so each requested distance can be
translated directly into a marker shift. For example, 10 cM corresponds to 20
marker intervals. Within each linkage group we pair markers separated by the
requested shift, calculate their correlations across the 1,000 individuals,
square the correlations, and summarize the resulting \(r^2\) values.

The evaluated distances range from 0.5 to 200 cM. A conventional continuous
x-axis compresses the short distances, making the first points and labels hard
to read. The figure therefore shows the nine evaluated distances as **ordered
categories with equal visual spacing**. The calculations still use their true
numeric distances in cM.

```{r linkage-disequilibrium, fig.cap="Linkage disequilibrium decay across the evaluated map distances. Distances are shown at equal visual spacing for readability; the dashed line represents the mean r-squared from 5,000 marker pairs belonging to different linkage groups."}
# Standardize each marker once so correlations can be calculated efficiently.
geno_scaled <- scale(geno_012)

ld_distances <- c(
  0.5,
  1,
  2.5,
  5,
  10,
  25,
  50,
  100,
  200
)

ld_summary <- data.frame(
  Distance_cM = ld_distances,
  Pairs = NA_integer_,
  Mean_r2 = NA_real_,
  Median_r2 = NA_real_
)

# Calculate within-linkage-group LD at each requested map distance.
for (d in seq_along(ld_distances)) {

  marker_shift <- as.integer(ld_distances[d] / 0.5)
  r2_all <- numeric()

  for (lg in 1:10) {

    markers <- which(genetic_map$LG == lg)

    marker_1 <- markers[
      1:(length(markers) - marker_shift)
    ]

    marker_2 <- markers[
      (1 + marker_shift):length(markers)
    ]

    r <- colSums(
      geno_scaled[, marker_1, drop = FALSE] *
        geno_scaled[, marker_2, drop = FALSE]
    ) / (nrow(geno_scaled) - 1)

    r2_all <- c(r2_all, r^2)
  }

  ld_summary$Pairs[d] <- length(r2_all)
  ld_summary$Mean_r2[d] <- mean(r2_all)
  ld_summary$Median_r2[d] <- median(r2_all)
}

# Sample marker pairs from different linkage groups for a background reference.
set.seed(SEED_GLOBAL + 101L)

n_inter_pairs <- 5000L

inter_i <- sample(
  seq_len(ncol(geno_012)),
  n_inter_pairs,
  replace = TRUE
)

inter_j <- sample(
  seq_len(ncol(geno_012)),
  n_inter_pairs,
  replace = TRUE
)

same_lg <- genetic_map$LG[inter_i] == genetic_map$LG[inter_j]

while (any(same_lg)) {

  inter_j[same_lg] <- sample(
    seq_len(ncol(geno_012)),
    sum(same_lg),
    replace = TRUE
  )

  same_lg <- genetic_map$LG[inter_i] == genetic_map$LG[inter_j]
}

inter_r <- colSums(
  geno_scaled[, inter_i, drop = FALSE] *
    geno_scaled[, inter_j, drop = FALSE]
) / (nrow(geno_scaled) - 1)

inter_r2 <- inter_r^2

interchromosomal_ld <- data.frame(
  Distance_cM = NA_real_,
  Pairs = length(inter_r2),
  Mean_r2 = mean(inter_r2),
  Median_r2 = median(inter_r2)
)

# Use ordered labels so the short-distance points remain legible.
ld_plot_data <- ld_summary
ld_plot_data$Distance_label <- factor(
  as.character(ld_plot_data$Distance_cM),
  levels = as.character(ld_distances)
)
ld_plot_data$Type <- "Within linkage group"

ld_reference_plot <- data.frame(
  Mean_r2 = interchromosomal_ld$Mean_r2,
  Type = "Between linkage groups"
)

p_ld <- ggplot(
  ld_plot_data,
  aes(
    x = Distance_label,
    y = Mean_r2,
    group = 1,
    color = Type
  )
) +
  geom_line(
    linewidth = 0.8
  ) +
  geom_point(
    size = 2.5
  ) +
  geom_hline(
    data = ld_reference_plot,
    aes(
      yintercept = Mean_r2,
      color = Type
    ),
    linetype = "dashed",
    linewidth = 0.8
  ) +
  scale_color_gdocs() +
  labs(
    x = "Map distance (cM)",
    y = expression(mean(r^2)),
    color = NULL
  ) +
  theme_gdocs()

print(p_ld)

ggsave(
  "output/simulated/m01/figures/ld_decay.tiff",
  plot = p_ld,
  device = "tiff",
  width = 9,
  height = 5.5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The figure emphasizes the observed pattern rather than the unequal numerical
spacing between distances. The exact distances and numerical summaries remain
available in the table below.

```{r linkage-disequilibrium-table}
ld_summary |>
  knitr::kable(
    format = "html",
    digits = 5,
    caption = "Within-linkage-group LD by map distance",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )

interchromosomal_ld |>
  knitr::kable(
    format = "html",
    digits = 5,
    caption = "LD between markers from different linkage groups",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

### 10.2 Interpretation of the LD pattern

The observed LD decreases strongly as map distance increases. Mean \(r^2\)
declines from
`r sprintf("%.4f", ld_summary$Mean_r2[ld_summary$Distance_cM == 0.5])` at
0.5 cM to
`r sprintf("%.4f", ld_summary$Mean_r2[ld_summary$Distance_cM == 50])` at
50 cM.

At longer distances, mean \(r^2\) remains
`r sprintf("%.4f", ld_summary$Mean_r2[ld_summary$Distance_cM == 100])` at
100 cM and
`r sprintf("%.4f", ld_summary$Mean_r2[ld_summary$Distance_cM == 200])` at
200 cM, compared with
`r sprintf("%.5f", interchromosomal_ld$Mean_r2)` for marker pairs from
different linkage groups.

The persistence of greater LD at 100-200 cM than between linkage groups is
therefore a property of the simulated recombination structure. It should be
kept in mind when genomic regions and prediction results are interpreted, but
it is not used here as a reason to remove markers.

## 11. Summary of the simulated data

The main quantities obtained above are collected in one compact table so that
the reader can verify the dimensions and basic properties of the data before
moving to the validation design.

```{r data-summary}
data_overview <- data.frame(
  Quantity = c(
    "Individuals",
    "Markers",
    "Simulation scenarios",
    "Linkage groups",
    "Markers per linkage group",
    "Missing genotype cells",
    "Duplicate individuals",
    "Duplicate markers",
    "Monomorphic markers",
    "Minimum MAF",
    "Mean r2 between linkage groups"
  ),
  Observed = c(
    nrow(geno_012),
    ncol(geno_012),
    ncol(phenotype),
    length(unique(genetic_map$LG)),
    paste(unique(as.integer(table(genetic_map$LG))), collapse = ", "),
    missing_genotypes,
    duplicate_individuals,
    duplicate_markers,
    monomorphic_markers,
    round(min(maf), 4),
    round(interchromosomal_ld$Mean_r2, 5)
  ),
  stringsAsFactors = FALSE
)

data_overview |>
  knitr::kable(
    format = "html",
    caption = "Summary of the simulated genomic data",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The table summarizes what was observed in the simulated population. The
scientific interpretation of heritability, CV, QTL distribution, and LD remains
in the corresponding sections rather than being reduced to generic pass/fail
messages.

```{r data-check-output, include=FALSE}
dim_ok <-
  identical(dim(phenotype), c(1000L, 6L)) &&
  identical(dim(true_genetic_value), c(1000L, 6L)) &&
  identical(dim(geno_012), c(1000L, 4010L))

finite_ok <-
  all(is.finite(phenotype)) &&
  all(is.finite(true_genetic_value))

genotype_codes_ok <- identical(
  as.numeric(sort(unique(as.vector(geno_012)))),
  c(0, 1, 2)
)

analysis_codes_ok <- identical(
  as.numeric(sort(unique(as.vector(geno_m101)))),
  c(-1, 0, 1)
)

map_ok <-
  nrow(genetic_map) == 4010L &&
  length(unique(genetic_map$LG)) == 10L &&
  all(table(genetic_map$LG) == 401L) &&
  all(tapply(genetic_map$Position_cM, genetic_map$LG, min) == 0) &&
  all(tapply(genetic_map$Position_cM, genetic_map$LG, max) == 200)

scenario_ok <-
  identical(
    as.numeric(control_table$QTL_N),
    c(8, 40, 80, 120, 240, 480)
  ) &&
  isTRUE(
    all.equal(
      as.numeric(control_table$H2_nominal),
      c(0.7, 0.7, 0.5, 0.5, 0.3, 0.3),
      tolerance = 1e-12
    )
  ) &&
  all(control_table$Model == 2)

qtl_indices_ok <- all(
  qtl_rows$QTL_index >= 1 &
    qtl_rows$QTL_index <= nrow(genetic_map)
)

reference_lg_ok <- all(
  qtl_by_lg[, c("LG9", "LG10")] == 0
)

row_alignment_ok <-
  nrow(phenotype) == nrow(true_genetic_value) &&
  nrow(phenotype) == nrow(geno_012)

parent_structure_ok <-
  nrow(parents) == 20L &&
  ncol(parents) == 405L &&
  all(parent_genotypes[parent_id == 1, , drop = FALSE] == 1) &&
  all(parent_genotypes[parent_id == 2, , drop = FALSE] == 0) &&
  all(table(parent_lg) == 2L) &&
  all(parent_nmarkers == 401)

f2_frequency_ok <- max(
  abs(
    as.numeric(geno_frequency) -
      c(0.25, 0.50, 0.25)
  )
) < 0.02

maf_ok <- min(maf) > 0.40

short_ld <- ld_summary$Mean_r2[
  ld_summary$Distance_cM == 0.5
]

medium_ld <- ld_summary$Mean_r2[
  ld_summary$Distance_cM == 10
]

ld_50 <- ld_summary$Mean_r2[
  ld_summary$Distance_cM == 50
]

ld_decay_ok <-
  short_ld > medium_ld &&
  medium_ld > ld_50 &&
  ld_50 > interchromosomal_ld$Mean_r2

data_check_log <- data.frame(
  Check = c(
    "Primary matrix dimensions",
    "Finite phenotype and true genetic values",
    "Genotype coding 0/1/2",
    "Analysis genotype coding -1/0/1",
    "Genetic map structure",
    "Simulation scenario parameters",
    "QTL indices within genetic map",
    "LG9 and LG10 without simulated QTL",
    "Row alignment",
    "Missing genotypes",
    "Duplicate individuals",
    "Duplicate markers",
    "Monomorphic markers",
    "F2 genotype proportions",
    "Minor allele frequency",
    "Parental structure",
    "Short-to-medium distance LD decay"
  ),
  Status = c(
    ifelse(dim_ok, "PASS", "BLOCKED"),
    ifelse(finite_ok, "PASS", "BLOCKED"),
    ifelse(genotype_codes_ok, "PASS", "BLOCKED"),
    ifelse(analysis_codes_ok, "PASS", "BLOCKED"),
    ifelse(map_ok, "PASS", "BLOCKED"),
    ifelse(scenario_ok, "PASS", "BLOCKED"),
    ifelse(qtl_indices_ok, "PASS", "BLOCKED"),
    ifelse(reference_lg_ok, "PASS", "BLOCKED"),
    ifelse(row_alignment_ok, "PASS", "BLOCKED"),
    ifelse(missing_genotypes == 0, "PASS", "BLOCKED"),
    ifelse(duplicate_individuals == 0, "PASS", "BLOCKED"),
    ifelse(duplicate_markers == 0, "PASS", "WARNING"),
    ifelse(monomorphic_markers == 0, "PASS", "WARNING"),
    ifelse(f2_frequency_ok, "PASS", "WARNING"),
    ifelse(maf_ok, "PASS", "WARNING"),
    ifelse(parent_structure_ok, "PASS", "BLOCKED"),
    ifelse(ld_decay_ok, "PASS", "WARNING")
  ),
  Details = c(
    "Expected: phenotype 1000 x 6, true genetic values 1000 x 6, genotype 1000 x 4010.",
    "Phenotypic and true genetic values must be finite.",
    paste("Observed:", paste(sort(unique(as.vector(geno_012))), collapse = ", ")),
    paste("Observed:", paste(sort(unique(as.vector(geno_m101))), collapse = ", ")),
    "Expected: 10 linkage groups, 401 markers per group, 0-200 cM.",
    "Expected QTL counts: 8, 40, 80, 120, 240, 480; nominal H2: 0.7, 0.7, 0.5, 0.5, 0.3, 0.3; Model = 2.",
    "All QTL indices must correspond to marker positions.",
    "LG9 and LG10 were intentionally simulated without QTL.",
    "Phenotype, true genetic values, and genotypes use the same 1,000 row positions.",
    paste(missing_genotypes, "missing genotype cells."),
    paste(duplicate_individuals, "duplicate individual genotype profiles."),
    paste(duplicate_markers, "duplicate marker profiles."),
    paste(monomorphic_markers, "monomorphic markers."),
    paste(
      "Observed proportions:",
      paste(round(as.numeric(geno_frequency), 5), collapse = ", ")
    ),
    sprintf(
      "MAF range %.4f to %.4f; median %.4f.",
      min(maf),
      max(maf),
      median(maf)
    ),
    "Two parents x ten linkage groups, with contrasting fixed marker states.",
    paste(
      "Mean r2 at 0.5, 10 and 50 cM:",
      paste(round(c(short_ld, medium_ld, ld_50), 4), collapse = ", "),
      "; interchromosomal mean r2:",
      round(interchromosomal_ld$Mean_r2, 5)
    )
  ),
  stringsAsFactors = FALSE
)
```

## 12. Shared data object for later modules

The quantities created above are saved together so that the next modules use
the same individuals, genotype coding, genetic map, scenario information, QTL
positions, and LD summaries. This prevents each prediction module from
re-reading and independently preparing the raw simulation files.

The three quantities with different scientific roles remain explicit:

- `phenotype` contains the response available for model fitting within each
  analysis set;
- `genotype_m101` contains the `-1/0/1` marker representation used by the
  prediction models;
- `true_genetic_value` is retained only for evaluating predictions on held-out
  individuals.

```{r save-shared-data, include=FALSE}
dir.create(
  "output/simulated/audit",
  showWarnings = FALSE,
  recursive = TRUE
)

write.csv(
  data_check_log,
  "output/simulated/audit/simulated_audit_summary.csv",
  row.names = FALSE
)

write.csv(
  scenario_summary,
  "output/simulated/audit/simulated_scenario_summary.csv",
  row.names = FALSE
)

write.csv(
  qtl_rows[, c("Scenario", "QTL_index", "LG", "Position_cM")],
  "output/simulated/audit/simulated_qtl_map.csv",
  row.names = FALSE
)

write.csv(
  ld_summary,
  "output/simulated/audit/simulated_ld_summary.csv",
  row.names = FALSE
)

simulated_common <- list(
  metadata = list(
    dataset = "simulated",
    seed = SEED_GLOBAL,
    n_individuals = nrow(geno_012),
    n_markers = ncol(geno_012),
    n_scenarios = ncol(phenotype),
    row_identifier = seq_len(nrow(geno_012)),
    row_alignment = "row position",
    genotype_coding = "-1/0/1"
  ),
  phenotype = phenotype,
  true_genetic_value = true_genetic_value,
  residual_value = residual_value,
  genotype_012 = geno_012,
  genotype_m101 = geno_m101,
  genetic_map = genetic_map[
    ,
    c(
      "LG",
      "Marker_within_LG",
      "Type",
      "Marker",
      "Position_cM",
      "Global_marker"
    )
  ],
  scenario_control = control_table,
  qtl_lists = qtl_lists,
  qtl_map = qtl_rows[, c("Scenario", "QTL_index", "LG", "Position_cM")],
  scenario_summary = scenario_summary,
  marker_qc = list(
    allele_frequency = allele_frequency,
    maf = maf,
    heterozygosity = heterozygosity
  ),
  ld_summary = ld_summary,
  interchromosomal_ld = interchromosomal_ld
)

saveRDS(
  simulated_common,
  "output/simulated/audit/preprocessed_simulated_common.rds",
  compress = "gzip"
)

cat(
  "Shared simulated-data object written to:",
  "output/simulated/audit/preprocessed_simulated_common.rds",
  "\n"
)
```

## 13. References

Cruz, C.D. (2013). GENES - a software package for analysis in experimental
statistics and quantitative genetics. *Acta Scientiarum*, 35(3), 271-276.

Cruz, C.D. (2016). Genes Software - extended and integrated with the R, Matlab
and Selegen. *Acta Scientiarum*, 38(4), 547-552.

## 14. Navigation

[Home](index.html) | [Next: Validation design](simulated_data_validation.html)
