---
title: "Simulated data: GBLUP-ADE and regularized ANN comparison"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    code_download: true
editor_options:
  chunk_output_type: console
---

## 1. Purpose

This module closes the simulated-data workflow by comparing GBLUP-ADE with the
four ANN variants on exactly the same outer-assessment individuals defined in
M02.

The comparison uses only predictions already produced in M03 and M04. No model
is refitted and no hyperparameter is selected here. The known **true simulated
genetic value** is the primary evaluation target; the **observed phenotype** is
reported as a secondary target because it also contains residual variation.

Three principles define the comparison:

- every method contributes the same 150 `Scenario x outer split` tasks;
- every task contains the same 200 held-out individuals for all methods;
- paired ANN-versus-GBLUP summaries are descriptive contrasts of repeated
  cross-validation results, not tests based on 150 statistically independent
  replicates.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)

options(stringsAsFactors = FALSE)

dir.create(
  "output/simulated/m05/figures",
  recursive = TRUE,
  showWarnings = FALSE
)
```

## 2. Inputs and matched evaluation design

### 2.1 Saved model results

```{r packages-and-inputs}
library(knitr)
library(ggplot2)
library(ggthemes)

GBLUP_RESULTS_RDS <-
  "output/simulated/gblup/simulated_gblup_ade_results.rds"
GBLUP_PREDICTIONS_RDS <-
  "output/simulated/gblup/simulated_gblup_ade_predictions.rds"
ANN_RESULTS_RDS <-
  "output/simulated/ann/final/simulated_ann_results.rds"
ANN_PREDICTIONS_RDS <-
  "output/simulated/ann/final/simulated_ann_predictions.rds"

gr <- readRDS(GBLUP_RESULTS_RDS)
gp <- readRDS(GBLUP_PREDICTIONS_RDS)
ar <- readRDS(ANN_RESULTS_RDS)
ap <- readRDS(ANN_PREDICTIONS_RDS)

expected_ann_models <- c(
  "ANN_0",
  "ANN_L1",
  "ANN_L2",
  "ANN_L1_L2"
)

model_order <- c(
  "GBLUP_ADE",
  expected_ann_models
)

input_summary <- data.frame(
  Object = c(
    "GBLUP-ADE task results",
    "GBLUP-ADE held-out predictions",
    "ANN task-by-model results",
    "ANN held-out predictions"
  ),
  Rows = c(
    nrow(gr),
    nrow(gp),
    nrow(ar),
    nrow(ap)
  ),
  stringsAsFactors = FALSE
)

input_summary |>
  knitr::kable(
    format = "html",
    caption = "Saved results entering the final model comparison",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The inputs contain 150 GBLUP-ADE task results, 600 ANN task-by-model results,
30,000 GBLUP-ADE held-out predictions, and 120,000 ANN held-out predictions.
The comparison therefore operates on completed model outputs rather than
reconstructing either modeling branch.

### 2.2 Matched tasks and assessment individuals

The next block checks the observed comparison structure directly. Prediction
records are aligned by scenario, repetition, outer fold, split identifier, task
identifier, and individual identifier before their targets are compared.

```{r matched-design}
comparison_keys <- c(
  "Scenario",
  "Repetition",
  "Fold",
  "SplitID",
  "TaskID",
  "ID"
)

model_structure_rows <- vector(
  "list",
  length(model_order)
)

for (model_index in seq_along(model_order)) {
  model <- model_order[model_index]

  if (model == "GBLUP_ADE") {
    result_data <- gr
    prediction_data <- gp
  } else {
    result_data <- ar[
      ar$Model == model,
      ,
      drop = FALSE
    ]
    prediction_data <- ap[
      ap$Model == model,
      ,
      drop = FALSE
    ]
  }

  predictions_per_task <- as.integer(
    table(prediction_data$TaskID)
  )

  model_structure_rows[[model_index]] <- data.frame(
    Model = model,
    Tasks = length(unique(result_data$TaskID)),
    Predictions = nrow(prediction_data),
    Min_predictions_per_task = min(predictions_per_task),
    Max_predictions_per_task = max(predictions_per_task),
    stringsAsFactors = FALSE
  )
}

model_structure <- do.call(
  rbind,
  model_structure_rows
)
rownames(model_structure) <- NULL

model_structure |>
  knitr::kable(
    format = "html",
    caption = "Observed task and assessment structure by model",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )

g_targets <- gp[
  , c(
    comparison_keys,
    "Observed_Phenotype",
    "True_Genetic_Value"
  )
]

paired_targets <- merge(
  g_targets,
  ap[
    , c(
      comparison_keys,
      "Model",
      "Observed_Phenotype",
      "True_Genetic_Value"
    )
  ],
  by = comparison_keys,
  suffixes = c("_GBLUP", "_ANN"),
  sort = FALSE
)

paired_targets$Phenotype_difference <- abs(
  paired_targets$Observed_Phenotype_GBLUP -
    paired_targets$Observed_Phenotype_ANN
)
paired_targets$TrueG_difference <- abs(
  paired_targets$True_Genetic_Value_GBLUP -
    paired_targets$True_Genetic_Value_ANN
)

target_match <- aggregate(
  cbind(
    Phenotype_difference,
    TrueG_difference
  ) ~ Model,
  data = paired_targets,
  FUN = max
)

target_match |>
  knitr::kable(
    format = "html",
    digits = 12,
    caption = "Maximum paired target difference relative to GBLUP-ADE",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

Every model contributes 150 tasks with 200 held-out predictions per task. The
maximum paired differences in phenotype and true genetic value are zero, so the
methods are evaluated on identical targets for identical held-out individuals.

## 3. Evaluation metrics

The same metrics are used in M03, M04, and this final comparison.

- **Pearson correlation** measures linear association between prediction and
  target. It is not a ranking metric and does not by itself measure calibration.
- **RMSE** and **MAE** measure prediction error on the original target scale;
  smaller values are better.
- **Bias** is `mean(prediction - target)`; values nearer zero indicate less mean
  prediction error.
- **Slope** is the prediction coefficient in `lm(target ~ prediction)`; values
  nearer one indicate better proportional calibration, while the regression
  intercept also contributes to calibration.
- **R-squared** is calculated as \(1-SSE/SST\). It is not squared Pearson
  correlation and can be negative for held-out predictions.

```{r metric-definitions, include=FALSE}
metrics <- c(
  "Genetic_Correlation",
  "Genetic_RMSE",
  "Genetic_MAE",
  "Genetic_Bias",
  "Genetic_Slope",
  "Genetic_R2",
  "Phenotype_Correlation",
  "Phenotype_RMSE",
  "Phenotype_MAE",
  "Phenotype_Bias",
  "Phenotype_Slope",
  "Phenotype_R2"
)

higher_better <- c(
  "Genetic_Correlation",
  "Genetic_R2",
  "Phenotype_Correlation",
  "Phenotype_R2"
)

lower_better <- c(
  "Genetic_RMSE",
  "Genetic_MAE",
  "Phenotype_RMSE",
  "Phenotype_MAE"
)

zero_target <- c(
  "Genetic_Bias",
  "Phenotype_Bias"
)

one_target <- c(
  "Genetic_Slope",
  "Phenotype_Slope"
)
```

For the paired contrasts below, positive performance advantage always favors the
ANN variant. Higher-is-better metrics use `ANN - GBLUP`; lower-is-better metrics
use `GBLUP - ANN`; bias is compared by distance from zero; and slope is compared
by distance from one.

## 4. Overall model performance

### 4.1 Overall means

```{r overall-performance}
analysis_cols <- c(
  "Scenario",
  "Repetition",
  "Fold",
  "SplitID",
  "TaskID",
  "Model",
  metrics
)

all_results <- rbind(
  gr[, analysis_cols],
  ar[, analysis_cols]
)

all_results$Model <- factor(
  all_results$Model,
  levels = model_order
)

overall <- aggregate(
  all_results[, metrics],
  by = list(
    Model = all_results$Model
  ),
  FUN = mean
)

overall$Model <- as.character(overall$Model)

overall_genetic_display <- overall[
  , c(
    "Model",
    "Genetic_Correlation",
    "Genetic_RMSE",
    "Genetic_MAE",
    "Genetic_R2"
  )
]

overall_genetic_display |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Primary overall performance against true genetic value",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )

overall_phenotype_display <- overall[
  , c(
    "Model",
    "Phenotype_Correlation",
    "Phenotype_RMSE",
    "Phenotype_MAE",
    "Phenotype_R2"
  )
]

overall_phenotype_display |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Secondary overall performance against observed phenotype",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )

best_corr_model <- overall$Model[
  which.max(overall$Genetic_Correlation)
]
best_rmse_model <- overall$Model[
  which.min(overall$Genetic_RMSE)
]
best_r2_model <- overall$Model[
  which.max(overall$Genetic_R2)
]

best_corr_value <- max(overall$Genetic_Correlation)
best_rmse_value <- min(overall$Genetic_RMSE)
best_r2_value <- max(overall$Genetic_R2)

ann_overall <- overall[
  overall$Model %in% expected_ann_models,
  ,
  drop = FALSE
]

best_ann_model <- ann_overall$Model[
  which.max(ann_overall$Genetic_Correlation)
]
best_ann_correlation <- max(
  ann_overall$Genetic_Correlation
)
```

Across the 150 matched outer tasks, the highest mean genetic Pearson correlation
is obtained by **`r best_corr_model`**
(`r sprintf("%.4f", best_corr_value)`), the lowest mean genetic RMSE by
**`r best_rmse_model`** (`r sprintf("%.4f", best_rmse_value)`), and the highest
mean genetic \(R^2\) by **`r best_r2_model`**
(`r sprintf("%.4f", best_r2_value)`). Among the ANN variants,
**`r best_ann_model`** has the highest mean genetic Pearson correlation
(`r sprintf("%.4f", best_ann_correlation)`). These are descriptive results for
the evaluated models and search spaces, not claims of universal model
superiority.

### 4.2 Distribution across outer tasks

The next figures show the 150 outer-task values for each method. Because the
repeated cross-validation tasks overlap in their training and assessment
samples, these distributions describe sensitivity to the held-out samples; they
are not interpreted as 150 statistically independent replicates.

```{r genetic-correlation-figure, fig.cap="Held-out true genetic-value Pearson correlation across the 150 outer tasks for GBLUP-ADE and the four ANN variants. Boxes summarize the task distribution; white diamonds indicate model means."}
plot_results <- all_results
plot_results$Model <- factor(
  plot_results$Model,
  levels = model_order,
  labels = c(
    "GBLUP-ADE",
    "ANN-0",
    "ANN-L1",
    "ANN-L2",
    "ANN-L1+L2"
  )
)

p_model_correlation <- ggplot(
  plot_results,
  aes(
    x = Model,
    y = Genetic_Correlation,
    fill = Model
  )
) +
  geom_boxplot(
    width = 0.58,
    outlier.shape = NA,
    alpha = 0.70
  ) +
  geom_point(
    position = position_jitter(
      width = 0.10,
      height = 0,
      seed = 20260821L + 501L
    ),
    alpha = 0.32,
    size = 1.3
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 3.2,
    fill = "white"
  ) +
  scale_fill_gdocs() +
  labs(
    x = NULL,
    y = "Pearson correlation with true genetic value",
    fill = NULL
  ) +
  theme_gdocs() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 20,
      hjust = 1
    )
  )

print(p_model_correlation)
```

```{r save-model-correlation-tiff, include=FALSE}
ggsave(
  "output/simulated/m05/figures/model_genetic_correlation_comparison.tiff",
  plot = p_model_correlation,
  device = "tiff",
  width = 8.5,
  height = 5.5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

```{r genetic-rmse-figure, fig.cap="Held-out true genetic-value RMSE across the 150 outer tasks for GBLUP-ADE and the four ANN variants. Lower values indicate smaller prediction error; white diamonds indicate model means."}
p_model_rmse <- ggplot(
  plot_results,
  aes(
    x = Model,
    y = Genetic_RMSE,
    fill = Model
  )
) +
  geom_boxplot(
    width = 0.58,
    outlier.shape = NA,
    alpha = 0.70
  ) +
  geom_point(
    position = position_jitter(
      width = 0.10,
      height = 0,
      seed = 20260821L + 502L
    ),
    alpha = 0.32,
    size = 1.3
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 3.2,
    fill = "white"
  ) +
  scale_fill_gdocs() +
  labs(
    x = NULL,
    y = "RMSE for true genetic value",
    fill = NULL
  ) +
  theme_gdocs() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 20,
      hjust = 1
    )
  )

print(p_model_rmse)
```

```{r save-model-rmse-tiff, include=FALSE}
ggsave(
  "output/simulated/m05/figures/model_genetic_rmse_comparison.tiff",
  plot = p_model_rmse,
  device = "tiff",
  width = 8.5,
  height = 5.5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

## 5. Performance by scenario

The scenario summary asks whether the overall pattern persists across C1-C6.
These scenarios jointly vary genetic complexity and nominal heritability, so
the sequence is not interpreted as the isolated causal effect of either
quantity.

```{r scenario-performance}
scenario_summary <- aggregate(
  all_results[, metrics],
  by = list(
    Scenario = all_results$Scenario,
    Model = all_results$Model
  ),
  FUN = mean
)

scenario_summary$Model <- as.character(
  scenario_summary$Model
)

scenario_display <- scenario_summary[
  , c(
    "Scenario",
    "Model",
    "Genetic_Correlation",
    "Genetic_RMSE",
    "Genetic_R2"
  )
]

scenario_display |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Primary genetic-value prediction metrics by scenario",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )

scenario_ids <- sort(
  unique(scenario_summary$Scenario)
)

scenario_winner_rows <- vector(
  "list",
  length(scenario_ids)
)

for (scenario_index in seq_along(scenario_ids)) {
  scenario <- scenario_ids[scenario_index]
  scenario_data <- scenario_summary[
    scenario_summary$Scenario == scenario,
    ,
    drop = FALSE
  ]

  scenario_winner_rows[[scenario_index]] <- data.frame(
    Scenario = scenario,
    Best_correlation = scenario_data$Model[
      which.max(scenario_data$Genetic_Correlation)
    ],
    Best_RMSE = scenario_data$Model[
      which.min(scenario_data$Genetic_RMSE)
    ],
    Best_R2 = scenario_data$Model[
      which.max(scenario_data$Genetic_R2)
    ],
    stringsAsFactors = FALSE
  )
}

scenario_winners <- do.call(
  rbind,
  scenario_winner_rows
)

corr_wins <- sum(
  scenario_winners$Best_correlation == "GBLUP_ADE"
)
rmse_wins <- sum(
  scenario_winners$Best_RMSE == "GBLUP_ADE"
)
r2_wins <- sum(
  scenario_winners$Best_R2 == "GBLUP_ADE"
)
```

GBLUP-ADE has the highest mean genetic Pearson correlation in
**`r corr_wins` of `r length(scenario_ids)` scenarios**, the lowest mean genetic
RMSE in **`r rmse_wins` of `r length(scenario_ids)`**, and the highest mean
genetic \(R^2\) in **`r r2_wins` of `r length(scenario_ids)`**.

## 6. Matched ANN versus GBLUP-ADE contrasts

The ANN-versus-GBLUP comparison is performed within each matched outer task.
Positive advantage always favors the ANN variant; negative advantage favors
GBLUP-ADE.

### 6.1 Overall matched contrasts

```{r matched-contrasts}
g <- gr[
  , c(
    "Scenario",
    "Repetition",
    "Fold",
    "SplitID",
    "TaskID",
    metrics
  )
]

names(g)[-(1:5)] <- paste0(
  names(g)[-(1:5)],
  "_GBLUP"
)

paired_rows <- list()
paired_row_index <- 0L

for (model in sort(unique(ar$Model))) {
  a <- ar[
    ar$Model == model,
    c(
      "Scenario",
      "Repetition",
      "Fold",
      "SplitID",
      "TaskID",
      metrics
    )
  ]

  matched <- merge(
    a,
    g,
    by = c(
      "Scenario",
      "Repetition",
      "Fold",
      "SplitID",
      "TaskID"
    ),
    all = FALSE,
    sort = FALSE
  )

  for (metric in metrics) {
    ann_value <- matched[[metric]]
    gblup_value <- matched[[paste0(metric, "_GBLUP")]]
    raw_difference <- ann_value - gblup_value

    if (metric %in% higher_better) {
      advantage <- ann_value - gblup_value
    } else if (metric %in% lower_better) {
      advantage <- gblup_value - ann_value
    } else if (metric %in% zero_target) {
      advantage <- abs(gblup_value) - abs(ann_value)
    } else {
      advantage <-
        abs(gblup_value - 1) -
        abs(ann_value - 1)
    }

    paired_row_index <- paired_row_index + 1L
    paired_rows[[paired_row_index]] <- data.frame(
      Model = model,
      Metric = metric,
      N = length(raw_difference),
      Mean_ANN = mean(ann_value),
      Mean_GBLUP = mean(gblup_value),
      Mean_raw_difference = mean(raw_difference),
      Median_raw_difference = median(raw_difference),
      Mean_performance_advantage = mean(advantage),
      Median_performance_advantage = median(advantage),
      ANN_better = sum(advantage > 0),
      Equal = sum(advantage == 0),
      GBLUP_better = sum(advantage < 0),
      stringsAsFactors = FALSE
    )
  }
}

paired <- do.call(
  rbind,
  paired_rows
)
rownames(paired) <- NULL

paired_primary <- paired[
  paired$Metric %in% c(
    "Genetic_Correlation",
    "Genetic_RMSE",
    "Genetic_R2"
  ),
  c(
    "Model",
    "Metric",
    "Mean_ANN",
    "Mean_GBLUP",
    "Mean_performance_advantage",
    "ANN_better",
    "Equal",
    "GBLUP_better"
  )
]

paired_primary |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Matched ANN versus GBLUP-ADE contrasts for primary genetic-value metrics",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

`ANN_better` and `GBLUP_better` count matched outer tasks, so each comparison
uses the same held-out sample for both methods. These counts describe how often
one method performs better within the repeated cross-validation design; they are
not used as independent observations for a naive fold-level significance test.

### 6.2 Matched contrasts by scenario

```{r matched-contrasts-by-scenario}
paired_scenario_rows <- list()
paired_scenario_index <- 0L

for (model in sort(unique(ar$Model))) {
  a <- ar[
    ar$Model == model,
    c(
      "Scenario",
      "Repetition",
      "Fold",
      "SplitID",
      "TaskID",
      metrics
    )
  ]

  matched <- merge(
    a,
    g,
    by = c(
      "Scenario",
      "Repetition",
      "Fold",
      "SplitID",
      "TaskID"
    ),
    all = FALSE,
    sort = FALSE
  )

  for (scenario in sort(unique(matched$Scenario))) {
    scenario_data <- matched[
      matched$Scenario == scenario,
      ,
      drop = FALSE
    ]

    for (metric in metrics) {
      ann_value <- scenario_data[[metric]]
      gblup_value <- scenario_data[[paste0(metric, "_GBLUP")]]

      if (metric %in% higher_better) {
        advantage <- ann_value - gblup_value
      } else if (metric %in% lower_better) {
        advantage <- gblup_value - ann_value
      } else if (metric %in% zero_target) {
        advantage <- abs(gblup_value) - abs(ann_value)
      } else {
        advantage <-
          abs(gblup_value - 1) -
          abs(ann_value - 1)
      }

      paired_scenario_index <-
        paired_scenario_index + 1L

      paired_scenario_rows[[paired_scenario_index]] <- data.frame(
        Scenario = scenario,
        Model = model,
        Metric = metric,
        N = nrow(scenario_data),
        Mean_advantage = mean(advantage),
        Median_advantage = median(advantage),
        ANN_better = sum(advantage > 0),
        Equal = sum(advantage == 0),
        GBLUP_better = sum(advantage < 0),
        stringsAsFactors = FALSE
      )
    }
  }
}

paired_scenario <- do.call(
  rbind,
  paired_scenario_rows
)
rownames(paired_scenario) <- NULL

paired_scenario_primary <- paired_scenario[
  paired_scenario$Metric == "Genetic_Correlation",
  c(
    "Scenario",
    "Model",
    "Mean_advantage",
    "ANN_better",
    "Equal",
    "GBLUP_better"
  )
]

paired_scenario_primary |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Matched genetic-correlation advantage by scenario; positive values favor ANN",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The scenario-level contrasts show whether differences are concentrated in a
subset of simulation conditions or recur across the full C1-C6 sequence. They
remain descriptive summaries of matched held-out tasks.

## 7. Regularization within the ANN branch

The four ANN variants share, within each outer task, the architecture and
learning rate selected with ANN_0 in M04. Their overall contrast therefore
describes the effect associated with the evaluated L1/L2 choices **conditional
on that shared structural selection rule**.

```{r ann-regularization}
ann_regularization_display <- overall[
  match(
    expected_ann_models,
    overall$Model
  ),
  c(
    "Model",
    "Genetic_Correlation",
    "Genetic_RMSE",
    "Genetic_MAE",
    "Genetic_R2"
  ),
  drop = FALSE
]

ann_regularization_display |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Overall primary genetic-value performance among ANN variants",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )

ann_l1 <- ann_overall[
  ann_overall$Model == "ANN_L1",
  ,
  drop = FALSE
]
ann_l1_l2 <- ann_overall[
  ann_overall$Model == "ANN_L1_L2",
  ,
  drop = FALSE
]
ann_l2 <- ann_overall[
  ann_overall$Model == "ANN_L2",
  ,
  drop = FALSE
]
ann_0 <- ann_overall[
  ann_overall$Model == "ANN_0",
  ,
  drop = FALSE
]
```

ANN-L1+L2 has the strongest overall ANN averages by a very small margin over
ANN-L1: genetic Pearson correlation is
`r sprintf("%.4f", ann_l1_l2$Genetic_Correlation)` versus
`r sprintf("%.4f", ann_l1$Genetic_Correlation)`, and genetic RMSE is
`r sprintf("%.4f", ann_l1_l2$Genetic_RMSE)` versus
`r sprintf("%.4f", ann_l1$Genetic_RMSE)`. ANN-L2 remains close to ANN-0
(correlations `r sprintf("%.4f", ann_l2$Genetic_Correlation)` and
`r sprintf("%.4f", ann_0$Genetic_Correlation)`, respectively). Within the
prespecified ANN search, the observed regularization gain is therefore associated
mainly with the L1 component.

## 8. Scientific conclusion

Within the prespecified validation design and ANN search space, GBLUP-ADE has
the strongest overall predictive performance for the known simulated genetic
value. It has the highest mean genetic Pearson correlation, the lowest mean
genetic RMSE, and the highest mean genetic \(R^2\), and the same three criteria
favor GBLUP-ADE across all six simulation scenarios.

Among the neural networks, ANN-L1+L2 is the strongest variant by a very small
margin over ANN-L1, whereas ANN-L2 remains close to ANN-0. The improvement
observed from regularization is therefore driven mainly by L1 under the bounded
architecture and penalty search evaluated here.

These conclusions are specific to the simulated scenarios, frozen repeated
cross-validation design, GBLUP-ADE specification, and prespecified ANN search
space used in this study. They do not establish a universal ranking between
genomic mixed models and neural networks.

```{r comparison-output-contract, include=FALSE}
WRITE_COMPARISON_OUTPUTS <- identical(
  Sys.getenv("WRITE_COMPARISON_OUTPUTS", "0"),
  "1"
)

COMPARISON_DIR <- "output/simulated/comparison"
OVERALL_CSV <- file.path(
  COMPARISON_DIR,
  "simulated_model_overall_summary.csv"
)
SCENARIO_CSV <- file.path(
  COMPARISON_DIR,
  "simulated_model_scenario_summary.csv"
)
PAIRED_CSV <- file.path(
  COMPARISON_DIR,
  "simulated_ann_vs_gblup_paired_summary.csv"
)
PAIRED_SCENARIO_CSV <- file.path(
  COMPARISON_DIR,
  "simulated_ann_vs_gblup_paired_by_scenario.csv"
)
```

```{r write-comparison-outputs, eval=WRITE_COMPARISON_OUTPUTS, include=FALSE}
dir.create(
  COMPARISON_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)
write.csv(
  overall,
  OVERALL_CSV,
  row.names = FALSE
)
write.csv(
  scenario_summary,
  SCENARIO_CSV,
  row.names = FALSE
)
write.csv(
  paired,
  PAIRED_CSV,
  row.names = FALSE
)
write.csv(
  paired_scenario,
  PAIRED_SCENARIO_CSV,
  row.names = FALSE
)
```

## 9. Navigation

[GBLUP-ADE](simulated_data_gblup.html) | [Regularized ANN](simulated_data_ann.html) | [Home](index.html)
