---
title: "Canonical comparison of GBLUP-ADE and regularized ANN models"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

# Purpose

This module is the **terminal comparison stage** of the canonical simulated-data
workflow. It compares the saved GBLUP-ADE results with the four saved ANN
variants after both modeling branches have been completed.

The page does **not** fit or tune any model. It consumes the canonical outputs
from `simulated_data_gblup.Rmd` and `simulated_data_ann.Rmd`, verifies direct
comparability under the frozen validation design, computes the descriptive
comparison tables, and checks them against the versioned canonical comparison
outputs.

The primary outcome is prediction of the **true simulated genetic value**.
Phenotype prediction is retained as a secondary outcome. Conclusions concerning
ANN performance are restricted to the **prespecified canonical ANN architecture
and hyperparameter search space**.

# Comparison contract

A direct comparison is allowed only when all of the following invariants hold:

1. GBLUP-ADE contains exactly 150 outer-CV tasks.
2. Each ANN variant contains the same 150 `TaskID`/`SplitID` combinations.
3. Scenario, repetition, fold, and split identifiers agree across methods.
4. Each task contains the same 200 outer-assessment individuals for GBLUP-ADE
   and every ANN variant.
5. `Observed_Phenotype` and `True_Genetic_Value` are identical for those paired
   individuals across methods.
6. All model fits and held-out predictions used in the comparison have status
   `ok`, and all reported comparison metrics and predictions are finite.
7. No ANN tuning, ANN final fitting, or GBLUP-ADE fitting is executed by this
   page.
8. Paired summaries are descriptive; repeated CV folds are not treated as
   statistically independent observations for naive significance testing.

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)

options(stringsAsFactors = FALSE)

find_project_root <- function(start = getwd()) {
  path <- normalizePath(start, winslash = "/", mustWork = TRUE)

  repeat {
    if (file.exists(file.path(path, "_workflowr.yml"))) {
      return(path)
    }

    parent <- dirname(path)

    if (identical(parent, path)) {
      stop("Could not locate workflowr project root.")
    }

    path <- parent
  }
}

PROJECT_ROOT <- find_project_root()
project_path <- function(...) file.path(PROJECT_ROOT, ...)

GBLUP_RESULTS_RDS <- project_path(
  "output", "simulated", "gblup", "simulated_gblup_ade_results.rds"
)
GBLUP_PREDICTIONS_RDS <- project_path(
  "output", "simulated", "gblup", "simulated_gblup_ade_predictions.rds"
)
ANN_RESULTS_RDS <- project_path(
  "output", "simulated", "ann", "final", "simulated_ann_results.rds"
)
ANN_PREDICTIONS_RDS <- project_path(
  "output", "simulated", "ann", "final", "simulated_ann_predictions.rds"
)

COMPARISON_DIR <- project_path("output", "simulated", "comparison")
OVERALL_CSV <- file.path(
  COMPARISON_DIR, "simulated_model_overall_summary.csv"
)
SCENARIO_CSV <- file.path(
  COMPARISON_DIR, "simulated_model_scenario_summary.csv"
)
PAIRED_CSV <- file.path(
  COMPARISON_DIR, "simulated_ann_vs_gblup_paired_summary.csv"
)
PAIRED_SCENARIO_CSV <- file.path(
  COMPARISON_DIR, "simulated_ann_vs_gblup_paired_by_scenario.csv"
)

WRITE_COMPARISON_OUTPUTS <- identical(
  Sys.getenv("WRITE_COMPARISON_OUTPUTS", "0"),
  "1"
)

metrics <- c(
  "Genetic_Correlation",
  "Genetic_RMSE",
  "Genetic_MAE",
  "Genetic_Bias",
  "Genetic_Slope",
  "Genetic_R2",
  "Phenotype_Correlation",
  "Phenotype_RMSE",
  "Phenotype_MAE",
  "Phenotype_Bias",
  "Phenotype_Slope",
  "Phenotype_R2"
)

higher_better <- c(
  "Genetic_Correlation",
  "Genetic_R2",
  "Phenotype_Correlation",
  "Phenotype_R2"
)
lower_better <- c(
  "Genetic_RMSE",
  "Genetic_MAE",
  "Phenotype_RMSE",
  "Phenotype_MAE"
)
zero_target <- c("Genetic_Bias", "Phenotype_Bias")
one_target <- c("Genetic_Slope", "Phenotype_Slope")

expected_ann_models <- c("ANN_0", "ANN_L1", "ANN_L2", "ANN_L1_L2")
```

# Canonical inputs

```{r load-canonical-inputs}
required_files <- c(
  GBLUP_RESULTS_RDS,
  GBLUP_PREDICTIONS_RDS,
  ANN_RESULTS_RDS,
  ANN_PREDICTIONS_RDS
)

if (!all(file.exists(required_files))) {
  stop(
    "Canonical model outputs are missing: ",
    paste(required_files[!file.exists(required_files)], collapse = ", ")
  )
}

gr <- readRDS(GBLUP_RESULTS_RDS)
gp <- readRDS(GBLUP_PREDICTIONS_RDS)
ar <- readRDS(ANN_RESULTS_RDS)
ap <- readRDS(ANN_PREDICTIONS_RDS)

cat("GBLUP result rows:", nrow(gr), "\n")
cat("GBLUP prediction rows:", nrow(gp), "\n")
cat("ANN result rows:", nrow(ar), "\n")
cat("ANN prediction rows:", nrow(ap), "\n")
cat("ANN models:", paste(sort(unique(ar$Model)), collapse = ", "), "\n")
```

# Compatibility audit

The result-level audit verifies the frozen task design. The prediction-level
audit independently verifies that the actual held-out individuals, observed
phenotypes, and true simulated genetic values are identical across methods.

```{r compatibility-audit}
required_result_cols <- c(
  "Scenario", "Repetition", "Fold", "SplitID", "TaskID",
  "Model", "Status", metrics
)

required_prediction_cols <- c(
  "Scenario", "Repetition", "Fold", "SplitID", "TaskID", "Model", "ID",
  "Observed_Phenotype", "True_Genetic_Value", "Predicted_Value", "Status"
)

missing_gr <- setdiff(required_result_cols, names(gr))
missing_ar <- setdiff(required_result_cols, names(ar))
missing_gp <- setdiff(required_prediction_cols, names(gp))
missing_ap <- setdiff(required_prediction_cols, names(ap))

if (length(missing_gr)) {
  stop("Missing GBLUP result columns: ", paste(missing_gr, collapse = ", "))
}
if (length(missing_ar)) {
  stop("Missing ANN result columns: ", paste(missing_ar, collapse = ", "))
}
if (length(missing_gp)) {
  stop("Missing GBLUP prediction columns: ", paste(missing_gp, collapse = ", "))
}
if (length(missing_ap)) {
  stop("Missing ANN prediction columns: ", paste(missing_ap, collapse = ", "))
}

result_key <- function(x) {
  paste(
    x$Scenario, x$Repetition, x$Fold, x$SplitID, x$TaskID,
    sep = "|"
  )
}

assessment_key <- function(x) {
  paste(
    x$Scenario, x$Repetition, x$Fold, x$SplitID, x$TaskID,
    as.character(x$ID),
    sep = "|"
  )
}

g_result_keys <- sort(result_key(gr))
g_assessment_keys <- sort(assessment_key(gp))
gp_ordered <- gp[order(assessment_key(gp)), , drop = FALSE]

ann_result_keys_match <- vapply(
  expected_ann_models,
  function(model) {
    z <- ar[ar$Model == model, , drop = FALSE]
    nrow(z) == 150L &&
      !anyDuplicated(z$TaskID) &&
      identical(sort(result_key(z)), g_result_keys)
  },
  logical(1)
)

ann_assessment_keys_match <- vapply(
  expected_ann_models,
  function(model) {
    z <- ap[ap$Model == model, , drop = FALSE]
    nrow(z) == 30000L &&
      !anyDuplicated(z[, c("TaskID", "ID")]) &&
      identical(sort(assessment_key(z)), g_assessment_keys)
  },
  logical(1)
)

ann_phenotype_match <- vapply(
  expected_ann_models,
  function(model) {
    z <- ap[ap$Model == model, , drop = FALSE]
    z <- z[order(assessment_key(z)), , drop = FALSE]
    isTRUE(all.equal(
      z$Observed_Phenotype,
      gp_ordered$Observed_Phenotype,
      tolerance = 1e-12,
      check.attributes = FALSE
    ))
  },
  logical(1)
)

ann_true_g_match <- vapply(
  expected_ann_models,
  function(model) {
    z <- ap[ap$Model == model, , drop = FALSE]
    z <- z[order(assessment_key(z)), , drop = FALSE]
    isTRUE(all.equal(
      z$True_Genetic_Value,
      gp_ordered$True_Genetic_Value,
      tolerance = 1e-12,
      check.attributes = FALSE
    ))
  },
  logical(1)
)

compatibility_checks <- data.frame(
  Check = c(
    "GBLUP has 150 result rows",
    "ANN has 600 result rows",
    "expected ANN models are present",
    "GBLUP TaskID values are unique",
    "each ANN model has identical result-level task keys",
    "GBLUP has 30000 held-out prediction rows",
    "ANN has 120000 held-out prediction rows",
    "GBLUP has exactly 200 predictions per task",
    "ANN has exactly 200 predictions per task and model",
    "GBLUP TaskID/ID pairs are unique",
    "each ANN model has identical outer-assessment individuals",
    "observed phenotypes are identical across paired assessments",
    "true genetic values are identical across paired assessments",
    "all GBLUP result statuses are ok",
    "all ANN result statuses are ok",
    "all GBLUP prediction statuses are ok",
    "all ANN prediction statuses are ok",
    "all GBLUP comparison metrics are finite",
    "all ANN comparison metrics are finite",
    "all GBLUP held-out predictions are finite",
    "all ANN held-out predictions are finite"
  ),
  PASS = c(
    nrow(gr) == 150L,
    nrow(ar) == 600L,
    identical(sort(unique(ar$Model)), sort(expected_ann_models)),
    !anyDuplicated(gr$TaskID),
    all(ann_result_keys_match),
    nrow(gp) == 30000L,
    nrow(ap) == 120000L,
    all(table(gp$TaskID) == 200L),
    all(table(interaction(ap$TaskID, ap$Model, drop = TRUE)) == 200L),
    !anyDuplicated(gp[, c("TaskID", "ID")]),
    all(ann_assessment_keys_match),
    all(ann_phenotype_match),
    all(ann_true_g_match),
    all(gr$Status == "ok"),
    all(ar$Status == "ok"),
    all(gp$Status == "ok"),
    all(ap$Status == "ok"),
    all(vapply(
      gr[, metrics, drop = FALSE],
      function(x) all(is.finite(x)),
      logical(1)
    )),
    all(vapply(
      ar[, metrics, drop = FALSE],
      function(x) all(is.finite(x)),
      logical(1)
    )),
    all(is.finite(gp$Predicted_Value)),
    all(is.finite(ap$Predicted_Value))
  ),
  stringsAsFactors = FALSE
)

compatibility_checks$Status <- ifelse(
  compatibility_checks$PASS,
  "PASS",
  "BLOCKED"
)

knitr::kable(
  compatibility_checks[, c("Check", "Status")],
  caption = "Compatibility audit for the canonical model comparison."
)

if (any(!compatibility_checks$PASS)) {
  stop("Canonical model-comparison compatibility audit failed.")
}
```

# Overall model performance

```{r overall-comparison}
analysis_cols <- c(
  "Scenario", "Repetition", "Fold", "SplitID", "TaskID", "Model", metrics
)

all_results <- rbind(gr[, analysis_cols], ar[, analysis_cols])

overall <- aggregate(
  all_results[, metrics],
  by = list(Model = all_results$Model),
  FUN = mean
)

overall_display <- overall[, c(
  "Model",
  "Genetic_Correlation",
  "Genetic_RMSE",
  "Genetic_MAE",
  "Genetic_R2",
  "Phenotype_Correlation",
  "Phenotype_RMSE",
  "Phenotype_MAE",
  "Phenotype_R2"
)]
overall_display[-1] <- lapply(overall_display[-1], round, 4)

knitr::kable(
  overall_display,
  caption = paste(
    "Overall held-out predictive performance.",
    "Genetic-value prediction is the primary outcome."
  )
)
```

# Performance by simulation scenario

```{r scenario-comparison}
scenario_summary <- aggregate(
  all_results[, metrics],
  by = list(
    Scenario = all_results$Scenario,
    Model = all_results$Model
  ),
  FUN = mean
)

scenario_display <- scenario_summary[, c(
  "Scenario", "Model", "Genetic_Correlation", "Genetic_RMSE", "Genetic_R2"
)]
scenario_display[, -(1:2)] <- lapply(
  scenario_display[, -(1:2), drop = FALSE],
  round,
  4
)

knitr::kable(
  scenario_display,
  caption = "Primary genetic-value prediction metrics by scenario."
)
```

# Paired ANN versus GBLUP-ADE comparison

A positive `performance_advantage` always means that the ANN variant performed
better than GBLUP-ADE for the metric under consideration. Correlations and
\(R^2\) are maximized, RMSE and MAE are minimized, bias is evaluated by distance
from zero, and slope is evaluated by distance from one.

```{r paired-comparison}
performance_advantage <- function(metric, ann_value, gblup_value) {
  if (metric %in% higher_better) {
    return(ann_value - gblup_value)
  }
  if (metric %in% lower_better) {
    return(gblup_value - ann_value)
  }
  if (metric %in% zero_target) {
    return(abs(gblup_value) - abs(ann_value))
  }
  if (metric %in% one_target) {
    return(abs(gblup_value - 1) - abs(ann_value - 1))
  }

  stop("Unclassified metric: ", metric)
}

g <- gr[, c("TaskID", "SplitID", metrics)]
names(g)[-(1:2)] <- paste0(names(g)[-(1:2)], "_GBLUP")

paired_list <- list()

for (model in sort(unique(ar$Model))) {
  a <- ar[
    ar$Model == model,
    c("TaskID", "SplitID", "Scenario", metrics)
  ]

  z <- merge(a, g, by = c("TaskID", "SplitID"), all = FALSE)

  if (nrow(z) != 150L) {
    stop(model, ": expected 150 paired tasks, got ", nrow(z))
  }

  for (metric in metrics) {
    ann_value <- z[[metric]]
    gblup_value <- z[[paste0(metric, "_GBLUP")]]
    raw_difference <- ann_value - gblup_value
    advantage <- performance_advantage(metric, ann_value, gblup_value)

    paired_list[[length(paired_list) + 1L]] <- data.frame(
      Model = model,
      Metric = metric,
      N = length(raw_difference),
      Mean_ANN = mean(ann_value),
      Mean_GBLUP = mean(gblup_value),
      Mean_raw_difference = mean(raw_difference),
      Median_raw_difference = median(raw_difference),
      Mean_performance_advantage = mean(advantage),
      Median_performance_advantage = median(advantage),
      ANN_better = sum(advantage > 0),
      Equal = sum(advantage == 0),
      GBLUP_better = sum(advantage < 0),
      stringsAsFactors = FALSE
    )
  }
}

paired <- do.call(rbind, paired_list)
rownames(paired) <- NULL

paired_primary <- paired[
  paired$Metric %in% c("Genetic_Correlation", "Genetic_RMSE", "Genetic_R2"),
  c(
    "Model", "Metric", "Mean_ANN", "Mean_GBLUP",
    "Mean_performance_advantage", "ANN_better", "Equal", "GBLUP_better"
  )
]
paired_primary[, 3:5] <- lapply(
  paired_primary[, 3:5, drop = FALSE],
  round,
  4
)

knitr::kable(
  paired_primary,
  caption = paste(
    "Paired ANN versus GBLUP-ADE comparison for the primary",
    "genetic-value metrics."
  )
)
```

# Paired comparison by scenario

```{r paired-by-scenario}
paired_scenario_list <- list()

for (model in sort(unique(ar$Model))) {
  a <- ar[
    ar$Model == model,
    c("TaskID", "SplitID", "Scenario", metrics)
  ]

  z <- merge(a, g, by = c("TaskID", "SplitID"), all = FALSE)

  for (scenario in sort(unique(z$Scenario))) {
    zs <- z[z$Scenario == scenario, , drop = FALSE]

    if (nrow(zs) != 25L) {
      stop(
        model, " / ", scenario,
        ": expected 25 paired tasks, got ", nrow(zs)
      )
    }

    for (metric in metrics) {
      ann_value <- zs[[metric]]
      gblup_value <- zs[[paste0(metric, "_GBLUP")]]
      advantage <- performance_advantage(metric, ann_value, gblup_value)

      paired_scenario_list[[length(paired_scenario_list) + 1L]] <- data.frame(
        Scenario = scenario,
        Model = model,
        Metric = metric,
        N = nrow(zs),
        Mean_advantage = mean(advantage),
        Median_advantage = median(advantage),
        ANN_better = sum(advantage > 0),
        Equal = sum(advantage == 0),
        GBLUP_better = sum(advantage < 0),
        stringsAsFactors = FALSE
      )
    }
  }
}

paired_scenario <- do.call(rbind, paired_scenario_list)
rownames(paired_scenario) <- NULL

paired_scenario_primary <- paired_scenario[
  paired_scenario$Metric == "Genetic_Correlation",
  c(
    "Scenario", "Model", "Mean_advantage",
    "ANN_better", "Equal", "GBLUP_better"
  )
]
paired_scenario_primary$Mean_advantage <- round(
  paired_scenario_primary$Mean_advantage,
  4
)

knitr::kable(
  paired_scenario_primary,
  caption = paste(
    "Paired genetic-correlation advantage by scenario.",
    "Positive values favor ANN."
  )
)
```

# Canonical comparison-output audit

The four versioned CSV files are canonical exported summaries. A normal render
recomputes the comparison in memory and verifies those exports. Set
`WRITE_COMPARISON_OUTPUTS=1` only for an intentional refresh of the comparison
CSVs; this flag never fits ANN or GBLUP-ADE models.

```{r write-comparison-outputs, eval=WRITE_COMPARISON_OUTPUTS}
dir.create(COMPARISON_DIR, recursive = TRUE, showWarnings = FALSE)
write.csv(overall, OVERALL_CSV, row.names = FALSE)
write.csv(scenario_summary, SCENARIO_CSV, row.names = FALSE)
write.csv(paired, PAIRED_CSV, row.names = FALSE)
write.csv(paired_scenario, PAIRED_SCENARIO_CSV, row.names = FALSE)
```

```{r verify-comparison-outputs}
canonical_files <- c(
  OVERALL_CSV,
  SCENARIO_CSV,
  PAIRED_CSV,
  PAIRED_SCENARIO_CSV
)

if (!all(file.exists(canonical_files))) {
  stop(
    "Canonical comparison CSV files are missing: ",
    paste(canonical_files[!file.exists(canonical_files)], collapse = ", ")
  )
}

frames_equal <- function(saved, computed, tolerance = 1e-12) {
  if (!identical(names(saved), names(computed)) || nrow(saved) != nrow(computed)) {
    return(FALSE)
  }

  for (name in names(saved)) {
    x <- saved[[name]]
    y <- computed[[name]]

    if (is.numeric(x) && is.numeric(y)) {
      if (!all(is.finite(x)) || !all(is.finite(y))) {
        return(FALSE)
      }
      if (!all(abs(x - y) <= tolerance)) {
        return(FALSE)
      }
    } else if (!identical(as.character(x), as.character(y))) {
      return(FALSE)
    }
  }

  TRUE
}

saved_overall <- read.csv(
  OVERALL_CSV, stringsAsFactors = FALSE, check.names = FALSE
)
saved_scenario <- read.csv(
  SCENARIO_CSV, stringsAsFactors = FALSE, check.names = FALSE
)
saved_paired <- read.csv(
  PAIRED_CSV, stringsAsFactors = FALSE, check.names = FALSE
)
saved_paired_scenario <- read.csv(
  PAIRED_SCENARIO_CSV, stringsAsFactors = FALSE, check.names = FALSE
)

output_checks <- data.frame(
  Output = basename(canonical_files),
  Status = ifelse(
    c(
      frames_equal(saved_overall, overall),
      frames_equal(saved_scenario, scenario_summary),
      frames_equal(saved_paired, paired),
      frames_equal(saved_paired_scenario, paired_scenario)
    ),
    "PASS",
    "BLOCKED"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  output_checks,
  caption = "Agreement between recomputed and versioned comparison tables."
)

if (any(output_checks$Status != "PASS")) {
  stop(
    "Canonical comparison exports do not match the recomputed results. ",
    "Review model outputs before using WRITE_COMPARISON_OUTPUTS=1."
  )
}
```

# Main findings

```{r main-findings, echo=FALSE}
best_model_corr <- overall$Model[which.max(overall$Genetic_Correlation)]

ann_overall <- overall[overall$Model != "GBLUP_ADE", , drop = FALSE]
best_ann_corr <- ann_overall$Model[
  which.max(ann_overall$Genetic_Correlation)
]

scenarios <- sort(unique(scenario_summary$Scenario))

gblup_corr_wins <- sum(vapply(
  scenarios,
  function(scenario) {
    z <- scenario_summary[
      scenario_summary$Scenario == scenario,
      ,
      drop = FALSE
    ]
    z$Model[which.max(z$Genetic_Correlation)] == "GBLUP_ADE"
  },
  logical(1)
))

gblup_rmse_wins <- sum(vapply(
  scenarios,
  function(scenario) {
    z <- scenario_summary[
      scenario_summary$Scenario == scenario,
      ,
      drop = FALSE
    ]
    z$Model[which.min(z$Genetic_RMSE)] == "GBLUP_ADE"
  },
  logical(1)
))
```

Within the prespecified canonical model and ANN architecture search space,
**`r best_model_corr`** has the highest overall mean genetic correlation.

Among the neural-network variants, **`r best_ann_corr`** has the highest overall
mean genetic correlation. ANN variants should be interpreted together with the
regularization pattern rather than from a single metric in isolation.

GBLUP-ADE has the highest mean genetic correlation in
**`r gblup_corr_wins` of `r length(scenarios)` scenarios** and the lowest mean
genetic RMSE in **`r gblup_rmse_wins` of `r length(scenarios)` scenarios**.

These results do not establish a universal ordering between genomic mixed
models and all possible neural-network architectures. They describe performance
under the frozen outer-CV design and the prespecified canonical ANN search
space used in this project.

# Reproducibility

A normal render reads the saved GBLUP-ADE and ANN result and prediction objects,
audits task-level and individual-level comparability, recomputes the comparison
summaries, and verifies the four canonical CSV exports. It does not run ANN
tuning, final ANN fitting, or GBLUP-ADE fitting.

The canonical exported comparison tables are stored under
`output/simulated/comparison/`.

```{r session-info}
sessionInfo()
```
