---
title: "Simulated data: regularized ANN"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

## 1. Purpose

This module fits four feed-forward neural-network variants to the same six
simulated scenarios and the same nested validation design established in M02:
`ANN_0`, `ANN_L1`, `ANN_L2`, and `ANN_L1_L2`.

The scientific separation is fixed throughout the module:

- the **outer analysis set** contains 800 individuals and is the only source of
  phenotype information used for tuning, epoch selection, and final fitting;
- the **outer assessment set** contains 200 individuals and remains untouched
  until the final prediction for that task already exists;
- the **four inner folds** are used only to select architecture, learning rate,
  and applicable L1/L2 penalties inside the outer analysis set;
- **true genetic values** do not participate in partition generation,
  preprocessing, tuning, epoch selection, fitting, or prediction and are used
  only for held-out evaluation.

Architecture and learning rate are selected with `ANN_0` for each outer task and
then held fixed across all four variants. The regularized variants subsequently
tune only their applicable penalties. This creates a controlled comparison of
regularization **conditional on the structure selected with ANN_0**; it does not
claim that each regularized family would select the same structure under an
independent architecture search.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)

options(stringsAsFactors = FALSE)

dir.create(
  "output/simulated/m04/figures",
  recursive = TRUE,
  showWarnings = FALSE
)
```

## 2. ANN variants and fixed settings

The four variants use the same multilayer perceptron and differ only in the
penalties applied to network weights. Bias terms are not penalized.

```{r packages-and-settings, warning=FALSE, message=FALSE}
library(knitr)
library(ggplot2)
library(ggthemes)

RUN_ANN_TUNING <- identical(
  Sys.getenv("RUN_ANN_TUNING", "0"),
  "1"
)
RUN_ANN <- identical(
  Sys.getenv("RUN_ANN", "0"),
  "1"
)
ANN_RUNTIME_REQUIRED <- RUN_ANN_TUNING || RUN_ANN

if (ANN_RUNTIME_REQUIRED) {
  suppressPackageStartupMessages(library(torch))
}

SEED_GLOBAL <- 20260821L
MAX_EPOCHS <- 2000L
PATIENCE <- 25L
MIN_DELTA <- 1e-6
EARLY_VALIDATION_FRACTION <- 0.15
DROPOUT <- 0
FINAL_EPOCH_MAX_EPOCHS <- 3000L

MODEL_LEVELS <- c(
  "ANN_0",
  "ANN_L1",
  "ANN_L2",
  "ANN_L1_L2"
)

ann_variants <- data.frame(
  Model = MODEL_LEVELS,
  L1 = c("No", "Yes", "No", "Yes"),
  L2 = c("No", "No", "Yes", "Yes"),
  Interpretation = c(
    "Unregularized reference ANN.",
    "ANN with L1 weight regularization.",
    "ANN with L2 weight regularization.",
    "ANN with simultaneous L1 and L2 weight regularization."
  ),
  stringsAsFactors = FALSE
)

ann_variants |>
  knitr::kable(
    format = "html",
    caption = "ANN variants evaluated under the same validation design",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )

fixed_controls <- data.frame(
  Control = c(
    "Activation",
    "Optimizer",
    "Dropout",
    "Maximum tuning epochs",
    "Patience",
    "Minimum improvement",
    "Early-validation fraction",
    "Maximum epoch-selection epochs"
  ),
  Value = c(
    "ReLU",
    "Adam",
    DROPOUT,
    MAX_EPOCHS,
    PATIENCE,
    MIN_DELTA,
    EARLY_VALIDATION_FRACTION,
    FINAL_EPOCH_MAX_EPOCHS
  ),
  stringsAsFactors = FALSE
)

fixed_controls |>
  knitr::kable(
    format = "html",
    caption = "Fixed ANN optimization controls",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The 2000-epoch and 3000-epoch values are **maximum optimization caps**, not
claims of numerical convergence and not fixed training lengths. Early stopping
can terminate an optimization run earlier when the prespecified patience rule is
met.

## 3. Data and validation structure

The ANN branch reuses the M01 data object and the M02 nested validation object
directly.

```{r inputs}
common <- readRDS(
  "output/simulated/audit/preprocessed_simulated_common.rds"
)
cv <- readRDS(
  "output/simulated/validation/simulated_cv_design.rds"
)

phenotype <- common$phenotype
geno <- common$genotype_m101
control <- common$scenario_control

ids <- seq_len(nrow(geno))
scenario_names <- colnames(phenotype)

input_summary <- data.frame(
  Object = c(
    "Phenotype",
    "Genotype",
    "Outer splits",
    "Inner splits"
  ),
  Structure = c(
    paste(dim(phenotype), collapse = " x "),
    paste(dim(geno), collapse = " x "),
    length(cv$outer_splits),
    sum(vapply(cv$inner_splits, length, integer(1)))
  ),
  Role = c(
    "Response used inside analysis subsets only.",
    "4,010 marker predictors used by all ANN variants.",
    "25 held-out outer evaluation partitions.",
    "100 inner partitions used for nested tuning."
  ),
  stringsAsFactors = FALSE
)

input_summary |>
  knitr::kable(
    format = "html",
    caption = "Objects entering the ANN workflow",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The same 1,000 individuals, 4,010 markers, six scenarios, 25 outer splits, and
100 inner splits are therefore used by all ANN variants. No ANN variant creates
its own validation partition.

## 4. Neural-network model and regularization

For standardized marker vector \(\mathbf{x}\), each hidden layer applies

\[
\mathbf{h}^{(\ell)}=
\operatorname{ReLU}
\left(
\mathbf{W}_{\ell}\mathbf{h}^{(\ell-1)}+
\mathbf{b}_{\ell}
\right).
\]

The final layer is linear because the response is continuous. Training minimizes

\[
\mathcal{L}=
\operatorname{MSE}(y,\hat y)
+
\lambda_1\sum_j |w_j|
+
\frac{\lambda_2}{2}\sum_j w_j^2.
\]

`ANN_0` sets both penalties to zero, `ANN_L1` uses only \(\lambda_1\),
`ANN_L2` uses only \(\lambda_2\), and `ANN_L1_L2` uses both. Adam is run with
`weight_decay = 0`, so L2 enters only through the explicit penalty shown above.

The runtime implementation below is evaluated only when tuning or final ANN
fitting is explicitly requested. The `initialize` and `forward` methods are the
network definition required by the `torch` API; the analysis itself does not use
separate local helper functions for preprocessing, fitting, ranking, or metric
calculation.

```{r runtime-and-model, eval=ANN_RUNTIME_REQUIRED}
cuda_available <- torch::cuda_is_available()
requested_device <- tolower(Sys.getenv("ANN_DEVICE", "auto"))

ANN_DEVICE_LABEL <- if (requested_device == "cuda") {
  "cuda"
} else if (requested_device == "cpu") {
  "cpu"
} else if (cuda_available) {
  "cuda"
} else {
  "cpu"
}

ANN_DEVICE <- torch::torch_device(ANN_DEVICE_LABEL)

TorchGenomicMLP <- torch::nn_module(
  "TorchGenomicMLP",
  initialize = function(input_dim, hidden_units) {
    self$linears <- torch::nn_module_list()
    previous <- as.integer(input_dim)

    for (hidden in hidden_units) {
      self$linears$append(
        torch::nn_linear(
          previous,
          as.integer(hidden)
        )
      )
      previous <- as.integer(hidden)
    }

    self$output <- torch::nn_linear(previous, 1L)
  },
  forward = function(x) {
    for (layer_index in seq_len(length(self$linears))) {
      x <- torch::nnf_relu(
        self$linears[[layer_index]](x)
      )
    }
    self$output(x)
  }
)
```

## 5. Nested hyperparameter tuning

### 5.1 Search space

The architecture catalog is intentionally bounded. Architecture and learning
rate are selected first with `ANN_0`; L1/L2 penalties are selected only after
that shared structure has been fixed for the outer task.

```{r search-space}
ARCHITECTURES <- list(
  `2` = c(2L),
  `4` = c(4L),
  `8` = c(8L),
  `16` = c(16L),
  `32` = c(32L),
  `8-4` = c(8L, 4L),
  `16-8` = c(16L, 8L),
  `32-16` = c(32L, 16L),
  `64-32` = c(64L, 32L)
)

LEARNING_RATES <- c(
  1.25e-4,
  2.5e-4,
  5e-4,
  1e-3
)

L1_VALUES <- c(
  1e-5,
  1e-4,
  1e-3,
  1e-2,
  3e-2,
  1e-1
)

L2_VALUES <- c(
  1e-5,
  1e-4,
  1e-3,
  1e-2
)

L1_L2_GRID <- expand.grid(
  L1 = L1_VALUES,
  L2 = L2_VALUES,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

structure_grid <- expand.grid(
  Architecture = names(ARCHITECTURES),
  Learning_rate = LEARNING_RATES,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)
structure_grid$Model <- "ANN_0"
structure_grid$L1 <- 0
structure_grid$L2 <- 0

l1_grid <- data.frame(
  Model = "ANN_L1",
  L1 = L1_VALUES,
  L2 = 0,
  stringsAsFactors = FALSE
)

l2_grid <- data.frame(
  Model = "ANN_L2",
  L1 = 0,
  L2 = L2_VALUES,
  stringsAsFactors = FALSE
)

l1l2_grid <- data.frame(
  Model = "ANN_L1_L2",
  L1 = L1_L2_GRID$L1,
  L2 = L1_L2_GRID$L2,
  stringsAsFactors = FALSE
)

search_summary <- data.frame(
  Stage = c(
    "Architecture + learning rate with ANN_0",
    "ANN_L1 penalty",
    "ANN_L2 penalty",
    "ANN_L1_L2 penalties"
  ),
  Candidates = c(
    nrow(structure_grid),
    nrow(l1_grid),
    nrow(l2_grid),
    nrow(l1l2_grid)
  ),
  Inner_fits_per_outer_task = 4L * c(
    nrow(structure_grid),
    nrow(l1_grid),
    nrow(l2_grid),
    nrow(l1l2_grid)
  ),
  stringsAsFactors = FALSE
)

search_summary |>
  knitr::kable(
    format = "html",
    caption = "Prespecified ANN tuning search",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )

CANDIDATES_PER_TASK <- sum(search_summary$Candidates)
INNER_FITS_PER_TASK <- 4L * CANDIDATES_PER_TASK
```

The search contains **70 candidate configurations per outer task**: 36
architecture-learning-rate combinations, 6 L1 values, 4 L2 values, and 24
positive L1-L2 pairs. With four inner folds, that yields **280 candidate-fold
fits per outer task** and **42,000 inner fits across the 150 scenario-by-split
outer tasks**.

Selection of `64-32` or another boundary value is interpreted descriptively as
boundary occupancy of this prespecified search space. It does not trigger a
post-hoc enlargement of the search.

### 5.2 Inner fitting and preprocessing

For every inner fold, the 600 inner-analysis individuals are split again only
for optimization: 15% form an early-validation subset and the remainder form the
fitting subset. The 200 frozen inner-assessment individuals remain untouched
until the candidate has been trained and its best early-validation checkpoint
has been restored.

Marker means, marker standard deviations, phenotype mean, and phenotype standard
deviation are estimated **only from the current fitting subset**. Those fitted
transformations are then applied unchanged to the early-validation and
inner-assessment observations. Markers with fitting-set standard deviation below
`1e-8` receive unit scale to avoid division by zero.

The complete tuning code below writes the same historical output schemas when
`RUN_ANN_TUNING=1`. The normal page build leaves this expensive stage disabled
and reads the existing saved outputs instead.

```{r tuning-paths, include=FALSE}
SELECTED_RDS <- "output/simulated/ann/tuning/simulated_ann_tuning_selected_configurations.rds"
CANDIDATE_RDS <- "output/simulated/ann/tuning/simulated_ann_tuning_candidate_summary.rds"
FOLD_RDS <- "output/simulated/ann/tuning/simulated_ann_tuning_fold_results.rds"
RUN_RDS <- "output/simulated/ann/tuning/simulated_ann_tuning_run_metadata.rds"
MANIFEST_RDS <- "output/simulated/ann/tuning/simulated_ann_tuning_manifest.rds"

SELECTED_CSV <- "output/simulated/ann/tuning/simulated_ann_tuning_selected_configurations.csv"
CANDIDATE_CSV <- "output/simulated/ann/tuning/simulated_ann_tuning_candidate_summary.csv"
FOLD_CSV <- "output/simulated/ann/tuning/simulated_ann_tuning_fold_results.csv"
MANIFEST_CSV <- "output/simulated/ann/tuning/simulated_ann_tuning_manifest.csv"

FINAL_RESULTS_RDS <- "output/simulated/ann/final/simulated_ann_results.rds"
FINAL_PREDICTIONS_RDS <- "output/simulated/ann/final/simulated_ann_predictions.rds"
FINAL_EPOCH_RDS <- "output/simulated/ann/final/simulated_ann_epoch_selection.rds"
FINAL_RUN_RDS <- "output/simulated/ann/final/simulated_ann_run_metadata.rds"

FINAL_RESULTS_CSV <- "output/simulated/ann/final/simulated_ann_results.csv"
FINAL_PREDICTIONS_CSV <- "output/simulated/ann/final/simulated_ann_predictions.csv"
FINAL_EPOCH_CSV <- "output/simulated/ann/final/simulated_ann_epoch_selection.csv"

TUNING_V1_COMPATIBILITY_TOKEN <- 2000L
ANN_TUNING_CONFIG_ID <- paste(
  "ANN_TUNING_CANONICAL_V1",
  paste(names(ARCHITECTURES), collapse = "-"),
  paste(LEARNING_RATES, collapse = "-"),
  paste(L1_VALUES, collapse = "-"),
  paste(L2_VALUES, collapse = "-"),
  MAX_EPOCHS,
  PATIENCE,
  MIN_DELTA,
  EARLY_VALIDATION_FRACTION,
  TUNING_V1_COMPATIBILITY_TOKEN,
  sep = "__"
)

FINAL_CONFIG_ID <- paste0(
  "ANN_FINAL_CANONICAL_V2__FROM__",
  ANN_TUNING_CONFIG_ID,
  "__FINAL_EPOCH_CAP_",
  FINAL_EPOCH_MAX_EPOCHS,
  "__BEST_EPOCH_REFIT"
)
```

```{r run-tuning, eval=RUN_ANN_TUNING}
dir.create(
  "output/simulated/ann/tuning",
  recursive = TRUE,
  showWarnings = FALSE
)

outer_map <- unique(
  cv$outer_assignments[
    , c("Repetition", "OuterFold", "SplitID")
  ]
)
outer_map <- outer_map[
  order(outer_map$Repetition, outer_map$OuterFold),
  ,
  drop = FALSE
]

task_table <- merge(
  data.frame(
    Scenario = scenario_names,
    stringsAsFactors = FALSE
  ),
  outer_map,
  by = NULL
)
task_table <- task_table[
  order(
    match(task_table$Scenario, scenario_names),
    task_table$Repetition,
    task_table$OuterFold
  ),
  ,
  drop = FALSE
]
task_table$Fold <- task_table$OuterFold
task_table$TaskID <- paste(
  task_table$Scenario,
  task_table$SplitID,
  sep = "_"
)
rownames(task_table) <- NULL

selected_task_rows <- vector("list", nrow(task_table))
candidate_task_rows <- vector("list", nrow(task_table))
fold_task_rows <- vector("list", nrow(task_table))
task_seconds <- numeric(nrow(task_table))

run_start <- proc.time()[["elapsed"]]

for (task_index in seq_len(nrow(task_table))) {
  task <- task_table[task_index, , drop = FALSE]
  scenario_index <- match(task$Scenario, scenario_names)
  inner <- cv$inner_splits[[task$SplitID]]
  task_start <- proc.time()[["elapsed"]]

  bundles <- vector("list", 4L)

  for (inner_fold in 1:4) {
    inner_split <- inner[[paste0("I", inner_fold)]]

    validation_seed <-
      SEED_GLOBAL +
      scenario_index * 1000000L +
      task$Repetition * 10000L +
      task$OuterFold * 1000L +
      inner_fold * 10L

    set.seed(validation_seed)
    n_validation <- as.integer(
      round(
        length(inner_split$analysis_ids) *
        EARLY_VALIDATION_FRACTION
      )
    )
    validation_ids <- sort(
      sample(
        inner_split$analysis_ids,
        n_validation,
        replace = FALSE
      )
    )
    fit_ids <- sort(
      setdiff(
        inner_split$analysis_ids,
        validation_ids
      )
    )

    fit_index <- match(fit_ids, ids)
    validation_index <- match(validation_ids, ids)
    assessment_index <- match(
      inner_split$assessment_ids,
      ids
    )

    x_center <- colMeans(
      geno[fit_index, , drop = FALSE]
    )
    x_scale <- apply(
      geno[fit_index, , drop = FALSE],
      2,
      sd
    )
    zero_variance <- x_scale < 1e-8
    x_scale[zero_variance] <- 1

    y_center <- mean(
      phenotype[fit_index, scenario_index]
    )
    y_scale <- max(
      sd(phenotype[fit_index, scenario_index]),
      1e-8
    )

    x_fit <- sweep(
      sweep(
        geno[fit_index, , drop = FALSE],
        2,
        x_center,
        "-"
      ),
      2,
      x_scale,
      "/"
    )
    x_validation <- sweep(
      sweep(
        geno[validation_index, , drop = FALSE],
        2,
        x_center,
        "-"
      ),
      2,
      x_scale,
      "/"
    )
    x_assessment <- sweep(
      sweep(
        geno[assessment_index, , drop = FALSE],
        2,
        x_center,
        "-"
      ),
      2,
      x_scale,
      "/"
    )

    y_fit <-
      (phenotype[fit_index, scenario_index] - y_center) /
      y_scale
    y_validation <-
      (phenotype[validation_index, scenario_index] - y_center) /
      y_scale

    bundles[[inner_fold]] <- list(
      x_fit = x_fit,
      x_validation = x_validation,
      x_assessment = x_assessment,
      y_fit = y_fit,
      y_validation = y_validation,
      y_assessment = phenotype[
        assessment_index,
        scenario_index
      ],
      y_center = y_center,
      y_scale = y_scale,
      zero_variance_markers = sum(zero_variance)
    )
  }

  base_seed <-
    SEED_GLOBAL +
    scenario_index * 10000000L +
    task$Repetition * 100000L +
    task$OuterFold * 10000L

  structure_summary_rows <- vector(
    "list",
    nrow(structure_grid)
  )
  structure_fold_rows <- vector(
    "list",
    nrow(structure_grid)
  )

  for (candidate_index in seq_len(nrow(structure_grid))) {
    candidate <- structure_grid[
      candidate_index,
      ,
      drop = FALSE
    ]

    fold_rows <- vector("list", 4L)

    for (inner_fold in 1:4) {
      bundle <- bundles[[inner_fold]]
      fit_seed <- base_seed + 1000000L + inner_fold

      set.seed(as.integer(fit_seed))
      torch::torch_manual_seed(as.integer(fit_seed))

      x_fit_tensor <- torch::torch_tensor(
        bundle$x_fit,
        dtype = torch::torch_float()
      )$to(device = ANN_DEVICE)
      y_fit_tensor <- torch::torch_tensor(
        matrix(bundle$y_fit, ncol = 1),
        dtype = torch::torch_float()
      )$to(device = ANN_DEVICE)
      x_validation_tensor <- torch::torch_tensor(
        bundle$x_validation,
        dtype = torch::torch_float()
      )$to(device = ANN_DEVICE)
      y_validation_tensor <- torch::torch_tensor(
        matrix(bundle$y_validation, ncol = 1),
        dtype = torch::torch_float()
      )$to(device = ANN_DEVICE)
      x_assessment_tensor <- torch::torch_tensor(
        bundle$x_assessment,
        dtype = torch::torch_float()
      )$to(device = ANN_DEVICE)

      model <- TorchGenomicMLP(
        ncol(bundle$x_fit),
        ARCHITECTURES[[candidate$Architecture]]
      )$to(device = ANN_DEVICE)

      optimizer <- torch::optim_adam(
        params = model$parameters,
        lr = candidate$Learning_rate,
        weight_decay = 0
      )

      best_validation_loss <- Inf
      checkpoint_epoch <- 1L
      state <- model$state_dict()
      checkpoint_state <- state
      for (state_name in names(state)) {
        checkpoint_state[[state_name]] <-
          state[[state_name]]$detach()$clone()
      }
      stale_epochs <- 0L
      last_epoch <- 0L
      stopped_by_patience <- FALSE

      if (ANN_DEVICE_LABEL == "cuda") {
        torch::cuda_synchronize()
      }
      fit_start <- proc.time()[["elapsed"]]

      for (epoch in seq_len(MAX_EPOCHS)) {
        last_epoch <- epoch
        model$train()
        optimizer$zero_grad()

        prediction <- model(x_fit_tensor)
        penalty_l1 <- prediction$sum() * 0
        penalty_l2 <- prediction$sum() * 0

        for (layer_index in seq_len(length(model$linears))) {
          weight <- model$linears[[layer_index]]$weight
          penalty_l1 <- penalty_l1 + weight$abs()$sum()
          penalty_l2 <- penalty_l2 + weight$pow(2)$sum()
        }
        weight <- model$output$weight
        penalty_l1 <- penalty_l1 + weight$abs()$sum()
        penalty_l2 <- penalty_l2 + weight$pow(2)$sum()

        loss <-
          torch::nnf_mse_loss(
            prediction,
            y_fit_tensor
          ) +
          candidate$L1 * penalty_l1 +
          (candidate$L2 / 2) * penalty_l2

        loss$backward()
        optimizer$step()

        model$eval()
        validation_loss <- torch::with_no_grad({
          torch::nnf_mse_loss(
            model(x_validation_tensor),
            y_validation_tensor
          )
        })
        validation_value <- as.numeric(
          validation_loss$to(
            device = torch::torch_device("cpu")
          )
        )

        improved <-
          validation_value <
          best_validation_loss - MIN_DELTA

        if (improved) {
          best_validation_loss <- validation_value
          checkpoint_epoch <- epoch
          state <- model$state_dict()
          checkpoint_state <- state
          for (state_name in names(state)) {
            checkpoint_state[[state_name]] <-
              state[[state_name]]$detach()$clone()
          }
          stale_epochs <- 0L
        } else {
          stale_epochs <- stale_epochs + 1L
        }

        if (stale_epochs >= PATIENCE) {
          stopped_by_patience <- TRUE
          break
        }
      }

      if (ANN_DEVICE_LABEL == "cuda") {
        torch::cuda_synchronize()
      }
      fit_seconds <-
        proc.time()[["elapsed"]] - fit_start

      model$load_state_dict(checkpoint_state)
      model$eval()

      prediction_start <- proc.time()[["elapsed"]]
      prediction_scaled <- torch::with_no_grad({
        as.numeric(
          model(x_assessment_tensor)$to(
            device = torch::torch_device("cpu")
          )
        )
      })
      prediction_seconds <-
        proc.time()[["elapsed"]] - prediction_start

      prediction_assessment <-
        prediction_scaled * bundle$y_scale +
        bundle$y_center
      observed_assessment <- bundle$y_assessment

      fold_rows[[inner_fold]] <- data.frame(
        Model = "ANN_0",
        CandidateID = candidate_index,
        Architecture = candidate$Architecture,
        Learning_rate = candidate$Learning_rate,
        L1 = 0,
        L2 = 0,
        InnerFold = inner_fold,
        RMSE = sqrt(
          mean(
            (prediction_assessment - observed_assessment)^2
          )
        ),
        Correlation = cor(
          prediction_assessment,
          observed_assessment
        ),
        MAE = mean(
          abs(
            prediction_assessment - observed_assessment
          )
        ),
        EarlyStop_epoch = as.integer(checkpoint_epoch),
        Last_epoch = as.integer(last_epoch),
        Stopped_by_patience = stopped_by_patience,
        Hit_epoch_ceiling =
          last_epoch >= MAX_EPOCHS &&
          !stopped_by_patience,
        Fit_seconds = fit_seconds,
        Prediction_seconds = prediction_seconds,
        Status = "ok",
        Error = NA_character_,
        stringsAsFactors = FALSE
      )
    }

    folds <- do.call(rbind, fold_rows)
    hidden <- ARCHITECTURES[[candidate$Architecture]]
    network_inputs <- c(ncol(geno), hidden)
    network_outputs <- c(hidden, 1L)

    structure_summary_rows[[candidate_index]] <- data.frame(
      Model = "ANN_0",
      CandidateID = candidate_index,
      Architecture = candidate$Architecture,
      Learning_rate = candidate$Learning_rate,
      L1 = 0,
      L2 = 0,
      Inner_folds_ok = 4L,
      Mean_RMSE = mean(folds$RMSE),
      Mean_Correlation = mean(folds$Correlation),
      Mean_MAE = mean(folds$MAE),
      Median_EarlyStop_epoch = median(folds$EarlyStop_epoch),
      Max_Last_epoch = max(folds$Last_epoch),
      N_epoch_ceiling = sum(folds$Hit_epoch_ceiling),
      Parameters = sum(
        network_inputs * network_outputs +
        network_outputs
      ),
      Penalty = 0,
      Status = "ok",
      stringsAsFactors = FALSE
    )

    structure_fold_rows[[candidate_index]] <- folds
  }

  structure_summary <- do.call(
    rbind,
    structure_summary_rows
  )
  structure_folds <- do.call(
    rbind,
    structure_fold_rows
  )

  structure_summary <- structure_summary[
    order(
      structure_summary$Mean_RMSE,
      -structure_summary$Mean_Correlation,
      structure_summary$Parameters,
      structure_summary$Penalty,
      structure_summary$Learning_rate,
      structure_summary$CandidateID
    ),
    ,
    drop = FALSE
  ]

  chosen_structure <- structure_summary[1, , drop = FALSE]
  selected_architecture <- chosen_structure$Architecture
  selected_learning_rate <- chosen_structure$Learning_rate

  penalty_grids <- list(
    ANN_L1 = l1_grid,
    ANN_L2 = l2_grid,
    ANN_L1_L2 = l1l2_grid
  )
  penalty_summary_all <- list()
  penalty_fold_all <- list()
  penalty_selected <- list()

  for (model_name in names(penalty_grids)) {
    grid <- penalty_grids[[model_name]]
    model_summary_rows <- vector("list", nrow(grid))
    model_fold_rows <- vector("list", nrow(grid))

    for (candidate_index in seq_len(nrow(grid))) {
      fold_rows <- vector("list", 4L)

      for (inner_fold in 1:4) {
        bundle <- bundles[[inner_fold]]
        fit_seed <- base_seed + 2000000L + inner_fold

        set.seed(as.integer(fit_seed))
        torch::torch_manual_seed(as.integer(fit_seed))

        x_fit_tensor <- torch::torch_tensor(
          bundle$x_fit,
          dtype = torch::torch_float()
        )$to(device = ANN_DEVICE)
        y_fit_tensor <- torch::torch_tensor(
          matrix(bundle$y_fit, ncol = 1),
          dtype = torch::torch_float()
        )$to(device = ANN_DEVICE)
        x_validation_tensor <- torch::torch_tensor(
          bundle$x_validation,
          dtype = torch::torch_float()
        )$to(device = ANN_DEVICE)
        y_validation_tensor <- torch::torch_tensor(
          matrix(bundle$y_validation, ncol = 1),
          dtype = torch::torch_float()
        )$to(device = ANN_DEVICE)
        x_assessment_tensor <- torch::torch_tensor(
          bundle$x_assessment,
          dtype = torch::torch_float()
        )$to(device = ANN_DEVICE)

        model <- TorchGenomicMLP(
          ncol(bundle$x_fit),
          ARCHITECTURES[[selected_architecture]]
        )$to(device = ANN_DEVICE)

        optimizer <- torch::optim_adam(
          params = model$parameters,
          lr = selected_learning_rate,
          weight_decay = 0
        )

        best_validation_loss <- Inf
        checkpoint_epoch <- 1L
        state <- model$state_dict()
        checkpoint_state <- state
        for (state_name in names(state)) {
          checkpoint_state[[state_name]] <-
            state[[state_name]]$detach()$clone()
        }
        stale_epochs <- 0L
        last_epoch <- 0L
        stopped_by_patience <- FALSE

        if (ANN_DEVICE_LABEL == "cuda") {
          torch::cuda_synchronize()
        }
        fit_start <- proc.time()[["elapsed"]]

        for (epoch in seq_len(MAX_EPOCHS)) {
          last_epoch <- epoch
          model$train()
          optimizer$zero_grad()

          prediction <- model(x_fit_tensor)
          penalty_l1 <- prediction$sum() * 0
          penalty_l2 <- prediction$sum() * 0

          for (layer_index in seq_len(length(model$linears))) {
            weight <- model$linears[[layer_index]]$weight
            penalty_l1 <- penalty_l1 + weight$abs()$sum()
            penalty_l2 <- penalty_l2 + weight$pow(2)$sum()
          }
          weight <- model$output$weight
          penalty_l1 <- penalty_l1 + weight$abs()$sum()
          penalty_l2 <- penalty_l2 + weight$pow(2)$sum()

          loss <-
            torch::nnf_mse_loss(
              prediction,
              y_fit_tensor
            ) +
            grid$L1[candidate_index] * penalty_l1 +
            (grid$L2[candidate_index] / 2) * penalty_l2

          loss$backward()
          optimizer$step()

          model$eval()
          validation_loss <- torch::with_no_grad({
            torch::nnf_mse_loss(
              model(x_validation_tensor),
              y_validation_tensor
            )
          })
          validation_value <- as.numeric(
            validation_loss$to(
              device = torch::torch_device("cpu")
            )
          )

          improved <-
            validation_value <
            best_validation_loss - MIN_DELTA

          if (improved) {
            best_validation_loss <- validation_value
            checkpoint_epoch <- epoch
            state <- model$state_dict()
            checkpoint_state <- state
            for (state_name in names(state)) {
              checkpoint_state[[state_name]] <-
                state[[state_name]]$detach()$clone()
            }
            stale_epochs <- 0L
          } else {
            stale_epochs <- stale_epochs + 1L
          }

          if (stale_epochs >= PATIENCE) {
            stopped_by_patience <- TRUE
            break
          }
        }

        if (ANN_DEVICE_LABEL == "cuda") {
          torch::cuda_synchronize()
        }
        fit_seconds <-
          proc.time()[["elapsed"]] - fit_start

        model$load_state_dict(checkpoint_state)
        model$eval()

        prediction_start <- proc.time()[["elapsed"]]
        prediction_scaled <- torch::with_no_grad({
          as.numeric(
            model(x_assessment_tensor)$to(
              device = torch::torch_device("cpu")
            )
          )
        })
        prediction_seconds <-
          proc.time()[["elapsed"]] - prediction_start

        prediction_assessment <-
          prediction_scaled * bundle$y_scale +
          bundle$y_center
        observed_assessment <- bundle$y_assessment

        fold_rows[[inner_fold]] <- data.frame(
          Model = model_name,
          CandidateID = candidate_index,
          Architecture = selected_architecture,
          Learning_rate = selected_learning_rate,
          L1 = grid$L1[candidate_index],
          L2 = grid$L2[candidate_index],
          InnerFold = inner_fold,
          RMSE = sqrt(
            mean(
              (prediction_assessment - observed_assessment)^2
            )
          ),
          Correlation = cor(
            prediction_assessment,
            observed_assessment
          ),
          MAE = mean(
            abs(
              prediction_assessment - observed_assessment
            )
          ),
          EarlyStop_epoch = as.integer(checkpoint_epoch),
          Last_epoch = as.integer(last_epoch),
          Stopped_by_patience = stopped_by_patience,
          Hit_epoch_ceiling =
            last_epoch >= MAX_EPOCHS &&
            !stopped_by_patience,
          Fit_seconds = fit_seconds,
          Prediction_seconds = prediction_seconds,
          Status = "ok",
          Error = NA_character_,
          stringsAsFactors = FALSE
        )
      }

      folds <- do.call(rbind, fold_rows)
      hidden <- ARCHITECTURES[[selected_architecture]]
      network_inputs <- c(ncol(geno), hidden)
      network_outputs <- c(hidden, 1L)

      model_summary_rows[[candidate_index]] <- data.frame(
        Model = model_name,
        CandidateID = candidate_index,
        Architecture = selected_architecture,
        Learning_rate = selected_learning_rate,
        L1 = grid$L1[candidate_index],
        L2 = grid$L2[candidate_index],
        Inner_folds_ok = 4L,
        Mean_RMSE = mean(folds$RMSE),
        Mean_Correlation = mean(folds$Correlation),
        Mean_MAE = mean(folds$MAE),
        Median_EarlyStop_epoch = median(folds$EarlyStop_epoch),
        Max_Last_epoch = max(folds$Last_epoch),
        N_epoch_ceiling = sum(folds$Hit_epoch_ceiling),
        Parameters = sum(
          network_inputs * network_outputs +
          network_outputs
        ),
        Penalty =
          grid$L1[candidate_index] +
          grid$L2[candidate_index],
        Status = "ok",
        stringsAsFactors = FALSE
      )

      model_fold_rows[[candidate_index]] <- folds
    }

    model_summary <- do.call(rbind, model_summary_rows)
    model_folds <- do.call(rbind, model_fold_rows)

    model_summary <- model_summary[
      order(
        model_summary$Mean_RMSE,
        -model_summary$Mean_Correlation,
        model_summary$Parameters,
        model_summary$Penalty,
        model_summary$Learning_rate,
        model_summary$CandidateID
      ),
      ,
      drop = FALSE
    ]

    penalty_summary_all[[model_name]] <- model_summary
    penalty_fold_all[[model_name]] <- model_folds
    penalty_selected[[model_name]] <-
      model_summary[1, , drop = FALSE]
  }

  selected_table <- rbind(
    data.frame(
      Model = "ANN_0",
      Architecture = selected_architecture,
      Learning_rate = selected_learning_rate,
      L1 = 0,
      L2 = 0,
      Inner_RMSE = chosen_structure$Mean_RMSE,
      Inner_Correlation = chosen_structure$Mean_Correlation,
      Inner_MAE = chosen_structure$Mean_MAE,
      stringsAsFactors = FALSE
    ),
    data.frame(
      Model = "ANN_L1",
      Architecture = selected_architecture,
      Learning_rate = selected_learning_rate,
      L1 = penalty_selected$ANN_L1$L1,
      L2 = penalty_selected$ANN_L1$L2,
      Inner_RMSE = penalty_selected$ANN_L1$Mean_RMSE,
      Inner_Correlation = penalty_selected$ANN_L1$Mean_Correlation,
      Inner_MAE = penalty_selected$ANN_L1$Mean_MAE,
      stringsAsFactors = FALSE
    ),
    data.frame(
      Model = "ANN_L2",
      Architecture = selected_architecture,
      Learning_rate = selected_learning_rate,
      L1 = penalty_selected$ANN_L2$L1,
      L2 = penalty_selected$ANN_L2$L2,
      Inner_RMSE = penalty_selected$ANN_L2$Mean_RMSE,
      Inner_Correlation = penalty_selected$ANN_L2$Mean_Correlation,
      Inner_MAE = penalty_selected$ANN_L2$Mean_MAE,
      stringsAsFactors = FALSE
    ),
    data.frame(
      Model = "ANN_L1_L2",
      Architecture = selected_architecture,
      Learning_rate = selected_learning_rate,
      L1 = penalty_selected$ANN_L1_L2$L1,
      L2 = penalty_selected$ANN_L1_L2$L2,
      Inner_RMSE = penalty_selected$ANN_L1_L2$Mean_RMSE,
      Inner_Correlation = penalty_selected$ANN_L1_L2$Mean_Correlation,
      Inner_MAE = penalty_selected$ANN_L1_L2$Mean_MAE,
      stringsAsFactors = FALSE
    )
  )

  tuning_summary <- rbind(
    structure_summary,
    penalty_summary_all$ANN_L1,
    penalty_summary_all$ANN_L2,
    penalty_summary_all$ANN_L1_L2
  )
  tuning_folds <- rbind(
    structure_folds,
    penalty_fold_all$ANN_L1,
    penalty_fold_all$ANN_L2,
    penalty_fold_all$ANN_L1_L2
  )

  for (column_name in c(
    "Scenario",
    "Repetition",
    "Fold",
    "SplitID",
    "TaskID"
  )) {
    selected_table[[column_name]] <- task[[column_name]]
    tuning_summary[[column_name]] <- task[[column_name]]
    tuning_folds[[column_name]] <- task[[column_name]]
  }

  selected_task_rows[[task_index]] <- selected_table
  candidate_task_rows[[task_index]] <- tuning_summary
  fold_task_rows[[task_index]] <- tuning_folds
  task_seconds[task_index] <-
    proc.time()[["elapsed"]] - task_start

  cat(
    sprintf(
      "[%03d/%03d] %s | %.1f s | arch=%s lr=%g\n",
      task_index,
      nrow(task_table),
      task$TaskID,
      task_seconds[task_index],
      selected_architecture,
      selected_learning_rate
    )
  )
}

selected_configurations <- do.call(
  rbind,
  selected_task_rows
)
candidate_summary <- do.call(
  rbind,
  candidate_task_rows
)
fold_results <- do.call(
  rbind,
  fold_task_rows
)

rownames(selected_configurations) <- NULL
rownames(candidate_summary) <- NULL
rownames(fold_results) <- NULL

run_seconds <-
  proc.time()[["elapsed"]] - run_start

manifest <- list(
  analysis = "canonical_simulated_ann_tuning",
  dataset = "simulated",
  mode = "full",
  pilot_task = NA_character_,
  config_id = ANN_TUNING_CONFIG_ID,
  seed_global = SEED_GLOBAL,
  models = MODEL_LEVELS,
  architectures = ARCHITECTURES,
  learning_rates = LEARNING_RATES,
  l1_values = L1_VALUES,
  l2_values = L2_VALUES,
  l1_l2_grid = L1_L2_GRID,
  optimizer = "Adam",
  activation = "ReLU",
  dropout = DROPOUT,
  max_epochs = MAX_EPOCHS,
  final_train_epochs = TUNING_V1_COMPATIBILITY_TOKEN,
  patience = PATIENCE,
  min_delta = MIN_DELTA,
  early_validation_fraction = EARLY_VALIDATION_FRACTION,
  primary_tuning_metric = "mean inner-assessment RMSE",
  secondary_tuning_metric = "mean inner-assessment correlation",
  epoch_used_in_candidate_ranking = FALSE,
  outer_assessment_used = FALSE,
  true_genetic_value_used = FALSE,
  architecture_lr_shared_across_models = TRUE,
  candidates_per_task = CANDIDATES_PER_TASK,
  inner_fits_per_task = INNER_FITS_PER_TASK,
  number_tasks = nrow(task_table),
  selected_rows = nrow(selected_configurations),
  candidate_rows = nrow(candidate_summary),
  fold_rows = nrow(fold_results),
  torch_version = as.character(packageVersion("torch")),
  device = ANN_DEVICE_LABEL,
  cuda_available = cuda_available,
  current_session_seconds = run_seconds,
  checkpoint_task_seconds = sum(task_seconds),
  timestamp = as.character(Sys.time())
)

manifest_table <- data.frame(
  Field = c(
    "analysis",
    "mode",
    "config_id",
    "candidates_per_task",
    "inner_fits_per_task",
    "tasks",
    "outer_assessment_used",
    "true_genetic_value_used"
  ),
  Value = c(
    manifest$analysis,
    manifest$mode,
    manifest$config_id,
    manifest$candidates_per_task,
    manifest$inner_fits_per_task,
    manifest$number_tasks,
    FALSE,
    FALSE
  ),
  stringsAsFactors = FALSE
)

saveRDS(
  selected_configurations,
  SELECTED_RDS,
  compress = "gzip"
)
saveRDS(
  candidate_summary,
  CANDIDATE_RDS,
  compress = "gzip"
)
saveRDS(
  fold_results,
  FOLD_RDS,
  compress = "gzip"
)
saveRDS(
  manifest,
  RUN_RDS,
  compress = "gzip"
)
saveRDS(
  manifest,
  MANIFEST_RDS,
  compress = "gzip"
)

write.csv(
  selected_configurations,
  SELECTED_CSV,
  row.names = FALSE
)
write.csv(
  candidate_summary,
  CANDIDATE_CSV,
  row.names = FALSE
)
write.csv(
  fold_results,
  FOLD_CSV,
  row.names = FALSE
)
write.csv(
  manifest_table,
  MANIFEST_CSV,
  row.names = FALSE
)
```

### 5.3 Candidate ranking

Within each tuning stage, candidate ranking follows the same deterministic
hierarchy:

1. lower mean inner-assessment RMSE;
2. higher mean inner-assessment **Pearson correlation**;
3. fewer network parameters;
4. smaller total penalty coefficient;
5. smaller learning rate;
6. candidate identifier.

Early-stop epoch information is stored as an optimization diagnostic but does
not participate in candidate ranking.

## 6. Selected configurations and tuning diagnostics

The normal page build reads the previously completed tuning outputs rather than
rerunning the 42,000 candidate-fold fits.

```{r saved-tuning-results}
selected_configurations <- readRDS(SELECTED_RDS)
candidate_summary <- readRDS(CANDIDATE_RDS)
fold_results <- readRDS(FOLD_RDS)
tuning_metadata <- readRDS(RUN_RDS)

saved_tuning_summary <- data.frame(
  Object = c(
    "Selected configurations",
    "Candidate summaries",
    "Inner candidate-fold results"
  ),
  Rows = c(
    nrow(selected_configurations),
    nrow(candidate_summary),
    nrow(fold_results)
  ),
  stringsAsFactors = FALSE
)

saved_tuning_summary |>
  knitr::kable(
    format = "html",
    caption = "Stored ANN tuning results",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

Architecture and learning rate are shared across all four ANN variants within an
outer task. Their frequencies therefore describe the structure selected before
model-specific penalty tuning.

```{r tuning-frequency-summaries}
task_structure <- unique(
  selected_configurations[
    , c("TaskID", "Architecture", "Learning_rate")
  ]
)

architecture_frequency <- as.data.frame(
  table(task_structure$Architecture),
  stringsAsFactors = FALSE
)
names(architecture_frequency) <- c(
  "Architecture",
  "Selected_tasks"
)

learning_rate_frequency <- as.data.frame(
  table(task_structure$Learning_rate),
  stringsAsFactors = FALSE
)
names(learning_rate_frequency) <- c(
  "Learning_rate",
  "Selected_tasks"
)

architecture_frequency |>
  knitr::kable(
    format = "html",
    caption = "Selected shared architecture across the 150 outer tasks",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )

learning_rate_frequency |>
  knitr::kable(
    format = "html",
    caption = "Selected shared learning rate across the 150 outer tasks",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The optimization diagnostics below are descriptive. Reaching an epoch cap or a
search boundary does not retrospectively alter the tuning grid.

```{r tuning-diagnostics}
optimization_summary <- data.frame(
  Diagnostic = c(
    "Inner candidate-fold fits",
    "Median best early-stop epoch",
    "Median last epoch",
    "Fits stopped by patience",
    "Fits reaching 2000-epoch cap"
  ),
  Value = c(
    nrow(fold_results),
    median(fold_results$EarlyStop_epoch, na.rm = TRUE),
    median(fold_results$Last_epoch, na.rm = TRUE),
    sum(fold_results$Stopped_by_patience, na.rm = TRUE),
    sum(fold_results$Hit_epoch_ceiling, na.rm = TRUE)
  ),
  stringsAsFactors = FALSE
)

optimization_summary |>
  knitr::kable(
    format = "html",
    caption = "ANN tuning optimization diagnostics",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

## 7. Final fitting

Nested tuning returns one architecture, learning rate, and applicable penalty
coefficient(s) for each `Scenario × outer split × ANN variant`. Final fitting
then uses those choices without reopening the hyperparameter search.

The final sequence is:

1. take the selected tuning configuration;
2. split only the 800 outer-analysis individuals into fitting and
   early-validation subsets;
3. select `Best_epoch` under the 3000-epoch optimization cap;
4. reinitialize the network;
5. refit on all 800 outer-analysis individuals for exactly `Best_epoch`;
6. predict the untouched 200 outer-assessment individuals;
7. only then access outer-assessment phenotype and true genetic value for
   evaluation.

The complete final-fitting code is shown directly below without local training,
preprocessing, or metric wrappers.

```{r run-final-ann, eval=RUN_ANN}
dir.create(
  "output/simulated/ann/final",
  recursive = TRUE,
  showWarnings = FALSE
)

outer_map <- unique(
  cv$outer_assignments[
    , c("Repetition", "OuterFold", "SplitID")
  ]
)
outer_map <- outer_map[
  order(outer_map$Repetition, outer_map$OuterFold),
  ,
  drop = FALSE
]

final_task_table <- merge(
  data.frame(
    Scenario = scenario_names,
    stringsAsFactors = FALSE
  ),
  outer_map,
  by = NULL
)
final_task_table <- final_task_table[
  order(
    match(final_task_table$Scenario, scenario_names),
    final_task_table$Repetition,
    final_task_table$OuterFold
  ),
  ,
  drop = FALSE
]
final_task_table$Fold <- final_task_table$OuterFold
final_task_table$TaskID <- paste(
  final_task_table$Scenario,
  final_task_table$SplitID,
  sep = "_"
)
rownames(final_task_table) <- NULL

result_task_rows <- vector(
  "list",
  nrow(final_task_table)
)
prediction_task_rows <- vector(
  "list",
  nrow(final_task_table)
)
epoch_task_rows <- vector(
  "list",
  nrow(final_task_table)
)
final_task_seconds <- numeric(
  nrow(final_task_table)
)

run_start <- proc.time()[["elapsed"]]

for (task_index in seq_len(nrow(final_task_table))) {
  task <- final_task_table[
    task_index,
    ,
    drop = FALSE
  ]
  scenario_index <- match(
    task$Scenario,
    scenario_names
  )
  outer <- cv$outer_splits[[task$SplitID]]
  selected_for_task <- selected_configurations[
    selected_configurations$TaskID == task$TaskID,
    ,
    drop = FALSE
  ]

  seed_base <-
    SEED_GLOBAL +
    scenario_index * 10000000L +
    task$Repetition * 100000L +
    task$OuterFold * 10000L

  split_seed <- seed_base + 8000000L
  epoch_seed <- seed_base + 8100000L
  final_fit_seed <- seed_base + 8200000L

  result_rows <- vector(
    "list",
    length(MODEL_LEVELS)
  )
  prediction_rows <- vector(
    "list",
    length(MODEL_LEVELS)
  )
  epoch_rows <- vector(
    "list",
    length(MODEL_LEVELS)
  )

  task_start <- proc.time()[["elapsed"]]

  for (model_index in seq_along(MODEL_LEVELS)) {
    model_name <- MODEL_LEVELS[model_index]
    config <- selected_for_task[
      selected_for_task$Model == model_name,
      ,
      drop = FALSE
    ]

    # Select Best_epoch using only the 800 outer-analysis individuals.
    set.seed(split_seed)
    n_validation <- as.integer(
      round(
        length(outer$analysis_ids) *
        EARLY_VALIDATION_FRACTION
      )
    )
    validation_ids <- sort(
      sample(
        outer$analysis_ids,
        n_validation,
        replace = FALSE
      )
    )
    fit_ids <- sort(
      setdiff(
        outer$analysis_ids,
        validation_ids
      )
    )

    fit_index <- match(fit_ids, ids)
    validation_index <- match(validation_ids, ids)

    x_center <- colMeans(
      geno[fit_index, , drop = FALSE]
    )
    x_scale <- apply(
      geno[fit_index, , drop = FALSE],
      2,
      sd
    )
    x_scale[x_scale < 1e-8] <- 1

    y_center <- mean(
      phenotype[fit_index, scenario_index]
    )
    y_scale <- max(
      sd(phenotype[fit_index, scenario_index]),
      1e-8
    )

    x_fit <- sweep(
      sweep(
        geno[fit_index, , drop = FALSE],
        2,
        x_center,
        "-"
      ),
      2,
      x_scale,
      "/"
    )
    x_validation <- sweep(
      sweep(
        geno[validation_index, , drop = FALSE],
        2,
        x_center,
        "-"
      ),
      2,
      x_scale,
      "/"
    )
    y_fit <-
      (phenotype[fit_index, scenario_index] - y_center) /
      y_scale
    y_validation <-
      (phenotype[validation_index, scenario_index] - y_center) /
      y_scale

    set.seed(as.integer(epoch_seed))
    torch::torch_manual_seed(as.integer(epoch_seed))

    x_fit_tensor <- torch::torch_tensor(
      x_fit,
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)
    y_fit_tensor <- torch::torch_tensor(
      matrix(y_fit, ncol = 1),
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)
    x_validation_tensor <- torch::torch_tensor(
      x_validation,
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)
    y_validation_tensor <- torch::torch_tensor(
      matrix(y_validation, ncol = 1),
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)

    model <- TorchGenomicMLP(
      ncol(x_fit),
      ARCHITECTURES[[config$Architecture]]
    )$to(device = ANN_DEVICE)
    optimizer <- torch::optim_adam(
      params = model$parameters,
      lr = config$Learning_rate,
      weight_decay = 0
    )

    best_validation_loss <- Inf
    best_epoch <- 1L
    state <- model$state_dict()
    best_state <- state
    for (state_name in names(state)) {
      best_state[[state_name]] <-
        state[[state_name]]$detach()$clone()
    }
    stale_epochs <- 0L
    last_epoch <- 0L
    stopped_by_patience <- FALSE

    if (ANN_DEVICE_LABEL == "cuda") {
      torch::cuda_synchronize()
    }
    epoch_fit_start <- proc.time()[["elapsed"]]

    for (epoch in seq_len(FINAL_EPOCH_MAX_EPOCHS)) {
      last_epoch <- epoch
      model$train()
      optimizer$zero_grad()

      prediction <- model(x_fit_tensor)
      penalty_l1 <- prediction$sum() * 0
      penalty_l2 <- prediction$sum() * 0

      for (layer_index in seq_len(length(model$linears))) {
        weight <- model$linears[[layer_index]]$weight
        penalty_l1 <- penalty_l1 + weight$abs()$sum()
        penalty_l2 <- penalty_l2 + weight$pow(2)$sum()
      }
      weight <- model$output$weight
      penalty_l1 <- penalty_l1 + weight$abs()$sum()
      penalty_l2 <- penalty_l2 + weight$pow(2)$sum()

      loss <-
        torch::nnf_mse_loss(
          prediction,
          y_fit_tensor
        ) +
        config$L1 * penalty_l1 +
        (config$L2 / 2) * penalty_l2

      loss$backward()
      optimizer$step()

      model$eval()
      validation_loss <- torch::with_no_grad({
        torch::nnf_mse_loss(
          model(x_validation_tensor),
          y_validation_tensor
        )
      })
      validation_value <- as.numeric(
        validation_loss$to(
          device = torch::torch_device("cpu")
        )
      )

      improved <-
        validation_value <
        best_validation_loss - MIN_DELTA

      if (improved) {
        best_validation_loss <- validation_value
        best_epoch <- epoch
        state <- model$state_dict()
        best_state <- state
        for (state_name in names(state)) {
          best_state[[state_name]] <-
            state[[state_name]]$detach()$clone()
        }
        stale_epochs <- 0L
      } else {
        stale_epochs <- stale_epochs + 1L
      }

      if (stale_epochs >= PATIENCE) {
        stopped_by_patience <- TRUE
        break
      }
    }

    if (ANN_DEVICE_LABEL == "cuda") {
      torch::cuda_synchronize()
    }
    epoch_fit_seconds <-
      proc.time()[["elapsed"]] - epoch_fit_start

    model$load_state_dict(best_state)
    model$eval()
    validation_prediction_scaled <- torch::with_no_grad({
      as.numeric(
        model(x_validation_tensor)$to(
          device = torch::torch_device("cpu")
        )
      )
    })
    validation_prediction <-
      validation_prediction_scaled * y_scale +
      y_center
    validation_observed <- phenotype[
      validation_index,
      scenario_index
    ]

    validation_rmse <- sqrt(
      mean(
        (validation_prediction - validation_observed)^2
      )
    )
    validation_correlation <- cor(
      validation_prediction,
      validation_observed
    )

    # Refit all 800 outer-analysis individuals for exactly Best_epoch.
    train_index <- match(outer$analysis_ids, ids)
    assessment_index <- match(
      outer$assessment_ids,
      ids
    )

    x_center <- colMeans(
      geno[train_index, , drop = FALSE]
    )
    x_scale <- apply(
      geno[train_index, , drop = FALSE],
      2,
      sd
    )
    zero_variance <- x_scale < 1e-8
    x_scale[zero_variance] <- 1

    y_center <- mean(
      phenotype[train_index, scenario_index]
    )
    y_scale <- max(
      sd(phenotype[train_index, scenario_index]),
      1e-8
    )

    x_train <- sweep(
      sweep(
        geno[train_index, , drop = FALSE],
        2,
        x_center,
        "-"
      ),
      2,
      x_scale,
      "/"
    )
    x_assessment <- sweep(
      sweep(
        geno[assessment_index, , drop = FALSE],
        2,
        x_center,
        "-"
      ),
      2,
      x_scale,
      "/"
    )
    y_train <-
      (phenotype[train_index, scenario_index] - y_center) /
      y_scale

    set.seed(as.integer(final_fit_seed))
    torch::torch_manual_seed(as.integer(final_fit_seed))

    x_train_tensor <- torch::torch_tensor(
      x_train,
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)
    y_train_tensor <- torch::torch_tensor(
      matrix(y_train, ncol = 1),
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)
    x_assessment_tensor <- torch::torch_tensor(
      x_assessment,
      dtype = torch::torch_float()
    )$to(device = ANN_DEVICE)

    model <- TorchGenomicMLP(
      ncol(x_train),
      ARCHITECTURES[[config$Architecture]]
    )$to(device = ANN_DEVICE)
    optimizer <- torch::optim_adam(
      params = model$parameters,
      lr = config$Learning_rate,
      weight_decay = 0
    )

    if (ANN_DEVICE_LABEL == "cuda") {
      torch::cuda_synchronize()
    }
    final_fit_start <- proc.time()[["elapsed"]]

    for (epoch in seq_len(as.integer(best_epoch))) {
      model$train()
      optimizer$zero_grad()

      prediction <- model(x_train_tensor)
      penalty_l1 <- prediction$sum() * 0
      penalty_l2 <- prediction$sum() * 0

      for (layer_index in seq_len(length(model$linears))) {
        weight <- model$linears[[layer_index]]$weight
        penalty_l1 <- penalty_l1 + weight$abs()$sum()
        penalty_l2 <- penalty_l2 + weight$pow(2)$sum()
      }
      weight <- model$output$weight
      penalty_l1 <- penalty_l1 + weight$abs()$sum()
      penalty_l2 <- penalty_l2 + weight$pow(2)$sum()

      loss <-
        torch::nnf_mse_loss(
          prediction,
          y_train_tensor
        ) +
        config$L1 * penalty_l1 +
        (config$L2 / 2) * penalty_l2

      loss$backward()
      optimizer$step()
    }

    if (ANN_DEVICE_LABEL == "cuda") {
      torch::cuda_synchronize()
    }
    final_fit_seconds <-
      proc.time()[["elapsed"]] - final_fit_start

    model$eval()
    prediction_start <- proc.time()[["elapsed"]]
    predicted_scaled <- torch::with_no_grad({
      as.numeric(
        model(x_assessment_tensor)$to(
          device = torch::torch_device("cpu")
        )
      )
    })
    prediction_seconds <-
      proc.time()[["elapsed"]] - prediction_start

    predicted_value <-
      predicted_scaled * y_scale + y_center

    # Held-out targets are accessed only after prediction already exists.
    observed <- phenotype[
      assessment_index,
      scenario_index
    ]
    true_genetic_value <- common$true_genetic_value[
      assessment_index,
      scenario_index
    ]

    genetic_correlation <- cor(
      predicted_value,
      true_genetic_value
    )
    genetic_rmse <- sqrt(
      mean(
        (predicted_value - true_genetic_value)^2
      )
    )
    genetic_mae <- mean(
      abs(
        predicted_value - true_genetic_value
      )
    )
    genetic_bias <- mean(
      predicted_value - true_genetic_value
    )
    genetic_slope <- as.numeric(
      coef(
        lm(
          true_genetic_value ~ predicted_value
        )
      )[2]
    )
    genetic_r2 <- 1 -
      sum(
        (true_genetic_value - predicted_value)^2
      ) /
      sum(
        (
          true_genetic_value -
          mean(true_genetic_value)
        )^2
      )

    phenotype_correlation <- cor(
      predicted_value,
      observed
    )
    phenotype_rmse <- sqrt(
      mean(
        (predicted_value - observed)^2
      )
    )
    phenotype_mae <- mean(
      abs(
        predicted_value - observed
      )
    )
    phenotype_bias <- mean(
      predicted_value - observed
    )
    phenotype_slope <- as.numeric(
      coef(
        lm(observed ~ predicted_value)
      )[2]
    )
    phenotype_r2 <- 1 -
      sum(
        (observed - predicted_value)^2
      ) /
      sum(
        (observed - mean(observed))^2
      )

    meta <- control[
      control$Scenario == task$Scenario,
      ,
      drop = FALSE
    ]

    result_rows[[model_index]] <- data.frame(
      Dataset = "simulated",
      Scenario = task$Scenario,
      QTL_N = meta$QTL_N,
      H2_nominal = meta$H2_nominal,
      Repetition = task$Repetition,
      Fold = task$OuterFold,
      SplitID = task$SplitID,
      TaskID = task$TaskID,
      Model = model_name,
      Architecture = config$Architecture,
      Learning_rate = config$Learning_rate,
      L1 = config$L1,
      L2 = config$L2,
      Best_epoch = as.integer(best_epoch),
      Epoch_Last = as.integer(last_epoch),
      Epoch_Stopped_by_patience = stopped_by_patience,
      Epoch_Hit_ceiling =
        last_epoch >= FINAL_EPOCH_MAX_EPOCHS &&
        !stopped_by_patience,
      Inner_RMSE = config$Inner_RMSE,
      Inner_Correlation = config$Inner_Correlation,
      Inner_MAE = config$Inner_MAE,
      N_train = 800L,
      N_test = 200L,
      N_pred_finite = sum(is.finite(predicted_value)),
      N_zero_variance_markers = sum(zero_variance),
      Genetic_Correlation = genetic_correlation,
      Genetic_RMSE = genetic_rmse,
      Genetic_MAE = genetic_mae,
      Genetic_Bias = genetic_bias,
      Genetic_Slope = genetic_slope,
      Genetic_R2 = genetic_r2,
      Phenotype_Correlation = phenotype_correlation,
      Phenotype_RMSE = phenotype_rmse,
      Phenotype_MAE = phenotype_mae,
      Phenotype_Bias = phenotype_bias,
      Phenotype_Slope = phenotype_slope,
      Phenotype_R2 = phenotype_r2,
      Epoch_selection_seconds = epoch_fit_seconds,
      Final_fit_seconds = final_fit_seconds,
      Predict_seconds = prediction_seconds,
      Status = "ok",
      Error = NA_character_,
      stringsAsFactors = FALSE
    )

    prediction_rows[[model_index]] <- data.frame(
      Dataset = "simulated",
      Scenario = task$Scenario,
      QTL_N = meta$QTL_N,
      H2_nominal = meta$H2_nominal,
      Repetition = task$Repetition,
      Fold = task$OuterFold,
      SplitID = task$SplitID,
      TaskID = task$TaskID,
      Model = model_name,
      Architecture = config$Architecture,
      Learning_rate = config$Learning_rate,
      L1 = config$L1,
      L2 = config$L2,
      Best_epoch = as.integer(best_epoch),
      ID = ids[assessment_index],
      Observed_Phenotype = observed,
      True_Genetic_Value = true_genetic_value,
      Predicted_Value = predicted_value,
      Status = "ok",
      stringsAsFactors = FALSE
    )

    epoch_rows[[model_index]] <- data.frame(
      Scenario = task$Scenario,
      Repetition = task$Repetition,
      Fold = task$OuterFold,
      SplitID = task$SplitID,
      TaskID = task$TaskID,
      Model = model_name,
      Architecture = config$Architecture,
      Learning_rate = config$Learning_rate,
      L1 = config$L1,
      L2 = config$L2,
      Best_epoch = as.integer(best_epoch),
      Last_epoch = as.integer(last_epoch),
      Stopped_by_patience = stopped_by_patience,
      Hit_epoch_ceiling =
        last_epoch >= FINAL_EPOCH_MAX_EPOCHS &&
        !stopped_by_patience,
      Validation_RMSE = validation_rmse,
      Validation_Correlation = validation_correlation,
      N_epoch_fit = length(fit_ids),
      N_epoch_validation = length(validation_ids),
      Fit_seconds = epoch_fit_seconds,
      stringsAsFactors = FALSE
    )
  }

  result_task_rows[[task_index]] <- do.call(
    rbind,
    result_rows
  )
  prediction_task_rows[[task_index]] <- do.call(
    rbind,
    prediction_rows
  )
  epoch_task_rows[[task_index]] <- do.call(
    rbind,
    epoch_rows
  )

  final_task_seconds[task_index] <-
    proc.time()[["elapsed"]] - task_start

  cat(
    sprintf(
      "[%03d/%03d] %s | final %.1f min\n",
      task_index,
      nrow(final_task_table),
      task$TaskID,
      final_task_seconds[task_index] / 60
    )
  )
}

results <- do.call(rbind, result_task_rows)
predictions <- do.call(rbind, prediction_task_rows)
epoch_selection <- do.call(rbind, epoch_task_rows)

rownames(results) <- NULL
rownames(predictions) <- NULL
rownames(epoch_selection) <- NULL

run_seconds <-
  proc.time()[["elapsed"]] - run_start

final_meta <- list(
  dataset = "simulated",
  analysis = "canonical_simulated_ann",
  mode = "full",
  pilot_task = NA_character_,
  config_id = FINAL_CONFIG_ID,
  tuning_config_id = ANN_TUNING_CONFIG_ID,
  seed_global = SEED_GLOBAL,
  models = MODEL_LEVELS,
  tuning_max_epochs = MAX_EPOCHS,
  final_epoch_max_epochs = FINAL_EPOCH_MAX_EPOCHS,
  patience = PATIENCE,
  min_delta = MIN_DELTA,
  early_validation_fraction = EARLY_VALIDATION_FRACTION,
  final_refit_rule =
    "Best_epoch selected inside outer-analysis, then refit on all 800",
  outer_assessment_used_in_tuning = FALSE,
  outer_assessment_used_in_epoch_selection = FALSE,
  true_genetic_value_used_in_selection = FALSE,
  number_tasks = nrow(final_task_table),
  model_task_rows = nrow(results),
  prediction_rows = nrow(predictions),
  epoch_selection_rows = nrow(epoch_selection),
  torch_version = as.character(packageVersion("torch")),
  device = ANN_DEVICE_LABEL,
  cuda_available = cuda_available,
  current_session_seconds = run_seconds,
  checkpoint_task_seconds = sum(final_task_seconds),
  timestamp = as.character(Sys.time())
)

saveRDS(
  results,
  FINAL_RESULTS_RDS,
  compress = "gzip"
)
saveRDS(
  predictions,
  FINAL_PREDICTIONS_RDS,
  compress = "gzip"
)
saveRDS(
  epoch_selection,
  FINAL_EPOCH_RDS,
  compress = "gzip"
)
saveRDS(
  final_meta,
  FINAL_RUN_RDS,
  compress = "gzip"
)

write.csv(
  results,
  FINAL_RESULTS_CSV,
  row.names = FALSE
)
write.csv(
  predictions,
  FINAL_PREDICTIONS_CSV,
  row.names = FALSE
)
write.csv(
  epoch_selection,
  FINAL_EPOCH_CSV,
  row.names = FALSE
)
```

## 8. Saved ANN results

The normal page build reads the completed final ANN outputs. This keeps the
scientific results fixed while allowing the explanation, tables, and figures to
be rebuilt without reexecuting tuning or final fitting.

```{r saved-final-results}
results <- readRDS(FINAL_RESULTS_RDS)
predictions <- readRDS(FINAL_PREDICTIONS_RDS)
epoch_selection <- readRDS(FINAL_EPOCH_RDS)
final_meta <- readRDS(FINAL_RUN_RDS)

final_output_summary <- data.frame(
  Object = c(
    "Model-task results",
    "Held-out predictions",
    "Epoch selections"
  ),
  Rows = c(
    nrow(results),
    nrow(predictions),
    nrow(epoch_selection)
  ),
  Unique_tasks = c(
    length(unique(results$TaskID)),
    length(unique(predictions$TaskID)),
    length(unique(epoch_selection$TaskID))
  ),
  stringsAsFactors = FALSE
)

final_output_summary |>
  knitr::kable(
    format = "html",
    caption = "Stored final ANN result structure",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The stored benchmark contains **600 model-task results**, **120,000 held-out
predictions**, and **600 epoch-selection records**. The historical `Status =
"ok"` field indicates that a fit returned and produced stored predictions; it is
not interpreted as a formal proof of numerical convergence.

## 9. Predictive performance

Recovery of the known true genetic value is the primary evaluation target.
Phenotype prediction is retained as a secondary target because phenotype also
contains residual variation.

```{r performance-summary}
ok <- results[
  results$Status == "ok",
  ,
  drop = FALSE
]

scenario_performance <- aggregate(
  cbind(
    Genetic_Correlation,
    Genetic_RMSE,
    Genetic_MAE,
    Genetic_Bias,
    Genetic_Slope,
    Genetic_R2,
    Phenotype_Correlation,
    Phenotype_RMSE,
    Phenotype_MAE,
    Phenotype_Bias,
    Phenotype_Slope,
    Phenotype_R2
  ) ~ Scenario + Model + QTL_N + H2_nominal,
  data = ok,
  FUN = mean,
  na.rm = TRUE
)

scenario_performance <- scenario_performance[
  order(
    match(scenario_performance$Scenario, scenario_names),
    match(scenario_performance$Model, MODEL_LEVELS)
  ),
  ,
  drop = FALSE
]

scenario_performance[
  , c(
    "Scenario",
    "Model",
    "QTL_N",
    "H2_nominal",
    "Genetic_Correlation",
    "Genetic_RMSE",
    "Genetic_MAE",
    "Genetic_Bias",
    "Genetic_Slope",
    "Genetic_R2"
  )
] |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Primary evaluation: mean outer-split ANN performance against true genetic value",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

Here, larger **Pearson correlation** means stronger linear association with the
known genetic signal. Smaller RMSE and MAE indicate smaller absolute prediction
error. Bias near zero and calibration slope near one describe calibration on the
held-out scale.

```{r phenotype-performance}
scenario_performance[
  , c(
    "Scenario",
    "Model",
    "Phenotype_Correlation",
    "Phenotype_RMSE",
    "Phenotype_MAE",
    "Phenotype_Bias",
    "Phenotype_Slope",
    "Phenotype_R2"
  )
] |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Secondary evaluation: mean outer-split ANN performance against phenotype",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

### 9.1 Held-out performance across outer splits

The next figures display all 25 outer-split results for each scenario and ANN
variant. Their purpose is descriptive: repeated outer folds overlap in the
individuals used for model fitting, so the 25 values are **not treated as 25
statistically independent replicates or as inferential uncertainty**.

```{r genetic-correlation-figure, fig.cap="Held-out true genetic-value Pearson correlation for the four ANN variants across the 25 outer splits in each scenario. White diamonds indicate model means within scenario."}
p_ann_correlation <- ggplot(
  ok,
  aes(
    x = Model,
    y = Genetic_Correlation,
    fill = Model
  )
) +
  geom_boxplot(
    width = 0.62,
    outlier.shape = NA,
    alpha = 0.70
  ) +
  geom_point(
    position = position_jitter(
      width = 0.10,
      height = 0,
      seed = SEED_GLOBAL + 401L
    ),
    alpha = 0.35,
    size = 1.4
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 2.8,
    fill = "white"
  ) +
  facet_wrap(~ Scenario, ncol = 3) +
  scale_fill_gdocs() +
  labs(
    x = "ANN variant",
    y = "Pearson correlation with true genetic value",
    fill = NULL
  ) +
  theme_gdocs() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 35,
      hjust = 1
    )
  )

print(p_ann_correlation)
```

```{r save-ann-correlation-tiff, include=FALSE}
ggsave(
  "output/simulated/m04/figures/ann_genetic_correlation_by_model.tiff",
  plot = p_ann_correlation,
  device = "tiff",
  width = 11,
  height = 7.5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

```{r genetic-rmse-figure, fig.cap="Held-out true genetic-value RMSE for the four ANN variants across the 25 outer splits in each scenario. Lower values indicate smaller prediction error; white diamonds indicate model means within scenario."}
p_ann_rmse <- ggplot(
  ok,
  aes(
    x = Model,
    y = Genetic_RMSE,
    fill = Model
  )
) +
  geom_boxplot(
    width = 0.62,
    outlier.shape = NA,
    alpha = 0.70
  ) +
  geom_point(
    position = position_jitter(
      width = 0.10,
      height = 0,
      seed = SEED_GLOBAL + 402L
    ),
    alpha = 0.35,
    size = 1.4
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 2.8,
    fill = "white"
  ) +
  facet_wrap(~ Scenario, ncol = 3) +
  scale_fill_gdocs() +
  labs(
    x = "ANN variant",
    y = "RMSE for true genetic value",
    fill = NULL
  ) +
  theme_gdocs() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 35,
      hjust = 1
    )
  )

print(p_ann_rmse)
```

```{r save-ann-rmse-tiff, include=FALSE}
ggsave(
  "output/simulated/m04/figures/ann_genetic_rmse_by_model.tiff",
  plot = p_ann_rmse,
  device = "tiff",
  width = 11,
  height = 7.5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

### 9.2 Overall ANN comparison

```{r overall-ann-summary}
overall_ann <- aggregate(
  cbind(
    Genetic_Correlation,
    Genetic_RMSE,
    Genetic_R2
  ) ~ Model,
  data = ok,
  FUN = mean,
  na.rm = TRUE
)

overall_ann <- overall_ann[
  match(MODEL_LEVELS, overall_ann$Model),
  ,
  drop = FALSE
]

overall_ann |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Overall ANN performance against true genetic value",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

Across all outer tasks, ANN-L1+L2 has the highest mean genetic correlation and
lowest mean genetic RMSE among the four ANN variants, but its advantage over
ANN-L1 is very small. ANN-L2 remains close to ANN_0. Within this bounded search
space, the observed regularization gain is therefore driven mainly by the L1
component. The formal task-matched comparison with GBLUP-ADE is reserved for
M05.

## 10. Navigation

[Previous: Validation design](simulated_data_validation.html) | [Parallel branch: GBLUP-ADE](simulated_data_gblup.html) | [Next: Model comparison](simulated_data_comparison.html)
