---
title: "Canonical comparison of GBLUP-ADE and regularized ANN models"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    code_download: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)

options(stringsAsFactors = FALSE)
```

```{r format-table, include=FALSE}
format_table <- function(
  x,
  caption = NULL,
  digits = getOption("digits"),
  row.names = NA,
  ...
) {
  table <- knitr::kable(
    x,
    format = "html",
    caption = caption,
    digits = digits,
    row.names = row.names,
    table.attr = paste0(
      'class="table table-striped table-hover table-bordered" ',
      'style="width:100%; margin-bottom:0;"'
    ),
    ...
  )

  knitr::asis_output(
    paste0(
      '<div style="width:100%; overflow-x:auto; margin-bottom:1.25rem;">',
      table,
      '</div>'
    )
  )
}
```

## 1. Objective

This module provides the terminal comparison between GBLUP-ADE and the four ANN
variants under the frozen outer-CV design. All competing methods are evaluated
on identical outer-assessment individuals.

Prediction of the known **true simulated genetic value** is the primary outcome.
Prediction of the **observed phenotype** is retained as a secondary outcome.
Results are summarized overall, by simulation scenario, and through task-matched
ANN versus GBLUP-ADE contrasts.

The module does not fit models, tune hyperparameters, or alter predictions. It
combines the canonical outputs from M03 and M04 and verifies that the comparison
is made on the same tasks, individuals, and evaluation targets.

## 2. Packages and canonical inputs

### 2.1 Packages

Only `knitr` is required directly for reader-facing table presentation in this
comparison module. The statistical summaries use base R functions.

```{r packages}
library(knitr) # Format reader-facing tables and R Markdown output.
```

The paths below identify the frozen GBLUP-ADE and ANN results consumed by this
module, together with the versioned comparison tables that may be regenerated
explicitly when `WRITE_COMPARISON_OUTPUTS=1`.

```{r comparison-paths}
PROJECT_ROOT <- "."
project_path <- function(...) file.path(PROJECT_ROOT, ...)

GBLUP_RESULTS_RDS <- project_path(
  "output", "simulated", "gblup", "simulated_gblup_ade_results.rds"
)
GBLUP_PREDICTIONS_RDS <- project_path(
  "output", "simulated", "gblup", "simulated_gblup_ade_predictions.rds"
)
ANN_RESULTS_RDS <- project_path(
  "output", "simulated", "ann", "final", "simulated_ann_results.rds"
)
ANN_PREDICTIONS_RDS <- project_path(
  "output", "simulated", "ann", "final", "simulated_ann_predictions.rds"
)

COMPARISON_DIR <- project_path("output", "simulated", "comparison")
OVERALL_CSV <- file.path(
  COMPARISON_DIR, "simulated_model_overall_summary.csv"
)
SCENARIO_CSV <- file.path(
  COMPARISON_DIR, "simulated_model_scenario_summary.csv"
)
PAIRED_CSV <- file.path(
  COMPARISON_DIR, "simulated_ann_vs_gblup_paired_summary.csv"
)
PAIRED_SCENARIO_CSV <- file.path(
  COMPARISON_DIR, "simulated_ann_vs_gblup_paired_by_scenario.csv"
)

WRITE_COMPARISON_OUTPUTS <- identical(
  Sys.getenv("WRITE_COMPARISON_OUTPUTS", "0"),
  "1"
)

metrics <- c(
  "Genetic_Correlation",
  "Genetic_RMSE",
  "Genetic_MAE",
  "Genetic_Bias",
  "Genetic_Slope",
  "Genetic_R2",
  "Phenotype_Correlation",
  "Phenotype_RMSE",
  "Phenotype_MAE",
  "Phenotype_Bias",
  "Phenotype_Slope",
  "Phenotype_R2"
)

expected_ann_models <- c(
  "ANN_0",
  "ANN_L1",
  "ANN_L2",
  "ANN_L1_L2"
)
```

### 2.2 GBLUP-ADE results

The GBLUP-ADE branch contributes one result row for each of the 150 frozen outer
tasks and one held-out prediction for each of the 200 assessment individuals in
each task.

### 2.3 ANN results

The ANN branch contributes the same 150 outer tasks for each of four ANN
variants, yielding 600 task-by-model result rows and 120,000 held-out
predictions.

```{r load-canonical-inputs}
gr <- readRDS(GBLUP_RESULTS_RDS)
gp <- readRDS(GBLUP_PREDICTIONS_RDS)
ar <- readRDS(ANN_RESULTS_RDS)
ap <- readRDS(ANN_PREDICTIONS_RDS)

input_rows <- data.frame(
  Object = c(
    "GBLUP task results", "GBLUP predictions",
    "ANN task-by-model results", "ANN predictions"
  ),
  Rows = c(nrow(gr), nrow(gp), nrow(ar), nrow(ap))
)
format_table(
  input_rows,
  caption = "Canonical model-comparison inputs",
  row.names = FALSE
)
```

The displayed row counts provide the first check that the comparison consumes
the finalized outputs from both modeling branches rather than refitting either
method.

## 3. Fair-comparison contract

The comparison is meaningful only if all methods use the same frozen
outer-assessment design. The audit below therefore verifies the task structure,
assessment sample sizes, and equality of the two evaluation targets before any
performance summary is interpreted.

```{r compatibility-audit}
comparison_keys <- c(
  "Scenario", "Repetition", "Fold", "SplitID", "TaskID", "ID"
)
g_targets <- gp[
  , c(comparison_keys, "Observed_Phenotype", "True_Genetic_Value")
]
paired_targets <- merge(
  g_targets,
  ap[, c(
    comparison_keys, "Model", "Observed_Phenotype", "True_Genetic_Value"
  )],
  by = comparison_keys,
  suffixes = c("_GBLUP", "_ANN"),
  sort = FALSE
)
paired_targets$Phenotype_difference <- abs(
  paired_targets$Observed_Phenotype_GBLUP -
    paired_targets$Observed_Phenotype_ANN
)
paired_targets$TrueG_difference <- abs(
  paired_targets$True_Genetic_Value_GBLUP -
    paired_targets$True_Genetic_Value_ANN
)

model_structure <- do.call(
  rbind,
  lapply(c("GBLUP_ADE", expected_ann_models), function(model) {
    if (model == "GBLUP_ADE") {
      result_data <- gr
      prediction_data <- gp
    } else {
      result_data <- ar[ar$Model == model, , drop = FALSE]
      prediction_data <- ap[ap$Model == model, , drop = FALSE]
    }
    predictions_per_task <- as.integer(table(prediction_data$TaskID))
    data.frame(
      Model = model,
      Tasks = length(unique(result_data$TaskID)),
      Predictions = nrow(prediction_data),
      Min_predictions_per_task = min(predictions_per_task),
      Max_predictions_per_task = max(predictions_per_task)
    )
  })
)

target_match <- aggregate(
  cbind(Phenotype_difference, TrueG_difference) ~ Model,
  data = paired_targets,
  FUN = max
)
```

### 3.1 Matched tasks

Each model must contribute the same frozen set of 150 `Scenario × outer split`
tasks. The table also reports the number of held-out predictions and the
assessment size observed within each task.

```{r comparison-task-structure}
format_table(
  model_structure,
  caption = "Observed task and assessment structure by model",
  row.names = FALSE
)
```

### 3.2 Matched assessment individuals

The prediction-level merge uses
`Scenario + Repetition + Fold + SplitID + TaskID + ID`. Consequently, agreement
of the paired targets is evaluated only after the same assessment individual has
been aligned across GBLUP-ADE and each ANN variant.

### 3.3 Phenotype agreement

For each ANN variant, the maximum absolute difference between its held-out
phenotype and the matched GBLUP-ADE phenotype should be zero.

### 3.4 True-G agreement

The same check is performed for the known simulated genetic value, which is the
primary target of this study.

```{r comparison-target-agreement}
format_table(
  target_match,
  digits = 12,
  caption = "Maximum paired target difference relative to GBLUP-ADE",
  row.names = FALSE
)
```

The compact contract below summarizes all four observed pairing properties in a
single table.

```{r compatibility-summary}
comparison_contract <- data.frame(
  Quantity = c(
    "Outer tasks per model",
    "Assessment predictions per task",
    "Maximum paired phenotype difference",
    "Maximum paired true-G difference"
  ),
  Observed = c(
    paste(range(model_structure$Tasks), collapse = " to "),
    paste(
      range(c(
        model_structure$Min_predictions_per_task,
        model_structure$Max_predictions_per_task
      )),
      collapse = " to "
    ),
    max(target_match$Phenotype_difference),
    max(target_match$TrueG_difference)
  )
)
format_table(
  comparison_contract,
  caption = "Observed contract for the task-matched model comparison",
  row.names = FALSE
)
```

These quantities describe the pairing structure rather than predictive
performance. They establish that the subsequent summaries compare matched tasks,
matched assessment individuals, and identical evaluation targets.

## 4. Evaluation metrics

The same metric definitions are applied separately to prediction of the true
simulated genetic value and to prediction of the observed phenotype.

### 4.1 Prediction metrics

- **Correlation** describes linear agreement between prediction and target.
- **RMSE** and **MAE** summarize prediction error on the original target scale.
- **R-squared** is calculated as \(1-SSE/SST\), using prediction errors relative
  to the target mean. It is not defined here as squared correlation and can be
  negative for genuinely held-out predictions.

### 4.2 Calibration metrics

- **Bias** is the mean prediction error,
  \(\operatorname{mean}(\text{prediction}-\text{target})\). Zero is the
  calibration target.
- **Slope** is the prediction coefficient in `lm(target ~ prediction)`. A value
  near one indicates appropriate proportional scaling, although the regression
  intercept also contributes to calibration.

Correlation, error metrics, and calibration diagnostics answer complementary
questions. No single metric is interpreted as establishing universal model
superiority.

### 4.3 Direction of improvement

For paired ANN-versus-GBLUP comparisons, each metric is assigned either a
direction of improvement or a calibration target.

```{r comparison-metric-rules}
higher_better <- c(
  "Genetic_Correlation",
  "Genetic_R2",
  "Phenotype_Correlation",
  "Phenotype_R2"
)

lower_better <- c(
  "Genetic_RMSE",
  "Genetic_MAE",
  "Phenotype_RMSE",
  "Phenotype_MAE"
)

zero_target <- c(
  "Genetic_Bias",
  "Phenotype_Bias"
)

one_target <- c(
  "Genetic_Slope",
  "Phenotype_Slope"
)

metric_rule <- c(
  setNames(rep("higher", length(higher_better)), higher_better),
  setNames(rep("lower", length(lower_better)), lower_better),
  setNames(rep("zero", length(zero_target)), zero_target),
  setNames(rep("one", length(one_target)), one_target)
)

metric_rule_table <- data.frame(
  Metric = names(metric_rule),
  Better_value = unname(metric_rule),
  stringsAsFactors = FALSE
)

format_table(
  metric_rule_table,
  caption =
    "Direction or calibration target used in paired model comparisons",
  row.names = FALSE
)
```

For `higher`, larger values are preferred; for `lower`, smaller values are
preferred. Bias is evaluated by distance from zero and slope by distance from
one.

## 5. Overall model performance

Overall performance is shown separately for the two evaluation targets so that
the primary simulated-data criterion is not visually mixed with the secondary
phenotype criterion. Both summaries average the same 150 matched outer-CV tasks
per model.

```{r overall-comparison-calculations}
analysis_cols <- c(
  "Scenario", "Repetition", "Fold", "SplitID", "TaskID", "Model", metrics
)

all_results <- rbind(gr[, analysis_cols], ar[, analysis_cols])

overall <- aggregate(
  all_results[, metrics],
  by = list(Model = all_results$Model),
  FUN = mean
)

overall_genetic_display <- overall[, c(
  "Model",
  "Genetic_Correlation",
  "Genetic_RMSE",
  "Genetic_MAE",
  "Genetic_R2"
)]
overall_genetic_display[-1] <- lapply(
  overall_genetic_display[-1],
  round,
  4
)

overall_phenotype_display <- overall[, c(
  "Model",
  "Phenotype_Correlation",
  "Phenotype_RMSE",
  "Phenotype_MAE",
  "Phenotype_R2"
)]
overall_phenotype_display[-1] <- lapply(
  overall_phenotype_display[-1],
  round,
  4
)
```

### 5.1 True genetic value

The primary summary compares each model with the known simulated genetic value.
Higher correlation and \(R^2\), together with lower RMSE and MAE, indicate better
held-out recovery of the genetic signal.

```{r overall-genetic-performance}
format_table(
  overall_genetic_display,
  caption =
    "Primary overall performance against true genetic value",
  row.names = FALSE
)
```

### 5.2 Phenotype

The secondary summary evaluates the same held-out predictions against the
observed phenotype, which also contains non-genetic variation.

```{r overall-phenotype-performance}
format_table(
  overall_phenotype_display,
  caption =
    "Secondary overall performance against observed phenotype",
  row.names = FALSE
)
```

```{r overall-interpretation}
best_corr_model <- overall$Model[
  which.max(overall$Genetic_Correlation)
]

best_rmse_model <- overall$Model[
  which.min(overall$Genetic_RMSE)
]

best_r2_model <- overall$Model[
  which.max(overall$Genetic_R2)
]

best_overall <- overall[
  overall$Model == best_corr_model,
  ,
  drop = FALSE
]

ann_overall_now <- overall[
  overall$Model != "GBLUP_ADE",
  ,
  drop = FALSE
]

best_ann_now <- ann_overall_now[
  which.max(
    ann_overall_now$Genetic_Correlation
  ),
  ,
  drop = FALSE
]
```

Across all 150 matched outer-CV tasks, **`r best_corr_model`** has the highest
mean genetic correlation
(`r sprintf("%.4f", best_overall$Genetic_Correlation)`), the lowest mean
genetic RMSE (`r sprintf("%.4f", best_overall$Genetic_RMSE)`), and the highest
mean genetic \(R^2\) (`r sprintf("%.4f", best_overall$Genetic_R2)`). Among the
ANN variants, **`r best_ann_now$Model`** has the highest mean genetic
correlation (`r sprintf("%.4f", best_ann_now$Genetic_Correlation)`). These
statements describe the evaluated model and search space; they do not establish
universal superiority.

## 6. Performance by scenario

The next table asks whether the overall ranking persists across the six
prespecified simulation scenarios. Only the primary genetic-value metrics are
shown here; the complete metric set remains available in the versioned
comparison outputs.

```{r scenario-comparison}
scenario_summary <- aggregate(
  all_results[, metrics],
  by = list(
    Scenario = all_results$Scenario,
    Model = all_results$Model
  ),
  FUN = mean
)

scenario_display <- scenario_summary[, c(
  "Scenario", "Model", "Genetic_Correlation", "Genetic_RMSE", "Genetic_R2"
)]
scenario_display[, -(1:2)] <- lapply(
  scenario_display[, -(1:2), drop = FALSE],
  round,
  4
)

format_table(
  scenario_display,
  caption =
    "Primary genetic-value prediction metrics by scenario",
  row.names = FALSE
)
```

```{r scenario-interpretation}
scenario_ids <- sort(
  unique(scenario_summary$Scenario)
)

scenario_winners <- do.call(
  rbind,
  lapply(
    scenario_ids,
    function(sc) {
      z <- scenario_summary[
        scenario_summary$Scenario == sc,
        ,
        drop = FALSE
      ]

      data.frame(
        Scenario = sc,
        Best_correlation =
          z$Model[which.max(z$Genetic_Correlation)],
        Best_RMSE =
          z$Model[which.min(z$Genetic_RMSE)],
        Best_R2 =
          z$Model[which.max(z$Genetic_R2)],
        stringsAsFactors = FALSE
      )
    }
  )
)

corr_wins <- sum(
  scenario_winners$Best_correlation == "GBLUP_ADE"
)

rmse_wins <- sum(
  scenario_winners$Best_RMSE == "GBLUP_ADE"
)

r2_wins <- sum(
  scenario_winners$Best_R2 == "GBLUP_ADE"
)
```

GBLUP-ADE has the best mean genetic correlation in
**`r corr_wins` of `r length(scenario_ids)` scenarios**, the lowest mean genetic
RMSE in **`r rmse_wins` of `r length(scenario_ids)`**, and the highest mean
genetic \(R^2\) in **`r r2_wins` of `r length(scenario_ids)`**. The C1-C6
sequence is interpreted as performance across the prespecified joint simulation
gradient, not as an isolated causal effect of QTL number or nominal
heritability.

## 7. Task-matched ANN vs GBLUP-ADE comparison

### 7.1 Overall task-matched contrasts

For the three primary genetic-value metrics, the paired advantage is defined so
that a positive value always favors the ANN variant:

- **Correlation advantage** = ANN correlation - GBLUP-ADE correlation.
- **RMSE advantage** = GBLUP-ADE RMSE - ANN RMSE.
- **\(R^2\) advantage** = ANN \(R^2\) - GBLUP-ADE \(R^2\).

The same convention is extended to MAE, bias, slope, and the secondary phenotype
metrics by comparing the appropriate direction or distance from the calibration
target.

```{r paired-comparison}
performance_advantage <- function(
  metric,
  ann_value,
  gblup_value
) {
  rule <- unname(
    metric_rule[metric]
  )

  candidate_advantages <- list(
    higher =
      ann_value - gblup_value,
    lower =
      gblup_value - ann_value,
    zero =
      abs(gblup_value) - abs(ann_value),
    one =
      abs(gblup_value - 1) -
      abs(ann_value - 1)
  )

  candidate_advantages[[rule]]
}

g <- gr[, c("TaskID", "SplitID", metrics)]
names(g)[-(1:2)] <- paste0(names(g)[-(1:2)], "_GBLUP")

paired_list <- list()

for (model in sort(unique(ar$Model))) {
  a <- ar[
    ar$Model == model,
    c("TaskID", "SplitID", "Scenario", metrics)
  ]

  z <- merge(a, g, by = c("TaskID", "SplitID"), all = FALSE)

  for (metric in metrics) {
    ann_value <- z[[metric]]
    gblup_value <- z[[paste0(metric, "_GBLUP")]]
    raw_difference <- ann_value - gblup_value
    advantage <- performance_advantage(metric, ann_value, gblup_value)

    paired_list[[length(paired_list) + 1L]] <- data.frame(
      Model = model,
      Metric = metric,
      N = length(raw_difference),
      Mean_ANN = mean(ann_value),
      Mean_GBLUP = mean(gblup_value),
      Mean_raw_difference = mean(raw_difference),
      Median_raw_difference = median(raw_difference),
      Mean_performance_advantage = mean(advantage),
      Median_performance_advantage = median(advantage),
      ANN_better = sum(advantage > 0),
      Equal = sum(advantage == 0),
      GBLUP_better = sum(advantage < 0),
      stringsAsFactors = FALSE
    )
  }
}

paired <- do.call(rbind, paired_list)
rownames(paired) <- NULL

paired_primary <- paired[
  paired$Metric %in% c("Genetic_Correlation", "Genetic_RMSE", "Genetic_R2"),
  c(
    "Model", "Metric", "Mean_ANN", "Mean_GBLUP",
    "Mean_performance_advantage", "ANN_better", "Equal", "GBLUP_better"
  )
]
paired_primary[, 3:5] <- lapply(
  paired_primary[, 3:5, drop = FALSE],
  round,
  4
)

format_table(
  paired_primary,
  caption = paste(
    "Paired ANN versus GBLUP-ADE comparison for the primary",
    "genetic-value metrics"
  ),
  row.names = FALSE
)
```

The `ANN_better` and `GBLUP_better` counts refer to the same frozen
outer-assessment tasks for both methods. They therefore describe within-task
predictive wins under directly matched test samples. These counts are
descriptive summaries of repeated cross-validation and are not treated as
independent replicates for a naive fold-level significance test.

### 7.2 By simulation scenario

To localize the paired pattern, genetic-correlation advantage is summarized
within each scenario over its 25 matched outer tasks. Positive values favor the
ANN variant and negative values favor GBLUP-ADE.

```{r paired-by-scenario}
paired_scenario_list <- list()

for (model in sort(unique(ar$Model))) {
  a <- ar[
    ar$Model == model,
    c("TaskID", "SplitID", "Scenario", metrics)
  ]

  z <- merge(a, g, by = c("TaskID", "SplitID"), all = FALSE)

  for (scenario in sort(unique(z$Scenario))) {
    zs <- z[z$Scenario == scenario, , drop = FALSE]

    for (metric in metrics) {
      ann_value <- zs[[metric]]
      gblup_value <- zs[[paste0(metric, "_GBLUP")]]
      advantage <- performance_advantage(metric, ann_value, gblup_value)

      paired_scenario_list[[length(paired_scenario_list) + 1L]] <- data.frame(
        Scenario = scenario,
        Model = model,
        Metric = metric,
        N = nrow(zs),
        Mean_advantage = mean(advantage),
        Median_advantage = median(advantage),
        ANN_better = sum(advantage > 0),
        Equal = sum(advantage == 0),
        GBLUP_better = sum(advantage < 0),
        stringsAsFactors = FALSE
      )
    }
  }
}

paired_scenario <- do.call(rbind, paired_scenario_list)
rownames(paired_scenario) <- NULL

paired_scenario_primary <- paired_scenario[
  paired_scenario$Metric == "Genetic_Correlation",
  c(
    "Scenario", "Model", "Mean_advantage",
    "ANN_better", "Equal", "GBLUP_better"
  )
]
paired_scenario_primary$Mean_advantage <- round(
  paired_scenario_primary$Mean_advantage,
  4
)

format_table(
  paired_scenario_primary,
  caption = paste(
    "Paired genetic-correlation advantage by scenario.",
    "Positive values favor ANN."
  ),
  row.names = FALSE
)
```

The scenario-level paired summaries show whether the overall pattern is
concentrated in a subset of simulation conditions or persists across the full
C1--C6 sequence. Positive mean advantage values favor the ANN variant; negative
values favor GBLUP-ADE. Together with the scenario means above, these paired
summaries provide a task-matched description of how consistently the methods
differ.

## 8. Regularization effects among ANN models

The overall comparison can also be read within the ANN branch alone. Restricting
the primary genetic-value summary to the four neural-network variants isolates
the effect associated with the prespecified L1 and L2 regularization choices,
without changing the architecture and learning-rate search contract established
in M04.

```{r ann-regularization-summary}
ann_regularization_display <- overall[
  match(expected_ann_models, overall$Model),
  c(
    "Model",
    "Genetic_Correlation",
    "Genetic_RMSE",
    "Genetic_MAE",
    "Genetic_R2"
  ),
  drop = FALSE
]

ann_regularization_display[-1] <- lapply(
  ann_regularization_display[-1],
  round,
  4
)

format_table(
  ann_regularization_display,
  caption = "Overall primary genetic-value performance among ANN variants",
  row.names = FALSE
)
```

The regularized ANN results show that the main improvement relative to `ANN_0`
is associated with the L1 component. `ANN_L1_L2` has the strongest overall ANN
averages by a very small margin over `ANN_L1`, whereas `ANN_L2` remains close to
the unregularized network. This is an interpretation of the evaluated bounded
search space, not a claim about all possible ANN regularization settings.

## 9. Versioned comparison outputs

The four canonical comparison tables are versioned under
`output/simulated/comparison/`. They are rewritten only when
`WRITE_COMPARISON_OUTPUTS=1`; the normal tutorial render leaves the stored
scientific outputs unchanged.

```{r write-comparison-outputs, eval=WRITE_COMPARISON_OUTPUTS}
dir.create(COMPARISON_DIR, recursive = TRUE, showWarnings = FALSE)
write.csv(overall, OVERALL_CSV, row.names = FALSE)
write.csv(scenario_summary, SCENARIO_CSV, row.names = FALSE)
write.csv(paired, PAIRED_CSV, row.names = FALSE)
write.csv(paired_scenario, PAIRED_SCENARIO_CSV, row.names = FALSE)
```

The following table reports the comparison files already available for reuse.

```{r verify-comparison-outputs}
canonical_files <- c(
  OVERALL_CSV, SCENARIO_CSV, PAIRED_CSV, PAIRED_SCENARIO_CSV
)
output_files <- data.frame(
  File = basename(canonical_files),
  Exists = file.exists(canonical_files),
  Size_KB = round(file.info(canonical_files)$size / 1024, 1)
)
format_table(
  output_files,
  caption = "Versioned comparison tables available for reuse",
  row.names = FALSE
)
```

## 10. Overall scientific conclusion

Within the prespecified canonical model and ANN architecture search space,
GBLUP-ADE provides the strongest overall predictive performance and leads the
primary genetic-value metrics across all six simulation scenarios. Among the
neural networks, ANN-L1+L2 has the best overall averages by a very small margin
over ANN-L1, while ANN-L2 remains close to the unregularized ANN.

These results support a specific conclusion for the frozen design evaluated
here; they do not establish a universal ranking between genomic mixed models
and all possible neural-network architectures.

## 11. Navigation

[GBLUP-ADE](simulated_data_gblup.html) | [Regularized ANN](simulated_data_ann.html) | [Home](index.html)
