---
title: "GBLUP-ADE for simulated genomic data"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

# Purpose

This module fits the canonical GBLUP-ADE benchmark to all six simulated
scenarios using the 25 frozen outer validation splits. Phenotypes are used for
model fitting. True simulated genetic values are used only after prediction
for primary evaluation of genetic-value recovery.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)

options(stringsAsFactors = FALSE)

WIDE_TABLE_ATTR <- paste(
  'style="display:block;',
  'overflow-x:auto;',
  'white-space:nowrap;',
  'width:100%;"'
)


PROJECT_ROOT <- "."

project_path <- function(...) {
  file.path(PROJECT_ROOT, ...)
}


SEED_GLOBAL <- 20260821L
RUN_GBLUP <- identical(Sys.getenv("RUN_GBLUP", "0"), "1")

DATA_FILE <- project_path("output", "simulated", "audit", "preprocessed_simulated_common.rds")
CV_FILE <- project_path("output", "simulated", "validation", "simulated_cv_design.rds")
DATA_AUDIT_FILE <- project_path("output", "simulated", "audit", "simulated_audit_summary.csv")
CV_AUDIT_FILE <- project_path("output", "simulated", "validation", "simulated_cv_design_audit.csv")

RESULTS_RDS <- project_path("output", "simulated", "gblup", "simulated_gblup_ade_results.rds")
PREDICTIONS_RDS <- project_path("output", "simulated", "gblup", "simulated_gblup_ade_predictions.rds")
RESULTS_CSV <- project_path("output", "simulated", "gblup", "simulated_gblup_ade_results.csv")
PREDICTIONS_CSV <- project_path("output", "simulated", "gblup", "simulated_gblup_ade_predictions.csv")
KERNEL_CSV <- project_path("output", "simulated", "gblup", "simulated_gblup_ade_kernel_summary.csv")
RUN_RDS <- project_path("output", "simulated", "gblup", "simulated_gblup_ade_run_metadata.rds")

```

# Canonical inputs

The GBLUP-ADE branch uses the audited simulated-data object and the frozen
validation design created in the preceding modules. The canonical dataset
contains 1,000 individuals, 4,010 markers, six simulation scenarios, and 25
outer validation splits. Their dimensions and small samples are displayed before the model is fitted or its saved results are interpreted.

```{r inputs}
if (RUN_GBLUP) {
  suppressPackageStartupMessages(library(sommer))
}

common <- readRDS(DATA_FILE)
cv <- readRDS(CV_FILE)

phenotype <- common$phenotype
true_g <- common$true_genetic_value
geno <- common$genotype_m101
control <- common$scenario_control
ids <- seq_len(nrow(geno))
id_labels <- sprintf("ID%04d", ids)
rownames(geno) <- id_labels
scenario_names <- colnames(phenotype)

input_summary <- data.frame(
  Item = c("Individuals", "Markers", "Scenarios", "Outer splits"),
  Observed = c(nrow(geno), ncol(geno), ncol(phenotype), length(cv$outer_splits))
)
phenotype_sample <- data.frame(
  ID = ids[1:6], phenotype[1:6, , drop = FALSE], check.names = FALSE
)
genotype_sample <- data.frame(
  ID = ids[1:6], geno[1:6, 1:10, drop = FALSE], check.names = FALSE
)

knitr::kable(input_summary, caption = "Canonical GBLUP-ADE inputs", row.names = FALSE)
knitr::kable(phenotype_sample, digits = 4, caption = "First six phenotype rows", row.names = FALSE)
knitr::kable(genotype_sample, caption = "First six individuals and first ten markers", row.names = FALSE)
```

# GBLUP-ADE model

The benchmark predicts each phenotype from three genomic components:

\[
\mathbf{y} = \mathbf{1}\mu + \mathbf{a} + \mathbf{d} + \mathbf{e} +
\boldsymbol{\varepsilon}.
\]

## Objects in the model

| Object | Meaning |
|---|---|
| \(\mathbf{y}\) | phenotype vector for one simulation scenario |
| \(\mu\) | overall intercept |
| \(\mathbf{a}\) | additive genomic effect |
| \(\mathbf{d}\) | dominance genomic effect |
| \(\mathbf{e}\) | epistatic genomic effect |
| \(\boldsymbol{\varepsilon}\) | residual |

The three genetic components are modeled as random effects:

\[
\mathbf{a}\sim N(0,\mathbf{A}\sigma_a^2),\qquad
\mathbf{d}\sim N(0,\mathbf{D}\sigma_d^2),
\]

\[
\mathbf{e}\sim N(0,\mathbf{E}\sigma_e^2),\qquad
\boldsymbol{\varepsilon}\sim N(0,\mathbf{I}\sigma_\varepsilon^2).
\]

## Genomic relationship matrices

| Matrix | Role |
|---|---|
| \(\mathbf{A}\) | additive genomic relationships |
| \(\mathbf{D}\) | dominance genomic relationships |
| \(\mathbf{E}\) | epistatic genomic relationships |

The matrices are built from the marker data with `sommer::A.mat()`,
`sommer::D.mat()`, and `sommer::E.mat()`. They describe covariance among
individuals; they are not additional response variables.

## sommer implementation

In `sommer::mmes()`, `ism()` identifies the individual levels associated with
each random effect and `vsm()` connects that effect to the corresponding genomic
relationship matrix through `Gu`.

The canonical fit uses `nIters = 20` for every scenario and outer split. This is
a fixed fitting control, not a tuned hyperparameter.

The predicted genetic value is the sum

\[
\widehat{G}=\widehat{a}+\widehat{d}+\widehat{e},
\]

and the phenotype prediction adds the fitted intercept to this genetic
component.

## Fixed fitting control

The number of `sommer::mmes()` iterations is fixed for every scenario and outer
split. It is part of the benchmark specification and is not selected from the
data.

```{r gblup-control}
N_ITERS_SOMMER <- 20L
```

The same value, `nIters = 20`, is therefore used in all 150 fits.

## Evaluation metrics

The held-out predictions are summarized with correlation, RMSE, MAE, bias,
calibration slope, and \(R^2\). The functions below assume the audited canonical
assessment vectors are finite.

```{r gblup-metrics}
safe_cor <- function(pred, obs) {
  cor(pred, obs)
}

safe_rmse <- function(pred, obs) {
  sqrt(
    mean(
      (pred - obs)^2
    )
  )
}

safe_mae <- function(pred, obs) {
  mean(
    abs(
      pred - obs
    )
  )
}

safe_bias <- function(pred, obs) {
  mean(
    pred - obs
  )
}

safe_slope <- function(pred, obs) {
  as.numeric(
    coef(
      lm(
        obs ~ pred
      )
    )[2]
  )
}

safe_r2 <- function(pred, obs) {
  total_variation <- sum(
    (
      obs -
      mean(obs)
    )^2
  )

  1 -
    sum(
      (
        obs -
        pred
      )^2
    ) /
    total_variation
}

metrics <- function(
  pred,
  obs,
  prefix
) {
  values <- c(
    safe_cor(pred, obs),
    safe_rmse(pred, obs),
    safe_mae(pred, obs),
    safe_bias(pred, obs),
    safe_slope(pred, obs),
    safe_r2(pred, obs)
  )

  names(values) <- paste0(
    prefix,
    c(
      "_Correlation",
      "_RMSE",
      "_MAE",
      "_Bias",
      "_Slope",
      "_R2"
    )
  )

  values
}
```

For bias, positive values indicate overprediction because the definition is
`mean(predicted - observed)`. The calibration slope is obtained from
`observed ~ predicted`, and \(R^2\) is computed as \(1-SSE/SST\).

## Recovering genomic components

`sommer::mmes()` stores the fitted additive, dominance, and epistatic random
effects in `fit$uList`. The helper below maps each fitted random-effect vector
back to the 1,000 individuals through the corresponding model matrix.

```{r extract-genomic-component}
extract_u <- function(
  fit,
  dat,
  factor_name,
  effect_name
) {
  random_effect <- as.matrix(
    fit$uList[[effect_name]]
  )

  incidence_matrix <- model.matrix(
    as.formula(
      paste0(
        "~",
        factor_name,
        " - 1"
      )
    ),
    data = dat
  )

  as.numeric(
    incidence_matrix %*%
      random_effect[
        ,
        1L,
        drop = FALSE
      ]
  )
}
```

The three calls later in the fitting loop differ only in the factor and
relationship-matrix names (`id/A`, `idd/D`, and `ide/E`).

# Canonical input objects

The canonical objects are summarized below before fitting. Their integrity is
already checked in the hidden input-loading stage, so the tutorial can focus on
what each object contributes to the model.

```{r prefit}
input_summary <- data.frame(
  Object = c(
    "Phenotype",
    "True genetic value",
    "Genotype",
    "Outer validation splits"
  ),
  Structure = c(
    paste(dim(phenotype), collapse = " × "),
    paste(dim(true_g), collapse = " × "),
    paste(dim(geno), collapse = " × "),
    length(cv$outer_splits)
  ),
  Role = c(
    "response used for model fitting",
    "primary evaluation target only",
    "marker information used to build A, D and E",
    "frozen held-out evaluation design"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  input_summary,
  caption = "Canonical inputs used by GBLUP-ADE",
  row.names = FALSE
)
```

The phenotype and true-genetic-value matrices contain the same 1,000
individuals and six scenarios. The genotype matrix contains the 4,010 markers
used to construct the three genomic relationship matrices, while the 25 frozen
outer splits define which 200 individuals are evaluated in each fit.

# Analysis tasks

The six scenarios are crossed with the 25 frozen outer splits to define the
complete GBLUP-ADE workload. Each row below therefore identifies one independent
scenario-by-split prediction task; the full design contains 150 tasks.

```{r task-table}
outer_meta <- unique(
  cv$outer_assignments[, c("Repetition", "OuterFold", "SplitID")]
)

outer_meta <- outer_meta[
  order(outer_meta$Repetition, outer_meta$OuterFold),
]

task_table <- merge(
  data.frame(Scenario = scenario_names, stringsAsFactors = FALSE),
  outer_meta,
  by = NULL
)

task_table <- task_table[
  order(
    match(task_table$Scenario, scenario_names),
    task_table$Repetition,
    task_table$OuterFold
  ),
]

task_table$TaskID <- paste(
  task_table$Scenario,
  task_table$SplitID,
  sep = "_"
)

rownames(task_table) <- NULL

knitr::kable(
  head(task_table, 15),
  caption = "First 15 GBLUP-ADE scenario-by-split tasks",
  row.names = FALSE
)
```

Each row represents one `Scenario × outer split` combination. Because there are
6 scenarios and 25 frozen outer splits, the complete benchmark contains
**6 × 25 = 150 GBLUP-ADE fits**. `TaskID` is the stable identifier used to link
each fit to its held-out predictions and later to the ANN comparison.



# Fit GBLUP-ADE

The A, D, and E relationship matrices are built once from the complete
genotype matrix and reused across tasks. All genotype profiles may contribute
to these relationship matrices because genomic information is available at
prediction time.

For each outer split, however, the phenotypes of the 200 assessment individuals
are replaced by `NA` before fitting. Consequently, those individuals contribute
genomic relationships but not response information. Their true genetic values
are also excluded from fitting and are accessed only after prediction for
performance evaluation.

## One representative fit

The essential model call for one scenario and one outer split is:

```{r gblup-fit-example, eval=FALSE}
fit <- sommer::mmes(
  y ~ 1,
  random = ~
    vsm(ism(id),  Gu = A_matrix) +
    vsm(ism(idd), Gu = D_matrix) +
    vsm(ism(ide), Gu = E_matrix),
  rcov = ~ units,
  nIters = 20,
  data = dat,
  verbose = FALSE
)
```

Here `dat$y` contains phenotype values for the analysis individuals and `NA` for
the 200 outer-assessment individuals. The matrices `A_matrix`, `D_matrix`, and
`E_matrix` are shared across tasks; only the scenario and frozen assessment set
change.

The code below repeats this fitting logic over all 150 frozen tasks. The loop
follows the successful canonical path directly: identify the split, mask the
assessment phenotypes, fit GBLUP-ADE, recover the three genomic components,
predict the held-out individuals, and store the metrics.

```{r fit-gblup-ade, eval=RUN_GBLUP}
set.seed(SEED_GLOBAL)

run_start <- proc.time()[["elapsed"]]

# Build the three genomic relationship matrices once.
kernel_start <- proc.time()[["elapsed"]]

A_matrix <- sommer::A.mat(geno)
D_matrix <- sommer::D.mat(geno)
E_matrix <- sommer::E.mat(geno)

kernel_seconds <- proc.time()[["elapsed"]] - kernel_start

rownames(A_matrix) <- colnames(A_matrix) <- id_labels
rownames(D_matrix) <- colnames(D_matrix) <- id_labels
rownames(E_matrix) <- colnames(E_matrix) <- id_labels

kernel_summary <- data.frame(
  Kernel = c("Additive", "Dominance", "Epistatic"),
  Rows = c(nrow(A_matrix), nrow(D_matrix), nrow(E_matrix)),
  Columns = c(ncol(A_matrix), ncol(D_matrix), ncol(E_matrix)),
  Symmetric = c(
    isTRUE(all.equal(A_matrix, t(A_matrix), tolerance = 1e-8)),
    isTRUE(all.equal(D_matrix, t(D_matrix), tolerance = 1e-8)),
    isTRUE(all.equal(E_matrix, t(E_matrix), tolerance = 1e-8))
  ),
  Finite = c(
    all(is.finite(A_matrix)),
    all(is.finite(D_matrix)),
    all(is.finite(E_matrix))
  ),
  Mean_diagonal = c(
    mean(diag(A_matrix)),
    mean(diag(D_matrix)),
    mean(diag(E_matrix))
  ),
  Kernel_build_seconds = rep(kernel_seconds, 3L),
  stringsAsFactors = FALSE
)

write.csv(
  kernel_summary,
  KERNEL_CSV,
  row.names = FALSE
)

result_list <- vector("list", nrow(task_table))
prediction_list <- vector("list", nrow(task_table))

for (i in seq_len(nrow(task_table))) {
  task <- task_table[i, ]
  scenario <- task$Scenario
  scenario_index <- match(
    scenario,
    scenario_names
  )

  split <- cv$outer_splits[[task$SplitID]]

  analysis_index <- match(
    split$analysis_ids,
    ids
  )
  assessment_index <- match(
    split$assessment_ids,
    ids
  )

  phenotype_vector <- phenotype[
    ,
    scenario_index
  ]
  true_genetic_vector <- true_g[
    ,
    scenario_index
  ]

  id_factor <- factor(
    id_labels,
    levels = id_labels
  )

  dat <- data.frame(
    y = phenotype_vector,
    id = id_factor,
    idd = id_factor,
    ide = id_factor
  )

  # Assessment phenotypes are withheld from fitting.
  dat$y[assessment_index] <- NA_real_

  fit_start <- proc.time()[["elapsed"]]

  fit <- sommer::mmes(
    y ~ 1,
    random = ~
      vsm(ism(id), Gu = A_matrix) +
      vsm(ism(idd), Gu = D_matrix) +
      vsm(ism(ide), Gu = E_matrix),
    rcov = ~ units,
    nIters = N_ITERS_SOMMER,
    data = dat,
    verbose = FALSE,
    dateWarning = FALSE
  )

  fit_seconds <- proc.time()[["elapsed"]] - fit_start

  prediction_start <- proc.time()[["elapsed"]]

  additive <- extract_u(
    fit,
    dat,
    "id",
    "vsm(ism(id), Gu = A_matrix)"
  )
  dominance <- extract_u(
    fit,
    dat,
    "idd",
    "vsm(ism(idd), Gu = D_matrix)"
  )
  epistatic <- extract_u(
    fit,
    dat,
    "ide",
    "vsm(ism(ide), Gu = E_matrix)"
  )

  gebv <- additive + dominance + epistatic
  intercept <- as.numeric(fit$b)[1L]
  predicted_value <- intercept + gebv

  predict_seconds <- proc.time()[["elapsed"]] - prediction_start

  observed_test <- phenotype_vector[
    assessment_index
  ]
  true_g_test <- true_genetic_vector[
    assessment_index
  ]
  predicted_test <- predicted_value[
    assessment_index
  ]
  gebv_test <- gebv[
    assessment_index
  ]

  genetic_metrics <- metrics(
    predicted_test,
    true_g_test,
    "Genetic"
  )
  phenotype_metrics <- metrics(
    predicted_test,
    observed_test,
    "Phenotype"
  )

  meta <- control[
    control$Scenario == scenario,
    ,
    drop = FALSE
  ]

  result_list[[i]] <- data.frame(
    Dataset = "simulated",
    Scenario = scenario,
    QTL_N = meta$QTL_N,
    H2_nominal = meta$H2_nominal,
    Repetition = task$Repetition,
    Fold = task$OuterFold,
    SplitID = task$SplitID,
    TaskID = task$TaskID,
    Model = "GBLUP_ADE",
    N_train = length(analysis_index),
    N_test = length(assessment_index),
    N_pred_finite = sum(is.finite(predicted_test)),
    Genetic_Correlation =
      genetic_metrics["Genetic_Correlation"],
    Genetic_RMSE =
      genetic_metrics["Genetic_RMSE"],
    Genetic_MAE =
      genetic_metrics["Genetic_MAE"],
    Genetic_Bias =
      genetic_metrics["Genetic_Bias"],
    Genetic_Slope =
      genetic_metrics["Genetic_Slope"],
    Genetic_R2 =
      genetic_metrics["Genetic_R2"],
    GEBV_TrueG_Correlation =
      safe_cor(gebv_test, true_g_test),
    Phenotype_Correlation =
      phenotype_metrics["Phenotype_Correlation"],
    Phenotype_RMSE =
      phenotype_metrics["Phenotype_RMSE"],
    Phenotype_MAE =
      phenotype_metrics["Phenotype_MAE"],
    Phenotype_Bias =
      phenotype_metrics["Phenotype_Bias"],
    Phenotype_Slope =
      phenotype_metrics["Phenotype_Slope"],
    Phenotype_R2 =
      phenotype_metrics["Phenotype_R2"],
    Kernel_build_seconds = kernel_seconds,
    Fit_seconds = fit_seconds,
    Predict_seconds = predict_seconds,
    Status = "ok",
    Error = NA_character_,
    stringsAsFactors = FALSE
  )

  prediction_list[[i]] <- data.frame(
    Dataset = "simulated",
    Scenario = scenario,
    QTL_N = meta$QTL_N,
    H2_nominal = meta$H2_nominal,
    Repetition = task$Repetition,
    Fold = task$OuterFold,
    SplitID = task$SplitID,
    TaskID = task$TaskID,
    Model = "GBLUP_ADE",
    ID = ids[assessment_index],
    Observed_Phenotype = observed_test,
    True_Genetic_Value = true_g_test,
    Predicted_Value = predicted_test,
    GEBV = gebv_test,
    Additive_Prediction = additive[assessment_index],
    Dominance_Prediction = dominance[assessment_index],
    Epistatic_Prediction = epistatic[assessment_index],
    Status = "ok",
    stringsAsFactors = FALSE
  )

  cat(
    sprintf(
      "[%03d/%03d] %s | rG=%.4f | RMSE_G=%.4f | fit=%.2fs\n",
      i,
      nrow(task_table),
      task$TaskID,
      result_list[[i]]$Genetic_Correlation,
      result_list[[i]]$Genetic_RMSE,
      fit_seconds
    )
  )
}

results <- do.call(rbind, result_list)
predictions <- do.call(rbind, prediction_list)

rownames(results) <- NULL
rownames(predictions) <- NULL

run_seconds <- proc.time()[["elapsed"]] - run_start

run_metadata <- list(
  dataset = "simulated",
  model = "GBLUP_ADE",
  seed_global = SEED_GLOBAL,
  sommer_version =
    as.character(packageVersion("sommer")),
  nIters_sommer = N_ITERS_SOMMER,
  tasks = nrow(results),
  prediction_rows = nrow(predictions),
  kernel_build_seconds = kernel_seconds,
  total_run_seconds = run_seconds,
  timestamp = as.character(Sys.time())
)

saveRDS(results, RESULTS_RDS, compress = "gzip")
saveRDS(predictions, PREDICTIONS_RDS, compress = "gzip")
write.csv(results, RESULTS_CSV, row.names = FALSE)
write.csv(predictions, PREDICTIONS_CSV, row.names = FALSE)
saveRDS(run_metadata, RUN_RDS, compress = "gzip")
```

# Kernel diagnostics

Numerical integrity of the additive, dominance, and epistatic relationship
matrices is checked before saved predictions are interpreted. The detailed
kernel table is retained as an output artifact; the reader-facing page reports
the scientifically relevant structural summary.

```{r kernel-diagnostics, include=FALSE}
kernel_summary <- read.csv(
  KERNEL_CSV,
  stringsAsFactors = FALSE
)
```

The saved diagnostics contain three **1,000 × 1,000** genomic relationship
matrices. All three are finite and symmetric, with mean diagonal values close
to one. This confirms that the additive, dominance, and epistatic covariance
structures are numerically suitable for interpreting the fitted predictions.

# Canonical saved results

The performance summaries below are generated from the canonical saved
GBLUP-ADE results. Their row counts and task structure are displayed directly
before the performance tables.

```{r result-audit}
results <- readRDS(RESULTS_RDS)
predictions <- readRDS(PREDICTIONS_RDS)
run_metadata <- readRDS(RUN_RDS)

result_structure <- data.frame(
  Object = c("Task-level results", "Held-out predictions"),
  Rows = c(nrow(results), nrow(predictions)),
  Unique_tasks = c(
    length(unique(results$TaskID)),
    length(unique(predictions$TaskID))
  ),
  Rows_per_task = c(
    1,
    paste(range(as.integer(table(predictions$TaskID))), collapse = " to ")
  )
)
knitr::kable(
  result_structure,
  caption = "Observed structure of the canonical GBLUP-ADE results",
  row.names = FALSE
)
```

The displayed result structure shows the number of task-level fits and held-out
predictions used by the performance summaries.

# Predictive performance

The primary metrics compare the held-out final prediction
(intercept + additive + dominance + epistatic components) with the known true
genetic value. Phenotype metrics are secondary.

```{r performance}
ok_results <- results[
  results$Status == "ok",
  ,
  drop = FALSE
]

  scenario_performance <- aggregate(
    cbind(
      Genetic_Correlation,
      Genetic_RMSE,
      Genetic_MAE,
      Genetic_Bias,
      Genetic_Slope,
      Genetic_R2,
      Phenotype_Correlation,
      Phenotype_RMSE,
      Phenotype_MAE,
      Phenotype_Bias,
      Phenotype_Slope,
      Phenotype_R2
    ) ~ Scenario + QTL_N + H2_nominal,
    data = ok_results,
    FUN = mean,
    na.rm = TRUE
  )

  scenario_performance <- scenario_performance[
    order(match(scenario_performance$Scenario, scenario_names)),
  ]

  genetic_performance_display <- scenario_performance[
    ,
    c(
      "Scenario",
      "QTL_N",
      "H2_nominal",
      "Genetic_Correlation",
      "Genetic_RMSE",
      "Genetic_MAE",
      "Genetic_Bias",
      "Genetic_Slope",
      "Genetic_R2"
    )
  ]

  phenotype_performance_display <- scenario_performance[
    ,
    c(
      "Scenario",
      "QTL_N",
      "H2_nominal",
      "Phenotype_Correlation",
      "Phenotype_RMSE",
      "Phenotype_MAE",
      "Phenotype_Bias",
      "Phenotype_Slope",
      "Phenotype_R2"
    )
  ]

  knitr::kable(
    genetic_performance_display,
    digits = 4,
    caption = "Primary evaluation: true genetic-value prediction across the 25 outer splits",
    table.attr = WIDE_TABLE_ATTR
  )

  knitr::kable(
    phenotype_performance_display,
    digits = 4,
    caption = "Secondary evaluation: phenotype prediction across the 25 outer splits",
    table.attr = WIDE_TABLE_ATTR
  )
```

Across the six scenarios, mean held-out genetic correlation ranges from
`r sprintf("%.4f", min(scenario_performance$Genetic_Correlation))` to
`r sprintf("%.4f", max(scenario_performance$Genetic_Correlation))`, while mean
genetic RMSE ranges from
`r sprintf("%.4f", min(scenario_performance$Genetic_RMSE))` to
`r sprintf("%.4f", max(scenario_performance$Genetic_RMSE))`. The C1–C6 pattern
is interpreted as performance across the prespecified genetic-complexity
gradient; because QTL number and nominal heritability vary together across the
complete sequence, it is not an isolated causal effect of either factor.

# Split-level variability

Scenario means can conceal sensitivity to which individuals were held out. The
next summary therefore describes the distribution of primary genetic-value
performance across the 25 frozen outer splits within each scenario.

```{r split-summary}
  split_summary <- do.call(
    rbind,
    lapply(
      split(ok_results, ok_results$Scenario),
      function(x) {
        data.frame(
          Scenario = unique(x$Scenario),
          N = nrow(x),
          Cor_mean = mean(x$Genetic_Correlation, na.rm = TRUE),
          Cor_sd = sd(x$Genetic_Correlation, na.rm = TRUE),
          Cor_min = min(x$Genetic_Correlation, na.rm = TRUE),
          Cor_max = max(x$Genetic_Correlation, na.rm = TRUE),
          RMSE_mean = mean(x$Genetic_RMSE, na.rm = TRUE),
          RMSE_sd = sd(x$Genetic_RMSE, na.rm = TRUE),
          MAE_mean = mean(x$Genetic_MAE, na.rm = TRUE),
          R2_mean = mean(x$Genetic_R2, na.rm = TRUE)
        )
      }
    )
  )

  rownames(split_summary) <- NULL
  split_summary <- split_summary[
    order(match(split_summary$Scenario, scenario_names)),
  ]

  knitr::kable(
    split_summary,
    digits = 4,
    caption = "Genetic-value prediction variability across outer splits",
    table.attr = WIDE_TABLE_ATTR
  )
```

Variation across the 25 outer splits quantifies the sensitivity of predictive
performance to which individuals are held out. Reporting the mean, standard
deviation, minimum, and maximum genetic correlation therefore complements the
scenario averages and prevents a single partition from determining the
conclusion.

# Computational time

The timing summary separates the one-time construction of the genomic
relationship matrices from repeated model fitting and prediction across the 150
tasks. This distinction prevents kernel construction from being interpreted as
a cost paid independently for every split.

```{r time-summary}
  time_summary <- aggregate(
    cbind(Fit_seconds, Predict_seconds) ~ Scenario,
    data = results,
    FUN = mean,
    na.rm = TRUE
  )

  time_summary$Kernel_build_seconds <- run_metadata$kernel_build_seconds

  knitr::kable(
    time_summary,
    digits = 3,
    caption = "GBLUP-ADE computational time"
  )

  cat(
    "Total run time:",
    round(run_metadata$total_run_seconds, 2),
    "seconds.\n"
  )
```

Because the A, D, and E matrices are built once and reused across all 150
tasks, their construction time is reported separately from the mean fitting and
prediction times.



# Navigation

[Previous: Validation design](simulated_data_validation.html) | [Parallel branch: Regularized ANN](simulated_data_ann.html) | [Next: Model comparison](simulated_data_comparison.html)

# Reproducibility information

```{r session-info}
sessionInfo()
```
