---
title: "Simulated genomic data: structure and preparation"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

## 1. Purpose

This module introduces the simulated genomic population used throughout the
prediction workflow. The data were generated with **GENES**, a software package
for experimental statistics, quantitative genetics, and genetic simulation
(Cruz, 2013, 2016).

The module has four reader-facing goals:

- **Identify the input data:** distinguish phenotype, true genetic value,
  genotype matrix, genetic map, parental genotypes, and simulation settings.
- **Make the data structure explicit:** show how individuals, scenarios,
  markers, linkage groups, and QTL are organized before model fitting.
- **Describe the simulated population:** summarize genetic signal, segregation,
  marker properties, and linkage disequilibrium (LD).
- **Prepare one shared object:** preserve the same individuals, marker coding,
  map, scenarios, and QTL for the validation, GBLUP-ADE, ANN, and comparison
  modules.

No genomic prediction model is fitted here. The purpose is to establish what the
data contain, how the main quantities are represented, and which objects are
passed unchanged to the next stages of the workflow.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)

options(stringsAsFactors = FALSE)

dir.create(
  "output/simulated/m01/figures",
  showWarnings = FALSE,
  recursive = TRUE
)
```

## 2. Packages and reproducibility

### 2.1 Packages

Three packages support different parts of the reader-facing analysis:

- **`knitr`** converts R objects into formatted HTML tables, allowing dimensions,
  simulation settings, genotype summaries, and LD values to be inspected
  directly in the page.
- **`ggplot2`** constructs every statistical and genomic figure, including the
  phenotype distributions, chromosome-like linkage-group map, QTL distribution,
  F2 segregation, and LD curve.
- **`ggthemes`** provides `theme_gdocs()` and the matching Google Docs color
  scales so all figures share the same visual language.

The comments in the code provide a short reminder of these roles when the reader
executes the chunk independently.

```{r packages, warning=FALSE, message=FALSE}
library(knitr)     # Format data summaries as readable HTML tables.
library(ggplot2)   # Construct the statistical and genomic figures.
library(ggthemes)  # Apply the Google Docs theme and matching color scales.
```

For figure output, two rules are applied consistently:

- figures shown in the HTML use `theme_gdocs()` and Google Docs color/fill
  scales whenever categories are distinguished by color;
- publication-quality TIFF copies are written to
  `output/simulated/m01/figures/`, while the `ggsave()` commands remain in
  `include=FALSE` chunks so file-export instructions do not interrupt the
  analytical code shown to the reader.

### 2.2 Random seed

Most quantities in this module are fixed because they are read directly from
the GENES simulation files. The random seed is needed for only one operation:

- **Fixed inputs:** phenotype, true genetic values, genotypes, genetic map,
  parental genotypes, and scenario definitions do not change with the seed.
- **Random operation:** 5,000 marker pairs from different linkage groups are
  sampled in Section 10 to obtain a reproducible background LD reference.

Setting the seed here therefore reproduces the LD reference without modifying
any of the simulated data.

```{r random-seed}
SEED_GLOBAL <- 20260821L
set.seed(SEED_GLOBAL)
```

## 3. Phenotypic and true genetic values

The first two GENES files describe complementary quantities for the same
1,000 individuals and six scenarios:

- **Phenotype:** the observed response that may be used for model fitting when an
  individual belongs to an analysis set.
- **True genetic value:** the known simulated genetic component, reserved for
  evaluating predictions rather than fitting or partition construction.
- **Alignment:** row `i` in both matrices refers to the same simulated
  individual, so row order must remain unchanged throughout the workflow.

### 3.1 Phenotypic values

The phenotypic file is read before any transformation so the reader can see
how the response is organized. The chunk performs three steps:

- read the matrix with individuals in rows and scenarios in columns;
- label the six scenario columns as C1-C6;
- reshape only a plotting copy and use boxplots to compare center, dispersion,
  and extreme values across scenarios.

The original phenotype matrix remains unchanged for all later calculations.

```{r phenotype, fig.cap="Distribution of phenotypic values across the six simulation scenarios."}
# Read phenotypes with individuals in rows and scenarios in columns.
phenotype <- as.matrix(
  read.table(
    "data/dados_simulados/DG#gen_F2_Vfen.txt",
    header = FALSE
  )
)

colnames(phenotype) <- paste0("C", 1:6)

# Reshape only a plotting copy; the original matrix remains unchanged.
phenotype_long <- stack(as.data.frame(phenotype))
names(phenotype_long) <- c("Phenotype", "Scenario")

p_phenotype <- ggplot(
  phenotype_long,
  aes(
    x = Scenario,
    y = Phenotype,
    fill = Scenario
  )
) +
  geom_boxplot(
    width = 0.65,
    show.legend = FALSE
  ) +
  scale_fill_gdocs() +
  labs(
    x = "Simulation scenario",
    y = "Phenotypic value"
  ) +
  theme_gdocs()

print(p_phenotype)
```

```{r save-phenotype-tiff, include=FALSE}
ggsave(
  "output/simulated/m01/figures/phenotype_distribution.tiff",
  plot = p_phenotype,
  device = "tiff",
  width = 8,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The phenotype object now establishes two facts used later:

- **Dimensions:** `r nrow(phenotype)` individuals are represented in each of
  `r ncol(phenotype)` scenarios.
- **Use in prediction:** phenotype is the response available to a model only for
  individuals belonging to the corresponding analysis set defined by the
  validation design.

### 3.2 True genetic values

Because the population is simulated, the genetic value underlying each
individual is known. Its role differs from the phenotype:

- it is **not** used to create validation partitions;
- it is **not** used to tune or fit the prediction models;
- it is retained as the external target used to evaluate predictions for
  held-out individuals.

The chunk reads the matrix with the same C1-C6 structure and creates a plotting
copy. Comparing its distributions with the phenotype figure separates variation
in the known genetic component from variation in the observed response.

```{r true-genetic-values, fig.cap="Distribution of true genetic values across the six simulation scenarios."}
# Read the known genetic values generated by GENES.
true_genetic_value <- as.matrix(
  read.table(
    "data/dados_simulados/DG#gen_F2_Vgen.dat",
    header = FALSE
  )
)

colnames(true_genetic_value) <- paste0("C", 1:6)

# Reshape a copy only for the reader-facing figure.
true_g_long <- stack(as.data.frame(true_genetic_value))
names(true_g_long) <- c("True_genetic_value", "Scenario")

p_true_g <- ggplot(
  true_g_long,
  aes(
    x = Scenario,
    y = True_genetic_value,
    fill = Scenario
  )
) +
  geom_boxplot(
    width = 0.65,
    show.legend = FALSE
  ) +
  scale_fill_gdocs() +
  labs(
    x = "Simulation scenario",
    y = "True genetic value"
  ) +
  theme_gdocs()

print(p_true_g)
```

```{r save-true-genetic-value-tiff, include=FALSE}
ggsave(
  "output/simulated/m01/figures/true_genetic_value_distribution.tiff",
  plot = p_true_g,
  device = "tiff",
  width = 8,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

This second response-related object has three properties to retain:

- **Same dimensions:** `r nrow(true_genetic_value)` individuals x six scenarios.
- **Same row order:** every row remains paired with the corresponding phenotype
  row.
- **Different analytical role:** the values are reserved for held-out evaluation
  rather than model fitting.

## 4. Genotypic data

### 4.1 Genotype matrix

The GENES genotype file must first be oriented in the same direction as the
response matrices. The reader should note four points:

- **Source orientation:** markers are stored in rows and individuals in columns.
- **Analysis orientation:** the matrix is transposed once, producing individuals
  in rows and markers in columns.
- **Observed coding:** the original genotype states are `0`, `1`, and `2`.
- **Displayed sample:** ten individuals and twelve markers spread across all
  4,010 marker columns are printed so the discrete matrix structure can be seen
  directly; a boxplot would not convey that structure.

The complete matrix, not the displayed sample, is retained for every subsequent
calculation.

```{r genotype-sample}
# Read the GENES marker file and orient it as individuals x markers.
geno_012 <- t(
  as.matrix(
    read.table(
      "data/dados_simulados/genoma_mapa.txt",
      header = FALSE
    )
  )
)

# Select markers distributed from the beginning to the end of the matrix.
marker_sample <- unique(
  round(
    seq(
      1,
      ncol(geno_012),
      length.out = 12
    )
  )
)

# Show a small real sample while retaining the complete matrix for analysis.
genotype_sample <- data.frame(
  Individual = 1:10,
  geno_012[
    1:10,
    marker_sample,
    drop = FALSE
  ],
  check.names = FALSE
)

names(genotype_sample)[-1] <- paste0("M", marker_sample)

genotype_sample |>
  knitr::kable(
    format = "html",
    caption = "Sample of 10 individuals and 12 markers distributed across the genotype matrix",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The displayed sample makes the structure of the complete genotype object
explicit:

- **Dimensions:** `r nrow(geno_012)` individuals x `r ncol(geno_012)` markers.
- **Observed states:** `r paste(sort(unique(as.vector(geno_012))), collapse = ", ")`.
- **Interpretation:** these three discrete states are the genotype classes
  generated for the simulated F2 population.

### 4.2 Genotype coding used in the prediction models

The prediction models use the same genotype calls centered on zero. The
recoding is deliberately simple:

- `0 -> -1`;
- `1 -> 0`;
- `2 -> 1`.

Only the numerical scale changes. No individual or marker is removed, and the
underlying genotype class is preserved. The table below makes the transformation
explicit before the recoded matrix is saved for downstream analyses.

```{r genotype-coding}
geno_m101 <- geno_012 - 1L

data.frame(
  Original = c(0, 1, 2),
  Analysis_coding = c(-1, 0, 1)
) |>
  knitr::kable(
    format = "html",
    caption = "Genotype coding used in the prediction analyses",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

After recoding, the prediction matrix preserves the data structure:

- the same 1,000 individuals remain in the same row order;
- the same 4,010 markers remain in the same column order;
- only the numerical coding changes from `0/1/2` to `-1/0/1`.

## 5. Genetic map

The genetic map supplies the genomic organization needed to interpret marker
positions, QTL locations, and LD. The GENES text file is read in two conceptual
parts:

- the first 12 lines describe the map globally;
- marker-level records start on line 13 and provide linkage group, marker order,
  marker name, and position in centimorgans (cM).

For later matching, a global marker index from 1 to 4,010 is added. This index is
the bridge between the QTL numbers stored in the scenario file and their
positions on the genetic map.

```{r genetic-map-data}
# Read the marker-level portion of the GENES genetic-map file.
genetic_map <- read.table(
  "data/dados_simulados/genoma.txt",
  header = FALSE,
  sep = ",",
  skip = 12,
  fill = TRUE,
  strip.white = TRUE,
  fileEncoding = "latin1",
  stringsAsFactors = FALSE
)

genetic_map <- genetic_map[, 1:5]

names(genetic_map) <- c(
  "LG",
  "Marker_within_LG",
  "Type",
  "Marker",
  "Position_cM"
)

# Convert the map fields used in numerical calculations.
genetic_map$LG <- as.integer(genetic_map$LG)
genetic_map$Marker_within_LG <- as.integer(genetic_map$Marker_within_LG)
genetic_map$Position_cM <- as.numeric(genetic_map$Position_cM)
genetic_map$Global_marker <- seq_len(nrow(genetic_map))

genetic_map$LG_label <- factor(
  paste0("LG", genetic_map$LG),
  levels = paste0("LG", 1:10)
)

map_summary <- data.frame(
  Linkage_group = paste0("LG", 1:10),
  Markers = as.integer(table(genetic_map$LG)),
  Start_cM = as.numeric(
    tapply(genetic_map$Position_cM, genetic_map$LG, min)
  ),
  End_cM = as.numeric(
    tapply(genetic_map$Position_cM, genetic_map$LG, max)
  )
)

map_summary |>
  knitr::kable(
    format = "html",
    caption = "Structure of the simulated genetic map",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The numerical map can be summarized before QTL are introduced:

- **10 linkage groups (LG1-LG10);**
- **401 markers per linkage group;**
- **0-200 cM** covered by each group;
- **0.5 cM** between consecutive markers;
- **4,010 markers** in the complete map.

A chromosome-like representation is used below to make this organization easier
to read. Each colored body represents one linkage group and the short white
ticks show every tenth marker. Displaying a regular subset of marker ticks keeps
the 0.5-cM structure visible without drawing 401 overlapping marks on each
linkage group.

```{r genetic-map-figure, fig.cap="Chromosome-like representation of the ten linkage groups in the simulated genetic map. Colors identify linkage groups and white ticks show every tenth marker position."}
# Create one chromosome-like body for each linkage group.
chromosome_map <- data.frame(
  LG = 1:10,
  LG_label = factor(
    paste0("LG", 1:10),
    levels = paste0("LG", 1:10)
  ),
  Start_cM = 0,
  End_cM = 200
)

# Show a regular subset of marker positions to keep the map legible.
marker_ticks <- genetic_map[
  genetic_map$Marker_within_LG %% 10 == 1,
  c("LG", "LG_label", "Position_cM")
]

p_genetic_map <- ggplot() +
  geom_segment(
    data = chromosome_map,
    aes(
      x = LG_label,
      xend = LG_label,
      y = Start_cM,
      yend = End_cM,
      color = LG_label
    ),
    linewidth = 12,
    lineend = "round",
    show.legend = FALSE
  ) +
  geom_point(
    data = marker_ticks,
    aes(
      x = LG_label,
      y = Position_cM
    ),
    shape = 95,
    size = 4.2,
    color = "white",
    alpha = 0.85
  ) +
  scale_color_gdocs() +
  scale_y_reverse(
    breaks = seq(0, 200, 25),
    expand = expansion(mult = c(0.035, 0.035))
  ) +
  labs(
    x = "Linkage group",
    y = "Position (cM)"
  ) +
  theme_gdocs()

print(p_genetic_map)
```

```{r save-genetic-map-tiff, include=FALSE}
ggsave(
  "output/simulated/m01/figures/genetic_map.tiff",
  plot = p_genetic_map,
  device = "tiff",
  width = 9,
  height = 6,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

With the marker map defined, the simulated genetic architecture can be added
without mixing the two levels of information:

- **Marker framework:** Section 5 establishes LG1-LG10, marker order, and map
  positions independently of the simulated QTL.
- **QTL architecture:** Section 6 places the QTL from each scenario on that fixed
  framework and compares their distribution across C1-C6.

Keeping these layers separate lets the reader first understand where markers are
located and then examine which of those positions were selected as QTL in each
scenario.

## 6. Simulation settings and QTL architecture

### 6.1 GENES software and genetic model

The six scenarios were generated with **GENES** (Cruz, 2013, 2016). The control
file specifies `Model = 2`, represented by

\[
Y_i =
\mu +
\sum_j a_j +
\sum_j \sum_{j'} \alpha_j\alpha_{j'} +
\varepsilon_i .
\]

The terms have distinct roles in the simulation:

- \(Y_i\): phenotype generated for individual \(i\);
- \(\mu\): overall mean;
- \(\sum_j a_j\): contributions associated with loci;
- \(\sum_j\sum_{j'}\alpha_j\alpha_{j'}\): pairwise interaction contributions
  between loci;
- \(\varepsilon_i\): residual component.

This equation describes **data generation**, not the GBLUP-ADE or ANN prediction
models fitted later.

### 6.2 Scenario settings and QTL distribution

The control file contains six repeated blocks, one for each scenario C1-C6.
The code reads the blocks directly because the fields occur in a fixed order.
The reader can follow the preparation in four steps:

- **Read the scenario settings:** number of QTL, specified heritability, genetic
  model, constant, scale, and CV setting.
- **Read the QTL indices:** each selected QTL is stored as an integer marker
  index in the control file.
- **Match QTL to the genetic map:** each index is converted to its linkage group
  and position in cM using the global marker index from Section 5.
- **Reuse one QTL table:** the matched positions are used for the figure, the
  count table, and the object passed to later modules.

```{r simulation-scenarios}
# Read the six scenario blocks written by GENES.
control_raw <- readLines(
  "data/dados_simulados/controlegenetico#1.dat",
  warn = FALSE,
  encoding = "latin1"
)

# Transliterate accented text so the fixed block positions are read consistently.
control_ascii <- iconv(
  control_raw,
  from = "latin1",
  to = "ASCII//TRANSLIT",
  sub = ""
)

scenario_start <- grep(
  "^Variavel",
  control_ascii
)

control_table <- data.frame(
  Scenario = paste0("C", 1:6),
  Variable = NA_real_,
  Total_loci = NA_real_,
  QTL_N = NA_real_,
  gmd = NA_real_,
  Model = NA_real_,
  H2_nominal = NA_real_,
  Constant = NA_real_,
  Scale = NA_real_,
  CV_setting = NA_real_
)

qtl_lists <- vector(
  "list",
  length(scenario_start)
)

# Recover the settings and QTL indices scenario by scenario.
for (i in seq_along(scenario_start)) {

  s <- scenario_start[i]

  control_table$Variable[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s]))
  )
  control_table$Total_loci[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s + 1]))
  )
  control_table$QTL_N[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s + 2]))
  )

  # QTL values are marker indices and are therefore kept as integers.
  qtl_lists[[i]] <- as.integer(
    scan(
      text = control_ascii[s + 3],
      what = numeric(),
      quiet = TRUE
    )
  )

  control_table$gmd[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s + 5]))
  )
  control_table$Model[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s + 6]))
  )
  control_table$H2_nominal[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s + 7]))
  )
  control_table$Constant[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s + 8]))
  )
  control_table$Scale[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s + 9]))
  )
  control_table$CV_setting[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s + 10]))
  )
}

# Match each QTL index to its linkage group and map position.
qtl_rows <- data.frame(
  Scenario = character(),
  QTL_index = integer(),
  LG = integer(),
  Position_cM = numeric()
)

for (i in seq_along(qtl_lists)) {

  qtl <- qtl_lists[[i]]

  qtl_rows <- rbind(
    qtl_rows,
    data.frame(
      Scenario = paste0("C", i),
      QTL_index = qtl,
      LG = genetic_map$LG[qtl],
      Position_cM = genetic_map$Position_cM[qtl]
    )
  )
}

qtl_rows$LG_label <- factor(
  paste0("LG", qtl_rows$LG),
  levels = paste0("LG", 1:10)
)

control_table |>
  knitr::kable(
    format = "html",
    digits = 3,
    caption = "Settings used to generate the six simulation scenarios in GENES",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The six scenarios differ in two main characteristics that should be interpreted
together:

- **QTL number increases:** 8, 40, 80, 120, 240, and 480 from C1 to C6;
- **specified heritability decreases:** 0.7 in C1-C2, 0.5 in C3-C4, and 0.3 in
  C5-C6;
- **other settings remain fixed:** `gmd = 0.5`, `Model = 2`, `Constant = 100`,
  `Scale = 1`, and `CV = 12`;
- **LG9 and LG10 contain no simulated QTL:** they retain markers but provide
  reference genomic regions without a true QTL signal.

The single QTL figure below is therefore organized by **scenario**. Each panel
uses the chromosome-like linkage-group representation from Section 5 and places
all QTL for that scenario on their mapped positions. This avoids repeating a
second combined QTL map while still showing both genomic location and the
increase in QTL density across C1-C6.

```{r qtl-map, fig.cap="Positions of simulated QTL in each scenario. The linkage-group color scheme is constant across panels; LG9 and LG10 remain without QTL in all six scenarios."}
# Draw a chromosome-like body for every linkage group in every panel.
qtl_reference <- expand.grid(
  Scenario = paste0("C", 1:6),
  LG_label = paste0("LG", 1:10),
  stringsAsFactors = FALSE
)

qtl_reference$LG_label <- factor(
  qtl_reference$LG_label,
  levels = paste0("LG", 1:10)
)

qtl_reference$Start_cM <- 0
qtl_reference$End_cM <- 200

p_qtl <- ggplot() +
  geom_segment(
    data = qtl_reference,
    aes(
      x = LG_label,
      xend = LG_label,
      y = Start_cM,
      yend = End_cM,
      color = LG_label
    ),
    linewidth = 3.6,
    lineend = "round",
    show.legend = FALSE
  ) +
  geom_point(
    data = qtl_rows,
    aes(
      x = LG_label,
      y = Position_cM
    ),
    shape = 21,
    size = 1.6,
    fill = "white",
    color = "black",
    stroke = 0.25
  ) +
  facet_wrap(
    ~Scenario,
    ncol = 2
  ) +
  scale_color_gdocs() +
  scale_y_reverse(
    breaks = seq(0, 200, 50)
  ) +
  labs(
    x = "Linkage group",
    y = "Position (cM)"
  ) +
  theme_gdocs() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

print(p_qtl)
```

```{r save-qtl-map-tiff, include=FALSE}
ggsave(
  "output/simulated/m01/figures/qtl_positions_by_scenario.tiff",
  plot = p_qtl,
  device = "tiff",
  width = 10,
  height = 8,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The figure gives the spatial pattern, while the table below provides the exact
counts needed to verify it:

- rows correspond to scenarios C1-C6;
- columns correspond to LG1-LG10;
- each cell is the number of QTL placed in that linkage group;
- the zero counts in LG9 and LG10 make their reference role explicit.

```{r qtl-counts}
qtl_table <- table(
  factor(qtl_rows$Scenario, levels = paste0("C", 1:6)),
  factor(qtl_rows$LG, levels = 1:10)
)

qtl_by_lg <- data.frame(
  Scenario = rownames(qtl_table),
  unclass(qtl_table),
  check.names = FALSE,
  row.names = NULL
)

names(qtl_by_lg)[-1] <- paste0("LG", 1:10)

qtl_by_lg |>
  knitr::kable(
    format = "html",
    caption = "Number of simulated QTL per linkage group",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

## 7. Phenotypic and genetic signal

Because both phenotype \(Y\) and true genetic value \(G\) are known in the
simulation, the remaining component can be calculated directly as

\[
E = Y - G.
\]

The section then describes the strength of the simulated genetic signal using
four complementary quantities:

- **Phenotype-genetic correlation:** how closely the observed phenotype follows
  the known genetic value.
- **Genetic-residual correlation:** whether the simulated genetic and remaining
  components show residual association.
- **Realized heritability:** the proportion obtained from the genetic and
  residual variances actually present in the generated population.
- **Location and dispersion summaries:** means and standard deviations used to
  describe the scale of each scenario.

Together these summaries provide context for why prediction difficulty differs
among scenarios before any model is fitted.

```{r phenotypic-genetic-signal}
scenario_names <- paste0("C", 1:6)
residual_value <- phenotype - true_genetic_value

cor_p_g <- numeric(6)
cor_g_e <- numeric(6)
h2_realized <- numeric(6)
var_g_over_var_p <- numeric(6)

# Calculate the signal summaries separately for each simulation scenario.
for (j in 1:6) {

  cor_p_g[j] <- cor(
    phenotype[, j],
    true_genetic_value[, j]
  )

  cor_g_e[j] <- cor(
    true_genetic_value[, j],
    residual_value[, j]
  )

  vg <- var(true_genetic_value[, j])
  ve <- var(residual_value[, j])

  h2_realized[j] <- vg / (vg + ve)
  var_g_over_var_p[j] <- vg / var(phenotype[, j])
}

# Gather the quantities required for interpretation and later reuse.
scenario_summary <- data.frame(
  Scenario = scenario_names,
  QTL_N = control_table$QTL_N,
  H2_nominal = control_table$H2_nominal,
  Phenotype_mean = colMeans(phenotype),
  Phenotype_sd = apply(phenotype, 2, sd),
  Genetic_mean = colMeans(true_genetic_value),
  Genetic_sd = apply(true_genetic_value, 2, sd),
  Residual_mean = colMeans(residual_value),
  Residual_sd = apply(residual_value, 2, sd),
  Cor_P_G = cor_p_g,
  Cor_G_E = cor_g_e,
  H2_realized = h2_realized,
  VarG_over_VarP = var_g_over_var_p,
  CV_empirical_pct = 100 * apply(phenotype, 2, sd) / colMeans(phenotype),
  CV_setting = control_table$CV_setting
)

scenario_summary[
  ,
  c(
    "Scenario",
    "QTL_N",
    "H2_nominal",
    "H2_realized",
    "Cor_P_G",
    "Phenotype_mean",
    "Genetic_mean",
    "Residual_sd"
  )
] |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Phenotypic and genetic signal by simulation scenario",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

Two points from this table are especially important for later prediction:

- **Genetic signal differs among scenarios:** phenotype-true-genetic-value
  correlation ranges from
  `r sprintf("%.3f", min(scenario_summary$Cor_P_G))` to
  `r sprintf("%.3f", max(scenario_summary$Cor_P_G))`.
- **Comparison context:** model performance should therefore be interpreted
  alongside scenario difficulty rather than assuming that C1-C6 contain the
  same amount of recoverable genetic signal.

### 7.1 Specified and realized heritability

Two heritability quantities are intentionally kept separate:

- **Specified heritability:** the value entered in GENES when a scenario was
  generated.
- **Realized heritability:** the value calculated afterward from the genetic and
  residual variances observed in this finite simulated population.

The comparison shows how closely each generated sample reflects its specified
setting without redefining the scenario itself.

```{r heritability-comparison, fig.cap="Heritability specified in GENES and realized heritability in the generated population."}
heritability_plot <- rbind(
  data.frame(
    Scenario = scenario_summary$Scenario,
    Heritability = scenario_summary$H2_nominal,
    Type = "GENES setting"
  ),
  data.frame(
    Scenario = scenario_summary$Scenario,
    Heritability = scenario_summary$H2_realized,
    Type = "Realized"
  )
)

h2_difference <- scenario_summary$H2_realized -
  scenario_summary$H2_nominal

p_heritability <- ggplot(
  heritability_plot,
  aes(
    x = Scenario,
    y = Heritability,
    group = Type,
    color = Type,
    shape = Type
  )
) +
  geom_line(
    linewidth = 0.8
  ) +
  geom_point(
    size = 2.5
  ) +
  scale_color_gdocs() +
  labs(
    x = "Simulation scenario",
    y = expression(H^2),
    color = NULL,
    shape = NULL
  ) +
  theme_gdocs()

print(p_heritability)
```

```{r save-heritability-tiff, include=FALSE}
ggsave(
  "output/simulated/m01/figures/heritability_specified_vs_realized.tiff",
  plot = p_heritability,
  device = "tiff",
  width = 8,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The comparison can be read directly from two quantities:

- **Absolute differences for C1-C6:**
  `r paste(sprintf("%.4f", abs(h2_difference)), collapse = ", ")`.
- **Largest difference:**
  `r scenario_summary$Scenario[which.max(abs(h2_difference))]`, with
  `r sprintf("%.4f", max(abs(h2_difference)))`.

These are realized properties of the generated population and do not change the
heritability values specified when the scenarios were simulated.

### 7.2 GENES CV setting and empirical phenotype CV

The value `CV = 12` stored by GENES and the empirical phenotype CV answer
different questions:

- **GENES CV setting:** an input recorded in the simulation control file;
- **empirical phenotype CV:** `100 x SD / mean`, calculated from the phenotype
  values actually generated in each scenario.

Displaying both prevents a simulation input from being mistaken for a
descriptive statistic of this particular finite population.

```{r cv-comparison, fig.cap="CV specified in GENES and empirical coefficient of variation calculated from the simulated phenotypes."}
cv_plot <- rbind(
  data.frame(
    Scenario = scenario_summary$Scenario,
    CV = scenario_summary$CV_setting,
    Type = "GENES setting"
  ),
  data.frame(
    Scenario = scenario_summary$Scenario,
    CV = scenario_summary$CV_empirical_pct,
    Type = "Empirical phenotype"
  )
)

cv_difference <- scenario_summary$CV_empirical_pct -
  scenario_summary$CV_setting

p_cv <- ggplot(
  cv_plot,
  aes(
    x = Scenario,
    y = CV,
    group = Type,
    color = Type,
    shape = Type
  )
) +
  geom_line(
    linewidth = 0.8
  ) +
  geom_point(
    size = 2.5
  ) +
  scale_color_gdocs() +
  labs(
    x = "Simulation scenario",
    y = "Coefficient of variation (%)",
    color = NULL,
    shape = NULL
  ) +
  theme_gdocs()

print(p_cv)
```

```{r save-cv-tiff, include=FALSE}
ggsave(
  "output/simulated/m01/figures/cv_genes_vs_empirical.tiff",
  plot = p_cv,
  device = "tiff",
  width = 8,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The distinction is visible numerically:

- **Empirical phenotype CV for C1-C6:**
  `r paste(sprintf("%.2f", scenario_summary$CV_empirical_pct), collapse = ", ")`%.
- **Difference from the GENES setting:**
  `r paste(sprintf("%.2f", cv_difference), collapse = ", ")` percentage points.

The GENES setting and the empirical phenotype CV are therefore retained as
separate quantities with separate interpretations.

## 8. Genotype properties

### 8.1 F2 segregation

The expected segregation follows directly from the simulated cross:

- **Parental contrast:** the two parents are homozygous for opposite marker
  states.
- **F2 expectation:** the three genotype classes should occur approximately in
  the ratio `1:2:1`, corresponding to proportions 0.25, 0.50, and 0.25.
- **Observed comparison:** all genotype calls in the 1,000 x 4,010 matrix are
  used to calculate the realized class proportions.

The figure therefore compares the complete simulated marker matrix with the
theoretical segregation expected for an F2 population.

```{r genotype-segregation, fig.cap="Observed genotype proportions compared with the expected 1:2:1 segregation of an F2 population."}
# Calculate the observed frequency of each genotype state.
geno_frequency <- prop.table(
  table(
    factor(
      as.vector(geno_012),
      levels = 0:2
    )
  )
)

segregation_plot <- rbind(
  data.frame(
    Genotype = factor(0:2, levels = 0:2),
    Proportion = as.numeric(geno_frequency),
    Type = "Observed"
  ),
  data.frame(
    Genotype = factor(0:2, levels = 0:2),
    Proportion = c(0.25, 0.50, 0.25),
    Type = "F2 expectation"
  )
)

p_segregation <- ggplot(
  segregation_plot,
  aes(
    x = Genotype,
    y = Proportion,
    fill = Type
  )
) +
  geom_col(
    position = "dodge"
  ) +
  scale_fill_gdocs() +
  labs(
    x = "Genotype coding",
    y = "Proportion",
    fill = NULL
  ) +
  theme_gdocs()

print(p_segregation)
```

```{r save-genotype-segregation-tiff, include=FALSE}
ggsave(
  "output/simulated/m01/figures/f2_genotype_proportions.tiff",
  plot = p_segregation,
  device = "tiff",
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The segregation figure is used for two related checks:

- **Observed pattern:** genotype classes 0, 1, and 2 occur close to the expected
  F2 proportions 0.25, 0.50, and 0.25.
- **Purpose:** the comparison confirms that the complete marker matrix reflects
  the segregation pattern expected from the simulated cross before it enters
  the prediction models.

### 8.2 Marker and individual properties

The genotype matrix is also described from several complementary angles:

- **Missing genotype cells:** absent marker calls.
- **Duplicate individuals:** rows with identical complete marker profiles.
- **Duplicate markers:** columns with identical profiles across all individuals.
- **Monomorphic markers:** markers with no genotype variation.
- **Minor allele frequency (MAF):** balance between the two alleles at each
  marker.
- **Heterozygosity:** proportion of individuals in the heterozygous class at
  each marker.

These quantities describe the simulated data. They are not used here as an
automatic filtering rule.

```{r genotype-checks}
# Count missing data and repeated complete genotype profiles.
missing_genotypes <- sum(is.na(geno_012))
duplicate_individuals <- sum(duplicated(geno_012, MARGIN = 1))
duplicate_markers <- sum(duplicated(geno_012, MARGIN = 2))

# Summarize marker-level allele frequency, MAF, heterozygosity, and variance.
allele_frequency <- colMeans(geno_012) / 2
maf <- pmin(allele_frequency, 1 - allele_frequency)
heterozygosity <- colMeans(geno_012 == 1)
marker_variance <- apply(geno_012, 2, var)
monomorphic_markers <- sum(marker_variance == 0)

marker_summary <- data.frame(
  Statistic = c(
    "Missing genotype cells",
    "Duplicate individuals",
    "Duplicate markers",
    "Monomorphic markers",
    "Minimum MAF",
    "Median MAF",
    "Maximum MAF",
    "Minimum heterozygosity",
    "Median heterozygosity",
    "Maximum heterozygosity"
  ),
  Value = c(
    missing_genotypes,
    duplicate_individuals,
    duplicate_markers,
    monomorphic_markers,
    min(maf),
    median(maf),
    max(maf),
    min(heterozygosity),
    median(heterozygosity),
    max(heterozygosity)
  )
)

marker_summary |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Summary of marker and individual properties",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

These summaries are descriptive rather than filtering rules. Consequently:

- no marker is removed because of these summaries in this module;
- the complete genotype matrix is passed downstream with the same individuals
  and markers described above.

## 9. Parental genotypes

The parental file links the F2 population to the two homozygous parents used
by GENES. The chunk reads four pieces of information:

- parent identifier;
- linkage-group identifier;
- reported number of markers;
- the 401 marker states within each linkage group.

The resulting summary checks that both parents cover all ten linkage groups and
shows the contrasting fixed states from which the three F2 genotype classes were
generated.

```{r parental-genotypes}
# Read the parental records across the ten linkage groups.
parents <- as.matrix(
  read.table(
    "data/dados_simulados/genoma_pais.txt",
    header = FALSE
  )
)

parent_id <- parents[, 1]
parent_lg <- parents[, 2]
parent_nmarkers <- parents[, 3]
parent_genotypes <- parents[, 5:405, drop = FALSE]

# Summarize the linkage-group coverage and fixed marker state of each parent.
parent_summary <- data.frame(
  Parent = c(1, 2),
  Linkage_groups = c(
    length(unique(parent_lg[parent_id == 1])),
    length(unique(parent_lg[parent_id == 2]))
  ),
  Marker_state = c(
    unique(
      as.vector(
        parent_genotypes[
          parent_id == 1,
          ,
          drop = FALSE
        ]
      )
    )[1],
    unique(
      as.vector(
        parent_genotypes[
          parent_id == 2,
          ,
          drop = FALSE
        ]
      )
    )[1]
  )
)

parent_summary |>
  knitr::kable(
    format = "html",
    caption = "Parental genotypes used to generate the simulated F2 population",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The parental summary establishes the simulated cross directly:

- **Coverage:** both parents are represented in all 10 linkage groups.
- **Parent 1:** fixed for marker state 1.
- **Parent 2:** fixed for marker state 0.
- **Implication:** these contrasting homozygous parental states generate the
  three genotype classes observed in the F2 population.

## 10. Linkage disequilibrium

Linkage disequilibrium (LD) measures non-random association between marker
genotypes. Here it is summarized by \(r^2\), the squared correlation between two
marker columns.

The analysis distinguishes two kinds of marker pairs:

- **Within linkage group:** both markers belong to the same LG. Because they
  share the same genetic map, their separation can be measured in cM and LD can
  be examined as a function of map distance.
- **Between linkage groups:** the two markers belong to different LGs. A
  within-group map distance is not defined for such pairs, so their mean
  \(r^2\) is used as a background reference for association between different
  linkage groups.

The LD calculation follows a direct sequence:

1. standardize the 4,010 marker columns across the 1,000 individuals;
2. exploit the regular 0.5-cM marker spacing to convert each requested distance
   into a marker shift (for example, 10 cM = 20 intervals);
3. calculate \(r^2\) for all valid within-LG pairs at 0.5, 1, 2.5, 5, 10, 25,
   50, 100, and 200 cM;
4. independently sample 5,000 between-LG pairs to obtain the background
   reference;
5. compare the distance-dependent within-LG curve with the between-LG reference.

The x-axis displays the nine evaluated distances as ordered categories with
equal visual spacing because a continuous 0.5-200 cM axis compresses the short
distances. This changes only the figure layout: every calculation still uses the
true numeric distance in cM.

```{r linkage-disequilibrium, fig.cap="LD across the evaluated within-linkage-group distances. The dashed line is the mean r-squared from 5,000 marker pairs drawn from different linkage groups. Distances are equally spaced on the displayed x-axis for readability, but calculations use the true distances in cM."}
# Standardize markers once so correlations can be calculated efficiently.
geno_scaled <- scale(geno_012)

ld_distances <- c(
  0.5,
  1,
  2.5,
  5,
  10,
  25,
  50,
  100,
  200
)

ld_summary <- data.frame(
  Distance_cM = ld_distances,
  Pairs = NA_integer_,
  Mean_r2 = NA_real_,
  Median_r2 = NA_real_
)

# Calculate LD for marker pairs separated by each distance within each LG.
for (d in seq_along(ld_distances)) {

  marker_shift <- as.integer(ld_distances[d] / 0.5)
  r2_all <- numeric()

  for (lg in 1:10) {

    markers <- which(genetic_map$LG == lg)

    marker_1 <- markers[
      1:(length(markers) - marker_shift)
    ]

    marker_2 <- markers[
      (1 + marker_shift):length(markers)
    ]

    r <- colSums(
      geno_scaled[, marker_1, drop = FALSE] *
        geno_scaled[, marker_2, drop = FALSE]
    ) / (nrow(geno_scaled) - 1)

    r2_all <- c(r2_all, r^2)
  }

  ld_summary$Pairs[d] <- length(r2_all)
  ld_summary$Mean_r2[d] <- mean(r2_all)
  ld_summary$Median_r2[d] <- median(r2_all)
}

# Draw marker pairs from different LGs for the background LD reference.
set.seed(SEED_GLOBAL + 101L)

n_inter_pairs <- 5000L

inter_i <- sample(
  seq_len(ncol(geno_012)),
  n_inter_pairs,
  replace = TRUE
)

inter_j <- sample(
  seq_len(ncol(geno_012)),
  n_inter_pairs,
  replace = TRUE
)

same_lg <- genetic_map$LG[inter_i] == genetic_map$LG[inter_j]

# Resample only second markers that accidentally fall in the same LG.
while (any(same_lg)) {

  inter_j[same_lg] <- sample(
    seq_len(ncol(geno_012)),
    sum(same_lg),
    replace = TRUE
  )

  same_lg <- genetic_map$LG[inter_i] == genetic_map$LG[inter_j]
}

inter_r <- colSums(
  geno_scaled[, inter_i, drop = FALSE] *
    geno_scaled[, inter_j, drop = FALSE]
) / (nrow(geno_scaled) - 1)

inter_r2 <- inter_r^2

interchromosomal_ld <- data.frame(
  Distance_cM = NA_real_,
  Pairs = length(inter_r2),
  Mean_r2 = mean(inter_r2),
  Median_r2 = median(inter_r2)
)

# Convert distances to ordered labels only for the figure.
ld_plot_data <- ld_summary
ld_plot_data$Distance_label <- factor(
  as.character(ld_plot_data$Distance_cM),
  levels = as.character(ld_distances)
)
ld_plot_data$Type <- "Within linkage group"

ld_reference_plot <- data.frame(
  Mean_r2 = interchromosomal_ld$Mean_r2,
  Type = "Between linkage groups"
)

p_ld <- ggplot(
  ld_plot_data,
  aes(
    x = Distance_label,
    y = Mean_r2,
    group = 1,
    color = Type
  )
) +
  geom_line(
    linewidth = 0.9
  ) +
  geom_point(
    size = 2.8
  ) +
  geom_hline(
    data = ld_reference_plot,
    aes(
      yintercept = Mean_r2,
      color = Type
    ),
    linetype = "dashed",
    linewidth = 0.9
  ) +
  scale_color_gdocs() +
  labs(
    x = "Map distance within linkage group (cM)",
    y = expression(mean(r^2)),
    color = NULL
  ) +
  theme_gdocs()

print(p_ld)
```

```{r save-ld-tiff, include=FALSE}
ggsave(
  "output/simulated/m01/figures/ld_decay.tiff",
  plot = p_ld,
  device = "tiff",
  width = 9,
  height = 5.5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The figure summarizes the pattern visually; the tables below retain the exact
numbers needed to interpret it:

- the first table reports within-LG mean and median \(r^2\) at each true map
  distance;
- the second reports the between-LG background reference from 5,000 sampled
  pairs.

```{r linkage-disequilibrium-table}
ld_summary |>
  knitr::kable(
    format = "html",
    digits = 5,
    caption = "Within-linkage-group LD by map distance",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )

interchromosomal_ld |>
  knitr::kable(
    format = "html",
    digits = 5,
    caption = "LD between markers from different linkage groups",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The observed pattern shows the expected strong decline over short and
intermediate distances. Mean \(r^2\) decreases from
`r sprintf("%.4f", ld_summary$Mean_r2[ld_summary$Distance_cM == 0.5])` at
0.5 cM to
`r sprintf("%.4f", ld_summary$Mean_r2[ld_summary$Distance_cM == 50])` at
50 cM. At 100 and 200 cM the means are
`r sprintf("%.4f", ld_summary$Mean_r2[ld_summary$Distance_cM == 100])` and
`r sprintf("%.4f", ld_summary$Mean_r2[ld_summary$Distance_cM == 200])`,
respectively, whereas the mean for markers from different linkage groups is
`r sprintf("%.5f", interchromosomal_ld$Mean_r2)`.

The fact that long-distance pairs from the same linkage group retain greater
association than the between-group reference is a property of this simulated
recombination structure. It is documented for interpretation and is not used to
remove markers.

## 11. Summary of the simulated data

Before moving to the validation design, the final table collects the main
structural facts established in the module:

- population size and number of simulation scenarios;
- total markers, linkage groups, and markers per linkage group;
- missing, duplicate, and monomorphic genotype patterns;
- minimum MAF;
- background LD between linkage groups.

The table is a compact description of the prepared data, while the biological
and statistical interpretation remains in the sections where each quantity is
introduced.

```{r data-summary}
data_overview <- data.frame(
  Quantity = c(
    "Individuals",
    "Markers",
    "Simulation scenarios",
    "Linkage groups",
    "Markers per linkage group",
    "Missing genotype cells",
    "Duplicate individuals",
    "Duplicate markers",
    "Monomorphic markers",
    "Minimum MAF",
    "Mean r2 between linkage groups"
  ),
  Observed = c(
    nrow(geno_012),
    ncol(geno_012),
    ncol(phenotype),
    length(unique(genetic_map$LG)),
    paste(
      unique(as.integer(table(genetic_map$LG))),
      collapse = ", "
    ),
    missing_genotypes,
    duplicate_individuals,
    duplicate_markers,
    monomorphic_markers,
    round(min(maf), 4),
    round(interchromosomal_ld$Mean_r2, 5)
  ),
  stringsAsFactors = FALSE
)

data_overview |>
  knitr::kable(
    format = "html",
    caption = "Summary of the simulated genomic data",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The summary reports what is present in the simulated population; the detailed
interpretation of QTL architecture, heritability, CV, and LD remains in the
sections where each quantity is introduced.

```{r data-check-output, include=FALSE}
dim_ok <-
  identical(dim(phenotype), c(1000L, 6L)) &&
  identical(dim(true_genetic_value), c(1000L, 6L)) &&
  identical(dim(geno_012), c(1000L, 4010L))

finite_ok <-
  all(is.finite(phenotype)) &&
  all(is.finite(true_genetic_value))

genotype_codes_ok <- identical(
  as.numeric(sort(unique(as.vector(geno_012)))),
  c(0, 1, 2)
)

analysis_codes_ok <- identical(
  as.numeric(sort(unique(as.vector(geno_m101)))),
  c(-1, 0, 1)
)

map_ok <-
  nrow(genetic_map) == 4010L &&
  length(unique(genetic_map$LG)) == 10L &&
  all(table(genetic_map$LG) == 401L) &&
  all(tapply(genetic_map$Position_cM, genetic_map$LG, min) == 0) &&
  all(tapply(genetic_map$Position_cM, genetic_map$LG, max) == 200)

scenario_ok <-
  identical(
    as.numeric(control_table$QTL_N),
    c(8, 40, 80, 120, 240, 480)
  ) &&
  isTRUE(
    all.equal(
      as.numeric(control_table$H2_nominal),
      c(0.7, 0.7, 0.5, 0.5, 0.3, 0.3),
      tolerance = 1e-12
    )
  ) &&
  all(control_table$Model == 2)

qtl_indices_ok <- all(
  qtl_rows$QTL_index >= 1 &
    qtl_rows$QTL_index <= nrow(genetic_map)
)

reference_lg_ok <- all(
  qtl_by_lg[, c("LG9", "LG10")] == 0
)

row_alignment_ok <-
  nrow(phenotype) == nrow(true_genetic_value) &&
  nrow(phenotype) == nrow(geno_012)

parent_structure_ok <-
  nrow(parents) == 20L &&
  ncol(parents) == 405L &&
  all(parent_genotypes[parent_id == 1, , drop = FALSE] == 1) &&
  all(parent_genotypes[parent_id == 2, , drop = FALSE] == 0) &&
  all(table(parent_lg) == 2L) &&
  all(parent_nmarkers == 401)

f2_frequency_ok <- max(
  abs(
    as.numeric(geno_frequency) -
      c(0.25, 0.50, 0.25)
  )
) < 0.02

maf_ok <- min(maf) > 0.40

short_ld <- ld_summary$Mean_r2[
  ld_summary$Distance_cM == 0.5
]

medium_ld <- ld_summary$Mean_r2[
  ld_summary$Distance_cM == 10
]

ld_50 <- ld_summary$Mean_r2[
  ld_summary$Distance_cM == 50
]

ld_decay_ok <-
  short_ld > medium_ld &&
  medium_ld > ld_50 &&
  ld_50 > interchromosomal_ld$Mean_r2

data_check_log <- data.frame(
  Check = c(
    "Primary matrix dimensions",
    "Finite phenotype and true genetic values",
    "Genotype coding 0/1/2",
    "Analysis genotype coding -1/0/1",
    "Genetic map structure",
    "Simulation scenario parameters",
    "QTL indices within genetic map",
    "LG9 and LG10 without simulated QTL",
    "Row alignment",
    "Missing genotypes",
    "Duplicate individuals",
    "Duplicate markers",
    "Monomorphic markers",
    "F2 genotype proportions",
    "Minor allele frequency",
    "Parental structure",
    "Short-to-medium distance LD decay"
  ),
  Status = c(
    ifelse(dim_ok, "PASS", "BLOCKED"),
    ifelse(finite_ok, "PASS", "BLOCKED"),
    ifelse(genotype_codes_ok, "PASS", "BLOCKED"),
    ifelse(analysis_codes_ok, "PASS", "BLOCKED"),
    ifelse(map_ok, "PASS", "BLOCKED"),
    ifelse(scenario_ok, "PASS", "BLOCKED"),
    ifelse(qtl_indices_ok, "PASS", "BLOCKED"),
    ifelse(reference_lg_ok, "PASS", "BLOCKED"),
    ifelse(row_alignment_ok, "PASS", "BLOCKED"),
    ifelse(missing_genotypes == 0, "PASS", "BLOCKED"),
    ifelse(duplicate_individuals == 0, "PASS", "BLOCKED"),
    ifelse(duplicate_markers == 0, "PASS", "WARNING"),
    ifelse(monomorphic_markers == 0, "PASS", "WARNING"),
    ifelse(f2_frequency_ok, "PASS", "WARNING"),
    ifelse(maf_ok, "PASS", "WARNING"),
    ifelse(parent_structure_ok, "PASS", "BLOCKED"),
    ifelse(ld_decay_ok, "PASS", "WARNING")
  ),
  Details = c(
    "Expected: phenotype 1000 x 6, true genetic values 1000 x 6, genotype 1000 x 4010.",
    "Phenotypic and true genetic values must be finite.",
    paste(
      "Observed:",
      paste(sort(unique(as.vector(geno_012))), collapse = ", ")
    ),
    paste(
      "Observed:",
      paste(sort(unique(as.vector(geno_m101))), collapse = ", ")
    ),
    "Expected: 10 linkage groups, 401 markers per group, 0-200 cM.",
    "Expected QTL counts: 8, 40, 80, 120, 240, 480; nominal H2: 0.7, 0.7, 0.5, 0.5, 0.3, 0.3; Model = 2.",
    "All QTL indices must correspond to marker positions.",
    "LG9 and LG10 were intentionally simulated without QTL.",
    "Phenotype, true genetic values, and genotypes use the same 1,000 row positions.",
    paste(missing_genotypes, "missing genotype cells."),
    paste(
      duplicate_individuals,
      "duplicate individual genotype profiles."
    ),
    paste(duplicate_markers, "duplicate marker profiles."),
    paste(monomorphic_markers, "monomorphic markers."),
    paste(
      "Observed proportions:",
      paste(round(as.numeric(geno_frequency), 5), collapse = ", ")
    ),
    sprintf(
      "MAF range %.4f to %.4f; median %.4f.",
      min(maf),
      max(maf),
      median(maf)
    ),
    "Two parents x ten linkage groups, with contrasting fixed marker states.",
    paste(
      "Mean r2 at 0.5, 10 and 50 cM:",
      paste(
        round(
          c(short_ld, medium_ld, ld_50),
          4
        ),
        collapse = ", "
      ),
      "; interchromosomal mean r2:",
      round(interchromosomal_ld$Mean_r2, 5)
    )
  ),
  stringsAsFactors = FALSE
)
```

## 12. Shared data object for later modules

The later modules should not independently reinterpret or reconstruct the
simulation files. One shared object is therefore saved with the quantities
prepared above.

Its most important roles are:

- **`phenotype`:** response available for fitting only within the appropriate
  analysis set;
- **`genotype_m101`:** `-1/0/1` marker representation used by the prediction
  models;
- **`true_genetic_value`:** known simulated target reserved for held-out
  evaluation;
- **map and scenario objects:** linkage groups, QTL, scenario definitions, marker
  summaries, and LD values reused unchanged downstream.

The saving code is hidden because it is workflow plumbing rather than a new
scientific calculation. The historical field name `CV_metadata` is preserved
only inside the saved objects and CSV schema for compatibility; the visible
text consistently calls this quantity the **GENES CV setting**.

```{r save-shared-data, include=FALSE}
dir.create(
  "output/simulated/audit",
  showWarnings = FALSE,
  recursive = TRUE
)

# Preserve the historical saved schema without exposing that terminology above.
control_table_saved <- control_table
names(control_table_saved)[
  names(control_table_saved) == "CV_setting"
] <- "CV_metadata"

scenario_summary_saved <- scenario_summary
names(scenario_summary_saved)[
  names(scenario_summary_saved) == "CV_setting"
] <- "CV_metadata"

write.csv(
  data_check_log,
  "output/simulated/audit/simulated_audit_summary.csv",
  row.names = FALSE
)

write.csv(
  scenario_summary_saved,
  "output/simulated/audit/simulated_scenario_summary.csv",
  row.names = FALSE
)

write.csv(
  qtl_rows[
    ,
    c(
      "Scenario",
      "QTL_index",
      "LG",
      "Position_cM"
    )
  ],
  "output/simulated/audit/simulated_qtl_map.csv",
  row.names = FALSE
)

write.csv(
  ld_summary,
  "output/simulated/audit/simulated_ld_summary.csv",
  row.names = FALSE
)

simulated_common <- list(
  metadata = list(
    dataset = "simulated",
    seed = SEED_GLOBAL,
    n_individuals = nrow(geno_012),
    n_markers = ncol(geno_012),
    n_scenarios = ncol(phenotype),
    row_identifier = seq_len(nrow(geno_012)),
    row_alignment = "row position",
    genotype_coding = "-1/0/1"
  ),
  phenotype = phenotype,
  true_genetic_value = true_genetic_value,
  residual_value = residual_value,
  genotype_012 = geno_012,
  genotype_m101 = geno_m101,
  genetic_map = genetic_map[
    ,
    c(
      "LG",
      "Marker_within_LG",
      "Type",
      "Marker",
      "Position_cM",
      "Global_marker"
    )
  ],
  scenario_control = control_table_saved,
  qtl_lists = qtl_lists,
  qtl_map = qtl_rows[
    ,
    c(
      "Scenario",
      "QTL_index",
      "LG",
      "Position_cM"
    )
  ],
  scenario_summary = scenario_summary_saved,
  marker_qc = list(
    allele_frequency = allele_frequency,
    maf = maf,
    heterozygosity = heterozygosity
  ),
  ld_summary = ld_summary,
  interchromosomal_ld = interchromosomal_ld
)

saveRDS(
  simulated_common,
  "output/simulated/audit/preprocessed_simulated_common.rds",
  compress = "gzip"
)
```

## 13. References

Cruz, C.D. (2013). GENES - a software package for analysis in experimental
statistics and quantitative genetics. *Acta Scientiarum*, 35(3), 271-276.

Cruz, C.D. (2016). Genes Software - extended and integrated with the R, Matlab
and Selegen. *Acta Scientiarum*, 38(4), 547-552.

## 14. Navigation

[Home](index.html) | [Next: Validation design](simulated_data_validation.html)
