---
title: "Simulated genomic data: structure and preparation"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

## 1. Purpose

This module introduces the simulated genomic data used throughout the genomic
prediction workflow and prepares the common objects required by the later
modules. The population was generated with the **GENES** software, which
provides tools for experimental statistics, quantitative genetics, and genetic
simulation (Cruz, 2013, 2016).

The sections follow the same sequence in which the data are used analytically.
We first inspect the phenotypic and true genetic values, then orient and recode
the marker matrix, reconstruct the genetic map, recover the six simulation
scenarios and their QTL, describe the genetic signal and genotype properties,
inspect the parental genotypes, and finally characterize linkage
disequilibrium (LD). This order allows the reader to see where each object comes
from before it is used in a calculation.

No genomic prediction model is fitted in this module. Its role is to document
the simulated population clearly and create one shared data object that is used
unchanged by the validation, GBLUP-ADE, ANN, and model-comparison modules.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)

options(stringsAsFactors = FALSE)

dir.create(
  "output/simulated/m01/figures",
  showWarnings = FALSE,
  recursive = TRUE
)
```

## 2. Packages and reproducibility

### 2.1 Packages

The analysis uses three packages, each for a distinct reader-facing task.
`knitr` converts compact R objects into formatted HTML tables so that dimensions,
scenario settings, and numerical summaries can be read directly in the page.
`ggplot2` constructs the distribution plots, genetic-map diagrams, segregation
comparison, and LD curve. `ggthemes` supplies `theme_gdocs()` and the Google Docs
color scales, giving the figures a consistent visual design across the module.

```{r packages, warning=FALSE, message=FALSE}
library(knitr)     # Format data summaries as readable HTML tables.
library(ggplot2)   # Construct the statistical and genomic figures.
library(ggthemes)  # Apply the Google Docs theme and matching color scales.
```

Every figure displayed in the page uses `theme_gdocs()` and a Google Docs color
or fill scale whenever groups are distinguished visually. A publication-quality
TIFF copy is also written to `output/simulated/m01/figures/`; the saving commands
are placed in hidden chunks so they do not interrupt the explanatory code shown
to the reader.

### 2.2 Random seed

The phenotype, true genetic values, genotypes, genetic map, and scenario
definitions are fixed inputs and therefore do not depend on a random seed.
Random sampling is required only when 5,000 pairs of markers from different
linkage groups are drawn later to obtain a background LD reference. Setting the
seed here makes that one stochastic step reproducible.

```{r random-seed}
SEED_GLOBAL <- 20260821L
set.seed(SEED_GLOBAL)
```

## 3. Phenotypic and true genetic values

The first two GENES files contain the two response-related quantities used in
the project. Both matrices have the same 1,000 individuals in rows and six
simulation scenarios in columns. They must remain aligned by row because the
phenotype is used for fitting whereas the true genetic value is reserved for
evaluating predictions.

### 3.1 Phenotypic values

The phenotypic matrix contains the observed response available to a prediction
model. After reading the file, we label the six columns C1-C6 and reshape a copy
only for plotting. The boxplots summarize center, dispersion, and possible
extreme values without changing the matrix used later in the analysis.

```{r phenotype, fig.cap="Distribution of phenotypic values across the six simulation scenarios."}
# Read phenotypes with individuals in rows and scenarios in columns.
phenotype <- as.matrix(
  read.table(
    "data/dados_simulados/DG#gen_F2_Vfen.txt",
    header = FALSE
  )
)

colnames(phenotype) <- paste0("C", 1:6)

# Reshape only a plotting copy; the original matrix remains unchanged.
phenotype_long <- stack(as.data.frame(phenotype))
names(phenotype_long) <- c("Phenotype", "Scenario")

p_phenotype <- ggplot(
  phenotype_long,
  aes(
    x = Scenario,
    y = Phenotype,
    fill = Scenario
  )
) +
  geom_boxplot(
    width = 0.65,
    show.legend = FALSE
  ) +
  scale_fill_gdocs() +
  labs(
    x = "Simulation scenario",
    y = "Phenotypic value"
  ) +
  theme_gdocs()

print(p_phenotype)
```

```{r save-phenotype-tiff, include=FALSE}
ggsave(
  "output/simulated/m01/figures/phenotype_distribution.tiff",
  plot = p_phenotype,
  device = "tiff",
  width = 8,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The file contains `r nrow(phenotype)` individuals and
`r ncol(phenotype)` scenarios. These are the responses that become available to
the model only when an individual belongs to the corresponding analysis set in
the validation design.

### 3.2 True genetic values

Because the population is simulated, the underlying genetic value of each
individual is known. These values are **not** used to create validation
partitions, tune models, or fit the final prediction equations. They are retained
as an external target for evaluating how closely predictions recover the
simulated genetic signal in held-out individuals.

The code below reads the matrix, keeps the same C1-C6 column labels, and produces
the corresponding distribution plot. Comparing this figure with the phenotype
figure helps the reader distinguish variation in the genetic component from
variation in the observed response.

```{r true-genetic-values, fig.cap="Distribution of true genetic values across the six simulation scenarios."}
# Read the known genetic values generated by GENES.
true_genetic_value <- as.matrix(
  read.table(
    "data/dados_simulados/DG#gen_F2_Vgen.dat",
    header = FALSE
  )
)

colnames(true_genetic_value) <- paste0("C", 1:6)

# Reshape a copy only for the reader-facing figure.
true_g_long <- stack(as.data.frame(true_genetic_value))
names(true_g_long) <- c("True_genetic_value", "Scenario")

p_true_g <- ggplot(
  true_g_long,
  aes(
    x = Scenario,
    y = True_genetic_value,
    fill = Scenario
  )
) +
  geom_boxplot(
    width = 0.65,
    show.legend = FALSE
  ) +
  scale_fill_gdocs() +
  labs(
    x = "Simulation scenario",
    y = "True genetic value"
  ) +
  theme_gdocs()

print(p_true_g)
```

```{r save-true-genetic-value-tiff, include=FALSE}
ggsave(
  "output/simulated/m01/figures/true_genetic_value_distribution.tiff",
  plot = p_true_g,
  device = "tiff",
  width = 8,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The matrix contains the same `r nrow(true_genetic_value)` individuals and six
scenarios as the phenotype matrix, preserving row-by-row correspondence between
the observed response and the known simulated genetic component.

## 4. Genotypic data

### 4.1 Genotype matrix

The GENES genotype file is organized with markers in rows and individuals in
columns. The prediction workflow uses the opposite orientation, so the matrix is
transposed once at reading time. From that point onward, every row represents an
individual and every column represents a marker.

A boxplot would hide the discrete structure of a marker matrix. Instead, the
table below displays actual genotype calls for ten individuals and twelve
markers spread across the full set of 4,010 markers. This makes the matrix
orientation and the original `0/1/2` coding visible before any recoding is
performed.

```{r genotype-sample}
# Read the GENES marker file and orient it as individuals x markers.
geno_012 <- t(
  as.matrix(
    read.table(
      "data/dados_simulados/genoma_mapa.txt",
      header = FALSE
    )
  )
)

# Select markers distributed from the beginning to the end of the matrix.
marker_sample <- unique(
  round(
    seq(
      1,
      ncol(geno_012),
      length.out = 12
    )
  )
)

# Show a small real sample while retaining the complete matrix for analysis.
genotype_sample <- data.frame(
  Individual = 1:10,
  geno_012[
    1:10,
    marker_sample,
    drop = FALSE
  ],
  check.names = FALSE
)

names(genotype_sample)[-1] <- paste0("M", marker_sample)

genotype_sample |>
  knitr::kable(
    format = "html",
    caption = "Sample of 10 individuals and 12 markers distributed across the genotype matrix",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The complete matrix contains `r nrow(geno_012)` individuals and
`r ncol(geno_012)` markers. The observed states are
`r paste(sort(unique(as.vector(geno_012))), collapse = ", ")`, which correspond
to the three genotype classes generated for the F2 population.

### 4.2 Genotype coding used in the prediction models

The later prediction models use the same genotype calls centered on zero.
Subtracting one converts `0/1/2` to `-1/0/1`; no individual or marker is removed
and no biological genotype call is changed. The small table makes the
transformation explicit before the recoded matrix is saved for downstream use.

```{r genotype-coding}
geno_m101 <- geno_012 - 1L

data.frame(
  Original = c(0, 1, 2),
  Analysis_coding = c(-1, 0, 1)
) |>
  knitr::kable(
    format = "html",
    caption = "Genotype coding used in the prediction analyses",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

Thus, `0 -> -1`, `1 -> 0`, and `2 -> 1`. The transformation changes the
numerical scale only.

## 5. Genetic map

The genetic map provides the physical organization needed to interpret markers
and QTL. The first 12 lines of the GENES file describe the map globally; the
marker-level records begin on line 13. For each marker we recover the linkage
group, its order within that group, its marker name, and its position in
centimorgans (cM). A global marker index is then added so the QTL indices from
the scenario file can be matched directly to this map.

```{r genetic-map-data}
# Read the marker-level portion of the GENES genetic-map file.
genetic_map <- read.table(
  "data/dados_simulados/genoma.txt",
  header = FALSE,
  sep = ",",
  skip = 12,
  fill = TRUE,
  strip.white = TRUE,
  fileEncoding = "latin1",
  stringsAsFactors = FALSE
)

genetic_map <- genetic_map[, 1:5]

names(genetic_map) <- c(
  "LG",
  "Marker_within_LG",
  "Type",
  "Marker",
  "Position_cM"
)

# Convert the map fields used in numerical calculations.
genetic_map$LG <- as.integer(genetic_map$LG)
genetic_map$Marker_within_LG <- as.integer(genetic_map$Marker_within_LG)
genetic_map$Position_cM <- as.numeric(genetic_map$Position_cM)
genetic_map$Global_marker <- seq_len(nrow(genetic_map))

genetic_map$LG_label <- factor(
  paste0("LG", genetic_map$LG),
  levels = paste0("LG", 1:10)
)

map_summary <- data.frame(
  Linkage_group = paste0("LG", 1:10),
  Markers = as.integer(table(genetic_map$LG)),
  Start_cM = as.numeric(
    tapply(genetic_map$Position_cM, genetic_map$LG, min)
  ),
  End_cM = as.numeric(
    tapply(genetic_map$Position_cM, genetic_map$LG, max)
  )
)

map_summary |>
  knitr::kable(
    format = "html",
    caption = "Structure of the simulated genetic map",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The map contains 10 linkage groups, each with 401 markers spanning 0-200 cM at
0.5-cM intervals. The complete map therefore contains 4,010 markers. In the
next section, the QTL indices defined by GENES are matched to these positions so
that the linkage groups can be represented as chromosome-like structures rather
than as simple vertical lines.

## 6. Simulation settings and QTL architecture

### 6.1 GENES software and genetic model

The six scenarios were generated with **GENES** (Cruz, 2013, 2016). The control
file identifies the genetic model as `Model = 2`. Using the expression supplied
with the simulation, the phenotype of individual \(i\) is represented by

\[
Y_i =
\mu +
\sum_j a_j +
\sum_j \sum_{j'} \alpha_j\alpha_{j'} +
\varepsilon_i .
\]

In this expression, \(Y_i\) is the simulated phenotype of individual \(i\),
\(\mu\) is the overall mean, \(\sum_j a_j\) represents contributions associated
with loci, the double sum represents pairwise interaction contributions between
loci, and \(\varepsilon_i\) is the residual component. The equation describes
how the simulated data were generated; it is distinct from the prediction
models fitted in later modules.

### 6.2 Scenario settings and QTL positions

The GENES control file contains six repeated blocks corresponding to scenarios
C1-C6. Within each block the fields occur in a fixed order. The code therefore
reads the required values directly from those positions, including the number
of selected QTL, their marker indices, the heritability specified for the
scenario, the genetic model, and the other simulation settings.

After the six blocks are read, each QTL index is matched to the genetic map from
Section 5. This provides its linkage group and position in cM. The same matched
table is used both for the figures and for the object saved for later modules.

```{r simulation-scenarios}
# Read the six scenario blocks written by GENES.
control_raw <- readLines(
  "data/dados_simulados/controlegenetico#1.dat",
  warn = FALSE,
  encoding = "latin1"
)

# Transliterate accented text so the fixed block positions are read consistently.
control_ascii <- iconv(
  control_raw,
  from = "latin1",
  to = "ASCII//TRANSLIT",
  sub = ""
)

scenario_start <- grep(
  "^Variavel",
  control_ascii
)

control_table <- data.frame(
  Scenario = paste0("C", 1:6),
  Variable = NA_real_,
  Total_loci = NA_real_,
  QTL_N = NA_real_,
  gmd = NA_real_,
  Model = NA_real_,
  H2_nominal = NA_real_,
  Constant = NA_real_,
  Scale = NA_real_,
  CV_setting = NA_real_
)

qtl_lists <- vector(
  "list",
  length(scenario_start)
)

# Recover the settings and QTL indices scenario by scenario.
for (i in seq_along(scenario_start)) {

  s <- scenario_start[i]

  control_table$Variable[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s]))
  )
  control_table$Total_loci[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s + 1]))
  )
  control_table$QTL_N[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s + 2]))
  )

  # QTL values are marker indices and are therefore kept as integers.
  qtl_lists[[i]] <- as.integer(
    scan(
      text = control_ascii[s + 3],
      what = numeric(),
      quiet = TRUE
    )
  )

  control_table$gmd[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s + 5]))
  )
  control_table$Model[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s + 6]))
  )
  control_table$H2_nominal[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s + 7]))
  )
  control_table$Constant[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s + 8]))
  )
  control_table$Scale[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s + 9]))
  )
  control_table$CV_setting[i] <- as.numeric(
    trimws(sub("^[^,]*,", "", control_ascii[s + 10]))
  )
}

# Match each QTL index to its linkage group and map position.
qtl_rows <- data.frame(
  Scenario = character(),
  QTL_index = integer(),
  LG = integer(),
  Position_cM = numeric()
)

for (i in seq_along(qtl_lists)) {

  qtl <- qtl_lists[[i]]

  qtl_rows <- rbind(
    qtl_rows,
    data.frame(
      Scenario = paste0("C", i),
      QTL_index = qtl,
      LG = genetic_map$LG[qtl],
      Position_cM = genetic_map$Position_cM[qtl]
    )
  )
}

qtl_rows$LG_label <- factor(
  paste0("LG", qtl_rows$LG),
  levels = paste0("LG", 1:10)
)

control_table |>
  knitr::kable(
    format = "html",
    digits = 3,
    caption = "Settings used to generate the six simulation scenarios in GENES",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The number of simulated QTL increases from 8 in C1 to 480 in C6, while the
specified heritability changes from 0.7 in C1-C2 to 0.5 in C3-C4 and 0.3 in
C5-C6. The remaining settings shown in the table are constant across scenarios.
The scenarios therefore become genetically more polygenic as the number of QTL
increases, while the specified heritability decreases across the same sequence.

To make the map biologically easier to read, each linkage group is represented
below as a chromosome-like colored body extending from 0 to 200 cM. Short white
ticks indicate every tenth marker; showing a subset of marker ticks avoids
turning 401 positions per linkage group into a visually solid band. Every QTL
position used in at least one scenario is retained and drawn as a diamond on
top of the corresponding linkage group.

LG9 and LG10 are deliberately different from LG1-LG8: GENES placed no QTL in
those two groups. They therefore provide genomic regions with markers but no
true simulated QTL signal, which is useful when later evaluating whether a
method prioritizes regions that actually contain simulated QTL.

```{r genetic-map-qtl, fig.cap="Chromosome-like representation of the ten linkage groups. Colored bodies represent linkage groups, white ticks show a regular subset of marker positions, and diamonds identify positions used as QTL in at least one scenario. LG9 and LG10 contain no simulated QTL."}
# Create one chromosome-like body for each linkage group.
chromosome_map <- data.frame(
  LG = 1:10,
  LG_label = factor(
    paste0("LG", 1:10),
    levels = paste0("LG", 1:10)
  ),
  Start_cM = 0,
  End_cM = 200
)

# Show every tenth marker as a readable tick on the chromosome body.
marker_ticks <- genetic_map[
  genetic_map$Marker_within_LG %% 10 == 1,
  c("LG", "LG_label", "Position_cM")
]

# A QTL position appearing in several scenarios is drawn once in the combined map.
qtl_union <- unique(
  qtl_rows[
    ,
    c("LG", "LG_label", "Position_cM")
  ]
)

p_genetic_map <- ggplot() +
  geom_segment(
    data = chromosome_map,
    aes(
      x = LG_label,
      xend = LG_label,
      y = Start_cM,
      yend = End_cM,
      color = LG_label
    ),
    linewidth = 12,
    lineend = "round",
    show.legend = FALSE
  ) +
  geom_point(
    data = marker_ticks,
    aes(
      x = LG_label,
      y = Position_cM
    ),
    shape = 95,
    size = 4.2,
    color = "white",
    alpha = 0.85
  ) +
  geom_point(
    data = qtl_union,
    aes(
      x = LG_label,
      y = Position_cM
    ),
    shape = 23,
    size = 2.4,
    fill = "white",
    color = "black",
    stroke = 0.45
  ) +
  scale_color_gdocs() +
  scale_y_reverse(
    breaks = seq(0, 200, 25),
    expand = expansion(mult = c(0.035, 0.035))
  ) +
  labs(
    x = "Linkage group",
    y = "Position (cM)"
  ) +
  theme_gdocs()

print(p_genetic_map)
```

```{r save-genetic-map-tiff, include=FALSE}
ggsave(
  "output/simulated/m01/figures/genetic_map_with_qtl.tiff",
  plot = p_genetic_map,
  device = "tiff",
  width = 9,
  height = 6,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The combined map shows where QTL can occur, but the six scenarios differ
substantially in QTL number. The next figure keeps the same linkage-group scale
and displays each scenario separately. This makes the increasing QTL density
from C1 to C6 visible while also confirming that LG9 and LG10 remain without
QTL in every scenario.

```{r qtl-map, fig.cap="Positions of simulated QTL in each scenario. The linkage-group color scheme is constant across panels; LG9 and LG10 remain without QTL in all six scenarios."}
# Draw a chromosome-like body for every linkage group in every panel.
qtl_reference <- expand.grid(
  Scenario = paste0("C", 1:6),
  LG_label = paste0("LG", 1:10),
  stringsAsFactors = FALSE
)

qtl_reference$LG_label <- factor(
  qtl_reference$LG_label,
  levels = paste0("LG", 1:10)
)

qtl_reference$Start_cM <- 0
qtl_reference$End_cM <- 200

p_qtl <- ggplot() +
  geom_segment(
    data = qtl_reference,
    aes(
      x = LG_label,
      xend = LG_label,
      y = Start_cM,
      yend = End_cM,
      color = LG_label
    ),
    linewidth = 3.6,
    lineend = "round",
    show.legend = FALSE
  ) +
  geom_point(
    data = qtl_rows,
    aes(
      x = LG_label,
      y = Position_cM
    ),
    shape = 21,
    size = 1.6,
    fill = "white",
    color = "black",
    stroke = 0.25
  ) +
  facet_wrap(
    ~Scenario,
    ncol = 2
  ) +
  scale_color_gdocs() +
  scale_y_reverse(
    breaks = seq(0, 200, 50)
  ) +
  labs(
    x = "Linkage group",
    y = "Position (cM)"
  ) +
  theme_gdocs() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

print(p_qtl)
```

```{r save-qtl-map-tiff, include=FALSE}
ggsave(
  "output/simulated/m01/figures/qtl_positions_by_scenario.tiff",
  plot = p_qtl,
  device = "tiff",
  width = 10,
  height = 8,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The exact counts behind the figure are useful when comparing scenarios. The
table below counts QTL by linkage group and makes the zeros in LG9 and LG10
explicit rather than relying only on visual inspection.

```{r qtl-counts}
qtl_table <- table(
  factor(qtl_rows$Scenario, levels = paste0("C", 1:6)),
  factor(qtl_rows$LG, levels = 1:10)
)

qtl_by_lg <- data.frame(
  Scenario = rownames(qtl_table),
  unclass(qtl_table),
  check.names = FALSE,
  row.names = NULL
)

names(qtl_by_lg)[-1] <- paste0("LG", 1:10)

qtl_by_lg |>
  knitr::kable(
    format = "html",
    caption = "Number of simulated QTL per linkage group",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

## 7. Phenotypic and genetic signal

Because the simulation provides both phenotype \(Y\) and true genetic value
\(G\), the remaining component can be calculated directly as

\[
E = Y - G.
\]

The following calculations describe how strongly the phenotype reflects the
known genetic signal in each scenario. We calculate the phenotype-genetic
correlation, the genetic-residual correlation, realized heritability, and basic
location and dispersion summaries. These quantities help explain why prediction
difficulty changes among scenarios before any model is fitted.

```{r phenotypic-genetic-signal}
scenario_names <- paste0("C", 1:6)
residual_value <- phenotype - true_genetic_value

cor_p_g <- numeric(6)
cor_g_e <- numeric(6)
h2_realized <- numeric(6)
var_g_over_var_p <- numeric(6)

# Calculate the signal summaries separately for each simulation scenario.
for (j in 1:6) {

  cor_p_g[j] <- cor(
    phenotype[, j],
    true_genetic_value[, j]
  )

  cor_g_e[j] <- cor(
    true_genetic_value[, j],
    residual_value[, j]
  )

  vg <- var(true_genetic_value[, j])
  ve <- var(residual_value[, j])

  h2_realized[j] <- vg / (vg + ve)
  var_g_over_var_p[j] <- vg / var(phenotype[, j])
}

# Gather the quantities required for interpretation and later reuse.
scenario_summary <- data.frame(
  Scenario = scenario_names,
  QTL_N = control_table$QTL_N,
  H2_nominal = control_table$H2_nominal,
  Phenotype_mean = colMeans(phenotype),
  Phenotype_sd = apply(phenotype, 2, sd),
  Genetic_mean = colMeans(true_genetic_value),
  Genetic_sd = apply(true_genetic_value, 2, sd),
  Residual_mean = colMeans(residual_value),
  Residual_sd = apply(residual_value, 2, sd),
  Cor_P_G = cor_p_g,
  Cor_G_E = cor_g_e,
  H2_realized = h2_realized,
  VarG_over_VarP = var_g_over_var_p,
  CV_empirical_pct = 100 * apply(phenotype, 2, sd) / colMeans(phenotype),
  CV_setting = control_table$CV_setting
)

scenario_summary[
  ,
  c(
    "Scenario",
    "QTL_N",
    "H2_nominal",
    "H2_realized",
    "Cor_P_G",
    "Phenotype_mean",
    "Genetic_mean",
    "Residual_sd"
  )
] |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Phenotypic and genetic signal by simulation scenario",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The phenotype-true-genetic-value correlation ranges from
`r sprintf("%.3f", min(scenario_summary$Cor_P_G))` to
`r sprintf("%.3f", max(scenario_summary$Cor_P_G))`. This variation shows that
the six scenarios do not contain the same amount of recoverable genetic signal,
which is important context for the model comparisons made later.

### 7.1 Specified and realized heritability

The heritability specified in GENES is a parameter used when generating a
scenario. Realized heritability is instead calculated from the genetic and
residual variances actually observed in this finite simulated population.
Comparing both values shows how closely each generated sample reflects its
specified setting without redefining the scenario itself.

```{r heritability-comparison, fig.cap="Heritability specified in GENES and realized heritability in the generated population."}
heritability_plot <- rbind(
  data.frame(
    Scenario = scenario_summary$Scenario,
    Heritability = scenario_summary$H2_nominal,
    Type = "GENES setting"
  ),
  data.frame(
    Scenario = scenario_summary$Scenario,
    Heritability = scenario_summary$H2_realized,
    Type = "Realized"
  )
)

h2_difference <- scenario_summary$H2_realized -
  scenario_summary$H2_nominal

p_heritability <- ggplot(
  heritability_plot,
  aes(
    x = Scenario,
    y = Heritability,
    group = Type,
    color = Type,
    shape = Type
  )
) +
  geom_line(
    linewidth = 0.8
  ) +
  geom_point(
    size = 2.5
  ) +
  scale_color_gdocs() +
  labs(
    x = "Simulation scenario",
    y = expression(H^2),
    color = NULL,
    shape = NULL
  ) +
  theme_gdocs()

print(p_heritability)
```

```{r save-heritability-tiff, include=FALSE}
ggsave(
  "output/simulated/m01/figures/heritability_specified_vs_realized.tiff",
  plot = p_heritability,
  device = "tiff",
  width = 8,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The absolute differences for C1-C6 are
`r paste(sprintf("%.4f", abs(h2_difference)), collapse = ", ")`. The largest
difference occurs in
`r scenario_summary$Scenario[which.max(abs(h2_difference))]`, with an absolute
difference of `r sprintf("%.4f", max(abs(h2_difference)))`.

### 7.2 GENES CV setting and empirical phenotype CV

GENES records `CV = 12` for all six scenarios. The empirical phenotype CV is a
different quantity: it is calculated after simulation from the standard
deviation and mean of the generated phenotype. Showing both values prevents the
input setting from being mistaken for a descriptive statistic calculated from
this particular finite population.

```{r cv-comparison, fig.cap="CV specified in GENES and empirical coefficient of variation calculated from the simulated phenotypes."}
cv_plot <- rbind(
  data.frame(
    Scenario = scenario_summary$Scenario,
    CV = scenario_summary$CV_setting,
    Type = "GENES setting"
  ),
  data.frame(
    Scenario = scenario_summary$Scenario,
    CV = scenario_summary$CV_empirical_pct,
    Type = "Empirical phenotype"
  )
)

cv_difference <- scenario_summary$CV_empirical_pct -
  scenario_summary$CV_setting

p_cv <- ggplot(
  cv_plot,
  aes(
    x = Scenario,
    y = CV,
    group = Type,
    color = Type,
    shape = Type
  )
) +
  geom_line(
    linewidth = 0.8
  ) +
  geom_point(
    size = 2.5
  ) +
  scale_color_gdocs() +
  labs(
    x = "Simulation scenario",
    y = "Coefficient of variation (%)",
    color = NULL,
    shape = NULL
  ) +
  theme_gdocs()

print(p_cv)
```

```{r save-cv-tiff, include=FALSE}
ggsave(
  "output/simulated/m01/figures/cv_genes_vs_empirical.tiff",
  plot = p_cv,
  device = "tiff",
  width = 8,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The empirical phenotype CV values are
`r paste(sprintf("%.2f", scenario_summary$CV_empirical_pct), collapse = ", ")`%.
Their differences from the GENES setting are
`r paste(sprintf("%.2f", cv_difference), collapse = ", ")` percentage points.
The two quantities are therefore retained and interpreted separately.

## 8. Genotype properties

### 8.1 F2 segregation

The two simulated parents are contrasting homozygotes, so an F2 population is
expected to approach a `1:2:1` segregation for the three genotype classes. The
next calculation uses every genotype call in the 1,000 x 4,010 matrix to compare
the observed proportions with the theoretical values 0.25, 0.50, and 0.25.

```{r genotype-segregation, fig.cap="Observed genotype proportions compared with the expected 1:2:1 segregation of an F2 population."}
# Calculate the observed frequency of each genotype state.
geno_frequency <- prop.table(
  table(
    factor(
      as.vector(geno_012),
      levels = 0:2
    )
  )
)

segregation_plot <- rbind(
  data.frame(
    Genotype = factor(0:2, levels = 0:2),
    Proportion = as.numeric(geno_frequency),
    Type = "Observed"
  ),
  data.frame(
    Genotype = factor(0:2, levels = 0:2),
    Proportion = c(0.25, 0.50, 0.25),
    Type = "F2 expectation"
  )
)

p_segregation <- ggplot(
  segregation_plot,
  aes(
    x = Genotype,
    y = Proportion,
    fill = Type
  )
) +
  geom_col(
    position = "dodge"
  ) +
  scale_fill_gdocs() +
  labs(
    x = "Genotype coding",
    y = "Proportion",
    fill = NULL
  ) +
  theme_gdocs()

print(p_segregation)
```

```{r save-genotype-segregation-tiff, include=FALSE}
ggsave(
  "output/simulated/m01/figures/f2_genotype_proportions.tiff",
  plot = p_segregation,
  device = "tiff",
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The observed proportions can be compared directly with the expected F2
proportions. This verifies that the genotype matrix has the segregation pattern
expected from the simulated cross before the markers enter prediction models.

### 8.2 Marker and individual properties

Several additional summaries describe the marker matrix from complementary
angles. Missing cells identify absent genotype calls; duplicated rows identify
individuals with identical complete marker profiles; duplicated columns identify
markers with identical profiles; monomorphic markers have no genotype
variation; minor allele frequency (MAF) describes allele balance; and
heterozygosity records the proportion of heterozygous individuals at each
marker.

These quantities are descriptive here. They are not used to construct a generic
filtering rule or to remove markers automatically.

```{r genotype-checks}
# Count missing data and repeated complete genotype profiles.
missing_genotypes <- sum(is.na(geno_012))
duplicate_individuals <- sum(duplicated(geno_012, MARGIN = 1))
duplicate_markers <- sum(duplicated(geno_012, MARGIN = 2))

# Summarize marker-level allele frequency, MAF, heterozygosity, and variance.
allele_frequency <- colMeans(geno_012) / 2
maf <- pmin(allele_frequency, 1 - allele_frequency)
heterozygosity <- colMeans(geno_012 == 1)
marker_variance <- apply(geno_012, 2, var)
monomorphic_markers <- sum(marker_variance == 0)

marker_summary <- data.frame(
  Statistic = c(
    "Missing genotype cells",
    "Duplicate individuals",
    "Duplicate markers",
    "Monomorphic markers",
    "Minimum MAF",
    "Median MAF",
    "Maximum MAF",
    "Minimum heterozygosity",
    "Median heterozygosity",
    "Maximum heterozygosity"
  ),
  Value = c(
    missing_genotypes,
    duplicate_individuals,
    duplicate_markers,
    monomorphic_markers,
    min(maf),
    median(maf),
    max(maf),
    min(heterozygosity),
    median(heterozygosity),
    max(heterozygosity)
  )
)

marker_summary |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Summary of marker and individual properties",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The complete genotype matrix is retained for downstream prediction with the same
individuals and markers described above.

## 9. Parental genotypes

The parental file connects the simulated F2 population to the two homozygous
parents from which it was generated. Each parent is represented once for every
linkage group. We read the parent identifier, linkage-group identifier, reported
number of markers, and the 401 marker states so the origin of the three F2
genotype classes can be checked directly.

```{r parental-genotypes}
# Read the parental records across the ten linkage groups.
parents <- as.matrix(
  read.table(
    "data/dados_simulados/genoma_pais.txt",
    header = FALSE
  )
)

parent_id <- parents[, 1]
parent_lg <- parents[, 2]
parent_nmarkers <- parents[, 3]
parent_genotypes <- parents[, 5:405, drop = FALSE]

# Summarize the linkage-group coverage and fixed marker state of each parent.
parent_summary <- data.frame(
  Parent = c(1, 2),
  Linkage_groups = c(
    length(unique(parent_lg[parent_id == 1])),
    length(unique(parent_lg[parent_id == 2]))
  ),
  Marker_state = c(
    unique(
      as.vector(
        parent_genotypes[
          parent_id == 1,
          ,
          drop = FALSE
        ]
      )
    )[1],
    unique(
      as.vector(
        parent_genotypes[
          parent_id == 2,
          ,
          drop = FALSE
        ]
      )
    )[1]
  )
)

parent_summary |>
  knitr::kable(
    format = "html",
    caption = "Parental genotypes used to generate the simulated F2 population",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

Both parents occur in all 10 linkage groups. Parent 1 is fixed for marker state
1 and parent 2 for marker state 0, providing the contrasting homozygous states
from which the F2 genotypes were generated.

## 10. Linkage disequilibrium

Linkage disequilibrium (LD) describes non-random association between marker
genotypes. Here it is summarized by \(r^2\), the squared correlation between
pairs of marker columns.

Two types of marker pairs are compared. **Within-linkage-group pairs** contain
two markers from the same linkage group. Because both markers lie on the same
genetic map, their separation can be expressed in cM and the change in LD with
distance can be studied directly. **Between-linkage-group pairs** contain
markers from different linkage groups. They do not share a within-group map
distance, so they are used here as a background reference for association
between markers that are not on the same linkage group.

The genetic map has 0.5-cM marker spacing. Therefore a requested distance can be
translated directly into a number of marker intervals: 10 cM, for example,
corresponds to 20 intervals. For each requested distance we pair markers within
each linkage group, calculate their genotype correlations across all
individuals, square those correlations, and summarize the resulting \(r^2\)
values. Separately, 5,000 pairs drawn from different linkage groups provide the
between-group reference.

The evaluated distances are 0.5, 1, 2.5, 5, 10, 25, 50, 100, and 200 cM. A
continuous x-axis compresses the short distances severely, so the displayed
curve uses these distances as ordered categories with equal visual spacing.
This changes only the figure layout; every LD calculation still uses the true
distance in cM.

```{r linkage-disequilibrium, fig.cap="LD across the evaluated within-linkage-group distances. The dashed line is the mean r-squared from 5,000 marker pairs drawn from different linkage groups. Distances are equally spaced on the displayed x-axis for readability, but calculations use the true distances in cM."}
# Standardize markers once so correlations can be calculated efficiently.
geno_scaled <- scale(geno_012)

ld_distances <- c(
  0.5,
  1,
  2.5,
  5,
  10,
  25,
  50,
  100,
  200
)

ld_summary <- data.frame(
  Distance_cM = ld_distances,
  Pairs = NA_integer_,
  Mean_r2 = NA_real_,
  Median_r2 = NA_real_
)

# Calculate LD for marker pairs separated by each distance within each LG.
for (d in seq_along(ld_distances)) {

  marker_shift <- as.integer(ld_distances[d] / 0.5)
  r2_all <- numeric()

  for (lg in 1:10) {

    markers <- which(genetic_map$LG == lg)

    marker_1 <- markers[
      1:(length(markers) - marker_shift)
    ]

    marker_2 <- markers[
      (1 + marker_shift):length(markers)
    ]

    r <- colSums(
      geno_scaled[, marker_1, drop = FALSE] *
        geno_scaled[, marker_2, drop = FALSE]
    ) / (nrow(geno_scaled) - 1)

    r2_all <- c(r2_all, r^2)
  }

  ld_summary$Pairs[d] <- length(r2_all)
  ld_summary$Mean_r2[d] <- mean(r2_all)
  ld_summary$Median_r2[d] <- median(r2_all)
}

# Draw marker pairs from different LGs for the background LD reference.
set.seed(SEED_GLOBAL + 101L)

n_inter_pairs <- 5000L

inter_i <- sample(
  seq_len(ncol(geno_012)),
  n_inter_pairs,
  replace = TRUE
)

inter_j <- sample(
  seq_len(ncol(geno_012)),
  n_inter_pairs,
  replace = TRUE
)

same_lg <- genetic_map$LG[inter_i] == genetic_map$LG[inter_j]

# Resample only second markers that accidentally fall in the same LG.
while (any(same_lg)) {

  inter_j[same_lg] <- sample(
    seq_len(ncol(geno_012)),
    sum(same_lg),
    replace = TRUE
  )

  same_lg <- genetic_map$LG[inter_i] == genetic_map$LG[inter_j]
}

inter_r <- colSums(
  geno_scaled[, inter_i, drop = FALSE] *
    geno_scaled[, inter_j, drop = FALSE]
) / (nrow(geno_scaled) - 1)

inter_r2 <- inter_r^2

interchromosomal_ld <- data.frame(
  Distance_cM = NA_real_,
  Pairs = length(inter_r2),
  Mean_r2 = mean(inter_r2),
  Median_r2 = median(inter_r2)
)

# Convert distances to ordered labels only for the figure.
ld_plot_data <- ld_summary
ld_plot_data$Distance_label <- factor(
  as.character(ld_plot_data$Distance_cM),
  levels = as.character(ld_distances)
)
ld_plot_data$Type <- "Within linkage group"

ld_reference_plot <- data.frame(
  Mean_r2 = interchromosomal_ld$Mean_r2,
  Type = "Between linkage groups"
)

p_ld <- ggplot(
  ld_plot_data,
  aes(
    x = Distance_label,
    y = Mean_r2,
    group = 1,
    color = Type
  )
) +
  geom_line(
    linewidth = 0.9
  ) +
  geom_point(
    size = 2.8
  ) +
  geom_hline(
    data = ld_reference_plot,
    aes(
      yintercept = Mean_r2,
      color = Type
    ),
    linetype = "dashed",
    linewidth = 0.9
  ) +
  scale_color_gdocs() +
  labs(
    x = "Map distance within linkage group (cM)",
    y = expression(mean(r^2)),
    color = NULL
  ) +
  theme_gdocs()

print(p_ld)
```

```{r save-ld-tiff, include=FALSE}
ggsave(
  "output/simulated/m01/figures/ld_decay.tiff",
  plot = p_ld,
  device = "tiff",
  width = 9,
  height = 5.5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The exact numerical summaries are shown below so that the equal visual spacing
used in the figure cannot be confused with equal genetic distances. The first
table reports LD for pairs from the same linkage group at each map distance;
the second reports the background reference calculated from pairs belonging to
different linkage groups.

```{r linkage-disequilibrium-table}
ld_summary |>
  knitr::kable(
    format = "html",
    digits = 5,
    caption = "Within-linkage-group LD by map distance",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )

interchromosomal_ld |>
  knitr::kable(
    format = "html",
    digits = 5,
    caption = "LD between markers from different linkage groups",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The observed pattern shows the expected strong decline over short and
intermediate distances. Mean \(r^2\) decreases from
`r sprintf("%.4f", ld_summary$Mean_r2[ld_summary$Distance_cM == 0.5])` at
0.5 cM to
`r sprintf("%.4f", ld_summary$Mean_r2[ld_summary$Distance_cM == 50])` at
50 cM. At 100 and 200 cM the means are
`r sprintf("%.4f", ld_summary$Mean_r2[ld_summary$Distance_cM == 100])` and
`r sprintf("%.4f", ld_summary$Mean_r2[ld_summary$Distance_cM == 200])`,
respectively, whereas the mean for markers from different linkage groups is
`r sprintf("%.5f", interchromosomal_ld$Mean_r2)`.

The fact that long-distance pairs from the same linkage group retain greater
association than the between-group reference is a property of this simulated
recombination structure. It is documented for interpretation and is not used to
remove markers.

## 11. Summary of the simulated data

The table below gathers the main structural quantities established in the
previous sections. It provides a compact checkpoint before the workflow moves
to validation design: population size, marker number, scenario number, linkage
group structure, missingness, repeated profiles, marker variation, MAF, and the
background LD level.

```{r data-summary}
data_overview <- data.frame(
  Quantity = c(
    "Individuals",
    "Markers",
    "Simulation scenarios",
    "Linkage groups",
    "Markers per linkage group",
    "Missing genotype cells",
    "Duplicate individuals",
    "Duplicate markers",
    "Monomorphic markers",
    "Minimum MAF",
    "Mean r2 between linkage groups"
  ),
  Observed = c(
    nrow(geno_012),
    ncol(geno_012),
    ncol(phenotype),
    length(unique(genetic_map$LG)),
    paste(
      unique(as.integer(table(genetic_map$LG))),
      collapse = ", "
    ),
    missing_genotypes,
    duplicate_individuals,
    duplicate_markers,
    monomorphic_markers,
    round(min(maf), 4),
    round(interchromosomal_ld$Mean_r2, 5)
  ),
  stringsAsFactors = FALSE
)

data_overview |>
  knitr::kable(
    format = "html",
    caption = "Summary of the simulated genomic data",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The summary reports what is present in the simulated population; the detailed
interpretation of QTL architecture, heritability, CV, and LD remains in the
sections where each quantity is introduced.

```{r data-check-output, include=FALSE}
dim_ok <-
  identical(dim(phenotype), c(1000L, 6L)) &&
  identical(dim(true_genetic_value), c(1000L, 6L)) &&
  identical(dim(geno_012), c(1000L, 4010L))

finite_ok <-
  all(is.finite(phenotype)) &&
  all(is.finite(true_genetic_value))

genotype_codes_ok <- identical(
  as.numeric(sort(unique(as.vector(geno_012)))),
  c(0, 1, 2)
)

analysis_codes_ok <- identical(
  as.numeric(sort(unique(as.vector(geno_m101)))),
  c(-1, 0, 1)
)

map_ok <-
  nrow(genetic_map) == 4010L &&
  length(unique(genetic_map$LG)) == 10L &&
  all(table(genetic_map$LG) == 401L) &&
  all(tapply(genetic_map$Position_cM, genetic_map$LG, min) == 0) &&
  all(tapply(genetic_map$Position_cM, genetic_map$LG, max) == 200)

scenario_ok <-
  identical(
    as.numeric(control_table$QTL_N),
    c(8, 40, 80, 120, 240, 480)
  ) &&
  isTRUE(
    all.equal(
      as.numeric(control_table$H2_nominal),
      c(0.7, 0.7, 0.5, 0.5, 0.3, 0.3),
      tolerance = 1e-12
    )
  ) &&
  all(control_table$Model == 2)

qtl_indices_ok <- all(
  qtl_rows$QTL_index >= 1 &
    qtl_rows$QTL_index <= nrow(genetic_map)
)

reference_lg_ok <- all(
  qtl_by_lg[, c("LG9", "LG10")] == 0
)

row_alignment_ok <-
  nrow(phenotype) == nrow(true_genetic_value) &&
  nrow(phenotype) == nrow(geno_012)

parent_structure_ok <-
  nrow(parents) == 20L &&
  ncol(parents) == 405L &&
  all(parent_genotypes[parent_id == 1, , drop = FALSE] == 1) &&
  all(parent_genotypes[parent_id == 2, , drop = FALSE] == 0) &&
  all(table(parent_lg) == 2L) &&
  all(parent_nmarkers == 401)

f2_frequency_ok <- max(
  abs(
    as.numeric(geno_frequency) -
      c(0.25, 0.50, 0.25)
  )
) < 0.02

maf_ok <- min(maf) > 0.40

short_ld <- ld_summary$Mean_r2[
  ld_summary$Distance_cM == 0.5
]

medium_ld <- ld_summary$Mean_r2[
  ld_summary$Distance_cM == 10
]

ld_50 <- ld_summary$Mean_r2[
  ld_summary$Distance_cM == 50
]

ld_decay_ok <-
  short_ld > medium_ld &&
  medium_ld > ld_50 &&
  ld_50 > interchromosomal_ld$Mean_r2

data_check_log <- data.frame(
  Check = c(
    "Primary matrix dimensions",
    "Finite phenotype and true genetic values",
    "Genotype coding 0/1/2",
    "Analysis genotype coding -1/0/1",
    "Genetic map structure",
    "Simulation scenario parameters",
    "QTL indices within genetic map",
    "LG9 and LG10 without simulated QTL",
    "Row alignment",
    "Missing genotypes",
    "Duplicate individuals",
    "Duplicate markers",
    "Monomorphic markers",
    "F2 genotype proportions",
    "Minor allele frequency",
    "Parental structure",
    "Short-to-medium distance LD decay"
  ),
  Status = c(
    ifelse(dim_ok, "PASS", "BLOCKED"),
    ifelse(finite_ok, "PASS", "BLOCKED"),
    ifelse(genotype_codes_ok, "PASS", "BLOCKED"),
    ifelse(analysis_codes_ok, "PASS", "BLOCKED"),
    ifelse(map_ok, "PASS", "BLOCKED"),
    ifelse(scenario_ok, "PASS", "BLOCKED"),
    ifelse(qtl_indices_ok, "PASS", "BLOCKED"),
    ifelse(reference_lg_ok, "PASS", "BLOCKED"),
    ifelse(row_alignment_ok, "PASS", "BLOCKED"),
    ifelse(missing_genotypes == 0, "PASS", "BLOCKED"),
    ifelse(duplicate_individuals == 0, "PASS", "BLOCKED"),
    ifelse(duplicate_markers == 0, "PASS", "WARNING"),
    ifelse(monomorphic_markers == 0, "PASS", "WARNING"),
    ifelse(f2_frequency_ok, "PASS", "WARNING"),
    ifelse(maf_ok, "PASS", "WARNING"),
    ifelse(parent_structure_ok, "PASS", "BLOCKED"),
    ifelse(ld_decay_ok, "PASS", "WARNING")
  ),
  Details = c(
    "Expected: phenotype 1000 x 6, true genetic values 1000 x 6, genotype 1000 x 4010.",
    "Phenotypic and true genetic values must be finite.",
    paste(
      "Observed:",
      paste(sort(unique(as.vector(geno_012))), collapse = ", ")
    ),
    paste(
      "Observed:",
      paste(sort(unique(as.vector(geno_m101))), collapse = ", ")
    ),
    "Expected: 10 linkage groups, 401 markers per group, 0-200 cM.",
    "Expected QTL counts: 8, 40, 80, 120, 240, 480; nominal H2: 0.7, 0.7, 0.5, 0.5, 0.3, 0.3; Model = 2.",
    "All QTL indices must correspond to marker positions.",
    "LG9 and LG10 were intentionally simulated without QTL.",
    "Phenotype, true genetic values, and genotypes use the same 1,000 row positions.",
    paste(missing_genotypes, "missing genotype cells."),
    paste(
      duplicate_individuals,
      "duplicate individual genotype profiles."
    ),
    paste(duplicate_markers, "duplicate marker profiles."),
    paste(monomorphic_markers, "monomorphic markers."),
    paste(
      "Observed proportions:",
      paste(round(as.numeric(geno_frequency), 5), collapse = ", ")
    ),
    sprintf(
      "MAF range %.4f to %.4f; median %.4f.",
      min(maf),
      max(maf),
      median(maf)
    ),
    "Two parents x ten linkage groups, with contrasting fixed marker states.",
    paste(
      "Mean r2 at 0.5, 10 and 50 cM:",
      paste(
        round(
          c(short_ld, medium_ld, ld_50),
          4
        ),
        collapse = ", "
      ),
      "; interchromosomal mean r2:",
      round(interchromosomal_ld$Mean_r2, 5)
    )
  ),
  stringsAsFactors = FALSE
)
```

## 12. Shared data object for later modules

All objects required by the later modules are saved together so validation and
model fitting use exactly the same individuals, marker coding, genetic map,
scenario definitions, QTL positions, and LD summaries. The phenotype remains
the response available for fitting within an analysis set, `genotype_m101`
contains the marker representation used by the models, and
`true_genetic_value` remains reserved for evaluation on held-out individuals.

The saving step is hidden because it is an implementation detail of the
workflow rather than a new scientific calculation shown to the reader. The
historical column name `CV_metadata` is restored only in the saved scenario
objects so previously versioned files and downstream code keep the same schema;
the visible analysis consistently refers to this quantity as the GENES CV
setting.

```{r save-shared-data, include=FALSE}
dir.create(
  "output/simulated/audit",
  showWarnings = FALSE,
  recursive = TRUE
)

# Preserve the historical saved schema without exposing that terminology above.
control_table_saved <- control_table
names(control_table_saved)[
  names(control_table_saved) == "CV_setting"
] <- "CV_metadata"

scenario_summary_saved <- scenario_summary
names(scenario_summary_saved)[
  names(scenario_summary_saved) == "CV_setting"
] <- "CV_metadata"

write.csv(
  data_check_log,
  "output/simulated/audit/simulated_audit_summary.csv",
  row.names = FALSE
)

write.csv(
  scenario_summary_saved,
  "output/simulated/audit/simulated_scenario_summary.csv",
  row.names = FALSE
)

write.csv(
  qtl_rows[
    ,
    c(
      "Scenario",
      "QTL_index",
      "LG",
      "Position_cM"
    )
  ],
  "output/simulated/audit/simulated_qtl_map.csv",
  row.names = FALSE
)

write.csv(
  ld_summary,
  "output/simulated/audit/simulated_ld_summary.csv",
  row.names = FALSE
)

simulated_common <- list(
  metadata = list(
    dataset = "simulated",
    seed = SEED_GLOBAL,
    n_individuals = nrow(geno_012),
    n_markers = ncol(geno_012),
    n_scenarios = ncol(phenotype),
    row_identifier = seq_len(nrow(geno_012)),
    row_alignment = "row position",
    genotype_coding = "-1/0/1"
  ),
  phenotype = phenotype,
  true_genetic_value = true_genetic_value,
  residual_value = residual_value,
  genotype_012 = geno_012,
  genotype_m101 = geno_m101,
  genetic_map = genetic_map[
    ,
    c(
      "LG",
      "Marker_within_LG",
      "Type",
      "Marker",
      "Position_cM",
      "Global_marker"
    )
  ],
  scenario_control = control_table_saved,
  qtl_lists = qtl_lists,
  qtl_map = qtl_rows[
    ,
    c(
      "Scenario",
      "QTL_index",
      "LG",
      "Position_cM"
    )
  ],
  scenario_summary = scenario_summary_saved,
  marker_qc = list(
    allele_frequency = allele_frequency,
    maf = maf,
    heterozygosity = heterozygosity
  ),
  ld_summary = ld_summary,
  interchromosomal_ld = interchromosomal_ld
)

saveRDS(
  simulated_common,
  "output/simulated/audit/preprocessed_simulated_common.rds",
  compress = "gzip"
)
```

## 13. References

Cruz, C.D. (2013). GENES - a software package for analysis in experimental
statistics and quantitative genetics. *Acta Scientiarum*, 35(3), 271-276.

Cruz, C.D. (2016). Genes Software - extended and integrated with the R, Matlab
and Selegen. *Acta Scientiarum*, 38(4), 547-552.

## 14. Navigation

[Home](index.html) | [Next: Validation design](simulated_data_validation.html)
