---
title: "Simulated data: GBLUP-ADE"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_depth: 3
    toc_float: true
    code_download: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

## 1. Purpose

This module fits the GBLUP-ADE benchmark to the six simulated scenarios under
the validation design defined in M02. Each scenario is evaluated on the same 25
outer held-out splits, giving **6 x 25 = 150 scenario-by-split fits**.

Four principles define the analysis:

- **Phenotype is the fitting response:** only phenotype values from the 800
  outer-analysis individuals are available to a model fit.
- **Outer assessment remains held out:** the corresponding 200 phenotype values
  are replaced by `NA` before fitting.
- **Genotypes are available for prediction:** marker profiles from all 1,000
  individuals can be used to construct genomic relationship matrices because no
  held-out phenotype or true genetic value enters those matrices.
- **True genetic value is an evaluation target:** it is accessed only after the
  held-out predictions have been generated.

The benchmark separates additive, dominance, and epistatic genomic components.
Its results are later compared with the regularized ANN branch under exactly the
same outer validation assignments.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)

options(stringsAsFactors = FALSE)

dir.create(
  "output/simulated/m03/figures",
  showWarnings = FALSE,
  recursive = TRUE
)
```

## 2. Packages and fixed fitting settings

### 2.1 Packages

Four packages support the module:

- **`knitr`** formats the displayed tables.
- **`ggplot2`** summarizes predictive performance across scenarios and outer
  splits.
- **`ggthemes`** applies the same Google Docs visual style used in M01 and M02.
- **`sommer`** constructs the genomic relationship matrices and fits GBLUP-ADE
  when a new run is explicitly requested.

```{r packages, warning=FALSE, message=FALSE}
RUN_GBLUP <- identical(Sys.getenv("RUN_GBLUP", "0"), "1")

library(knitr)     # Format tables shown in the page.
library(ggplot2)   # Plot held-out predictive performance.
library(ggthemes)  # Apply the Google Docs theme and palettes.

if (RUN_GBLUP) {
  suppressPackageStartupMessages(
    library(sommer) # Build genomic kernels and fit GBLUP-ADE.
  )
}
```

### 2.2 Fixed settings

A normal page build uses the saved GBLUP-ADE results rather than refitting all
150 models. Setting the environment variable `RUN_GBLUP=1` activates the full
fitting workflow.

The statistical benchmark itself has two fixed controls:

- **Global seed:** `20260821`.
- **`sommer::mmes()` iterations:** 20 for every scenario-by-split fit.

The iteration count is part of the prespecified benchmark and is not tuned from
outer-assessment performance.

```{r gblup-control}
SEED_GLOBAL <- 20260821L
N_ITERS_SOMMER <- 20L

fitting_control <- data.frame(
  Parameter = c(
    "Global seed",
    "Run new GBLUP-ADE fits",
    "sommer iterations per fit"
  ),
  Value = c(
    SEED_GLOBAL,
    RUN_GBLUP,
    N_ITERS_SOMMER
  ),
  Role = c(
    "Reproduce the fitting workflow.",
    "Control whether all 150 fits are executed.",
    "Use the same fixed fitting control for every task."
  )
)

fitting_control |>
  knitr::kable(
    format = "html",
    caption = "Fixed GBLUP-ADE execution and fitting settings",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

## 3. Inputs and held-out information

The model branch starts from two objects prepared earlier in the workflow:

- the shared simulated-data object from M01;
- the nested validation design from M02.

The next chunk reads these objects directly and makes their roles explicit.

```{r inputs}
common <- readRDS(
  "output/simulated/audit/preprocessed_simulated_common.rds"
)
cv <- readRDS(
  "output/simulated/validation/simulated_cv_design.rds"
)

phenotype <- common$phenotype
true_g <- common$true_genetic_value
geno <- common$genotype_m101
control <- common$scenario_control

ids <- seq_len(nrow(geno))
id_labels <- sprintf("ID%04d", ids)
rownames(geno) <- id_labels
scenario_names <- colnames(phenotype)

input_summary <- data.frame(
  Object = c(
    "Phenotype",
    "True genetic value",
    "Genotype",
    "Outer validation splits"
  ),
  Structure = c(
    paste(dim(phenotype), collapse = " x "),
    paste(dim(true_g), collapse = " x "),
    paste(dim(geno), collapse = " x "),
    length(cv$outer_splits)
  ),
  Role = c(
    "Response used for fitting after outer assessment is masked.",
    "Primary held-out evaluation target only.",
    "Marker information used to construct A, D, and E.",
    "Common held-out evaluation design used by all models."
  ),
  stringsAsFactors = FALSE
)

input_summary |>
  knitr::kable(
    format = "html",
    caption = "Objects entering the GBLUP-ADE workflow",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The dimensions establish the data available to every task:

- **1,000 individuals** are aligned across phenotype, true genetic value, and
  genotype matrices;
- phenotype and true genetic value contain the same **six scenarios**;
- the genotype matrix contains **4,010 markers**;
- the validation object supplies the same **25 outer splits** used throughout
  the project.

The crucial separation is by information type, not by availability of genotype.
The 200 outer-assessment individuals retain their marker profiles but their
phenotypes are hidden before model fitting, and their true genetic values are
not read for evaluation until predictions already exist.

## 4. GBLUP-ADE model

### 4.1 Statistical model

For one scenario, the benchmark is

\[
\mathbf{y} = \mathbf{1}\mu + \mathbf{a} + \mathbf{d} + \mathbf{e} +
\boldsymbol{\varepsilon},
\]

where

\[
\mathbf{a}\sim N(0,\mathbf{A}\sigma_a^2),\qquad
\mathbf{d}\sim N(0,\mathbf{D}\sigma_d^2),
\]

\[
\mathbf{e}\sim N(0,\mathbf{E}\sigma_e^2),\qquad
\boldsymbol{\varepsilon}\sim N(0,\mathbf{I}\sigma_\varepsilon^2).
\]

The components have distinct roles:

- \(\mu\): overall intercept;
- \(\mathbf{a}\): additive genomic effect;
- \(\mathbf{d}\): dominance genomic effect;
- \(\mathbf{e}\): epistatic genomic effect;
- \(\boldsymbol{\varepsilon}\): residual variation.

### 4.2 Genomic relationship matrices

The three random effects are associated with genomic relationship matrices
constructed from the same marker matrix:

```{r relationship-matrix-roles}
relationship_matrices <- data.frame(
  Matrix = c("A", "D", "E"),
  Genetic_component = c(
    "Additive",
    "Dominance",
    "Epistatic"
  ),
  Constructor = c(
    "sommer::A.mat()",
    "sommer::D.mat()",
    "sommer::E.mat()"
  ),
  Interpretation = c(
    "Additive genomic covariance among individuals.",
    "Dominance genomic covariance among individuals.",
    "Epistatic genomic covariance used by this benchmark."
  )
)

relationship_matrices |>
  knitr::kable(
    format = "html",
    caption = "Genomic relationship matrices used by GBLUP-ADE",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

These matrices describe genomic covariance among individuals. They do not use
phenotype or true genetic value, and they are built once and reused across all
150 fits.

### 4.3 Prediction from the fitted components

In `sommer::mmes()`, `ism()` identifies the individual levels associated with a
random effect and `vsm()` links that effect to its genomic relationship matrix
through `Gu`.

The fitted genomic component is

\[
\widehat{G}=
\widehat{a}+\widehat{d}+\widehat{e}.
\]

The final prediction stored by this benchmark is

\[
\widehat{y}=
\widehat{\mu}+\widehat{G}.
\]

Thus the output preserves both the total prediction and the additive,
dominance, epistatic, and summed genomic components for every held-out
individual.

## 5. Evaluation metrics

Two targets are evaluated after prediction:

- **Primary target — true genetic value:** measures recovery of the known
  simulated genetic signal.
- **Secondary target — phenotype:** measures prediction of the observed response,
  which also contains residual variation.

For each target, the stored metrics are correlation, RMSE, MAE, bias,
calibration slope, and \(R^2\). Bias is defined as
`mean(predicted - observed)`, so positive values indicate overprediction. The
calibration slope is estimated from `observed ~ predicted`, and
\(R^2=1-SSE/SST\).

```{r gblup-metrics}
safe_cor <- function(pred, obs) {
  cor(pred, obs)
}

safe_rmse <- function(pred, obs) {
  sqrt(
    mean(
      (pred - obs)^2
    )
  )
}

safe_mae <- function(pred, obs) {
  mean(
    abs(
      pred - obs
    )
  )
}

safe_bias <- function(pred, obs) {
  mean(
    pred - obs
  )
}

safe_slope <- function(pred, obs) {
  as.numeric(
    coef(
      lm(
        obs ~ pred
      )
    )[2]
  )
}

safe_r2 <- function(pred, obs) {
  total_variation <- sum(
    (
      obs -
      mean(obs)
    )^2
  )

  1 -
    sum(
      (
        obs -
        pred
      )^2
    ) /
    total_variation
}

metrics <- function(
  pred,
  obs,
  prefix
) {
  values <- c(
    safe_cor(pred, obs),
    safe_rmse(pred, obs),
    safe_mae(pred, obs),
    safe_bias(pred, obs),
    safe_slope(pred, obs),
    safe_r2(pred, obs)
  )

  names(values) <- paste0(
    prefix,
    c(
      "_Correlation",
      "_RMSE",
      "_MAE",
      "_Bias",
      "_Slope",
      "_R2"
    )
  )

  values
}
```

## 6. Scenario-by-split workload

The six scenarios are crossed with the 25 outer splits from M02. This produces
150 distinct fitting tasks while preserving the same held-out individuals used
by the ANN branch.

```{r task-table}
outer_meta <- unique(
  cv$outer_assignments[
    , c("Repetition", "OuterFold", "SplitID")
  ]
)

outer_meta <- outer_meta[
  order(
    outer_meta$Repetition,
    outer_meta$OuterFold
  ),
]

task_table <- merge(
  data.frame(
    Scenario = scenario_names,
    stringsAsFactors = FALSE
  ),
  outer_meta,
  by = NULL
)

task_table <- task_table[
  order(
    match(task_table$Scenario, scenario_names),
    task_table$Repetition,
    task_table$OuterFold
  ),
]

task_table$TaskID <- paste(
  task_table$Scenario,
  task_table$SplitID,
  sep = "_"
)
rownames(task_table) <- NULL

head(task_table, 15) |>
  knitr::kable(
    format = "html",
    caption = "First 15 of the 150 GBLUP-ADE scenario-by-split tasks",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

`TaskID` links each fit to its 200 held-out predictions and later provides the
common scenario-by-split key used for model comparison.

## 7. Fitting procedure

The fitting sequence is the same for every task:

1. construct A, D, and E once from the complete genotype matrix;
2. identify the 800 analysis and 200 assessment individuals for the task;
3. replace the 200 assessment phenotypes by `NA`;
4. fit the additive + dominance + epistatic mixed model;
5. recover the three fitted genomic components;
6. form the held-out prediction;
7. access phenotype and true genetic value only for the 200 assessment
   individuals and calculate the stored metrics.

The same genomic kernels are reused for all tasks. Only the scenario response
and outer assessment set change from one fit to another.

### 7.1 Recovering the genomic components

`sommer::mmes()` stores the random-effect estimates in `fit$uList`. The helper
below maps a fitted effect back to the 1,000 individual positions through its
incidence matrix.

```{r extract-genomic-component}
extract_u <- function(
  fit,
  dat,
  factor_name,
  effect_name
) {
  random_effect <- as.matrix(
    fit$uList[[effect_name]]
  )

  incidence_matrix <- model.matrix(
    as.formula(
      paste0(
        "~",
        factor_name,
        " - 1"
      )
    ),
    data = dat
  )

  as.numeric(
    incidence_matrix %*%
      random_effect[
        ,
        1L,
        drop = FALSE
      ]
  )
}
```

The fitting loop calls this function for `id/A`, `idd/D`, and `ide/E` to recover
additive, dominance, and epistatic predictions, respectively.

### 7.2 One model call

For one scenario and one outer split, the mixed-model call is:

```{r gblup-fit-example, eval=FALSE}
fit <- sommer::mmes(
  y ~ 1,
  random = ~
    vsm(ism(id),  Gu = A_matrix) +
    vsm(ism(idd), Gu = D_matrix) +
    vsm(ism(ide), Gu = E_matrix),
  rcov = ~ units,
  nIters = 20,
  data = dat,
  verbose = FALSE
)
```

At this point `dat$y` contains phenotype values for the 800 outer-analysis
individuals and `NA` for the 200 outer-assessment individuals.

### 7.3 Complete 150-task workflow

The full code is retained so the saved benchmark can be regenerated explicitly.
It runs only when `RUN_GBLUP=1`; the standard page build skips these expensive
fits and reads the existing saved results instead.

```{r fit-gblup-ade, eval=RUN_GBLUP}
set.seed(SEED_GLOBAL)

run_start <- proc.time()[["elapsed"]]

# Build A, D, and E once from the marker matrix.
kernel_start <- proc.time()[["elapsed"]]

A_matrix <- sommer::A.mat(geno)
D_matrix <- sommer::D.mat(geno)
E_matrix <- sommer::E.mat(geno)

kernel_seconds <- proc.time()[["elapsed"]] - kernel_start

rownames(A_matrix) <- colnames(A_matrix) <- id_labels
rownames(D_matrix) <- colnames(D_matrix) <- id_labels
rownames(E_matrix) <- colnames(E_matrix) <- id_labels

kernel_summary <- data.frame(
  Kernel = c(
    "Additive",
    "Dominance",
    "Epistatic"
  ),
  Rows = c(
    nrow(A_matrix),
    nrow(D_matrix),
    nrow(E_matrix)
  ),
  Columns = c(
    ncol(A_matrix),
    ncol(D_matrix),
    ncol(E_matrix)
  ),
  Symmetric = c(
    isTRUE(
      all.equal(
        A_matrix,
        t(A_matrix),
        tolerance = 1e-8
      )
    ),
    isTRUE(
      all.equal(
        D_matrix,
        t(D_matrix),
        tolerance = 1e-8
      )
    ),
    isTRUE(
      all.equal(
        E_matrix,
        t(E_matrix),
        tolerance = 1e-8
      )
    )
  ),
  Finite = c(
    all(is.finite(A_matrix)),
    all(is.finite(D_matrix)),
    all(is.finite(E_matrix))
  ),
  Mean_diagonal = c(
    mean(diag(A_matrix)),
    mean(diag(D_matrix)),
    mean(diag(E_matrix))
  ),
  Kernel_build_seconds = rep(
    kernel_seconds,
    3L
  ),
  stringsAsFactors = FALSE
)

result_list <- vector(
  "list",
  nrow(task_table)
)
prediction_list <- vector(
  "list",
  nrow(task_table)
)

for (i in seq_len(nrow(task_table))) {

  task <- task_table[i, ]
  scenario <- task$Scenario
  scenario_index <- match(
    scenario,
    scenario_names
  )

  # Recover the fixed outer split defined in M02.
  split <- cv$outer_splits[[task$SplitID]]

  analysis_index <- match(
    split$analysis_ids,
    ids
  )
  assessment_index <- match(
    split$assessment_ids,
    ids
  )

  phenotype_vector <- phenotype[
    ,
    scenario_index
  ]
  true_genetic_vector <- true_g[
    ,
    scenario_index
  ]

  id_factor <- factor(
    id_labels,
    levels = id_labels
  )

  dat <- data.frame(
    y = phenotype_vector,
    id = id_factor,
    idd = id_factor,
    ide = id_factor
  )

  # Remove all response information from the outer assessment set.
  dat$y[assessment_index] <- NA_real_

  fit_start <- proc.time()[["elapsed"]]

  # Fit the prespecified additive + dominance + epistatic model.
  fit <- sommer::mmes(
    y ~ 1,
    random = ~
      vsm(ism(id),  Gu = A_matrix) +
      vsm(ism(idd), Gu = D_matrix) +
      vsm(ism(ide), Gu = E_matrix),
    rcov = ~ units,
    nIters = N_ITERS_SOMMER,
    data = dat,
    verbose = FALSE,
    dateWarning = FALSE
  )

  fit_seconds <-
    proc.time()[["elapsed"]] -
    fit_start

  prediction_start <- proc.time()[["elapsed"]]

  # Recover additive, dominance, and epistatic effects for all individuals.
  additive <- extract_u(
    fit,
    dat,
    "id",
    "vsm(ism(id), Gu = A_matrix)"
  )
  dominance <- extract_u(
    fit,
    dat,
    "idd",
    "vsm(ism(idd), Gu = D_matrix)"
  )
  epistatic <- extract_u(
    fit,
    dat,
    "ide",
    "vsm(ism(ide), Gu = E_matrix)"
  )

  gebv <-
    additive +
    dominance +
    epistatic

  intercept <- as.numeric(fit$b)[1L]
  predicted_value <- intercept + gebv

  predict_seconds <-
    proc.time()[["elapsed"]] -
    prediction_start

  # Evaluate only the 200 held-out individuals.
  observed_test <- phenotype_vector[
    assessment_index
  ]
  true_g_test <- true_genetic_vector[
    assessment_index
  ]
  predicted_test <- predicted_value[
    assessment_index
  ]
  gebv_test <- gebv[
    assessment_index
  ]

  genetic_metrics <- metrics(
    predicted_test,
    true_g_test,
    "Genetic"
  )
  phenotype_metrics <- metrics(
    predicted_test,
    observed_test,
    "Phenotype"
  )

  meta <- control[
    control$Scenario == scenario,
    ,
    drop = FALSE
  ]

  result_list[[i]] <- data.frame(
    Dataset = "simulated",
    Scenario = scenario,
    QTL_N = meta$QTL_N,
    H2_nominal = meta$H2_nominal,
    Repetition = task$Repetition,
    Fold = task$OuterFold,
    SplitID = task$SplitID,
    TaskID = task$TaskID,
    Model = "GBLUP_ADE",
    N_train = length(analysis_index),
    N_test = length(assessment_index),
    N_pred_finite = sum(is.finite(predicted_test)),
    Genetic_Correlation =
      genetic_metrics["Genetic_Correlation"],
    Genetic_RMSE =
      genetic_metrics["Genetic_RMSE"],
    Genetic_MAE =
      genetic_metrics["Genetic_MAE"],
    Genetic_Bias =
      genetic_metrics["Genetic_Bias"],
    Genetic_Slope =
      genetic_metrics["Genetic_Slope"],
    Genetic_R2 =
      genetic_metrics["Genetic_R2"],
    GEBV_TrueG_Correlation =
      safe_cor(gebv_test, true_g_test),
    Phenotype_Correlation =
      phenotype_metrics["Phenotype_Correlation"],
    Phenotype_RMSE =
      phenotype_metrics["Phenotype_RMSE"],
    Phenotype_MAE =
      phenotype_metrics["Phenotype_MAE"],
    Phenotype_Bias =
      phenotype_metrics["Phenotype_Bias"],
    Phenotype_Slope =
      phenotype_metrics["Phenotype_Slope"],
    Phenotype_R2 =
      phenotype_metrics["Phenotype_R2"],
    Kernel_build_seconds = kernel_seconds,
    Fit_seconds = fit_seconds,
    Predict_seconds = predict_seconds,
    Status = "ok",
    Error = NA_character_,
    stringsAsFactors = FALSE
  )

  prediction_list[[i]] <- data.frame(
    Dataset = "simulated",
    Scenario = scenario,
    QTL_N = meta$QTL_N,
    H2_nominal = meta$H2_nominal,
    Repetition = task$Repetition,
    Fold = task$OuterFold,
    SplitID = task$SplitID,
    TaskID = task$TaskID,
    Model = "GBLUP_ADE",
    ID = ids[assessment_index],
    Observed_Phenotype = observed_test,
    True_Genetic_Value = true_g_test,
    Predicted_Value = predicted_test,
    GEBV = gebv_test,
    Additive_Prediction = additive[assessment_index],
    Dominance_Prediction = dominance[assessment_index],
    Epistatic_Prediction = epistatic[assessment_index],
    Status = "ok",
    stringsAsFactors = FALSE
  )

  cat(
    sprintf(
      "[%03d/%03d] %s | rG=%.4f | RMSE_G=%.4f | fit=%.2fs\n",
      i,
      nrow(task_table),
      task$TaskID,
      result_list[[i]]$Genetic_Correlation,
      result_list[[i]]$Genetic_RMSE,
      fit_seconds
    )
  )
}

results <- do.call(
  rbind,
  result_list
)
predictions <- do.call(
  rbind,
  prediction_list
)
rownames(results) <- NULL
rownames(predictions) <- NULL

run_seconds <-
  proc.time()[["elapsed"]] -
  run_start

run_metadata <- list(
  dataset = "simulated",
  model = "GBLUP_ADE",
  seed_global = SEED_GLOBAL,
  sommer_version = as.character(
    packageVersion("sommer")
  ),
  nIters_sommer = N_ITERS_SOMMER,
  tasks = nrow(results),
  prediction_rows = nrow(predictions),
  kernel_build_seconds = kernel_seconds,
  total_run_seconds = run_seconds,
  timestamp = as.character(Sys.time())
)
```

```{r save-gblup-results, eval=RUN_GBLUP, include=FALSE}
write.csv(
  kernel_summary,
  "output/simulated/gblup/simulated_gblup_ade_kernel_summary.csv",
  row.names = FALSE
)

saveRDS(
  results,
  "output/simulated/gblup/simulated_gblup_ade_results.rds",
  compress = "gzip"
)
saveRDS(
  predictions,
  "output/simulated/gblup/simulated_gblup_ade_predictions.rds",
  compress = "gzip"
)
write.csv(
  results,
  "output/simulated/gblup/simulated_gblup_ade_results.csv",
  row.names = FALSE
)
write.csv(
  predictions,
  "output/simulated/gblup/simulated_gblup_ade_predictions.csv",
  row.names = FALSE
)
saveRDS(
  run_metadata,
  "output/simulated/gblup/simulated_gblup_ade_run_metadata.rds",
  compress = "gzip"
)
```

## 8. Genomic-kernel diagnostics

The saved kernel summary records numerical properties of the three relationship
matrices from the completed benchmark run.

```{r kernel-diagnostics}
kernel_summary <- read.csv(
  "output/simulated/gblup/simulated_gblup_ade_kernel_summary.csv",
  stringsAsFactors = FALSE
)

kernel_display <- kernel_summary[
  , c(
    "Kernel",
    "Rows",
    "Columns",
    "Symmetric",
    "Finite",
    "Mean_diagonal"
  )
]

kernel_display |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Numerical structure of the GBLUP-ADE genomic kernels",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The three kernels are **1,000 x 1,000**, finite, and symmetric. Their mean
diagonals are close to one, confirming the numerical structure expected for the
relationship matrices used in the fitted benchmark.

## 9. Saved results and predictive performance

The normal page build now reads the completed GBLUP-ADE outputs. This keeps the
scientific results fixed while allowing the explanation, tables, and figures to
be rebuilt quickly.

```{r saved-results}
results <- readRDS(
  "output/simulated/gblup/simulated_gblup_ade_results.rds"
)
predictions <- readRDS(
  "output/simulated/gblup/simulated_gblup_ade_predictions.rds"
)
run_metadata <- readRDS(
  "output/simulated/gblup/simulated_gblup_ade_run_metadata.rds"
)

result_structure <- data.frame(
  Object = c(
    "Task-level results",
    "Held-out predictions"
  ),
  Rows = c(
    nrow(results),
    nrow(predictions)
  ),
  Unique_tasks = c(
    length(unique(results$TaskID)),
    length(unique(predictions$TaskID))
  ),
  Rows_per_task = c(
    1,
    paste(
      range(
        as.integer(
          table(predictions$TaskID)
        )
      ),
      collapse = " to "
    )
  )
)

result_structure |>
  knitr::kable(
    format = "html",
    caption = "Stored GBLUP-ADE result structure",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

The stored benchmark contains one task-level result row for each of the 150
scenario-by-split fits and 200 held-out predictions per task.

### 9.1 Scenario means

The primary and secondary metrics are averaged across the 25 outer splits within
each scenario.

```{r performance-summary}
ok_results <- results[
  results$Status == "ok",
  ,
  drop = FALSE
]

scenario_performance <- aggregate(
  cbind(
    Genetic_Correlation,
    Genetic_RMSE,
    Genetic_MAE,
    Genetic_Bias,
    Genetic_Slope,
    Genetic_R2,
    Phenotype_Correlation,
    Phenotype_RMSE,
    Phenotype_MAE,
    Phenotype_Bias,
    Phenotype_Slope,
    Phenotype_R2
  ) ~ Scenario + QTL_N + H2_nominal,
  data = ok_results,
  FUN = mean,
  na.rm = TRUE
)

scenario_performance <- scenario_performance[
  order(
    match(
      scenario_performance$Scenario,
      scenario_names
    )
  ),
]

genetic_performance_display <- scenario_performance[
  , c(
    "Scenario",
    "QTL_N",
    "H2_nominal",
    "Genetic_Correlation",
    "Genetic_RMSE",
    "Genetic_MAE",
    "Genetic_Bias",
    "Genetic_Slope",
    "Genetic_R2"
  )
]

phenotype_performance_display <- scenario_performance[
  , c(
    "Scenario",
    "QTL_N",
    "H2_nominal",
    "Phenotype_Correlation",
    "Phenotype_RMSE",
    "Phenotype_MAE",
    "Phenotype_Bias",
    "Phenotype_Slope",
    "Phenotype_R2"
  )
]
```

#### True genetic value

```{r genetic-performance}
genetic_performance_display |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Primary evaluation: true genetic-value prediction across the 25 outer splits",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

Across the six scenarios, mean held-out genetic correlation ranges from
`r sprintf("%.4f", min(scenario_performance$Genetic_Correlation))` to
`r sprintf("%.4f", max(scenario_performance$Genetic_Correlation))`, while mean
genetic RMSE ranges from
`r sprintf("%.4f", min(scenario_performance$Genetic_RMSE))` to
`r sprintf("%.4f", max(scenario_performance$Genetic_RMSE))`.

The C1-C6 sequence represents a prespecified genetic-complexity gradient. QTL
number and nominal heritability change together across that sequence, so the
scenario pattern should not be interpreted as the isolated causal effect of
either quantity.

#### Phenotype

```{r phenotype-performance}
phenotype_performance_display |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "Secondary evaluation: phenotype prediction across the 25 outer splits",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

Phenotype is retained as a secondary target because it contains both genetic and
residual variation. The known simulated genetic value remains the primary target
for evaluating recovery of the simulated genetic signal.

### 9.2 Held-out performance across outer splits

Scenario means do not show how much performance changes with the particular
individuals held out. The next two figures therefore display all 25 outer-split
values within each scenario.

```{r genetic-correlation-figure, fig.cap="Held-out true genetic-value correlation for GBLUP-ADE across the 25 outer splits in each scenario. Boxes summarize the split distribution; white diamonds indicate scenario means."}
p_genetic_correlation <- ggplot(
  ok_results,
  aes(
    x = Scenario,
    y = Genetic_Correlation,
    fill = Scenario
  )
) +
  geom_boxplot(
    width = 0.58,
    outlier.shape = NA,
    alpha = 0.70
  ) +
  geom_jitter(
    width = 0.10,
    height = 0,
    alpha = 0.45,
    size = 1.8
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 3.2,
    fill = "white"
  ) +
  scale_fill_gdocs() +
  labs(
    x = "Simulation scenario",
    y = "Correlation with true genetic value",
    fill = NULL
  ) +
  theme_gdocs() +
  theme(
    legend.position = "none"
  )

print(p_genetic_correlation)
```

```{r save-genetic-correlation-tiff, include=FALSE}
ggsave(
  "output/simulated/m03/figures/gblup_genetic_correlation_by_scenario.tiff",
  plot = p_genetic_correlation,
  device = "tiff",
  width = 8.5,
  height = 5.5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

```{r genetic-rmse-figure, fig.cap="Held-out true genetic-value RMSE for GBLUP-ADE across the 25 outer splits in each scenario. Lower values indicate smaller prediction error; white diamonds indicate scenario means."}
p_genetic_rmse <- ggplot(
  ok_results,
  aes(
    x = Scenario,
    y = Genetic_RMSE,
    fill = Scenario
  )
) +
  geom_boxplot(
    width = 0.58,
    outlier.shape = NA,
    alpha = 0.70
  ) +
  geom_jitter(
    width = 0.10,
    height = 0,
    alpha = 0.45,
    size = 1.8
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 3.2,
    fill = "white"
  ) +
  scale_fill_gdocs() +
  labs(
    x = "Simulation scenario",
    y = "RMSE for true genetic value",
    fill = NULL
  ) +
  theme_gdocs() +
  theme(
    legend.position = "none"
  )

print(p_genetic_rmse)
```

```{r save-genetic-rmse-tiff, include=FALSE}
ggsave(
  "output/simulated/m03/figures/gblup_genetic_rmse_by_scenario.tiff",
  plot = p_genetic_rmse,
  device = "tiff",
  width = 8.5,
  height = 5.5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

The figures complement the scenario means in two ways:

- **correlation:** larger values indicate stronger agreement in ranking and
  linear association with the known genetic signal;
- **RMSE:** smaller values indicate less absolute prediction error on the scale
  of the true genetic value.

The spread within each scenario records sensitivity to the outer assessment
sample rather than a new source of model tuning.

## 10. Split-level variability

A compact numeric summary of the same 25 outer-split distributions is retained
for reproducibility and direct comparison across scenarios.

```{r split-summary}
split_summary <- do.call(
  rbind,
  lapply(
    split(
      ok_results,
      ok_results$Scenario
    ),
    function(x) {
      data.frame(
        Scenario = unique(x$Scenario),
        N = nrow(x),
        Cor_mean = mean(
          x$Genetic_Correlation,
          na.rm = TRUE
        ),
        Cor_sd = sd(
          x$Genetic_Correlation,
          na.rm = TRUE
        ),
        Cor_min = min(
          x$Genetic_Correlation,
          na.rm = TRUE
        ),
        Cor_max = max(
          x$Genetic_Correlation,
          na.rm = TRUE
        ),
        RMSE_mean = mean(
          x$Genetic_RMSE,
          na.rm = TRUE
        ),
        RMSE_sd = sd(
          x$Genetic_RMSE,
          na.rm = TRUE
        ),
        MAE_mean = mean(
          x$Genetic_MAE,
          na.rm = TRUE
        ),
        R2_mean = mean(
          x$Genetic_R2,
          na.rm = TRUE
        )
      )
    }
  )
)

rownames(split_summary) <- NULL
split_summary <- split_summary[
  order(
    match(
      split_summary$Scenario,
      scenario_names
    )
  ),
]

split_summary |>
  knitr::kable(
    format = "html",
    digits = 4,
    caption = "True genetic-value prediction variability across outer splits",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )
```

Reporting the mean, standard deviation, minimum, and maximum correlation prevents
one favorable or unfavorable split from determining the interpretation of a
scenario.

## 11. Computational time

Kernel construction is a one-time operation, whereas model fitting and
prediction are repeated across the 150 tasks. The timing summary keeps those
costs separate.

```{r time-summary}
time_summary <- aggregate(
  cbind(
    Fit_seconds,
    Predict_seconds
  ) ~ Scenario,
  data = results,
  FUN = mean,
  na.rm = TRUE
)

time_summary$Kernel_build_seconds <-
  run_metadata$kernel_build_seconds

time_summary |>
  knitr::kable(
    format = "html",
    digits = 3,
    caption = "GBLUP-ADE computational time",
    row.names = FALSE,
    table.attr = 'class="table table-striped table-hover table-bordered"'
  )

cat(
  "Total run time:",
  round(
    run_metadata$total_run_seconds,
    2
  ),
  "seconds.\n"
)
```

Because A, D, and E are built once and reused, their construction time should
not be interpreted as a cost incurred separately for every outer split.

## 12. Navigation

[Previous: Validation design](simulated_data_validation.html) | [Parallel branch: Regularized ANN](simulated_data_ann.html) | [Next: Model comparison](simulated_data_comparison.html)
