---
title: "Regularized artificial neural networks for simulated genomic data"
author:
  - Weverton Gomes da Costa
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

# Purpose

This module is the **canonical ANN workflow** for the simulated genomic-data
branch. ANN hyperparameter tuning and final outer-CV fitting are internal
stages of one modeling method, parallel to GBLUP in the public pipeline.

The four variants are `ANN_0`, `ANN_L1`, `ANN_L2`, and `ANN_L1_L2`
(displayed as **ANN-L1+L2**). They use the same multilayer perceptron and differ
only in L1/L2 weight penalties.

# Tuning contract

Only four quantities are treated as hyperparameters:

1. hidden-layer architecture;
2. Adam learning rate;
3. L1 penalty coefficient;
4. L2 penalty coefficient.

The following are **fixed optimization controls**, not tuned quantities:

- activation: ReLU;
- optimizer: Adam;
- dropout: 0;
- maximum epochs available to an early-stopped fit: 2000;
- patience: 25;
- minimum validation improvement: \(10^{-6}\);
- early-validation fraction: 0.15.

## Optimization concepts used in this workflow

These controls describe **how each candidate network is optimized**; they are
distinct from the scientific comparison among ANN variants.

- **Adam** is the gradient-based optimizer used to update the network weights.
  It adapts the magnitude of parameter updates from the recent history of the
  gradients. In this workflow Adam is fixed as the optimizer; it is not compared
  with alternative optimization algorithms.
- The **learning rate** controls the size of the optimization steps taken by
  Adam. It is a tuned quantity and is selected together with network
  architecture using `ANN_0` within each outer task.
- An **epoch** is one complete pass through the fitting observations used in a
  particular network fit. The number of epochs completed is therefore an
  optimization quantity, not a new cross-validation replicate.
- **Early stopping** monitors loss on the separate early-validation subset
  during optimization and stops training when additional epochs no longer
  produce sufficient improvement. The frozen inner-assessment and outer-
  assessment observations are not used for this stopping decision.
- **Patience = 25** means that optimization stops after 25 consecutive epochs
  without a qualifying improvement in early-validation loss.
- **Minimum improvement (`min_delta`) = \(10^{-6}\)** defines the decrease in
  early-validation loss required to count as an improvement and reset the
  patience counter. Changes smaller than this threshold do not reset patience.

Thus, early stopping determines how long an individual optimization run is
allowed to continue, whereas nested cross-validation determines which
prespecified architecture, learning rate, and penalty configuration is
selected.

The 2000-epoch value is a **convergence ceiling**, not a fixed final training
length. Epoch diagnostics are excluded from hyperparameter ranking. After
hyperparameters are selected, a separate early-stopping fit uses only the
outer-analysis sample to select `Best_epoch`; the final network is then
reinitialized and refit on all 800 outer-analysis individuals for exactly
`Best_epoch`.

# Architecture range

The canonical architecture catalog defines a bounded small-to-moderate
feed-forward network search space. The candidate hidden-layer structures are
`2`, `4`, `8`, `16`, `32`, `8-4`, `16-8`, `32-16`, and `64-32`.

The `64-32` architecture is the prespecified upper bound of the canonical
search space. Consequently, all ANN results and the subsequent comparison with
GBLUP-ADE are interpreted **within the prespecified canonical model/architecture
search space**, rather than as a comparison against every possible neural-network
architecture.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
options(stringsAsFactors = FALSE)

WIDE_TABLE_ATTR <- paste(
  'style="display:block;',
  'overflow-x:auto;',
  'white-space:nowrap;',
  'width:100%;"'
)

find_project_root <- function(start = getwd()) {
  path <- normalizePath(start, winslash = "/", mustWork = TRUE)

  repeat {
    if (file.exists(file.path(path, "_workflowr.yml"))) {
      return(path)
    }

    parent <- dirname(path)

    if (identical(parent, path)) {
      stop("Could not locate workflowr project root.")
    }

    path <- parent
  }
}

PROJECT_ROOT <- find_project_root()

project_path <- function(...) {
  file.path(PROJECT_ROOT, ...)
}


SEED_GLOBAL <- 20260821L

RUN_ANN_TUNING <- identical(
  Sys.getenv("RUN_ANN_TUNING", "0"),
  "1"
)
RUN_ANN <- identical(
  Sys.getenv("RUN_ANN", "0"),
  "1"
)
ANN_RUNTIME_REQUIRED <- RUN_ANN_TUNING || RUN_ANN

ANN_TUNING_FORCE <- identical(
  Sys.getenv("ANN_TUNING_FORCE", "0"),
  "1"
)
ANN_TUNING_PILOT <- identical(
  Sys.getenv("ANN_TUNING_PILOT", "0"),
  "1"
)
ANN_TUNING_PILOT_TASK <- Sys.getenv(
  "ANN_TUNING_PILOT_TASK",
  "C1_R01_F01"
)

MAX_EPOCHS <- 2000L
# Compatibility token only: preserves the exact config_id of the completed
# ANN_TUNING_CANONICAL_V1 run. It does not control final refit length.
TUNING_V1_COMPATIBILITY_TOKEN <- 2000L
PATIENCE <- 25L
MIN_DELTA <- 1e-6
EARLY_VALIDATION_FRACTION <- 0.15
DROPOUT <- 0

ARCHITECTURES <- list(
  `2` = c(2L),
  `4` = c(4L),
  `8` = c(8L),
  `16` = c(16L),
  `32` = c(32L),
  `8-4` = c(8L, 4L),
  `16-8` = c(16L, 8L),
  `32-16` = c(32L, 16L),
  `64-32` = c(64L, 32L)
)

LEARNING_RATES <- c(
  1.25e-4,
  2.5e-4,
  5e-4,
  1e-3
)

L1_VALUES <- c(
  1e-5,
  1e-4,
  1e-3,
  1e-2,
  3e-2,
  1e-1
)

L2_VALUES <- c(
  1e-5,
  1e-4,
  1e-3,
  1e-2
)

# Full Cartesian grid for the simultaneous L1+L2 model.
# Both coefficients are strictly positive in ANN_L1_L2.
L1_L2_GRID <- expand.grid(
  L1 = L1_VALUES,
  L2 = L2_VALUES,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

MODEL_LEVELS <- c(
  "ANN_0",
  "ANN_L1",
  "ANN_L2",
  "ANN_L1_L2"
)

DATA_FILE <- project_path("output", "simulated", "audit", "preprocessed_simulated_common.rds")
CV_FILE <- project_path("output", "simulated", "validation", "simulated_cv_design.rds")
DATA_AUDIT_FILE <- project_path("output", "simulated", "audit", "simulated_audit_summary.csv")
CV_AUDIT_FILE <- project_path("output", "simulated", "validation", "simulated_cv_design_audit.csv")

CHECKPOINT_DIR <- project_path(".local", "checkpoints", "canonical", "ann_tuning")

OUTPUT_STEM <- if (ANN_TUNING_PILOT) {
  project_path(".local", "pilot", "ann", "tuning", "simulated_ann_tuning_pilot")
} else {
  project_path("output", "simulated", "ann", "tuning", "simulated_ann_tuning")
}

CANDIDATE_RDS <- paste0(OUTPUT_STEM, "_candidate_summary.rds")
FOLD_RDS <- paste0(OUTPUT_STEM, "_fold_results.rds")
SELECTED_RDS <- paste0(OUTPUT_STEM, "_selected_configurations.rds")
RUN_RDS <- paste0(OUTPUT_STEM, "_run_metadata.rds")
MANIFEST_RDS <- paste0(OUTPUT_STEM, "_manifest.rds")

CANDIDATE_CSV <- paste0(OUTPUT_STEM, "_candidate_summary.csv")
FOLD_CSV <- paste0(OUTPUT_STEM, "_fold_results.csv")
SELECTED_CSV <- paste0(OUTPUT_STEM, "_selected_configurations.csv")
MANIFEST_CSV <- paste0(OUTPUT_STEM, "_manifest.csv")

ANN_TUNING_CONFIG_ID <- paste(
  "ANN_TUNING_CANONICAL_V1",
  paste(names(ARCHITECTURES), collapse = "-"),
  paste(LEARNING_RATES, collapse = "-"),
  paste(L1_VALUES, collapse = "-"),
  paste(L2_VALUES, collapse = "-"),
  MAX_EPOCHS,
  PATIENCE,
  MIN_DELTA,
  EARLY_VALIDATION_FRACTION,
  TUNING_V1_COMPATIBILITY_TOKEN,
  sep = "__"
)

if (RUN_ANN_TUNING) {
  dir.create(
    CHECKPOINT_DIR,
    recursive = TRUE,
    showWarnings = FALSE
  )
  dir.create(
    dirname(CANDIDATE_RDS),
    recursive = TRUE,
    showWarnings = FALSE
  )
}
```

```{r packages, include=FALSE}
if (ANN_RUNTIME_REQUIRED) {
  if (!requireNamespace("torch", quietly = TRUE)) {
    stop(
      paste(
        "Package 'torch' is required when",
        "RUN_ANN_TUNING=1 or RUN_ANN=1."
      )
    )
  }

  suppressPackageStartupMessages(
    library(torch)
  )

  backend_ok <- tryCatch({
    z <- torch::torch_tensor(
      c(1, 2),
      dtype = torch::torch_float()
    )
    all(as.numeric(z) == c(1, 2))
  }, error = function(e) FALSE)

  if (!backend_ok) {
    stop(
      "torch is installed but the libtorch backend is unavailable."
    )
  }

  cuda_available <- isTRUE(
    tryCatch(
      torch::cuda_is_available(),
      error = function(e) FALSE
    )
  )

  requested_device <- tolower(
    Sys.getenv("ANN_DEVICE", "auto")
  )

  if (!requested_device %in% c("auto", "cpu", "cuda")) {
    stop("ANN_DEVICE must be auto, cpu, or cuda.")
  }

  if (requested_device == "cuda" && !cuda_available) {
    stop(
      "ANN_DEVICE=cuda requested but CUDA is unavailable."
    )
  }

  ANN_DEVICE_LABEL <- if (
    requested_device == "cuda" ||
    (requested_device == "auto" && cuda_available)
  ) {
    "cuda"
  } else {
    "cpu"
  }

  ANN_DEVICE <- torch::torch_device(
    ANN_DEVICE_LABEL
  )

  sync_device <- function() {
    if (ANN_DEVICE_LABEL == "cuda") {
      try(
        torch::cuda_synchronize(),
        silent = TRUE
      )
    }
    invisible(NULL)
  }

  set_ann_seed <- function(seed) {
    set.seed(as.integer(seed))
    torch::torch_manual_seed(
      as.integer(seed)
    )
    invisible(NULL)
  }

  torch_version_display <-
    as.character(packageVersion("torch"))
  backend_display <- backend_ok
  cuda_display <- cuda_available
  device_display <- ANN_DEVICE_LABEL
} else {
  backend_ok <- NA
  cuda_available <- NA
  ANN_DEVICE_LABEL <-
    "not initialized (saved-output render)"
  ANN_DEVICE <- NULL

  sync_device <- function() {
    invisible(NULL)
  }

  set_ann_seed <- function(seed) {
    stop(
      paste(
        "ANN runtime is not initialized.",
        "Set RUN_ANN_TUNING=1 or RUN_ANN=1."
      )
    )
  }

  torch_version_display <-
    "not initialized (saved-output render)"
  backend_display <-
    "not initialized (saved-output render)"
  cuda_display <-
    "not initialized (saved-output render)"
  device_display <- ANN_DEVICE_LABEL
}

environment_table <- data.frame(
  Item = c(
    "torch version",
    "backend available",
    "CUDA available",
    "device",
    "MAX_EPOCHS",
    "TUNING_V1_COMPATIBILITY_TOKEN",
    "patience",
    "minimum improvement",
    "early-validation fraction",
    "dropout"
  ),
  Value = c(
    torch_version_display,
    backend_display,
    cuda_display,
    device_display,
    MAX_EPOCHS,
    TUNING_V1_COMPATIBILITY_TOKEN,
    PATIENCE,
    MIN_DELTA,
    EARLY_VALIDATION_FRACTION,
    DROPOUT
  )
)

knitr::kable(
  environment_table,
  caption = "Fixed ANN computational protocol"
)
```

# Canonical inputs

This module uses the audited common preprocessing object and the frozen nested
validation design established in the preceding modules. These fixed inputs keep
model tuning and outer-CV evaluation aligned across all four ANN variants.
Their dimensions, genotype coding, scenario names, and frozen split counts are
verified internally before any ANN tuning summary is interpreted.

```{r inputs, include=FALSE}
required_files <- c(
  DATA_FILE,
  CV_FILE,
  DATA_AUDIT_FILE,
  CV_AUDIT_FILE
)

if (!all(file.exists(required_files))) {
  stop(
    "Missing required file(s): ",
    paste(
      required_files[!file.exists(required_files)],
      collapse = ", "
    )
  )
}

data_audit <- read.csv(
  DATA_AUDIT_FILE,
  stringsAsFactors = FALSE
)
cv_audit <- read.csv(
  CV_AUDIT_FILE,
  stringsAsFactors = FALSE
)

if (any(data_audit$Status == "BLOCKED")) {
  stop(
    "Simulated-data audit contains BLOCKED checks."
  )
}
if (any(cv_audit$Status == "BLOCKED")) {
  stop(
    "Validation-design audit contains BLOCKED checks."
  )
}

common <- readRDS(DATA_FILE)
cv <- readRDS(CV_FILE)

phenotype <- common$phenotype
geno <- common$genotype_m101

ids <- as.integer(
  common$metadata$row_identifier
)
if (is.null(ids) || !length(ids)) {
  ids <- seq_len(nrow(geno))
}

scenario_names <- colnames(phenotype)
if (is.null(scenario_names)) {
  scenario_names <- paste0("C", 1:6)
}

input_checks <- data.frame(
  Check = c(
    "Phenotype is 1000 x 6",
    "Genotype is 1000 x 4010",
    "Genotype coding is -1/0/1",
    "25 frozen outer splits",
    "100 frozen inner splits",
    "Scenario names are C1-C6"
  ),
  PASS = c(
    identical(
      dim(phenotype),
      c(1000L, 6L)
    ),
    identical(
      dim(geno),
      c(1000L, 4010L)
    ),
    identical(
      sort(unique(as.vector(geno))),
      -1:1
    ),
    length(cv$outer_splits) == 25L,
    sum(vapply(
      cv$inner_splits,
      length,
      integer(1)
    )) == 100L,
    identical(
      scenario_names,
      paste0("C", 1:6)
    )
  )
)

input_checks$Status <- ifelse(
  input_checks$PASS,
  "PASS",
  "BLOCKED"
)

knitr::kable(
  input_checks[, c("Check", "Status")],
  caption = "Canonical ANN tuning pre-fit audit"
)

if (any(!input_checks$PASS)) {
  stop(
    "Blocking ANN tuning pre-fit check detected."
  )
}
```

# Candidate grids

Architecture and learning rate are selected **once per outer task using
`ANN_0`**. The selected structure is then shared by all four ANN variants. This
keeps the comparison between regularization variants focused on L1/L2
penalties rather than allowing each model family to change network capacity.

After structure selection:

- `ANN_L1` tunes only L1;
- `ANN_L2` tunes only L2;
- `ANN_L1_L2` tunes the full positive L1 × L2 Cartesian grid.

```{r grids}
structure_grid <- expand.grid(
  Architecture = names(ARCHITECTURES),
  Learning_rate = LEARNING_RATES,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)
structure_grid$Model <- "ANN_0"
structure_grid$L1 <- 0
structure_grid$L2 <- 0

l1_grid <- data.frame(
  Model = "ANN_L1",
  L1 = L1_VALUES,
  L2 = 0,
  stringsAsFactors = FALSE
)

l2_grid <- data.frame(
  Model = "ANN_L2",
  L1 = 0,
  L2 = L2_VALUES,
  stringsAsFactors = FALSE
)

l1l2_grid <- data.frame(
  Model = "ANN_L1_L2",
  L1 = L1_L2_GRID$L1,
  L2 = L1_L2_GRID$L2,
  stringsAsFactors = FALSE
)

grid_summary <- data.frame(
  Stage = c(
    "Shared architecture + learning rate",
    "ANN_L1 penalty",
    "ANN_L2 penalty",
    "ANN_L1_L2 penalties"
  ),
  Candidates = c(
    nrow(structure_grid),
    nrow(l1_grid),
    nrow(l2_grid),
    nrow(l1l2_grid)
  )
)

grid_summary$Inner_fits_per_outer_task <-
  4L * grid_summary$Candidates

knitr::kable(
  grid_summary,
  caption = "Canonical ANN tuning grid"
)

CANDIDATES_PER_TASK <- sum(
  grid_summary$Candidates
)
INNER_FITS_PER_TASK <- 4L *
  CANDIDATES_PER_TASK

cat(
  "Candidates per outer task:",
  CANDIDATES_PER_TASK,
  "\n"
)
cat(
  "Inner candidate-fold fits per outer task:",
  INNER_FITS_PER_TASK,
  "\n"
)
cat(
  "Planned inner fits for all 150 outer tasks:",
  150L * INNER_FITS_PER_TASK,
  "\n"
)
```

The grid contains 70 candidate configurations per outer task. With four frozen inner folds, this corresponds to 280 candidate-fold fits per task and 42,000 fits across the 150 scenario-by-split tasks. Architecture and learning rate are selected using `ANN_0` and then held common across the four ANN variants within each task, so their subsequent contrast isolates the L1/L2 penalty choice.

# Within-fold preprocessing

For every inner fit, marker means and standard deviations are estimated only
from the fitting subset. Those same parameters are then applied to the
corresponding early-validation and inner-assessment individuals. Phenotypes are
standardized from the fitting subset in the same way, and predictions are
returned to the original phenotype scale before calculating performance
metrics. Markers with effectively zero fitting-set variance receive unit scale
to avoid numerical division by zero.

This ordering prevents validation or assessment observations from influencing
the preprocessing parameters used to fit the network.

```{r support-functions, include=FALSE}
safe_cor <- function(p, o) {
  z <- is.finite(p) & is.finite(o)

  if (
    sum(z) < 3L ||
    sd(p[z]) == 0 ||
    sd(o[z]) == 0
  ) {
    return(NA_real_)
  }

  cor(p[z], o[z])
}

safe_rmse <- function(p, o) {
  z <- is.finite(p) & is.finite(o)

  if (!any(z)) {
    return(NA_real_)
  }

  sqrt(
    mean(
      (p[z] - o[z])^2
    )
  )
}

safe_mae <- function(p, o) {
  z <- is.finite(p) & is.finite(o)

  if (!any(z)) {
    return(NA_real_)
  }

  mean(
    abs(
      p[z] - o[z]
    )
  )
}

parameter_count <- function(
  input_dim,
  hidden_units
) {
  inputs <- c(
    input_dim,
    hidden_units
  )
  outputs <- c(
    hidden_units,
    1L
  )

  sum(
    inputs * outputs +
      outputs
  )
}

make_validation_split <- function(
  x,
  frac,
  seed
) {
  set.seed(seed)

  n_validation <- max(
    1L,
    as.integer(
      round(
        length(x) * frac
      )
    )
  )

  validation <- sort(
    sample(
      x,
      n_validation
    )
  )

  list(
    train = sort(
      setdiff(
        x,
        validation
      )
    ),
    validation = validation
  )
}

fit_x_scaler <- function(x) {
  center <- colMeans(x)
  scale <- apply(
    x,
    2,
    sd
  )

  zero <- !is.finite(scale) |
    scale < 1e-8

  scale[zero] <- 1

  list(
    center = center,
    scale = scale,
    zero = zero
  )
}

scale_x <- function(x, scaler) {
  sweep(
    sweep(
      x,
      2,
      scaler$center,
      "-"
    ),
    2,
    scaler$scale,
    "/"
  )
}

fit_y_scaler <- function(y) {
  scale <- sd(y)

  if (
    !is.finite(scale) ||
    scale < 1e-8
  ) {
    scale <- 1
  }

  list(
    center = mean(y),
    scale = scale
  )
}

scale_y <- function(y, scaler) {
  (
    y - scaler$center
  ) / scaler$scale
}

unscale_y <- function(y, scaler) {
  y * scaler$scale +
    scaler$center
}
```

# Shared ANN implementation

For standardized marker vector \(\mathbf{x}\), hidden layers apply affine
transformations followed by ReLU activation,

\[
\mathbf{h}^{(0)}=\mathbf{x}, \qquad
\mathbf{h}^{(\ell)}=\operatorname{ReLU}(\mathbf{W}_{\ell}\mathbf{h}^{(\ell-1)}+\mathbf{b}_{\ell}),
\]

with a linear output layer

\[
\hat{y}=\mathbf{w}_{o}^{\mathsf{T}}\mathbf{h}^{(L)}+b_o.
\]

Training minimizes

\[
\mathcal{L}=
\operatorname{MSE}(y,\hat{y})+
\lambda_1\sum_j |w_j|+
\frac{\lambda_2}{2}\sum_j w_j^2.
\]

In this objective, \(y\) is the observed phenotype on the standardized scale
defined from the current fitting subset, and \(\hat{y}\) is the corresponding
network prediction on that same scale during optimization. The terms \(w_j\)
denote the penalized network weights, while \(\lambda_1\) and \(\lambda_2\)
control the strengths of the L1 and L2 penalties, respectively. Bias parameters
are excluded from both penalty sums.

Standardization is used only to fit the network on a common numerical scale.
Before inner-assessment or outer-assessment performance is calculated,
predictions are transformed back with the fitting-set phenotype mean and
standard deviation, so the reported prediction errors are expressed on the
original phenotype scale.

`ANN_0` sets both penalty coefficients to zero, `ANN_L1` uses a selected
positive \(\lambda_1\) with \(\lambda_2=0\), `ANN_L2` uses a selected
positive \(\lambda_2\) with \(\lambda_1=0\), and `ANN_L1_L2` uses selected
positive values of both coefficients. Penalties are applied to network weights but not biases. Adam uses
`weight_decay = 0`, so the L2 contribution enters only through the explicit
`weight_penalty()` term shown below.

```{r model-definition, eval=ANN_RUNTIME_REQUIRED}
TorchGenomicMLP <- torch::nn_module(
  "TorchGenomicMLP_CanonicalTuning",

  initialize = function(
    input_dim,
    hidden_units
  ) {
    self$linears <-
      torch::nn_module_list()

    previous <-
      as.integer(input_dim)

    for (h in hidden_units) {
      self$linears$append(
        torch::nn_linear(
          previous,
          as.integer(h)
        )
      )
      previous <- as.integer(h)
    }

    self$output <-
      torch::nn_linear(
        previous,
        1L
      )
  },

  forward = function(x) {
    if (length(self$linears) > 0L) {
      for (
        i in seq_len(
          length(self$linears)
        )
      ) {
        x <- torch::nnf_relu(
          self$linears[[i]](x)
        )
      }
    }

    self$output(x)
  }
)

weight_penalty <- function(
  model,
  l1,
  l2
) {
  weights <- lapply(
    seq_len(
      length(model$linears)
    ),
    function(i) {
      model$linears[[i]]$weight
    }
  )

  weights[[length(weights) + 1L]] <-
    model$output$weight

  zero <- weights[[1]]$sum() * 0
  penalty_l1 <- zero
  penalty_l2 <- zero

  for (w in weights) {
    penalty_l1 <-
      penalty_l1 +
      w$abs()$sum()

    penalty_l2 <-
      penalty_l2 +
      w$pow(2)$sum()
  }

  l1 * penalty_l1 +
    (l2 / 2) * penalty_l2
}

clone_state <- function(model) {
  state <- model$state_dict()

  copy <- lapply(
    state,
    function(x) {
      x$detach()$clone()
    }
  )

  names(copy) <- names(state)

  copy
}
```

# Inner-CV training

The inner analysis set is divided deterministically into:

- a fitting subset;
- an early-validation subset.

The frozen inner assessment set remains untouched until prediction of that
candidate. Preprocessing is estimated from the fitting subset only.

Early stopping is an optimization mechanism. Its epoch diagnostics are saved
but excluded from the tuning rule.

```{r training-functions}
prepare_bundle <- function(
  fit_ids,
  validation_ids,
  assessment_ids,
  scenario_index
) {
  fit_index <- match(
    fit_ids,
    ids
  )
  validation_index <- match(
    validation_ids,
    ids
  )
  assessment_index <- match(
    assessment_ids,
    ids
  )

  if (
    anyNA(
      c(
        fit_index,
        validation_index,
        assessment_index
      )
    )
  ) {
    stop(
      "ANN ID alignment failed."
    )
  }

  x_scaler <- fit_x_scaler(
    geno[
      fit_index,
      ,
      drop = FALSE
    ]
  )

  y_scaler <- fit_y_scaler(
    phenotype[
      fit_index,
      scenario_index
    ]
  )

  list(
    x_fit = scale_x(
      geno[
        fit_index,
        ,
        drop = FALSE
      ],
      x_scaler
    ),
    x_validation = scale_x(
      geno[
        validation_index,
        ,
        drop = FALSE
      ],
      x_scaler
    ),
    x_assessment = scale_x(
      geno[
        assessment_index,
        ,
        drop = FALSE
      ],
      x_scaler
    ),
    y_fit = scale_y(
      phenotype[
        fit_index,
        scenario_index
      ],
      y_scaler
    ),
    y_validation = scale_y(
      phenotype[
        validation_index,
        scenario_index
      ],
      y_scaler
    ),
    y_assessment = phenotype[
      assessment_index,
      scenario_index
    ],
    y_scaler = y_scaler,
    zero_variance_markers =
      sum(x_scaler$zero)
  )
}

train_inner_candidate <- function(
  bundle,
  architecture,
  learning_rate,
  l1,
  l2,
  seed
) {
  set_ann_seed(seed)

  x_fit <- torch::torch_tensor(
    bundle$x_fit,
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  y_fit <- torch::torch_tensor(
    matrix(
      bundle$y_fit,
      ncol = 1
    ),
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  x_validation <- torch::torch_tensor(
    bundle$x_validation,
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  y_validation <- torch::torch_tensor(
    matrix(
      bundle$y_validation,
      ncol = 1
    ),
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  x_assessment <- torch::torch_tensor(
    bundle$x_assessment,
    dtype = torch::torch_float()
  )$to(device = ANN_DEVICE)

  model <- TorchGenomicMLP(
    ncol(bundle$x_fit),
    ARCHITECTURES[[architecture]]
  )$to(device = ANN_DEVICE)

  optimizer <- torch::optim_adam(
    params = model$parameters,
    lr = learning_rate,
    weight_decay = 0
  )

  best_validation_loss <- Inf
  checkpoint_epoch <- NA_integer_
  checkpoint_state <- NULL
  stale <- 0L
  last_epoch <- 0L
  stopped_by_patience <- FALSE

  sync_device()
  fit_start <-
    proc.time()[["elapsed"]]

  for (
    epoch in seq_len(MAX_EPOCHS)
  ) {
    last_epoch <- epoch

    model$train()
    optimizer$zero_grad()

    prediction <-
      model(x_fit)

    loss <-
      torch::nnf_mse_loss(
        prediction,
        y_fit
      ) +
      weight_penalty(
        model,
        l1,
        l2
      )

    loss$backward()
    optimizer$step()

    model$eval()

    validation_loss <-
      torch::with_no_grad({
        torch::nnf_mse_loss(
          model(x_validation),
          y_validation
        )
      })

    validation_value <-
      as.numeric(
        validation_loss$to(
          device =
            torch::torch_device("cpu")
        )
      )

    if (
      is.finite(validation_value) &&
      validation_value <
        best_validation_loss -
        MIN_DELTA
    ) {
      best_validation_loss <-
        validation_value

      checkpoint_epoch <-
        epoch

      checkpoint_state <-
        clone_state(model)

      stale <- 0L
    } else {
      stale <- stale + 1L
    }

    if (stale >= PATIENCE) {
      stopped_by_patience <- TRUE
      break
    }
  }

  sync_device()
  fit_seconds <-
    proc.time()[["elapsed"]] -
    fit_start

  if (
    is.null(checkpoint_state) ||
    is.na(checkpoint_epoch)
  ) {
    stop(
      "No finite early-validation checkpoint."
    )
  }

  model$load_state_dict(
    checkpoint_state
  )
  model$eval()

  sync_device()
  prediction_start <-
    proc.time()[["elapsed"]]

  prediction_scaled <-
    torch::with_no_grad({
      as.numeric(
        model(x_assessment)$to(
          device =
            torch::torch_device("cpu")
        )
      )
    })

  sync_device()
  prediction_seconds <-
    proc.time()[["elapsed"]] -
    prediction_start

  list(
    prediction = unscale_y(
      prediction_scaled,
      bundle$y_scaler
    ),
    early_stop_epoch =
      as.integer(checkpoint_epoch),
    last_epoch =
      as.integer(last_epoch),
    stopped_by_patience =
      stopped_by_patience,
    hit_epoch_ceiling =
      last_epoch >= MAX_EPOCHS &&
      !stopped_by_patience,
    fit_seconds = fit_seconds,
    prediction_seconds =
      prediction_seconds,
    zero_variance_markers =
      bundle$zero_variance_markers
  )
}
```

# Candidate evaluation and selection

Candidate ranking is deterministic and does **not** include epoch information.

The ranking hierarchy is:

1. lower mean inner-assessment RMSE;
2. higher mean inner-assessment correlation;
3. fewer network parameters;
4. smaller total penalty coefficient;
5. smaller learning rate;
6. candidate ID.

```{r candidate-functions}
evaluate_candidate <- function(
  bundles,
  model_name,
  architecture,
  learning_rate,
  l1,
  l2,
  seed_base,
  candidate_id
) {
  rows <- vector(
    "list",
    4L
  )

  for (k in 1:4) {
    fitted <- tryCatch(
      train_inner_candidate(
        bundles[[k]],
        architecture,
        learning_rate,
        l1,
        l2,
        seed_base + k
      ),
      error = function(e) e
    )

    if (
      inherits(
        fitted,
        "error"
      )
    ) {
      rows[[k]] <- data.frame(
        Model = model_name,
        CandidateID = candidate_id,
        Architecture = architecture,
        Learning_rate = learning_rate,
        L1 = l1,
        L2 = l2,
        InnerFold = k,
        RMSE = NA_real_,
        Correlation = NA_real_,
        MAE = NA_real_,
        EarlyStop_epoch = NA_integer_,
        Last_epoch = NA_integer_,
        Stopped_by_patience = NA,
        Hit_epoch_ceiling = NA,
        Fit_seconds = NA_real_,
        Prediction_seconds = NA_real_,
        Status = "error",
        Error = conditionMessage(fitted),
        stringsAsFactors = FALSE
      )
    } else {
      rows[[k]] <- data.frame(
        Model = model_name,
        CandidateID = candidate_id,
        Architecture = architecture,
        Learning_rate = learning_rate,
        L1 = l1,
        L2 = l2,
        InnerFold = k,
        RMSE = safe_rmse(
          fitted$prediction,
          bundles[[k]]$y_assessment
        ),
        Correlation = safe_cor(
          fitted$prediction,
          bundles[[k]]$y_assessment
        ),
        MAE = safe_mae(
          fitted$prediction,
          bundles[[k]]$y_assessment
        ),
        EarlyStop_epoch =
          fitted$early_stop_epoch,
        Last_epoch =
          fitted$last_epoch,
        Stopped_by_patience =
          fitted$stopped_by_patience,
        Hit_epoch_ceiling =
          fitted$hit_epoch_ceiling,
        Fit_seconds =
          fitted$fit_seconds,
        Prediction_seconds =
          fitted$prediction_seconds,
        Status = "ok",
        Error = NA_character_,
        stringsAsFactors = FALSE
      )
    }
  }

  folds <- do.call(
    rbind,
    rows
  )

  ok <- folds$Status == "ok"

  summary <- data.frame(
    Model = model_name,
    CandidateID = candidate_id,
    Architecture = architecture,
    Learning_rate = learning_rate,
    L1 = l1,
    L2 = l2,
    Inner_folds_ok = sum(ok),
    Mean_RMSE = if (all(ok)) {
      mean(folds$RMSE)
    } else {
      NA_real_
    },
    Mean_Correlation = if (all(ok)) {
      mean(folds$Correlation)
    } else {
      NA_real_
    },
    Mean_MAE = if (all(ok)) {
      mean(folds$MAE)
    } else {
      NA_real_
    },
    Median_EarlyStop_epoch = if (all(ok)) {
      median(folds$EarlyStop_epoch)
    } else {
      NA_real_
    },
    Max_Last_epoch = if (all(ok)) {
      max(folds$Last_epoch)
    } else {
      NA_real_
    },
    N_epoch_ceiling = if (all(ok)) {
      sum(folds$Hit_epoch_ceiling)
    } else {
      NA_integer_
    },
    Status = if (all(ok)) {
      "ok"
    } else {
      "incomplete"
    },
    stringsAsFactors = FALSE
  )

  list(
    summary = summary,
    folds = folds
  )
}

choose_candidate <- function(x) {
  eligible <- x[
    x$Status == "ok" &
      is.finite(x$Mean_RMSE),
    ,
    drop = FALSE
  ]

  if (!nrow(eligible)) {
    stop(
      "No complete ANN candidate is available."
    )
  }

  eligible$Parameters <- vapply(
    eligible$Architecture,
    function(architecture) {
      parameter_count(
        ncol(geno),
        ARCHITECTURES[[architecture]]
      )
    },
    numeric(1)
  )

  eligible$Penalty <-
    eligible$L1 +
    eligible$L2

  eligible <- eligible[
    order(
      eligible$Mean_RMSE,
      -eligible$Mean_Correlation,
      eligible$Parameters,
      eligible$Penalty,
      eligible$Learning_rate,
      eligible$CandidateID
    ),
    ,
    drop = FALSE
  ]

  # Epoch diagnostics are intentionally absent from the ranking above.
  eligible[
    1,
    ,
    drop = FALSE
  ]
}
```

# Outer-task table

The six scenarios share the same 25 frozen outer splits, producing 150
`Scenario × outer split` tuning tasks.

```{r tasks, echo=FALSE}
outer_map <- unique(
  cv$outer_assignments[
    ,
    c(
      "Repetition",
      "OuterFold",
      "SplitID"
    )
  ]
)

outer_map <- outer_map[
  order(
    outer_map$Repetition,
    outer_map$OuterFold
  ),
  ,
  drop = FALSE
]

task_table <- merge(
  data.frame(
    Scenario = scenario_names,
    stringsAsFactors = FALSE
  ),
  outer_map,
  by = NULL
)

task_table <- task_table[
  order(
    match(
      task_table$Scenario,
      scenario_names
    ),
    task_table$Repetition,
    task_table$OuterFold
  ),
  ,
  drop = FALSE
]

task_table$TaskID <- paste(
  task_table$Scenario,
  task_table$SplitID,
  sep = "_"
)

task_table$Fold <-
  task_table$OuterFold

rownames(task_table) <- NULL

if (nrow(task_table) != 150L) {
  stop(
    "Expected exactly 150 ANN tuning tasks."
  )
}

full_task_table <- task_table

if (ANN_TUNING_PILOT) {
  if (
    !ANN_TUNING_PILOT_TASK %in%
      full_task_table$TaskID
  ) {
    stop(
      paste(
        "Unknown ANN_TUNING_PILOT_TASK:",
        ANN_TUNING_PILOT_TASK
      )
    )
  }

  task_table <- full_task_table[
    full_task_table$TaskID ==
      ANN_TUNING_PILOT_TASK,
    ,
    drop = FALSE
  ]
}

EXPECTED_TASKS <- nrow(task_table)
EXPECTED_SELECTED_ROWS <-
  EXPECTED_TASKS *
  length(MODEL_LEVELS)
EXPECTED_CANDIDATE_ROWS <-
  EXPECTED_TASKS *
  CANDIDATES_PER_TASK
EXPECTED_FOLD_ROWS <-
  EXPECTED_TASKS *
  INNER_FITS_PER_TASK

cat(
  "Mode:",
  if (ANN_TUNING_PILOT) {
    "PILOT"
  } else {
    "FULL"
  },
  "\n"
)

cat(
  "Tasks:",
  EXPECTED_TASKS,
  "\n"
)

if (ANN_TUNING_PILOT) {
  cat(
    "Pilot task:",
    ANN_TUNING_PILOT_TASK,
    "\n"
  )
}

cat(
  "Configuration ID:",
  ANN_TUNING_CONFIG_ID,
  "\n"
)

cat(
  "Expected selected rows:",
  EXPECTED_SELECTED_ROWS,
  "\n"
)
cat(
  "Expected candidate rows:",
  EXPECTED_CANDIDATE_ROWS,
  "\n"
)
cat(
  "Expected fold rows:",
  EXPECTED_FOLD_ROWS,
  "\n"
)

knitr::kable(
  head(
    task_table,
    15
  ),
  caption = if (
    ANN_TUNING_PILOT
  ) {
    "Canonical ANN tuning pilot task"
  } else {
    "First 15 canonical ANN tuning tasks"
  }
)
```

# Canonical nested tuning

This section evaluates only inner-CV candidates. It never predicts the outer
assessment set.

```{r run-tuning, eval=RUN_ANN_TUNING, echo=FALSE}
run_start <-
  proc.time()[["elapsed"]]

task_outputs <- vector(
  "list",
  nrow(task_table)
)

for (
  task_index in seq_len(
    nrow(task_table)
  )
) {
  task <- task_table[
    task_index,
    ,
    drop = FALSE
  ]

  scenario <- task$Scenario
  scenario_index <- match(
    scenario,
    scenario_names
  )
  split_id <- task$SplitID

  checkpoint_file <- file.path(
    CHECKPOINT_DIR,
    paste0(
      task$TaskID,
      ".rds"
    )
  )

  if (
    file.exists(checkpoint_file) &&
    !ANN_TUNING_FORCE
  ) {
    checkpoint <-
      readRDS(checkpoint_file)

    if (
      identical(
        checkpoint$Status,
        "ok"
      ) &&
      identical(
        checkpoint$ConfigID,
        ANN_TUNING_CONFIG_ID
      )
    ) {
      task_outputs[[task_index]] <-
        checkpoint

      cat(
        sprintf(
          "[%03d/%03d] %s | checkpoint reused\n",
          task_index,
          nrow(task_table),
          task$TaskID
        )
      )

      next
    } else {
      cat(
        sprintf(
          "[%03d/%03d] %s | incompatible checkpoint ignored\n",
          task_index,
          nrow(task_table),
          task$TaskID
        )
      )
    }
  }

  task_start <-
    proc.time()[["elapsed"]]

  inner <- cv$inner_splits[[split_id]]

  if (length(inner) != 4L) {
    stop(
      "Expected four frozen inner folds."
    )
  }

  bundles <- vector(
    "list",
    4L
  )

  for (k in 1:4) {
    inner_split <- inner[[paste0("I", k)]]

    validation_seed <-
      SEED_GLOBAL +
      scenario_index * 1000000L +
      task$Repetition * 10000L +
      task$OuterFold * 1000L +
      k * 10L

    validation <- make_validation_split(
      inner_split$analysis_ids,
      EARLY_VALIDATION_FRACTION,
      validation_seed
    )

    bundles[[k]] <- prepare_bundle(
      validation$train,
      validation$validation,
      inner_split$assessment_ids,
      scenario_index
    )
  }

  base_seed <-
    SEED_GLOBAL +
    scenario_index * 10000000L +
    task$Repetition * 100000L +
    task$OuterFold * 10000L

  structure_summary_rows <- vector(
    "list",
    nrow(structure_grid)
  )
  structure_fold_rows <- vector(
    "list",
    nrow(structure_grid)
  )

  for (
    candidate_index in seq_len(
      nrow(structure_grid)
    )
  ) {
    candidate <-
      structure_grid[
        candidate_index,
        ,
        drop = FALSE
      ]

    evaluated <- evaluate_candidate(
      bundles = bundles,
      model_name = "ANN_0",
      architecture =
        candidate$Architecture,
      learning_rate =
        candidate$Learning_rate,
      l1 = 0,
      l2 = 0,
      seed_base =
        base_seed + 1000000L,
      candidate_id =
        candidate_index
    )

    structure_summary_rows[[candidate_index]] <-
      evaluated$summary

    structure_fold_rows[[candidate_index]] <-
      evaluated$folds
  }

  structure_summary <- do.call(
    rbind,
    structure_summary_rows
  )
  structure_folds <- do.call(
    rbind,
    structure_fold_rows
  )

  chosen_structure <-
    choose_candidate(
      structure_summary
    )

  selected_architecture <-
    chosen_structure$Architecture

  selected_learning_rate <-
    chosen_structure$Learning_rate

  penalty_grids <- list(
    ANN_L1 = l1_grid,
    ANN_L2 = l2_grid,
    ANN_L1_L2 = l1l2_grid
  )

  penalty_summaries <- list()
  penalty_folds <- list()
  penalty_selected <- list()

  for (
    model_index in seq_along(
      penalty_grids
    )
  ) {
    model_name <-
      names(penalty_grids)[
        model_index
      ]

    grid <-
      penalty_grids[[model_index]]

    candidate_summaries <- vector(
      "list",
      nrow(grid)
    )
    candidate_folds <- vector(
      "list",
      nrow(grid)
    )

    for (
      candidate_index in seq_len(
        nrow(grid)
      )
    ) {
      evaluated <- evaluate_candidate(
        bundles = bundles,
        model_name = model_name,
        architecture =
          selected_architecture,
        learning_rate =
          selected_learning_rate,
        l1 =
          grid$L1[candidate_index],
        l2 =
          grid$L2[candidate_index],
        seed_base =
          base_seed + 2000000L,
        candidate_id =
          candidate_index
      )

      candidate_summaries[[candidate_index]] <-
        evaluated$summary

      candidate_folds[[candidate_index]] <-
        evaluated$folds
    }

    penalty_summaries[[model_name]] <-
      do.call(
        rbind,
        candidate_summaries
      )

    penalty_folds[[model_name]] <-
      do.call(
        rbind,
        candidate_folds
      )

    penalty_selected[[model_name]] <-
      choose_candidate(
        penalty_summaries[[model_name]]
      )
  }

  selected_table <- rbind(
    data.frame(
      Model = "ANN_0",
      Architecture =
        selected_architecture,
      Learning_rate =
        selected_learning_rate,
      L1 = 0,
      L2 = 0,
      Inner_RMSE =
        chosen_structure$Mean_RMSE,
      Inner_Correlation =
        chosen_structure$Mean_Correlation,
      Inner_MAE =
        chosen_structure$Mean_MAE,
      stringsAsFactors = FALSE
    ),
    do.call(
      rbind,
      lapply(
        names(penalty_selected),
        function(model_name) {
          selected_candidate <-
            penalty_selected[[model_name]]

          data.frame(
            Model = model_name,
            Architecture =
              selected_architecture,
            Learning_rate =
              selected_learning_rate,
            L1 =
              selected_candidate$L1,
            L2 =
              selected_candidate$L2,
            Inner_RMSE =
              selected_candidate$Mean_RMSE,
            Inner_Correlation =
              selected_candidate$Mean_Correlation,
            Inner_MAE =
              selected_candidate$Mean_MAE,
            stringsAsFactors = FALSE
          )
        }
      )
    )
  )

  for (
    column_name in c(
      "Scenario",
      "Repetition",
      "Fold",
      "SplitID",
      "TaskID"
    )
  ) {
    selected_table[[column_name]] <-
      task[[column_name]]
  }

  tuning_summary <- rbind(
    structure_summary,
    do.call(
      rbind,
      penalty_summaries
    )
  )

  tuning_folds <- rbind(
    structure_folds,
    do.call(
      rbind,
      penalty_folds
    )
  )

  for (
    column_name in c(
      "Scenario",
      "Repetition",
      "Fold",
      "SplitID",
      "TaskID"
    )
  ) {
    tuning_summary[[column_name]] <-
      task[[column_name]]

    tuning_folds[[column_name]] <-
      task[[column_name]]
  }

  task_seconds <-
    proc.time()[["elapsed"]] -
    task_start

  checkpoint <- list(
    Status = "ok",
    ConfigID =
      ANN_TUNING_CONFIG_ID,
    TaskID = task$TaskID,
    selected = selected_table,
    tuning_summary = tuning_summary,
    tuning_folds = tuning_folds,
    task_seconds = task_seconds
  )

  saveRDS(
    checkpoint,
    checkpoint_file,
    compress = "gzip"
  )

  task_outputs[[task_index]] <-
    checkpoint

  cat(
    sprintf(
      "[%03d/%03d] %s | %.1f s | arch=%s lr=%g\n",
      task_index,
      nrow(task_table),
      task$TaskID,
      task_seconds,
      selected_architecture,
      selected_learning_rate
    )
  )

  gc()
}

current_session_seconds <-
  proc.time()[["elapsed"]] -
  run_start

selected_configurations <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(x) {
      x$selected
    }
  )
)

candidate_summary <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(x) {
      x$tuning_summary
    }
  )
)

fold_results <- do.call(
  rbind,
  lapply(
    task_outputs,
    function(x) {
      x$tuning_folds
    }
  )
)

checkpoint_seconds <- sum(
  vapply(
    task_outputs,
    function(x) {
      x$task_seconds
    },
    numeric(1)
  )
)

manifest <- list(
  analysis =
    "canonical_simulated_ann_tuning",
  dataset = "simulated",
  mode = if (
    ANN_TUNING_PILOT
  ) {
    "pilot"
  } else {
    "full"
  },
  pilot_task = if (
    ANN_TUNING_PILOT
  ) {
    ANN_TUNING_PILOT_TASK
  } else {
    NA_character_
  },
  config_id =
    ANN_TUNING_CONFIG_ID,
  seed_global = SEED_GLOBAL,
  models = MODEL_LEVELS,
  architectures = ARCHITECTURES,
  learning_rates =
    LEARNING_RATES,
  l1_values = L1_VALUES,
  l2_values = L2_VALUES,
  l1_l2_grid = L1_L2_GRID,
  optimizer = "Adam",
  activation = "ReLU",
  dropout = DROPOUT,
  max_epochs = MAX_EPOCHS,
  final_train_epochs =
    TUNING_V1_COMPATIBILITY_TOKEN,
  patience = PATIENCE,
  min_delta = MIN_DELTA,
  early_validation_fraction =
    EARLY_VALIDATION_FRACTION,
  primary_tuning_metric =
    "mean inner-assessment RMSE",
  secondary_tuning_metric =
    "mean inner-assessment correlation",
  epoch_used_in_candidate_ranking =
    FALSE,
  outer_assessment_used =
    FALSE,
  true_genetic_value_used =
    FALSE,
  architecture_lr_shared_across_models =
    TRUE,
  candidates_per_task =
    CANDIDATES_PER_TASK,
  inner_fits_per_task =
    INNER_FITS_PER_TASK,
  number_tasks =
    EXPECTED_TASKS,
  selected_rows =
    nrow(selected_configurations),
  candidate_rows =
    nrow(candidate_summary),
  fold_rows =
    nrow(fold_results),
  torch_version =
    as.character(
      packageVersion("torch")
    ),
  device = ANN_DEVICE_LABEL,
  cuda_available =
    cuda_available,
  current_session_seconds =
    current_session_seconds,
  checkpoint_task_seconds =
    checkpoint_seconds,
  timestamp =
    as.character(Sys.time())
)

manifest_table <- data.frame(
  Field = c(
    "analysis",
    "mode",
    "config_id",
    "architectures",
    "learning_rates",
    "L1_values",
    "L2_values",
    "combined_grid_size",
    "MAX_EPOCHS",
    "TUNING_V1_COMPATIBILITY_TOKEN",
    "patience",
    "dropout",
    "optimizer",
    "activation",
    "candidates_per_task",
    "inner_fits_per_task",
    "tasks",
    "epoch_used_in_candidate_ranking",
    "outer_assessment_used",
    "true_genetic_value_used"
  ),
  Value = c(
    manifest$analysis,
    manifest$mode,
    manifest$config_id,
    paste(
      names(ARCHITECTURES),
      collapse = " | "
    ),
    paste(
      LEARNING_RATES,
      collapse = " | "
    ),
    paste(
      L1_VALUES,
      collapse = " | "
    ),
    paste(
      L2_VALUES,
      collapse = " | "
    ),
    nrow(L1_L2_GRID),
    MAX_EPOCHS,
    TUNING_V1_COMPATIBILITY_TOKEN,
    PATIENCE,
    DROPOUT,
    "Adam",
    "ReLU",
    CANDIDATES_PER_TASK,
    INNER_FITS_PER_TASK,
    EXPECTED_TASKS,
    FALSE,
    FALSE,
    FALSE
  ),
  stringsAsFactors = FALSE
)

saveRDS(
  selected_configurations,
  SELECTED_RDS,
  compress = "gzip"
)
saveRDS(
  candidate_summary,
  CANDIDATE_RDS,
  compress = "gzip"
)
saveRDS(
  fold_results,
  FOLD_RDS,
  compress = "gzip"
)
saveRDS(
  manifest,
  RUN_RDS,
  compress = "gzip"
)
saveRDS(
  manifest,
  MANIFEST_RDS,
  compress = "gzip"
)

write.csv(
  selected_configurations,
  SELECTED_CSV,
  row.names = FALSE
)
write.csv(
  candidate_summary,
  CANDIDATE_CSV,
  row.names = FALSE
)
write.csv(
  fold_results,
  FOLD_CSV,
  row.names = FALSE
)
write.csv(
  manifest_table,
  MANIFEST_CSV,
  row.names = FALSE
)
```

# Saved-result integrity

The saved tuning outputs are checked against the canonical tuning contract
before selected configurations or diagnostics are displayed. Configuration ID,
row counts, finite metrics, exclusion of outer assessment/true genetic values,
and the shared architecture/learning-rate rule remain blocking integrity
conditions.

```{r saved-integrity, include=FALSE}
available <- all(
  file.exists(
    c(
      SELECTED_RDS,
      CANDIDATE_RDS,
      FOLD_RDS,
      RUN_RDS
    )
  )
)

if (!available) {
  cat(
    "Canonical ANN tuning outputs are unavailable; tuning summaries cannot be displayed.\n"
  )
} else {
  selected_configurations <-
    readRDS(SELECTED_RDS)
  candidate_summary <-
    readRDS(CANDIDATE_RDS)
  fold_results <-
    readRDS(FOLD_RDS)
  run_metadata <-
    readRDS(RUN_RDS)

  integrity <- data.frame(
    Check = c(
      "Configuration ID matches",
      "Selected rows",
      "Candidate rows",
      "Fold rows",
      "All candidate rows are ok",
      "All fold rows are ok",
      "All candidate RMSE values are finite",
      "All fold RMSE values are finite",
      "Epoch excluded from ranking",
      "Outer assessment excluded",
      "True genetic values excluded",
      "Architecture/LR shared within task"
    ),
    Status = c(
      if (
        identical(
          run_metadata$config_id,
          ANN_TUNING_CONFIG_ID
        )
      ) "PASS" else "BLOCKED",
      if (
        nrow(selected_configurations) ==
        EXPECTED_SELECTED_ROWS
      ) "PASS" else "BLOCKED",
      if (
        nrow(candidate_summary) ==
        EXPECTED_CANDIDATE_ROWS
      ) "PASS" else "BLOCKED",
      if (
        nrow(fold_results) ==
        EXPECTED_FOLD_ROWS
      ) "PASS" else "BLOCKED",
      if (
        all(
          candidate_summary$Status ==
            "ok"
        )
      ) "PASS" else "BLOCKED",
      if (
        all(
          fold_results$Status ==
            "ok"
        )
      ) "PASS" else "BLOCKED",
      if (
        all(
          is.finite(
            candidate_summary$Mean_RMSE
          )
        )
      ) "PASS" else "BLOCKED",
      if (
        all(
          is.finite(
            fold_results$RMSE
          )
        )
      ) "PASS" else "BLOCKED",
      if (
        identical(
          run_metadata$epoch_used_in_candidate_ranking,
          FALSE
        )
      ) "PASS" else "BLOCKED",
      if (
        identical(
          run_metadata$outer_assessment_used,
          FALSE
        )
      ) "PASS" else "BLOCKED",
      if (
        identical(
          run_metadata$true_genetic_value_used,
          FALSE
        )
      ) "PASS" else "BLOCKED",
      {
        task_structure <- aggregate(
          cbind(
            Architecture_code =
              as.numeric(
                factor(
                  selected_configurations$Architecture,
                  levels =
                    names(ARCHITECTURES)
                )
              ),
            Learning_rate =
              selected_configurations$Learning_rate
          ) ~ TaskID,
          data =
            selected_configurations,
          FUN = function(z) {
            length(unique(z))
          }
        )

        if (
          all(
            task_structure$Architecture_code ==
              1
          ) &&
          all(
            task_structure$Learning_rate ==
              1
          )
        ) {
          "PASS"
        } else {
          "BLOCKED"
        }
      }
    ),
    stringsAsFactors = FALSE
  )

  knitr::kable(
    integrity,
    caption =
      "Canonical ANN tuning saved-result integrity"
  )

  if (
    any(
      integrity$Status ==
        "BLOCKED"
    )
  ) {
    stop(
      "Canonical ANN tuning failed a blocking integrity check."
    )
  }
}
```

# Selected configurations

The summaries below show how often the 150 outer tasks selected each shared
architecture and learning rate, followed by the selected penalty values for the
regularized variants. A small configuration sample then illustrates the
task-level selections without reproducing all 600 rows.

```{r selected-summary, echo=FALSE}
if (available) {
  task_level <- unique(
    selected_configurations[
      ,
      c(
        "TaskID",
        "Architecture",
        "Learning_rate"
      )
    ]
  )

  architecture_frequency <- as.data.frame(
    table(
      Architecture = task_level$Architecture
    ),
    stringsAsFactors = FALSE
  )
  names(architecture_frequency)[2] <- "Selected_tasks"
  architecture_frequency <- architecture_frequency[
    order(
      match(
        architecture_frequency$Architecture,
        names(ARCHITECTURES)
      )
    ),
    ,
    drop = FALSE
  ]

  learning_rate_frequency <- as.data.frame(
    table(
      Learning_rate = task_level$Learning_rate
    ),
    stringsAsFactors = FALSE
  )
  names(learning_rate_frequency)[2] <- "Selected_tasks"

  l1_frequency <- selected_configurations[
    selected_configurations$Model %in%
      c("ANN_L1", "ANN_L1_L2"),
    c("Model", "L1"),
    drop = FALSE
  ]
  l1_frequency$Selected_tasks <- 1L
  l1_frequency <- aggregate(
    Selected_tasks ~ Model + L1,
    data = l1_frequency,
    FUN = sum
  )

  l2_frequency <- selected_configurations[
    selected_configurations$Model %in%
      c("ANN_L2", "ANN_L1_L2"),
    c("Model", "L2"),
    drop = FALSE
  ]
  l2_frequency$Selected_tasks <- 1L
  l2_frequency <- aggregate(
    Selected_tasks ~ Model + L2,
    data = l2_frequency,
    FUN = sum
  )

  knitr::kable(
    architecture_frequency,
    caption = "Selected shared architecture across the 150 outer tasks"
  )

  knitr::kable(
    learning_rate_frequency,
    caption = "Selected shared learning rate across the 150 outer tasks"
  )

  knitr::kable(
    l1_frequency,
    digits = 6,
    caption = "Selected L1 coefficients by regularized ANN variant"
  )

  knitr::kable(
    l2_frequency,
    digits = 6,
    caption = "Selected L2 coefficients by regularized ANN variant"
  )

  selected_display <- head(
    selected_configurations[
      order(
        selected_configurations$Scenario,
        selected_configurations$Repetition,
        selected_configurations$Fold,
        match(
          selected_configurations$Model,
          MODEL_LEVELS
        )
      ),
      c(
        "Scenario",
        "Repetition",
        "Fold",
        "Model",
        "Architecture",
        "Learning_rate",
        "L1",
        "L2",
        "Inner_RMSE",
        "Inner_Correlation"
      ),
      drop = FALSE
    ],
    20
  )

  knitr::kable(
    selected_display,
    digits = 6,
    caption = "First 20 selected canonical ANN configurations",
    table.attr = WIDE_TABLE_ATTR
  )
}
```

The selection frequencies summarize how often each architecture, learning rate, and penalty coefficient is preferred across the frozen outer tasks. Because architecture and learning rate are shared by the four ANN variants within each task, differences among `ANN_0`, `ANN_L1`, `ANN_L2`, and `ANN_L1_L2` are not produced by giving the regularized variants a different structural search space.

# Optimization diagnostics

These diagnostics describe how the fixed early-stopping protocol behaved over
the canonical tuning fits. They do not enter candidate ranking and do not
change the tuning grid.

```{r optimization-diagnostics, echo=FALSE}
if (available) {
  selected_keys <- merge(
    selected_configurations[
      ,
      c(
        "TaskID",
        "Model",
        "Architecture",
        "Learning_rate",
        "L1",
        "L2"
      )
    ],
    fold_results,
    by = c(
      "TaskID",
      "Model",
      "Architecture",
      "Learning_rate",
      "L1",
      "L2"
    ),
    all.x = TRUE,
    sort = FALSE
  )

  optimization_summary <- data.frame(
    Diagnostic = c(
      "Inner candidate-fold fits",
      "Median best early-stop epoch",
      "Median last epoch",
      "Fits stopped by patience",
      "Fits reaching 2000-epoch ceiling",
      "Selected-candidate inner folds",
      "Selected-candidate folds reaching ceiling"
    ),
    Value = c(
      as.character(nrow(fold_results)),
      sprintf("%.1f", median(fold_results$EarlyStop_epoch, na.rm = TRUE)),
      sprintf("%.1f", median(fold_results$Last_epoch, na.rm = TRUE)),
      sprintf(
        "%d / %d",
        sum(fold_results$Stopped_by_patience, na.rm = TRUE),
        nrow(fold_results)
      ),
      sprintf(
        "%d / %d",
        sum(fold_results$Hit_epoch_ceiling, na.rm = TRUE),
        nrow(fold_results)
      ),
      as.character(nrow(selected_keys)),
      sprintf(
        "%d / %d",
        sum(selected_keys$Hit_epoch_ceiling, na.rm = TRUE),
        nrow(selected_keys)
      )
    ),
    stringsAsFactors = FALSE
  )

  knitr::kable(
    optimization_summary,
    caption = "Canonical ANN tuning optimization diagnostics"
  )
}
```

# Boundary description

Boundary occupancy is reported descriptively because the upper architecture
`64-32` is a deliberate methodological cap. Selection of `64-32` indicates that
the bounded search space was active for those tasks, but it does not trigger
post hoc enlargement of the canonical grid.

Consequently, boundary occupancy is treated as a limitation of the prespecified
canonical comparison rather than as justification for extending the search
space after observing the results.

```{r boundary-description, echo=FALSE}
if (available) {
  task_level <- unique(
    selected_configurations[
      ,
      c(
        "TaskID",
        "Scenario",
        "Architecture",
        "Learning_rate"
      )
    ]
  )

  boundary_table <- data.frame(
    Boundary = c(
      "Architecture = 64-32",
      "Learning rate = 1.25e-4",
      "ANN_L1: L1 = 1e-1",
      "ANN_L1_L2: L1 = 1e-1",
      "ANN_L2: L2 = 1e-2",
      "ANN_L1_L2: L2 = 1e-2"
    ),
    Selected = c(
      sum(
        task_level$Architecture ==
          "64-32"
      ),
      sum(
        abs(
          task_level$Learning_rate -
            min(LEARNING_RATES)
        ) < 1e-14
      ),
      sum(
        selected_configurations$Model ==
          "ANN_L1" &
        abs(
          selected_configurations$L1 -
            max(L1_VALUES)
        ) < 1e-14
      ),
      sum(
        selected_configurations$Model ==
          "ANN_L1_L2" &
        abs(
          selected_configurations$L1 -
            max(L1_VALUES)
        ) < 1e-14
      ),
      sum(
        selected_configurations$Model ==
          "ANN_L2" &
        abs(
          selected_configurations$L2 -
            max(L2_VALUES)
        ) < 1e-14
      ),
      sum(
        selected_configurations$Model ==
          "ANN_L1_L2" &
        abs(
          selected_configurations$L2 -
            max(L2_VALUES)
        ) < 1e-14
      )
    ),
    Denominator = c(
      150L,
      150L,
      150L,
      150L,
      150L,
      150L
    ),
    stringsAsFactors = FALSE
  )

  boundary_table$Proportion <-
    boundary_table$Selected /
    boundary_table$Denominator

  knitr::kable(
    boundary_table,
    digits = 4,
    caption =
      "Descriptive canonical tuning-boundary occupancy"
  )
}
```

# From tuning to final fitting

Nested tuning produces one selected architecture, learning rate, and applicable penalty coefficient(s) for each `Scenario × outer split × ANN variant`. The final stage consumes those selections directly under the same frozen validation design; hyperparameter selection is not repeated.

# Final ANN fitting

The tuning stage uses a 2000-epoch convergence ceiling. Final epoch selection
uses a separate 3000-epoch safety ceiling so that patience-based stopping can
operate within the outer-analysis sample without imposing a fixed training
length. The 3000-epoch value is only an upper bound: each final network is
reinitialized and refit on all 800 outer-analysis individuals for exactly its
selected `Best_epoch`.

The final stage consumes the selected hyperparameters from nested tuning; it
does not repeat model selection.

For each `Scenario × outer split × ANN variant`, `Best_epoch` is selected with
early stopping using only the 800 outer-analysis individuals. The network is
then reinitialized, refit on all 800 for exactly `Best_epoch`, and only then
applied to the untouched 200 outer-assessment individuals.

```{r final-ann-contract, include=FALSE}
ANN_FORCE <- identical(Sys.getenv("ANN_FORCE", "0"), "1")
ANN_PILOT <- identical(Sys.getenv("ANN_PILOT", "0"), "1")
ANN_PILOT_TASK <- Sys.getenv("ANN_PILOT_TASK", "C1_R01_F01")

# Separate safety ceiling used only for Best_epoch selection within the
# outer-analysis sample; final networks are refit for exactly Best_epoch.
FINAL_EPOCH_MAX_EPOCHS <- 3000L

FINAL_CHECKPOINT_DIR <- project_path(".local", "checkpoints", "canonical", "ann_final")

FINAL_OUTPUT_STEM <- if (ANN_PILOT) {
  project_path(".local", "pilot", "ann", "final", "simulated_ann_pilot")
} else {
  project_path("output", "simulated", "ann", "final", "simulated_ann")
}

FINAL_RESULTS_RDS <- paste0(FINAL_OUTPUT_STEM, "_results.rds")
FINAL_PREDICTIONS_RDS <- paste0(FINAL_OUTPUT_STEM, "_predictions.rds")
FINAL_EPOCH_RDS <- paste0(FINAL_OUTPUT_STEM, "_epoch_selection.rds")
FINAL_RUN_RDS <- paste0(FINAL_OUTPUT_STEM, "_run_metadata.rds")

FINAL_RESULTS_CSV <- paste0(FINAL_OUTPUT_STEM, "_results.csv")
FINAL_PREDICTIONS_CSV <- paste0(FINAL_OUTPUT_STEM, "_predictions.csv")
FINAL_EPOCH_CSV <- paste0(FINAL_OUTPUT_STEM, "_epoch_selection.csv")

if (RUN_ANN) {
  dir.create(
    FINAL_CHECKPOINT_DIR,
    recursive = TRUE,
    showWarnings = FALSE
  )
  dir.create(
    dirname(FINAL_RESULTS_RDS),
    recursive = TRUE,
    showWarnings = FALSE
  )
}

FINAL_CONFIG_ID <- paste0(
  "ANN_FINAL_CANONICAL_V2__FROM__",
  ANN_TUNING_CONFIG_ID,
  "__FINAL_EPOCH_CAP_",
  FINAL_EPOCH_MAX_EPOCHS,
  "__BEST_EPOCH_REFIT"
)

knitr::kable(data.frame(
  Item=c("Tuning source","Epoch ceiling","Final epoch rule","Outer assessment in tuning","Outer assessment in epoch selection","True G in selection"),
  Value=c("ANN_TUNING_CANONICAL_V1",FINAL_EPOCH_MAX_EPOCHS,"Best_epoch selected inside outer-analysis; refit all 800",FALSE,FALSE,FALSE)
), caption="Canonical final ANN contract")
```

<!-- Source section: final-fit helper functions -->

```{r final-ann-functions, include=FALSE}
final_safe_bias <- function(p,o){z<-is.finite(p)&is.finite(o);if(!any(z))return(NA_real_);mean(p[z]-o[z])}
final_safe_slope <- function(p,o){z<-is.finite(p)&is.finite(o);if(sum(z)<3L||sd(p[z])==0)return(NA_real_);as.numeric(coef(lm(o[z]~p[z]))[2])}
final_safe_r2 <- function(p,o){z<-is.finite(p)&is.finite(o);if(sum(z)<2L)return(NA_real_);d<-sum((o[z]-mean(o[z]))^2);if(d==0)return(NA_real_);1-sum((o[z]-p[z])^2)/d}
final_metrics <- function(p,o,prefix){z<-c(safe_cor(p,o),safe_rmse(p,o),safe_mae(p,o),final_safe_bias(p,o),final_safe_slope(p,o),final_safe_r2(p,o));names(z)<-paste0(prefix,c("_Correlation","_RMSE","_MAE","_Bias","_Slope","_R2"));z}

select_final_epoch <- function(analysis_ids,scenario_index,architecture,learning_rate,l1,l2,split_seed,fit_seed){
  iv<-make_validation_split(analysis_ids,EARLY_VALIDATION_FRACTION,split_seed)
  fi<-match(iv$train,ids);vi<-match(iv$validation,ids);if(anyNA(c(fi,vi)))stop("Final epoch-selection ID alignment failed.")
  xs<-fit_x_scaler(geno[fi,,drop=FALSE]);ys<-fit_y_scaler(phenotype[fi,scenario_index])
  xf<-scale_x(geno[fi,,drop=FALSE],xs);xv<-scale_x(geno[vi,,drop=FALSE],xs);yf<-scale_y(phenotype[fi,scenario_index],ys);yv<-scale_y(phenotype[vi,scenario_index],ys)
  set_ann_seed(fit_seed)
  xt<-torch::torch_tensor(xf,dtype=torch::torch_float())$to(device=ANN_DEVICE);yt<-torch::torch_tensor(matrix(yf,ncol=1),dtype=torch::torch_float())$to(device=ANN_DEVICE)
  xv_t<-torch::torch_tensor(xv,dtype=torch::torch_float())$to(device=ANN_DEVICE);yv_t<-torch::torch_tensor(matrix(yv,ncol=1),dtype=torch::torch_float())$to(device=ANN_DEVICE)
  model<-TorchGenomicMLP(ncol(xf),ARCHITECTURES[[architecture]])$to(device=ANN_DEVICE);opt<-torch::optim_adam(params=model$parameters,lr=learning_rate,weight_decay=0)
  best<-Inf;best_epoch<-NA_integer_;best_state<-NULL;stale<-0L;last<-0L;stopped<-FALSE;sync_device();t0<-proc.time()[["elapsed"]]
  for(epoch in seq_len(FINAL_EPOCH_MAX_EPOCHS)){last<-epoch;model$train();opt$zero_grad();pr<-model(xt);loss<-torch::nnf_mse_loss(pr,yt)+weight_penalty(model,l1,l2);loss$backward();opt$step();model$eval();vl<-torch::with_no_grad({torch::nnf_mse_loss(model(xv_t),yv_t)});v<-as.numeric(vl$to(device=torch::torch_device("cpu")));if(is.finite(v)&&v<best-MIN_DELTA){best<-v;best_epoch<-epoch;best_state<-clone_state(model);stale<-0L}else stale<-stale+1L;if(stale>=PATIENCE){stopped<-TRUE;break}}
  sync_device();secs<-proc.time()[["elapsed"]]-t0;if(is.null(best_state)||is.na(best_epoch))stop("No finite final epoch-selection solution.")
  model$load_state_dict(best_state);model$eval();pv_s<-torch::with_no_grad({as.numeric(model(xv_t)$to(device=torch::torch_device("cpu")))});pv<-unscale_y(pv_s,ys);ov<-phenotype[vi,scenario_index]
  list(best_epoch=as.integer(best_epoch),last_epoch=as.integer(last),stopped_by_patience=stopped,hit_epoch_ceiling=last>=FINAL_EPOCH_MAX_EPOCHS&&!stopped,validation_rmse=safe_rmse(pv,ov),validation_correlation=safe_cor(pv,ov),fit_seconds=secs,n_fit=length(iv$train),n_validation=length(iv$validation))
}

train_final_fixed <- function(train_ids,assessment_ids,scenario_index,architecture,learning_rate,l1,l2,epochs,seed){
  tr<-match(train_ids,ids);te<-match(assessment_ids,ids);if(anyNA(c(tr,te)))stop("Final ANN ID alignment failed.")
  xs<-fit_x_scaler(geno[tr,,drop=FALSE]);ys<-fit_y_scaler(phenotype[tr,scenario_index]);xtr<-scale_x(geno[tr,,drop=FALSE],xs);xte<-scale_x(geno[te,,drop=FALSE],xs);ytr<-scale_y(phenotype[tr,scenario_index],ys)
  set_ann_seed(seed);xt<-torch::torch_tensor(xtr,dtype=torch::torch_float())$to(device=ANN_DEVICE);yt<-torch::torch_tensor(matrix(ytr,ncol=1),dtype=torch::torch_float())$to(device=ANN_DEVICE);xe<-torch::torch_tensor(xte,dtype=torch::torch_float())$to(device=ANN_DEVICE)
  model<-TorchGenomicMLP(ncol(xtr),ARCHITECTURES[[architecture]])$to(device=ANN_DEVICE);opt<-torch::optim_adam(params=model$parameters,lr=learning_rate,weight_decay=0);sync_device();t0<-proc.time()[["elapsed"]]
  for(epoch in seq_len(as.integer(epochs))){model$train();opt$zero_grad();pr<-model(xt);loss<-torch::nnf_mse_loss(pr,yt)+weight_penalty(model,l1,l2);loss$backward();opt$step()}
  sync_device();fit_seconds<-proc.time()[["elapsed"]]-t0;model$eval();sync_device();t1<-proc.time()[["elapsed"]];ps<-torch::with_no_grad({as.numeric(model(xe)$to(device=torch::torch_device("cpu")))});sync_device();pred_seconds<-proc.time()[["elapsed"]]-t1
  list(prediction=unscale_y(ps,ys),fit_seconds=fit_seconds,prediction_seconds=pred_seconds,zero_variance_markers=sum(xs$zero))
}
```

<!-- Source section: final outer-CV tasks -->

```{r final-ann-tasks, include=FALSE}
om<-unique(cv$outer_assignments[,c("Repetition","OuterFold","SplitID")]);om<-om[order(om$Repetition,om$OuterFold),]
final_task_table<-merge(data.frame(Scenario=scenario_names),om,by=NULL);final_task_table<-final_task_table[order(match(final_task_table$Scenario,scenario_names),final_task_table$Repetition,final_task_table$OuterFold),];final_task_table$Fold<-final_task_table$OuterFold;final_task_table$TaskID<-paste(final_task_table$Scenario,final_task_table$SplitID,sep="_");rownames(final_task_table)<-NULL
if(nrow(final_task_table)!=150L)stop("Expected 150 canonical ANN outer tasks.")
if(ANN_PILOT){if(!ANN_PILOT_TASK%in%final_task_table$TaskID)stop("Unknown ANN_PILOT_TASK: ",ANN_PILOT_TASK);final_task_table<-final_task_table[final_task_table$TaskID==ANN_PILOT_TASK,,drop=FALSE]}
knitr::kable(head(final_task_table,15),caption=if(ANN_PILOT)"Canonical ANN final pilot task"else"First 15 canonical ANN final tasks")
```

<!-- Source section: final fitting and outer prediction -->

```{r run-final-ann, eval=RUN_ANN, echo=FALSE}
req<-c(project_path("output","simulated","ann","tuning","simulated_ann_tuning_selected_configurations.rds"),project_path("output","simulated","ann","tuning","simulated_ann_tuning_run_metadata.rds"),project_path("output","simulated","ann","tuning","simulated_ann_tuning_manifest.rds"));if(!all(file.exists(req)))stop("Canonical tuning outputs are missing. Run RUN_ANN_TUNING=1 first. Missing: ",paste(req[!file.exists(req)],collapse=", "))
sel<-readRDS(req[1]);tm<-readRDS(req[2]);if(!identical(tm$config_id,ANN_TUNING_CONFIG_ID))stop("Saved tuning config_id is incompatible with canonical ANN page.");if(!identical(tm$outer_assessment_used,FALSE)||!identical(tm$true_genetic_value_used,FALSE))stop("Saved tuning provenance failed.");if(nrow(sel)!=600L)stop("Expected 600 canonical selected configurations.")
true_g<-common$true_genetic_value;control<-common$scenario_control;if(!identical(dim(true_g),c(1000L,6L)))stop("True G must be 1000 x 6.")
run_start<-proc.time()[["elapsed"]];outs<-vector("list",nrow(final_task_table))
for(ti in seq_len(nrow(final_task_table))){
  task<-final_task_table[ti,,drop=FALSE];sj<-match(task$Scenario,scenario_names);outer<-cv$outer_splits[[task$SplitID]];if(length(outer$analysis_ids)!=800L||length(outer$assessment_ids)!=200L||length(intersect(outer$analysis_ids,outer$assessment_ids)))stop("Outer split integrity failure: ",task$TaskID)
  st<-sel[sel$TaskID==task$TaskID,,drop=FALSE];if(nrow(st)!=4L||!setequal(st$Model,MODEL_LEVELS))stop("Expected one config/model: ",task$TaskID)
  cf<-file.path(FINAL_CHECKPOINT_DIR,paste0(task$TaskID,".rds"));if(file.exists(cf)&&!ANN_FORCE){cp<-readRDS(cf);if(identical(cp$Status,"ok")&&identical(cp$ConfigID,FINAL_CONFIG_ID)&&identical(cp$TuningConfigID,ANN_TUNING_CONFIG_ID)){outs[[ti]]<-cp;cat(sprintf("[%03d/%03d] %s | final checkpoint reused\n",ti,nrow(final_task_table),task$TaskID));next}}
  t0<-proc.time()[["elapsed"]];base<-SEED_GLOBAL+sj*10000000L+task$Repetition*100000L+task$OuterFold*10000L;rr<-pp<-ee<-vector("list",4L);split_seed<-base+8000000L;epoch_seed<-base+8100000L;fit_seed<-base+8200000L
  for(mi in seq_along(MODEL_LEVELS)){
    ml<-MODEL_LEVELS[mi];cfg<-st[st$Model==ml,,drop=FALSE];ef<-select_final_epoch(outer$analysis_ids,sj,cfg$Architecture,cfg$Learning_rate,cfg$L1,cfg$L2,split_seed,epoch_seed);ff<-train_final_fixed(outer$analysis_ids,outer$assessment_ids,sj,cfg$Architecture,cfg$Learning_rate,cfg$L1,cfg$L2,ef$best_epoch,fit_seed)
    ai<-match(outer$assessment_ids,ids);obs<-phenotype[ai,sj];gt<-true_g[ai,sj];pred<-ff$prediction;mg<-final_metrics(pred,gt,"Genetic");mp<-final_metrics(pred,obs,"Phenotype");meta<-control[control$Scenario==task$Scenario,,drop=FALSE]
    rr[[mi]]<-data.frame(Dataset="simulated",Scenario=task$Scenario,QTL_N=meta$QTL_N,H2_nominal=meta$H2_nominal,Repetition=task$Repetition,Fold=task$OuterFold,SplitID=task$SplitID,TaskID=task$TaskID,Model=ml,Architecture=cfg$Architecture,Learning_rate=cfg$Learning_rate,L1=cfg$L1,L2=cfg$L2,Best_epoch=ef$best_epoch,Epoch_Last=ef$last_epoch,Epoch_Stopped_by_patience=ef$stopped_by_patience,Epoch_Hit_ceiling=ef$hit_epoch_ceiling,Inner_RMSE=cfg$Inner_RMSE,Inner_Correlation=cfg$Inner_Correlation,Inner_MAE=cfg$Inner_MAE,N_train=800L,N_test=200L,N_pred_finite=sum(is.finite(pred)),N_zero_variance_markers=ff$zero_variance_markers,Genetic_Correlation=mg["Genetic_Correlation"],Genetic_RMSE=mg["Genetic_RMSE"],Genetic_MAE=mg["Genetic_MAE"],Genetic_Bias=mg["Genetic_Bias"],Genetic_Slope=mg["Genetic_Slope"],Genetic_R2=mg["Genetic_R2"],Phenotype_Correlation=mp["Phenotype_Correlation"],Phenotype_RMSE=mp["Phenotype_RMSE"],Phenotype_MAE=mp["Phenotype_MAE"],Phenotype_Bias=mp["Phenotype_Bias"],Phenotype_Slope=mp["Phenotype_Slope"],Phenotype_R2=mp["Phenotype_R2"],Epoch_selection_seconds=ef$fit_seconds,Final_fit_seconds=ff$fit_seconds,Predict_seconds=ff$prediction_seconds,Status=if(all(is.finite(pred)))"ok"else"prediction_nonfinite",Error=NA_character_)
    pp[[mi]]<-data.frame(Dataset="simulated",Scenario=task$Scenario,QTL_N=meta$QTL_N,H2_nominal=meta$H2_nominal,Repetition=task$Repetition,Fold=task$OuterFold,SplitID=task$SplitID,TaskID=task$TaskID,Model=ml,Architecture=cfg$Architecture,Learning_rate=cfg$Learning_rate,L1=cfg$L1,L2=cfg$L2,Best_epoch=ef$best_epoch,ID=ids[ai],Observed_Phenotype=obs,True_Genetic_Value=gt,Predicted_Value=pred,Status="ok")
    ee[[mi]]<-data.frame(Scenario=task$Scenario,Repetition=task$Repetition,Fold=task$OuterFold,SplitID=task$SplitID,TaskID=task$TaskID,Model=ml,Architecture=cfg$Architecture,Learning_rate=cfg$Learning_rate,L1=cfg$L1,L2=cfg$L2,Best_epoch=ef$best_epoch,Last_epoch=ef$last_epoch,Stopped_by_patience=ef$stopped_by_patience,Hit_epoch_ceiling=ef$hit_epoch_ceiling,Validation_RMSE=ef$validation_rmse,Validation_Correlation=ef$validation_correlation,N_epoch_fit=ef$n_fit,N_epoch_validation=ef$n_validation,Fit_seconds=ef$fit_seconds)
  }
  cp<-list(Status="ok",ConfigID=FINAL_CONFIG_ID,TuningConfigID=ANN_TUNING_CONFIG_ID,TaskID=task$TaskID,results=do.call(rbind,rr),predictions=do.call(rbind,pp),epoch_selection=do.call(rbind,ee),task_seconds=proc.time()[["elapsed"]]-t0);saveRDS(cp,cf,compress="gzip");outs[[ti]]<-cp;cat(sprintf("[%03d/%03d] %s | final %.1f min\n",ti,nrow(final_task_table),task$TaskID,cp$task_seconds/60));gc(verbose=FALSE)
}
results<-do.call(rbind,lapply(outs,`[[`,"results"));predictions<-do.call(rbind,lapply(outs,`[[`,"predictions"));epoch_selection<-do.call(rbind,lapply(outs,`[[`,"epoch_selection"));rownames(results)<-rownames(predictions)<-rownames(epoch_selection)<-NULL
run_seconds<-proc.time()[["elapsed"]]-run_start;checkpoint_seconds<-sum(vapply(outs,function(z)as.numeric(z$task_seconds),numeric(1)),na.rm=TRUE)
final_meta<-list(dataset="simulated",analysis="canonical_simulated_ann",mode=if(ANN_PILOT)"pilot"else"full",pilot_task=if(ANN_PILOT)ANN_PILOT_TASK else NA_character_,config_id=FINAL_CONFIG_ID,tuning_config_id=ANN_TUNING_CONFIG_ID,seed_global=SEED_GLOBAL,models=MODEL_LEVELS,tuning_max_epochs=MAX_EPOCHS,final_epoch_max_epochs=FINAL_EPOCH_MAX_EPOCHS,patience=PATIENCE,min_delta=MIN_DELTA,early_validation_fraction=EARLY_VALIDATION_FRACTION,final_refit_rule="Best_epoch selected inside outer-analysis, then refit on all 800",outer_assessment_used_in_tuning=FALSE,outer_assessment_used_in_epoch_selection=FALSE,true_genetic_value_used_in_selection=FALSE,number_tasks=nrow(final_task_table),model_task_rows=nrow(results),prediction_rows=nrow(predictions),epoch_selection_rows=nrow(epoch_selection),torch_version=as.character(packageVersion("torch")),device=ANN_DEVICE_LABEL,cuda_available=cuda_available,current_session_seconds=run_seconds,checkpoint_task_seconds=checkpoint_seconds,timestamp=as.character(Sys.time()))
saveRDS(results,FINAL_RESULTS_RDS,compress="gzip");saveRDS(predictions,FINAL_PREDICTIONS_RDS,compress="gzip");saveRDS(epoch_selection,FINAL_EPOCH_RDS,compress="gzip");saveRDS(final_meta,FINAL_RUN_RDS,compress="gzip");write.csv(results,FINAL_RESULTS_CSV,row.names=FALSE);write.csv(predictions,FINAL_PREDICTIONS_CSV,row.names=FALSE);write.csv(epoch_selection,FINAL_EPOCH_CSV,row.names=FALSE)
```

# Saved final-result integrity

Before final predictive summaries are shown, the 600 model-task rows, 120,000
held-out predictions, 600 epoch-selection rows, finite predictions, selected
epoch range, and leakage-prevention flags are checked internally against the
canonical final-fitting contract.

```{r final-ann-integrity, include=FALSE}
final_available<-all(file.exists(c(FINAL_RESULTS_RDS,FINAL_PREDICTIONS_RDS,FINAL_EPOCH_RDS,FINAL_RUN_RDS)))
if(!final_available){cat("Canonical final ANN outputs are unavailable; final summaries cannot be displayed.\n")}else{
  results<-readRDS(FINAL_RESULTS_RDS);predictions<-readRDS(FINAL_PREDICTIONS_RDS);epoch_selection<-readRDS(FINAL_EPOCH_RDS);final_meta<-readRDS(FINAL_RUN_RDS);nt<-if(identical(final_meta$mode,"pilot"))1L else 150L;nr<-nt*4L;np<-nr*200L
  audit<-data.frame(Check=c("Final config ID","Tuning config ID","Model-task rows","Prediction rows","Epoch-selection rows","4 models/task","200 predictions/model-task","All model tasks ok","All predictions finite","Best_epoch in range","Outer excluded from tuning","Outer excluded from epoch selection","True G excluded from selection"),Status=c(if(identical(final_meta$config_id,FINAL_CONFIG_ID))"PASS"else"BLOCKED",if(identical(final_meta$tuning_config_id,ANN_TUNING_CONFIG_ID))"PASS"else"BLOCKED",if(nrow(results)==nr)"PASS"else"BLOCKED",if(nrow(predictions)==np)"PASS"else"BLOCKED",if(nrow(epoch_selection)==nr)"PASS"else"BLOCKED",if(all(table(results$TaskID)==4L))"PASS"else"BLOCKED",if(all(table(interaction(predictions$TaskID,predictions$Model,drop=TRUE))==200L))"PASS"else"BLOCKED",if(all(results$Status=="ok"))"PASS"else"BLOCKED",if(all(is.finite(predictions$Predicted_Value)))"PASS"else"BLOCKED",if(all(epoch_selection$Best_epoch>=1L&epoch_selection$Best_epoch<=FINAL_EPOCH_MAX_EPOCHS))"PASS"else"BLOCKED",if(identical(final_meta$outer_assessment_used_in_tuning,FALSE))"PASS"else"BLOCKED",if(identical(final_meta$outer_assessment_used_in_epoch_selection,FALSE))"PASS"else"BLOCKED",if(identical(final_meta$true_genetic_value_used_in_selection,FALSE))"PASS"else"BLOCKED"))
  knitr::kable(audit,caption="Canonical final ANN saved-result integrity");if(any(audit$Status=="BLOCKED"))stop("Canonical final ANN failed integrity audit.");cat("\nFinal Best_epoch summary:\n");print(summary(epoch_selection$Best_epoch));cat("\nFinal epoch-selection fits hitting",FINAL_EPOCH_MAX_EPOCHS,"ceiling:",sum(epoch_selection$Hit_epoch_ceiling,na.rm=TRUE),"/",nrow(epoch_selection),"\n")
}
```

```{r final-epoch-interpretation, echo=FALSE, results="asis"}
if (final_available) {
  cat(sprintf(
    "Across the %d final epoch-selection fits, %d stopped by patience and %d reached the %d-epoch ceiling without a patience stop. Thus, the final ceiling acts as a safety bound rather than a fixed training length; each network is subsequently refit for its selected `Best_epoch`.\n",
    nrow(epoch_selection),
    sum(epoch_selection$Stopped_by_patience, na.rm = TRUE),
    sum(epoch_selection$Hit_epoch_ceiling, na.rm = TRUE),
    FINAL_EPOCH_MAX_EPOCHS
  ))
}
```

# Final ANN predictive performance

Final performance is summarized separately for the two evaluation targets.
Recovery of the known **true genetic value** is the primary simulated-data
criterion; prediction of the **observed phenotype** is secondary. Both tables
average the same 25 frozen outer splits for each scenario and ANN variant.

```{r final-ann-performance, echo=FALSE}
if (final_available) {
  ok <- results[
    results$Status == "ok",
    ,
    drop = FALSE
  ]

  perf <- aggregate(
    cbind(
      Genetic_Correlation,
      Genetic_RMSE,
      Genetic_MAE,
      Genetic_Bias,
      Genetic_Slope,
      Genetic_R2,
      Phenotype_Correlation,
      Phenotype_RMSE,
      Phenotype_MAE,
      Phenotype_Bias,
      Phenotype_Slope,
      Phenotype_R2
    ) ~ Scenario + Model + QTL_N + H2_nominal,
    data = ok,
    FUN = mean,
    na.rm = TRUE
  )

  perf <- perf[
    order(
      match(perf$Scenario, scenario_names),
      match(perf$Model, MODEL_LEVELS)
    ),
    ,
    drop = FALSE
  ]

  knitr::kable(
    perf[
      ,
      c(
        "Scenario",
        "Model",
        "QTL_N",
        "H2_nominal",
        "Genetic_Correlation",
        "Genetic_RMSE",
        "Genetic_MAE",
        "Genetic_Bias",
        "Genetic_Slope",
        "Genetic_R2"
      )
    ],
    digits = 4,
    caption = "Primary evaluation: mean outer-CV ANN performance against true genetic value",
    table.attr = WIDE_TABLE_ATTR
  )

  knitr::kable(
    perf[
      ,
      c(
        "Scenario",
        "Model",
        "QTL_N",
        "H2_nominal",
        "Phenotype_Correlation",
        "Phenotype_RMSE",
        "Phenotype_MAE",
        "Phenotype_Bias",
        "Phenotype_Slope",
        "Phenotype_R2"
      )
    ],
    digits = 4,
    caption = "Secondary evaluation: mean outer-CV ANN performance against observed phenotype",
    table.attr = WIDE_TABLE_ATTR
  )
}
```

```{r final-performance-interpretation, echo=FALSE, results="asis"}
if (final_available) {
  overall_ann <- aggregate(
    cbind(Genetic_Correlation, Genetic_RMSE, Genetic_R2) ~ Model,
    data = results[results$Status == "ok", , drop = FALSE],
    FUN = mean,
    na.rm = TRUE
  )
  z <- overall_ann[match(c("ANN_0", "ANN_L1", "ANN_L2", "ANN_L1_L2"), overall_ann$Model), ]
  cat(sprintf(
    "Across all outer-CV tasks, ANN-L1+L2 has the highest mean genetic correlation (%.4f) and lowest mean genetic RMSE (%.4f) among the four ANN variants. ANN-L1 is effectively tied (correlation %.4f; RMSE %.4f), whereas ANN-L2 remains close to the unregularized ANN (correlations %.4f and %.4f, respectively). Thus, within the prespecified canonical model/architecture search space, the observable regularization gain is driven mainly by the L1 component. The formal comparison with GBLUP-ADE is presented in the next module.\n",
    z$Genetic_Correlation[4], z$Genetic_RMSE[4],
    z$Genetic_Correlation[2], z$Genetic_RMSE[2],
    z$Genetic_Correlation[3], z$Genetic_Correlation[1]
  ))
}
```

# Navigation

[Previous: Validation design](simulated_data_validation.html) | [Parallel branch: GBLUP-ADE](simulated_data_gblup.html) | [Next: Model comparison](simulated_data_comparison.html)

# Reproducibility information

```{r session-info}
sessionInfo()
```
